---
title: "Fecundity process investigation"
author: "Niklas Neisse"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: hide
---

# Introduction

This report documents the fecundity process as it is currently implemented for
*Eriogonum longifolium*. It is intended as an exploratory description of the
current data structure and fitted fecundity components.

The report focuses on:

1. flowering probability,
2. number of scapes conditional on flowering,
3. the scapes-to-recruits transition,
4. recruit size distribution, and
5. how these components are combined in the IPM fecundity process.

Only the currently selected models are fitted here. No model selection is
repeated in this report.


# Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center')

library(MASS)
library(tidyverse)
library(lme4)

set.seed(100)
options(stringsAsFactors = FALSE)

v_head <- 'archbold'
v_species <- 'Eriogonum longifolium'
v_years_re <- c()

v_sp_abb <- tolower(
  gsub(
    ' ',
    '',
    paste(
      substr(
        unlist(strsplit(v_species, ' ')),
        1, 2),
      collapse = '')))

v_ggp_suffix <- paste(
  tools::toTitleCase(v_head),
  '-',
  v_species)

dir_data <- file.path(
  'C:', 'code', 'RUPDemo_IPMs',
  v_head,
  'data',
  v_sp_abb)

data_file <- file.path(
  dir_data,
  paste0(
    'ab_',
    v_sp_abb,
    '_df_workdata_260820.csv'))

if (!file.exists(data_file)) {
  stop(
    'Data file not found: ',
    data_file,
    '\nKnit this report from the project root or update data_file.')
}

ctrl_glmer <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(
    maxfun = 2e5))

ctrl_re <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(
    maxfun = 2e5))
```


# Data

The same working-data file and basic data preparation used in the current IPM
analysis are used here.

```{r data}
disturbance_case_levels <- c(
  'Undisturbed',
  'Current-year fire',
  'Previous-year fire',
  'Fire in both years')

df <- read.csv(
  data_file) %>%
  mutate(
    disturbance = as.numeric(
      dist_transition),
    disturbance_prev = as.numeric(
      disturbance_prev),
    year = as.integer(year),
    row_type = as.character(
      row_type),
    disturbance_case = factor(
      paste0(
        disturbance,
        disturbance_prev),
      levels = c(
        '00', '10', '01', '11'),
      labels = disturbance_case_levels)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))

df_ind <- df %>%
  filter(
    row_type == 'individual')

df_re_all <- df %>%
  filter(
    row_type == 'recruitment')

df_re <- df_re_all %>%
  filter(
    recruitment_complete,
    !is.na(recruits_simple))
```


# Fecundity as currently implemented

The current fecundity process has three fitted demographic components followed
by a recruit-size distribution:

- flowering probability is modeled at the individual-year level;
- scape number is modeled at the individual-year level conditional on flowering;
- scapes and recruits are summed to the site-population-year level for the
  scapes-to-recruits model;
- the fitted recruitment relationship is converted back to recruits per scape
  before entering the individual-level IPM fecundity calculation;
- recruits are distributed across recruit size using the observed seedling-size
  distribution.

Previous-year fire is used in flowering and scape production. Current-year fire
is used in the scapes-to-recruits component.


# Flowering probability

## Flowering data

```{r flowering-data}
df_fl <- df %>%
  filter(
    state == 'active',
    !is.na(flower),
    size_t0 > 0,
    is.finite(logsize_t0),
    is.finite(logsize_t0_2),
    is.finite(logsize_t0_3),
    !is.na(disturbance_prev)) %>%
  mutate(
    year = factor(year),
    disturbance_prev = as.numeric(
      disturbance_prev))
```


## Selected flowering model

The currently selected flowering model contains cubic size dependence,
previous-year fire, and a year random intercept.

```{r flowering-model}
mod_fl_bestfit <- glmer(
  flower ~
    logsize_t0 +
    logsize_t0_2 +
    logsize_t0_3 +
    disturbance_prev +
    (1 | year),
  data = df_fl,
  family = binomial,
  control = ctrl_glmer)

formula(mod_fl_bestfit)
summary(mod_fl_bestfit)
```


## Flowering data coverage

The first table shows individual-year records. The second removes repeated
individuals within a quadrat-year and therefore shows the disturbance coverage
at the quadrat-year level.

```{r flowering-coverage}
df_fl %>%
  count(
    disturbance_prev,
    name = 'n_individual_years') %>%
  as_tibble() %>%
  print(n = Inf)

df_fl %>%
  distinct(
    site,
    pop,
    qu,
    year,
    disturbance_prev) %>%
  count(
    disturbance_prev,
    name = 'n_quadrat_years') %>%
  as_tibble() %>%
  print(n = Inf)
```


## Mean flowering relationship

The mean relationship uses fixed effects only. The year random effect is
excluded with `re.form = NA`.

```{r flowering-mean, fig.width=7, fig.height=5}
df_fl_mean_pred <- tidyr::expand_grid(
  logsize_t0 = seq(
    min(
      df_fl$logsize_t0,
      na.rm = TRUE),
    max(
      df_fl$logsize_t0,
      na.rm = TRUE),
    length.out = 100),
  disturbance_prev = c(0, 1)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3,
    year = factor(
      levels(df_fl$year)[1],
      levels = levels(df_fl$year)))

df_fl_mean_pred <- df_fl_mean_pred %>%
  mutate(
    flower = predict(
      mod_fl_bestfit,
      newdata = df_fl_mean_pred,
      type = 'response',
      re.form = NA,
      allow.new.levels = TRUE))

df_fl_mean_binned <- df_fl %>%
  mutate(
    bin = cut(
      logsize_t0,
      breaks = 10)) %>%
  group_by(
    bin,
    disturbance_prev) %>%
  summarise(
    logsize_t0 = mean(
      logsize_t0,
      na.rm = TRUE),
    flower = mean(
      flower,
      na.rm = TRUE),
    n = n(),
    .groups = 'drop') %>%
  filter(
    !is.na(logsize_t0),
    n > 0)

fig_fl_mean <- ggplot() +
  geom_point(
    data = df_fl_mean_binned,
    aes(
      logsize_t0,
      flower,
      color = factor(
        disturbance_prev))) +
  geom_line(
    data = df_fl_mean_pred,
    aes(
      logsize_t0,
      flower,
      color = factor(
        disturbance_prev),
      linetype = factor(
        disturbance_prev)),
    linewidth = 1) +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  scale_linetype_manual(
    values = c(
      '0' = 'solid',
      '1' = 'dashed'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  labs(
    title = 'Mean flowering probability',
    subtitle = 'Fixed effects only - no year effect',
    x = expression(
      'log(diameter)'[t0]),
    y = 'Flowering probability') +
  coord_cartesian(
    ylim = c(0, 1)) +
  theme_bw()

fig_fl_mean
```


## Flowering relationships by year

The same model is shown for all years in one panel. The corresponding year
random effect is included with `re.form = NULL`.

```{r flowering-years, fig.width=11, fig.height=10}
df_fl_year_pred <- map_dfr(
  levels(df_fl$year),
  function(year_i) {
    df_i <- df_fl %>%
      filter(year == year_i)

    x <- seq(
      min(df_i$logsize_t0, na.rm = TRUE),
      max(df_i$logsize_t0, na.rm = TRUE),
      length.out = 100)

    pred_i <- tidyr::expand_grid(
  logsize_t0 = x,
  disturbance_prev = c(0, 1)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3,
    year = factor(
      year_i,
      levels = levels(df_fl$year)))

pred_i %>%
  mutate(
    flower = predict(
      mod_fl_bestfit,
      newdata = pred_i,
      type = 'response',
      re.form = NULL,
      allow.new.levels = TRUE))
  })

df_fl_year_obs <- df_fl %>%
  group_by(year, disturbance_prev) %>%
  mutate(
    bin = cut(
      logsize_t0,
      breaks = 8)) %>%
  group_by(
    year,
    disturbance_prev,
    bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    flower = mean(flower, na.rm = TRUE),
    n = n(),
    .groups = 'drop') %>%
  filter(
    !is.na(logsize_t0),
    n > 0)

fig_fl_year <- ggplot() +
  geom_point(
    data = df_fl_year_obs,
    aes(
      logsize_t0,
      flower,
      color = factor(disturbance_prev)),
    size = 1.2) +
  geom_line(
    data = df_fl_year_pred,
    aes(
      logsize_t0,
      flower,
      color = factor(disturbance_prev),
      linetype = factor(disturbance_prev)),
    linewidth = 0.7) +
  facet_wrap(
    ~ year,
    ncol = 5) +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  scale_linetype_manual(
    values = c(
      '0' = 'solid',
      '1' = 'dashed'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  coord_cartesian(
    ylim = c(0, 1)) +
  labs(
    title = 'Flowering probability by year',
    x = expression('log(diameter)'[t0]),
    y = 'Flowering probability') +
  theme_bw() +
  theme(
    legend.position = 'bottom')

fig_fl_year
```

# Number of scapes conditional on flowering

## Scape-number data

```{r scape-data}
df_fl_cond <- df_fl %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr == round(fl_nr),
    !is.na(disturbance_prev)) %>%
  mutate(
    year = factor(year))
```


## Selected scape-number model

The currently selected model contains cubic size dependence, previous-year
fire, and a year random intercept.

```{r scape-model}
mod_fl_n_bestfit <- glmer.nb(
  fl_nr ~
    logsize_t0 +
    logsize_t0_2 +
    logsize_t0_3 +
    disturbance_prev +
    (1 | year),
  data = df_fl_cond,
  control = ctrl_glmer)

formula(mod_fl_n_bestfit)
summary(mod_fl_n_bestfit)
```


## Scape-number data coverage

```{r scape-coverage}
df_fl_cond %>%
  count(
    disturbance_prev,
    name = 'n_flowering_individual_years') %>%
  as_tibble() %>%
  print(n = Inf)

df_fl_cond %>%
  distinct(
    site,
    pop,
    qu,
    year,
    disturbance_prev) %>%
  count(
    disturbance_prev,
    name = 'n_quadrat_years') %>%
  as_tibble() %>%
  print(n = Inf)
```


## Mean scape-number relationship

The mean relationship uses fixed effects only.

```{r scape-mean, fig.width=7, fig.height=5}
df_fl_n_mean_pred <- tidyr::expand_grid(
  logsize_t0 = seq(
    min(
      df_fl_cond$logsize_t0,
      na.rm = TRUE),
    max(
      df_fl_cond$logsize_t0,
      na.rm = TRUE),
    length.out = 100),
  disturbance_prev = c(0, 1)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3,
    year = factor(
      levels(df_fl_cond$year)[1],
      levels = levels(
        df_fl_cond$year)))

df_fl_n_mean_pred <- df_fl_n_mean_pred %>%
  mutate(
    fl_nr = predict(
      mod_fl_n_bestfit,
      newdata = df_fl_n_mean_pred,
      type = 'response',
      re.form = NA,
      allow.new.levels = TRUE))

df_fl_n_mean_binned <- df_fl_cond %>%
  mutate(
    bin = cut(
      logsize_t0,
      breaks = 10)) %>%
  group_by(
    bin,
    disturbance_prev) %>%
  summarise(
    logsize_t0 = mean(
      logsize_t0,
      na.rm = TRUE),
    fl_nr = mean(
      fl_nr,
      na.rm = TRUE),
    n = n(),
    .groups = 'drop') %>%
  filter(
    !is.na(logsize_t0),
    n > 0)

fig_fl_n_mean <- ggplot() +
  geom_point(
    data = df_fl_n_mean_binned,
    aes(
      logsize_t0,
      fl_nr,
      color = factor(
        disturbance_prev))) +
  geom_line(
    data = df_fl_n_mean_pred,
    aes(
      logsize_t0,
      fl_nr,
      color = factor(
        disturbance_prev),
      linetype = factor(
        disturbance_prev)),
    linewidth = 1) +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  scale_linetype_manual(
    values = c(
      '0' = 'solid',
      '1' = 'dashed'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  labs(
    title = 'Mean scape number conditional on flowering',
    subtitle = 'Fixed effects only - no year effect',
    x = expression(
      'log(diameter)'[t0]),
    y = 'Number of flowering scapes') +
  theme_bw()

fig_fl_n_mean
```


## Scape-number relationships by year

The selected scape-number model is shown for all years in one panel. The
corresponding year random effect is included.

```{r scape-years, fig.width=11, fig.height=10}
df_fl_n_year_pred <- map_dfr(
  levels(df_fl_cond$year),
  function(year_i) {
    df_i <- df_fl_cond %>%
      filter(year == year_i)

    x <- seq(
      min(df_i$logsize_t0, na.rm = TRUE),
      max(df_i$logsize_t0, na.rm = TRUE),
      length.out = 100)

    pred_i <- tidyr::expand_grid(
  logsize_t0 = x,
  disturbance_prev = c(0, 1)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3,
    year = factor(
      year_i,
      levels = levels(df_fl_cond$year)))

pred_i %>%
  mutate(
    fl_nr = predict(
      mod_fl_n_bestfit,
      newdata = pred_i,
      type = 'response',
      re.form = NULL,
      allow.new.levels = TRUE))
  })

df_fl_n_year_obs <- df_fl_cond %>%
  group_by(year, disturbance_prev) %>%
  mutate(
    bin = cut(
      logsize_t0,
      breaks = 8)) %>%
  group_by(
    year,
    disturbance_prev,
    bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    fl_nr = mean(fl_nr, na.rm = TRUE),
    n = n(),
    .groups = 'drop') %>%
  filter(
    !is.na(logsize_t0),
    n > 0)

fig_fl_n_year <- ggplot() +
  geom_point(
    data = df_fl_n_year_obs,
    aes(
      logsize_t0,
      fl_nr,
      color = factor(disturbance_prev)),
    size = 1.2) +
  geom_line(
    data = df_fl_n_year_pred,
    aes(
      logsize_t0,
      fl_nr,
      color = factor(disturbance_prev),
      linetype = factor(disturbance_prev)),
    linewidth = 0.7) +
  facet_wrap(
    ~ year,
    ncol = 5,
    scales = 'free_y') +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  scale_linetype_manual(
    values = c(
      '0' = 'solid',
      '1' = 'dashed'),
    labels = c(
      '0' = 'No previous fire',
      '1' = 'Previous-year fire'),
    name = 'Fire') +
  labs(
    title = 'Scape number conditional on flowering by year',
    x = expression('log(diameter)'[t0]),
    y = 'Number of flowering scapes') +
  theme_bw() +
  theme(
    legend.position = 'bottom')

fig_fl_n_year
```

# Scape bookkeeping

The recruitment rows contain `nr_scapes`, while individual plant records contain
`fl_nr`. The following check compares both quantities after individual scape
counts are summed to the quadrat-year level.

```{r scape-bookkeeping}
df_scapes_ind_quad <- df_ind %>%
  filter(
    !is.na(fl_nr)) %>%
  group_by(
    site,
    pop,
    qu,
    year) %>%
  summarise(
    n_individual_records = n(),
    scapes_from_individuals = sum(
      fl_nr,
      na.rm = TRUE),
    .groups = 'drop')

df_scapes_re_quad <- df_re %>%
  filter(
    !is.na(nr_scapes)) %>%
  group_by(
    site,
    pop,
    qu,
    year) %>%
  summarise(
    scapes_from_recruitment = sum(
      nr_scapes,
      na.rm = TRUE),
    recruits = sum(
      recruits_simple,
      na.rm = TRUE),
    .groups = 'drop')

df_scape_compare <- full_join(
  df_scapes_ind_quad,
  df_scapes_re_quad,
  by = c(
    'site',
    'pop',
    'qu',
    'year')) %>%
  mutate(
    difference =
      scapes_from_recruitment -
      scapes_from_individuals)

df_scape_compare %>%
  summarise(
    n_quadrat_years = n(),
    n_both_available = sum(
      !is.na(
        scapes_from_individuals) &
        !is.na(
          scapes_from_recruitment)),
    mean_individual_scapes = mean(
      scapes_from_individuals,
      na.rm = TRUE),
    mean_recruitment_scapes = mean(
      scapes_from_recruitment,
      na.rm = TRUE),
    mean_difference = mean(
      difference,
      na.rm = TRUE),
    max_abs_difference = max(
      abs(difference),
      na.rm = TRUE),
    correlation = cor(
      scapes_from_individuals,
      scapes_from_recruitment,
      use = 'complete.obs')) %>%
  as_tibble() %>%
  print(n = Inf)
```


# Scapes to recruits

## Recruitment data and aggregation

The current scapes-to-recruits model is fitted after aggregation to
site-population-year. Scapes and recruits are summed over recruitment quadrats.

A site-population-year is classified as disturbed when at least one contributing
quadrat has `disturbance == 1`. This is the current aggregation used in the IPM.

```{r recruitment-data}
df_sc2re <- df_re %>%
  filter(
    !is.na(nr_scapes),
    !is.na(recruits_simple)) %>%
  group_by(
    site,
    pop,
    year) %>%
  summarise(
    total_scapes = sum(
      nr_scapes),
    recruit_count = sum(
      recruits_simple),
    disturbance = case_when(
      any(
        disturbance == 1,
        na.rm = TRUE) ~ 1,
      any(
        is.na(disturbance)) ~ NA_real_,
      TRUE ~ 0),
    disturbance_prev = case_when(
      any(
        disturbance_prev == 1,
        na.rm = TRUE) ~ 1,
      any(
        is.na(disturbance_prev)) ~ NA_real_,
      TRUE ~ 0),
    .groups = 'drop')

df_sc2re_mod <- df_sc2re %>%
  filter(
    !is.na(total_scapes),
    !is.na(recruit_count),
    !is.na(disturbance),
    !(year %in% v_years_re)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(
      disturbance),
    disturbance_prev = as.numeric(
      disturbance_prev))
```


## Selected scapes-to-recruits model

The currently selected recruitment model uses log-transformed scape number,
current-year disturbance, their interaction, and a year random intercept.

```{r recruitment-model}
mod_re_bestfit <- glmer.nb(
  recruit_count ~
    log1p(total_scapes) *
    disturbance +
    (1 | year),
  data = df_sc2re_mod,
  control = ctrl_re)

formula(mod_re_bestfit)
summary(mod_re_bestfit)

re_year_sd <- if (
    'year' %in%
      names(
        VarCorr(
          mod_re_bestfit))) {
  attr(
    VarCorr(
      mod_re_bestfit)$year,
    'stddev')[1]
} else {
  0
}

tibble(
  singular = isSingular(
    mod_re_bestfit),
  year_intercept_sd = as.numeric(
    re_year_sd)) %>%
  print(n = Inf)
```


## Recruitment disturbance coverage

```{r recruitment-coverage}
df_re %>%
  count(
    disturbance,
    name = 'n_recruitment_quadrat_years') %>%
  as_tibble() %>%
  print(n = Inf)

df_sc2re_mod %>%
  count(
    disturbance,
    name = 'n_site_pop_years') %>%
  as_tibble() %>%
  print(n = Inf)
```

The first table describes recruitment disturbance information at the
quadrat-year level. The second shows how many observations remain after
aggregation to site-population-year.


## Mean scapes-to-recruits relationship

The mean relationship uses fixed effects only.

```{r recruitment-mean, fig.width=7, fig.height=5}
df_sc2re_mean_pred <- tidyr::expand_grid(
  total_scapes = seq(
    0,
    max(
      df_sc2re_mod$total_scapes,
      na.rm = TRUE),
    length.out = 100),
  disturbance = c(0, 1)) %>%
  mutate(
    year = factor(
      levels(
        df_sc2re_mod$year)[1],
      levels = levels(
        df_sc2re_mod$year)))

df_sc2re_mean_pred <- df_sc2re_mean_pred %>%
  mutate(
    recruit_count = predict(
      mod_re_bestfit,
      newdata = df_sc2re_mean_pred,
      type = 'response',
      re.form = NA,
      allow.new.levels = TRUE))

fig_sc2re_mean <- ggplot(
  df_sc2re_mod,
  aes(
    total_scapes,
    recruit_count,
    color = factor(
      disturbance))) +
  geom_jitter(
    height = 0.2,
    width = 0.5,
    alpha = 0.4) +
  geom_line(
    data = df_sc2re_mean_pred,
    aes(
      total_scapes,
      recruit_count,
      color = factor(
        disturbance),
      linetype = factor(
        disturbance)),
    linewidth = 1.2) +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No current fire',
      '1' = 'Current-year fire'),
    name = 'Fire') +
  scale_linetype_manual(
    values = c(
      '0' = 'solid',
      '1' = 'dashed'),
    labels = c(
      '0' = 'No current fire',
      '1' = 'Current-year fire'),
    name = 'Fire') +
  labs(
    title = 'Mean scapes-to-recruits relationship',
    subtitle = 'Fixed effects only - no year effect',
    x = expression(
      'Total flowering scapes '[t0]),
    y = expression(
      'Number of recruits '[t1])) +
  theme_bw()

fig_sc2re_mean
```


## Scapes-to-recruits relationships by year

The selected recruitment model is shown for all years in one panel. The
corresponding year random effect is included.

```{r recruitment-years, fig.width=11, fig.height=10}
df_re_year_pred <- map_dfr(
  levels(df_sc2re_mod$year),
  function(year_i) {
    df_i <- df_sc2re_mod %>%
      filter(year == year_i)

    x <- seq(
      0,
      max(df_i$total_scapes, na.rm = TRUE),
      length.out = 100)

    pred_i <- tidyr::expand_grid(
  total_scapes = x,
  disturbance = c(0, 1)) %>%
  mutate(
    year = factor(
      year_i,
      levels = levels(df_sc2re_mod$year)))

pred_i %>%
  mutate(
    recruit_count = predict(
      mod_re_bestfit,
      newdata = pred_i,
      type = 'response',
      re.form = NULL,
      allow.new.levels = TRUE))
  })

fig_re_year <- ggplot(
  df_sc2re_mod,
  aes(
    total_scapes,
    recruit_count,
    color = factor(disturbance))) +
  geom_point(
    alpha = 0.75,
    size = 1.3) +
  geom_line(
    data = df_re_year_pred,
    aes(
      total_scapes,
      recruit_count,
      color = factor(disturbance),
      linetype = factor(disturbance)),
    linewidth = 0.7) +
  facet_wrap(
    ~ year,
    ncol = 5,
    scales = 'free') +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No current fire',
      '1' = 'Current-year fire'),
    name = 'Fire') +
  scale_linetype_manual(
    values = c(
      '0' = 'solid',
      '1' = 'dashed'),
    labels = c(
      '0' = 'No current fire',
      '1' = 'Current-year fire'),
    name = 'Fire') +
  labs(
    title = 'Scapes-to-recruits relationship by year',
    x = expression('Total flowering scapes '[t0]),
    y = expression('Number of recruits '[t1])) +
  theme_bw() +
  theme(
    legend.position = 'bottom')

fig_re_year
```

# Recruitment aggregation checks

## Mixed-fire site-population-years

The following table identifies site-population-years in which both disturbed
and undisturbed quadrats contribute before aggregation.

```{r mixed-fire}
df_f2r_fire_trace <- df_re %>%
  filter(
    !is.na(nr_scapes),
    !is.na(recruits_simple)) %>%
  group_by(
    site,
    pop,
    year) %>%
  summarise(
    n_rows = n(),
    n_quadrats = n_distinct(qu),
    n_fire_0 = sum(
      disturbance == 0,
      na.rm = TRUE),
    n_fire_1 = sum(
      disturbance == 1,
      na.rm = TRUE),
    n_fire_na = sum(
      is.na(disturbance)),
    mixed_fire =
      n_fire_0 > 0 &
      n_fire_1 > 0,
    proportion_fire = if_else(
      n_fire_0 + n_fire_1 > 0,
      n_fire_1 /
        (n_fire_0 + n_fire_1),
      NA_real_),
    disturbance_used = case_when(
      any(
        disturbance == 1,
        na.rm = TRUE) ~ 1,
      any(
        is.na(disturbance)) ~ NA_real_,
      TRUE ~ 0),
    total_scapes = sum(
      nr_scapes,
      na.rm = TRUE),
    recruit_count = sum(
      recruits_simple,
      na.rm = TRUE),
    .groups = 'drop')

df_f2r_fire_trace %>%
  summarise(
    n_site_pop_years = n(),
    n_mixed_fire = sum(
      mixed_fire,
      na.rm = TRUE),
    p_mixed_fire = mean(
      mixed_fire,
      na.rm = TRUE),
    mean_prop_fire_when_disturbed = mean(
      proportion_fire[
        disturbance_used == 1],
      na.rm = TRUE),
    min_prop_fire_when_disturbed = min(
      proportion_fire[
        disturbance_used == 1],
      na.rm = TRUE),
    max_prop_fire_when_disturbed = max(
      proportion_fire[
        disturbance_used == 1],
      na.rm = TRUE)) %>%
  as_tibble() %>%
  print(n = Inf)

df_f2r_fire_trace %>%
  filter(
    mixed_fire) %>%
  as_tibble() %>%
  print(n = Inf)
```


## Quadrat-level scapes and recruits

This is an empirical check of the data before site-population-year aggregation.
It does not replace the current fitted model.

```{r quadrat-f2r}
df_f2r_quad <- df_re %>%
  filter(
    !is.na(nr_scapes),
    !is.na(recruits_simple)) %>%
  transmute(
    site,
    pop,
    qu,
    year,
    disturbance,
    disturbance_prev,
    scapes = nr_scapes,
    recruits = recruits_simple)

f2r_quad_summary <- df_f2r_quad %>%
  summarise(
    n_quadrat_years = n(),
    n_zero_scapes = sum(
      scapes == 0),
    p_zero_scapes = mean(
      scapes == 0),
    n_positive_recruits = sum(
      recruits > 0),
    p_positive_recruits = mean(
      recruits > 0),
    n_zero_scapes_positive_recruits = sum(
      scapes == 0 &
        recruits > 0),
    p_zero_scapes_positive_recruits = mean(
      scapes == 0 &
        recruits > 0))

f2r_quad_summary %>%
  as_tibble() %>%
  print(n = Inf)

df_f2r_quad %>%
  filter(
    !is.na(disturbance)) %>%
  group_by(
    disturbance) %>%
  summarise(
    n_quadrat_years = n(),
    total_scapes = sum(
      scapes,
      na.rm = TRUE),
    total_recruits = sum(
      recruits,
      na.rm = TRUE),
    recruits_per_scape =
      total_recruits /
      total_scapes,
    mean_recruits = mean(
      recruits,
      na.rm = TRUE),
    median_recruits = median(
      recruits,
      na.rm = TRUE),
    p_zero_scapes = mean(
      scapes == 0),
    p_zero_scapes_positive_recruits = mean(
      scapes == 0 &
        recruits > 0),
    .groups = 'drop') %>%
  as_tibble() %>%
  print(n = Inf)
```


## Quadrat-level empirical scapes-to-recruits plot

```{r quadrat-f2r-plot, fig.width=7, fig.height=5}
fig_f2r_quad <- ggplot(
  df_f2r_quad %>%
    filter(
      !is.na(disturbance)),
  aes(
    x = scapes,
    y = recruits,
    color = factor(
      disturbance))) +
  geom_point(
    alpha = 0.45) +
  scale_color_manual(
    values = c(
      '0' = 'black',
      '1' = 'red'),
    labels = c(
      '0' = 'No current fire',
      '1' = 'Current-year fire'),
    name = 'Fire') +
  theme_bw() +
  labs(
    title = 'Quadrat-level scapes to recruits',
    x = 'Number of scapes',
    y = 'Number of recruits')

fig_f2r_quad
```


# Conversion back to individual fecundity

The recruitment model is fitted at site-population-year level. The IPM,
however, requires expected recruitment from an individual plant.

The current implementation therefore predicts recruitment across the common
observed distribution of site-population-year scape totals and divides the sum
of predicted recruits by the total number of observed scapes. This produces one
mean recruits-per-scape conversion for each current-fire state.

```{r recruit-conversion}
get_recr_per_scape <- function(
    disturbance_i,
    year_i = NULL) {

  df_i <- df_sc2re_mod %>%
    mutate(
      disturbance =
        disturbance_i)

  if (is.null(year_i)) {
    re_form_i <- NA
  } else if (
      as.character(year_i) %in%
        levels(df_sc2re_mod$year)) {

    df_i <- df_i %>%
      mutate(
        year = factor(
          as.character(year_i),
          levels = levels(
            df_sc2re_mod$year)))

    re_form_i <- NULL
  } else {
    re_form_i <- NA
  }

  pred_i <- predict(
    mod_re_bestfit,
    newdata = df_i,
    type = 'response',
    re.form = re_form_i,
    allow.new.levels = TRUE)

  total_scapes_i <- sum(
    df_i$total_scapes,
    na.rm = TRUE)

  if (
      !is.finite(
        total_scapes_i) ||
      total_scapes_i <= 0) {
    return(0)
  }

  sum(
    pred_i,
    na.rm = TRUE) /
    total_scapes_i
}

recr_per_scape_0 <-
  get_recr_per_scape(0)

recr_per_scape_1 <-
  get_recr_per_scape(1)

tibble(
  disturbance = c(
    'No current fire',
    'Current-year fire'),
  recruits_per_scape = c(
    recr_per_scape_0,
    recr_per_scape_1)) %>%
  print(n = Inf)
```

For an individual of size \(x\), this conversion is combined with flowering
probability and expected scape number:

\[
E(R \mid x) =
P(\mathrm{flower}\mid x)
\times
E(\mathrm{scapes}\mid \mathrm{flower},x)
\times
E(\mathrm{recruits\ per\ scape}).
\]

Thus, the site-population-year recruitment model is brought back to the
individual fecundity calculation through the recruits-per-scape term.


# Recruit size distribution

Recruit size is taken from individual records classified as seedlings.

```{r recruit-size-data}
df_re_size <- df_ind %>%
  filter(
    recruit_type == 'seedling',
    size_t0 > 0,
    is.finite(logsize_t0))

recr_sz <- mean(
  df_re_size$logsize_t0,
  na.rm = TRUE)

recr_sd <- sd(
  df_re_size$logsize_t0,
  na.rm = TRUE)

recr_min <- min(
  df_re_size$logsize_t0,
  na.rm = TRUE)

recr_max <- max(
  df_re_size$logsize_t0,
  na.rm = TRUE)

normal_mass_observed_range <-
  pnorm(
    recr_max,
    mean = recr_sz,
    sd = recr_sd) -
  pnorm(
    recr_min,
    mean = recr_sz,
    sd = recr_sd)

tibble(
  n_seedlings = nrow(
    df_re_size),
  mean_logsize = recr_sz,
  sd_logsize = recr_sd,
  minimum_logsize = recr_min,
  maximum_logsize = recr_max,
  fitted_normal_mass_inside_observed_range =
    normal_mass_observed_range) %>%
  print(n = Inf)
```


## Observed recruit-size distribution

The histogram shows the observed seedling-size distribution. The line is the
normal distribution defined by the observed mean and standard deviation, which
is the distributional form used for recruits in the IPM.

```{r recruit-size-plot, fig.width=7, fig.height=5}
df_re_size_density <- tibble(
  logsize_t0 = seq(
    recr_min,
    recr_max,
    length.out = 300)) %>%
  mutate(
    density = dnorm(
      logsize_t0,
      mean = recr_sz,
      sd = recr_sd))

fig_re_size <- ggplot(
  df_re_size,
  aes(
    logsize_t0)) +
  geom_histogram(
    aes(
      y = after_stat(
        density)),
    bins = 25,
    fill = 'grey80',
    color = 'black') +
  geom_line(
    data = df_re_size_density,
    aes(
      logsize_t0,
      density),
    linewidth = 1) +
  theme_bw() +
  labs(
    title = 'Recruit size distribution',
    subtitle = 'Observed seedlings and fitted normal distribution',
    x = expression(
      'log(diameter)'),
    y = 'Density')

fig_re_size
```


# Fecundity bookkeeping summary

The current fecundity process can therefore be summarized as follows.

```{r fecundity-summary}
fecundity_bookkeeping <- tibble(
  component = c(
    'Flowering probability',
    'Scapes conditional on flowering',
    'Scapes to recruits',
    'Recruit size'),
  fitted_or_estimated_level = c(
    'Individual-year',
    'Individual-year',
    'Site-population-year',
    'Individual seedlings'),
  disturbance = c(
    'Previous-year fire',
    'Previous-year fire',
    'Current-year fire',
    'None'),
  IPM_quantity = c(
    'Probability of flowering',
    'Expected scapes conditional on flowering',
    'Recruits per scape',
    'Recruit size density'))

fecundity_bookkeeping %>%
  print(n = Inf)
```

The fecundity contribution of an individual of size \(x\) to recruits of size
\(y\) is represented by:

\[
F(y,x) =
P(\mathrm{flower}\mid x)
\times
E(\mathrm{scapes}\mid \mathrm{flower},x)
\times
E(\mathrm{recruits\ per\ scape})
\times
f_{\mathrm{recruit\ size}}(y).
\]

This section describes the current implementation only. The preceding tables
and figures document the observational levels, disturbance coverage, fitted
mean relationships, yearly relationships, recruitment aggregation, and recruit
size distribution used in that implementation.
