# Updated data {#data_debugged}

This chapter documents the updated *Polygala lewtonii* dataset and the
construction of the revised working data. The new dataset is not simply an
extension of the data described in Chapter \@ref(data). It adds observations
from more recent years, reorganizes the original data structure, and provides
additional information at the individual, quadrat, and site levels.
Consequently, the complete data-processing workflow was reviewed and updated.

The main aim of this chapter is to describe the newly provided files, compare
them with the data used in Chapter \@ref(data), document the decisions made
during data cleaning, and explain how the final working dataset was produced.

## Overview

The updated data were provided as an Excel workbook containing four principal
worksheets: `Demog_longform`, `Demog`, `QuadLevel`, and `SiteHistory`. Each
worksheet is accompanied by separate metadata describing its variables and
coding.

The demographic records now extend through 2025. In addition to the longer
study period, the updated workbook separates repeated demographic
observations, quadrat-level information, and site-level management history
more clearly than the dataset used in Chapter \@ref(data). These structural
changes required us to reassess plant identity, annual transitions,
recruitment, mortality, growth, and disturbance exposure before constructing
the revised working dataset.


```{r 8_setup}
library(dplyr)

file_workbook <- paste0(
  "C:/code/RUPDemo_IPMs/archbold/data/pole/",
  "Pol_lew_CCS_2025_complete.xlsx")

df_workbook_sheets <- tibble(
  sheet = readxl::excel_sheets(file_workbook)) %>%
  mutate(
    type = if_else(
      str_detect(sheet, "metadata$"),
      "Metadata", "Data"),
    dataset = str_remove(sheet, " metadata$")) %>%
  dplyr::select(dataset, type, sheet)

knitr::kable(
  df_workbook_sheets,
  col.names = c("Dataset", "Worksheet type", "Worksheet name"))
```

`Demog_longform` contains the repeated demographic censuses,
whereas `Demog` provides the demographic records in a wider format.
`QuadLevel` contains information recorded at the quadrat level, and
`SiteHistory` documents disturbance and management history at the site level.

```{r data-debugged-dataset-dimensions}
data_sheets <- c(
  "Demog_longform", "Demog", "QuadLevel", "SiteHistory")

df_dataset_dimensions <- map_dfr(data_sheets, function(sheet_i) {
  data_i <- readxl::read_excel(file_workbook, sheet = sheet_i)

  tibble(
    dataset = sheet_i,
    rows = nrow(data_i),
    columns = ncol(data_i))
})

knitr::kable(
  df_dataset_dimensions,
  col.names = c("Dataset", "Rows", "Columns"))
```

```{r data-debugged-read-sheets, include=FALSE}
df_demog_long <- readxl::read_excel(
  file_workbook,
  sheet = "Demog_longform",
  na = c("", "NA")) %>%
  janitor::clean_names()

df_demog_wide <- readxl::read_excel(
  file_workbook,
  sheet = "Demog",
  na = c("", "NA")) %>%
  janitor::clean_names()

df_quad <- readxl::read_excel(
  file_workbook,
  sheet = "QuadLevel",
  na = c("", "NA")) %>%
  janitor::clean_names()

df_site <- readxl::read_excel(
  file_workbook,
  sheet = "SiteHistory",
  col_types = c("numeric", "text", "text", "text")) %>%
  janitor::clean_names() %>%
  mutate(
    date = case_when(
      is.na(date) ~ NA_character_,
      str_detect(date, "^\\d+$") ~ format(
        as.Date(
          as.numeric(date),
          origin = "1899-12-30"),
        "%m/%d/%Y"),
      TRUE ~ date))
```


## Data worksheets


### Demog_longform

`Demog_longform` is the main demographic dataset used in this analysis. It
contains 43,869 observations and 12 variables. Each row represents one
individual plant during one quarterly census. Plants are identified by a
unique individual ID and linked to their quadrat and patch.

Survival was recorded during the quarterly censuses in March, June, September,
and December. Stage, height, maximum crown diameter, stem number, and flowering
stem number were generally recorded during the annual March census. The first
annual census was carried out in April 2001.

Compared with the dataset used in Chapter \@ref(data), the updated long-format
dataset extends the demographic record through 2025 and already contains a
consistent `unique_ID`.

```{r data-debugged-show-demog-long}
df_demog_long %>%
  slice_head(n = 10) %>%
  knitr::kable()
```

### Demog

`Demog` contains 7,424 rows and 309 columns. It stores the demographic history
in wide format, with one row per individual and separate columns for each
census date. For example, survival, stage, height, crown diameter, stem number,
and flowering stem number are stored in variables such as `surv0325`,
`stg0325`, and `ht0325`.

The two demographic worksheets therefore contain much of the same underlying
information but organize it differently. `Demog`
retains additional information such as cohort, plant position, comments,
annual-survival variables, and historical post-burn measurements.


### QuadLevel

`QuadLevel` contains one row for each of the 220 permanent quadrats. It includes
the quadrat coordinates, treatment block, patch, monitoring level, recent
plant counts, flowering-stem counts, stem counts, occupancy, and measurements
of the 2022 prescribed burn.

This worksheet is particularly important for understanding the change in
sampling after June 2022. Some quadrats continued under Level 3 individual
demographic monitoring, whereas the remaining quadrats were reduced to Level 2
count-based monitoring. The Level 2 observations are stored in this worksheet
rather than in the individual demographic dataset.


```{r data-debugged-show-quad}
df_quad %>%
  dplyr::select(
    patch, quad, lvl3, n2025, flst0325, stm0325,
    n2024, flst0324, stm0324, n2023,
    postburn_status0322, postburn_pct0322) %>%
  dplyr::slice_head(n = 10) %>%
  knitr::kable()
```

### SiteHistory

`SiteHistory` provides a chronological record of major natural disturbances
and management events at the study site. The worksheet includes prescribed
burns in 2001, 2007, 2009, 2015, 2019, 2020, and 2022, several hurricanes in
2004, Hurricane Irma in 2017, and hardwood reduction in 2023.

This is an important addition to the data described in Chapter \@ref(data),
which only included the four prescribed burns up to 2015. Fire is central to
the demography of *P. lewtonii*: established plants are commonly killed by
fire, while large seedling cohorts can recruit from the persistent soil
seedbank after burning. Other disturbances, including hurricanes and hardwood
removal, may also alter survival, recruitment, vegetation cover, and the
availability of open habitat. Disturbance history is therefore treated as a
central part of the demographic data rather than only as background site
information.

Note to the data provider: The date recorded for the 2015 prescribed burn incorrectly gives the year 2025. This is a known typographical error, and the event is treated as a 2015
burn throughout the analysis.

```{r data-debugged-show-site-history}
(df_site) %>% print()
```

## Differences from the previous dataset

The updated dataset differs from the data used in Chapter \@ref(data) in both
its time span and structure. The earlier dataset covered the monitoring period
from 2001 to 2017, whereas the updated demographic records extend through
2025.

```{r data-debugged-differences}
df_data_differences <- tribble(
  ~Aspect, ~`Previous dataset`, ~`Updated dataset`,
  "Study period",
  "2001--2017",
  "2001--2025",
  "Plant identity",
  paste(
    "Identity reconstructed from management unit, quadrat, plant tag,",
    "angle, and distance"),
  "A unique individual identifier is supplied",
  "Demographic structure",
  "Quarterly observations stored in one main table",
  paste(
    "Quarterly long-format data and individual-level wide-format data",
    "are supplied separately"),
  "Spatial information",
  "Stored together with the demographic observations",
  paste(
    "Additional quadrat-level information is supplied in",
    "`QuadLevel`"),
  "Disturbance history",
  "Four prescribed burns through 2015",
  paste(
    "Prescribed burns through 2022 and additional disturbance history",
    "are supplied"),
  "Monitoring design",
  "Individual demographic monitoring across the permanent quadrats",
  paste(
    "Level 3 individual monitoring and Level 2 count monitoring are",
    "distinguished after 2022"))

knitr::kable(df_data_differences)
```

The supplied unique individual identifier is an improvement. In the
previous analysis, plant tags were not always unique through time and plant
identity therefore had to be reconstructed from the tag and its position
within the quadrat. The updated `unique_ID` variable removes this step and is
used as the individual identifier throughout the revised data preparation.

The updated workbook also separates information that was previously combined
within the demographic records. Repeated demographic observations are taken
from `Demog_longform`, while additional individual, quadrat, and disturbance
information can be obtained from the other worksheets. This provides more
information, but also requires the different spatial and temporal levels of
the data to be handled explicitly.

The revised working-data procedure was therefore rebuilt rather than simply
applied to the additional years. In particular, recruitment timing, annual
survival, growth transitions, and disturbance exposure were checked again
against the quarterly census structure.


## Construction of the annual demographic dataset

The demographic data contain quarterly observations, but the IPMs describe
transitions between annual censuses. The annual dataset was therefore created
by retaining records with a recorded stage. Stage, plant size, stem number, and
flowering stem number were normally measured during the March census. The
initial annual census was conducted in April 2001.

Annual census records were identified by the presence of a stage
classification. Unlike filtering by height or crown diameter, this approach
also retains individuals with incomplete size measurements, including
confirmed recruits that could not be assigned a volume.

```{r data-debugged-annual-data}
df_annual <- df_demog_long %>%
  rename(id = unique_id) %>%
  filter(!is.na(stg)) %>%
  arrange(id, year, month)

n_annual_duplicates <- df_annual %>%
  count(id, year) %>%
  filter(n > 1) %>%
  nrow()

df_annual_summary <- df_annual %>%
  summarise(
    records = n(),
    individuals = n_distinct(id),
    quadrats = n_distinct(quad),
    sites = n_distinct(patch),
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE))

knitr::kable(
  df_annual_summary,
  col.names = c(
    "Annual records", "Individuals", "Quadrats", "Sites",
    "First year", "Last year"))
```

Each row in the resulting dataset represents one individual at one annual
census. No duplicated individual-year records were found
(`r n_annual_duplicates` cases).

```{r data-debugged-annual-census-timing}
df_census_timing <- df_annual %>%
  group_by(month) %>%
  summarise(
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE),
    years = n_distinct(year),
    records = n(),
    .groups = "drop")

knitr::kable(
  df_census_timing,
  col.names = c(
    "Census month", "First year", "Last year",
    "Number of years", "Annual records"))
```

The April records belong exclusively to the first census in 2001. All
subsequent annual records were collected in March. The annual transition
represented by a row therefore generally begins in March of year \(t\) and
ends in March of year \(t+1\). The first transition begins in April 2001 and
ends in March 2002.


## Recruitment

Recruitment was identified from the quarterly survival code `qsurv == 5`,
which records the first observation of an individual as a seedling. Recruitment
was assigned to the annual census at the end of the corresponding census
interval.

Seedlings recorded during March were assigned to that census year. Seedlings
first recorded from June to December were assigned to the following annual
census. Recruitment in 2001 was left as unknown because the first census
represents the beginning of the study rather than a complete annual
recruitment interval.

```{r data-debugged-recruitment}
df_recruit <- df_demog_long %>%
  filter(qsurv == 5) %>%
  transmute(
    id = unique_id,
    recruit_year = case_when(
      year == 2001 & month == 4 ~ 2001,
      month > 3 ~ year + 1,
      TRUE ~ year)) %>%
  distinct(id, recruit_year) %>%
  mutate(recruit_qsurv = 1)

df_annual_recruit <- df_annual %>%
  left_join(
    df_recruit,
    by = c("id", "year" = "recruit_year")) %>%
  mutate(
    recruit = case_when(
      year == 2001 ~ NA_real_,
      recruit_qsurv == 1 ~ 1,
      TRUE ~ 0))

df_recruit_summary <- df_annual_recruit %>%
  summarise(
    confirmed_recruits = sum(recruit == 1, na.rm = TRUE),
    non_recruits = sum(recruit == 0, na.rm = TRUE),
    baseline_unknown = sum(is.na(recruit)))

knitr::kable(
  df_recruit_summary,
  col.names = c(
    "Confirmed recruits", "Non-recruits",
    "Unknown baseline records"))
```

The previous preparation combined three criteria: the immediately preceding
quarterly survival code, the current quarterly survival code, and annual stage
1. This could miss seedlings first observed earlier in the annual interval if
later quarterly observations occurred before the annual census.

The revised method searches the complete interval for a confirmed seedling
record and does not depend on the immediately preceding observation. Annual
stage is not used to define recruitment. Individuals recorded as new adults
(`qsurv == 3`) are also not classified as recruits because their seedling
status during that year cannot be confirmed. :contentReference[oaicite:0]{index=0}



## Survival and growth transitions

Survival and growth were defined over the interval between consecutive annual
censuses. All recorded deaths (`qsurv == 0`) occurring after the census in year
\(t\) and no later than the census in year \(t+1\) were assigned to that annual
transition.

A recorded death was treated as direct evidence of mortality during the annual
transition. In a small number of cases, an individual was coded as dead during
one quarterly census but was recorded again at a later quarterly census within
the same annual interval. However, the individual was absent from the next
full annual census and did not appear in subsequent annual records. The later
quarterly observation was therefore not interpreted as survival to \(t+1\).

The annual transition containing the death was classified as
`survives = 0`. Any later quarterly size measurement within that interval was
not used as the individual's size at \(t+1\), so the transition was also
excluded from the growth data.

| Census | Observation | Interpretation |
|:--|:--|:--|
| March, year \(t\) | Individual measured during the annual census | Alive at \(t\) |
| June, year \(t\) | Individual coded as dead (`qsurv = 0`) | Death occurred during the transition |
| September, year \(t\) | Individual recorded again during a quarterly census | Does not override the death record |
| March, year \(t+1\) | Individual absent from the annual census | `survives = 0`; no growth transition |

```{r data-debugged-survival-growth}
df_death_interval <- df_demog_long %>%
  filter(!is.na(stg)) %>%
  transmute(
    id = unique_id, year,
    census_time = year * 12 + month,
    target_time = (year + 1) * 12 + 3) %>%
  left_join(
    df_demog_long %>%
      filter(qsurv == 0) %>%
      transmute(
        id = unique_id,
        death_time = year * 12 + month),
    by = "id", relationship = "many-to-many") %>%
  group_by(id, year) %>%
  summarise(
    death_in_interval = any(
      !is.na(death_time) &
        death_time > census_time &
        death_time <= target_time),
    .groups = "drop")

df_transitions <- df_annual_recruit %>%
  left_join(df_death_interval, by = c("id", "year")) %>%
  mutate(
    death_in_interval = replace_na(death_in_interval, FALSE)) %>%
  group_by(id) %>%
  arrange(year, month, .by_group = TRUE) %>%
  mutate(
    next_year = lead(year),
    last_observed_year = max(year, na.rm = TRUE),
    annual_transition = coalesce(next_year == year + 1, FALSE),
    observed_later = last_observed_year > year,
    valid_growth_transition =
      annual_transition & !death_in_interval,
    survives = case_when(
      death_in_interval ~ 0,
      annual_transition ~ 1,
      observed_later ~ 1,
      TRUE ~ NA_real_)) %>%
  ungroup()

df_transition_summary <- df_transitions %>%
  summarise(
    recorded_deaths = sum(death_in_interval),
    observed_survival = sum(
      annual_transition & !death_in_interval),
    inferred_survival = sum(
      observed_later & !annual_transition &
        !death_in_interval),
    unknown_survival = sum(is.na(survives)),
    growth_transitions = sum(valid_growth_transition))

knitr::kable(
  df_transition_summary,
  col.names = c(
    "Recorded deaths",
    "Observed survival",
    "Survival inferred from later observations",
    "Unknown survival",
    "Valid growth transitions"))
```

When an individual was not recorded in year \(t+1\), but was observed again in
a later year, it was considered to have survived the first transition. This
accounts for known gaps in annual sampling. Growth was not reconstructed across
these gaps because no consecutive size measurement was available.

Growth measurements at \(t+1\) were retained only when the individual was
recorded in the following annual census and no death occurred during the
transition.



## Disturbance

Disturbance was assigned at the quadrat level rather than the site level.
Prescribed burns did not affect every quadrat within a site equally. In the
updated data, several site-year combinations contain both disturbed and
undisturbed quadrats. Aggregating disturbance to the site level would therefore
remove observed spatial variation.

The disturbance variable describes whether a quadrat was affected by fire
during an annual transition. A recorded fire effect was coded as 1 and an
explicitly unburned quadrat was coded as 0. A missing classification was
retained as `NA`, because it provides no evidence that the quadrat was
unburned.

### Source of the 2022 fire classification

Fire information through 2020 was obtained from the historical post-burn
variables in `Demog`. The 2022 fire classification was instead obtained from
`QuadLevel`.

The metadata for `postburn_status0322` distinguishes three possible values:
1 for burned, 0 for unburned, and `NA` for a missing plot. However, the workbook
does not establish why a classification is missing. A missing value could
represent a missing field observation, incomplete data entry, or another
unrecorded reason. It is therefore described here simply as having no recorded
classification.

The two worksheets were compared before selecting the source of the 2022
variable.

```{r data-debugged-disturbance-source}
df_dist_2022_demog <- df_demog_wide %>%
  filter(!is.na(quad)) %>%
  group_by(quad) %>%
  summarise(
    dist_2022_demog = if_else(
      all(is.na(postburn_status0322)),
      NA_real_,
      as.numeric(
        any(postburn_status0322 == 1, na.rm = TRUE))),
    .groups = "drop")

df_dist_2022 <- df_quad %>%
  transmute(
    quad,
    dist_2022 = as.numeric(postburn_status0322))

df_2022_source_summary <- df_dist_2022 %>%
  left_join(
    df_dist_2022_demog,
    by = "quad") %>%
  summarise(
    quadrats = n(),
    missing_in_quadlevel = sum(is.na(dist_2022)),
    missing_in_demog = sum(is.na(dist_2022_demog)),
    available_in_quadlevel_but_missing_in_demog = sum(
      !is.na(dist_2022) &
        is.na(dist_2022_demog)),
    conflicting_values = sum(
      !is.na(dist_2022) &
        !is.na(dist_2022_demog) &
        dist_2022 != dist_2022_demog))

knitr::kable(
  df_2022_source_summary,
  col.names = c(
    "Quadrats",
    "Missing in QuadLevel",
    "Missing in Demog",
    "Available only in QuadLevel",
    "Conflicting values"))
```

`QuadLevel` contained an explicit 2022 classification for
`r df_2022_source_summary$available_in_quadlevel_but_missing_in_demog`
quadrats that lacked the corresponding information in `Demog`. There were
`r df_2022_source_summary$conflicting_values` conflicts among quadrats with
information in both worksheets. `QuadLevel` was therefore used as the source
of the 2022 fire classification.

Only `r df_2022_source_summary$missing_in_quadlevel` actual quadrat lacked a
recorded 2022 classification. This quadrat was retained as `NA` rather than
being added to the explicitly unburned group.

### Assignment to annual transitions

For the fires through 2020, a positive post-burn category was coded as 1 and a
recorded category of zero was coded as 0. If all relevant measurements were
missing during a known fire event, the classification remained `NA`. The 2022
classification was joined from `QuadLevel`.

```{r data-debugged-disturbance}
df_dist_event <- df_demog_wide %>%
  filter(!is.na(quad)) %>%
  group_by(quad) %>%
  summarise(
    dist_2001 = if_else(
      all(is.na(postburn_status1201)), NA_real_,
      as.numeric(any(
        postburn_status1201 > 0,
        na.rm = TRUE))),
    dist_2007 = if_else(
      all(is.na(postburn_pct1207)), NA_real_,
      as.numeric(any(
        postburn_pct1207 > 0,
        na.rm = TRUE))),
    dist_2009 = if_else(
      all(is.na(postburn_pct0609)), NA_real_,
      as.numeric(any(
        postburn_pct0609 > 0,
        na.rm = TRUE))),
    dist_2015 = if_else(
      all(is.na(postburn_pct0615)), NA_real_,
      as.numeric(any(
        postburn_pct0615 > 0,
        na.rm = TRUE))),
    dist_2019 = if_else(
      all(is.na(postburn_pct0319)), NA_real_,
      as.numeric(any(
        postburn_pct0319 > 0,
        na.rm = TRUE))),
    dist_2020 = if_else(
      all(is.na(postburn_pct0120)), NA_real_,
      as.numeric(any(
        postburn_pct0120 > 0,
        na.rm = TRUE))),
    .groups = "drop") %>%
  left_join(
    df_dist_2022,
    by = "quad")

df_dist_transition <- bind_rows(
  df_dist_event %>%
    transmute(
      quad,
      year = 2001,
      dist_transition = dist_2001),
  df_dist_event %>%
    transmute(
      quad,
      year = 2007,
      dist_transition = dist_2007),
  df_dist_event %>%
    transmute(
      quad,
      year = 2009,
      dist_transition = dist_2009),
  df_dist_event %>%
    transmute(
      quad,
      year = 2015,
      dist_transition = dist_2015),
  df_dist_event %>%
    transmute(
      quad,
      year = 2018,
      dist_transition = dist_2019),
  df_dist_event %>%
    transmute(
      quad,
      year = 2019,
      dist_transition = dist_2020),
  df_dist_event %>%
    transmute(
      quad,
      year = 2021,
      dist_transition = dist_2022))

df_transitions_dist <- df_transitions %>%
  left_join(
    df_dist_transition,
    by = c("quad", "year")) %>%
  mutate(
    dist_transition = case_when(
      year %in% c(
        2001, 2007, 2009, 2015,
        2018, 2019, 2021) ~
        dist_transition,
      TRUE ~ 0))

n_mixed_site_years <- df_transitions_dist %>%
  filter(!is.na(dist_transition)) %>%
  distinct(
    patch,
    year,
    dist_transition) %>%
  count(patch, year) %>%
  filter(n > 1) %>%
  nrow()

df_disturbance_mapping <- tribble(
  ~`Fire year`, ~`Source`, ~`Source variable`,
  ~`Transition year`,
  2001, "Demog", "postburn_status1201", 2001,
  2007, "Demog", "postburn_pct1207", 2007,
  2009, "Demog", "postburn_pct0609", 2009,
  2015, "Demog", "postburn_pct0615", 2015,
  2019, "Demog", "postburn_pct0319", 2018,
  2020, "Demog", "postburn_pct0120", 2019,
  2022, "QuadLevel", "postburn_status0322", 2021) %>%
  left_join(
    df_transitions_dist %>%
      filter(is.na(dist_transition)) %>%
      count(
        year,
        name = "Missing annual records"),
    by = c("Transition year" = "year")) %>%
  mutate(
    `Missing annual records` = replace_na(
      `Missing annual records`,
      0L))

knitr::kable(df_disturbance_mapping)
```

The fire event was assigned to the annual transition during which it occurred.
The fires in 2001, 2007, 2009, and 2015 occurred after the annual census and
were assigned to transitions beginning in those years. The fires in 2019,
2020, and 2022 were assigned to the transitions ending in those years.

Disturbance differed among quadrats within
`r n_mixed_site_years` site-year combinations, supporting its retention at the
quadrat level. In total,
`r sum(is.na(df_transitions_dist$dist_transition))` annual records have no
recorded fire classification during a known fire transition. The number
associated with each fire is reported in the final column of the table.




## Final working dataset

The cleaned annual records were exported as the working dataset used by the
demographic models. The dataset retains the site, quadrat, individual, and year
identifiers so that observations can be analysed at the appropriate spatial
and temporal levels.

```{r data-debugged-final-working-data}
file_workdata <- file.path(
  dirname(file_workbook),
  "ab_pole_df_workdata_260519.csv")

df_work <- readr::read_csv(
  file_workdata,
  show_col_types = FALSE)

df_work_summary <- df_work %>%
  summarise(
    records = n(),
    individuals = n_distinct(id),
    quadrats = n_distinct(quad),
    sites = n_distinct(site),
    first_year = min(year, na.rm = TRUE),
    last_year = max(year, na.rm = TRUE),
    recruits = sum(recruit == 1, na.rm = TRUE),
    deaths = sum(survives == 0, na.rm = TRUE))

knitr::kable(
  df_work_summary,
  col.names = c(
    "Records", "Individuals", "Quadrats", "Sites",
    "First year", "Last year", "Recruits", "Deaths"))
```

Each row represents one individual at one annual census. Variables measured at
the beginning of the transition are labelled with `t0`, while measurements
from the following annual census are labelled with `t1`. A `t1` measurement is
only available when the individual has a valid transition to the next annual
census.

The final dataset includes height, crown diameter, stem number, flowering stem
number, stage, recruitment, survival, growth, and quadrat-level disturbance.
Plant volume was calculated as a cylindrical size index:

\[
\text{volume} =
\pi \left(\frac{\text{crown diameter}}{2}\right)^2
\times \text{height}.
\]

Log-transformed height and volume variables, including quadratic and cubic
terms, were added for later comparison of nonlinear vital-rate models.

```{r data-debugged-working-missingness}
df_work_missing <- tibble(
  variable = c(
    "survives", "size_t0", "size_t1",
    "volume_t0", "volume_t1", "dist_transition"),
  missing = c(
    sum(is.na(df_work$survives)),
    sum(is.na(df_work$size_t0)),
    sum(is.na(df_work$size_t1)),
    sum(is.na(df_work$volume_t0)),
    sum(is.na(df_work$volume_t1)),
    sum(is.na(df_work$dist_transition)))) %>%
  mutate(percent = 100 * missing / nrow(df_work))

knitr::kable(
  df_work_missing,
  digits = 2,
  col.names = c("Variable", "Missing records", "Percent missing"))
```

Missing `size_t1` and `volume_t1` values mainly represent individuals without a
valid consecutive annual transition. Missing `volume_t0` values occur mainly
among recruits for which crown diameter was not measured. These recruits are
retained in the dataset because they still contribute to recruitment counts,
even when they cannot be used in volume-based vital-rate models.

Disturbance remains missing when no fire classification is recorded for a
quadrat during a known fire event. The available data do not establish why a
classification is missing. These observations were therefore retained as
unknown rather than being classified as undisturbed.

## Sampling inventory

The following figure shows the quadrat-year combinations represented in the
individual demographic data. Each point indicates that at least one individual
record occurs for that quadrat and year. The figure therefore shows the
temporal coverage of the individual-level dataset, including gaps in annual
records and the reduction in individual demographic monitoring after 2022.

```{r data-debugged-sampling-inventory, fig.width=10, fig.height=7}
df_sampling_inventory <- df_demog_long %>%
  filter(
    !is.na(quad),
    !is.na(year)) %>%
  distinct(quad, year)

fig_sampling <- ggplot(
  df_sampling_inventory,
  aes(
    x = year,
    y = quad)) +
  geom_point(size = 0.8) +
  scale_x_continuous(
    breaks = sort(
      unique(df_sampling_inventory$year))) +
  theme_bw() +
  labs(
    title = "Sampling inventory",
    x = "Year",
    y = "Quadrat") +
  theme(
    axis.text.y = element_text(size = 5),
    axis.text.x = element_text(
      angle = 90,
      hjust = 1))

fig_sampling
```

Because this figure is constructed from individual records in
`Demog_longform`, the absence of a point means that no individual demographic
record is available for that quadrat-year. It does not necessarily demonstrate
that the quadrat was not visited. This distinction is especially important
after 2022, when Level 2 quadrats continued to provide quadrat-level counts but
no longer contributed complete individual demographic records.


## Monitoring levels after 2022

Following the 2022 field season, monitoring was reduced in a subset of the
permanent quadrats. The variable `lvl3` identifies whether a quadrat continued
under Level 3 individual demographic monitoring or was monitored using the
reduced Level 2 protocol.

The two monitoring levels contain different types of information:

```{r data-debugged-monitoring-level-definition}
df_level_definition <- tribble(
  ~Feature, ~`Level 2`, ~`Level 3`,
  "Observation unit",
  "Whole quadrat",
  "Individually marked plants",
  "Information recorded",
  paste(
    "Total plants, flowering stems and stems",
    "within the quadrat"),
  paste(
    "Survival, recruitment, stage, size and flowering",
    "of individual plants"),
  "Main worksheet",
  "`QuadLevel`",
  "`Demog_longform`",
  "Use in this project",
  "Population counts and monitoring effort",
  "Annual transitions and vital-rate models")

knitr::kable(df_level_definition)
```

Level 3 monitoring follows individual plants through time. It therefore
provides the information required to estimate individual survival, growth,
flowering and recruitment. These records are stored in `Demog_longform` and
form the basis of the annual working dataset.

Level 2 monitoring does not follow individual plants. Instead, the field team
records quadrat-level totals, including the number of plants, flowering stems
and stems. These observations are stored in `QuadLevel`. Level 2 quadrats
therefore remain part of the monitoring programme, but they cannot provide
individual annual transitions.

The distribution of quadrats between the two monitoring levels and the
availability of recent count data are shown below.

```{r data-debugged-monitoring-level-summary}
df_level_summary <- df_quad %>%
  mutate(
    monitoring_level = case_when(
      lvl3 == 0 ~ "Level 2",
      lvl3 == 1 ~ "Level 3",
      TRUE ~ "Unknown")) %>%
  group_by(monitoring_level) %>%
  summarise(
    quadrats = n_distinct(quad),
    sites = n_distinct(patch),
    count_2023 = sum(!is.na(n2023)),
    count_2024 = sum(!is.na(n2024)),
    count_2025 = sum(!is.na(n2025)),
    .groups = "drop")

knitr::kable(
  df_level_summary,
  col.names = c(
    "Monitoring level", "Quadrats", "Sites",
    "Quadrats with 2023 counts",
    "Quadrats with 2024 counts",
    "Quadrats with 2025 counts"))
```

The following table shows which monitoring levels contributed individual
demographic records from 2023 onward.

```{r data-debugged-monitoring-level-demography}
df_level_demography <- df_demog_long %>%
  filter(year >= 2023) %>%
  dplyr::select(unique_id, quad, year) %>%
  left_join(
    df_quad %>%
      dplyr::select(quad, lvl3),
    by = "quad") %>%
  mutate(
    monitoring_level = case_when(
      lvl3 == 0 ~ "Level 2",
      lvl3 == 1 ~ "Level 3",
      TRUE ~ "Unknown")) %>%
  group_by(monitoring_level) %>%
  summarise(
    demographic_records = n(),
    individuals = n_distinct(unique_id),
    quadrats = n_distinct(quad),
    .groups = "drop")

knitr::kable(
  df_level_demography,
  col.names = c(
    "Monitoring level", "Demographic records",
    "Individuals", "Quadrats"))
```

This distinction is important when interpreting changes in the number of
records after 2022. A reduction in individual records does not necessarily
represent a decline in the population. It can also reflect the change from
Level 3 individual monitoring to Level 2 quadrat-level counts.

The working dataset is constructed from `Demog_longform` and therefore uses
individual-level observations. Level 2 data must be analysed separately or
explicitly incorporated as quadrat-level counts. Raw population totals from
the working dataset should consequently not be compared across the monitoring
change without accounting for the number and type of quadrats sampled.
  
