# Disturbance, year, site projections, and a two-stage IPM {#dist-yr-site-2stage-update}

This chapter builds the updated year-specific integral projection model for
*Polygala lewtonii*. It follows the organization of the corresponding
*Crotalaria avonensis* chapter while retaining the vital rates and biological
assumptions specific to *P. lewtonii*.

The model contains five demographic processes:

1. survival;
2. growth in log volume;
3. flowering probability;
4. flowering-stem number conditional on flowering; and
5. recruitment from flowering stems in year $t_0$ to recruits in year $t_1$.

Survival, growth, flowering, and flowering-stem production vary among years.
Recruitment has a random slope for recruitment year but **no site random
effect**. Site is retained for aggregating flowering stems and recruits, applying
site-specific fire histories, constructing observed initial size distributions,
and validating projected population growth.

The recruitment relationship is forced through $(0,0)$ on the `log1p` scale.
Both the undisturbed and disturbed slopes, including every recruitment-year
random-slope realization, therefore predict zero recruits when flowering-stem
production is zero.

The IPM remains a two-stage, continuous-size model. Its block structure is

\[
\mathbf{K} =
\begin{bmatrix}
\mathbf{0} & \mathbf{F}_{A\rightarrow R} \\
\mathbf{T}_{R\rightarrow A} & \mathbf{T}_{A\rightarrow A}
\end{bmatrix},
\]

where recruits have their own survival and growth transition and adults produce
new recruits.

## Required objects and model settings

This chapter uses the directory and species objects created in the introduction.
It then loads the processed annual workdata directly, because the earlier data
chapter does not yet contain every variable required by this updated model.

```{r dist_yr_site_2stage_required_objects}
required_objects <- c('dir_data', 'v_sp_abb', 'v_ggp_suffix')
missing_objects <- required_objects[!purrr::map_lgl(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop(
    'The following objects are missing. Run the introduction first: ',
    paste(missing_objects, collapse = ', '))
}

v_years_re <- c()
v_mod_set_su <- c(6)
v_mod_set_gr <- c(9)
v_mod_set_fl <- c()
v_mod_set_fl_n <- c(2)
v_mod_set_re <- c()

```

## Data and disturbance timing

The chapter loads the processed annual workdata used by the standalone model. Current-year disturbance affects survival, growth, and recruitment. Flowering probability and flowering-stem number use disturbance in the preceding year.

```{r dist_yr_site_2stage_data}
# Load workdata ----

df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv')))



# Add disturbance from the exact previous year ----

df_dist_prev <- df %>%
  distinct(site, year, dist_transition) %>%
  transmute(
    site,
    year = year + 1,
    disturbance_prev = dist_transition)

df <- df %>%
  select(-any_of("disturbance_prev")) %>%
  left_join(df_dist_prev, by = c("site", "year")) %>%
  mutate(
    disturbance_prev = replace_na(disturbance_prev, 0),
    disturbance = factor(dist_transition, levels = c(0, 1)),
    disturbance_prev = factor(disturbance_prev, levels = c(0, 1)),
    year = as.integer(year),
    site = factor(site))

if (length(v_years_re) > 0) {
  df <- df %>% filter(!year %in% v_years_re)
}


# Controls ---------------------------------------------------------------------
ctrl_glmer <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))
```

## Survival

Survival is modeled separately for recruits and established plants through recruit terms in the candidate set. All candidate models contain a random year intercept.

```{r dist_yr_site_2stage_survival_data}
# Survival data ----------------------------------------------------------------
df_su <- df %>%
  filter(
    !is.na(survives),
    !is.na(logvol_t0),
    !is.na(logvol_t0_2),
    !is.na(logvol_t0_3),
    !is.na(recruits),
    size_t0 != 0) %>%
  mutate(
    year = factor(year),
    recruits = factor(recruits),
    disturbance = factor(disturbance, levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage, recruits)

# Survival model ---------------------------------------------------------------
mod_su_0 <- glmer(
  survives ~ disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_1 <- glmer(
  survives ~ logvol_t0 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_2 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_3 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_01 <- glmer(
  survives ~ recruits + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_11 <- glmer(
  survives ~ logvol_t0 + recruits + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_21 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + recruits + disturbance +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_31 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + recruits +
    disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_12 <- glmer(
  survives ~ logvol_t0 * recruits + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_22 <- glmer(
  survives ~ logvol_t0 * recruits + logvol_t0_2:recruits +
    disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_32 <- glmer(
  survives ~ logvol_t0 * recruits + logvol_t0_2:recruits +
    logvol_t0_3:recruits + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mods_su <- list(
  mod_su_0, mod_su_1, mod_su_2, mod_su_3,
  mod_su_01, mod_su_11, mod_su_21, mod_su_31,
  mod_su_12, mod_su_22, mod_su_32)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef <- coef(mod_su_best)$year

mod_su_best
summary(mod_su_best)
mods_su_dAIC
```

The following panels show conditional year-specific predictions, including the fitted year random effects.

```{r dist_yr_site_2stage_survival_plot, fig.width=10, fig.height=12}
# Survival plots by year -------------------------------------------------------
make_su_year_plot <- function(year_i) {
  df_i <- df_su %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    recruits = levels(df_su$recruits),
    disturbance = levels(df_su$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      recruits = factor(recruits, levels = levels(df_su$recruits)),
      disturbance = factor(disturbance, levels = levels(df_su$disturbance)),
      year = factor(year_i, levels = levels(df_su$year)))
  
  pred_i <- pred_i %>%
    mutate(
      survives = predict(
        mod_su_best,
        newdata = pred_i,
        type = "response",
        re.form = NULL))
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, recruits, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      survives = mean(survives, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, survives, color = recruits, shape = disturbance),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, survives, color = recruits, linetype = disturbance),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = '#BBB857', '1' = '#3666DC')) +
    scale_linetype_manual(values = c('0' = 'solid', '1' = 'dashed')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = expression('Survival probability'[t1])) +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

su_yrs <- lapply(levels(df_su$year), make_su_year_plot)
fig_su_years <- wrap_plots(su_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_su_years
```

## Growth

Growth is measured as log volume at the next annual census. Candidate models allow recruit-specific growth and year-specific intercepts and slopes.

```{r dist_yr_site_2stage_growth_data}
# Growth data ------------------------------------------------------------------
df_gr <- df %>%
  filter(
    !is.na(size_t1), !is.na(size_t0),
    !is.na(logsize_t0), !is.na(logsize_t1),
    !is.na(logsize_t0_2), !is.na(logsize_t0_3),
    !is.na(volume_t0), !is.na(volume_t1),
    !is.na(logvol_t0), !is.na(logvol_t1),
    !is.na(logvol_t0_2), !is.na(logvol_t0_3),
    !is.na(recruits),
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1)) %>%
  mutate(
    year = factor(year),
    recruits = factor(recruits),
    disturbance = factor(disturbance, levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage, recruits)

fig_gr_raw <- ggplot(df_gr, aes(logvol_t0, logvol_t1)) +
  geom_point(alpha = 0.5, pch = 16, size = 0.7) +
  geom_abline(intercept = 0, slope = 1) +
  theme_bw() +
  labs(
    title = 'Growth',
    subtitle = v_ggp_suffix,
    x = expression('log(volume)'[t0]),
    y = expression('log(volume)'[t1])) +
  theme(plot.subtitle = element_text(size = 8))

fig_gr_raw

# Growth model -----------------------------------------------------------------
mod_gr_0 <- lmer(
  logvol_t1 ~ disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_1 <- lmer(
  logvol_t1 ~ logvol_t0 + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_2 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_3 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_01 <- lmer(
  logvol_t1 ~ recruits + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_11 <- lmer(
  logvol_t1 ~ logvol_t0 + recruits + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_21 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + recruits + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_31 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    recruits + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_12 <- lmer(
  logvol_t1 ~ logvol_t0 * recruits + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_22 <- lmer(
  logvol_t1 ~ logvol_t0 * recruits + logvol_t0_2:recruits +
    disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_221 <- lmer(
  logvol_t1 ~ logvol_t0 * recruits + logvol_t0_2 + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_32 <- lmer(
  logvol_t1 ~ logvol_t0 * recruits + logvol_t0_2:recruits +
    logvol_t0_3:recruits + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)

mods_gr <- list(
  mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3,
  mod_gr_01, mod_gr_11, mod_gr_21, mod_gr_31,
  mod_gr_12, mod_gr_22, mod_gr_221, mod_gr_32)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef <- coef(mod_gr_best)$year

mod_gr_best
summary(mod_gr_best)
mods_gr_dAIC
```

The growth panels compare observed and predicted next-year log volume.

```{r dist_yr_site_2stage_growth_plot, fig.width=10, fig.height=12}
# Growth plots by year ---------------------------------------------------------
make_gr_year_plot <- function(year_i) {
  df_i <- df_gr %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    recruits = levels(df_gr$recruits),
    disturbance = levels(df_gr$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      recruits = factor(recruits, levels = levels(df_gr$recruits)),
      disturbance = factor(disturbance, levels = levels(df_gr$disturbance)),
      year = factor(year_i, levels = levels(df_gr$year)))
  
  pred_i <- pred_i %>%
    mutate(
      logvol_t1 = predict(
        mod_gr_best,
        newdata = pred_i,
        re.form = NULL))
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, recruits, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      logvol_t1 = mean(logvol_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, logvol_t1, color = recruits, shape = disturbance),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, logvol_t1, color = recruits, linetype = disturbance),
      linewidth = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = 'blue', lty = 2) +
    scale_color_manual(values = c('0' = '#BBB857', '1' = '#3666DC')) +
    scale_linetype_manual(values = c('0' = 'solid', '1' = 'dashed')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = expression('log(volume)'[t1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

gr_yrs <- lapply(levels(df_gr$year), make_gr_year_plot)
fig_gr_years <- wrap_plots(gr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Growth - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_gr_years

# Growth variance --------------------------------------------------------------
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

## Flowering probability

Flowering uses the disturbance state from the preceding year. Recruits are retained because some observed recruits flowered during the same annual census.

```{r dist_yr_site_2stage_flowering_data}
# Flower data ------------------------------------------------------------------
# Recruits should theoretically be excluded from flowering probability and flower-number models.
# That exclusion is not supported by the data, because 35 recruits did flower, 
# including several fairly large individuals and some stage-3 plants.
df_fl <- df %>%
  filter(
    !is.na(flower),
    !is.na(logvol_t0),
    !is.na(logvol_t0_2),
    !is.na(logvol_t0_3)) %>%
  mutate(
    year = factor(year),
    disturbance_prev = factor(
      disturbance_prev, levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, flower, fl_nr, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance_prev, stage, recruits)

# Flower model -----------------------------------------------------------------
mod_fl_00 <- glmer(
  flower ~ 1 + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_01 <- glmer(
  flower ~ disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_10 <- glmer(
  flower ~ logvol_t0 + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_11 <- glmer(
  flower ~ logvol_t0 + disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_20 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_21 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + disturbance_prev +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_30 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_31 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mods_fl <- list(mod_fl_00, mod_fl_01, mod_fl_10, mod_fl_11, 
                mod_fl_20, mod_fl_21, mod_fl_30, mod_fl_31)
mods_fl_dAIC <- bbmle::AICctab(mods_fl, weights = TRUE, sort = FALSE)$dAIC
mods_fl_sorted <- order(mods_fl_dAIC)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_best <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef <- coef(mod_fl_best)$year

mod_fl_best
summary(mod_fl_best)
mods_fl_dAIC
```

Conditional predictions are displayed for every modeled year and both previous-year disturbance states.

```{r dist_yr_site_2stage_flowering_plot, fig.width=10, fig.height=12}
# Flowering plots by year ------------------------------------------------------
make_fl_year_plot <- function(year_i) {
  df_i <- df_fl %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance_prev = levels(df_fl$disturbance_prev)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_prev = factor(
        disturbance_prev, levels = levels(df_fl$disturbance_prev)),
      year = factor(year_i, levels = levels(df_fl$year)))
  
  pred_i <- pred_i %>%
    mutate(
      flower = predict(
        mod_fl_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance_prev) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      flower = mean(flower, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, flower, color = disturbance_prev),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, flower, color = disturbance_prev),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = 'Flowering probability') +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_yrs <- lapply(levels(df_fl$year), make_fl_year_plot)
fig_fl_years <- wrap_plots(fl_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flowering - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_years
```

## Flowering-stem number

Flowering-stem number is modeled conditional on flowering with a negative binomial GLMM and a random year intercept.

```{r dist_yr_site_2stage_flower_number}
# Flower number conditional on flowering data ----------------------------------
df_fl_cond <- df_fl %>%
  filter(flower == 1, !is.na(fl_nr), fl_nr > 0, fl_nr == round(fl_nr)) %>%
  mutate(year = factor(year))

# Flower number model ----------------------------------------------------------
mod_fl_n_0 <- glmer.nb(
  fl_nr ~ disturbance_prev + (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_1 <- glmer.nb(
  fl_nr ~ logvol_t0 + disturbance_prev + (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_2 <- glmer.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + disturbance_prev +
    (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_3 <- glmer.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance_prev +
    (1 | year),
  data = df_fl_cond, control = ctrl_glmer)

mods_fl_n <- list(mod_fl_n_0, mod_fl_n_1, mod_fl_n_2, mod_fl_n_3) # 
mods_fl_n_dAIC <- bbmle::AICctab(
  mods_fl_n, weights = TRUE, sort = FALSE)$dAIC
mods_fl_n_sorted <- order(mods_fl_n_dAIC)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_bestfit <- mods_fl_n_sorted[1]
  v_mod_fl_n_index <- mod_fl_n_index_bestfit - 1
} else {
  mod_fl_n_index_bestfit <- v_mod_set_fl_n + 1
  v_mod_fl_n_index <- v_mod_set_fl_n
}

mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_bestfit]]
mod_fl_n_ranef <- coef(mod_fl_n_best)$year

mod_fl_n_best
summary(mod_fl_n_best)
mods_fl_n_dAIC

# Flower number plots by year --------------------------------------------------
make_fl_n_year_plot <- function(year_i) {
  df_i <- df_fl_cond %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance_prev = levels(df_fl_cond$disturbance_prev)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_prev = factor(
        disturbance_prev, levels = levels(df_fl_cond$disturbance_prev)),
      year = factor(year_i, levels = levels(df_fl_cond$year)))
  
  pred_i <- pred_i %>%
    mutate(
      fl_nr = predict(
        mod_fl_n_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance_prev) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      fl_nr = mean(fl_nr, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, fl_nr, color = disturbance_prev),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, fl_nr, color = disturbance_prev),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = 'Number of flowering stems') +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_n_yrs <- lapply(levels(df_fl_cond$year), make_fl_n_year_plot)
fig_fl_n_years <- wrap_plots(fl_n_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flower number - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_n_years
```

## Flowering stems to recruits

Recruitment is aggregated by site and transition year. Site identifies the observational unit, but it is not fitted as a random effect. Total flowering stems in year $t_0$ predict recruits in year $t_1$.

```{r dist_yr_site_2stage_recruitment_data}
# Flowering stems to recruits --------------------------------------------------
# Biological note:
# P. lewtonii has post-fire seedbank recruitment, so recruitment is modeled from
# site-year flowering stems at t0 to recruits at t1, with fire as a fixed effect.
df_fs2r <- df %>%
  group_by(site, year) %>%
  summarise(
    fs_t0 = sum(fl_nr, na.rm = TRUE),
    disturbance_t0 = factor(
      max(as.numeric(as.character(disturbance)), na.rm = TRUE),
      levels = c(0, 1)),
    .groups = "drop") %>%
  mutate(year_t1 = year + 1) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(site, year, name = "re_t1"),
    by = c("site", "year_t1" = "year")) %>%
  mutate(
    re_t1 = replace_na(re_t1, 0L),
    site = factor(site),
    year_t0 = factor(year),
    year_t1 = factor(year_t1),
    fs_t0_log = log1p(fs_t0),
    re_t1_log = log1p(re_t1))

# Keep the old mean-script recruitment plot as a quick check.
fig_fs2r_raw <- ggplot(
  df_fs2r,
  aes(x = fs_t0, y = re_t1, color = disturbance_t0)) +
  geom_jitter(height = 0.2, width = 0.5, alpha = 0.4) +
  scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
  theme_bw() +
  labs(
    title = 'Recruits t1 by flowering stems t0',
    subtitle = v_ggp_suffix,
    x = 'Total flowering stems at site in year t',
    y = 'Number of recruits at site in year t + 1',
    color = 'Fire')

fig_fs2r_raw
```

### Recruitment models through the origin

Both candidate models contain a random flowering-stem slope for recruitment year. Neither contains a fixed intercept, a disturbance main effect, or a random year intercept. Consequently, every mean and year-specific relationship passes through $(0,0)$ on the `log1p` scale. The second model allows disturbance to alter the flowering-stem slope.

```{r dist_yr_site_2stage_recruitment_models}
# Recruitment model ------------------------------------------------------------
ctrl_re <- lmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 2e5))

df_fs2r <- df_fs2r %>%
  mutate(disturbance = as.numeric(as.character(disturbance_t0)))

mod_re_0 <- lmer(
  re_t1_log ~ 0 + fs_t0_log + (0 + fs_t0_log | year_t1),
  data = df_fs2r, REML = FALSE, control = ctrl_re)

mod_re_1 <- lmer(
  re_t1_log ~ 0 + fs_t0_log + fs_t0_log:disturbance +
    (0 + fs_t0_log | year_t1),
  data = df_fs2r, REML = FALSE, control = ctrl_re)


mods_re <- list(mod_re_0, mod_re_1)

mods_re_dAIC <- bbmle::AICctab(mods_re, weights = TRUE, sort = FALSE)$dAIC
mods_re_sorted <- order(mods_re_dAIC)

if (length(v_mod_set_re) == 0) {
  mod_re_index_bestfit <- mods_re_sorted[1]
  v_mod_re_index <- mod_re_index_bestfit - 1
} else {
  mod_re_index_bestfit <- v_mod_set_re + 1
  v_mod_re_index <- v_mod_set_re
}

mod_re_best <- mods_re[[mod_re_index_bestfit]]

mod_re_best
summary(mod_re_best)
mods_re_dAIC
```

Sites are retained as point colors only. The fitted lines vary by recruitment year and disturbance, not by site.

```{r dist_yr_site_2stage_recruitment_plot, fig.width=10, fig.height=12}
# Recruitment plots by recruit year -------------------------------------------
predict_re_safe <- function(model, newdata, re.form = NULL) {
  if (inherits(model, 'merMod')) {
    out <- predict(
      model, newdata = newdata, re.form = re.form, allow.new.levels = TRUE)
  } else {
    out <- predict(model, newdata = newdata)
  }
  return(out)
}

shape_values_re <- c(16, 17, 15, 18, 3, 4, 7, 8, 0, 1, 2, 5, 6, 9)
site_shapes_re <- setNames(
  shape_values_re[seq_along(levels(df_fs2r$site))],
  levels(df_fs2r$site))

make_re_year_plot <- function(year_i) {
  df_i <- df_fs2r %>% filter(year_t1 == year_i)
  
  x <- seq(
    0, max(df_i$fs_t0_log, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    fs_t0_log = x,
    disturbance_t0 = levels(df_fs2r$disturbance_t0)) %>%
    mutate(
      disturbance = as.numeric(as.character(disturbance_t0)),
      year_t1 = factor(year_i, levels = levels(df_fs2r$year_t1)),
      disturbance_t0 = factor(
        disturbance_t0, levels = levels(df_fs2r$disturbance_t0)))
  
  pred_i <- pred_i %>%
    mutate(
      pred_re = predict_re_safe(
        mod_re_best, pred_i, re.form = NULL))
  
  ggplot() +
    geom_point(
      data = df_i,
      aes(fs_t0_log, re_t1_log, color = site, shape = site),
      alpha = 0.75, size = 2) +
    geom_line(
      data = pred_i,
      aes(
        fs_t0_log, pred_re,
        group = disturbance_t0,
        linetype = disturbance_t0),
      inherit.aes = FALSE, color = 'black', linewidth = 0.8) +
    scale_color_viridis_d(option = 'turbo', end = 0.95) +
    scale_shape_manual(values = site_shapes_re) +
    scale_linetype_manual(values = c('0' = 'solid', '1' = 'dashed')) +
    theme_bw() +
    labs(
      title = year_i,
      x = expression('log(1 + flowering stems)'[t0]),
      y = expression('log(1 + recruits)'[t1]),
      color = 'Site',
      shape = 'Site',
      linetype = 'Fire') +
    theme(text = element_text(size = 5), legend.position = 'none')
}

re_yrs <- lapply(levels(df_fs2r$year_t1), make_re_year_plot)
fig_re_years <- wrap_plots(re_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Recruitment - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_re_years
```

## Recruit-size distribution

Observed recruit sizes define the size distribution used to place new recruits on the IPM mesh.

```{r dist_yr_site_2stage_recruit_size}
# Recruit size distribution ----------------------------------------------------
df_recr <- df %>%
  filter(recruits == 1, volume_t0 > 0, is.finite(logvol_t0))

rc_sz <- data.frame(
  mean = mean(df_recr$logvol_t0, na.rm = TRUE),
  sd = sd(df_recr$logvol_t0, na.rm = TRUE))
```

## Parameter extraction

Fixed effects, conditional year coefficients, and recruitment-year random-slope deviations are stored separately before each annual IPM is assembled.

```{r dist_yr_site_2stage_parameter_helpers}
# Parameter extraction helpers -------------------------------------------------
empty_coef_df <- function() {
  data.frame(coefficient = character(), value = numeric())
}

is_mixed_model <- function(model) {
  inherits(model, 'merMod')
}

get_fixef_safe <- function(model) {
  if (is_mixed_model(model)) {
    fixef(model)
  } else {
    coef(model)
  }
}

get_group_coef_safe <- function(model, group_var) {
  if (!is_mixed_model(model)) {
    return(NULL)
  }
  coef_list <- coef(model)
  if (!(group_var %in% names(coef_list))) {
    return(NULL)
  }
  coef_list[[group_var]]
}

find_model_term <- function(model_terms, aliases) {
  aliases <- unlist(aliases)
  for (alias in aliases) {
    if (grepl('^regex:', alias)) {
      pattern <- sub('^regex:', '', alias)
      hit <- grep(pattern, model_terms, value = TRUE)
    } else {
      hit <- model_terms[model_terms == alias]
    }
    if (length(hit) > 0) {
      return(hit[1])
    }
  }
  NA_character_
}

bind_coef_rows <- function(x) {
  x <- x[!vapply(x, is.null, logical(1))]
  if (length(x) == 0) {
    return(empty_coef_df())
  }
  out <- do.call(rbind, x)
  rownames(out) <- NULL
  out
}

check_duplicate_coefficients <- function(x, object_name = 'parameter table') {
  dup <- x$coefficient[duplicated(x$coefficient)]
  if (length(dup) > 0) {
    stop(
      object_name,
      ' contains duplicated coefficient names: ',
      paste(unique(dup), collapse = ', '))
  }
  invisible(TRUE)
}

coef_df_to_list <- function(x) {
  if (nrow(x) == 0) {
    return(list())
  }
  check_duplicate_coefficients(x)
  x %>%
    mutate(coefficient = as.character(coefficient)) %>%
    pivot_wider(names_from = coefficient, values_from = value) %>%
    as.list()
}

extract_fixed_pars <- function(model, term_map, fill_missing = TRUE) {
  model_coefs <- get_fixef_safe(model)
  model_terms <- names(model_coefs)
  out <- lapply(names(term_map), function(coef_name) {
    model_term <- find_model_term(model_terms, term_map[[coef_name]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- 0
      } else {
        return(NULL)
      }
    } else {
      value <- unname(model_coefs[model_term])
    }
    data.frame(coefficient = coef_name, value = value)
  })
  bind_coef_rows(out)
}

extract_group_pars <- function(model, group_var, term_map, fill_missing = TRUE) {
  coef_matrix <- get_group_coef_safe(model, group_var)
  if (is.null(coef_matrix)) {
    return(empty_coef_df())
  }
  model_terms <- colnames(coef_matrix)
  out <- lapply(names(term_map), function(coef_prefix) {
    model_term <- find_model_term(model_terms, term_map[[coef_prefix]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- rep(0, nrow(coef_matrix))
      } else {
        return(NULL)
      }
    } else {
      value <- coef_matrix[, model_term]
    }
    data.frame(
      coefficient = paste0(coef_prefix, '_', rownames(coef_matrix)),
      value = value)
  })
  bind_coef_rows(out)
}

extract_re_ranef_devs <- function(model, group_var, group_label) {
  if (!inherits(model, 'merMod')) {
    return(empty_coef_df())
  }
  ranef_list <- ranef(model)
  if (!(group_var %in% names(ranef_list))) {
    return(empty_coef_df())
  }
  ranef_matrix <- ranef_list[[group_var]]
  out <- list()
  if ('(Intercept)' %in% colnames(ranef_matrix)) {
    out[['intercept']] <- data.frame(
      coefficient = paste0('re_u0_', group_label, '_',
                           rownames(ranef_matrix)),
      value = ranef_matrix[, '(Intercept)'])
  }
  if ('fs_t0_log' %in% colnames(ranef_matrix)) {
    out[['slope']] <- data.frame(
      coefficient = paste0('re_u1_', group_label, '_',
                           rownames(ranef_matrix)),
      value = ranef_matrix[, 'fs_t0_log'])
  }
  bind_coef_rows(out)
}

# Term maps --------------------------------------------------------------------
size_term_map <- list(
  b0 = c('(Intercept)'),
  b1 = c('logvol_t0'),
  b2 = c('logvol_t0_2', 'I(logvol_t0^2)'),
  b3 = c('logvol_t0_3', 'I(logvol_t0^3)'),
  bf = c('disturbance1', 'regex:^disturbance'),
  br = c('recruits1', 'regex:^recruits1$'),
  b1_br = c('logvol_t0:recruits1', 'recruits1:logvol_t0'),
  b2_br0 = c('logvol_t0_2:recruits0', 'recruits0:logvol_t0_2'),
  b2_br1 = c('logvol_t0_2:recruits1', 'recruits1:logvol_t0_2'),
  b3_br0 = c('logvol_t0_3:recruits0', 'recruits0:logvol_t0_3'),
  b3_br1 = c('logvol_t0_3:recruits1', 'recruits1:logvol_t0_3'))

# Flowering disturbance term map
lagged_size_term_map <- list(
  b0 = c("(Intercept)"),
  b1 = c("logvol_t0"),
  b2 = c("logvol_t0_2", "I(logvol_t0^2)"),
  b3 = c("logvol_t0_3", "I(logvol_t0^3)"),
  bf = c(
    "disturbance_prev1",
    "regex:^disturbance_prev"))

make_term_map <- function(prefix, map) {
  out <- map
  names(out) <- paste0(prefix, names(out))
  out
}

su_term_map <- make_term_map('surv_', size_term_map)
gr_term_map <- make_term_map('grow_', size_term_map)
fl_term_map  <- make_term_map("fl_",  lagged_size_term_map)
fln_term_map <- make_term_map("fln_", lagged_size_term_map)
re_term_map <- list(
  re_b0 = c('(Intercept)'),
  re_b1 = c('fs_t0_log'),
  re_bf = c('fs_t0_log:disturbance', 'disturbance:fs_t0_log'))

# Fixed parameters -------------------------------------------------------------
su_fe <- extract_fixed_pars(mod_su_best, su_term_map, fill_missing = TRUE)
gr_fe <- extract_fixed_pars(mod_gr_best, gr_term_map, fill_missing = TRUE)
fl_fe <- extract_fixed_pars(mod_fl_best, fl_term_map, fill_missing = TRUE)
fln_fe <- extract_fixed_pars(mod_fl_n_best, fln_term_map, fill_missing = TRUE)
re_fe <- extract_fixed_pars(mod_re_best, re_term_map, fill_missing = TRUE)


# Constants --------------------------------------------------------------------
gr_var_coef <- coef(mod_gr_var)

constants <- tibble::tribble(
  ~coefficient, ~value,
  'recr_sz', rc_sz$mean,
  'recr_sd', rc_sz$sd,
  'a', as.numeric(gr_var_coef[1]),
  'b', as.numeric(gr_var_coef[2]),
  'L', min(c(df_gr$logvol_t0, df_fl$logvol_t0), na.rm = TRUE) - 0.1,
  'U', max(c(df_gr$logvol_t0, df_fl$logvol_t0), na.rm = TRUE) + 0.1,
  'mat_siz', 200,
  'fs_t0_ref', mean(df_fs2r$fs_t0, na.rm = TRUE),
  'fs_t0_ref_log', log1p(mean(df_fs2r$fs_t0, na.rm = TRUE)),
  'mod_su_index', v_mod_su_index,
  'mod_gr_index', v_mod_gr_index,
  'mod_fl_index', v_mod_fl_index,
  'mod_fl_n_index', v_mod_fl_n_index,
  'mod_re_index', v_mod_re_index) %>%
  mutate(coefficient = as.character(coefficient), value = as.numeric(value))

pars_cons <- bind_coef_rows(list(su_fe, gr_fe, fl_fe, fln_fe, re_fe,
                                 constants))
check_duplicate_coefficients(pars_cons, object_name = 'pars_cons')
pars_all_mean <- coef_df_to_list(pars_cons)
pars_mean <- pars_all_mean

# Year-varying parameters ------------------------------------------------------
su_out_yr <- extract_group_pars(mod_su_best, 'year', su_term_map,
                                fill_missing = TRUE)
gr_out_yr <- extract_group_pars(mod_gr_best, 'year', gr_term_map,
                                fill_missing = TRUE)
fl_out_yr <- extract_group_pars(mod_fl_best, 'year', fl_term_map,
                                fill_missing = TRUE)
fln_out_yr <- extract_group_pars(mod_fl_n_best, 'year', fln_term_map,
                                 fill_missing = TRUE)

pars_var <- bind_coef_rows(list(su_out_yr, gr_out_yr, fl_out_yr,
                                fln_out_yr))
check_duplicate_coefficients(pars_var, object_name = 'pars_var')
pars_all_year <- coef_df_to_list(pars_var)


# Recruitment-year random-effect deviations ---------------------------
pars_re_year_t1 <- extract_re_ranef_devs(
  mod_re_best, 'year_t1', 'year_t1') %>%
  mutate(coefficient = as.character(coefficient))
check_duplicate_coefficients(
  pars_re_year_t1, object_name = 'pars_re_year_t1')
pars_all_re_year_t1 <- coef_df_to_list(pars_re_year_t1)
```

## IPM functions

The flowering-stem-to-recruit relationship is converted back with the direct inverse `log1p` transformation. No residual-variance correction is added, because that correction would introduce positive recruitment at a zero linear predictor. Disturbance changes the recruitment slope rather than shifting an intercept.

```{r dist_yr_site_2stage_ipm_functions}
# IPM helper functions ---------------------------------------------------------
inv_logit <- function(x) {
  plogis(x)
}

get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    return(pars[[par_name]])
  }
  default
}

make_ipm_pars <- function(
    pars_mean,
    pars_year = NULL,
    pars_re_year_t1 = NULL,
    year = NULL,
    re_year_t1 = NULL) {
  pars <- pars_mean
  
  if (!is.null(year) && !is.null(pars_year)) {
    year <- as.character(year)
    year_hits <- grep(paste0('_', year, '$'), names(pars_year), value = TRUE)
    
    for (nm in year_hits) {
      base_nm <- sub(paste0('_', year, '$'), '', nm)
      pars[[base_nm]] <- pars_year[[nm]]
    }
  }
  
  if (is.null(re_year_t1) && !is.null(year)) {
    re_year_t1 <- as.numeric(as.character(year)) + 1
  }
  
  if (!is.null(re_year_t1) && !is.null(pars_re_year_t1)) {
    re_year_t1 <- as.character(re_year_t1)
    pars$re_b0 <- get_par(pars, 're_b0', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u0_year_t1_', re_year_t1), 0)
    pars$re_b1 <- get_par(pars, 're_b1', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u1_year_t1_', re_year_t1), 0)
  }
  
  pars
}

surv_lp <- function(x, pars, stage = c('adult', 'recruit'), disturbance = 0) {
  stage <- match.arg(stage)
  eta <- get_par(pars, 'surv_b0', 0) +
    get_par(pars, 'surv_b1', 0) * x +
    get_par(pars, 'surv_b2', 0) * x^2 +
    get_par(pars, 'surv_b3', 0) * x^3 +
    get_par(pars, 'surv_bf', 0) * disturbance
  if (stage == 'recruit') {
    eta <- eta +
      get_par(pars, 'surv_br', 0) +
      get_par(pars, 'surv_b1_br', 0) * x +
      get_par(pars, 'surv_b2_br1', 0) * x^2 +
      get_par(pars, 'surv_b3_br1', 0) * x^3
  } else {
    eta <- eta +
      get_par(pars, 'surv_b2_br0', 0) * x^2 +
      get_par(pars, 'surv_b3_br0', 0) * x^3
  }
  eta
}

sx <- function(x, pars, stage = c('adult', 'recruit'), disturbance = 0) {
  inv_logit(surv_lp(x, pars, stage, disturbance))
}

grow_mu <- function(x, pars, stage = c('adult', 'recruit'), disturbance = 0) {
  stage <- match.arg(stage)
  mu <- get_par(pars, 'grow_b0', 0) +
    get_par(pars, 'grow_b1', 0) * x +
    get_par(pars, 'grow_b2', 0) * x^2 +
    get_par(pars, 'grow_b3', 0) * x^3 +
    get_par(pars, 'grow_bf', 0) * disturbance
  if (stage == 'recruit') {
    mu <- mu +
      get_par(pars, 'grow_br', 0) +
      get_par(pars, 'grow_b1_br', 0) * x +
      get_par(pars, 'grow_b2_br1', 0) * x^2 +
      get_par(pars, 'grow_b3_br1', 0) * x^3
  } else {
    mu <- mu +
      get_par(pars, 'grow_b2_br0', 0) * x^2 +
      get_par(pars, 'grow_b3_br0', 0) * x^3
  }
  mu
}

grow_sd <- function(x, pars) {
  sqrt(pars$a * exp(pars$b * x))
}

gxy <- function(x, y, pars, stage = c('adult', 'recruit'), disturbance = 0) {
  stage <- match.arg(stage)
  dnorm(
    y,
    mean = grow_mu(x, pars, stage = stage, disturbance = disturbance),
    sd = grow_sd(x, pars))
}

fl_x <- function(x, pars, disturbance_prev = 0) {
  eta <- get_par(pars, "fl_b0", 0) +
    get_par(pars, "fl_b1", 0) * x +
    get_par(pars, "fl_b2", 0) * x^2 +
    get_par(pars, "fl_b3", 0) * x^3 +
    get_par(pars, "fl_bf", 0) * disturbance_prev
  inv_logit(eta)
}

fl_n_x <- function(x, pars, disturbance_prev = 0) {
  eta <- get_par(pars, "fln_b0", 0) +
    get_par(pars, "fln_b1", 0) * x +
    get_par(pars, "fln_b2", 0) * x^2 +
    get_par(pars, "fln_b3", 0) * x^3 +
    get_par(pars, "fln_bf", 0) * disturbance_prev
  exp(eta)
}

fs_x <- function(x, pars, disturbance_prev = 0) {
  fl_x(x, pars, disturbance_prev) *
    fl_n_x(x, pars, disturbance_prev)
}

re_total_ref <- function(pars, disturbance = 0) {
  re_slope <- get_par(pars, 're_b1', 0) +
    get_par(pars, 're_bf', 0) * disturbance
  re_log <- re_slope * get_par(pars, 'fs_t0_ref_log', 0)
  re_value <- expm1(re_log)
  pmax(re_value, 0)
}

rx <- function(
    x,
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  fs_value <- fs_x(
    x, pars, disturbance_prev = disturbance_prev)
  re_ref <- re_total_ref(
    pars, disturbance = disturbance)
  re_per_fs <- re_ref /
    pmax(get_par(pars, "fs_t0_ref", 0), .Machine$double.eps)
  fs_value * re_per_fs
}

re_y_dist <- function(y, pars, h = NULL) {
  dens <- dnorm(y, mean = pars$recr_sz, sd = pars$recr_sd)
  if (!is.null(h)) {
    dens <- dens / sum(dens * h)
  }
  dens
}

fyx <- function(
    y,
    x,
    pars,
    h,
    disturbance = 0,
    disturbance_prev = 0) {
  Pvec <- re_y_dist(y, pars, h = h)
  Rvec <- rx(
    x,
    pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
  outer(Pvec, Rvec) * h
}
```

## Two-stage kernel

The projection kernel keeps recruits and adults as separate continuous-size stages. Recruits have their own survival and growth transition and enter the adult stage after one projection step. Adults both persist and produce recruits.

```{r dist_yr_site_2stage_kernel}
# Two-stage kernel -------------------------------------------------------------
kernel <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  S_rec <- sx(y, pars, stage = 'recruit', disturbance = disturbance)
  S_ad <- sx(y, pars, stage = 'adult', disturbance = disturbance)
  G_rec <- matrix(0, n, n)
  G_ad <- matrix(0, n, n)
  G_rec[] <- t(outer(
    y, y, Vectorize(function(x, y) {
      gxy(x, y, pars, stage = 'recruit', disturbance = disturbance)
    }))) * h
  G_ad[] <- t(outer(
    y, y, Vectorize(function(x, y) {
      gxy(x, y, pars, stage = 'adult', disturbance = disturbance)
    }))) * h
  T_rec <- matrix(0, n, n)
  T_ad <- matrix(0, n, n)
  for (i in 1:n) {
    if (i <= n / 2) {
      G_rec[1, i] <- G_rec[1, i] + 1 - sum(G_rec[, i])
      G_ad[1, i] <- G_ad[1, i] + 1 - sum(G_ad[, i])
    } else {
      G_rec[n, i] <- G_rec[n, i] + 1 - sum(G_rec[, i])
      G_ad[n, i] <- G_ad[n, i] + 1 - sum(G_ad[, i])
    }
    T_rec[, i] <- G_rec[, i] * S_rec[i]
    T_ad[, i] <- G_ad[, i] * S_ad[i]
  }
  
  F_ad <- fyx(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
  
  K_rec_rec <- matrix(0, n, n)
  K_rec_ad <- F_ad
  K_ad_rec <- T_rec
  K_ad_ad <- T_ad
  k_yx <- rbind(cbind(K_rec_rec, K_rec_ad), cbind(K_ad_rec, K_ad_ad))
  list(
    k_yx = k_yx,
    K_rec_rec = K_rec_rec,
    K_rec_ad = K_rec_ad,
    K_ad_rec = K_ad_rec,
    K_ad_ad = K_ad_ad,
    F_ad = F_ad,
    T_rec = T_rec,
    T_ad = T_ad,
    G_rec = G_rec,
    G_ad = G_ad,
    S_rec = S_rec,
    S_ad = S_ad,
    meshpts = y,
    h = h,
    L = L,
    U = U)
}

lambda_ipm <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  Re(eigen(
    kernel(
      pars,
      disturbance = disturbance,
      disturbance_prev = disturbance_prev)$k_yx)$values[1])
}
```

## Mean and year-specific population growth

Four disturbance scenarios separate current-year effects from lagged effects on flowering and flowering-stem production.

```{r dist_yr_site_2stage_lambda}
# Mean IPMs --------------------------------------------------------------------
lambda_ipm(pars_mean, disturbance = 0, disturbance_prev = 0)
lambda_ipm(pars_mean, disturbance = 0, disturbance_prev = 1)
lambda_ipm(pars_mean, disturbance = 1, disturbance_prev = 1)
lambda_ipm(pars_mean, disturbance = 1, disturbance_prev = 0)



# Year-specific IPMs -----------------------------------------------------------
lambda_ipm_year <- function(
    year,
    disturbance = 0,
    disturbance_prev = 0,
    re_year_t1 = NULL) {
  pars_i <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = year,
    re_year_t1 = re_year_t1)
  
  lambda_ipm(
    pars_i,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
}

# Year-specific disturbance scenarios ----

ipm_years <- sort(unique(as.integer(as.character(df_su$year))))

lambda_year <- data.frame(
  year = ipm_years,
  lambda_undisturbed = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 0,
      disturbance_prev = 0)}),
  lambda_current_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 1,
      disturbance_prev = 0)}),
  lambda_previous_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 0,
      disturbance_prev = 1)}),
  lambda_consecutive_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 1,
      disturbance_prev = 1)})
  )

fig_lambda_year <- lambda_year %>%
  pivot_longer(
    cols = -year,
    names_to = 'disturbance',
    values_to = 'lambda') %>%
  mutate(
    disturbance = recode(
      disturbance,
      lambda_undisturbed = 'Undisturbed',
      lambda_current_fire = 'Current-year fire',
      lambda_previous_fire = 'Previous-year fire',
      lambda_consecutive_fire = 'Fire in both years')) %>%
  ggplot(aes(year, lambda, color = disturbance)) +
  geom_hline(yintercept = 1, linetype = 'dashed') +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(
    title = 'Year-specific asymptotic lambda',
    subtitle = v_ggp_suffix,
    x = 'Year',
    y = expression(lambda),
    color = 'Disturbance')

fig_lambda_year
```

## Site-year projections

Site-specific observed size distributions and fire histories are used for projection and validation. These site-level projections do not add a site random effect to any vital rate.

```{r dist_yr_site_2stage_site_projection}
# Observed and projected site-year population growth ---------------------------
# Site-year disturbance lookup ----

fire_lookup_site <- df %>%
  group_by(site, year) %>%
  summarise(
    disturbance_num = max(
      as.numeric(as.character(disturbance)),
      na.rm = TRUE),
    disturbance_prev_num = max(
      as.numeric(as.character(disturbance_prev)),
      na.rm = TRUE),
    .groups = 'drop')

# Get site-year disturbance state ----

get_disturbance_site_y <- function(site_i, yr, variable) {
  out <- fire_lookup_site %>%
    filter(site == site_i, year == yr) %>%
    pull({{ variable }})
  
  if (length(out) == 0 || is.na(out)) {
    out <- 0
  }
  
  out
}

# Observed site-year counts in the two-stage IPM state
# Sized plants are counted directly. Unsized recruits are counted as recruit
# individuals and are allocated over the recruit size distribution for projection.
df_counts_site <- df %>%
  group_by(site, year) %>%
  summarise(
    n_sized = sum(is.finite(logvol_t0), na.rm = TRUE),
    n_unsized_recruits = sum(!is.finite(logvol_t0) & recruits == 1,
                             na.rm = TRUE),
    n_ipm_state = n_sized + n_unsized_recruits,
    .groups = 'drop')

df_obs_pgr_site <- df_counts_site %>%
  rename(n_t0 = n_ipm_state) %>%
  left_join(
    df_counts_site %>%
      transmute(site, year = year - 1, n_t1 = n_ipm_state),
    by = c('site', 'year')) %>%
  filter(!is.na(n_t1), n_t0 > 0, n_t1 > 0) %>%
  left_join(fire_lookup_site, by = c('site', 'year')) %>%
  mutate(
    disturbance_num = replace_na(disturbance_num, 0),
    disturbance_prev_num = replace_na(disturbance_prev_num, 0),
    obs_pgr = n_t1 / n_t0)

make_initial_n_site <- function(year0, site_i, pars) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  breaks <- seq(L, U, length.out = n + 1)
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  df0 <- df %>% filter(year == year0, site == site_i)
  rec_sized <- df0 %>%
    filter(recruits == 1, is.finite(logvol_t0)) %>%
    pull(logvol_t0)
  ad_sized <- df0 %>%
    filter((is.na(recruits) | recruits != 1), is.finite(logvol_t0)) %>%
    pull(logvol_t0)
  rec_counts <- hist(
    pmin(pmax(rec_sized, L), U),
    breaks = breaks, plot = FALSE, include.lowest = TRUE)$counts
  ad_counts <- hist(
    pmin(pmax(ad_sized, L), U),
    breaks = breaks, plot = FALSE, include.lowest = TRUE)$counts
  rec_density <- rec_counts / h
  ad_density <- ad_counts / h
  n_unsized_recruits <- df0 %>%
    summarise(n = sum(!is.finite(logvol_t0) & recruits == 1, na.rm = TRUE)) %>%
    pull(n)
  if (n_unsized_recruits > 0) {
    rec_density <- rec_density + n_unsized_recruits *
      re_y_dist(y, pars, h = h)
  }
  c(rec_density, ad_density)
}

# Project one site-year ----

project_one_site_year <- function(yr, site_i) {
  
  disturbance_y <- get_disturbance_site_y(
    site_i,
    yr,
    disturbance_num)
  
  disturbance_prev_y <- get_disturbance_site_y(
    site_i,
    yr,
    disturbance_prev_num)
  
  pars_y <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = yr,
    re_year_t1 = yr + 1)
  
  n_obs <- make_initial_n_site(
    year0 = yr,
    site_i = site_i,
    pars = pars_y)
  
  K <- kernel(
    pars = pars_y,
    disturbance = disturbance_y,
    disturbance_prev = disturbance_prev_y)$k_yx
  
  h <- (pars_y$U - pars_y$L) / pars_y$mat_siz
  n_initial <- sum(n_obs) * h
  
  if (n_initial <= 0) {
    return(data.frame(
      year = yr,
      site = site_i,
      disturbance_num = disturbance_y,
      disturbance_prev_num = disturbance_prev_y,
      n_obs_model = n_initial,
      n_proj_model = NA_real_,
      asym_lambda = NA_real_,
      proj_lambda = NA_real_))
  }
  
  n_proj <- K %*% n_obs
  
  data.frame(
    year = yr,
    site = site_i,
    disturbance_num = disturbance_y,
    disturbance_prev_num = disturbance_prev_y,
    n_obs_model = n_initial,
    n_proj_model = sum(n_proj) * h,
    asym_lambda = Re(eigen(K)$values[1]),
    proj_lambda = as.numeric(
      (sum(n_proj) * h) / n_initial))
}

df_proj_site <- bind_rows(
  lapply(seq_len(nrow(df_obs_pgr_site)), function(i) {
    project_one_site_year(
      yr = df_obs_pgr_site$year[i],
      site_i = df_obs_pgr_site$site[i])
  }))


# Join projected and observed site-year values ----
df_compare_site <- df_obs_pgr_site %>%
  left_join(
    df_proj_site,
    by = c(
      'year',
      'site',
      'disturbance_num',
      'disturbance_prev_num')) %>%
  mutate(
    error_asymptotic_vs_obs = asym_lambda - obs_pgr,
    error_projected_vs_obs = proj_lambda - obs_pgr)

df_compare_site %>% print(n = 100)

# Whole-population annual comparison

df_compare <- df_compare_site %>%
  group_by(year) %>%
  summarise(
    asym_lambda = weighted.mean(
      asym_lambda, w = n_obs_model, na.rm = TRUE),
    n_t0 = sum(n_t0, na.rm = TRUE),
    n_t1 = sum(n_t1, na.rm = TRUE),
    obs_pgr = n_t1 / n_t0,
    n_obs_model = sum(n_obs_model, na.rm = TRUE),
    n_proj_model = sum(n_proj_model, na.rm = TRUE),
    proj_lambda = n_proj_model / n_obs_model,
    disturbance = if_else(
      any(disturbance_num == 1, na.rm = TRUE), 'Fire', 'No fire'),
    .groups = 'drop') %>%
  mutate(
    disturbance = factor(disturbance, levels = c('No fire', 'Fire')))

# Observed vs modeled plot

df_plot <- df_compare %>%
  select(year, obs_pgr, asym_lambda, proj_lambda, disturbance) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Abundance-weighted asymptotic lambda',
      proj_lambda = 'Projected lambda from observed site size distributions'))

fig_mod_vs_obs <- ggplot(
  df_plot,
  aes(x = lambda, y = obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = 'Observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs

# Log-transformed observed vs modeled plot

df_plot_log <- df_plot %>%
  filter(obs_pgr > 0, lambda > 0) %>%
  mutate(
    log_obs_pgr = log(obs_pgr),
    log_lambda = log(lambda))

fig_mod_vs_obs_log <- ggplot(
  df_plot_log,
  aes(x = log_lambda, y = log_obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = paste(v_ggp_suffix, '- log-transformed lambda'),
    x = expression('log modeled ' * lambda),
    y = 'log observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs_log
```

## Projection summaries

The final tables compare observed population growth with asymptotic and one-step projected growth at annual and site levels.

```{r dist_yr_site_2stage_summary}
# Summary statistics -----------------------------------------------------------
df_compare_summary <- df_compare %>%
  summarise(
    n_years = n(),
    arithmetic_mean_obs_pgr = mean(obs_pgr, na.rm = TRUE),
    geometric_mean_obs_pgr = exp(mean(log(obs_pgr), na.rm = TRUE)),
    arithmetic_mean_asym_lambda = mean(asym_lambda, na.rm = TRUE),
    geometric_mean_asym_lambda = exp(mean(log(asym_lambda), na.rm = TRUE)),
    arithmetic_mean_proj_lambda = mean(proj_lambda, na.rm = TRUE),
    geometric_mean_proj_lambda = exp(mean(log(proj_lambda), na.rm = TRUE)),
    mean_error_asymptotic_vs_obs = mean(asym_lambda - obs_pgr,
                                        na.rm = TRUE),
    mean_error_projected_vs_obs = mean(proj_lambda - obs_pgr, na.rm = TRUE),
    rmse_asymptotic_vs_obs = sqrt(mean((asym_lambda - obs_pgr)^2,
                                       na.rm = TRUE)),
    rmse_projected_vs_obs = sqrt(mean((proj_lambda - obs_pgr)^2,
                                      na.rm = TRUE)))

df_compare_summary

df_compare %>% print(n = 100)

# Site-level summary -----------------------------------------------------------
df_compare_site_summary <- df_compare_site %>%
  filter(
    !is.na(obs_pgr),
    !is.na(asym_lambda),
    !is.na(proj_lambda),
    obs_pgr > 0,
    asym_lambda > 0,
    proj_lambda > 0) %>%
  group_by(site) %>%
  summarise(
    n_year_transitions = n(),
    lambda_obs_geometric = exp(mean(log(obs_pgr), na.rm = TRUE)),
    lambda_obs_arithmetic = mean(obs_pgr, na.rm = TRUE),
    lambda_asymptotic_geometric = exp(mean(log(asym_lambda),
                                           na.rm = TRUE)),
    lambda_asymptotic_arithmetic = mean(asym_lambda, na.rm = TRUE),
    lambda_projected_geometric = exp(mean(log(proj_lambda),
                                          na.rm = TRUE)),
    lambda_projected_arithmetic = mean(proj_lambda, na.rm = TRUE),
    error_projected_geo_vs_obs_geo =
      lambda_projected_geometric - lambda_obs_geometric,
    rmse_projected_vs_obs = sqrt(mean((proj_lambda - obs_pgr)^2,
                                      na.rm = TRUE)),
    mean_n_initial = mean(n_obs_model, na.rm = TRUE),
    mean_disturbance = mean(disturbance_num, na.rm = TRUE),
    .groups = 'drop') %>%
  arrange(as.numeric(as.character(site)))

df_compare_site_summary %>% print(n = 100, width = Inf)
```

## Model interpretation

The final model has the following structure:

- survival and growth use current-year disturbance;
- flowering probability and flowering-stem number use disturbance in the
  preceding year;
- flowering stems in year $t_0$ predict recruits in year $t_1$;
- recruitment varies among recruitment years through a random slope;
- recruitment has no site random effect;
- site remains part of the observational and projection design; and
- the projection kernel is a $2n \times 2n$ recruit-adult block kernel.

The term “site” in this chapter title therefore refers to site-level aggregation,
initial conditions, disturbance histories, and validation. It does not indicate
an additional site random effect in the recruitment vital rate.
