# Mean IPM with fire for *Polygala lewtonii* {#mean-ipm-fire}

## What this mean IPM does

This chapter develops a deterministic mean integral projection model (IPM) for
*Polygala lewtonii* using demographic records from 2001 through 2025. The model
pools observations across years and sites to estimate average vital-rate
relationships, while retaining fire as an explicit quadrat-level predictor.
It is therefore a **mean model**: it describes expected demographic performance
under specified fire conditions rather than reconstructing the demographic
conditions of each individual year or site.

The continuous state variable is log plant volume. For a plant of size $z$ in
year $t$, the IPM projects the density of plant sizes in year $t+1$ using a
survival-growth component and a fertility component,

$$
K(z',z) = P(z',z) + F(z',z).
$$

Current-year fire affects survival, growth, and the quadrat-level conversion of
flowering stems to recruits. Fire in the previous year affects flowering
probability and flowering-stem production. The model therefore distinguishes
four fire states:

1. no fire in either year;
2. current-year fire only;
3. previous-year fire only; and
4. fire in both years.

A fifth kernel averages these four state-specific kernels using their observed
frequencies among quadrat-years. The model reports both the dominant eigenvalue
of each kernel, which is the asymptotic population growth rate, and the
one-step growth rate obtained by projecting the mean observed size
distribution.

## Main assumptions and interpretation

The model makes the following assumptions.

1. **Mean demographic relationships.** Survival, growth, flowering, and
   recruitment relationships are pooled across years and sites. No year or
   site random effects are included. Year and site are retained only for data
   construction, disturbance timing, recruitment transitions, and validation.

2. **Annual and time-homogeneous transitions.** Each kernel represents an
   annual transition under one fixed combination of current and previous-year
   fire. Within a fire state, the fitted vital-rate relationships are assumed
   not to change through time.

3. **Quadrat-level fire.** Fire exposure is assigned at the quadrat-year level.
   Current fire and previous-year fire remain separate because they affect
   different demographic processes.

4. **Size is the continuous plant state.** Individual heterogeneity other than
   size and fire exposure is not modeled explicitly.

5. **Phenomenological recruitment model.** Total flowering stems in quadrat
   $q$ during year $t$ predict recruits in the same quadrat during year $t+1$.
   Recruitment may occur when no flowering stems are recorded because the
   intercept represents emergence from the persistent seedbank. The negative
   flowering-stem coefficient, especially after fire, is interpreted as a
   possible density or establishment effect rather than as direct evidence of
   parent-specific fecundity.

6. **No explicit seedbank stage.** The persistent seedbank is represented
   indirectly by the recruitment intercept and fire response. Predicted
   quadrat recruitment is converted to recruits per flowering stem at a
   quadrat-level reference stem abundance and allocated among plants according
   to expected flowering-stem production. This keeps the kernel linear but
   does not imply known parentage.

7. **Deterministic projection.** The kernels do not include environmental
   stochasticity, demographic stochasticity, or parameter uncertainty.

8. **Observed fire-state mixture.** The expected-regime kernel uses observed
   quadrat-year frequencies as fixed weights. The fire-in-both-years state is
   retained, but it is supported by only one observed quadrat-year and must be
   interpreted cautiously.

9. **Mean initial size distribution.** The one-step projected growth rate uses
   the arithmetic mean of annual observed size-frequency vectors. Recruits
   without a usable size measurement are distributed across the fitted recruit
   size distribution.

10. **Observed-growth validation is incomplete.** Annual abundance change is
    calculated only from quadrats containing individual records in both
    consecutive years. Historically sampled but empty quadrats cannot be
    reconstructed consistently, so this comparison describes matched occupied
    quadrats rather than a complete census of all sampled quadrats.

The chapter contains the full analysis code. It does not source the completed
IPM script as a shortcut.

## Set up the analysis

```{r mean-ipm-settings}
set.seed(100)
options(stringsAsFactors = FALSE)
```


### Model specification

```{r mean-ipm-specification}
v_head <- 'archbold'
v_species <- 'Polygala lewtonii'
v_years_re <- c()

v_sp_abb <- tolower(
  gsub(
    ' ', '',
    paste(
      substr(unlist(strsplit(v_species, ' ')), 1, 2),
      collapse = '')))

v_script_prefix <- v_head
v_ggp_suffix <- paste(tools::toTitleCase(v_head), '-', v_species)

# Empty vectors select the best-supported polynomial by AICc.
v_mod_set_su <- c()
v_mod_set_gr <- c()
v_mod_set_fl <- c()
v_mod_set_fl_n <- c()
v_mod_set_re <- c()
```


## Data and metadata

```{r mean-ipm-original-data}
df_og <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_original_260519.csv')))

df_site <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_sitehist_260519.csv')))

df_meta <- data.frame(
  variable = c(
    'unique_ID', 'quad', 'patch', 'year', 'month', 'date', 'qsurv',
    'stg', 'ht', 'mcd', 'st', 'flst'),
  description = c(
    'Unique identifier for each individual plant; Numeric',
    'Name of circular quadrat with radius of 0.25m; Numeric',
    'Name of an aggregate of several quads; Numeric',
    'Year data were collected; Numeric',
    paste(
      'Month data were collected for March, June, September, and',
      'December. April was also used in the first sampling; Numeric'),
    'Combination of year and month; MMYY',
    paste(
      'Quarterly survival code. 0 or 1 refer to survival from the',
      'previous quarter. If a plant was dropped from monitoring,',
      'survival is NA and these plants should NOT be considered dead.',
      '0 = dead; 1 = alive; 2 = missing; 3 = new adult; 5 = seedling'),
    paste(
      'Stage class from March census. If plant was a seedling during',
      'the previous year it is considered a seedling here.',
      '1 = putative seedling; 2 = vegetative; 3 = reproductive adult'),
    'Height from March census in cm; Numeric',
    'Maximum crown diameter from March census in cm; Numeric',
    'Total number of stems; Numeric',
    'Number of flowering stems; Numeric'),
  stringsAsFactors = FALSE)
```

### Load the completed workdata

The workdata already contains current-year disturbance and disturbance in the
exact previous year. These fields are retained directly rather than rebuilt
from the subset of observed individual rows.

```{r mean-ipm-workdata}
df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv'))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.integer(year),
    site = factor(site)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))
```

## Construct the vital-rate data sets

### Survival

```{r mean-ipm-survival-data}
df_su <- df %>%
  filter(
    !is.na(survives),
    size_t0 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance, stage)
```

### Growth

```{r mean-ipm-growth-data}
df_gr <- df %>%
  filter(
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance, stage)
```

### Flowering probability and flowering-stem number

Observed recruits are retained because some recruits flowered.

```{r mean-ipm-flowering-data}
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance_prev)) %>%
  dplyr::select(
    id, site, year, size_t0, flower, fl_nr, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance_prev, stage, recruits)

df_fl_cond <- df_fl %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr == round(fl_nr))
```

### Quadrat-level flowering stems and recruits

Total flowering stems in quadrat $q$ during year $t$ predict recruits in the
same quadrat during year $t+1$. A quadrat transition is retained only when the
same quadrat has individual records in both years.

```{r mean-ipm-recruitment-data}
df_quad_year_sampled <- df %>%
  distinct(site, quad, year)

df_fs2r <- df %>%
  group_by(site, quad, year) %>%
  summarise(
    fs_t0 = sum(fl_nr, na.rm = TRUE),
    disturbance = if_else(
      all(is.na(disturbance)),
      NA_real_,
      max(disturbance, na.rm = TRUE)),
    .groups = 'drop') %>%
  mutate(
    year_t1 = year + 1) %>%
  semi_join(
    df_quad_year_sampled %>%
      transmute(
        site,
        quad,
        year_t1 = year),
    by = c(
      'site',
      'quad',
      'year_t1')) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(
        site,
        quad,
        year,
        name = 're_t1'),
    by = c(
      'site',
      'quad',
      'year_t1' = 'year')) %>%
  mutate(
    re_t1 = replace_na(re_t1, 0L),
    site = factor(site),
    fs_t0_log = log1p(fs_t0),
    re_t1_log = log1p(re_t1)) %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !is.na(disturbance),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re))

df_fs2r %>%
  summarise(
    quadrats = n_distinct(quad),
    transitions = n(),
    zero_stems = sum(fs_t0 == 0),
    zero_recruits = sum(re_t1 == 0),
    burned_transitions = sum(disturbance == 1),
    unburned_transitions = sum(disturbance == 0))
```

## Fit the mean vital-rate models

Candidate models use intercept-only, fire-only, and first- through third-order
size polynomials, each with and without the relevant fire term. Empty model-set
vectors select the model with the lowest AICc.

### Survival model

```{r mean-ipm-survival-model}
mod_su_00 <- glm(
  survives ~ 1,
  data = df_su,
  family = binomial)

mod_su_0 <- glm(
  survives ~ disturbance,
  data = df_su,
  family = binomial)

mod_su_10 <- glm(
  survives ~ logvol_t0,
  data = df_su,
  family = binomial)

mod_su_1 <- glm(
  survives ~ logvol_t0 + disturbance,
  data = df_su,
  family = binomial)

mod_su_20 <- glm(
  survives ~ logvol_t0 + logvol_t0_2,
  data = df_su,
  family = binomial)

mod_su_2 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance,
  data = df_su,
  family = binomial)

mod_su_30 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_su,
  family = binomial)

mod_su_3 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance,
  data = df_su,
  family = binomial)

mods_su <- list(
  mod_su_00, mod_su_0, mod_su_10, mod_su_1,
  mod_su_20, mod_su_2, mod_su_30, mod_su_3)

mods_su_dAICc <- bbmle::AICctab(
  mods_su,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_su_sorted <- order(mods_su_dAICc)

if (length(v_mod_set_su) == 0) {
  mod_su_index_best <- mods_su_sorted[1]
} else {
  mod_su_index_best <- v_mod_set_su + 1
}

v_mod_su_index <- floor((mod_su_index_best - 1) / 2)
mod_su_best <- mods_su[[mod_su_index_best]]

summary(mod_su_best)
mods_su_dAICc
```

### Growth model

```{r mean-ipm-growth-model}
mod_gr_00 <- lm(
  logvol_t1 ~ 1,
  data = df_gr)

mod_gr_0 <- lm(
  logvol_t1 ~ disturbance,
  data = df_gr)

mod_gr_10 <- lm(
  logvol_t1 ~ logvol_t0,
  data = df_gr)

mod_gr_1 <- lm(
  logvol_t1 ~ logvol_t0 + disturbance,
  data = df_gr)

mod_gr_20 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2,
  data = df_gr)

mod_gr_2 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance,
  data = df_gr)

mod_gr_30 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_gr)

mod_gr_3 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance,
  data = df_gr)

mods_gr <- list(
  mod_gr_00, mod_gr_0, mod_gr_10, mod_gr_1,
  mod_gr_20, mod_gr_2, mod_gr_30, mod_gr_3)

mods_gr_dAICc <- bbmle::AICctab(
  mods_gr,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_gr_sorted <- order(mods_gr_dAICc)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_best <- mods_gr_sorted[1]
} else {
  mod_gr_index_best <- v_mod_set_gr + 1
}

v_mod_gr_index <- floor((mod_gr_index_best - 1) / 2)
mod_gr_best <- mods_gr[[mod_gr_index_best]]

summary(mod_gr_best)
mods_gr_dAICc
```

### Growth variance

Growth variance is modeled as an exponential function of the fitted growth
mean.

```{r mean-ipm-growth-variance}
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(
    maxiter = 1000,
    tol = 1e-6,
    warnOnly = TRUE))

summary(mod_gr_var)
```

### Flowering-probability model

```{r mean-ipm-flowering-probability-model}
mod_fl_00 <- glm(
  flower ~ 1,
  data = df_fl,
  family = binomial)

mod_fl_0 <- glm(
  flower ~ disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_10 <- glm(
  flower ~ logvol_t0,
  data = df_fl,
  family = binomial)

mod_fl_1 <- glm(
  flower ~ logvol_t0 + disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_20 <- glm(
  flower ~ logvol_t0 + logvol_t0_2,
  data = df_fl,
  family = binomial)

mod_fl_2 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_30 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_fl,
  family = binomial)

mod_fl_3 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev,
  data = df_fl,
  family = binomial)

mods_fl <- list(
  mod_fl_00, mod_fl_0, mod_fl_10, mod_fl_1,
  mod_fl_20, mod_fl_2, mod_fl_30, mod_fl_3)

mods_fl_dAICc <- bbmle::AICctab(
  mods_fl,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_fl_sorted <- order(mods_fl_dAICc)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_best <- mods_fl_sorted[1]
} else {
  mod_fl_index_best <- v_mod_set_fl + 1
}

v_mod_fl_index <- floor((mod_fl_index_best - 1) / 2)
mod_fl_best <- mods_fl[[mod_fl_index_best]]

summary(mod_fl_best)
mods_fl_dAICc
```

### Flowering-stem-number model

The conditional model includes only observed flowering plants with positive
integer flowering-stem counts.

```{r mean-ipm-flowering-number-model}
mod_fl_n_00 <- MASS::glm.nb(
  fl_nr ~ 1,
  data = df_fl_cond)

mod_fl_n_0 <- MASS::glm.nb(
  fl_nr ~ disturbance_prev,
  data = df_fl_cond)

mod_fl_n_10 <- MASS::glm.nb(
  fl_nr ~ logvol_t0,
  data = df_fl_cond)

mod_fl_n_1 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_20 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2,
  data = df_fl_cond)

mod_fl_n_2 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_30 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_fl_cond)

mod_fl_n_3 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev,
  data = df_fl_cond)

mods_fl_n <- list(
  mod_fl_n_00, mod_fl_n_0, mod_fl_n_10, mod_fl_n_1,
  mod_fl_n_20, mod_fl_n_2, mod_fl_n_30, mod_fl_n_3)

mods_fl_n_dAICc <- bbmle::AICctab(
  mods_fl_n,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_fl_n_sorted <- order(mods_fl_n_dAICc)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_best <- mods_fl_n_sorted[1]
} else {
  mod_fl_n_index_best <- v_mod_set_fl_n + 1
}

v_mod_fl_n_index <- floor((mod_fl_n_index_best - 1) / 2)
mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_best]]

summary(mod_fl_n_best)
mods_fl_n_dAICc
```

### Flowering-stem-to-recruit model

The recruitment candidates allow a seedbank intercept, a current-fire effect,
a flowering-stem effect, and their interaction. No year or site random effects
are fitted.

```{r mean-ipm-recruitment-model}
mod_re_00 <- lm(
  re_t1_log ~ 1,
  data = df_fs2r)

mod_re_0 <- lm(
  re_t1_log ~ disturbance,
  data = df_fs2r)

mod_re_10 <- lm(
  re_t1_log ~ fs_t0_log,
  data = df_fs2r)

mod_re_1 <- lm(
  re_t1_log ~ fs_t0_log + disturbance,
  data = df_fs2r)

mod_re_20 <- lm(
  re_t1_log ~ fs_t0_log * disturbance,
  data = df_fs2r)

mods_re <- list(
  mod_re_00,
  mod_re_0,
  mod_re_10,
  mod_re_1,
  mod_re_20)

mods_re_dAICc <- bbmle::AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_re_sorted <- order(mods_re_dAICc)

if (length(v_mod_set_re) == 0) {
  mod_re_index_best <- mods_re_sorted[1]
} else {
  mod_re_index_best <- v_mod_set_re + 1
}

v_mod_re_index <- mod_re_index_best - 1
mod_re_best <- mods_re[[mod_re_index_best]]

summary(mod_re_best)
mods_re_dAICc
```

### Recruitment-model support

```{r mean-ipm-recruitment-support}
df_recruit_model_support <- tibble(
  model = c(
    'Intercept only',
    'Disturbance only',
    'Flowering stems',
    'Flowering stems + disturbance',
    'Full interaction'),
  dAICc = as.numeric(mods_re_dAICc)) %>%
  mutate(
    weight = exp(-0.5 * dAICc) /
      sum(exp(-0.5 * dAICc))) %>%
  arrange(dAICc)

knitr::kable(
  df_recruit_model_support,
  digits = c(NA, 2, 3),
  caption = paste(
    'AICc support for quadrat-level flowering-stem-to-recruit',
    'models.'))
```

The full interaction model is selected. Its fitted form is

$$
\log(1 + R_{q,t+1}) =
\beta_0 + \beta_1\log(1 + M_{q,t}) + \beta_d d_{q,t}
+ \beta_{1d}\log(1 + M_{q,t})d_{q,t} + \varepsilon,
$$

where $M_{q,t}$ is total flowering stems, $R_{q,t+1}$ is the number of recruits
in the same quadrat in the following year, and $d_{q,t}$ indicates current
fire.

### Recruitment slopes and uncertainty

```{r mean-ipm-recruitment-slopes}
re_coef <- coef(mod_re_20)
re_vcov <- vcov(mod_re_20)

slope_unburned <- re_coef['fs_t0_log']

se_unburned <- sqrt(
  re_vcov[
    'fs_t0_log',
    'fs_t0_log'])

slope_burned <-
  re_coef['fs_t0_log'] +
  re_coef['fs_t0_log:disturbance']

se_burned <- sqrt(
  re_vcov[
    'fs_t0_log',
    'fs_t0_log'] +
  re_vcov[
    'fs_t0_log:disturbance',
    'fs_t0_log:disturbance'] +
  2 * re_vcov[
    'fs_t0_log',
    'fs_t0_log:disturbance'])

df_recruit_slopes <- tibble(
  disturbance = c(0, 1),
  slope = c(
    slope_unburned,
    slope_burned),
  lower_95 = c(
    slope_unburned - 1.96 * se_unburned,
    slope_burned - 1.96 * se_burned),
  upper_95 = c(
    slope_unburned + 1.96 * se_unburned,
    slope_burned + 1.96 * se_burned))

knitr::kable(
  df_recruit_slopes,
  digits = 3,
  caption = paste(
    'Flowering-stem slopes from the full interaction recruitment',
    'model.'))
```

Recruitment can occur when no flowering stems are observed. This is consistent
with emergence from the persistent seedbank.

```{r mean-ipm-zero-stem-recruitment}
df_fs2r_zero_stems <- df_fs2r %>%
  filter(fs_t0 == 0) %>%
  group_by(disturbance) %>%
  summarise(
    transitions = n(),
    transitions_with_recruits = sum(re_t1 > 0),
    proportion_with_recruits = mean(re_t1 > 0),
    total_recruits = sum(re_t1),
    mean_recruits = mean(re_t1),
    median_recruits = median(re_t1),
    max_recruits = max(re_t1),
    .groups = 'drop')

knitr::kable(
  df_fs2r_zero_stems,
  digits = 3,
  caption = paste(
    'Recruitment in quadrat transitions with no recorded flowering',
    'stems.'))
```

## Plot the selected vital-rate models

```{r mean-ipm-plot-functions}
make_mean_prediction <- function(data, model, disturbance_var) {
  x <- seq(
    min(data$logvol_t0, na.rm = TRUE),
    max(data$logvol_t0, na.rm = TRUE),
    length.out = 200)

  pred <- expand_grid(
    logvol_t0 = x,
    disturbance_value = c(0, 1)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_plot = factor(
        disturbance_value,
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  pred[[disturbance_var]] <- pred$disturbance_value
  pred$prediction <- predict(
    model,
    newdata = pred,
    type = 'response')

  pred
}

make_mean_plot <- function(
    data, model, response, disturbance_var, title, y_lab,
    probability = FALSE) {
  pred <- make_mean_prediction(
    data = data,
    model = model,
    disturbance_var = disturbance_var)

  data_plot <- data %>%
    mutate(
      disturbance_plot = factor(
        .data[[disturbance_var]],
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  point_layer <- if (probability) {
    geom_jitter(
      data = data_plot,
      aes(
        x = logvol_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7,
      width = 0.05,
      height = 0.05)
  } else {
    geom_point(
      data = data_plot,
      aes(
        x = logvol_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7)
  }

  p <- ggplot() +
    point_layer +
    geom_line(
      data = pred,
      aes(
        x = logvol_t0,
        y = prediction,
        color = disturbance_plot),
      linewidth = 0.8) +
    scale_color_manual(
      values = c(
        'No fire' = 'black',
        'Fire' = 'red')) +
    theme_bw() +
    labs(
      title = title,
      subtitle = v_ggp_suffix,
      x = expression('log(volume)'[t0]),
      y = y_lab,
      color = NULL)

  if (probability) {
    p <- p + coord_cartesian(ylim = c(0, 1))
  }

  p
}
```

```{r mean-ipm-vital-rate-figures, fig.width=9, fig.height=7, fig.cap='Selected mean survival, growth, flowering-probability, and conditional flowering-stem models. Current fire is used for survival and growth, and previous-year fire is used for flowering.'}
fig_su_mean <- make_mean_plot(
  data = df_su,
  model = mod_su_best,
  response = 'survives',
  disturbance_var = 'disturbance',
  title = 'Mean survival',
  y_lab = 'Survival probability',
  probability = TRUE)

fig_gr_mean <- make_mean_plot(
  data = df_gr,
  model = mod_gr_best,
  response = 'logvol_t1',
  disturbance_var = 'disturbance',
  title = 'Mean growth',
  y_lab = expression('log(volume)'[t1])) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = 'dashed')

fig_fl_mean <- make_mean_plot(
  data = df_fl,
  model = mod_fl_best,
  response = 'flower',
  disturbance_var = 'disturbance_prev',
  title = 'Mean flowering probability',
  y_lab = 'Flowering probability',
  probability = TRUE)

fig_fl_n_mean <- make_mean_plot(
  data = df_fl_cond,
  model = mod_fl_n_best,
  response = 'fl_nr',
  disturbance_var = 'disturbance_prev',
  title = 'Mean flowering-stem number',
  y_lab = 'Number of flowering stems')

fig_vital_rates_mean <-
  fig_su_mean + fig_gr_mean + fig_fl_mean + fig_fl_n_mean +
  plot_layout(ncol = 2, guides = 'collect') &
  theme(legend.position = 'bottom')

fig_vital_rates_mean
```

### Flowering-stem-to-recruit relationship

```{r mean-ipm-recruitment-figure, fig.width=7, fig.height=5, fig.cap='Quadrat-level flowering stems in year t and recruits in the same quadrat in year t + 1 under current-fire and no-fire conditions.'}
df_fs2r_pred <- expand_grid(
  fs_t0_log = seq(
    0,
    max(df_fs2r$fs_t0_log, na.rm = TRUE),
    length.out = 200),
  disturbance = c(0, 1)) %>%
  mutate(
    re_t1_log = predict(
      mod_re_best,
      newdata = .),
    fs_t0 = expm1(fs_t0_log),
    re_t1 = pmax(0, expm1(re_t1_log)),
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c('No fire', 'Fire')))

df_fs2r_plot <- df_fs2r %>%
  mutate(
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c('No fire', 'Fire')))

fig_fs2r_mean <- ggplot(
  df_fs2r_plot,
  aes(
    x = fs_t0,
    y = re_t1,
    color = disturbance_plot)) +
  geom_point(alpha = 0.4, size = 1.3) +
  geom_line(
    data = df_fs2r_pred,
    aes(
      x = fs_t0,
      y = re_t1,
      color = disturbance_plot),
    linewidth = 0.8) +
  scale_x_continuous(
  transform = scales::transform_pseudo_log(
    sigma = 1)) +
  scale_color_manual(
    values = c(
      'No fire' = 'black',
      'Fire' = 'red')) +
  theme_bw() +
  labs(
    title = 'Mean flowering-stem-to-recruit relationship',
    subtitle = v_ggp_suffix,
    x = expression('Total flowering stems'[t]),
    y = expression('Recruits'[t + 1]),
    color = NULL)

fig_fs2r_mean
```

## Recruit size distribution

```{r mean-ipm-recruit-size}
df_recr <- df %>%
  filter(
    recruits == 1,
    volume_t0 > 0,
    is.finite(logvol_t0))

rc_sz <- df_recr %>%
  summarise(
    mean = mean(logvol_t0, na.rm = TRUE),
    sd = sd(logvol_t0, na.rm = TRUE))

rc_sz
```

## Extract the selected model parameters

The extraction functions return zero for terms that are absent from the
selected model. The same IPM functions can therefore use any selected candidate
without manual coefficient changes.

```{r mean-ipm-parameter-extraction}
size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logvol_t0',
  b2 = 'logvol_t0_2',
  b3 = 'logvol_t0_3',
  bd = 'disturbance')

lagged_size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logvol_t0',
  b2 = 'logvol_t0_2',
  b3 = 'logvol_t0_3',
  bd = 'disturbance_prev')

extract_terms <- function(model, prefix, term_map) {
  coefs <- stats::coef(model)

  values <- vapply(term_map, function(term_i) {
    if (term_i %in% names(coefs)) {
      unname(coefs[[term_i]])
    } else {
      0
    }
  }, numeric(1))

  names(values) <- paste0(prefix, names(term_map))
  values
}

extract_coef <- function(model, term, default = 0) {
  coefs <- stats::coef(model)

  if (term %in% names(coefs)) {
    unname(coefs[[term]])
  } else {
    default
  }
}

grow_var_coef <- stats::coef(mod_gr_var)

mesh_limits <- range(
  c(
    df_gr$logvol_t0,
    df_gr$logvol_t1,
    df_recr$logvol_t0),
  na.rm = TRUE,
  finite = TRUE)

mesh_limits <- c(
  mesh_limits[1] - 0.1,
  mesh_limits[2] + 0.1)

pars_mean <- as.list(c(
  extract_terms(
    mod_su_best,
    'surv_',
    size_term_map),
  extract_terms(
    mod_gr_best,
    'grow_',
    size_term_map),
  extract_terms(
    mod_fl_best,
    'fl_',
    lagged_size_term_map),
  extract_terms(
    mod_fl_n_best,
    'fln_',
    lagged_size_term_map),
  re_b0 = extract_coef(
    mod_re_best,
    '(Intercept)'),
  re_b1 = extract_coef(
    mod_re_best,
    'fs_t0_log'),
  re_bd = extract_coef(
    mod_re_best,
    'disturbance'),
  re_bf = extract_coef(
    mod_re_best,
    'fs_t0_log:disturbance'),
  re_sigma = stats::sigma(mod_re_best),
  recr_mean = rc_sz$mean,
  recr_sd = rc_sz$sd,
  grow_var_a = unname(grow_var_coef[['a']]),
  grow_var_b = unname(grow_var_coef[['b']]),
  fs_ref = mean(df_fs2r$fs_t0, na.rm = TRUE),
  L = mesh_limits[1],
  U = mesh_limits[2],
  mat_siz = 200,
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_fl_index = v_mod_fl_index,
  mod_fl_n_index = v_mod_fl_n_index,
  mod_re_index = v_mod_re_index))

pars_mean_table <- tibble(
  parameter = names(pars_mean),
  value = as.numeric(unlist(pars_mean)))

knitr::kable(
  pars_mean_table,
  digits = 5,
  caption = 'Parameters passed to the mean IPM.')
```

## Define the IPM functions

### General parameter and linear-predictor functions

```{r mean-ipm-general-functions}
get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    pars[[par_name]]
  } else {
    default
  }
}

vital_lp <- function(x, pars, prefix, disturbance = 0) {
  get_par(pars, paste0(prefix, 'b0')) +
    get_par(pars, paste0(prefix, 'b1')) * x +
    get_par(pars, paste0(prefix, 'b2')) * x^2 +
    get_par(pars, paste0(prefix, 'b3')) * x^3 +
    get_par(pars, paste0(prefix, 'bd')) * disturbance
}
```

### Survival

```{r mean-ipm-survival-function}
sx <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'surv_',
    disturbance = disturbance))
}
```

### Growth

```{r mean-ipm-growth-functions}
grow_mu <- function(x, pars, disturbance = 0) {
  vital_lp(
    x = x,
    pars = pars,
    prefix = 'grow_',
    disturbance = disturbance)
}

grow_sd <- function(mean_value, pars) {
  variance <- get_par(pars, 'grow_var_a') *
    exp(get_par(pars, 'grow_var_b') * mean_value)

  sqrt(pmax(variance, .Machine$double.eps))
}

gxy <- function(x, y, pars, disturbance = 0) {
  mean_value <- grow_mu(
    x = x,
    pars = pars,
    disturbance = disturbance)

  sd_value <- grow_sd(
    mean_value = mean_value,
    pars = pars)

  dnorm(
    y,
    mean = mean_value,
    sd = sd_value)
}
```

### Flowering probability and conditional flowering-stem number

```{r mean-ipm-flowering-functions}
fl_x <- function(x, pars, disturbance_prev = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fl_',
    disturbance = disturbance_prev))
}

fln_x <- function(x, pars, disturbance_prev = 0) {
  exp(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fln_',
    disturbance = disturbance_prev))
}

fs_x <- function(x, pars, disturbance_prev = 0) {
  fl_x(
    x = x,
    pars = pars,
    disturbance_prev = disturbance_prev) *
    fln_x(
      x = x,
      pars = pars,
      disturbance_prev = disturbance_prev)
}
```

### Flowering-stem-to-recruit conversion

The coefficient extraction supports intercept-only, disturbance-only,
flowering-stem, additive, and interaction models. Recruitment is not forced to
zero when flowering stems equal zero because the fitted intercept represents
seedbank emergence.

```{r mean-ipm-recruitment-functions}
predict_recruits_from_stems <- function(
    flowering_stems, pars, disturbance = 0) {
  stems_nonnegative <- pmax(flowering_stems, 0)

  slope <- get_par(pars, 're_b1') +
    get_par(pars, 're_bf') * disturbance

  eta <- get_par(pars, 're_b0') +
    get_par(pars, 're_bd') * disturbance +
    slope * log1p(stems_nonnegative)

  recruits <- expm1(
    eta + 0.5 * get_par(pars, 're_sigma')^2)

  pmax(recruits, 0)
}

recruits_per_stem <- function(pars, disturbance = 0) {
  fs_ref <- get_par(pars, 'fs_ref', 0)

  if (!is.finite(fs_ref) || fs_ref <= 0) {
    return(0)
  }

  predict_recruits_from_stems(
    flowering_stems = fs_ref,
    pars = pars,
    disturbance = disturbance) / fs_ref
}

rx <- function(
    x, pars, disturbance = 0, disturbance_prev = 0) {
  fs_x(
    x = x,
    pars = pars,
    disturbance_prev = disturbance_prev) *
    recruits_per_stem(
      pars = pars,
      disturbance = disturbance)
}
```

The fitted model predicts nonzero recruitment from the seedbank when no current
flowering stems are recorded.

```{r mean-ipm-zero-stem-predictions}
tibble(
  disturbance = c(0, 1),
  predicted_recruits_zero_stems =
    predict_recruits_from_stems(
      flowering_stems = 0,
      pars = pars_mean,
      disturbance = disturbance))
```

### Recruit size distribution and fertility matrix

```{r mean-ipm-fertility-functions}
recr_y <- function(y, pars) {
  dnorm(
    y,
    mean = get_par(pars, 'recr_mean'),
    sd = get_par(pars, 'recr_sd'))
}

fy <- function(
    y, x, pars, h, disturbance = 0, disturbance_prev = 0) {
  Pvec <- recr_y(y, pars)
  recruit_mass <- sum(Pvec * h)

  if (!is.finite(recruit_mass) || recruit_mass <= 0) {
    stop('Recruit size distribution has no mass inside the mesh')
  }

  Pvec <- Pvec / recruit_mass

  Rvec <- rx(
    x = x,
    pars = pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)

  outer(Pvec, Rvec) * h
}
```

### Full kernel

```{r mean-ipm-kernel-function}
kernel <- function(
    pars, disturbance = 0, disturbance_prev = 0) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  Fmat <- fy(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)

  Smat <- sx(
    x = y,
    pars = pars,
    disturbance = disturbance)

  Gmat <- matrix(0, n, n)
  Gmat[] <- t(
    outer(
      y,
      y,
      Vectorize(function(x, y) {
        gxy(
          x = x,
          y = y,
          pars = pars,
          disturbance = disturbance)
      }))) * h

  Tmat <- matrix(0, n, n)

  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  k_yx <- Fmat + Tmat

  list(
    k_yx = k_yx,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Smat = Smat,
    meshpts = y,
    h = h)
}

lambda_ipm <- function(
    pars, disturbance = 0, disturbance_prev = 0) {
  K <- kernel(
    pars = pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)$k_yx

  Re(eigen(K, only.values = TRUE)$values[1])
}
```

## Construct the mean observed initial size distribution

The initial vector is the arithmetic mean of annual size-frequency vectors.
Unsized recruits are distributed over the fitted recruit-size distribution.

```{r mean-ipm-initial-distribution}
make_initial_n_mean <- function(pars, df_init = df) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  breaks <- seq(L, U, length.out = n + 1)
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  recruit_density <- recr_y(y, pars)
  recruit_density <- recruit_density /
    sum(recruit_density * h)

  years_i <- df_init %>%
    filter(!is.na(year)) %>%
    distinct(year) %>%
    arrange(year) %>%
    pull(year)

  counts_by_year <- purrr::map(years_i, function(year_i) {
    df_i <- df_init %>%
      filter(year == year_i)

    x <- df_i %>%
      filter(is.finite(logvol_t0)) %>%
      pull(logvol_t0)

    size_counts <- hist(
      pmin(pmax(x, L), U),
      breaks = breaks,
      plot = FALSE,
      include.lowest = TRUE)$counts

    n_unsized_recruits <- df_i %>%
      summarise(
        n = sum(
          !is.finite(logvol_t0) & recruits == 1,
          na.rm = TRUE)) %>%
      pull(n)

    size_counts +
      n_unsized_recruits * recruit_density * h
  })

  mean_counts <- Reduce(`+`, counts_by_year) /
    length(counts_by_year)

  mean_counts / h
}
```

## Quadrat-level flowering-stem reference

The recruitment model is fitted to flowering-stem totals per quadrat. The
reference stem abundance used by the fertility conversion must therefore also
be on the quadrat scale.

```{r mean-ipm-stem-reference}
df_fs2r_previous_fire <- df_fs2r %>%
  left_join(
    df %>%
      distinct(
        site,
        quad,
        year,
        disturbance_prev),
    by = c(
      'site',
      'quad',
      'year')) %>%
  filter(!is.na(disturbance_prev))

df_stem_reference <- df_fs2r_previous_fire %>%
  group_by(disturbance_prev) %>%
  summarise(
    fs_ref = mean(fs_t0),
    n_quadrat_transitions = n(),
    .groups = 'drop')

get_stem_reference <- function(disturbance_prev_i) {
  df_stem_reference %>%
    filter(
      disturbance_prev == disturbance_prev_i) %>%
    pull(fs_ref)
}

mean_stems_no_previous_fire <- get_stem_reference(0)
mean_stems_previous_fire <- get_stem_reference(1)

knitr::kable(
  df_stem_reference,
  digits = 3,
  caption = paste(
    'Mean quadrat-level flowering-stem abundance by previous-year',
    'fire state.'))
```

## Observed disturbance regime

Each unique quadrat-year receives equal weight. The four combinations preserve
the distinction between current fire and fire in the previous year.

```{r mean-ipm-disturbance-regime}
df_disturbance_regime <- df %>%
  filter(
    !is.na(quad),
    !is.na(year),
    !is.na(disturbance),
    !is.na(disturbance_prev)) %>%
  distinct(
    quad,
    year,
    disturbance,
    disturbance_prev) %>%
  count(
    disturbance,
    disturbance_prev,
    name = 'n_quadrat_years') %>%
  complete(
    disturbance = c(0, 1),
    disturbance_prev = c(0, 1),
    fill = list(n_quadrat_years = 0)) %>%
  mutate(
    p_state = n_quadrat_years /
      sum(n_quadrat_years)) %>%
  arrange(
    disturbance,
    disturbance_prev)

knitr::kable(
  df_disturbance_regime,
  digits = c(0, 0, 0, 4),
  caption = 'Observed quadrat-year frequencies of the four fire states.')

get_state_probability <- function(
    disturbance_i, disturbance_prev_i) {
  df_disturbance_regime %>%
    filter(
      disturbance == disturbance_i,
      disturbance_prev == disturbance_prev_i) %>%
    pull(p_state)
}

p_00 <- get_state_probability(0, 0)
p_01 <- get_state_probability(0, 1)
p_10 <- get_state_probability(1, 0)
p_11 <- get_state_probability(1, 1)

p_previous_fire <- p_01 + p_11

mean_stems_regime <-
  (1 - p_previous_fire) * mean_stems_no_previous_fire +
  p_previous_fire * mean_stems_previous_fire

tibble(
  reference = c(
    'No previous fire',
    'Previous fire',
    'Observed disturbance regime'),
  flowering_stems = c(
    mean_stems_no_previous_fire,
    mean_stems_previous_fire,
    mean_stems_regime))
```

## Build the state-specific and expected-regime kernels

Each state-specific kernel uses the flowering-stem reference corresponding to
its previous-year-fire state. All components of the expected-regime kernel use
the weighted reference stem abundance for the observed disturbance regime.

```{r mean-ipm-state-kernels}
pars_00 <- pars_mean
pars_00$fs_ref <- mean_stems_no_previous_fire

pars_01 <- pars_mean
pars_01$fs_ref <- mean_stems_previous_fire

pars_10 <- pars_mean
pars_10$fs_ref <- mean_stems_no_previous_fire

pars_11 <- pars_mean
pars_11$fs_ref <- mean_stems_previous_fire

pars_regime <- pars_mean
pars_regime$fs_ref <- mean_stems_regime

K_00 <- kernel(
  pars = pars_00,
  disturbance = 0,
  disturbance_prev = 0)$k_yx

K_01 <- kernel(
  pars = pars_01,
  disturbance = 0,
  disturbance_prev = 1)$k_yx

K_10 <- kernel(
  pars = pars_10,
  disturbance = 1,
  disturbance_prev = 0)$k_yx

K_11 <- kernel(
  pars = pars_11,
  disturbance = 1,
  disturbance_prev = 1)$k_yx

K_regime_00 <- kernel(
  pars = pars_regime,
  disturbance = 0,
  disturbance_prev = 0)$k_yx

K_regime_01 <- kernel(
  pars = pars_regime,
  disturbance = 0,
  disturbance_prev = 1)$k_yx

K_regime_10 <- kernel(
  pars = pars_regime,
  disturbance = 1,
  disturbance_prev = 0)$k_yx

K_regime_11 <- kernel(
  pars = pars_regime,
  disturbance = 1,
  disturbance_prev = 1)$k_yx

K_regime <-
  p_00 * K_regime_00 +
  p_01 * K_regime_01 +
  p_10 * K_regime_10 +
  p_11 * K_regime_11
```

## Calculate asymptotic and projected population growth

The asymptotic growth rate is the dominant eigenvalue of the kernel. The
projected growth rate is the one-step change from the mean observed size
distribution.

```{r mean-ipm-projection-functions}
project_kernel <- function(K, pars, label, df_init = df) {
  n_obs <- make_initial_n_mean(pars, df_init)
  h <- (pars$U - pars$L) / pars$mat_siz
  n_initial <- sum(n_obs) * h

  n_proj <- K %*% n_obs
  n_projected <- sum(n_proj) * h

  tibble(
    lambda_type = label,
    asym_lambda = Re(
      eigen(
        K,
        only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial,
    n_initial = n_initial,
    n_projected = n_projected)
}

make_state_lambda <- function(
    K, pars, label, disturbance, disturbance_prev, p_state) {
  project_kernel(
    K = K,
    pars = pars,
    label = label) %>%
    mutate(
      disturbance = disturbance,
      disturbance_prev = disturbance_prev,
      observed_state_probability = p_state,
      flowering_stems_ref = pars$fs_ref,
      recruits_ref = predict_recruits_from_stems(
        flowering_stems = pars$fs_ref,
        pars = pars,
        disturbance = disturbance),
      recruits_per_stem = if_else(
        flowering_stems_ref > 0,
        recruits_ref / flowering_stems_ref,
        0))
}
```

```{r mean-ipm-lambda-results}
df_lambda_states <- bind_rows(
  make_state_lambda(
    K = K_00,
    pars = pars_00,
    label = 'Undisturbed',
    disturbance = 0,
    disturbance_prev = 0,
    p_state = p_00),
  make_state_lambda(
    K = K_10,
    pars = pars_10,
    label = 'Current-year fire',
    disturbance = 1,
    disturbance_prev = 0,
    p_state = p_10),
  make_state_lambda(
    K = K_01,
    pars = pars_01,
    label = 'Previous-year fire',
    disturbance = 0,
    disturbance_prev = 1,
    p_state = p_01),
  make_state_lambda(
    K = K_11,
    pars = pars_11,
    label = 'Fire in both years',
    disturbance = 1,
    disturbance_prev = 1,
    p_state = p_11))

recruits_ref_regime <-
  (p_00 + p_01) * predict_recruits_from_stems(
    flowering_stems = mean_stems_regime,
    pars = pars_regime,
    disturbance = 0) +
  (p_10 + p_11) * predict_recruits_from_stems(
    flowering_stems = mean_stems_regime,
    pars = pars_regime,
    disturbance = 1)

df_lambda_regime <- project_kernel(
  K = K_regime,
  pars = pars_regime,
  label = 'Expected disturbance regime') %>%
  mutate(
    disturbance = NA_real_,
    disturbance_prev = NA_real_,
    observed_state_probability = 1,
    flowering_stems_ref = mean_stems_regime,
    recruits_ref = recruits_ref_regime,
    recruits_per_stem = if_else(
      flowering_stems_ref > 0,
      recruits_ref / flowering_stems_ref,
      0))

df_lambda_mean <- bind_rows(
  df_lambda_states,
  df_lambda_regime)

knitr::kable(
  df_lambda_mean,
  digits = 3,
  caption = paste(
    'Asymptotic and one-step projected population growth under the',
    'four fire states and the observed disturbance regime.'))
```

The fire-in-both-years row is retained for completeness, but its independent
interpretation is not supported because only one quadrat-year contributed to
that state.

## Observed annual abundance change in matched occupied quadrats

Counts include sized plants and recruits without a usable size measurement.
Only quadrats with individual records in both consecutive years are retained.
Historically sampled empty quadrats cannot be added because they have no
individual rows in the workdata.

```{r mean-ipm-observed-growth}
df_counts_quad_year <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year)) %>%
  group_by(site, quad, year) %>%
  summarise(
    n_sized = sum(
      is.finite(logvol_t0),
      na.rm = TRUE),
    n_unsized_recruits = sum(
      !is.finite(logvol_t0) & recruits == 1,
      na.rm = TRUE),
    n_t0 = n_sized + n_unsized_recruits,
    .groups = 'drop') %>%
  arrange(site, quad, year) %>%
  group_by(site, quad) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n_t0),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_obs_growth_year <- df_counts_quad_year %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(n_t0),
    n_t1 = sum(n_t1),
    n_quadrats = n(),
    n_sites = n_distinct(site),
    obs_lambda = n_t1 / n_t0,
    .groups = 'drop') %>%
  filter(n_t0 > 0)

df_obs_growth_year
```

```{r mean-ipm-observed-growth-summary}
df_obs_growth_summary <- df_obs_growth_year %>%
  summarise(
    n_transitions = n(),
    lambda_arithmetic = mean(
      obs_lambda,
      na.rm = TRUE),
    lambda_geometric = exp(
      mean(
        log(obs_lambda),
        na.rm = TRUE)),
    lambda_median = median(
      obs_lambda,
      na.rm = TRUE),
    lambda_min = min(
      obs_lambda,
      na.rm = TRUE),
    lambda_max = max(
      obs_lambda,
      na.rm = TRUE))

knitr::kable(
  df_obs_growth_summary,
  digits = 3,
  caption = paste(
    'Summary of annual abundance change in matched occupied',
    'quadrats.'))
```

## Compare observed and modeled population growth

```{r mean-ipm-growth-comparison-figure, fig.width=9, fig.height=5, fig.cap='Annual abundance change in matched occupied quadrats and asymptotic and one-step projected growth rates from the mean IPM.'}
df_model_lines <- df_lambda_mean %>%
  dplyr::select(
    lambda_type,
    asym_lambda,
    proj_lambda) %>%
  pivot_longer(
    cols = c(
      asym_lambda,
      proj_lambda),
    names_to = 'metric',
    values_to = 'lambda') %>%
  mutate(
    metric = recode(
      metric,
      asym_lambda = 'Asymptotic',
      proj_lambda = 'Projected'))

x_point <- max(
  df_obs_growth_year$year,
  na.rm = TRUE) + 0.4

fig_lambda_mean <- ggplot(
  df_obs_growth_year,
  aes(
    x = year,
    y = obs_lambda)) +
  geom_hline(
    yintercept = 1,
    linetype = 'dotted') +
  geom_line(color = 'black') +
  geom_point(
    size = 2,
    color = 'black') +
  geom_hline(
    data = df_model_lines,
    aes(
      yintercept = lambda,
      color = lambda_type),
    linewidth = 0.8) +
  geom_point(
    data = df_model_lines,
    aes(
      x = x_point,
      y = lambda,
      color = lambda_type,
      shape = metric),
    size = 3) +
  expand_limits(x = x_point + 1) +
  theme_bw() +
  labs(
    title = 'Observed annual growth and mean-IPM lambda',
    subtitle = v_ggp_suffix,
    x = 'Year t0',
    y = expression(lambda),
    color = NULL,
    shape = NULL)

fig_lambda_mean
```

## Kernel diagnostics

```{r mean-ipm-kernel-diagnostics}
df_kernel_diagnostics <- tibble(
  kernel = c(
    'Undisturbed',
    'Current-year fire',
    'Previous-year fire',
    'Fire in both years',
    'Expected disturbance regime'),
  min_entry = c(
    min(K_00),
    min(K_10),
    min(K_01),
    min(K_11),
    min(K_regime)),
  max_entry = c(
    max(K_00),
    max(K_10),
    max(K_01),
    max(K_11),
    max(K_regime)),
  any_nonfinite = c(
    any(!is.finite(K_00)),
    any(!is.finite(K_10)),
    any(!is.finite(K_01)),
    any(!is.finite(K_11)),
    any(!is.finite(K_regime))))

knitr::kable(
  df_kernel_diagnostics,
  digits = c(NA, 6, 6, NA),
  caption = paste(
    'Numerical diagnostics for the state-specific and expected-regime',
    'kernels.'))
```

All kernel entries should be finite and non-negative. These checks must be
repeated whenever the vital-rate models, mesh limits, recruitment conversion,
or fire-state weighting are changed.

## Interpretation of the mean model

The fitted model represents a fire-responsive perennial with a persistent
seedbank. Current fire increases observed recruitment but can simultaneously
reduce the performance of established plants. The strongest positive response
occurs in the year following fire, when flowering and the resulting fertility
contribution reflect previous-year disturbance.

The flowering-stem-to-recruit relationship is a quadrat-level relationship.
Recruitment in quadrats without current flowering is attributed to the
persistent seedbank. The negative flowering-stem relationship, especially in
burned quadrats, is consistent with fewer open establishment sites where many
productive adults are already present, while additional seed production may
feed the seedbank instead of appearing immediately as recruits. This remains a
phenomenological interpretation; the model does not identify parent plants or
separate seed production, seed survival, germination, and establishment.

The expected-regime kernel combines the observed frequencies of the four fire
states. It is therefore a deterministic summary of average demographic
performance under the observed disturbance regime, not a stochastic fire model
and not a simulation of a specific historical fire sequence.

The comparison with observed annual abundance change is useful but limited.
The workdata contain individual records, so quadrats sampled with zero plants
are absent from the historical series. The observed geometric mean therefore
describes matched occupied quadrats and may omit some local extinction and
recolonization events.

## Save key outputs

The following commands are retained from the analysis script and can be
uncommented when the outputs should be written to disk.

```{r mean-ipm-save-outputs, eval=FALSE}
# saveRDS(
#   pars_mean,
#   file.path(
#     dir_result,
#     'pole_mean_fire_pars.rds'))

# saveRDS(
#   df_lambda_mean,
#   file.path(
#     dir_result,
#     'pole_mean_fire_lambda.rds'))

# saveRDS(
#   df_obs_growth_year,
#   file.path(
#     dir_result,
#     'pole_mean_fire_observed_growth.rds'))
```
