# Mean IPM with one disturbance state for *Polygala lewtonii* {#mean-ipm-dist1}

## What this mean IPM does

This chapter develops the same deterministic mean integral projection model
(IPM) for *Polygala lewtonii* as Chapter 9, but simplifies how fire is handled
inside the IPM. The vital-rate models themselves are fitted exactly as before:
survival and growth use current disturbance, flowering probability and
flowering-stem number use previous-year disturbance, and quadrat-level
recruitment uses current disturbance.

The change occurs only after the vital-rate models have been fitted and their
coefficients have been extracted. From that point onward, the IPM uses one
shared binary disturbance state for every demographic process. In other words,
a kernel is either evaluated with `disturbance = 0` for all vital rates or with
`disturbance = 1` for all vital rates. The coefficients fitted to
`disturbance_prev` in the flowering models are therefore deliberately treated
as if they belonged to the same disturbance level as the other vital rates.

The continuous state variable is log plant volume. For a plant of size $z$ in
year $t$, the IPM projects the density of plant sizes in year $t+1$ using a
survival-growth component and a fertility component,

$$
K(z',z) = P(z',z) + F(z',z).
$$

This simplified IPM therefore has only two disturbance-specific kernels:

1. an undisturbed kernel, in which all vital rates receive `disturbance = 0`;
2. a fire kernel, in which all vital rates receive `disturbance = 1`.

A third kernel averages these two kernels using the observed quadrat-year
frequency of the shared disturbance variable. The model reports both the
dominant eigenvalue of each kernel, which is the asymptotic population growth
rate, and the one-step growth rate obtained by projecting the mean observed
size distribution.

## Main assumptions and interpretation

The model makes the following assumptions.

1. **Mean demographic relationships.** Survival, growth, flowering, and
   recruitment relationships are pooled across years and sites. No year or
   site random effects are included. Year and site are retained only for data
   construction, disturbance timing, recruitment transitions, and validation.

2. **Annual and time-homogeneous transitions.** Each kernel represents an
   annual transition under one fixed disturbance state. Within a disturbance
   state, the fitted vital-rate relationships are assumed not to change through
   time.

3. **The fitted disturbance timing is unchanged.** Survival and growth are
   fitted with current disturbance, flowering probability and flowering-stem
   number are fitted with previous-year disturbance, and the flowering-stem-to-
   recruit model is fitted with current disturbance. No disturbance variable is
   changed in the model-fitting stage.

4. **One shared disturbance state in the IPM.** After coefficient extraction,
   the distinction between current and previous-year disturbance is discarded.
   The same binary `disturbance` value is passed to survival, growth,
   flowering, flowering-stem production, and recruitment. This is a deliberate
   simplification for the IPM and should not be interpreted as preserving the
   observed timing of fire effects.

5. **Size is the continuous plant state.** Individual heterogeneity other than
   size and disturbance exposure is not modeled explicitly.

6. **Phenomenological recruitment model.** Total flowering stems in quadrat
   $q$ during year $t$ predict recruits in the same quadrat during year $t+1$.
   Recruitment may occur when no flowering stems are recorded because the
   intercept represents emergence from the persistent seedbank. The negative
   flowering-stem coefficient, especially after fire, is interpreted as a
   possible density or establishment effect rather than as direct evidence of
   parent-specific fecundity.

7. **No explicit seedbank stage.** The persistent seedbank is represented
   indirectly by the recruitment intercept and disturbance response. Predicted
   quadrat recruitment is converted to recruits per flowering stem at a
   quadrat-level reference stem abundance and allocated among plants according
   to expected flowering-stem production. This keeps the kernel linear but does
   not imply known parentage.

8. **Deterministic projection.** The kernels do not include environmental
   stochasticity, demographic stochasticity, or parameter uncertainty.

9. **Observed two-state disturbance mixture.** The expected-regime kernel uses
   the observed frequencies of `disturbance = 0` and `disturbance = 1` among
   quadrat-years as fixed weights. It no longer distinguishes current fire from
   previous-year fire inside the IPM.

10. **Mean initial size distribution.** The one-step projected growth rate uses
    the arithmetic mean of annual observed size-frequency vectors. Recruits
    without a usable size measurement are distributed across the fitted recruit
    size distribution.

11. **Observed-growth validation is incomplete.** Annual abundance change is
    calculated only from quadrats containing individual records in both
    consecutive years. Historically sampled but empty quadrats cannot be
    reconstructed consistently, so this comparison describes matched occupied
    quadrats rather than a complete census of all sampled quadrats.

The chapter retains the full model-fitting analysis from Chapter 9. Only the
IPM-side disturbance handling is changed.

## Set up the analysis

```{r mean-ipm-dist1-settings}
set.seed(100)
options(stringsAsFactors = FALSE)
```


### Model specification

```{r mean-ipm-dist1-specification}
v_head <- 'archbold'
v_species <- 'Polygala lewtonii'
v_years_re <- c()

v_sp_abb <- tolower(
  gsub(
    ' ', '',
    paste(
      substr(unlist(strsplit(v_species, ' ')), 1, 2),
      collapse = '')))

v_script_prefix <- v_head
v_ggp_suffix <- paste(tools::toTitleCase(v_head), '-', v_species)

# Empty vectors select the best-supported polynomial by AICc.
v_mod_set_su <- c()
v_mod_set_gr <- c()
v_mod_set_fl <- c()
v_mod_set_fl_n <- c()
v_mod_set_re <- c()
```


## Data and metadata

```{r mean-ipm-dist1-original-data}
df_og <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_original_260519.csv')))

df_site <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_sitehist_260519.csv')))

df_meta <- data.frame(
  variable = c(
    'unique_ID', 'quad', 'patch', 'year', 'month', 'date', 'qsurv',
    'stg', 'ht', 'mcd', 'st', 'flst'),
  description = c(
    'Unique identifier for each individual plant; Numeric',
    'Name of circular quadrat with radius of 0.25m; Numeric',
    'Name of an aggregate of several quads; Numeric',
    'Year data were collected; Numeric',
    paste(
      'Month data were collected for March, June, September, and',
      'December. April was also used in the first sampling; Numeric'),
    'Combination of year and month; MMYY',
    paste(
      'Quarterly survival code. 0 or 1 refer to survival from the',
      'previous quarter. If a plant was dropped from monitoring,',
      'survival is NA and these plants should NOT be considered dead.',
      '0 = dead; 1 = alive; 2 = missing; 3 = new adult; 5 = seedling'),
    paste(
      'Stage class from March census. If plant was a seedling during',
      'the previous year it is considered a seedling here.',
      '1 = putative seedling; 2 = vegetative; 3 = reproductive adult'),
    'Height from March census in cm; Numeric',
    'Maximum crown diameter from March census in cm; Numeric',
    'Total number of stems; Numeric',
    'Number of flowering stems; Numeric'),
  stringsAsFactors = FALSE)
```

### Load the completed workdata

The workdata already contains current-year disturbance and disturbance in the
exact previous year. These fields are retained directly rather than rebuilt
from the subset of observed individual rows.

```{r mean-ipm-dist1-workdata}
df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv'))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.integer(year),
    site = factor(site)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))
```

## Construct the vital-rate data sets

### Survival

```{r mean-ipm-dist1-survival-data}
df_su <- df %>%
  filter(
    !is.na(survives),
    size_t0 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance, stage)
```

### Growth

```{r mean-ipm-dist1-growth-data}
df_gr <- df %>%
  filter(
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance, stage)
```

### Flowering probability and flowering-stem number

Observed recruits are retained because some recruits flowered.

```{r mean-ipm-dist1-flowering-data}
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance_prev)) %>%
  dplyr::select(
    id, site, year, size_t0, flower, fl_nr, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance_prev, stage, recruits)

df_fl_cond <- df_fl %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr == round(fl_nr))
```

### Quadrat-level flowering stems and recruits

Total flowering stems in quadrat $q$ during year $t$ predict recruits in the
same quadrat during year $t+1$. A quadrat transition is retained only when the
same quadrat has individual records in both years.

```{r mean-ipm-dist1-recruitment-data}
df_quad_year_sampled <- df %>%
  distinct(site, quad, year)

df_fs2r <- df %>%
  group_by(site, quad, year) %>%
  summarise(
    fs_t0 = sum(fl_nr, na.rm = TRUE),
    disturbance = if_else(
      all(is.na(disturbance)),
      NA_real_,
      max(disturbance, na.rm = TRUE)),
    .groups = 'drop') %>%
  mutate(
    year_t1 = year + 1) %>%
  semi_join(
    df_quad_year_sampled %>%
      transmute(
        site,
        quad,
        year_t1 = year),
    by = c(
      'site',
      'quad',
      'year_t1')) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(
        site,
        quad,
        year,
        name = 're_t1'),
    by = c(
      'site',
      'quad',
      'year_t1' = 'year')) %>%
  mutate(
    re_t1 = replace_na(re_t1, 0L),
    site = factor(site),
    fs_t0_log = log1p(fs_t0),
    re_t1_log = log1p(re_t1)) %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !is.na(disturbance),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re))

df_fs2r %>%
  summarise(
    quadrats = n_distinct(quad),
    transitions = n(),
    zero_stems = sum(fs_t0 == 0),
    zero_recruits = sum(re_t1 == 0),
    burned_transitions = sum(disturbance == 1),
    unburned_transitions = sum(disturbance == 0))
```

## Fit the mean vital-rate models

Candidate models use intercept-only, fire-only, and first- through third-order
size polynomials, each with and without the relevant fire term. Empty model-set
vectors select the model with the lowest AICc.

### Survival model

```{r mean-ipm-dist1-survival-model}
mod_su_00 <- glm(
  survives ~ 1,
  data = df_su,
  family = binomial)

mod_su_0 <- glm(
  survives ~ disturbance,
  data = df_su,
  family = binomial)

mod_su_10 <- glm(
  survives ~ logvol_t0,
  data = df_su,
  family = binomial)

mod_su_1 <- glm(
  survives ~ logvol_t0 + disturbance,
  data = df_su,
  family = binomial)

mod_su_20 <- glm(
  survives ~ logvol_t0 + logvol_t0_2,
  data = df_su,
  family = binomial)

mod_su_2 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance,
  data = df_su,
  family = binomial)

mod_su_30 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_su,
  family = binomial)

mod_su_3 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance,
  data = df_su,
  family = binomial)

mods_su <- list(
  mod_su_00, mod_su_0, mod_su_10, mod_su_1,
  mod_su_20, mod_su_2, mod_su_30, mod_su_3)

mods_su_dAICc <- bbmle::AICctab(
  mods_su,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_su_sorted <- order(mods_su_dAICc)

if (length(v_mod_set_su) == 0) {
  mod_su_index_best <- mods_su_sorted[1]
} else {
  mod_su_index_best <- v_mod_set_su + 1
}

v_mod_su_index <- floor((mod_su_index_best - 1) / 2)
mod_su_best <- mods_su[[mod_su_index_best]]

summary(mod_su_best)
mods_su_dAICc
```

### Growth model

```{r mean-ipm-dist1-growth-model}
mod_gr_00 <- lm(
  logvol_t1 ~ 1,
  data = df_gr)

mod_gr_0 <- lm(
  logvol_t1 ~ disturbance,
  data = df_gr)

mod_gr_10 <- lm(
  logvol_t1 ~ logvol_t0,
  data = df_gr)

mod_gr_1 <- lm(
  logvol_t1 ~ logvol_t0 + disturbance,
  data = df_gr)

mod_gr_20 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2,
  data = df_gr)

mod_gr_2 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance,
  data = df_gr)

mod_gr_30 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_gr)

mod_gr_3 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance,
  data = df_gr)

mods_gr <- list(
  mod_gr_00, mod_gr_0, mod_gr_10, mod_gr_1,
  mod_gr_20, mod_gr_2, mod_gr_30, mod_gr_3)

mods_gr_dAICc <- bbmle::AICctab(
  mods_gr,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_gr_sorted <- order(mods_gr_dAICc)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_best <- mods_gr_sorted[1]
} else {
  mod_gr_index_best <- v_mod_set_gr + 1
}

v_mod_gr_index <- floor((mod_gr_index_best - 1) / 2)
mod_gr_best <- mods_gr[[mod_gr_index_best]]

summary(mod_gr_best)
mods_gr_dAICc
```

### Growth variance

Growth variance is modeled as an exponential function of the fitted growth
mean.

```{r mean-ipm-dist1-growth-variance}
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(
    maxiter = 1000,
    tol = 1e-6,
    warnOnly = TRUE))

summary(mod_gr_var)
```

### Flowering-probability model

```{r mean-ipm-dist1-flowering-probability-model}
mod_fl_00 <- glm(
  flower ~ 1,
  data = df_fl,
  family = binomial)

mod_fl_0 <- glm(
  flower ~ disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_10 <- glm(
  flower ~ logvol_t0,
  data = df_fl,
  family = binomial)

mod_fl_1 <- glm(
  flower ~ logvol_t0 + disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_20 <- glm(
  flower ~ logvol_t0 + logvol_t0_2,
  data = df_fl,
  family = binomial)

mod_fl_2 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_30 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_fl,
  family = binomial)

mod_fl_3 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev,
  data = df_fl,
  family = binomial)

mods_fl <- list(
  mod_fl_00, mod_fl_0, mod_fl_10, mod_fl_1,
  mod_fl_20, mod_fl_2, mod_fl_30, mod_fl_3)

mods_fl_dAICc <- bbmle::AICctab(
  mods_fl,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_fl_sorted <- order(mods_fl_dAICc)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_best <- mods_fl_sorted[1]
} else {
  mod_fl_index_best <- v_mod_set_fl + 1
}

v_mod_fl_index <- floor((mod_fl_index_best - 1) / 2)
mod_fl_best <- mods_fl[[mod_fl_index_best]]

summary(mod_fl_best)
mods_fl_dAICc
```

### Flowering-stem-number model

The conditional model includes only observed flowering plants with positive
integer flowering-stem counts.

```{r mean-ipm-dist1-flowering-number-model}
mod_fl_n_00 <- MASS::glm.nb(
  fl_nr ~ 1,
  data = df_fl_cond)

mod_fl_n_0 <- MASS::glm.nb(
  fl_nr ~ disturbance_prev,
  data = df_fl_cond)

mod_fl_n_10 <- MASS::glm.nb(
  fl_nr ~ logvol_t0,
  data = df_fl_cond)

mod_fl_n_1 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_20 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2,
  data = df_fl_cond)

mod_fl_n_2 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_30 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_fl_cond)

mod_fl_n_3 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev,
  data = df_fl_cond)

mods_fl_n <- list(
  mod_fl_n_00, mod_fl_n_0, mod_fl_n_10, mod_fl_n_1,
  mod_fl_n_20, mod_fl_n_2, mod_fl_n_30, mod_fl_n_3)

mods_fl_n_dAICc <- bbmle::AICctab(
  mods_fl_n,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_fl_n_sorted <- order(mods_fl_n_dAICc)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_best <- mods_fl_n_sorted[1]
} else {
  mod_fl_n_index_best <- v_mod_set_fl_n + 1
}

v_mod_fl_n_index <- floor((mod_fl_n_index_best - 1) / 2)
mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_best]]

summary(mod_fl_n_best)
mods_fl_n_dAICc
```

### Flowering-stem-to-recruit model

The recruitment candidates allow a seedbank intercept, a current-fire effect,
a flowering-stem effect, and their interaction. No year or site random effects
are fitted.

```{r mean-ipm-dist1-recruitment-model}
mod_re_00 <- lm(
  re_t1_log ~ 1,
  data = df_fs2r)

mod_re_0 <- lm(
  re_t1_log ~ disturbance,
  data = df_fs2r)

mod_re_10 <- lm(
  re_t1_log ~ fs_t0_log,
  data = df_fs2r)

mod_re_1 <- lm(
  re_t1_log ~ fs_t0_log + disturbance,
  data = df_fs2r)

mod_re_20 <- lm(
  re_t1_log ~ fs_t0_log * disturbance,
  data = df_fs2r)

mods_re <- list(
  mod_re_00,
  mod_re_0,
  mod_re_10,
  mod_re_1,
  mod_re_20)

mods_re_dAICc <- bbmle::AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_re_sorted <- order(mods_re_dAICc)

if (length(v_mod_set_re) == 0) {
  mod_re_index_best <- mods_re_sorted[1]
} else {
  mod_re_index_best <- v_mod_set_re + 1
}

v_mod_re_index <- mod_re_index_best - 1
mod_re_best <- mods_re[[mod_re_index_best]]

summary(mod_re_best)
mods_re_dAICc
```

### Recruitment-model support

```{r mean-ipm-dist1-recruitment-support}
df_recruit_model_support <- tibble(
  model = c(
    'Intercept only',
    'Disturbance only',
    'Flowering stems',
    'Flowering stems + disturbance',
    'Full interaction'),
  dAICc = as.numeric(mods_re_dAICc)) %>%
  mutate(
    weight = exp(-0.5 * dAICc) /
      sum(exp(-0.5 * dAICc))) %>%
  arrange(dAICc)

knitr::kable(
  df_recruit_model_support,
  digits = c(NA, 2, 3),
  caption = paste(
    'AICc support for quadrat-level flowering-stem-to-recruit',
    'models.'))
```

The full interaction model is selected. Its fitted form is

$$
\log(1 + R_{q,t+1}) =
\beta_0 + \beta_1\log(1 + M_{q,t}) + \beta_d d_{q,t}
+ \beta_{1d}\log(1 + M_{q,t})d_{q,t} + \varepsilon,
$$

where $M_{q,t}$ is total flowering stems, $R_{q,t+1}$ is the number of recruits
in the same quadrat in the following year, and $d_{q,t}$ indicates current
fire.

### Recruitment slopes and uncertainty

```{r mean-ipm-dist1-recruitment-slopes}
re_coef <- coef(mod_re_20)
re_vcov <- vcov(mod_re_20)

slope_unburned <- re_coef['fs_t0_log']

se_unburned <- sqrt(
  re_vcov[
    'fs_t0_log',
    'fs_t0_log'])

slope_burned <-
  re_coef['fs_t0_log'] +
  re_coef['fs_t0_log:disturbance']

se_burned <- sqrt(
  re_vcov[
    'fs_t0_log',
    'fs_t0_log'] +
  re_vcov[
    'fs_t0_log:disturbance',
    'fs_t0_log:disturbance'] +
  2 * re_vcov[
    'fs_t0_log',
    'fs_t0_log:disturbance'])

df_recruit_slopes <- tibble(
  disturbance = c(0, 1),
  slope = c(
    slope_unburned,
    slope_burned),
  lower_95 = c(
    slope_unburned - 1.96 * se_unburned,
    slope_burned - 1.96 * se_burned),
  upper_95 = c(
    slope_unburned + 1.96 * se_unburned,
    slope_burned + 1.96 * se_burned))

knitr::kable(
  df_recruit_slopes,
  digits = 3,
  caption = paste(
    'Flowering-stem slopes from the full interaction recruitment',
    'model.'))
```

Recruitment can occur when no flowering stems are observed. This is consistent
with emergence from the persistent seedbank.

```{r mean-ipm-dist1-zero-stem-recruitment}
df_fs2r_zero_stems <- df_fs2r %>%
  filter(fs_t0 == 0) %>%
  group_by(disturbance) %>%
  summarise(
    transitions = n(),
    transitions_with_recruits = sum(re_t1 > 0),
    proportion_with_recruits = mean(re_t1 > 0),
    total_recruits = sum(re_t1),
    mean_recruits = mean(re_t1),
    median_recruits = median(re_t1),
    max_recruits = max(re_t1),
    .groups = 'drop')

knitr::kable(
  df_fs2r_zero_stems,
  digits = 3,
  caption = paste(
    'Recruitment in quadrat transitions with no recorded flowering',
    'stems.'))
```

## Plot the selected vital-rate models

```{r mean-ipm-dist1-plot-functions}
make_mean_prediction <- function(data, model, disturbance_var) {
  x <- seq(
    min(data$logvol_t0, na.rm = TRUE),
    max(data$logvol_t0, na.rm = TRUE),
    length.out = 200)

  pred <- expand_grid(
    logvol_t0 = x,
    disturbance_value = c(0, 1)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_plot = factor(
        disturbance_value,
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  pred[[disturbance_var]] <- pred$disturbance_value
  pred$prediction <- predict(
    model,
    newdata = pred,
    type = 'response')

  pred
}

make_mean_plot <- function(
    data, model, response, disturbance_var, title, y_lab,
    probability = FALSE) {
  pred <- make_mean_prediction(
    data = data,
    model = model,
    disturbance_var = disturbance_var)

  data_plot <- data %>%
    mutate(
      disturbance_plot = factor(
        .data[[disturbance_var]],
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  point_layer <- if (probability) {
    geom_jitter(
      data = data_plot,
      aes(
        x = logvol_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7,
      width = 0.05,
      height = 0.05)
  } else {
    geom_point(
      data = data_plot,
      aes(
        x = logvol_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7)
  }

  p <- ggplot() +
    point_layer +
    geom_line(
      data = pred,
      aes(
        x = logvol_t0,
        y = prediction,
        color = disturbance_plot),
      linewidth = 0.8) +
    scale_color_manual(
      values = c(
        'No fire' = 'black',
        'Fire' = 'red')) +
    theme_bw() +
    labs(
      title = title,
      subtitle = v_ggp_suffix,
      x = expression('log(volume)'[t0]),
      y = y_lab,
      color = NULL)

  if (probability) {
    p <- p + coord_cartesian(ylim = c(0, 1))
  }

  p
}
```

```{r mean-ipm-dist1-vital-rate-figures, fig.width=9, fig.height=7, fig.cap='Selected mean survival, growth, flowering-probability, and conditional flowering-stem models. Current fire is used for survival and growth, and previous-year fire is used for flowering.'}
fig_su_mean <- make_mean_plot(
  data = df_su,
  model = mod_su_best,
  response = 'survives',
  disturbance_var = 'disturbance',
  title = 'Mean survival',
  y_lab = 'Survival probability',
  probability = TRUE)

fig_gr_mean <- make_mean_plot(
  data = df_gr,
  model = mod_gr_best,
  response = 'logvol_t1',
  disturbance_var = 'disturbance',
  title = 'Mean growth',
  y_lab = expression('log(volume)'[t1])) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = 'dashed')

fig_fl_mean <- make_mean_plot(
  data = df_fl,
  model = mod_fl_best,
  response = 'flower',
  disturbance_var = 'disturbance_prev',
  title = 'Mean flowering probability',
  y_lab = 'Flowering probability',
  probability = TRUE)

fig_fl_n_mean <- make_mean_plot(
  data = df_fl_cond,
  model = mod_fl_n_best,
  response = 'fl_nr',
  disturbance_var = 'disturbance_prev',
  title = 'Mean flowering-stem number',
  y_lab = 'Number of flowering stems')

fig_vital_rates_mean <-
  fig_su_mean + fig_gr_mean + fig_fl_mean + fig_fl_n_mean +
  plot_layout(ncol = 2, guides = 'collect') &
  theme(legend.position = 'bottom')

fig_vital_rates_mean
```

### Flowering-stem-to-recruit relationship

```{r mean-ipm-dist1-recruitment-figure, fig.width=7, fig.height=5, fig.cap='Quadrat-level flowering stems in year t and recruits in the same quadrat in year t + 1 under current-fire and no-fire conditions.'}
df_fs2r_pred <- expand_grid(
  fs_t0_log = seq(
    0,
    max(df_fs2r$fs_t0_log, na.rm = TRUE),
    length.out = 200),
  disturbance = c(0, 1)) %>%
  mutate(
    re_t1_log = predict(
      mod_re_best,
      newdata = .),
    fs_t0 = expm1(fs_t0_log),
    re_t1 = pmax(0, expm1(re_t1_log)),
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c('No fire', 'Fire')))

df_fs2r_plot <- df_fs2r %>%
  mutate(
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c('No fire', 'Fire')))

fig_fs2r_mean <- ggplot(
  df_fs2r_plot,
  aes(
    x = fs_t0,
    y = re_t1,
    color = disturbance_plot)) +
  geom_point(alpha = 0.4, size = 1.3) +
  geom_line(
    data = df_fs2r_pred,
    aes(
      x = fs_t0,
      y = re_t1,
      color = disturbance_plot),
    linewidth = 0.8) +
  scale_x_continuous(
  transform = scales::transform_pseudo_log(
    sigma = 1)) +
  scale_color_manual(
    values = c(
      'No fire' = 'black',
      'Fire' = 'red')) +
  theme_bw() +
  labs(
    title = 'Mean flowering-stem-to-recruit relationship',
    subtitle = v_ggp_suffix,
    x = expression('Total flowering stems'[t]),
    y = expression('Recruits'[t + 1]),
    color = NULL)

fig_fs2r_mean
```

## Recruit size distribution

```{r mean-ipm-dist1-recruit-size}
df_recr <- df %>%
  filter(
    recruits == 1,
    volume_t0 > 0,
    is.finite(logvol_t0))

rc_sz <- df_recr %>%
  summarise(
    mean = mean(logvol_t0, na.rm = TRUE),
    sd = sd(logvol_t0, na.rm = TRUE))

rc_sz
```

## Extract the selected model parameters

The extraction functions return zero for terms that are absent from the
selected model. The same IPM functions can therefore use any selected candidate
without manual coefficient changes.

```{r mean-ipm-dist1-parameter-extraction}
size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logvol_t0',
  b2 = 'logvol_t0_2',
  b3 = 'logvol_t0_3',
  bd = 'disturbance')

lagged_size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logvol_t0',
  b2 = 'logvol_t0_2',
  b3 = 'logvol_t0_3',
  bd = 'disturbance_prev')

extract_terms <- function(model, prefix, term_map) {
  coefs <- stats::coef(model)

  values <- vapply(term_map, function(term_i) {
    if (term_i %in% names(coefs)) {
      unname(coefs[[term_i]])
    } else {
      0
    }
  }, numeric(1))

  names(values) <- paste0(prefix, names(term_map))
  values
}

extract_coef <- function(model, term, default = 0) {
  coefs <- stats::coef(model)

  if (term %in% names(coefs)) {
    unname(coefs[[term]])
  } else {
    default
  }
}

grow_var_coef <- stats::coef(mod_gr_var)

mesh_limits <- range(
  c(
    df_gr$logvol_t0,
    df_gr$logvol_t1,
    df_recr$logvol_t0),
  na.rm = TRUE,
  finite = TRUE)

mesh_limits <- c(
  mesh_limits[1] - 0.1,
  mesh_limits[2] + 0.1)

pars_mean <- as.list(c(
  extract_terms(
    mod_su_best,
    'surv_',
    size_term_map),
  extract_terms(
    mod_gr_best,
    'grow_',
    size_term_map),
  extract_terms(
    mod_fl_best,
    'fl_',
    lagged_size_term_map),
  extract_terms(
    mod_fl_n_best,
    'fln_',
    lagged_size_term_map),
  re_b0 = extract_coef(
    mod_re_best,
    '(Intercept)'),
  re_b1 = extract_coef(
    mod_re_best,
    'fs_t0_log'),
  re_bd = extract_coef(
    mod_re_best,
    'disturbance'),
  re_bf = extract_coef(
    mod_re_best,
    'fs_t0_log:disturbance'),
  re_sigma = stats::sigma(mod_re_best),
  recr_mean = rc_sz$mean,
  recr_sd = rc_sz$sd,
  grow_var_a = unname(grow_var_coef[['a']]),
  grow_var_b = unname(grow_var_coef[['b']]),
  fs_ref = mean(df_fs2r$fs_t0, na.rm = TRUE),
  L = mesh_limits[1],
  U = mesh_limits[2],
  mat_siz = 200,
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_fl_index = v_mod_fl_index,
  mod_fl_n_index = v_mod_fl_n_index,
  mod_re_index = v_mod_re_index))

pars_mean_table <- tibble(
  parameter = names(pars_mean),
  value = as.numeric(unlist(pars_mean)))

knitr::kable(
  pars_mean_table,
  digits = 5,
  caption = 'Parameters passed to the mean IPM.')
```

## Define the IPM functions

The extracted flowering coefficients still come from models fitted with
`disturbance_prev`. From this point onward, however, their disturbance
coefficient is used with the same generic `disturbance` argument as all other
vital rates.

### General parameter and linear-predictor functions

```{r mean-ipm-dist1-general-functions}
get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    pars[[par_name]]
  } else {
    default
  }
}

vital_lp <- function(x, pars, prefix, disturbance = 0) {
  get_par(pars, paste0(prefix, 'b0')) +
    get_par(pars, paste0(prefix, 'b1')) * x +
    get_par(pars, paste0(prefix, 'b2')) * x^2 +
    get_par(pars, paste0(prefix, 'b3')) * x^3 +
    get_par(pars, paste0(prefix, 'bd')) * disturbance
}
```

### Survival

```{r mean-ipm-dist1-survival-function}
sx <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'surv_',
    disturbance = disturbance))
}
```

### Growth

```{r mean-ipm-dist1-growth-functions}
grow_mu <- function(x, pars, disturbance = 0) {
  vital_lp(
    x = x,
    pars = pars,
    prefix = 'grow_',
    disturbance = disturbance)
}

grow_sd <- function(mean_value, pars) {
  variance <- get_par(pars, 'grow_var_a') *
    exp(get_par(pars, 'grow_var_b') * mean_value)

  sqrt(pmax(variance, .Machine$double.eps))
}

gxy <- function(x, y, pars, disturbance = 0) {
  mean_value <- grow_mu(
    x = x,
    pars = pars,
    disturbance = disturbance)

  sd_value <- grow_sd(
    mean_value = mean_value,
    pars = pars)

  dnorm(
    y,
    mean = mean_value,
    sd = sd_value)
}
```

### Flowering probability and conditional flowering-stem number

The fitted flowering models still use `disturbance_prev`; only their IPM
functions are changed so that the extracted disturbance coefficient is driven
by the shared disturbance state.

```{r mean-ipm-dist1-flowering-functions}
fl_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fl_',
    disturbance = disturbance))
}

fln_x <- function(x, pars, disturbance = 0) {
  exp(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fln_',
    disturbance = disturbance))
}

fs_x <- function(x, pars, disturbance = 0) {
  fl_x(
    x = x,
    pars = pars,
    disturbance = disturbance) *
    fln_x(
      x = x,
      pars = pars,
      disturbance = disturbance)
}
```

### Flowering-stem-to-recruit conversion

The coefficient extraction supports intercept-only, disturbance-only,
flowering-stem, additive, and interaction models. Recruitment is not forced to
zero when flowering stems equal zero because the fitted intercept represents
seedbank emergence.

```{r mean-ipm-dist1-recruitment-functions}
predict_recruits_from_stems <- function(
    flowering_stems, pars, disturbance = 0) {
  stems_nonnegative <- pmax(flowering_stems, 0)

  slope <- get_par(pars, 're_b1') +
    get_par(pars, 're_bf') * disturbance

  eta <- get_par(pars, 're_b0') +
    get_par(pars, 're_bd') * disturbance +
    slope * log1p(stems_nonnegative)

  recruits <- expm1(
    eta + 0.5 * get_par(pars, 're_sigma')^2)

  pmax(recruits, 0)
}

recruits_per_stem <- function(pars, disturbance = 0) {
  fs_ref <- get_par(pars, 'fs_ref', 0)

  if (!is.finite(fs_ref) || fs_ref <= 0) {
    return(0)
  }

  predict_recruits_from_stems(
    flowering_stems = fs_ref,
    pars = pars,
    disturbance = disturbance) / fs_ref
}

rx <- function(x, pars, disturbance = 0) {
  fs_x(
    x = x,
    pars = pars,
    disturbance = disturbance) *
    recruits_per_stem(
      pars = pars,
      disturbance = disturbance)
}
```

The fitted model predicts nonzero recruitment from the seedbank when no current
flowering stems are recorded.

```{r mean-ipm-dist1-zero-stem-predictions}
tibble(
  disturbance = c(0, 1),
  predicted_recruits_zero_stems =
    predict_recruits_from_stems(
      flowering_stems = 0,
      pars = pars_mean,
      disturbance = disturbance))
```

### Recruit size distribution and fertility matrix

```{r mean-ipm-dist1-fertility-functions}
recr_y <- function(y, pars) {
  dnorm(
    y,
    mean = get_par(pars, 'recr_mean'),
    sd = get_par(pars, 'recr_sd'))
}

fy <- function(y, x, pars, h, disturbance = 0) {
  Pvec <- recr_y(y, pars)
  recruit_mass <- sum(Pvec * h)

  if (!is.finite(recruit_mass) || recruit_mass <= 0) {
    stop('Recruit size distribution has no mass inside the mesh')
  }

  Pvec <- Pvec / recruit_mass

  Rvec <- rx(
    x = x,
    pars = pars,
    disturbance = disturbance)

  outer(Pvec, Rvec) * h
}
```

### Full kernel

Only one disturbance argument is now accepted by the fertility and full-kernel
functions. That value is propagated to every vital-rate component.

```{r mean-ipm-dist1-kernel-function}
kernel <- function(pars, disturbance = 0) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  Fmat <- fy(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance)

  Smat <- sx(
    x = y,
    pars = pars,
    disturbance = disturbance)

  Gmat <- matrix(0, n, n)
  Gmat[] <- t(
    outer(
      y,
      y,
      Vectorize(function(x, y) {
        gxy(
          x = x,
          y = y,
          pars = pars,
          disturbance = disturbance)
      }))) * h

  Tmat <- matrix(0, n, n)

  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  k_yx <- Fmat + Tmat

  list(
    k_yx = k_yx,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Smat = Smat,
    meshpts = y,
    h = h)
}

lambda_ipm <- function(pars, disturbance = 0) {
  K <- kernel(
    pars = pars,
    disturbance = disturbance)$k_yx

  Re(eigen(K, only.values = TRUE)$values[1])
}
```

## Construct the mean observed initial size distribution

The initial vector is the arithmetic mean of annual size-frequency vectors.
Unsized recruits are distributed over the fitted recruit-size distribution.

```{r mean-ipm-dist1-initial-distribution}
make_initial_n_mean <- function(pars, df_init = df) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  breaks <- seq(L, U, length.out = n + 1)
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  recruit_density <- recr_y(y, pars)
  recruit_density <- recruit_density /
    sum(recruit_density * h)

  years_i <- df_init %>%
    filter(!is.na(year)) %>%
    distinct(year) %>%
    arrange(year) %>%
    pull(year)

  counts_by_year <- purrr::map(years_i, function(year_i) {
    df_i <- df_init %>%
      filter(year == year_i)

    x <- df_i %>%
      filter(is.finite(logvol_t0)) %>%
      pull(logvol_t0)

    size_counts <- hist(
      pmin(pmax(x, L), U),
      breaks = breaks,
      plot = FALSE,
      include.lowest = TRUE)$counts

    n_unsized_recruits <- df_i %>%
      summarise(
        n = sum(
          !is.finite(logvol_t0) & recruits == 1,
          na.rm = TRUE)) %>%
      pull(n)

    size_counts +
      n_unsized_recruits * recruit_density * h
  })

  mean_counts <- Reduce(`+`, counts_by_year) /
    length(counts_by_year)

  mean_counts / h
}
```

## Quadrat-level flowering-stem reference

The recruitment model is fitted to flowering-stem totals per quadrat. Because
the IPM now uses one disturbance state, the reference stem abundance is also
summarized directly by that same `disturbance` variable.

```{r mean-ipm-dist1-stem-reference}
df_stem_reference <- df_fs2r %>%
  group_by(disturbance) %>%
  summarise(
    fs_ref = mean(fs_t0),
    n_quadrat_transitions = n(),
    .groups = 'drop')

get_stem_reference <- function(disturbance_i) {
  df_stem_reference %>%
    filter(disturbance == disturbance_i) %>%
    pull(fs_ref)
}

mean_stems_no_fire <- get_stem_reference(0)
mean_stems_fire <- get_stem_reference(1)

knitr::kable(
  df_stem_reference,
  digits = 3,
  caption = paste(
    'Mean quadrat-level flowering-stem abundance by shared',
    'disturbance state.'))
```

## Observed disturbance regime

Each unique quadrat-year receives equal weight. Only the shared current
`disturbance` variable is used here, so the expected regime contains two states
rather than the four current/previous-fire combinations used in Chapter 9.

```{r mean-ipm-dist1-disturbance-regime}
df_disturbance_regime <- df %>%
  filter(
    !is.na(quad),
    !is.na(year),
    !is.na(disturbance)) %>%
  distinct(
    quad,
    year,
    disturbance) %>%
  count(
    disturbance,
    name = 'n_quadrat_years') %>%
  complete(
    disturbance = c(0, 1),
    fill = list(n_quadrat_years = 0)) %>%
  mutate(
    p_state = n_quadrat_years /
      sum(n_quadrat_years)) %>%
  arrange(disturbance)

knitr::kable(
  df_disturbance_regime,
  digits = c(0, 0, 4),
  caption = 'Observed quadrat-year frequencies of the two disturbance states.')

get_state_probability <- function(disturbance_i) {
  df_disturbance_regime %>%
    filter(disturbance == disturbance_i) %>%
    pull(p_state)
}

p_0 <- get_state_probability(0)
p_1 <- get_state_probability(1)

mean_stems_regime <-
  p_0 * mean_stems_no_fire +
  p_1 * mean_stems_fire

tibble(
  reference = c(
    'No fire',
    'Fire',
    'Observed disturbance regime'),
  flowering_stems = c(
    mean_stems_no_fire,
    mean_stems_fire,
    mean_stems_regime))
```

## Build the state-specific and expected-regime kernels

The IPM now has only an undisturbed and a fire kernel. In each kernel, the same
disturbance value is supplied to every vital-rate function. The expected-regime
kernel is their observed-frequency-weighted average.

```{r mean-ipm-dist1-state-kernels}
pars_0 <- pars_mean
pars_0$fs_ref <- mean_stems_no_fire

pars_1 <- pars_mean
pars_1$fs_ref <- mean_stems_fire

# All mixture components use the same expected total flowering-stem production
# under the observed disturbance regime.
pars_regime <- pars_mean
pars_regime$fs_ref <- mean_stems_regime

K_0 <- kernel(
  pars = pars_0,
  disturbance = 0)$k_yx

K_1 <- kernel(
  pars = pars_1,
  disturbance = 1)$k_yx

K_regime_0 <- kernel(
  pars = pars_regime,
  disturbance = 0)$k_yx

K_regime_1 <- kernel(
  pars = pars_regime,
  disturbance = 1)$k_yx

K_regime <-
  p_0 * K_regime_0 +
  p_1 * K_regime_1
```

## Calculate asymptotic and projected population growth

The asymptotic growth rate is the dominant eigenvalue of the kernel. The
projected growth rate is the one-step change from the mean observed size
distribution.

```{r mean-ipm-dist1-projection-functions}
project_kernel <- function(K, pars, label, df_init = df) {
  n_obs <- make_initial_n_mean(pars, df_init)
  h <- (pars$U - pars$L) / pars$mat_siz
  n_initial <- sum(n_obs) * h

  n_proj <- K %*% n_obs
  n_projected <- sum(n_proj) * h

  tibble(
    lambda_type = label,
    asym_lambda = Re(
      eigen(
        K,
        only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial,
    n_initial = n_initial,
    n_projected = n_projected)
}

make_state_lambda <- function(
    K, pars, label, disturbance, p_state) {
  project_kernel(
    K = K,
    pars = pars,
    label = label) %>%
    mutate(
      disturbance = disturbance,
      observed_state_probability = p_state,
      flowering_stems_ref = pars$fs_ref,
      recruits_ref = predict_recruits_from_stems(
        flowering_stems = pars$fs_ref,
        pars = pars,
        disturbance = disturbance),
      recruits_per_stem = if_else(
        flowering_stems_ref > 0,
        recruits_ref / flowering_stems_ref,
        0))
}
```

```{r mean-ipm-dist1-lambda-results}
df_lambda_states <- bind_rows(
  make_state_lambda(
    K = K_0,
    pars = pars_0,
    label = 'Undisturbed',
    disturbance = 0,
    p_state = p_0),
  make_state_lambda(
    K = K_1,
    pars = pars_1,
    label = 'Fire',
    disturbance = 1,
    p_state = p_1))

recruits_ref_regime <-
  p_0 * predict_recruits_from_stems(
    flowering_stems = mean_stems_regime,
    pars = pars_regime,
    disturbance = 0) +
  p_1 * predict_recruits_from_stems(
    flowering_stems = mean_stems_regime,
    pars = pars_regime,
    disturbance = 1)

df_lambda_regime <- project_kernel(
  K = K_regime,
  pars = pars_regime,
  label = 'Expected disturbance regime') %>%
  mutate(
    disturbance = NA_real_,
    observed_state_probability = 1,
    flowering_stems_ref = mean_stems_regime,
    recruits_ref = recruits_ref_regime,
    recruits_per_stem = if_else(
      flowering_stems_ref > 0,
      recruits_ref / flowering_stems_ref,
      0))

df_lambda_mean <- bind_rows(
  df_lambda_states,
  df_lambda_regime)

knitr::kable(
  df_lambda_mean,
  digits = 3,
  caption = paste(
    'Asymptotic and one-step projected population growth under the',
    'two shared disturbance states and the observed disturbance regime.'))
```

## Observed annual abundance change in matched occupied quadrats

Counts include sized plants and recruits without a usable size measurement.
Only quadrats with individual records in both consecutive years are retained.
Historically sampled empty quadrats cannot be added because they have no
individual rows in the workdata.

```{r mean-ipm-dist1-observed-growth}
df_counts_quad_year <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year)) %>%
  group_by(site, quad, year) %>%
  summarise(
    n_sized = sum(
      is.finite(logvol_t0),
      na.rm = TRUE),
    n_unsized_recruits = sum(
      !is.finite(logvol_t0) & recruits == 1,
      na.rm = TRUE),
    n_t0 = n_sized + n_unsized_recruits,
    .groups = 'drop') %>%
  arrange(site, quad, year) %>%
  group_by(site, quad) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n_t0),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_obs_growth_year <- df_counts_quad_year %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(n_t0),
    n_t1 = sum(n_t1),
    n_quadrats = n(),
    n_sites = n_distinct(site),
    obs_lambda = n_t1 / n_t0,
    .groups = 'drop') %>%
  filter(n_t0 > 0)

df_obs_growth_year
```

```{r mean-ipm-dist1-observed-growth-summary}
df_obs_growth_summary <- df_obs_growth_year %>%
  summarise(
    n_transitions = n(),
    lambda_arithmetic = mean(
      obs_lambda,
      na.rm = TRUE),
    lambda_geometric = exp(
      mean(
        log(obs_lambda),
        na.rm = TRUE)),
    lambda_median = median(
      obs_lambda,
      na.rm = TRUE),
    lambda_min = min(
      obs_lambda,
      na.rm = TRUE),
    lambda_max = max(
      obs_lambda,
      na.rm = TRUE))

knitr::kable(
  df_obs_growth_summary,
  digits = 3,
  caption = paste(
    'Summary of annual abundance change in matched occupied',
    'quadrats.'))
```

## Compare observed and modeled population growth

```{r mean-ipm-dist1-growth-comparison-figure, fig.width=9, fig.height=5, fig.cap='Annual abundance change in matched occupied quadrats and asymptotic and one-step projected growth rates from the mean IPM.'}
df_model_lines <- df_lambda_mean %>%
  dplyr::select(
    lambda_type,
    asym_lambda,
    proj_lambda) %>%
  pivot_longer(
    cols = c(
      asym_lambda,
      proj_lambda),
    names_to = 'metric',
    values_to = 'lambda') %>%
  mutate(
    metric = recode(
      metric,
      asym_lambda = 'Asymptotic',
      proj_lambda = 'Projected'))

x_point <- max(
  df_obs_growth_year$year,
  na.rm = TRUE) + 0.4

fig_lambda_mean <- ggplot(
  df_obs_growth_year,
  aes(
    x = year,
    y = obs_lambda)) +
  geom_hline(
    yintercept = 1,
    linetype = 'dotted') +
  geom_line(color = 'black') +
  geom_point(
    size = 2,
    color = 'black') +
  geom_hline(
    data = df_model_lines,
    aes(
      yintercept = lambda,
      color = lambda_type),
    linewidth = 0.8) +
  geom_point(
    data = df_model_lines,
    aes(
      x = x_point,
      y = lambda,
      color = lambda_type,
      shape = metric),
    size = 3) +
  expand_limits(x = x_point + 1) +
  theme_bw() +
  labs(
    title = 'Observed annual growth and mean-IPM lambda',
    subtitle = v_ggp_suffix,
    x = 'Year t0',
    y = expression(lambda),
    color = NULL,
    shape = NULL)

fig_lambda_mean
```

## Kernel diagnostics

```{r mean-ipm-dist1-kernel-diagnostics}
df_kernel_diagnostics <- tibble(
  kernel = c(
    'Undisturbed',
    'Fire',
    'Expected disturbance regime'),
  min_entry = c(
    min(K_0),
    min(K_1),
    min(K_regime)),
  max_entry = c(
    max(K_0),
    max(K_1),
    max(K_regime)),
  any_nonfinite = c(
    any(!is.finite(K_0)),
    any(!is.finite(K_1)),
    any(!is.finite(K_regime))))

knitr::kable(
  df_kernel_diagnostics,
  digits = c(NA, 6, 6, NA),
  caption = paste(
    'Numerical diagnostics for the two state-specific and expected-regime',
    'kernels.'))
```

All kernel entries should be finite and non-negative. These checks must be
repeated whenever the vital-rate models, mesh limits, recruitment conversion,
or disturbance-state weighting are changed.

## Interpretation of the mean model

The fitted vital-rate models retain the same biological timing as in Chapter 9:
current disturbance affects survival, growth, and the quadrat-level recruitment
model, whereas previous-year disturbance affects flowering probability and
flowering-stem number. The simplification introduced here begins only when those
fitted coefficients are assembled into the IPM.

Inside the IPM, a single binary disturbance state is imposed on all demographic
processes simultaneously. Consequently, the `Fire` kernel should be interpreted
as a deliberately aligned disturbance scenario, not as a literal annual fire
transition supported by the original timing of every vital rate. In particular,
the flowering disturbance coefficients were estimated from previous-year fire
but are evaluated at the same shared state as current-fire effects in the other
vital rates.

The flowering-stem-to-recruit relationship remains a quadrat-level
relationship. Recruitment in quadrats without current flowering is attributed
to the persistent seedbank. The model does not identify parent plants or
separate seed production, seed survival, germination, and establishment.

The expected-regime kernel combines the observed frequencies of the two shared
disturbance states. It is therefore a deterministic summary under this imposed
two-state disturbance representation, not a stochastic fire model and not a
simulation of the historical sequence of current and previous-year fire.

The comparison with observed annual abundance change is unchanged from Chapter
9 and remains limited by the available workdata. Quadrats sampled with zero
plants are absent from the historical individual-level series, so the observed
geometric mean describes matched occupied quadrats and may omit some local
extinction and recolonization events.

## Save key outputs

The following commands are retained from the analysis script and can be
uncommented when the outputs should be written to disk.

```{r mean-ipm-dist1-save-outputs, eval=FALSE}
# saveRDS(
#   pars_mean,
#   file.path(
#     dir_result,
#     'pole_mean_dist1_pars.rds'))

# saveRDS(
#   df_lambda_mean,
#   file.path(
#     dir_result,
#     'pole_mean_dist1_lambda.rds'))

# saveRDS(
#   df_obs_growth_year,
#   file.path(
#     dir_result,
#     'pole_mean_dist1_observed_growth.rds'))
```
