# Historical integral projection models {#mean-ipm-hist-kuss}

## Background

Standard integral projection models (IPMs) describe changes in a continuously
structured population between consecutive time steps. If $x$ denotes
individual size at time $t$ and $y$ denotes size at time $t+1$, population
projection can be written as

$$
n(y,t+1)
=
\int_L^U K(y,x)n(x,t)\,dx,
$$

where the projection kernel is commonly separated into a survival-growth
component, $P(y,x)$, and a fecundity component, $F(y,x)$:

$$
K(y,x) = P(y,x) + F(y,x).
$$

After discretizing the continuous size distribution into $m$ mesh points,
this becomes the familiar matrix projection

$$
\mathbf{n}_{t+1}
=
\mathbf{K}\mathbf{n}_t,
$$

with

$$
\mathbf{K}
=
\mathbf{P}+\mathbf{F}.
$$

An important assumption of this standard, or *ahistorical*, formulation is
that the demographic future of an individual is completely described by its
state at time $t$. Conditional on current size, information from earlier time
steps is not retained. Thus, two individuals of equal size at time $t$ are
treated as demographically equivalent even if they reached that size through
different histories.

Historical population models relax this assumption. Previous states or events
may influence demographic performance after accounting for current state
(Ehrlén 2000; Shefferson et al. 2021). In a general second-order historical
model, demographic transitions can therefore depend on the sequence

$$
z_{t-1}
\rightarrow
z_t
\rightarrow
z_{t+1},
$$

rather than only on

$$
z_t
\rightarrow
z_{t+1}.
$$

The relevant history need not necessarily be previous individual size. It may
also represent previous reproductive status, damage, resource allocation, or
environmental conditions that produce carry-over effects.

Here, two different historical formulations are considered. The first is the
*time-lagged IPM* of Kuss et al. (2008), in which fecundity operates with an
additional one-year delay. The second is an *environmental historical IPM* in
which the population retains information about fire during the previous
interval because different demographic processes respond to fire over
different time periods.

Although both approaches enlarge the population state and produce a
megamatrix, the biological meaning of the additional state differs between
them.


## Readings
**Kuss, P., Rees, M., Ægisdóttir, H. H., Ellner, S. P., & Stöcklin, J. (2008). Evolutionary demography of long‐lived monocarpic perennials: a time‐lagged integral projection model. Journal of Ecology, 96(4), 821-832.**

1. “*So far, the existing class of models assume that demographic rates are only influenced by current individual state, say size and/or age. In many species, however, the decision to flower is made many months (e.g. Agave deserti; Tissue & Nobel 1990) or years (Frasera speciosa; Taylor & Inouye 1985) before flowering occurs. Additionally, flowering individuals of monocarpic perennials often have a different morphology, for example, withered or absent rosette leaves, such that size measurements in the year of flowering cannot be compared with the years prior to flowering (e.g. Campanula sibirica, C. spicata; Hegi 1975). In these latter cases, an additional time-lag needs to be incorporated into the models in order to faithfully represent the demography of a study species.*”

2. “*Our objectives in this study are therefore (i) to present a theoretical and computational framework for a time-lagged IPM which allows demographic transitions to depend on an individual’s state either in the current year or the year before, (ii) to demonstrate that the novel combination of field methods (permanent plots and herb chronology) can be used to parameterize efficiently and validate IPMs with data from the long-lived monocarpic C. thyrsoides collected from a relatively short census period, (iii) to use the model to predict the age-structure observed in the field and determine the evolutionarily stable flowering strategy, and (iv) to draw general conclusions about the evolution of the flowering strategies in long-lived vs. short-lived monocarpic plant species.*”

3. “*However, in many species a complication arises because either (i) rosette size cannot be measured in the year the plants flower or (ii) the decision to flower is made the year before plants flower. As a consequence the probability of flowering and seed production in year t has to be predicted as a function of plant size in year t − 1, and any seedlings produced then enter the population in year t + 1. Therefore the time-lagged IPM becomes*”

4. “*The matrix K is useful for calculating various measures of population growth; however, it is inappropriate for sensitivity and elasticity analyses because the identity matrix, I, in the lower left hand corner is simply a computational device and not a biological transition.*”


**Ehrlén, J. (2000). The dynamics of plant populations: does the history of individuals matter?. Ecology, 81(6), 1675-1684.**

1. “*At the level of individuals, historical effects in which past conditions influence future performance are particularly likely to occur in long-lived organisms that store resources between seasons and that form organs months or years before their elaboration. Such carry-over mechanisms have been documented in several perennial plant species, but the implications for population processes are poorly known. In this study, I examine how the history of individuals influences their future performance, population dynamics, and life cycle structure in the long-lived herb Lathyrus vernus. Overall effects of plant history on population dynamics, in terms of growth rate, reproductive values, stable stage distribution, and elasticities, are examined by comparing an ordinary first-order matrix model with a second-order matrix model. In the latter, not only the present state of individuals, but also their past state is allowed to influence future fate.*”

2. “*In perennial plants, the current phenotype of an individual and its response to environmental change have been shown to reflect the individual’s history, i.e., its past phenotype and environment (Primack and Hall 1990, Bullock et al. 1993, Geber et al. 1997). Such carry-over mechanisms imply that the performance of individuals can be influenced both by past environmental conditions, such as herbivory or abiotic factors (Taylor and Inouye 1985, Nobel 1987, Bullock et al. 1993, 1995, Crone 1997, Rose et al. 1998), and by past internal plant events such as morphological development and resource allocation patterns (Karlsson 1986, Karlsson et al. 1990, Primack and Hall 1990, Geber et al. 1997). The mechanisms by which past events can influence plant phenotypes and their ability to respond to environmental variation include resource storage and a temporal separation between organ formation and elaboration. Resource stores are widely recognized as important to the ecology of perennial plants (Chapin et al. 1990), and organ preformation is widespread among angiosperms (Geber et al. 1997). Hence, carry-over mechanisms and historical effects are likely to occur in many perennial plant species.*”

3. “*There are several ways in which historical events may influence the future performance of plants that currently belong to the same stage. Such historical effects may arise because the costs of allocating resources to one function, e.g., reproduction, in terms of decreased allocation to other functions, are not expressed until after more than one year (e.g., Primack and Hall 1990). Moreover, in plants that have the potential to accumulate resources and store them among seasons, plant size may be partly restored after it has been depressed during one season, e.g., by herbivory. In addition, correlations between past and current performance can be mediated by the external environment. If size changes of individuals reflect altered environmental conditions that persist for more than one season, then individuals experiencing a certain size change during one year are likely to experience a similar change during the next year.*”

4. “*Conventional matrix models are first-order models, i.e., the stage distribution at time t + 1 depends solely on the stage distribution at time t.[...] When the Markovian process is extended to be second order, the stage distribution at time t + 1 becomes dependent upon the distribution at time t − 1 as well as that at time t, and on second-order transition probabilities (Tanner et al. 1996). The matrix of transition probabilities in a second-order model is no longer two-dimensional, but instead becomes three-dimensional. In analogy with the conventional notation, the equation for the system can be expressed as: x(t + 1, t) = Ax(t, t − 1), where A is the three-dimensional transition probability matrix and x(t + 1, t) is a two-dimensional matrix describing the joint stage distributions at times t + 1 and t. Thus, the element aijk of A represents the probability that an individual of stages k and j in years t − 1 and t, respectively, transitions to stages j and i in years t and t + 1.*”

5. “*In ‘historical’ models, each transition from year t to year t + 1 was calculated separately for the seven stages of individuals in year t − 1 (Table 2). To allow for the analytical procedures developed for first-order models, three-dimensional matrices can be converted to a two-dimensional form and two-dimensional stage vectors can be transformed into an one-dimensional form. In this two-dimensional matrix, the columns represent the combinations of individual stages in years t − 1 and t (j, k), and rows represent the combination of individual stages in years t and t + 1 (i, j). Hence, the three-dimensional matrix for L. vernus was converted to a two-dimensional form, including transitions from all combinations of stage year t − 1 and year t to all combinations of stage year t and stage year t + 1.*”


**Shefferson, R. P., Kurokawa, S., & Ehrlén, J. (2021). lefko3: Analysing individual history through size‐classified matrix population models. Methods in Ecology and Evolution, 12(2), 378-382.**

1. “*Matrix projection models typically assume that the histories of individuals do not influence population dynamics. Specifically, the state of an individual in time t + 1 is solely dependent on the state of that individual at time t. We refer to such MPMs as ahistorical MPMs, or ahMPMs, and use the term MPM to refer generally to any kind of MPM. This assumption is a temporal interpretation of the assumption of independence and identical distribution (iid) in vital rates, and it allows the creation of low-dimension MPMs in which columns represent life-history stage (or age) in time t and rows represent stage (or age) in time t + 1 (Caswell, 2001). When stage-based MPMs are characterized by transitions between life-history stages, yielding a matrix in which the transition probabilities and fecundity of an individual in time t + 1 are governed solely by its condition in time t (Lefkovitch, 1965). Integral projection models (IPMs) are function-based MPMs in which size follows a Gaussian distribution (Ellner & Rees, 2006). Neither traditional MPMs nor IPMs incorporate individual history.*”

2. “*Matrix projection models typically assume that the histories of individuals do not influence population dynamics. Specifically, the state of an individual in time t + 1 is solely dependent on the state of that individual at time t. We refer to such MPMs as ahistorical MPMs, or ahMPMs, and use the term MPM to refer generally to any kind of MPM. This assumption is a temporal interpretation of the assumption of independence and identical distribution (iid) in vital rates, and it allows the creation of low-dimension MPMs in which columns represent life-history stage (or age) in time t and rows represent stage (or age) in time t + 1 (Caswell, 2001). When stage-based MPMs are characterized by transitions between life-history stages, yielding a matrix in which the transition probabilities and fecundity of an individual in time t + 1 are governed solely by its condition in time t (Lefkovitch, 1965). Integral projection models (IPMs) are function-based MPMs in which size follows a Gaussian distribution (Ellner & Rees, 2006). Neither traditional MPMs nor IPMs incorporate individual history.*”

3. “*In an ahMPM, rows and columns correspond to the states of individuals in times t and t + 1 (Table 1a). hMPMs add time t − 1 into the two-dimensional matrix by recharacterizing the rows and columns of the matrix so that they correspond to stage-pairs defined across two consecutive times (Table 1b). Redefining stages in this way leads to large, sparse matrices. An ahMPM with m stages will have m2 elements, all of which may be non-zero. A hMPM with m stages will have m2 stage-pairs on each axis and m4 elements, of which m3 elements are potentially estimable and (m − 1)m3 elements are structural zeroes. Structural zeroes result from the fact that row stage pairs and column stage pairs both include stage in time t, and so stage at time t must be equal in both for a transition to be estimable.*”

4. “*Function modelsearch() takes historical vertical datasets and estimates historical and ahistorical linear models of all necessary vital rates to create function-based MPMs, including complex IPMs.[...] Historical global models can include size and reproductive status in times t and t − 1 and their interactions.*”


## Time-lagged IPM

Kuss et al. (2008) developed a time-lagged IPM for a demographic system in
which reproduction could not be related to individual size in the same year.
Survival and growth were determined by size at time $t$, whereas reproduction
contributing to the population at $t+1$ depended on size at $t-1$.

Kuss et al. (2008): “demographic transitions to depend on an individual’s
state either in the current year or the year before”

The resulting population projection is

$$
\mathbf{n}_{t+1}
=
\mathbf{P}\mathbf{n}_t
+
\mathbf{F}\mathbf{n}_{t-1}.
$$

The state of the system must therefore contain both the current and previous
population vectors:

$$
\mathbf{N}_t
=
\begin{pmatrix}
\mathbf{n}_t \\
\mathbf{n}_{t-1}
\end{pmatrix}.
$$

Projection can then be written as

$$
\begin{pmatrix}
\mathbf{n}_{t+1} \\
\mathbf{n}_t
\end{pmatrix}
=
\begin{pmatrix}
\mathbf{P} & \mathbf{F} \\
\mathbf{I} & \mathbf{0}
\end{pmatrix}
\begin{pmatrix}
\mathbf{n}_t \\
\mathbf{n}_{t-1}
\end{pmatrix}.
$$

Thus, the historical projection matrix is

$$
\mathbf{K}_{\mathrm{lag}}
=
\begin{pmatrix}
\mathbf{P} & \mathbf{F} \\
\mathbf{I} & \mathbf{0}
\end{pmatrix}.
$$

The four blocks have different roles. $\mathbf{P}$ contains survival-growth
transitions from $t$ to $t+1$, whereas $\mathbf{F}$ produces recruits at
$t+1$ from the population at $t-1$. The identity matrix $\mathbf{I}$ transfers
the current population into the historical position for the next projection,
and $\mathbf{0}$ prevents the stored population from contributing directly to
the following state.

Importantly, $\mathbf{I}$ is only a bookkeeping device. It does not represent
a biological demographic transition.
Kuss et al. (2008): “simply a computational device and not a biological
transition”

### Implementation for *Polygala lewtonii*

In the implementation used here, the survival-growth matrix is called
`Tmat` and corresponds to $\mathbf{P}$ above, while the fecundity matrix is
called `Fmat`. With disturbance state $d$, the model is therefore

$$
\mathbf{K}_{\mathrm{lag}}(d)
=
\begin{pmatrix}
\mathbf{T}_d & \mathbf{F}_d \\
\mathbf{I} & \mathbf{0}
\end{pmatrix}.
$$

For 200 size mesh points, both $\mathbf{T}_d$ and $\mathbf{F}_d$ are
$200\times200$ matrices, producing a $400\times400$ time-lagged
megamatrix.

This first implementation deliberately uses a simplified disturbance
representation. A single disturbance state $d$ is supplied to the demographic
components when constructing each kernel. The purpose of this model is
therefore to isolate the consequences of introducing the one-year fecundity
lag and to reproduce the basic megamatrix formulation of Kuss et al. (2008).

The second historical model removes this simplification and explicitly retains
the different timing of fire effects among the vital rates.

## Environmental historical IPM

For *Polygala lewtonii*, the relevant historical dependence arises naturally
from the timing of fire relative to the annual census. Let

$$
D_{t-1}
$$

denote disturbance during the interval from $t-1$ to $t$, and

$$
D_t
$$

denote disturbance during the interval from $t$ to $t+1$.

Yet again: Ehrlén (2000): “the current phenotype of an individual and its response to
environmental change have been shown to reflect the individual’s history,
i.e., its past phenotype and environment”

These two disturbance states affect different demographic processes.
Survival and growth from $t$ to $t+1$ depend on $D_t$. Recruitment from
flowering stems at $t$ to recruits at $t+1$ also depends on $D_t$. In
contrast, flowering probability and flowering-stem production observed at
$t$ depend on disturbance during the preceding interval, $D_{t-1}$.

The demographic kernel for one projection interval must therefore be
conditional on both disturbance states. With binary disturbance,
$D\in\{0,1\}$, four kernels are required:

$$
\mathbf{K}_{00},\qquad
\mathbf{K}_{01},\qquad
\mathbf{K}_{10},\qquad
\mathbf{K}_{11},
$$

where the first subscript denotes current disturbance $D_t$ and the second
denotes previous disturbance $D_{t-1}$.

For example,

$$
\mathbf{K}_{10}
$$

describes demographic transitions during a year with current fire
$(D_t=1)$ that follows an unburned previous interval $(D_{t-1}=0)$.

Each state-specific kernel can be written as

$$
\mathbf{K}_{ij}
=
\mathbf{P}_i
+
\mathbf{F}_{ij},
$$

where

$$
i=D_t,\qquad j=D_{t-1}.
$$

The survival-growth component depends only on current disturbance,

$$
P_i(y,x)
=
s(x\mid D_t=i)
\,g(y,x\mid D_t=i),
$$

whereas fecundity depends on both disturbance intervals,

$$
F_{ij}(y,x)
=
r(y)\,
b(x\mid D_{t-1}=j,D_t=i).
$$

Here, $r(y)$ describes the recruit-size distribution and
$b(x\mid D_{t-1},D_t)$ represents expected recruitment generated by an
individual of size $x$. In the fitted model, flowering probability and
flowering-stem production contribute the dependence on $D_{t-1}$, whereas
the flowering-stem-to-recruit relationship contributes the dependence on
$D_t$.

### Carrying the disturbance history forward

The model must also specify how disturbance states themselves change through
time. Conditional transition probabilities are estimated from the observed
fire histories:

$$
p_{i|j}
=
P(D_t=i\mid D_{t-1}=j).
$$

For the two-state disturbance process,

$$
p_{0|0}+p_{1|0}=1
$$

and

$$
p_{0|1}+p_{1|1}=1.
$$

The population at time $t$ is separated according to the disturbance state
experienced during the preceding interval:

$$
\mathbf{N}_t
=
\begin{pmatrix}
\mathbf{n}_t^{(0)} \
\mathbf{n}_t^{(1)}
\end{pmatrix},
$$

where

$$
\mathbf{n}_t^{(0)}
$$

contains individuals for which $D_{t-1}=0$, and

$$
\mathbf{n}_t^{(1)}
$$

contains individuals for which $D_{t-1}=1$.

The environmental historical projection is then

$$
\begin{pmatrix}
\mathbf{n}_{t+1}^{(0)} \\
\mathbf{n}_{t+1}^{(1)}
\end{pmatrix}
=
\begin{pmatrix}
p_{0|0}\mathbf{K}_{00}
&
p_{0|1}\mathbf{K}_{01}
\\
p_{1|0}\mathbf{K}_{10}
&
p_{1|1}\mathbf{K}_{11}
\end{pmatrix}
\begin{pmatrix}
\mathbf{n}_t^{(0)} \\
\mathbf{n}_t^{(1)}
\end{pmatrix}.
$$

Thus, the environmental historical megamatrix is

$$
\mathbf{K}_{\mathrm{env}}
=
\begin{pmatrix}
p_{0|0}\mathbf{K}_{00}
&
p_{0|1}\mathbf{K}_{01}
\\
p_{1|0}\mathbf{K}_{10}
&
p_{1|1}\mathbf{K}_{11}
\end{pmatrix}.
$$

After projection, the upper block contains individuals whose new previous
disturbance state is $D_t=0$, while the lower block contains individuals whose
new previous disturbance state is $D_t=1$. In this way, the disturbance
experienced during one projection becomes the historical disturbance state
used in the next.

As in the time-lagged model, 200 size mesh points and two historical states
produce a $400\times400$ megamatrix. However, the interpretation of those
400 states is fundamentally different.

## Time-lagged versus environmental historical IPMs

Both models extend the ordinary IPM by augmenting the population state, but
they retain different kinds of information.

In the time-lagged model,

$$
\mathbf{N}_t
=
\begin{pmatrix}
\mathbf{n}_t \\
\mathbf{n}_{t-1}
\end{pmatrix},
$$

and the lower half of the state vector is a stored copy of the previous
population. The identity block exists solely to carry this information
forward. Fecundity is then applied to the lagged population.

In the environmental historical model,

$$
\mathbf{N}_t
=
\begin{pmatrix}
\mathbf{n}_t^{(D_{t-1}=0)} \\
\mathbf{n}_t^{(D_{t-1}=1)}
\end{pmatrix}.
$$

Here, both halves contain real individuals at time $t$. They are mutually
exclusive components of the current population, distinguished by the
environment experienced during the preceding interval. No identity
bookkeeping block is required. Instead, conditional disturbance-transition
probabilities move individuals between environmental-history states.

The distinction can therefore be summarized as

$$
\text{time-lagged model:}
\qquad
\text{memory of the population at }t-1,
$$

versus

$$
\text{environmental historical model:}
\qquad
\text{memory of the environment during }t-1\rightarrow t.
$$

Neither formulation is identical to a fully individual-history IPM in which
vital rates depend directly on both current and previous individual size,
such as

$$
z_{t+1}
\sim
z_t + z_{t-1}.
$$

In such models, the individual state is defined by the pair
$(z_{t-1},z_t)$, and transitions occur between consecutive state pairs
(Ehrlén 2000; Shefferson et al. 2021). The models implemented here instead
focus on two specific forms of temporal dependence that are directly relevant
to the demographic structure of the *Polygala lewtonii* IPM.


## Data and metadata

```{r}
df_og <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_original_260519.csv')))

df_site <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_sitehist_260519.csv')))

df_meta <- data.frame(
  variable = c(
    'unique_ID', 'quad', 'patch', 'year', 'month', 'date', 'qsurv',
    'stg', 'ht', 'mcd', 'st', 'flst'),
  description = c(
    'Unique identifier for each individual plant; Numeric',
    'Name of circular quadrat with radius of 0.25m; Numeric',
    'Name of an aggregate of several quads; Numeric',
    'Year data were collected; Numeric',
    paste(
      'Month data were collected for March, June, September, and',
      'December. April was also used in the first sampling; Numeric'),
    'Combination of year and month; MMYY',
    paste(
      'Quarterly survival code. 0 or 1 refer to survival from the',
      'previous quarter. If a plant was dropped from monitoring,',
      'survival is NA and these plants should NOT be considered dead.',
      '0 = dead; 1 = alive; 2 = missing; 3 = new adult; 5 = seedling'),
    paste(
      'Stage class from March census. If plant was a seedling during',
      'the previous year it is considered a seedling here.',
      '1 = putative seedling; 2 = vegetative; 3 = reproductive adult'),
    'Height from March census in cm; Numeric',
    'Maximum crown diameter from March census in cm; Numeric',
    'Total number of stems; Numeric',
    'Number of flowering stems; Numeric'),
  stringsAsFactors = FALSE)
```

### Load the completed workdata

The workdata already contains current-year disturbance and disturbance in the
exact previous year. These fields are retained directly rather than rebuilt
from the subset of observed individual rows.

```{r}
df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv'))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.integer(year),
    site = factor(site)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))
```

## Construct the vital-rate data sets

### Survival

```{r}
df_su <- df %>%
  filter(
    !is.na(survives),
    size_t0 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance, stage)
```

### Growth

```{r}
df_gr <- df %>%
  filter(
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance, stage)
```

### Flowering probability and flowering-stem number

Observed recruits are retained because some recruits flowered.

```{r}
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance_prev)) %>%
  dplyr::select(
    id, site, year, size_t0, flower, fl_nr, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3, disturbance_prev, stage, recruits)

df_fl_cond <- df_fl %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr == round(fl_nr))
```

### Quadrat-level flowering stems and recruits

Total flowering stems in quadrat $q$ during year $t$ predict recruits in the
same quadrat during year $t+1$. A quadrat transition is retained only when the
same quadrat has individual records in both years.

```{r}
df_quad_year_sampled <- df %>%
  distinct(site, quad, year)

df_fs2r <- df %>%
  group_by(site, quad, year) %>%
  summarise(
    fs_t0 = sum(fl_nr, na.rm = TRUE),
    disturbance = if_else(
      all(is.na(disturbance)),
      NA_real_,
      max(disturbance, na.rm = TRUE)),
    .groups = 'drop') %>%
  mutate(
    year_t1 = year + 1) %>%
  semi_join(
    df_quad_year_sampled %>%
      transmute(
        site,
        quad,
        year_t1 = year),
    by = c(
      'site',
      'quad',
      'year_t1')) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(
        site,
        quad,
        year,
        name = 're_t1'),
    by = c(
      'site',
      'quad',
      'year_t1' = 'year')) %>%
  mutate(
    re_t1 = replace_na(re_t1, 0L),
    site = factor(site),
    fs_t0_log = log1p(fs_t0),
    re_t1_log = log1p(re_t1)) %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !is.na(disturbance),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re))

df_fs2r %>%
  summarise(
    quadrats = n_distinct(quad),
    transitions = n(),
    zero_stems = sum(fs_t0 == 0),
    zero_recruits = sum(re_t1 == 0),
    burned_transitions = sum(disturbance == 1),
    unburned_transitions = sum(disturbance == 0))
```

## Fit the mean vital-rate models

Candidate models use intercept-only, fire-only, and first- through third-order
size polynomials, each with and without the relevant fire term. Empty model-set
vectors select the model with the lowest AICc.

### Survival model

```{r}
mod_su_00 <- glm(
  survives ~ 1,
  data = df_su,
  family = binomial)

mod_su_0 <- glm(
  survives ~ disturbance,
  data = df_su,
  family = binomial)

mod_su_10 <- glm(
  survives ~ logvol_t0,
  data = df_su,
  family = binomial)

mod_su_1 <- glm(
  survives ~ logvol_t0 + disturbance,
  data = df_su,
  family = binomial)

mod_su_20 <- glm(
  survives ~ logvol_t0 + logvol_t0_2,
  data = df_su,
  family = binomial)

mod_su_2 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance,
  data = df_su,
  family = binomial)

mod_su_30 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_su,
  family = binomial)

mod_su_3 <- glm(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance,
  data = df_su,
  family = binomial)

mods_su <- list(
  mod_su_00, mod_su_0, mod_su_10, mod_su_1,
  mod_su_20, mod_su_2, mod_su_30, mod_su_3)

mods_su_dAICc <- bbmle::AICctab(
  mods_su,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_su_sorted <- order(mods_su_dAICc)

if (length(v_mod_set_su) == 0) {
  mod_su_index_best <- mods_su_sorted[1]
} else {
  mod_su_index_best <- v_mod_set_su + 1
}

v_mod_su_index <- floor((mod_su_index_best - 1) / 2)
mod_su_best <- mods_su[[mod_su_index_best]]

summary(mod_su_best)
mods_su_dAICc
```

### Growth model

```{r}
mod_gr_00 <- lm(
  logvol_t1 ~ 1,
  data = df_gr)

mod_gr_0 <- lm(
  logvol_t1 ~ disturbance,
  data = df_gr)

mod_gr_10 <- lm(
  logvol_t1 ~ logvol_t0,
  data = df_gr)

mod_gr_1 <- lm(
  logvol_t1 ~ logvol_t0 + disturbance,
  data = df_gr)

mod_gr_20 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2,
  data = df_gr)

mod_gr_2 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance,
  data = df_gr)

mod_gr_30 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_gr)

mod_gr_3 <- lm(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance,
  data = df_gr)

mods_gr <- list(
  mod_gr_00, mod_gr_0, mod_gr_10, mod_gr_1,
  mod_gr_20, mod_gr_2, mod_gr_30, mod_gr_3)

mods_gr_dAICc <- bbmle::AICctab(
  mods_gr,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_gr_sorted <- order(mods_gr_dAICc)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_best <- mods_gr_sorted[1]
} else {
  mod_gr_index_best <- v_mod_set_gr + 1
}

v_mod_gr_index <- floor((mod_gr_index_best - 1) / 2)
mod_gr_best <- mods_gr[[mod_gr_index_best]]

summary(mod_gr_best)
mods_gr_dAICc
```

### Growth variance

Growth variance is modeled as an exponential function of the fitted growth
mean.

```{r}
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(
    maxiter = 1000,
    tol = 1e-6,
    warnOnly = TRUE))

summary(mod_gr_var)
```

### Flowering-probability model

```{r}
mod_fl_00 <- glm(
  flower ~ 1,
  data = df_fl,
  family = binomial)

mod_fl_0 <- glm(
  flower ~ disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_10 <- glm(
  flower ~ logvol_t0,
  data = df_fl,
  family = binomial)

mod_fl_1 <- glm(
  flower ~ logvol_t0 + disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_20 <- glm(
  flower ~ logvol_t0 + logvol_t0_2,
  data = df_fl,
  family = binomial)

mod_fl_2 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + disturbance_prev,
  data = df_fl,
  family = binomial)

mod_fl_30 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_fl,
  family = binomial)

mod_fl_3 <- glm(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev,
  data = df_fl,
  family = binomial)

mods_fl <- list(
  mod_fl_00, mod_fl_0, mod_fl_10, mod_fl_1,
  mod_fl_20, mod_fl_2, mod_fl_30, mod_fl_3)

mods_fl_dAICc <- bbmle::AICctab(
  mods_fl,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_fl_sorted <- order(mods_fl_dAICc)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_best <- mods_fl_sorted[1]
} else {
  mod_fl_index_best <- v_mod_set_fl + 1
}

v_mod_fl_index <- floor((mod_fl_index_best - 1) / 2)
mod_fl_best <- mods_fl[[mod_fl_index_best]]

summary(mod_fl_best)
mods_fl_dAICc
```

### Flowering-stem-number model

The conditional model includes only observed flowering plants with positive
integer flowering-stem counts.

```{r}
mod_fl_n_00 <- MASS::glm.nb(
  fl_nr ~ 1,
  data = df_fl_cond)

mod_fl_n_0 <- MASS::glm.nb(
  fl_nr ~ disturbance_prev,
  data = df_fl_cond)

mod_fl_n_10 <- MASS::glm.nb(
  fl_nr ~ logvol_t0,
  data = df_fl_cond)

mod_fl_n_1 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_20 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2,
  data = df_fl_cond)

mod_fl_n_2 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_30 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3,
  data = df_fl_cond)

mod_fl_n_3 <- MASS::glm.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev,
  data = df_fl_cond)

mods_fl_n <- list(
  mod_fl_n_00, mod_fl_n_0, mod_fl_n_10, mod_fl_n_1,
  mod_fl_n_20, mod_fl_n_2, mod_fl_n_30, mod_fl_n_3)

mods_fl_n_dAICc <- bbmle::AICctab(
  mods_fl_n,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_fl_n_sorted <- order(mods_fl_n_dAICc)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_best <- mods_fl_n_sorted[1]
} else {
  mod_fl_n_index_best <- v_mod_set_fl_n + 1
}

v_mod_fl_n_index <- floor((mod_fl_n_index_best - 1) / 2)
mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_best]]

summary(mod_fl_n_best)
mods_fl_n_dAICc
```

### Flowering-stem-to-recruit model

The recruitment candidates allow a seedbank intercept, a current-fire effect,
a flowering-stem effect, and their interaction. No year or site random effects
are fitted.

```{r}
mod_re_00 <- lm(
  re_t1_log ~ 1,
  data = df_fs2r)

mod_re_0 <- lm(
  re_t1_log ~ disturbance,
  data = df_fs2r)

mod_re_10 <- lm(
  re_t1_log ~ fs_t0_log,
  data = df_fs2r)

mod_re_1 <- lm(
  re_t1_log ~ fs_t0_log + disturbance,
  data = df_fs2r)

mod_re_20 <- lm(
  re_t1_log ~ fs_t0_log * disturbance,
  data = df_fs2r)

mods_re <- list(
  mod_re_00,
  mod_re_0,
  mod_re_10,
  mod_re_1,
  mod_re_20)

mods_re_dAICc <- bbmle::AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_re_sorted <- order(mods_re_dAICc)

if (length(v_mod_set_re) == 0) {
  mod_re_index_best <- mods_re_sorted[1]
} else {
  mod_re_index_best <- v_mod_set_re + 1
}

v_mod_re_index <- mod_re_index_best - 1
mod_re_best <- mods_re[[mod_re_index_best]]

summary(mod_re_best)
mods_re_dAICc
```

### Recruitment-model support

```{r}
df_recruit_model_support <- tibble(
  model = c(
    'Intercept only',
    'Disturbance only',
    'Flowering stems',
    'Flowering stems + disturbance',
    'Full interaction'),
  dAICc = as.numeric(mods_re_dAICc)) %>%
  mutate(
    weight = exp(-0.5 * dAICc) /
      sum(exp(-0.5 * dAICc))) %>%
  arrange(dAICc)

knitr::kable(
  df_recruit_model_support,
  digits = c(NA, 2, 3),
  caption = paste(
    'AICc support for quadrat-level flowering-stem-to-recruit',
    'models.'))
```

The full interaction model is selected. Its fitted form is

$$
\log(1 + R_{q,t+1}) =
\beta_0 + \beta_1\log(1 + M_{q,t}) + \beta_d d_{q,t}
+ \beta_{1d}\log(1 + M_{q,t})d_{q,t} + \varepsilon,
$$

where $M_{q,t}$ is total flowering stems, $R_{q,t+1}$ is the number of recruits
in the same quadrat in the following year, and $d_{q,t}$ indicates current
fire.

### Recruitment slopes and uncertainty

```{r}
re_coef <- coef(mod_re_20)
re_vcov <- vcov(mod_re_20)

slope_unburned <- re_coef['fs_t0_log']

se_unburned <- sqrt(
  re_vcov[
    'fs_t0_log',
    'fs_t0_log'])

slope_burned <-
  re_coef['fs_t0_log'] +
  re_coef['fs_t0_log:disturbance']

se_burned <- sqrt(
  re_vcov[
    'fs_t0_log',
    'fs_t0_log'] +
  re_vcov[
    'fs_t0_log:disturbance',
    'fs_t0_log:disturbance'] +
  2 * re_vcov[
    'fs_t0_log',
    'fs_t0_log:disturbance'])

df_recruit_slopes <- tibble(
  disturbance = c(0, 1),
  slope = c(
    slope_unburned,
    slope_burned),
  lower_95 = c(
    slope_unburned - 1.96 * se_unburned,
    slope_burned - 1.96 * se_burned),
  upper_95 = c(
    slope_unburned + 1.96 * se_unburned,
    slope_burned + 1.96 * se_burned))

knitr::kable(
  df_recruit_slopes,
  digits = 3,
  caption = paste(
    'Flowering-stem slopes from the full interaction recruitment',
    'model.'))
```

Recruitment can occur when no flowering stems are observed. This is consistent
with emergence from the persistent seedbank.

```{r}
df_fs2r_zero_stems <- df_fs2r %>%
  filter(fs_t0 == 0) %>%
  group_by(disturbance) %>%
  summarise(
    transitions = n(),
    transitions_with_recruits = sum(re_t1 > 0),
    proportion_with_recruits = mean(re_t1 > 0),
    total_recruits = sum(re_t1),
    mean_recruits = mean(re_t1),
    median_recruits = median(re_t1),
    max_recruits = max(re_t1),
    .groups = 'drop')

knitr::kable(
  df_fs2r_zero_stems,
  digits = 3,
  caption = paste(
    'Recruitment in quadrat transitions with no recorded flowering',
    'stems.'))
```

## Plot the selected vital-rate models

```{r}
make_mean_prediction <- function(data, model, disturbance_var) {
  x <- seq(
    min(data$logvol_t0, na.rm = TRUE),
    max(data$logvol_t0, na.rm = TRUE),
    length.out = 200)

  pred <- expand_grid(
    logvol_t0 = x,
    disturbance_value = c(0, 1)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_plot = factor(
        disturbance_value,
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  pred[[disturbance_var]] <- pred$disturbance_value
  pred$prediction <- predict(
    model,
    newdata = pred,
    type = 'response')

  pred
}

make_mean_plot <- function(
    data, model, response, disturbance_var, title, y_lab,
    probability = FALSE) {
  pred <- make_mean_prediction(
    data = data,
    model = model,
    disturbance_var = disturbance_var)

  data_plot <- data %>%
    mutate(
      disturbance_plot = factor(
        .data[[disturbance_var]],
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  point_layer <- if (probability) {
    geom_jitter(
      data = data_plot,
      aes(
        x = logvol_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7,
      width = 0.05,
      height = 0.05)
  } else {
    geom_point(
      data = data_plot,
      aes(
        x = logvol_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7)
  }

  p <- ggplot() +
    point_layer +
    geom_line(
      data = pred,
      aes(
        x = logvol_t0,
        y = prediction,
        color = disturbance_plot),
      linewidth = 0.8) +
    scale_color_manual(
      values = c(
        'No fire' = 'black',
        'Fire' = 'red')) +
    theme_bw() +
    labs(
      title = title,
      subtitle = v_ggp_suffix,
      x = expression('log(volume)'[t0]),
      y = y_lab,
      color = NULL)

  if (probability) {
    p <- p + coord_cartesian(ylim = c(0, 1))
  }

  p
}
```

```{r, fig.width=9, fig.height=7, fig.cap='Selected mean survival, growth, flowering-probability, and conditional flowering-stem models. Current fire is used for survival and growth, and previous-year fire is used for flowering.'}
fig_su_mean <- make_mean_plot(
  data = df_su,
  model = mod_su_best,
  response = 'survives',
  disturbance_var = 'disturbance',
  title = 'Mean survival',
  y_lab = 'Survival probability',
  probability = TRUE)

fig_gr_mean <- make_mean_plot(
  data = df_gr,
  model = mod_gr_best,
  response = 'logvol_t1',
  disturbance_var = 'disturbance',
  title = 'Mean growth',
  y_lab = expression('log(volume)'[t1])) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = 'dashed')

fig_fl_mean <- make_mean_plot(
  data = df_fl,
  model = mod_fl_best,
  response = 'flower',
  disturbance_var = 'disturbance_prev',
  title = 'Mean flowering probability',
  y_lab = 'Flowering probability',
  probability = TRUE)

fig_fl_n_mean <- make_mean_plot(
  data = df_fl_cond,
  model = mod_fl_n_best,
  response = 'fl_nr',
  disturbance_var = 'disturbance_prev',
  title = 'Mean flowering-stem number',
  y_lab = 'Number of flowering stems')

fig_vital_rates_mean <-
  fig_su_mean + fig_gr_mean + fig_fl_mean + fig_fl_n_mean +
  plot_layout(ncol = 2, guides = 'collect') &
  theme(legend.position = 'bottom')

fig_vital_rates_mean
```

### Flowering-stem-to-recruit relationship

```{r, fig.width=7, fig.height=5, fig.cap='Quadrat-level flowering stems in year t and recruits in the same quadrat in year t + 1 under current-fire and no-fire conditions.'}
df_fs2r_pred <- expand_grid(
  fs_t0_log = seq(
    0,
    max(df_fs2r$fs_t0_log, na.rm = TRUE),
    length.out = 200),
  disturbance = c(0, 1)) %>%
  mutate(
    re_t1_log = predict(
      mod_re_best,
      newdata = .),
    fs_t0 = expm1(fs_t0_log),
    re_t1 = pmax(0, expm1(re_t1_log)),
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c('No fire', 'Fire')))

df_fs2r_plot <- df_fs2r %>%
  mutate(
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c('No fire', 'Fire')))

fig_fs2r_mean <- ggplot(
  df_fs2r_plot,
  aes(
    x = fs_t0,
    y = re_t1,
    color = disturbance_plot)) +
  geom_point(alpha = 0.4, size = 1.3) +
  geom_line(
    data = df_fs2r_pred,
    aes(
      x = fs_t0,
      y = re_t1,
      color = disturbance_plot),
    linewidth = 0.8) +
  scale_x_continuous(
  transform = scales::transform_pseudo_log(
    sigma = 1)) +
  scale_color_manual(
    values = c(
      'No fire' = 'black',
      'Fire' = 'red')) +
  theme_bw() +
  labs(
    title = 'Mean flowering-stem-to-recruit relationship',
    subtitle = v_ggp_suffix,
    x = expression('Total flowering stems'[t]),
    y = expression('Recruits'[t + 1]),
    color = NULL)

fig_fs2r_mean
```

## Recruit size distribution

```{r}
df_recr <- df %>%
  filter(
    recruits == 1,
    volume_t0 > 0,
    is.finite(logvol_t0))

rc_sz <- df_recr %>%
  summarise(
    mean = mean(logvol_t0, na.rm = TRUE),
    sd = sd(logvol_t0, na.rm = TRUE))

rc_sz
```

## Extract the selected model parameters

The extraction functions return zero for terms that are absent from the
selected model. The same IPM functions can therefore use any selected candidate
without manual coefficient changes.

```{r}
size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logvol_t0',
  b2 = 'logvol_t0_2',
  b3 = 'logvol_t0_3',
  bd = 'disturbance')

lagged_size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logvol_t0',
  b2 = 'logvol_t0_2',
  b3 = 'logvol_t0_3',
  bd = 'disturbance_prev')

extract_terms <- function(model, prefix, term_map) {
  coefs <- stats::coef(model)

  values <- vapply(term_map, function(term_i) {
    if (term_i %in% names(coefs)) {
      unname(coefs[[term_i]])
    } else {
      0
    }
  }, numeric(1))

  names(values) <- paste0(prefix, names(term_map))
  values
}

extract_coef <- function(model, term, default = 0) {
  coefs <- stats::coef(model)

  if (term %in% names(coefs)) {
    unname(coefs[[term]])
  } else {
    default
  }
}

grow_var_coef <- stats::coef(mod_gr_var)

mesh_limits <- range(
  c(
    df_gr$logvol_t0,
    df_gr$logvol_t1,
    df_recr$logvol_t0),
  na.rm = TRUE,
  finite = TRUE)

mesh_limits <- c(
  mesh_limits[1] - 0.1,
  mesh_limits[2] + 0.1)

pars_mean <- as.list(c(
  extract_terms(
    mod_su_best,
    'surv_',
    size_term_map),
  extract_terms(
    mod_gr_best,
    'grow_',
    size_term_map),
  extract_terms(
    mod_fl_best,
    'fl_',
    lagged_size_term_map),
  extract_terms(
    mod_fl_n_best,
    'fln_',
    lagged_size_term_map),
  re_b0 = extract_coef(
    mod_re_best,
    '(Intercept)'),
  re_b1 = extract_coef(
    mod_re_best,
    'fs_t0_log'),
  re_bd = extract_coef(
    mod_re_best,
    'disturbance'),
  re_bf = extract_coef(
    mod_re_best,
    'fs_t0_log:disturbance'),
  re_sigma = stats::sigma(mod_re_best),
  recr_mean = rc_sz$mean,
  recr_sd = rc_sz$sd,
  grow_var_a = unname(grow_var_coef[['a']]),
  grow_var_b = unname(grow_var_coef[['b']]),
  fs_ref = mean(df_fs2r$fs_t0, na.rm = TRUE),
  L = mesh_limits[1],
  U = mesh_limits[2],
  mat_siz = 200,
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_fl_index = v_mod_fl_index,
  mod_fl_n_index = v_mod_fl_n_index,
  mod_re_index = v_mod_re_index))

pars_mean_table <- tibble(
  parameter = names(pars_mean),
  value = as.numeric(unlist(pars_mean)))

knitr::kable(
  pars_mean_table,
  digits = 5,
  caption = 'Parameters passed to the mean IPM.')
```

## Define the IPM functions

The extracted flowering coefficients still come from models fitted with
`disturbance_prev`. From this point onward, however, their disturbance
coefficient is used with the same generic `disturbance` argument as all other
vital rates.

### General parameter and linear-predictor functions

```{r}
get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    pars[[par_name]]
  } else {
    default
  }
}

vital_lp <- function(x, pars, prefix, disturbance = 0) {
  get_par(pars, paste0(prefix, 'b0')) +
    get_par(pars, paste0(prefix, 'b1')) * x +
    get_par(pars, paste0(prefix, 'b2')) * x^2 +
    get_par(pars, paste0(prefix, 'b3')) * x^3 +
    get_par(pars, paste0(prefix, 'bd')) * disturbance
}
```

###Survival
```{r}
sx <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'surv_',
    disturbance = disturbance))
}
```

###Growth
```{r}
grow_mu <- function(x, pars, disturbance = 0) {
  vital_lp(
    x = x,
    pars = pars,
    prefix = 'grow_',
    disturbance = disturbance)
}

grow_sd <- function(mean_value, pars) {
  variance <- get_par(pars, 'grow_var_a') *
    exp(get_par(pars, 'grow_var_b') * mean_value)

  sqrt(pmax(variance, .Machine$double.eps))
}

gxy <- function(x, y, pars, disturbance = 0) {
  mean_value <- grow_mu(
    x = x,
    pars = pars,
    disturbance = disturbance)

  sd_value <- grow_sd(
    mean_value = mean_value,
    pars = pars)

  dnorm(y, mean = mean_value, sd = sd_value)
}
```


### Flowering probability and flowering-stem number

For this initial Kuss-style implementation, disturbance is intentionally
treated as a single common state within the IPM. Although the flowering
models were fitted using disturbance_prev, their extracted disturbance
coefficients are evaluated using the same generic disturbance state used
for the other vital rates.

This is a simplifying assumption used to implement and examine the
time-lagged matrix structure. It does not yet reproduce the actual
difference in fire timing among the vital rates.


```{r}
fl_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fl_',
    disturbance = disturbance))
}

fln_x <- function(x, pars, disturbance = 0) {
  exp(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fln_',
    disturbance = disturbance))
}

fs_x <- function(x, pars, disturbance = 0) {
  fl_x(
    x = x,
    pars = pars,
    disturbance = disturbance) *
    fln_x(
      x = x,
      pars = pars,
      disturbance = disturbance)
}
```

###Recruitment
```{r}
predict_recruits_from_stems <- function(
    flowering_stems, pars, disturbance = 0) {
  stems_nonnegative <- pmax(flowering_stems, 0)

  slope <- get_par(pars, 're_b1') +
    get_par(pars, 're_bf') * disturbance

  eta <- get_par(pars, 're_b0') +
    get_par(pars, 're_bd') * disturbance +
    slope * log1p(stems_nonnegative)

  recruits <- expm1(
    eta + 0.5 * get_par(pars, 're_sigma')^2)

  pmax(recruits, 0)
}

recruits_per_stem <- function(pars, disturbance = 0) {
  fs_ref <- get_par(pars, 'fs_ref', 0)

  if (!is.finite(fs_ref) || fs_ref <= 0) {
    return(0)
  }

  predict_recruits_from_stems(
    flowering_stems = fs_ref,
    pars = pars,
    disturbance = disturbance) / fs_ref
}

rx <- function(x, pars, disturbance = 0) {
  fs_x(
    x = x,
    pars = pars,
    disturbance = disturbance) *
    recruits_per_stem(
      pars = pars,
      disturbance = disturbance)
}

recr_y <- function(y, pars) {
  dnorm(
    y,
    mean = get_par(pars, 'recr_mean'),
    sd = get_par(pars, 'recr_sd'))
}

fy <- function(y, x, pars, h, disturbance = 0) {
  Pvec <- recr_y(y, pars)
  recruit_mass <- sum(Pvec * h)

  if (!is.finite(recruit_mass) || recruit_mass <= 0) {
    stop('Recruit size distribution has no mass inside the mesh')
  }

  Pvec <- Pvec / recruit_mass

  Rvec <- rx(
    x = x,
    pars = pars,
    disturbance = disturbance)

  outer(Pvec, Rvec) * h
}
```

## Construct the time-lagged IPM
The standard mean IPM combines the survival-growth and fecundity matrices
within a single projection matrix,

$$
\mathbf{K} = \mathbf{P} + \mathbf{F}.
$$

In the code below, `Tmat` corresponds to the survival-growth matrix
$\mathbf{P}$ of Kuss et al. (2008). For comparison with the original
ahistorical model, `k_yx` retains the standard matrix

$$
\mathbf{K} = \mathbf{P} + \mathbf{F}.
$$

The historical model instead places survival-growth and fecundity in
different blocks of a larger projection matrix:

$$
\mathbf{K}_H =
\begin{pmatrix}
\mathbf{P} & \mathbf{F} \\
\mathbf{I} & \mathbf{0}
\end{pmatrix}.
$$

With 200 mesh points, the ordinary IPM is a $200 \times 200$ matrix,
whereas the historical IPM is a $400 \times 400$ matrix.

```{r}
kernel <- function(pars, disturbance = 0) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  Fmat <- fy(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance)

  Smat <- sx(
    x = y,
    pars = pars,
    disturbance = disturbance)

  Gmat <- matrix(0, n, n)
  Gmat[] <- t(
    outer(
      y,
      y,
      Vectorize(function(x, y) {
        gxy(
          x = x,
          y = y,
          pars = pars,
          disturbance = disturbance)
      }))) * h

  Tmat <- matrix(0, n, n)

  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  k_yx <- Fmat + Tmat

  Imat <- diag(n)
  Zmat <- matrix(0, n, n)

  K_hist <- rbind(
    cbind(Tmat, Fmat),
    cbind(Imat, Zmat))

  list(
    k_yx = k_yx,
    K_hist = K_hist,
    Fmat = Fmat,
    Tmat = Tmat,
    Imat = Imat,
    Zmat = Zmat,
    Gmat = Gmat,
    Smat = Smat,
    meshpts = y,
    h = h)
}
```


The lower-left identity matrix is not a demographic transition. It is a
bookkeeping operator that transfers the current population vector into the
historical component of the state vector for the following projection
(Kuss et al. 2008).

##Historical asymptotic population growth rate

Population growth is calculated from the dominant eigenvalue of the full
historical megamatrix rather than from the ordinary
$\mathbf{P}+\mathbf{F}$ matrix.

```{r}
lambda_ipm <- function(pars, disturbance = 0) {
  K <- kernel(
    pars = pars,
    disturbance = disturbance)$K_hist

  Re(eigen(K, only.values = TRUE)$values[1])
}
```

##Observed structures
### Mean observed size distribution

```{r}
make_initial_n_mean <- function(pars, df_init = df) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  breaks <- seq(L, U, length.out = n + 1)
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  recruit_density <- recr_y(y, pars)
  recruit_density <- recruit_density / sum(recruit_density * h)
  
  years_i <- df_init %>%
    filter(!is.na(year)) %>%
    distinct(year) %>%
    arrange(year) %>%
    pull(year)
  
  counts_by_year <- purrr::map(years_i, function(year_i) {
    df_i <- df_init %>%
      filter(year == year_i)
    
    x <- df_i %>%
      filter(is.finite(logvol_t0)) %>%
      pull(logvol_t0)
    
    size_counts <- hist(
      pmin(pmax(x, L), U),
      breaks = breaks,
      plot = FALSE,
      include.lowest = TRUE)$counts
    
    n_unsized_recruits <- df_i %>%
      summarise(
        n = sum(
          !is.finite(logvol_t0) & recruits == 1,
          na.rm = TRUE)) %>%
      pull(n)
    
    size_counts + n_unsized_recruits * recruit_density * h
  })
  
  mean_counts <- Reduce(`+`, counts_by_year) / length(counts_by_year)
  mean_counts / h
}
```

### Quadrat-level flowering-stem reference
```{r}
# The flowering-stem-to-recruit model is fitted to quadrat-year totals.
# Its flowering-stem reference must therefore also be on the quadrat scale.
df_stem_reference <- df_fs2r %>%
  group_by(disturbance) %>%
  summarise(
    fs_ref = mean(fs_t0),
    n_quadrat_transitions = n(),
    .groups = "drop")

get_stem_reference <- function(disturbance_i) {
  df_stem_reference %>%
    filter(disturbance == disturbance_i) %>%
    pull(fs_ref)
}

mean_stems_no_fire <- get_stem_reference(0)
mean_stems_fire <- get_stem_reference(1)

df_stem_reference
```


### Mean disturbance exposure
```{r}
# Each sampled quadrat-year receives equal weight. The IPM uses one shared
# disturbance state for all vital rates.
df_disturbance_regime <- df %>%
  filter(
    !is.na(quad),
    !is.na(year),
    !is.na(disturbance)) %>%
  distinct(
    quad,
    year,
    disturbance) %>%
  count(
    disturbance,
    name = 'n_quadrat_years') %>%
  complete(
    disturbance = c(0, 1),
    fill = list(n_quadrat_years = 0)) %>%
  mutate(
    p_state = n_quadrat_years /
      sum(n_quadrat_years)) %>%
  arrange(disturbance)

df_disturbance_regime

get_state_probability <- function(disturbance_i) {
  df_disturbance_regime %>%
    filter(disturbance == disturbance_i) %>%
    pull(p_state)
}

p_0 <- get_state_probability(0)
p_1 <- get_state_probability(1)

mean_stems_regime <-
  p_0 * mean_stems_no_fire +
  p_1 * mean_stems_fire

```


## Disturbance-specific historical kernels

Separate historical megamatrices are constructed for undisturbed and burned
conditions. The same disturbance state is applied to all vital-rate
functions in this initial implementation.

The flowering-stem-to-recruit relationship was fitted at the quadrat level.
The mean number of flowering stems is therefore also calculated at the
quadrat level separately for burned and unburned transitions.


```{r}
pars_0 <- pars_mean
pars_0$fs_ref <- mean_stems_no_fire

pars_1 <- pars_mean
pars_1$fs_ref <- mean_stems_fire

# All mixture components use the same expected total flowering-stem production
# under the observed disturbance regime.
pars_regime <- pars_mean
pars_regime$fs_ref <- mean_stems_regime

K_0 <- kernel(
  pars = pars_0,
  disturbance = 0)$K_hist

K_1 <- kernel(
  pars = pars_1,
  disturbance = 1)$K_hist

K_regime_0 <- kernel(
  pars = pars_regime,
  disturbance = 0)$K_hist

K_regime_1 <- kernel(
  pars = pars_regime,
  disturbance = 1)$K_hist

K_regime <-
  p_0 * K_regime_0 +
  p_1 * K_regime_1

```


## Projection of the historical population vector

The historical state vector contains two size distributions,

$$
\mathbf{N}_t
=
\begin{pmatrix}
\mathbf{n}_t \\
\mathbf{n}_{t-1}
\end{pmatrix}.
$$

For the mean IPM, both components are initialized using the mean observed
size distribution. Thus, at the first projection,

$$
\mathbf{n}_{t-1} = \mathbf{n}_t = \bar{\mathbf{n}}.
$$

This initialization implies that the first projected population is
identical to that obtained from the corresponding ahistorical IPM because

$$
\mathbf{P}\mathbf{n}_t +
\mathbf{F}\mathbf{n}_{t-1}
=
(\mathbf{P}+\mathbf{F})\mathbf{n}_t
$$

when $\mathbf{n}_{t-1}=\mathbf{n}_t$.

Only the upper half of the projected state vector represents the population
at $t+1$. The lower half is the stored population at $t$ and must therefore
not be counted as additional individuals.

```{r}
project_kernel <- function(K, pars, label, df_init = df) {
  n <- as.integer(pars$mat_siz)
  n_current <- make_initial_n_mean(pars, df_init)
  n_previous <- n_current
  n_hist <- c(n_current, n_previous)

  h <- (pars$U - pars$L) / n
  n_initial <- sum(n_current) * h

  n_proj_hist <- K %*% n_hist
  n_next <- n_proj_hist[1:n]
  n_projected <- sum(n_next) * h

  tibble(
    lambda_type = label,
    asym_lambda = Re(eigen(K, only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial,
    n_initial = n_initial,
    n_projected = n_projected)
}


```


The asymptotic growth rate is calculated from the complete historical
megamatrix. In contrast, the one-step projected growth rate describes the
change from the observed mean population distribution during the first
projection. Because the initial current and historical size distributions
are identical, the first projection is equivalent to the corresponding
ahistorical projection. Historical effects emerge during subsequent
iterations and in the asymptotic growth rate.

The historical asymptotic growth rates were approximately 0.956 under
undisturbed conditions, 0.646 under fire, and 0.930 under the expected
disturbance regime. All three values were below one, indicating declining
long-term population growth under these deterministic mean conditions.


```{r}
df_lambda_states <- bind_rows(
  make_state_lambda(
    K = K_0,
    pars = pars_0,
    label = 'Undisturbed',
    disturbance = 0,
    p_state = p_0),
  make_state_lambda(
    K = K_1,
    pars = pars_1,
    label = 'Fire',
    disturbance = 1,
    p_state = p_1))

recruits_ref_regime <-
  p_0 * predict_recruits_from_stems(
    flowering_stems = mean_stems_regime,
    pars = pars_regime,
    disturbance = 0) +
  p_1 * predict_recruits_from_stems(
    flowering_stems = mean_stems_regime,
    pars = pars_regime,
    disturbance = 1)

df_lambda_regime <- project_kernel(
  K = K_regime,
  pars = pars_regime,
  label = 'Expected disturbance regime') %>%
  mutate(
    disturbance = NA_real_,
    observed_state_probability = 1,
    flowering_stems_ref = mean_stems_regime,
    recruits_ref = recruits_ref_regime,
    recruits_per_stem = if_else(
      flowering_stems_ref > 0,
      recruits_ref / flowering_stems_ref,
      0))

df_lambda_mean <- bind_rows(
  df_lambda_states,
  df_lambda_regime)

knitr::kable(
  df_lambda_mean,
  digits = 3,
  caption = paste(
    'Asymptotic historical and one-step projected population growth',
    'rates under the disturbance scenarios.'))
```

### Observed annual abundance change
```{r}
# Counts include sized plants and recruits without a usable size measurement.
# Only quadrats with individual records in both consecutive years are retained.
df_counts_quad_year <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year)) %>%
  group_by(site, quad, year) %>%
  summarise(
    n_sized = sum(
      is.finite(logvol_t0),
      na.rm = TRUE),
    n_unsized_recruits = sum(
      !is.finite(logvol_t0) & recruits == 1,
      na.rm = TRUE),
    n_t0 = n_sized + n_unsized_recruits,
    .groups = "drop") %>%
  arrange(site, quad, year) %>%
  group_by(site, quad) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n_t0),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_obs_growth_year <- df_counts_quad_year %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(n_t0),
    n_t1 = sum(n_t1),
    n_quadrats = n(),
    n_sites = n_distinct(site),
    obs_lambda = n_t1 / n_t0,
    .groups = "drop") %>%
  filter(n_t0 > 0)

df_obs_growth_year


df_obs_growth_summary <- df_obs_growth_year %>%
  summarise(
    n_transitions = n(),
    lambda_arithmetic = mean(obs_lambda, na.rm = TRUE),
    lambda_geometric = exp(mean(log(obs_lambda), na.rm = TRUE)),
    lambda_median = median(obs_lambda, na.rm = TRUE),
    lambda_min = min(obs_lambda, na.rm = TRUE),
    lambda_max = max(obs_lambda, na.rm = TRUE))

df_obs_growth_summary

df_model_lines <- df_lambda_mean %>%
  dplyr::select(lambda_type, asym_lambda, proj_lambda) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'metric',
    values_to = 'lambda') %>%
  mutate(
    metric = recode(
      metric,
      asym_lambda = 'Historical asymptotic',
      proj_lambda = 'One-step projected')
  )

x_point <- max(df_obs_growth_year$year, na.rm = TRUE) + 0.4

fig_lambda_hist <- ggplot(
  df_obs_growth_year,
  aes(x = year, y = obs_lambda)) +
  geom_hline(
    yintercept = 1,
    linetype = 'dotted') +
  geom_line(color = 'black') +
  geom_point(size = 2, color = 'black') +
  geom_hline(
    data = df_model_lines,
    aes(
      yintercept = lambda,
      color = lambda_type),
    linewidth = 0.8) +
  geom_point(
    data = df_model_lines,
    aes(
      x = x_point,
      y = lambda,
      color = lambda_type,
      shape = metric),
    size = 3) +
  expand_limits(x = x_point + 1) +
  theme_bw() +
  labs(
    title = 'Observed annual growth and historical mean-IPM lambda',
    subtitle = v_ggp_suffix,
    x = 'Year t0',
    y = expression(lambda),
    color = NULL,
    shape = NULL)

fig_lambda_hist
```
