# Yearly historical kuss projection model {#yr-ipm-hist-kuss}
## Introduction

This analysis develops a **year-specific historical integral projection model
(IPM)** for *Polygala lewtonii* at Archbold. The model combines annual variation
in demographic rates with the delayed contribution of reproduction to
recruitment.

Survival, growth, flowering, flowering-stem production, and recruitment are
first modeled separately. Year-specific demographic variation is introduced
through the year effects from the fitted mixed models. This means that each
annual IPM uses the demographic conditions estimated for that particular year,
rather than applying one mean set of vital-rate parameters to the full time
series.

Disturbance is incorporated as a single common fire state within the IPM.
Survival, growth, flowering, flowering-stem production, and recruitment are
therefore evaluated under either

- no disturbance (`disturbance = 0`), or
- disturbance (`disturbance = 1`).

This deliberately places disturbance on the same level for all demographic
processes inside the IPM. The underlying flowering and flower-number models are
still fitted using previous-year disturbance, as in the original demographic
models, but their disturbance coefficients are applied to the common
disturbance state when the IPM is constructed.

The historical structure is needed because flowering in year \(t\) contributes
to recruitment in year \(t+1\). The population state therefore contains both
the current and previous size distributions,

\[
\mathbf{n}^{H}_t =
\begin{pmatrix}
\mathbf{n}_t \\
\mathbf{n}_{t-1}
\end{pmatrix}.
\]

The corresponding historical kernel is

\[
\mathbf{K}^{H}_t =
\begin{pmatrix}
\mathbf{T}_t & \mathbf{F}_t \\
\mathbf{I}   & \mathbf{0}
\end{pmatrix},
\]

where \(\mathbf{T}_t\) contains survival and growth, \(\mathbf{F}_t\) contains
fertility, \(\mathbf{I}\) transfers the current population into the lagged
state, and \(\mathbf{0}\) prevents the lagged state from persisting beyond one
year. Projection therefore gives

\[
\begin{pmatrix}
\mathbf{n}_{t+1} \\
\mathbf{n}_t
\end{pmatrix}
=
\begin{pmatrix}
\mathbf{T}_t & \mathbf{F}_t \\
\mathbf{I}   & \mathbf{0}
\end{pmatrix}
\begin{pmatrix}
\mathbf{n}_t \\
\mathbf{n}_{t-1}
\end{pmatrix}.
\]

For the **year-specific model**, a separate historical kernel is constructed
for each year using that year's demographic parameters. For every year, two
asymptotic scenarios are evaluated: one assuming no fire and one assuming
fire. The dominant eigenvalue of the corresponding historical kernel gives the
year-specific asymptotic population growth rate under that disturbance state.

The yearly model is also used for one-step projections of the observed
population. For each quadrat-year transition, the observed size distribution
in year \(t\) is combined with the observed size distribution from year
\(t-1\). The historical kernel for year \(t\), using the observed disturbance
state of that quadrat, is then applied once to obtain the predicted population
in year \(t+1\).

The projected population growth rate is calculated as

\[
\lambda_{\mathrm{proj},t}
=
\frac{N_{\mathrm{projected},t+1}}
     {N_{\mathrm{observed},t}}.
\]

This differs from the asymptotic growth rate. The projected lambda depends on
the actual observed population structure in years \(t\) and \(t-1\), whereas
the asymptotic lambda is determined by the dominant eigenvalue of the annual
historical kernel and represents the long-term growth rate that would result
if the demographic conditions and disturbance state of that year were held
constant.

Observed population growth from \(t\) to \(t+1\) is then compared with both the
projected and asymptotic growth rates. This allows the year-specific model to
be evaluated against the realized annual population dynamics while retaining
the historical lag between flowering and recruitment.



## Working dataset
```{r}
df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv'))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.integer(year),
    site = factor(site)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))
```

### Controls
```{r}
ctrl_glmer <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))
```

## Vital rates
### Survival
```{r}
df_su <- df %>%
  filter(
    !is.na(survives),
    size_t0 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = factor(disturbance, levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage)


# Survival model ---------------------------------------------------------------
mod_su_0 <- glmer(
  survives ~ disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_10 <- glmer(
  survives ~ logvol_t0 + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_11 <- glmer(
  survives ~ logvol_t0 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_20 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_21 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_30 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_31 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)


mods_su <- list(
  mod_su_0, mod_su_10, mod_su_11, 
  mod_su_20, mod_su_21, mod_su_30, mod_su_31)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef <- coef(mod_su_best)$year

mod_su_best
summary(mod_su_best)
mods_su_dAIC

```


```{r}
# Survival plots by year -------------------------------------------------------
make_su_year_plot <- function(year_i) {
  df_i <- df_su %>% filter(year == year_i)
  
  x <- seq(
    min(df_i$logvol_t0, na.rm = TRUE),
    max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance = levels(df_su$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance = factor(
        disturbance, levels = levels(df_su$disturbance)),
      year = factor(year_i, levels = levels(df_su$year)))
  
  pred_i$survives <- predict(
    mod_su_best, newdata = pred_i,
    type = "response", re.form = NULL)
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      survives = mean(survives, na.rm = TRUE),
      n = n(), .groups = "drop") %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, survives, color = disturbance), size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, survives, color = disturbance,
          linetype = disturbance),
      linewidth = 0.7) +
    scale_color_manual(values = c("0" = "black", "1" = "red")) +
    labs(
      title = year_i,
      x = expression("log(volume)"[t0]),
      y = expression("Survival probability"[t1])) +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = "none")
}

su_yrs <- lapply(levels(df_su$year), make_su_year_plot)
fig_su_years <- wrap_plots(su_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_su_years
```

### Grwoth
```{r}
# Growth data ------------------------------------------------------------------
df_gr <- df %>%
  filter(
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = factor(
      disturbance,
      levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage)



# Growth model -----------------------------------------------------------------
mod_gr_0 <- lmer(
  logvol_t1 ~ disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_10 <- lmer(
  logvol_t1 ~ logvol_t0 + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_11 <- lmer(
  logvol_t1 ~ logvol_t0 + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_20 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_21 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_30 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_31 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)


mods_gr <- list(
  mod_gr_0, mod_gr_10, mod_gr_11, mod_gr_20, mod_gr_21, 
  mod_gr_30, mod_gr_31)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef <- coef(mod_gr_best)$year

mod_gr_best
summary(mod_gr_best)
mods_gr_dAIC
```


```{r}
# Growth plots by year ---------------------------------------------------------
make_gr_year_plot <- function(year_i) {
  df_i <- df_gr %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance = levels(df_gr$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance = factor(
        disturbance, levels = levels(df_gr$disturbance)),
      year = factor(year_i, levels = levels(df_gr$year)))
  
  pred_i <- pred_i %>%
    mutate(
      logvol_t1 = predict(
        mod_gr_best,
        newdata = pred_i,
        re.form = NULL))
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      logvol_t1 = mean(logvol_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, logvol_t1, color = disturbance),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, logvol_t1, color = disturbance,
          linetype = disturbance),
      linewidth = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = 'blue', lty = 2) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = expression('log(volume)'[t1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

gr_yrs <- lapply(levels(df_gr$year), make_gr_year_plot)
fig_gr_years <- wrap_plots(gr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Growth - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_gr_years
```


```{r}
# Growth variance --------------------------------------------------------------
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

### Flowering probability
```{r}
# Flower data ------------------------------------------------------------------
# Recruits should theoretically be excluded from flowering probability and flower-number models.
# That exclusion is not supported by the data, because 35 recruits did flower, 
# including several fairly large individuals and some stage-3 plants.
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance_prev)) %>%
  mutate(
    year = factor(year),
    disturbance_prev = factor(
      disturbance_prev,
      levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, flower, fl_nr, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance_prev, stage, recruits)


# Flower model -----------------------------------------------------------------
mod_fl_00 <- glmer(
  flower ~ 1 + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_01 <- glmer(
  flower ~ disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_10 <- glmer(
  flower ~ logvol_t0 + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_11 <- glmer(
  flower ~ logvol_t0 + disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_20 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_21 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + disturbance_prev +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_30 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_31 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mods_fl <- list(mod_fl_00, mod_fl_01, mod_fl_10, mod_fl_11, 
                mod_fl_20, mod_fl_21, mod_fl_30, mod_fl_31)
mods_fl_dAIC <- bbmle::AICctab(mods_fl, weights = TRUE, sort = FALSE)$dAIC
mods_fl_sorted <- order(mods_fl_dAIC)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_best <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef <- coef(mod_fl_best)$year

mod_fl_best
summary(mod_fl_best)
mods_fl_dAIC
```


```{r}
# Flowering plots by year ------------------------------------------------------
make_fl_year_plot <- function(year_i) {
  df_i <- df_fl %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance_prev = levels(df_fl$disturbance_prev)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_prev = factor(
        disturbance_prev, levels = levels(df_fl$disturbance_prev)),
      year = factor(year_i, levels = levels(df_fl$year)))
  
  pred_i <- pred_i %>%
    mutate(
      flower = predict(
        mod_fl_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance_prev) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      flower = mean(flower, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, flower, color = disturbance_prev),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, flower, color = disturbance_prev),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = 'Flowering probability') +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_yrs <- lapply(levels(df_fl$year), make_fl_year_plot)
fig_fl_years <- wrap_plots(fl_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flowering - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_years
```

### Numbers of flowers
```{r}
# Flower number conditional on flowering data ----------------------------------
df_fl_cond <- df_fl %>%
  filter(flower == 1, !is.na(fl_nr), fl_nr > 0, fl_nr == round(fl_nr)) %>%
  mutate(year = factor(year))


# Flower number model ----------------------------------------------------------
mod_fl_n_0 <- glmer.nb(
  fl_nr ~ disturbance_prev + (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_1 <- glmer.nb(
  fl_nr ~ logvol_t0 + disturbance_prev + (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_2 <- glmer.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + disturbance_prev +
    (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_3 <- glmer.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance_prev +
    (1 | year),
  data = df_fl_cond, control = ctrl_glmer)

mods_fl_n <- list(mod_fl_n_0, mod_fl_n_1, mod_fl_n_2, mod_fl_n_3) # 
mods_fl_n_dAIC <- bbmle::AICctab(
  mods_fl_n, weights = TRUE, sort = FALSE)$dAIC
mods_fl_n_sorted <- order(mods_fl_n_dAIC)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_bestfit <- mods_fl_n_sorted[1]
  v_mod_fl_n_index <- mod_fl_n_index_bestfit - 1
} else {
  mod_fl_n_index_bestfit <- v_mod_set_fl_n + 1
  v_mod_fl_n_index <- v_mod_set_fl_n
}

mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_bestfit]]
mod_fl_n_ranef <- coef(mod_fl_n_best)$year

mod_fl_n_best
summary(mod_fl_n_best)
mods_fl_n_dAIC
```


```{r}
# Flower number plots by year --------------------------------------------------
make_fl_n_year_plot <- function(year_i) {
  df_i <- df_fl_cond %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance_prev = levels(df_fl_cond$disturbance_prev)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_prev = factor(
        disturbance_prev, levels = levels(df_fl_cond$disturbance_prev)),
      year = factor(year_i, levels = levels(df_fl_cond$year)))
  
  pred_i <- pred_i %>%
    mutate(
      fl_nr = predict(
        mod_fl_n_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance_prev) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      fl_nr = mean(fl_nr, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, fl_nr, color = disturbance_prev),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, fl_nr, color = disturbance_prev),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = 'Number of flowering stems') +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_n_yrs <- lapply(levels(df_fl_cond$year), make_fl_n_year_plot)
fig_fl_n_years <- wrap_plots(fl_n_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flower number - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_n_years
```

### Flwoers to recuit transition
```{r}
# Flowering stems to recruits --------------------------------------------------
# Total flowering stems in quadrat q during year t predict recruits in the
# same quadrat during year t + 1. Current fire is retained at the quadrat level.
df_quad_year_sampled <- df %>%
  distinct(
    site,
    quad,
    year)

df_fs2r <- df %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    fs_t0 = sum(
      fl_nr,
      na.rm = TRUE),
    disturbance = first(
      disturbance[!is.na(disturbance)],
      default = NA_real_),
    .groups = 'drop') %>%
  mutate(
    year_t1 = year + 1) %>%
  semi_join(
    df_quad_year_sampled %>%
      transmute(
        site,
        quad,
        year_t1 = year),
    by = c(
      'site',
      'quad',
      'year_t1')) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(
        site,
        quad,
        year,
        name = 're_t1'),
    by = c(
      'site',
      'quad',
      'year_t1' = 'year')) %>%
  mutate(
    re_t1 = replace_na(re_t1, 0L),
    fs_t0_log = log1p(fs_t0),
    re_t1_log = log1p(re_t1)) %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !is.na(disturbance),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re)) %>%
  mutate(
    site = factor(site),
    year_t0 = factor(year),
    year_t1 = factor(year_t1),
    disturbance_t0 = factor(
      disturbance,
      levels = c(0, 1)))

# Keep the old mean-script recruitment plot as a quick check.
fig_fs2r_raw <- ggplot(
  df_fs2r,
  aes(x = fs_t0, y = re_t1, color = disturbance_t0)) +
  geom_jitter(height = 0.2, width = 0.5, alpha = 0.4) +
  scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
  theme_bw() +
  labs(
    title = 'Recruits t1 by flowering stems t0',
    subtitle = v_ggp_suffix,
    x = 'Total flowering stems in quadrat in year t',
    y = 'Number of recruits in quadrat in year t + 1',
    color = 'Fire')

fig_fs2r_raw
```


```{r}
# Flowering-stem reference by disturbance --------------------------------------
df_stem_reference <- df_fs2r %>%
  group_by(disturbance) %>%
  summarise(
    fs_t0_ref = mean(fs_t0),
    n_quadrat_transitions = n(), .groups = 'drop')

df_stem_reference
```


```{r}
# Flower 2 Recruit model -------------------------------------------------------
ctrl_re <- lmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

df_fs2r <- df_fs2r %>%
  mutate(
    disturbance =
      as.numeric(as.character(disturbance_t0)))

# Intercept only
mod_re_00 <- lmer(
  re_t1_log ~ 1 +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Disturbance only
mod_re_0 <- lmer(
  re_t1_log ~ disturbance +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Flowering stems only
mod_re_10 <- lmer(
  re_t1_log ~ fs_t0_log +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Flowering stems + disturbance
mod_re_1 <- lmer(
  re_t1_log ~ fs_t0_log + disturbance +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Flowering stems x disturbance
mod_re_20 <- lmer(
  re_t1_log ~ fs_t0_log * disturbance +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

mods_re <- list(
  mod_re_00,
  mod_re_0,
  mod_re_10,
  mod_re_1,
  mod_re_20)

mods_re_dAIC <- bbmle::AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAIC

mods_re_sorted <- order(mods_re_dAIC)

if (length(v_mod_set_re) == 0) {
  mod_re_index_bestfit <- mods_re_sorted[1]
} else {
  mod_re_index_bestfit <- v_mod_set_re + 1
}

v_mod_re_index <- mod_re_index_bestfit - 1
mod_re_best <- mods_re[[mod_re_index_bestfit]]

mod_re_best
summary(mod_re_best)
mods_re_dAIC
```


```{r}
# Recruitment plots by recruit year -------------------------------------------
predict_re_safe <- function(model, newdata, re.form = NULL) {
  if (inherits(model, 'merMod')) {
    out <- predict(
      model, newdata = newdata, re.form = re.form, allow.new.levels = TRUE)
  } else {
    out <- predict(model, newdata = newdata)
  }
  return(out)
}

shape_values_re <- c(16, 17, 15, 18, 3, 4, 7, 8, 0, 1, 2, 5, 6, 9)
site_shapes_re <- setNames(
  shape_values_re[seq_along(levels(df_fs2r$site))],
  levels(df_fs2r$site))

make_re_year_plot <- function(year_i) {
  df_i <- df_fs2r %>% filter(year_t1 == year_i)
  
  x <- seq(
    0, max(df_i$fs_t0_log, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    fs_t0_log = x,
    disturbance_t0 = levels(df_fs2r$disturbance_t0)) %>%
    mutate(
      disturbance =
        as.numeric(as.character(disturbance_t0)),
      year_t1 = factor(year_i, levels = levels(df_fs2r$year_t1)),
      disturbance_t0 = factor(
        disturbance_t0, levels = levels(df_fs2r$disturbance_t0)))
  
  pred_i <- pred_i %>%
    mutate(
      pred_re = predict_re_safe(
        mod_re_best, pred_i, re.form = NULL))
  
  ggplot() +
    geom_point(
      data = df_i,
      aes(fs_t0_log, re_t1_log, color = site, shape = site),
      alpha = 0.75, size = 2) +
    geom_line(
      data = pred_i,
      aes(
        fs_t0_log, pred_re,
        group = disturbance_t0,
        linetype = disturbance_t0),
      inherit.aes = FALSE, color = 'black', linewidth = 0.8) +
    scale_color_viridis_d(option = 'turbo', end = 0.95) +
    scale_shape_manual(values = site_shapes_re) +
    scale_linetype_manual(values = c('0' = 'solid', '1' = 'dashed')) +
    theme_bw() +
    labs(
      title = year_i,
      x = expression('log(1 + flowering stems)'[t0]),
      y = expression('log(1 + recruits)'[t1]),
      color = 'Site',
      shape = 'Site',
      linetype = 'Fire') +
    theme(text = element_text(size = 5), legend.position = 'none')
}

re_yrs <- lapply(levels(df_fs2r$year_t1), make_re_year_plot)
fig_re_years <- wrap_plots(re_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Recruitment - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_re_years
```

### Recruits
```{r}
# Recruit size distribution ----------------------------------------------------
df_recr <- df %>%
  filter(recruits == 1, volume_t0 > 0, is.finite(logvol_t0))

rc_sz <- data.frame(
  mean = mean(df_recr$logvol_t0, na.rm = TRUE),
  sd = sd(df_recr$logvol_t0, na.rm = TRUE))
```

## Parameter extraction
```{r}
# Parameter extraction helpers -------------------------------------------------
empty_coef_df <- function() {
  data.frame(coefficient = character(), value = numeric())
}

is_mixed_model <- function(model) {
  inherits(model, 'merMod')
}

get_fixef_safe <- function(model) {
  if (is_mixed_model(model)) {
    fixef(model)
  } else {
    coef(model)
  }
}

get_group_coef_safe <- function(model, group_var) {
  if (!is_mixed_model(model)) {
    return(NULL)
  }
  coef_list <- coef(model)
  if (!(group_var %in% names(coef_list))) {
    return(NULL)
  }
  coef_list[[group_var]]
}

find_model_term <- function(model_terms, aliases) {
  aliases <- unlist(aliases)
  for (alias in aliases) {
    if (grepl('^regex:', alias)) {
      pattern <- sub('^regex:', '', alias)
      hit <- grep(pattern, model_terms, value = TRUE)
    } else {
      hit <- model_terms[model_terms == alias]
    }
    if (length(hit) > 0) {
      return(hit[1])
    }
  }
  NA_character_
}

bind_coef_rows <- function(x) {
  x <- x[!vapply(x, is.null, logical(1))]
  if (length(x) == 0) {
    return(empty_coef_df())
  }
  out <- do.call(rbind, x)
  rownames(out) <- NULL
  out
}

check_duplicate_coefficients <- function(x, object_name = 'parameter table') {
  dup <- x$coefficient[duplicated(x$coefficient)]
  if (length(dup) > 0) {
    stop(
      object_name,
      ' contains duplicated coefficient names: ',
      paste(unique(dup), collapse = ', '))
  }
  invisible(TRUE)
}

coef_df_to_list <- function(x) {
  if (nrow(x) == 0) {
    return(list())
  }
  check_duplicate_coefficients(x)
  x %>%
    mutate(coefficient = as.character(coefficient)) %>%
    pivot_wider(names_from = coefficient, values_from = value) %>%
    as.list()
}

extract_fixed_pars <- function(model, term_map, fill_missing = TRUE) {
  model_coefs <- get_fixef_safe(model)
  model_terms <- names(model_coefs)
  out <- lapply(names(term_map), function(coef_name) {
    model_term <- find_model_term(model_terms, term_map[[coef_name]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- 0
      } else {
        return(NULL)
      }
    } else {
      value <- unname(model_coefs[model_term])
    }
    data.frame(coefficient = coef_name, value = value)
  })
  bind_coef_rows(out)
}

extract_group_pars <- function(model, group_var, term_map, fill_missing = TRUE) {
  coef_matrix <- get_group_coef_safe(model, group_var)
  if (is.null(coef_matrix)) {
    return(empty_coef_df())
  }
  model_terms <- colnames(coef_matrix)
  out <- lapply(names(term_map), function(coef_prefix) {
    model_term <- find_model_term(model_terms, term_map[[coef_prefix]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- rep(0, nrow(coef_matrix))
      } else {
        return(NULL)
      }
    } else {
      value <- coef_matrix[, model_term]
    }
    data.frame(
      coefficient = paste0(coef_prefix, '_', rownames(coef_matrix)),
      value = value)
  })
  bind_coef_rows(out)
}

extract_re_ranef_devs <- function(model, group_var, group_label) {
  if (!inherits(model, 'merMod')) {
    return(empty_coef_df())
  }
  ranef_list <- ranef(model)
  if (!(group_var %in% names(ranef_list))) {
    return(empty_coef_df())
  }
  ranef_matrix <- ranef_list[[group_var]]
  out <- list()
  if ('(Intercept)' %in% colnames(ranef_matrix)) {
    out[['intercept']] <- data.frame(
      coefficient = paste0('re_u0_', group_label, '_',
                           rownames(ranef_matrix)),
      value = ranef_matrix[, '(Intercept)'])
  }
  if ('fs_t0_log' %in% colnames(ranef_matrix)) {
    out[['slope']] <- data.frame(
      coefficient = paste0('re_u1_', group_label, '_',
                           rownames(ranef_matrix)),
      value = ranef_matrix[, 'fs_t0_log'])
  }
  bind_coef_rows(out)
}
```

### Term maps 
```{r}
size_term_map <- list(
  b0 = c('(Intercept)'),
  b1 = c('logvol_t0'),
  b2 = c('logvol_t0_2', 'I(logvol_t0^2)'),
  b3 = c('logvol_t0_3', 'I(logvol_t0^3)'),
  bf = c('disturbance1', 'regex:^disturbance'))

# Flowering disturbance term map
lagged_size_term_map <- list(
  b0 = c("(Intercept)"),
  b1 = c("logvol_t0"),
  b2 = c("logvol_t0_2", "I(logvol_t0^2)"),
  b3 = c("logvol_t0_3", "I(logvol_t0^3)"),
  bf = c(
    "disturbance_prev1",
    "regex:^disturbance_prev"))

make_term_map <- function(prefix, map) {
  out <- map
  names(out) <- paste0(prefix, names(out))
  out
}

su_term_map <- make_term_map('surv_', size_term_map)
gr_term_map <- make_term_map('grow_', size_term_map)
fl_term_map  <- make_term_map("fl_",  lagged_size_term_map)
fln_term_map <- make_term_map("fln_", lagged_size_term_map)
re_term_map <- list(
  re_b0 = c('(Intercept)'),
  re_b1 = c('fs_t0_log'),
  re_bd = c('disturbance'),
  re_bf = c(
    'fs_t0_log:disturbance',
    'disturbance:fs_t0_log'))
```


### Fixed parameters
```{r}
su_fe <- extract_fixed_pars(mod_su_best, su_term_map, fill_missing = TRUE)
gr_fe <- extract_fixed_pars(mod_gr_best, gr_term_map, fill_missing = TRUE)
fl_fe <- extract_fixed_pars(mod_fl_best, fl_term_map, fill_missing = TRUE)
fln_fe <- extract_fixed_pars(mod_fl_n_best, fln_term_map, fill_missing = TRUE)
re_fe <- extract_fixed_pars(mod_re_best, re_term_map, fill_missing = TRUE)
```

### Constants
```{r}
gr_var_coef <- coef(mod_gr_var)

constants <- tibble::tribble(
  ~coefficient, ~value,
  'recr_sz', rc_sz$mean,
  'recr_sd', rc_sz$sd,
  'a', as.numeric(gr_var_coef[1]),
  'b', as.numeric(gr_var_coef[2]),
  'L', min(c(df_gr$logvol_t0, df_fl$logvol_t0), na.rm = TRUE) - 0.1,
  'U', max(c(df_gr$logvol_t0, df_fl$logvol_t0), na.rm = TRUE) + 0.1,
  'mat_siz', 200,
  'fs_t0_ref_0', df_stem_reference %>%
    filter(disturbance == 0) %>% pull(fs_t0_ref),
  'fs_t0_ref_1', df_stem_reference %>%
    filter(disturbance == 1) %>% pull(fs_t0_ref),
  're_sigma', sigma(mod_re_best),
  'mod_su_index', v_mod_su_index,
  'mod_gr_index', v_mod_gr_index,
  'mod_fl_index', v_mod_fl_index,
  'mod_fl_n_index', v_mod_fl_n_index,
  'mod_re_index', v_mod_re_index) %>%
  mutate(coefficient = as.character(coefficient), value = as.numeric(value))

pars_cons <- bind_coef_rows(list(su_fe, gr_fe, fl_fe, fln_fe, re_fe,
                                 constants))
check_duplicate_coefficients(pars_cons, object_name = 'pars_cons')
pars_all_mean <- coef_df_to_list(pars_cons)
pars_mean <- pars_all_mean
```

### Year-varying parameters
```{r}
su_out_yr <- extract_group_pars(mod_su_best, 'year', su_term_map,
                                fill_missing = TRUE)
gr_out_yr <- extract_group_pars(mod_gr_best, 'year', gr_term_map,
                                fill_missing = TRUE)
fl_out_yr <- extract_group_pars(mod_fl_best, 'year', fl_term_map,
                                fill_missing = TRUE)
fln_out_yr <- extract_group_pars(mod_fl_n_best, 'year', fln_term_map,
                                 fill_missing = TRUE)

pars_var <- bind_coef_rows(list(su_out_yr, gr_out_yr, fl_out_yr,
                                fln_out_yr))
check_duplicate_coefficients(pars_var, object_name = 'pars_var')
pars_all_year <- coef_df_to_list(pars_var)
```

### Recruitment-year random-effect deviations
```{r}
pars_re_year_t1 <- extract_re_ranef_devs(
  mod_re_best, 'year_t1', 'year_t1') %>%
  mutate(coefficient = as.character(coefficient))
check_duplicate_coefficients(
  pars_re_year_t1, object_name = 'pars_re_year_t1')
pars_all_re_year_t1 <- coef_df_to_list(pars_re_year_t1)
```

## IPM helper functions
```{r}
inv_logit <- function(x) {
  plogis(x)
}

get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    return(pars[[par_name]])
  }
  default
}

make_ipm_pars <- function(
    pars_mean,
    pars_year = NULL,
    pars_re_year_t1 = NULL,
    year = NULL,
    re_year_t1 = NULL) {
  pars <- pars_mean
  
  if (!is.null(year) && !is.null(pars_year)) {
    year <- as.character(year)
    year_hits <- grep(paste0('_', year, '$'), names(pars_year), value = TRUE)
    
    for (nm in year_hits) {
      base_nm <- sub(paste0('_', year, '$'), '', nm)
      pars[[base_nm]] <- pars_year[[nm]]
    }
  }
  
  if (is.null(re_year_t1) && !is.null(year)) {
    re_year_t1 <- as.numeric(as.character(year)) + 1
  }
  
  if (!is.null(re_year_t1) && !is.null(pars_re_year_t1)) {
    re_year_t1 <- as.character(re_year_t1)
    pars$re_b0 <- get_par(pars, 're_b0', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u0_year_t1_', re_year_t1), 0)
    pars$re_b1 <- get_par(pars, 're_b1', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u1_year_t1_', re_year_t1), 0)
  }
  
  pars
}

surv_lp <- function(x, pars, disturbance = 0) {
  get_par(pars, 'surv_b0', 0) +
    get_par(pars, 'surv_b1', 0) * x +
    get_par(pars, 'surv_b2', 0) * x^2 +
    get_par(pars, 'surv_b3', 0) * x^3 +
    get_par(pars, 'surv_bf', 0) * disturbance
}

sx <- function(x, pars, disturbance = 0) {
  inv_logit(surv_lp(x, pars, disturbance))
}

grow_mu <- function(x, pars, disturbance = 0) {
  get_par(pars, 'grow_b0', 0) +
    get_par(pars, 'grow_b1', 0) * x +
    get_par(pars, 'grow_b2', 0) * x^2 +
    get_par(pars, 'grow_b3', 0) * x^3 +
    get_par(pars, 'grow_bf', 0) * disturbance
}

grow_sd <- function(x, pars) {
  sqrt(pars$a * exp(pars$b * x))
}

gxy <- function(x, y, pars, disturbance = 0) {
  dnorm(
    y,
    mean = grow_mu(x, pars, disturbance),
    sd = grow_sd(x, pars))
}

# Flowering probability
fl_x <- function(x, pars, disturbance = 0) {
  eta <- get_par(pars, 'fl_b0', 0) +
    get_par(pars, 'fl_b1', 0) * x +
    get_par(pars, 'fl_b2', 0) * x^2 +
    get_par(pars, 'fl_b3', 0) * x^3 +
    get_par(pars, 'fl_bf', 0) * disturbance
  
  inv_logit(eta)
}


# Flowering-stem number conditional on flowering
fl_n_x <- function(x, pars, disturbance = 0) {
  eta <- get_par(pars, 'fln_b0', 0) +
    get_par(pars, 'fln_b1', 0) * x +
    get_par(pars, 'fln_b2', 0) * x^2 +
    get_par(pars, 'fln_b3', 0) * x^3 +
    get_par(pars, 'fln_bf', 0) * disturbance
  
  exp(eta)
}


# Expected flowering stems per individual
fs_x <- function(x, pars, disturbance = 0) {
  fl_x(x, pars, disturbance) * fl_n_x(x, pars, disturbance)
}


# Flowering-stem reference
get_fs_ref <- function(pars, disturbance = 0) {
  if (disturbance == 1) {
    get_par(pars, 'fs_t0_ref_1')
  } else {
    get_par(pars, 'fs_t0_ref_0')
  }
}


# Flowering-stem-to-recruit conversion
re_total_ref <- function(pars, disturbance = 0) {
  re_intercept <- get_par(pars, 're_b0', 0) +
    get_par(pars, 're_bd', 0) * disturbance
  
  re_slope <- get_par(pars, 're_b1', 0) +
    get_par(pars, 're_bf', 0) * disturbance
  
  fs_ref <- get_fs_ref(pars, disturbance)
  
  re_log <- re_intercept + re_slope * log1p(fs_ref)
  
  re_value <- expm1(
    re_log + 0.5 * get_par(pars, 're_sigma', 0)^2)
  
  pmax(re_value, 0)
}


rx <- function(x, pars, disturbance = 0) {
  fs_value <- fs_x(x, pars, disturbance)
  fs_ref <- get_fs_ref(pars, disturbance)
  re_ref <- re_total_ref(pars, disturbance)
  
  re_per_fs <- re_ref / pmax(fs_ref, .Machine$double.eps)
  
  fs_value * re_per_fs
}


# Recruit size distribution
re_y_dist <- function(y, pars, h = NULL) {
  dens <- dnorm(y, mean = pars$recr_sz, sd = pars$recr_sd)
  
  if (!is.null(h)) {
    dens <- dens / sum(dens * h)
  }
  
  dens
}


# Fertility matrix
fyx <- function(y, x, pars, h, disturbance = 0) {
  Pvec <- re_y_dist(y, pars, h = h)
  Rvec <- rx(x, pars, disturbance)
  
  outer(Pvec, Rvec) * h
}
```


## Kernel: historical size-structured
```{r}
kernel <- function(pars, disturbance = 0) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  S <- sx(y, pars, disturbance = disturbance)
  
  G <- t(outer(
    y, y, Vectorize(function(x, y) {
      gxy(x, y, pars, disturbance = disturbance)
    }))) * h
  
  for (i in 1:n) {
    if (i <= n / 2) {
      G[1, i] <- G[1, i] + 1 - sum(G[, i])
    } else {
      G[n, i] <- G[n, i] + 1 - sum(G[, i])
    }
  }
  
  T <- sweep(G, 2, S, '*')
  
  F <- fyx(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance)
  
  k_yx <- T + F
  
  Imat <- diag(n)
  Zmat <- matrix(0, n, n)
  
  K_hist <- rbind(
    cbind(T, F),
    cbind(Imat, Zmat))
  
  list(
    k_yx = k_yx,
    K_hist = K_hist,
    T = T,
    F = F,
    Imat = Imat,
    Zmat = Zmat,
    G = G,
    S = S,
    meshpts = y,
    h = h,
    L = L,
    U = U)
}

lambda_ipm <- function(pars, disturbance = 0) {
  K <- kernel(
    pars = pars,
    disturbance = disturbance)$K_hist
  
  Re(eigen(K, only.values = TRUE)$values[1])
}
```


## Mean IPMs
```{r}
lambda_ipm(pars_mean, disturbance = 0)
lambda_ipm(pars_mean, disturbance = 1)
```


## Year-specific IPMs
```{r}
lambda_ipm_year <- function(
    year,
    disturbance = 0,
    re_year_t1 = NULL) {
  pars_i <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = year,
    re_year_t1 = re_year_t1)
  
  lambda_ipm(
    pars_i,
    disturbance = disturbance)
}

# Year-specific disturbance scenarios ----
ipm_years <- sort(unique(as.integer(as.character(df_su$year))))

lambda_year <- data.frame(
  year = ipm_years,
  lambda_undisturbed = sapply(ipm_years, function(i) {
    lambda_ipm_year(year = i, disturbance = 0)}),
  lambda_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(year = i, disturbance = 1)})
)


fig_lambda_year <- lambda_year %>%
  pivot_longer(
    cols = -year,
    names_to = 'disturbance',
    values_to = 'lambda') %>%
  mutate(
    disturbance = recode(
      disturbance,
      lambda_undisturbed = 'Undisturbed',
      lambda_fire = 'Fire')) %>%
  ggplot(aes(year, lambda, color = disturbance)) +
  geom_hline(yintercept = 1, linetype = 'dashed') +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(
    title = 'Year-specific asymptotic lambda',
    subtitle = v_ggp_suffix,
    x = 'Year',
    y = expression(lambda),
    color = 'Disturbance')

fig_lambda_year
```

## Historical kernel diagnostics by year
```{r}
df_hist_kernel_check <- bind_rows(
  lapply(ipm_years, function(year_i) {
    pars_i <- make_ipm_pars(
      pars_mean = pars_all_mean,
      pars_year = pars_all_year,
      pars_re_year_t1 = pars_all_re_year_t1,
      year = year_i,
      re_year_t1 = year_i + 1)
    
    bind_rows(
      lapply(c(0, 1), function(disturbance_i) {
        kern_i <- kernel(
          pars = pars_i,
          disturbance = disturbance_i)
        
        n <- pars_i$mat_siz
        K <- kern_i$K_hist
        
        tibble(
          year = year_i,
          disturbance = disturbance_i,
          n_rows = nrow(K),
          n_cols = ncol(K),
          identity_block = isTRUE(all.equal(
            K[(n + 1):(2 * n), 1:n],
            diag(n))),
          zero_block = all(
            K[(n + 1):(2 * n), (n + 1):(2 * n)] == 0),
          any_nonfinite = any(!is.finite(K)))
      }))
  }))

df_hist_kernel_check
```


## Observed and projected population growth
Observed growth is calculated from quadrats represented in t - 1, t, and t + 1. Each quadrat is projected with its observed disturbance at year t.


### Quadrat-year disturbance lookup
```{r}
fire_lookup_quad <- df %>%
  group_by(site, quad, year) %>%
  summarise(
    disturbance_num = first(
      disturbance[!is.na(disturbance)],
      default = NA_real_),
    .groups = 'drop')
```

### Quadrat-year abundance
Counts include sized plants and recruits without a usable size measurement.
```{r}
df_counts_quad_year <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year)) %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    n_sized = sum(
      is.finite(logvol_t0),
      na.rm = TRUE),
    n_unsized_recruits = sum(
      !is.finite(logvol_t0) & recruits == 1,
      na.rm = TRUE),
    n_ipm_state = n_sized + n_unsized_recruits,
    .groups = 'drop')
```

### Matched consecutive quadrat transitions
```{r}
df_obs_pgr_quad <- df_counts_quad_year %>%
  arrange(site, quad, year) %>%
  group_by(site, quad) %>%
  mutate(
    year_tm1 = lag(year),
    n_tm1 = lag(n_ipm_state),
    year_t1 = lead(year),
    n_t1 = lead(n_ipm_state),
    year_gap_prev = year - year_tm1,
    year_gap_next = year_t1 - year) %>%
  ungroup() %>%
  filter(
    year_gap_prev == 1,
    year_gap_next == 1,
    n_ipm_state > 0) %>%
  rename(n_t0 = n_ipm_state) %>%
  left_join(
    fire_lookup_quad,
    by = c('site', 'quad', 'year')) %>%
  filter(!is.na(disturbance_num)) %>%
  mutate(obs_pgr = n_t1 / n_t0)
```

### Initial quadrat size distribution
```{r}
make_initial_n_quad <- function(
    year0,
    site_i,
    quad_i,
    pars) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  breaks <- seq(
    L,
    U,
    length.out = n + 1)
  
  b <- L + c(0:n) * h
  
  y <- 0.5 *
    (b[1:n] + b[2:(n + 1)])
  
  df0 <- df %>%
    filter(
      year == year0,
      site == site_i,
      quad == quad_i)
  
  sizes <- df0 %>%
    filter(
      is.finite(logvol_t0)) %>%
    pull(logvol_t0)
  
  size_counts <- hist(
    pmin(
      pmax(sizes, L),
      U),
    breaks = breaks,
    plot = FALSE,
    include.lowest = TRUE)$counts
  
  n_density <- size_counts / h
  
  n_unsized_recruits <- df0 %>%
    summarise(
      n = sum(
        !is.finite(logvol_t0) &
          recruits == 1,
        na.rm = TRUE)) %>%
    pull(n)
  
  if (n_unsized_recruits > 0) {
    n_density <- n_density +
      n_unsized_recruits *
      re_y_dist(
        y,
        pars,
        h = h)
  }
  
  n_density
}
```

### Project one quadrat-year
```{r}
project_one_quad_year <- function(
    yr,
    site_i,
    quad_i,
    disturbance_y) {
  
  pars_y <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = yr,
    re_year_t1 = yr + 1)
  
  n_current <- make_initial_n_quad(
    year0 = yr,
    site_i = site_i,
    quad_i = quad_i,
    pars = pars_y)
  
  n_previous <- make_initial_n_quad(
    year0 = yr - 1,
    site_i = site_i,
    quad_i = quad_i,
    pars = pars_y)
  
  n_hist <- c(n_current, n_previous)
  
  K <- kernel(
    pars = pars_y,
    disturbance = disturbance_y)$K_hist
  
  n <- pars_y$mat_siz
  h <- (pars_y$U - pars_y$L) / n
  
  n_initial <- sum(n_current) * h
  
  if (n_initial <= 0) {
    return(
      data.frame(
        year = yr,
        site = site_i,
        quad = quad_i,
        disturbance_num = disturbance_y,
        n_obs_model = n_initial,
        n_proj_model = NA_real_,
        asym_lambda = NA_real_,
        proj_lambda = NA_real_))
  }
  
  n_proj_hist <- as.numeric(K %*% n_hist)
  n_next <- n_proj_hist[1:n]
  n_projected <- sum(n_next) * h
  
  data.frame(
    year = yr,
    site = site_i,
    quad = quad_i,
    disturbance_num = disturbance_y,
    n_obs_model = n_initial,
    n_proj_model = n_projected,
    asym_lambda = Re(eigen(K, only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial)
}
```


### Project all matched quadrat transitions
```{r}
df_proj_quad <- bind_rows(
  lapply(
    seq_len(nrow(df_obs_pgr_quad)),
    function(i) {
      
      project_one_quad_year(
        yr = df_obs_pgr_quad$year[i],
        site_i = df_obs_pgr_quad$site[i],
        quad_i = df_obs_pgr_quad$quad[i],
        disturbance_y = df_obs_pgr_quad$disturbance_num[i])
    }))
```


### Quadrat-level observed and modeled growth
```{r}
df_compare_quad <- df_obs_pgr_quad %>%
  left_join(
    df_proj_quad,
    by = c(
      'year',
      'site',
      'quad',
      'disturbance_num')) %>%
  mutate(
    error_asymptotic_vs_obs =
      asym_lambda - obs_pgr,
    error_projected_vs_obs =
      proj_lambda - obs_pgr)
```

### Site-year comparison from matched quadrats
```{r}
df_compare_site <- df_compare_quad %>%
  group_by(
    site,
    year) %>%
  summarise(
    asym_lambda = weighted.mean(
      asym_lambda,
      w = n_obs_model,
      na.rm = TRUE),
    n_t0 = sum(
      n_t0,
      na.rm = TRUE),
    n_t1 = sum(
      n_t1,
      na.rm = TRUE),
    n_obs_model = sum(
      n_obs_model,
      na.rm = TRUE),
    n_proj_model = sum(
      n_proj_model,
      na.rm = TRUE),
    n_quadrats = n(),
    n_burned_quadrats = sum(
      disturbance_num == 1,
      na.rm = TRUE),
    disturbance_num = as.numeric(
      any(
        disturbance_num == 1,
        na.rm = TRUE)),
    .groups = 'drop') %>%
  mutate(
    obs_pgr = n_t1 / n_t0,
    proj_lambda = n_proj_model / n_obs_model,
    error_asymptotic_vs_obs =
      asym_lambda - obs_pgr,
    error_projected_vs_obs =
      proj_lambda - obs_pgr)

df_compare_site %>%
  print(
    n = 100,
    width = Inf)
```

### Whole-population annual comparison
Annual totals contain only quadrats represented in t - 1, t, and t + 1.
```{r}
df_compare <- df_compare_quad %>%
  group_by(year) %>%
  summarise(
    asym_lambda = weighted.mean(
      asym_lambda,
      w = n_obs_model,
      na.rm = TRUE),
    n_t0 = sum(
      n_t0,
      na.rm = TRUE),
    n_t1 = sum(
      n_t1,
      na.rm = TRUE),
    n_obs_model = sum(
      n_obs_model,
      na.rm = TRUE),
    n_proj_model = sum(
      n_proj_model,
      na.rm = TRUE),
    n_quadrats = n(),
    n_sites = n_distinct(site),
    n_burned_quadrats = sum(
      disturbance_num == 1,
      na.rm = TRUE),
    p_burned_quadrats = mean(
      disturbance_num == 1,
      na.rm = TRUE),
    disturbance = if_else(
      any(
        disturbance_num == 1,
        na.rm = TRUE),
      'Fire',
      'No fire'),
    .groups = 'drop') %>%
  mutate(
    obs_pgr = n_t1 / n_t0,
    proj_lambda = n_proj_model / n_obs_model,
    disturbance = factor(
      disturbance,
      levels = c(
        'No fire',
        'Fire')))

# Observed vs modeled plot

df_plot <- df_compare %>%
  select(year, obs_pgr, asym_lambda, proj_lambda, disturbance) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Abundance-weighted asymptotic lambda',
      proj_lambda = 'Projected lambda from observed quadrat size distributions'))

fig_mod_vs_obs <- ggplot(
  df_plot,
  aes(x = lambda, y = obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = 'Observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs
```

### Log-transformed observed vs modeled plot
```{r}
df_plot_log <- df_plot %>%
  filter(obs_pgr > 0, lambda > 0) %>%
  mutate(
    log_obs_pgr = log(obs_pgr),
    log_lambda = log(lambda))

fig_mod_vs_obs_log <- ggplot(
  df_plot_log,
  aes(x = log_lambda, y = log_obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = paste(v_ggp_suffix, '- log-transformed lambda'),
    x = expression('log modeled ' * lambda),
    y = 'log observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs_log
```


### Summary statistics
```{r}
df_compare_summary <- df_compare %>%
  summarise(
    n_years = n(),
    arithmetic_mean_obs_pgr = mean(obs_pgr, na.rm = TRUE),
    geometric_mean_obs_pgr = exp(mean(log(obs_pgr), na.rm = TRUE)),
    arithmetic_mean_asym_lambda = mean(asym_lambda, na.rm = TRUE),
    geometric_mean_asym_lambda = exp(mean(log(asym_lambda), na.rm = TRUE)),
    arithmetic_mean_proj_lambda = mean(proj_lambda, na.rm = TRUE),
    geometric_mean_proj_lambda = exp(mean(log(proj_lambda), na.rm = TRUE)),
    mean_error_asymptotic_vs_obs = mean(
      asym_lambda - obs_pgr, na.rm = TRUE),
    mean_error_projected_vs_obs = mean(
      proj_lambda - obs_pgr, na.rm = TRUE),
    percent_bias_asymptotic_vs_obs = 100 * sum(
      asym_lambda - obs_pgr, na.rm = TRUE) /
      sum(obs_pgr, na.rm = TRUE),
    percent_bias_projected_vs_obs = 100 * sum(
      proj_lambda - obs_pgr, na.rm = TRUE) /
      sum(obs_pgr, na.rm = TRUE),
    rmse_asymptotic_vs_obs = sqrt(mean(
      (asym_lambda - obs_pgr)^2, na.rm = TRUE)),
    rmse_projected_vs_obs = sqrt(mean(
      (proj_lambda - obs_pgr)^2, na.rm = TRUE))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "statistic",
    values_to = "value")

df_compare_summary

df_compare %>% print(n = 100)
```


### Site-level summary
```{r}
df_compare_site_summary <- df_compare_site %>%
  filter(
    !is.na(obs_pgr),
    !is.na(asym_lambda),
    !is.na(proj_lambda),
    obs_pgr > 0,
    asym_lambda > 0,
    proj_lambda > 0) %>%
  group_by(site) %>%
  summarise(
    n_year_transitions = n(),
    lambda_obs_geometric = exp(mean(log(obs_pgr), na.rm = TRUE)),
    lambda_obs_arithmetic = mean(obs_pgr, na.rm = TRUE),
    lambda_asymptotic_geometric = exp(mean(log(asym_lambda),
                                           na.rm = TRUE)),
    lambda_asymptotic_arithmetic = mean(asym_lambda, na.rm = TRUE),
    lambda_projected_geometric = exp(mean(log(proj_lambda),
                                          na.rm = TRUE)),
    lambda_projected_arithmetic = mean(proj_lambda, na.rm = TRUE),
    error_projected_geo_vs_obs_geo =
      lambda_projected_geometric - lambda_obs_geometric,
    rmse_projected_vs_obs = sqrt(mean((proj_lambda - obs_pgr)^2,
                                      na.rm = TRUE)),
    mean_n_initial = mean(n_obs_model, na.rm = TRUE),
    mean_disturbance = mean(disturbance_num, na.rm = TRUE),
    .groups = 'drop') %>%
  arrange(as.numeric(as.character(site)))

df_compare_site_summary %>% print(n = 100, width = Inf)

# 
# # Save key outputs -------------------------------------------------------------
# # Uncomment if needed.
# # saveRDS(pars_all_mean,
# #         file.path(dir_result, 'pole_year_fire_pars_mean.rds'))
# # saveRDS(pars_all_year,
# #         file.path(dir_result, 'pole_year_fire_pars_year.rds'))
# # saveRDS(df_compare,
# #         file.path(dir_result, 'pole_year_fire_lambda_compare.rds'))
# # saveRDS(df_compare_site,
# #         file.path(dir_result, 'pole_year_fire_lambda_compare_site.rds'))
```
