# Yearly historical envi projection model {#yr-ipm-hist-envi}

## Year-specific environmental-history IPM

This analysis develops a **year-specific environmental-history integral
projection model (IPM)** for *Polygala lewtonii*. In contrast to the historical
population-state model, where the population vector explicitly contains both
the current and previous population size distributions, the historical
information in this model is carried by the **disturbance state**.

The timing of disturbance differs among the demographic processes. Survival and
growth from year \(t\) to \(t+1\) depend on disturbance during the current
interval, denoted \(d_t\). Recruitment from flowering stems in year \(t\) to
recruits in year \(t+1\) can also respond to this current disturbance. In
contrast, flowering probability and flowering-stem production in year \(t\)
depend on disturbance during the preceding interval, denoted \(d_{t-1}\).

The demographic kernel therefore depends on both the current and previous
disturbance states:

$$
K_t(d_t,d_{t-1}) =
T_t(d_t) + F_t(d_t,d_{t-1}),
$$

where \(T_t\) contains survival and growth and \(F_t\) contains recruitment.
Survival and growth depend only on current disturbance, while fertility depends
on both disturbance states because flowering and flowering-stem production
respond to previous disturbance and the flowering-stem-to-recruit transition
can respond to current disturbance.

This creates four possible demographic kernels:

$$
K_{00,t}, \qquad
K_{10,t}, \qquad
K_{01,t}, \qquad
K_{11,t},
$$

where the first index represents current disturbance and the second index
represents previous disturbance. Thus,

- \(K_{00,t}\): no fire in either interval,
- \(K_{10,t}\): fire in the current interval only,
- \(K_{01,t}\): fire in the previous interval only, and
- \(K_{11,t}\): fire in both intervals.


### One-step projection of the observed population

For the observed annual projection, the historical disturbance state is known.
Each quadrat has an observed current disturbance \(d_t\) and previous
disturbance \(d_{t-1}\). The appropriate year-specific kernel is therefore
selected directly and applied to the observed size distribution in year \(t\):

$$
\mathbf{n}_{t+1}
=
K_t(d_t,d_{t-1})\mathbf{n}_t.
$$

For example, a quadrat that burned during the current interval but did not burn
during the preceding interval is projected using

$$
\mathbf{n}_{t+1}
=
K_{10,t}\mathbf{n}_t.
$$

A quadrat that burned in the previous interval but not during the current
interval is projected using

$$
\mathbf{n}_{t+1}
=
K_{01,t}\mathbf{n}_t.
$$

Importantly, the previous population vector \(\mathbf{n}_{t-1}\) is **not**
required in this formulation. The reproductive contribution in year \(t\)
originates from individuals that are members of the current population
\(\mathbf{n}_t\). The historical information required by the model is instead
contained in the previous disturbance state \(d_{t-1}\).

The resulting projected annual population growth rate is

$$
\lambda_{\mathrm{proj},t}
=
\frac{N_{\mathrm{projected},t+1}}
     {N_{\mathrm{observed},t}}.
$$

This projected lambda depends on the observed population structure and the
actual disturbance history of each quadrat and can therefore be compared
directly with the observed population growth rate from \(t\) to \(t+1\).

### Long-term environmental history

The four state-specific kernels can also be examined individually. However,
their dominant eigenvalues do not all represent biologically possible
long-term disturbance sequences. In particular, repeatedly applying
\(K_{10}\) would imply that every year has current fire but no fire in the
previous year, while repeatedly applying \(K_{01}\) would imply the opposite.
These histories cannot persist because the current disturbance state in one
year necessarily becomes the previous disturbance state in the following year.

For long-term population growth, the disturbance history must therefore be
updated explicitly. If

$$
p_{j|i} = P(d_t=j \mid d_{t-1}=i),
$$

the disturbance process can be represented by transition probabilities between
the two environmental states.

The corresponding environmental-history megakernel is

$$
K_{\mathrm{env},t}
=
\begin{pmatrix}
p_{0|0}K_{00,t} & p_{0|1}K_{01,t} \\
p_{1|0}K_{10,t} & p_{1|1}K_{11,t}
\end{pmatrix}.
$$

The two population blocks represent individuals whose previous disturbance
state is no fire or fire. After projection, individuals are automatically
transferred into the appropriate disturbance-history state for the following
year.

The dominant eigenvalue of this environmental-history kernel provides a
long-term growth rate that respects the temporal sequence of disturbance.
Thus, the model separates two related questions:

1. **Observed one-step projection:** what population growth is predicted from
   year \(t\) to \(t+1\), given the observed population structure and observed
   disturbance history?

2. **Environmental-history asymptotic growth:** what population growth is
   expected when the disturbance process continues according to its observed
   transition probabilities?




## Working dataset
```{r}
df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv'))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.integer(year),
    site = factor(site)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))
```



```{r}
# Controls ---------------------------------------------------------------------
ctrl_glmer <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))
```

## Vital rate models
### Survival
```{r}
# Survival data ----------------------------------------------------------------
df_su <- df %>%
  filter(
    !is.na(survives),
    size_t0 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = factor(disturbance, levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage)
```

#### Survival model
```{r}
mod_su_0 <- glmer(
  survives ~ disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_10 <- glmer(
  survives ~ logvol_t0 + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_11 <- glmer(
  survives ~ logvol_t0 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_20 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_21 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_30 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_31 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)


mods_su <- list(
  mod_su_0, mod_su_10, mod_su_11, 
  mod_su_20, mod_su_21, mod_su_30, mod_su_31)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef <- coef(mod_su_best)$year

mod_su_best
summary(mod_su_best)
mods_su_dAIC
```

#### Survival plots by year
```{r, fig.width=10, fig.height=12}
make_su_year_plot <- function(year_i) {
  df_i <- df_su %>% filter(year == year_i)
  
  x <- seq(
    min(df_i$logvol_t0, na.rm = TRUE),
    max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance = levels(df_su$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance = factor(
        disturbance, levels = levels(df_su$disturbance)),
      year = factor(year_i, levels = levels(df_su$year)))
  
  pred_i$survives <- predict(
    mod_su_best, newdata = pred_i,
    type = "response", re.form = NULL)
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      survives = mean(survives, na.rm = TRUE),
      n = n(), .groups = "drop") %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, survives, color = disturbance), size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, survives, color = disturbance,
          linetype = disturbance),
      linewidth = 0.7) +
    scale_color_manual(values = c("0" = "black", "1" = "red")) +
    labs(
      title = year_i,
      x = expression("log(volume)"[t0]),
      y = expression("Survival probability"[t1])) +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = "none")
}

su_yrs <- lapply(levels(df_su$year), make_su_year_plot)
fig_su_years <- wrap_plots(su_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_su_years
```

### Growth
```{r}
# Growth data ------------------------------------------------------------------
df_gr <- df %>%
  filter(
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = factor(
      disturbance,
      levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage)
```


#### Growth model
```{r}
mod_gr_0 <- lmer(
  logvol_t1 ~ disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_10 <- lmer(
  logvol_t1 ~ logvol_t0 + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_11 <- lmer(
  logvol_t1 ~ logvol_t0 + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_20 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_21 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_30 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_31 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)


mods_gr <- list(
  mod_gr_0, mod_gr_10, mod_gr_11, mod_gr_20, mod_gr_21, 
  mod_gr_30, mod_gr_31)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef <- coef(mod_gr_best)$year

mod_gr_best
summary(mod_gr_best)
mods_gr_dAIC
```

#### Growth plots by year
```{r, fig.width=10, fig.height=12}
make_gr_year_plot <- function(year_i) {
  df_i <- df_gr %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance = levels(df_gr$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance = factor(
        disturbance, levels = levels(df_gr$disturbance)),
      year = factor(year_i, levels = levels(df_gr$year)))
  
  pred_i <- pred_i %>%
    mutate(
      logvol_t1 = predict(
        mod_gr_best,
        newdata = pred_i,
        re.form = NULL))
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      logvol_t1 = mean(logvol_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, logvol_t1, color = disturbance),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, logvol_t1, color = disturbance,
          linetype = disturbance),
      linewidth = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = 'blue', lty = 2) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = expression('log(volume)'[t1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

gr_yrs <- lapply(levels(df_gr$year), make_gr_year_plot)
fig_gr_years <- wrap_plots(gr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Growth - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_gr_years
```

#### Growth variance
```{r}
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

### Flowering probability
```{r}
# Flower data ------------------------------------------------------------------
# Recruits should theoretically be excluded from flowering probability and flower-number models.
# That exclusion is not supported by the data, because 35 recruits did flower, 
# including several fairly large individuals and some stage-3 plants.
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance_prev)) %>%
  mutate(
    year = factor(year),
    disturbance_prev = factor(
      disturbance_prev,
      levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, flower, fl_nr, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance_prev, stage, recruits)
```

#### Flower model
```{r}
mod_fl_00 <- glmer(
  flower ~ 1 + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_01 <- glmer(
  flower ~ disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_10 <- glmer(
  flower ~ logvol_t0 + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_11 <- glmer(
  flower ~ logvol_t0 + disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_20 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_21 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + disturbance_prev +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_30 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)
mod_fl_31 <- glmer(
  flower ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance_prev + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mods_fl <- list(mod_fl_00, mod_fl_01, mod_fl_10, mod_fl_11, 
                mod_fl_20, mod_fl_21, mod_fl_30, mod_fl_31)
mods_fl_dAIC <- bbmle::AICctab(mods_fl, weights = TRUE, sort = FALSE)$dAIC
mods_fl_sorted <- order(mods_fl_dAIC)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_best <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef <- coef(mod_fl_best)$year

mod_fl_best
summary(mod_fl_best)
mods_fl_dAIC
```

#### Flowering plots by year
```{r, fig.width=10, fig.height=12}
make_fl_year_plot <- function(year_i) {
  df_i <- df_fl %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance_prev = levels(df_fl$disturbance_prev)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_prev = factor(
        disturbance_prev, levels = levels(df_fl$disturbance_prev)),
      year = factor(year_i, levels = levels(df_fl$year)))
  
  pred_i <- pred_i %>%
    mutate(
      flower = predict(
        mod_fl_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance_prev) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      flower = mean(flower, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, flower, color = disturbance_prev),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, flower, color = disturbance_prev),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = 'Flowering probability') +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_yrs <- lapply(levels(df_fl$year), make_fl_year_plot)
fig_fl_years <- wrap_plots(fl_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flowering - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_years
```

### Number of flowers
```{r}
# Flower number conditional on flowering data ----------------------------------
df_fl_cond <- df_fl %>%
  filter(flower == 1, !is.na(fl_nr), fl_nr > 0, fl_nr == round(fl_nr)) %>%
  mutate(year = factor(year))
```

#### Flower number model
```{r}
mod_fl_n_0 <- glmer.nb(
  fl_nr ~ disturbance_prev + (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_1 <- glmer.nb(
  fl_nr ~ logvol_t0 + disturbance_prev + (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_2 <- glmer.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + disturbance_prev +
    (1 | year),
  data = df_fl_cond, control = ctrl_glmer)
mod_fl_n_3 <- glmer.nb(
  fl_nr ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance_prev +
    (1 | year),
  data = df_fl_cond, control = ctrl_glmer)

mods_fl_n <- list(mod_fl_n_0, mod_fl_n_1, mod_fl_n_2, mod_fl_n_3) # 
mods_fl_n_dAIC <- bbmle::AICctab(
  mods_fl_n, weights = TRUE, sort = FALSE)$dAIC
mods_fl_n_sorted <- order(mods_fl_n_dAIC)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_bestfit <- mods_fl_n_sorted[1]
  v_mod_fl_n_index <- mod_fl_n_index_bestfit - 1
} else {
  mod_fl_n_index_bestfit <- v_mod_set_fl_n + 1
  v_mod_fl_n_index <- v_mod_set_fl_n
}

mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_bestfit]]
mod_fl_n_ranef <- coef(mod_fl_n_best)$year

mod_fl_n_best
summary(mod_fl_n_best)
mods_fl_n_dAIC
```

#### Flower number plots by year
```{r, fig.width=10, fig.height=12}
make_fl_n_year_plot <- function(year_i) {
  df_i <- df_fl_cond %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance_prev = levels(df_fl_cond$disturbance_prev)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance_prev = factor(
        disturbance_prev, levels = levels(df_fl_cond$disturbance_prev)),
      year = factor(year_i, levels = levels(df_fl_cond$year)))
  
  pred_i <- pred_i %>%
    mutate(
      fl_nr = predict(
        mod_fl_n_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance_prev) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      fl_nr = mean(fl_nr, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, fl_nr, color = disturbance_prev),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, fl_nr, color = disturbance_prev),
      linewidth = 0.7) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = 'Number of flowering stems') +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_n_yrs <- lapply(levels(df_fl_cond$year), make_fl_n_year_plot)
fig_fl_n_years <- wrap_plots(fl_n_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flower number - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_n_years
```

### Number of flowers to recruit transition
```{r, fig.width=5, fig.height=8}
# Flowering stems to recruits --------------------------------------------------
# Total flowering stems in quadrat q during year t predict recruits in the
# same quadrat during year t + 1. Current fire is retained at the quadrat level.
df_quad_year_sampled <- df %>%
  distinct(
    site,
    quad,
    year)

df_fs2r <- df %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    fs_t0 = sum(
      fl_nr,
      na.rm = TRUE),
    disturbance = first(
      disturbance[!is.na(disturbance)],
      default = NA_real_),
    .groups = 'drop') %>%
  mutate(
    year_t1 = year + 1) %>%
  semi_join(
    df_quad_year_sampled %>%
      transmute(
        site,
        quad,
        year_t1 = year),
    by = c(
      'site',
      'quad',
      'year_t1')) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(
        site,
        quad,
        year,
        name = 're_t1'),
    by = c(
      'site',
      'quad',
      'year_t1' = 'year')) %>%
  mutate(
    re_t1 = replace_na(re_t1, 0L),
    fs_t0_log = log1p(fs_t0),
    re_t1_log = log1p(re_t1)) %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !is.na(disturbance),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re)) %>%
  mutate(
    site = factor(site),
    year_t0 = factor(year),
    year_t1 = factor(year_t1),
    disturbance_t0 = factor(
      disturbance,
      levels = c(0, 1)))

# Keep the old mean-script recruitment plot as a quick check.
fig_fs2r_raw <- ggplot(
  df_fs2r,
  aes(x = fs_t0, y = re_t1, color = disturbance_t0)) +
  geom_jitter(height = 0.2, width = 0.5, alpha = 0.4) +
  scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
  theme_bw() +
  labs(
    title = 'Recruits t1 by flowering stems t0',
    subtitle = v_ggp_suffix,
    x = 'Total flowering stems in quadrat in year t',
    y = 'Number of recruits in quadrat in year t + 1',
    color = 'Fire')

fig_fs2r_raw
```

#### Flowering-stem reference by previous-year fire
```{r}
df_fs2r_previous_fire <- df_fs2r %>%
  left_join(
    df %>%
      distinct(
        site,
        quad,
        year,
        disturbance_prev),
    by = c(
      'site',
      'quad',
      'year')) %>%
  filter(!is.na(disturbance_prev))

df_stem_reference <- df_fs2r_previous_fire %>%
  group_by(disturbance_prev) %>%
  summarise(
    fs_t0_ref = mean(fs_t0),
    n_quadrat_transitions = n(),
    .groups = 'drop')

df_stem_reference
```

#### Flower 2 Recruit model
```{r}
ctrl_re <- lmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

df_fs2r <- df_fs2r %>%
  mutate(
    disturbance =
      as.numeric(as.character(disturbance_t0)))

# Intercept only
mod_re_00 <- lmer(
  re_t1_log ~ 1 +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Disturbance only
mod_re_0 <- lmer(
  re_t1_log ~ disturbance +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Flowering stems only
mod_re_10 <- lmer(
  re_t1_log ~ fs_t0_log +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Flowering stems + disturbance
mod_re_1 <- lmer(
  re_t1_log ~ fs_t0_log + disturbance +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

# Flowering stems x disturbance
mod_re_20 <- lmer(
  re_t1_log ~ fs_t0_log * disturbance +
    (1 | year_t1),
  data = df_fs2r,
  REML = FALSE,
  control = ctrl_re)

mods_re <- list(
  mod_re_00,
  mod_re_0,
  mod_re_10,
  mod_re_1,
  mod_re_20)

mods_re_dAIC <- bbmle::AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAIC

mods_re_sorted <- order(mods_re_dAIC)

if (length(v_mod_set_re) == 0) {
  mod_re_index_bestfit <- mods_re_sorted[1]
} else {
  mod_re_index_bestfit <- v_mod_set_re + 1
}

v_mod_re_index <- mod_re_index_bestfit - 1
mod_re_best <- mods_re[[mod_re_index_bestfit]]

mod_re_best
summary(mod_re_best)
mods_re_dAIC
```


#### Recruitment plots by recruit year
```{r, fig.width=10, fig.height=12}
predict_re_safe <- function(model, newdata, re.form = NULL) {
  if (inherits(model, 'merMod')) {
    out <- predict(
      model, newdata = newdata, re.form = re.form, allow.new.levels = TRUE)
  } else {
    out <- predict(model, newdata = newdata)
  }
  return(out)
}

shape_values_re <- c(16, 17, 15, 18, 3, 4, 7, 8, 0, 1, 2, 5, 6, 9)
site_shapes_re <- setNames(
  shape_values_re[seq_along(levels(df_fs2r$site))],
  levels(df_fs2r$site))

make_re_year_plot <- function(year_i) {
  df_i <- df_fs2r %>% filter(year_t1 == year_i)
  
  x <- seq(
    0, max(df_i$fs_t0_log, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    fs_t0_log = x,
    disturbance_t0 = levels(df_fs2r$disturbance_t0)) %>%
    mutate(
      disturbance =
        as.numeric(as.character(disturbance_t0)),
      year_t1 = factor(year_i, levels = levels(df_fs2r$year_t1)),
      disturbance_t0 = factor(
        disturbance_t0, levels = levels(df_fs2r$disturbance_t0)))
  
  pred_i <- pred_i %>%
    mutate(
      pred_re = predict_re_safe(
        mod_re_best, pred_i, re.form = NULL))
  
  ggplot() +
    geom_point(
      data = df_i,
      aes(fs_t0_log, re_t1_log, color = site, shape = site),
      alpha = 0.75, size = 2) +
    geom_line(
      data = pred_i,
      aes(
        fs_t0_log, pred_re,
        group = disturbance_t0,
        linetype = disturbance_t0),
      inherit.aes = FALSE, color = 'black', linewidth = 0.8) +
    scale_color_viridis_d(option = 'turbo', end = 0.95) +
    scale_shape_manual(values = site_shapes_re) +
    scale_linetype_manual(values = c('0' = 'solid', '1' = 'dashed')) +
    theme_bw() +
    labs(
      title = year_i,
      x = expression('log(1 + flowering stems)'[t0]),
      y = expression('log(1 + recruits)'[t1]),
      color = 'Site',
      shape = 'Site',
      linetype = 'Fire') +
    theme(text = element_text(size = 5), legend.position = 'none')
}

re_yrs <- lapply(levels(df_fs2r$year_t1), make_re_year_plot)
fig_re_years <- wrap_plots(re_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Recruitment - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_re_years
```

### Recruitment
```{r}
# Recruit size distribution ----------------------------------------------------
df_recr <- df %>%
  filter(recruits == 1, volume_t0 > 0, is.finite(logvol_t0))

rc_sz <- data.frame(
  mean = mean(df_recr$logvol_t0, na.rm = TRUE),
  sd = sd(df_recr$logvol_t0, na.rm = TRUE))
```


## Parameter extraction
### Parameter extraction helpers
```{r}
empty_coef_df <- function() {
  data.frame(coefficient = character(), value = numeric())
}

is_mixed_model <- function(model) {
  inherits(model, 'merMod')
}

get_fixef_safe <- function(model) {
  if (is_mixed_model(model)) {
    fixef(model)
  } else {
    coef(model)
  }
}

get_group_coef_safe <- function(model, group_var) {
  if (!is_mixed_model(model)) {
    return(NULL)
  }
  coef_list <- coef(model)
  if (!(group_var %in% names(coef_list))) {
    return(NULL)
  }
  coef_list[[group_var]]
}

find_model_term <- function(model_terms, aliases) {
  aliases <- unlist(aliases)
  for (alias in aliases) {
    if (grepl('^regex:', alias)) {
      pattern <- sub('^regex:', '', alias)
      hit <- grep(pattern, model_terms, value = TRUE)
    } else {
      hit <- model_terms[model_terms == alias]
    }
    if (length(hit) > 0) {
      return(hit[1])
    }
  }
  NA_character_
}

bind_coef_rows <- function(x) {
  x <- x[!vapply(x, is.null, logical(1))]
  if (length(x) == 0) {
    return(empty_coef_df())
  }
  out <- do.call(rbind, x)
  rownames(out) <- NULL
  out
}

check_duplicate_coefficients <- function(x, object_name = 'parameter table') {
  dup <- x$coefficient[duplicated(x$coefficient)]
  if (length(dup) > 0) {
    stop(
      object_name,
      ' contains duplicated coefficient names: ',
      paste(unique(dup), collapse = ', '))
  }
  invisible(TRUE)
}

coef_df_to_list <- function(x) {
  if (nrow(x) == 0) {
    return(list())
  }
  check_duplicate_coefficients(x)
  x %>%
    mutate(coefficient = as.character(coefficient)) %>%
    pivot_wider(names_from = coefficient, values_from = value) %>%
    as.list()
}

extract_fixed_pars <- function(model, term_map, fill_missing = TRUE) {
  model_coefs <- get_fixef_safe(model)
  model_terms <- names(model_coefs)
  out <- lapply(names(term_map), function(coef_name) {
    model_term <- find_model_term(model_terms, term_map[[coef_name]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- 0
      } else {
        return(NULL)
      }
    } else {
      value <- unname(model_coefs[model_term])
    }
    data.frame(coefficient = coef_name, value = value)
  })
  bind_coef_rows(out)
}

extract_group_pars <- function(model, group_var, term_map, fill_missing = TRUE) {
  coef_matrix <- get_group_coef_safe(model, group_var)
  if (is.null(coef_matrix)) {
    return(empty_coef_df())
  }
  model_terms <- colnames(coef_matrix)
  out <- lapply(names(term_map), function(coef_prefix) {
    model_term <- find_model_term(model_terms, term_map[[coef_prefix]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- rep(0, nrow(coef_matrix))
      } else {
        return(NULL)
      }
    } else {
      value <- coef_matrix[, model_term]
    }
    data.frame(
      coefficient = paste0(coef_prefix, '_', rownames(coef_matrix)),
      value = value)
  })
  bind_coef_rows(out)
}

extract_re_ranef_devs <- function(model, group_var, group_label) {
  if (!inherits(model, 'merMod')) {
    return(empty_coef_df())
  }
  ranef_list <- ranef(model)
  if (!(group_var %in% names(ranef_list))) {
    return(empty_coef_df())
  }
  ranef_matrix <- ranef_list[[group_var]]
  out <- list()
  if ('(Intercept)' %in% colnames(ranef_matrix)) {
    out[['intercept']] <- data.frame(
      coefficient = paste0('re_u0_', group_label, '_',
                           rownames(ranef_matrix)),
      value = ranef_matrix[, '(Intercept)'])
  }
  if ('fs_t0_log' %in% colnames(ranef_matrix)) {
    out[['slope']] <- data.frame(
      coefficient = paste0('re_u1_', group_label, '_',
                           rownames(ranef_matrix)),
      value = ranef_matrix[, 'fs_t0_log'])
  }
  bind_coef_rows(out)
}
```



### Term maps
```{r}
size_term_map <- list(
  b0 = c('(Intercept)'),
  b1 = c('logvol_t0'),
  b2 = c('logvol_t0_2', 'I(logvol_t0^2)'),
  b3 = c('logvol_t0_3', 'I(logvol_t0^3)'),
  bf = c('disturbance1', 'regex:^disturbance'))

# Flowering disturbance term map
lagged_size_term_map <- list(
  b0 = c("(Intercept)"),
  b1 = c("logvol_t0"),
  b2 = c("logvol_t0_2", "I(logvol_t0^2)"),
  b3 = c("logvol_t0_3", "I(logvol_t0^3)"),
  bf = c(
    "disturbance_prev1",
    "regex:^disturbance_prev"))

make_term_map <- function(prefix, map) {
  out <- map
  names(out) <- paste0(prefix, names(out))
  out
}

su_term_map <- make_term_map('surv_', size_term_map)
gr_term_map <- make_term_map('grow_', size_term_map)
fl_term_map  <- make_term_map("fl_",  lagged_size_term_map)
fln_term_map <- make_term_map("fln_", lagged_size_term_map)
re_term_map <- list(
  re_b0 = c('(Intercept)'),
  re_b1 = c('fs_t0_log'),
  re_bd = c('disturbance'),
  re_bf = c(
    'fs_t0_log:disturbance',
    'disturbance:fs_t0_log'))
```

### Fixed parameters
```{r}
su_fe <- extract_fixed_pars(mod_su_best, su_term_map, fill_missing = TRUE)
gr_fe <- extract_fixed_pars(mod_gr_best, gr_term_map, fill_missing = TRUE)
fl_fe <- extract_fixed_pars(mod_fl_best, fl_term_map, fill_missing = TRUE)
fln_fe <- extract_fixed_pars(mod_fl_n_best, fln_term_map, fill_missing = TRUE)
re_fe <- extract_fixed_pars(mod_re_best, re_term_map, fill_missing = TRUE)
```

### Constants
```{r}
gr_var_coef <- coef(mod_gr_var)

constants <- tibble::tribble(
  ~coefficient, ~value,
  'recr_sz', rc_sz$mean,
  'recr_sd', rc_sz$sd,
  'a', as.numeric(gr_var_coef[1]),
  'b', as.numeric(gr_var_coef[2]),
  'L', min(c(df_gr$logvol_t0, df_fl$logvol_t0), na.rm = TRUE) - 0.1,
  'U', max(c(df_gr$logvol_t0, df_fl$logvol_t0), na.rm = TRUE) + 0.1,
  'mat_siz', 200,
  'fs_t0_ref_0', df_stem_reference %>% filter(disturbance_prev == 0) %>% pull(fs_t0_ref),
  'fs_t0_ref_1', df_stem_reference %>% filter(disturbance_prev == 1) %>% pull(fs_t0_ref),
  're_sigma', sigma(mod_re_best),
  'mod_su_index', v_mod_su_index,
  'mod_gr_index', v_mod_gr_index,
  'mod_fl_index', v_mod_fl_index,
  'mod_fl_n_index', v_mod_fl_n_index,
  'mod_re_index', v_mod_re_index) %>%
  mutate(coefficient = as.character(coefficient), value = as.numeric(value))

pars_cons <- bind_coef_rows(list(su_fe, gr_fe, fl_fe, fln_fe, re_fe,
                                 constants))
check_duplicate_coefficients(pars_cons, object_name = 'pars_cons')
pars_all_mean <- coef_df_to_list(pars_cons)
pars_mean <- pars_all_mean
```

### Year-varying parameters
```{r}
su_out_yr <- extract_group_pars(mod_su_best, 'year', su_term_map,
                                fill_missing = TRUE)
gr_out_yr <- extract_group_pars(mod_gr_best, 'year', gr_term_map,
                                fill_missing = TRUE)
fl_out_yr <- extract_group_pars(mod_fl_best, 'year', fl_term_map,
                                fill_missing = TRUE)
fln_out_yr <- extract_group_pars(mod_fl_n_best, 'year', fln_term_map,
                                 fill_missing = TRUE)

pars_var <- bind_coef_rows(list(su_out_yr, gr_out_yr, fl_out_yr,
                                fln_out_yr))
check_duplicate_coefficients(pars_var, object_name = 'pars_var')
pars_all_year <- coef_df_to_list(pars_var)
```

### Recruitment-year random-effect deviations
```{r}
pars_re_year_t1 <- extract_re_ranef_devs(
  mod_re_best, 'year_t1', 'year_t1') %>%
  mutate(coefficient = as.character(coefficient))
check_duplicate_coefficients(
  pars_re_year_t1, object_name = 'pars_re_year_t1')
pars_all_re_year_t1 <- coef_df_to_list(pars_re_year_t1)

```

## Disturbance transition probabilities
```{r}
# Current disturbance refers to t -> t + 1 and previous disturbance to
# t - 1 -> t.

df_disturbance_regime <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year),
    !is.na(disturbance),
    !is.na(disturbance_prev)) %>%
  distinct(
    site,
    quad,
    year,
    disturbance,
    disturbance_prev) %>%
  count(
    disturbance,
    disturbance_prev,
    name = 'n_quadrat_years') %>%
  complete(
    disturbance = c(0, 1),
    disturbance_prev = c(0, 1),
    fill = list(n_quadrat_years = 0)) %>%
  arrange(
    disturbance_prev,
    disturbance)

df_disturbance_regime


df_disturbance_transition <- df_disturbance_regime %>%
  group_by(disturbance_prev) %>%
  mutate(
    p_transition = n_quadrat_years /
      sum(n_quadrat_years)) %>%
  ungroup()

df_disturbance_transition


get_transition_probability <- function(
    disturbance_i,
    disturbance_prev_i) {
  
  df_disturbance_transition %>%
    filter(
      disturbance == disturbance_i,
      disturbance_prev == disturbance_prev_i) %>%
    pull(p_transition)
}


p_0_given_0 <- get_transition_probability(
  disturbance_i = 0,
  disturbance_prev_i = 0)

p_1_given_0 <- get_transition_probability(
  disturbance_i = 1,
  disturbance_prev_i = 0)

p_0_given_1 <- get_transition_probability(
  disturbance_i = 0,
  disturbance_prev_i = 1)

p_1_given_1 <- get_transition_probability(
  disturbance_i = 1,
  disturbance_prev_i = 1)


df_disturbance_transition
```

## IPM helper functions
```{r}
inv_logit <- function(x) {
  plogis(x)
}

get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    return(pars[[par_name]])
  }
  default
}

make_ipm_pars <- function(
    pars_mean,
    pars_year = NULL,
    pars_re_year_t1 = NULL,
    year = NULL,
    re_year_t1 = NULL) {
  pars <- pars_mean
  
  if (!is.null(year) && !is.null(pars_year)) {
    year <- as.character(year)
    year_hits <- grep(paste0('_', year, '$'), names(pars_year), value = TRUE)
    
    for (nm in year_hits) {
      base_nm <- sub(paste0('_', year, '$'), '', nm)
      pars[[base_nm]] <- pars_year[[nm]]
    }
  }
  
  if (is.null(re_year_t1) && !is.null(year)) {
    re_year_t1 <- as.numeric(as.character(year)) + 1
  }
  
  if (!is.null(re_year_t1) && !is.null(pars_re_year_t1)) {
    re_year_t1 <- as.character(re_year_t1)
    pars$re_b0 <- get_par(pars, 're_b0', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u0_year_t1_', re_year_t1), 0)
    pars$re_b1 <- get_par(pars, 're_b1', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u1_year_t1_', re_year_t1), 0)
  }
  
  pars
}

surv_lp <- function(x, pars, disturbance = 0) {
  get_par(pars, 'surv_b0', 0) +
    get_par(pars, 'surv_b1', 0) * x +
    get_par(pars, 'surv_b2', 0) * x^2 +
    get_par(pars, 'surv_b3', 0) * x^3 +
    get_par(pars, 'surv_bf', 0) * disturbance
}

sx <- function(x, pars, disturbance = 0) {
  inv_logit(surv_lp(x, pars, disturbance))
}

grow_mu <- function(x, pars, disturbance = 0) {
  get_par(pars, 'grow_b0', 0) +
    get_par(pars, 'grow_b1', 0) * x +
    get_par(pars, 'grow_b2', 0) * x^2 +
    get_par(pars, 'grow_b3', 0) * x^3 +
    get_par(pars, 'grow_bf', 0) * disturbance
}

grow_sd <- function(x, pars) {
  sqrt(pars$a * exp(pars$b * x))
}

gxy <- function(x, y, pars, disturbance = 0) {
  dnorm(
    y,
    mean = grow_mu(x, pars, disturbance),
    sd = grow_sd(x, pars))
}

fl_x <- function(x, pars, disturbance_prev = 0) {
  eta <- get_par(pars, "fl_b0", 0) +
    get_par(pars, "fl_b1", 0) * x +
    get_par(pars, "fl_b2", 0) * x^2 +
    get_par(pars, "fl_b3", 0) * x^3 +
    get_par(pars, "fl_bf", 0) * disturbance_prev
  inv_logit(eta)
}

fl_n_x <- function(x, pars, disturbance_prev = 0) {
  eta <- get_par(pars, "fln_b0", 0) +
    get_par(pars, "fln_b1", 0) * x +
    get_par(pars, "fln_b2", 0) * x^2 +
    get_par(pars, "fln_b3", 0) * x^3 +
    get_par(pars, "fln_bf", 0) * disturbance_prev
  exp(eta)
}

fs_x <- function(x, pars, disturbance_prev = 0) {
  fl_x(x, pars, disturbance_prev) *
    fl_n_x(x, pars, disturbance_prev)
}

get_fs_ref <- function(
    pars,
    disturbance_prev = 0) {
  
  if (disturbance_prev == 1) {
    get_par(pars, 'fs_t0_ref_1')
  } else {
    get_par(pars, 'fs_t0_ref_0')
  }
}

re_total_ref <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  
  re_intercept <- get_par(pars, 're_b0', 0) +
    get_par(pars, 're_bd', 0) * disturbance
  
  re_slope <- get_par(pars, 're_b1', 0) +
    get_par(pars, 're_bf', 0) * disturbance
  
  fs_ref <- get_fs_ref(
    pars,
    disturbance_prev = disturbance_prev)
  
  re_log <- re_intercept +
    re_slope * log1p(fs_ref)
  
  re_value <- expm1(
    re_log +
      0.5 * get_par(pars, 're_sigma', 0)^2)
  
  pmax(re_value, 0)
}

rx <- function(
    x,
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  
  fs_value <- fs_x(
    x,
    pars,
    disturbance_prev = disturbance_prev)
  
  fs_ref <- get_fs_ref(
    pars,
    disturbance_prev = disturbance_prev)
  
  re_ref <- re_total_ref(
    pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
  
  re_per_fs <- re_ref /
    pmax(
      fs_ref,
      .Machine$double.eps)
  
  fs_value * re_per_fs
}

re_y_dist <- function(y, pars, h = NULL) {
  dens <- dnorm(y, mean = pars$recr_sz, sd = pars$recr_sd)
  if (!is.null(h)) {
    dens <- dens / sum(dens * h)
  }
  dens
}

fyx <- function(
    y,
    x,
    pars,
    h,
    disturbance = 0,
    disturbance_prev = 0) {
  Pvec <- re_y_dist(y, pars, h = h)
  Rvec <- rx(
    x,
    pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
  outer(Pvec, Rvec) * h
}
```


## Single-state size-structured kernel
```{r}
kernel <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  S <- sx(y, pars, disturbance = disturbance)
  
  G <- t(outer(
    y, y, Vectorize(function(x, y) {
      gxy(x, y, pars, disturbance = disturbance)
    }))) * h
  
  for (i in 1:n) {
    if (i <= n / 2) {
      G[1, i] <- G[1, i] + 1 - sum(G[, i])
    } else {
      G[n, i] <- G[n, i] + 1 - sum(G[, i])
    }
  }
  
  T <- sweep(G, 2, S, '*')
  
  F <- fyx(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
  
  k_yx <- T + F
  
  list(
    k_yx = k_yx,
    T = T,
    F = F,
    G = G,
    S = S,
    meshpts = y,
    h = h,
    L = L,
    U = U)
}

lambda_ipm <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  Re(eigen(
    kernel(
      pars,
      disturbance = disturbance,
      disturbance_prev = disturbance_prev)$k_yx)$values[1])
}
```

## Environmental-history kernel
```{r}
kernel_env_hist <- function(pars) {
  
  K_00 <- kernel(
    pars = pars,
    disturbance = 0,
    disturbance_prev = 0)$k_yx
  
  K_10 <- kernel(
    pars = pars,
    disturbance = 1,
    disturbance_prev = 0)$k_yx
  
  K_01 <- kernel(
    pars = pars,
    disturbance = 0,
    disturbance_prev = 1)$k_yx
  
  K_11 <- kernel(
    pars = pars,
    disturbance = 1,
    disturbance_prev = 1)$k_yx
  
  K_env <- rbind(
    cbind(
      p_0_given_0 * K_00,
      p_0_given_1 * K_01),
    cbind(
      p_1_given_0 * K_10,
      p_1_given_1 * K_11))
  
  list(
    K_env = K_env,
    K_00 = K_00,
    K_10 = K_10,
    K_01 = K_01,
    K_11 = K_11)
}
```


### Environmental-history asymptotic lambda
```{r}
lambda_env_hist <- function(pars) {
  
  K <- kernel_env_hist(
    pars = pars)$K_env
  
  Re(
    eigen(
      K,
      only.values = TRUE)$values[1])
}
```


## Mean IPMs
```{r}
lambda_ipm(pars_mean, disturbance = 0, disturbance_prev = 0)
lambda_ipm(pars_mean, disturbance = 0, disturbance_prev = 1)
lambda_ipm(pars_mean, disturbance = 1, disturbance_prev = 1)
lambda_ipm(pars_mean, disturbance = 1, disturbance_prev = 0)

lambda_env_hist(pars_mean)
```


## Year-specific IPMs
```{r}
lambda_ipm_year <- function(
    year,
    disturbance = 0,
    disturbance_prev = 0,
    re_year_t1 = NULL) {
  pars_i <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = year,
    re_year_t1 = re_year_t1)
  
  lambda_ipm(
    pars_i,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
}
```

### Year-specific environmental-history lambda
```{r}
lambda_env_hist_year <- function(
    year,
    re_year_t1 = NULL) {
  
  pars_i <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = year,
    re_year_t1 = re_year_t1)
  
  lambda_env_hist(
    pars = pars_i)
}
```


### Year-specific disturbance scenarios
```{r, fig.width=5, fig.height=8}
ipm_years <- sort(unique(as.integer(as.character(df_su$year))))

lambda_year <- data.frame(
  year = ipm_years,
  lambda_undisturbed = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 0,
      disturbance_prev = 0)}),
  lambda_current_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 1,
      disturbance_prev = 0)}),
  lambda_previous_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 0,
      disturbance_prev = 1)}),
  lambda_consecutive_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 1,
      disturbance_prev = 1)}),
  lambda_environmental_history = sapply(ipm_years, function(i) {
    lambda_env_hist_year(
      year = i)})
)

fig_lambda_year <- lambda_year %>%
  pivot_longer(
    cols = -year,
    names_to = 'disturbance',
    values_to = 'lambda') %>%
  mutate(
    disturbance = recode(
      disturbance,
      lambda_undisturbed = 'Undisturbed',
      lambda_current_fire = 'Current-year fire',
      lambda_previous_fire = 'Previous-year fire',
      lambda_consecutive_fire = 'Fire in both years',
      lambda_environmental_history = 'Environmental-history regime')) %>%
  ggplot(aes(year, lambda, color = disturbance)) +
  geom_hline(yintercept = 1, linetype = 'dashed') +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(
    title = 'Year-specific asymptotic lambda',
    subtitle = v_ggp_suffix,
    x = 'Year',
    y = expression(lambda),
    color = 'Disturbance')

fig_lambda_year
```

## Observed and projected population growth
Observed growth is calculated from the same quadrats represented in both consecutive years. Each quadrat is projected with its own current- and previous-year disturbance state.


### Quadrat-year disturbance lookup
```{r}
fire_lookup_quad <- df %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    disturbance_num = first(
      disturbance[!is.na(disturbance)],
      default = NA_real_),
    disturbance_prev_num = first(
      disturbance_prev[!is.na(disturbance_prev)],
      default = NA_real_),
    .groups = 'drop')
```

### Quadrat-year abundance
```{r}
# Counts include sized plants and recruits without a usable size measurement.
df_counts_quad_year <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year)) %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    n_sized = sum(
      is.finite(logvol_t0),
      na.rm = TRUE),
    n_unsized_recruits = sum(
      !is.finite(logvol_t0) & recruits == 1,
      na.rm = TRUE),
    n_ipm_state = n_sized + n_unsized_recruits,
    .groups = 'drop')
```

### Matched consecutive quadrat transitions
```{r}
df_obs_pgr_quad <- df_counts_quad_year %>%
  arrange(
    site,
    quad,
    year) %>%
  group_by(
    site,
    quad) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n_ipm_state),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(
    year_gap == 1,
    n_ipm_state > 0) %>%
  rename(
    n_t0 = n_ipm_state) %>%
  left_join(
    fire_lookup_quad,
    by = c(
      'site',
      'quad',
      'year')) %>%
  filter(
    !is.na(disturbance_num),
    !is.na(disturbance_prev_num)) %>%
  mutate(
    obs_pgr = n_t1 / n_t0)
```

### Initial quadrat size distribution
```{r}
make_initial_n_quad <- function(
    year0,
    site_i,
    quad_i,
    pars) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  breaks <- seq(
    L,
    U,
    length.out = n + 1)
  
  b <- L + c(0:n) * h
  
  y <- 0.5 *
    (b[1:n] + b[2:(n + 1)])
  
  df0 <- df %>%
    filter(
      year == year0,
      site == site_i,
      quad == quad_i)
  
  sizes <- df0 %>%
    filter(
      is.finite(logvol_t0)) %>%
    pull(logvol_t0)
  
  size_counts <- hist(
    pmin(
      pmax(sizes, L),
      U),
    breaks = breaks,
    plot = FALSE,
    include.lowest = TRUE)$counts
  
  n_density <- size_counts / h
  
  n_unsized_recruits <- df0 %>%
    summarise(
      n = sum(
        !is.finite(logvol_t0) &
          recruits == 1,
        na.rm = TRUE)) %>%
    pull(n)
  
  if (n_unsized_recruits > 0) {
    n_density <- n_density +
      n_unsized_recruits *
      re_y_dist(
        y,
        pars,
        h = h)
  }
  
  n_density
}
```

### Project one quadrat-year
```{r}
project_one_quad_year <- function(
    yr,
    site_i,
    quad_i,
    disturbance_y,
    disturbance_prev_y) {
  
  pars_y <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = yr,
    re_year_t1 = yr + 1)
  
  n_obs <- make_initial_n_quad(
    year0 = yr,
    site_i = site_i,
    quad_i = quad_i,
    pars = pars_y)
  
  K <- kernel(
    pars = pars_y,
    disturbance = disturbance_y,
    disturbance_prev = disturbance_prev_y)$k_yx
  
  h <- (pars_y$U - pars_y$L) /
    pars_y$mat_siz
  
  n_initial <- sum(n_obs) * h
  
  if (n_initial <= 0) {
    return(
      data.frame(
        year = yr,
        site = site_i,
        quad = quad_i,
        disturbance_num = disturbance_y,
        disturbance_prev_num = disturbance_prev_y,
        n_obs_model = n_initial,
        n_proj_model = NA_real_,
        state_asym_lambda = NA_real_,
        proj_lambda = NA_real_))
  }
  
  n_proj <- K %*% n_obs
  
  n_projected <- sum(n_proj) * h
  
  data.frame(
    year = yr,
    site = site_i,
    quad = quad_i,
    disturbance_num = disturbance_y,
    disturbance_prev_num = disturbance_prev_y,
    n_obs_model = n_initial,
    n_proj_model = n_projected,
    state_asym_lambda = Re(
      eigen(
        K,
        only.values = TRUE)$values[1]),
    proj_lambda = as.numeric(
      n_projected / n_initial))
}
```

### Project all matched quadrat transitions
```{r}
df_proj_quad <- bind_rows(
  lapply(
    seq_len(nrow(df_obs_pgr_quad)),
    function(i) {
      
      project_one_quad_year(
        yr = df_obs_pgr_quad$year[i],
        site_i = df_obs_pgr_quad$site[i],
        quad_i = df_obs_pgr_quad$quad[i],
        disturbance_y =
          df_obs_pgr_quad$disturbance_num[i],
        disturbance_prev_y =
          df_obs_pgr_quad$disturbance_prev_num[i])
    }))
```

### Quadrat-level observed and modeled growth
```{r}
df_compare_quad <- df_obs_pgr_quad %>%
  left_join(
    df_proj_quad,
    by = c(
      'year',
      'site',
      'quad',
      'disturbance_num',
      'disturbance_prev_num')) %>%
  mutate(
    error_state_asymptotic_vs_obs =
      state_asym_lambda - obs_pgr,
    error_projected_vs_obs =
      proj_lambda - obs_pgr)
```

### Disturbance-state kernel used for each quadrat-year
```{r}
df_compare_quad <- df_compare_quad %>%
  mutate(
    kernel_state = case_when(
      disturbance_num == 0 & disturbance_prev_num == 0 ~ 'K_00',
      disturbance_num == 1 & disturbance_prev_num == 0 ~ 'K_10',
      disturbance_num == 0 & disturbance_prev_num == 1 ~ 'K_01',
      disturbance_num == 1 & disturbance_prev_num == 1 ~ 'K_11',
      TRUE ~ NA_character_))


df_kernel_state_use <- df_compare_quad %>%
  count(
    year,
    kernel_state,
    name = 'n_quadrat_transitions') %>%
  arrange(
    year,
    kernel_state)

df_kernel_state_use
```

### Site-year comparison from matched quadrats
```{r}
df_compare_site <- df_compare_quad %>%
  group_by(
    site,
    year) %>%
  summarise(
    n_t0 = sum(
      n_t0,
      na.rm = TRUE),
    n_t1 = sum(
      n_t1,
      na.rm = TRUE),
    n_obs_model = sum(
      n_obs_model,
      na.rm = TRUE),
    n_proj_model = sum(
      n_proj_model,
      na.rm = TRUE),
    n_quadrats = n(),
    n_burned_quadrats = sum(
      disturbance_num == 1,
      na.rm = TRUE),
    disturbance_num = as.numeric(
      any(
        disturbance_num == 1,
        na.rm = TRUE)),
    disturbance_prev_num = as.numeric(
      any(
        disturbance_prev_num == 1,
        na.rm = TRUE)),
    .groups = 'drop') %>%
  left_join(
    lambda_year %>%
      dplyr::select(
        year,
        lambda_environmental_history),
    by = 'year') %>%
  rename(
    asym_lambda = lambda_environmental_history) %>%
  mutate(
    obs_pgr = n_t1 / n_t0,
    proj_lambda = n_proj_model / n_obs_model,
    error_asymptotic_vs_obs =
      asym_lambda - obs_pgr,
    error_projected_vs_obs =
      proj_lambda - obs_pgr)

df_compare_site %>%
  print(
    n = 100,
    width = Inf)
```

## Whole-population annual comparison
```{r, fig.width=5, fig.height=8}
# Annual totals contain only quadrats represented in both t and t + 1.
df_compare <- df_compare_quad %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(
      n_t0,
      na.rm = TRUE),
    n_t1 = sum(
      n_t1,
      na.rm = TRUE),
    n_obs_model = sum(
      n_obs_model,
      na.rm = TRUE),
    n_proj_model = sum(
      n_proj_model,
      na.rm = TRUE),
    n_quadrats = n(),
    n_sites = n_distinct(site),
    n_burned_quadrats = sum(
      disturbance_num == 1,
      na.rm = TRUE),
    p_burned_quadrats = mean(
      disturbance_num == 1,
      na.rm = TRUE),
    disturbance = if_else(
      any(
        disturbance_num == 1,
        na.rm = TRUE),
      'Fire',
      'No fire'),
    .groups = 'drop') %>%
  left_join(
    lambda_year %>%
      dplyr::select(
        year,
        lambda_environmental_history),
    by = 'year') %>%
  rename(
    asym_lambda = lambda_environmental_history) %>%
  mutate(
    obs_pgr = n_t1 / n_t0,
    proj_lambda = n_proj_model / n_obs_model,
    disturbance = factor(
      disturbance,
      levels = c(
        'No fire',
        'Fire')))

# Observed vs modeled plot

df_plot <- df_compare %>%
  dplyr::select(year, obs_pgr, asym_lambda, proj_lambda, disturbance) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Environmental-history asymptotic lambda',
      proj_lambda = 'Projected lambda from observed quadrat size distributions'))

fig_mod_vs_obs <- ggplot(
  df_plot,
  aes(x = lambda, y = obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = 'Observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs
```

### Log-transformed observed vs modeled plot
```{r}
df_plot_log <- df_plot %>%
  filter(obs_pgr > 0, lambda > 0) %>%
  mutate(
    log_obs_pgr = log(obs_pgr),
    log_lambda = log(lambda))

fig_mod_vs_obs_log <- ggplot(
  df_plot_log,
  aes(x = log_lambda, y = log_obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = paste(v_ggp_suffix, '- log-transformed lambda'),
    x = expression('log modeled ' * lambda),
    y = 'log observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs_log
```

## Summary statistics
```{r}
df_compare_summary <- df_compare %>%
  summarise(
    n_years = n(),
    arithmetic_mean_obs_pgr =
      mean(obs_pgr, na.rm = TRUE),
    geometric_mean_obs_pgr =
      exp(mean(log(obs_pgr), na.rm = TRUE)),
    arithmetic_mean_env_hist_lambda =
      mean(asym_lambda, na.rm = TRUE),
    geometric_mean_env_hist_lambda =
      exp(mean(log(asym_lambda), na.rm = TRUE)),
    arithmetic_mean_proj_lambda =
      mean(proj_lambda, na.rm = TRUE),
    geometric_mean_proj_lambda =
      exp(mean(log(proj_lambda), na.rm = TRUE)),
    mean_error_env_hist_vs_obs =
      mean(
        asym_lambda - obs_pgr,
        na.rm = TRUE),
    mean_error_projected_vs_obs =
      mean(
        proj_lambda - obs_pgr,
        na.rm = TRUE),
    percent_bias_env_hist_vs_obs =
      100 * sum(
        asym_lambda - obs_pgr,
        na.rm = TRUE) /
      sum(
        obs_pgr,
        na.rm = TRUE),
    percent_bias_projected_vs_obs =
      100 * sum(
        proj_lambda - obs_pgr,
        na.rm = TRUE) /
      sum(
        obs_pgr,
        na.rm = TRUE),
    rmse_env_hist_vs_obs =
      sqrt(
        mean(
          (asym_lambda - obs_pgr)^2,
          na.rm = TRUE)),
    rmse_projected_vs_obs =
      sqrt(
        mean(
          (proj_lambda - obs_pgr)^2,
          na.rm = TRUE))) %>%
  pivot_longer(
    cols = everything(),
    names_to = 'statistic',
    values_to = 'value')

df_compare_summary

df_compare %>% print(n = 100)
```

### Site-level summary
```{r}
df_compare_site_summary <- df_compare_site %>%
  filter(
    !is.na(obs_pgr),
    !is.na(asym_lambda),
    !is.na(proj_lambda),
    obs_pgr > 0,
    asym_lambda > 0,
    proj_lambda > 0) %>%
  group_by(site) %>%
  summarise(
    n_year_transitions = n(),
    lambda_obs_geometric =
      exp(mean(log(obs_pgr), na.rm = TRUE)),
    lambda_obs_arithmetic =
      mean(obs_pgr, na.rm = TRUE),
    lambda_env_hist_geometric =
      exp(mean(log(asym_lambda), na.rm = TRUE)),
    lambda_env_hist_arithmetic =
      mean(asym_lambda, na.rm = TRUE),
    lambda_projected_geometric =
      exp(mean(log(proj_lambda), na.rm = TRUE)),
    lambda_projected_arithmetic =
      mean(proj_lambda, na.rm = TRUE),
    error_env_hist_geo_vs_obs_geo =
      lambda_env_hist_geometric -
      lambda_obs_geometric,
    error_projected_geo_vs_obs_geo =
      lambda_projected_geometric -
      lambda_obs_geometric,
    rmse_env_hist_vs_obs =
      sqrt(mean(
        (asym_lambda - obs_pgr)^2,
        na.rm = TRUE)),
    rmse_projected_vs_obs =
      sqrt(mean(
        (proj_lambda - obs_pgr)^2,
        na.rm = TRUE)),
    mean_n_initial =
      mean(n_obs_model, na.rm = TRUE),
    mean_disturbance =
      mean(disturbance_num, na.rm = TRUE),
    .groups = 'drop') %>%
  arrange(as.numeric(as.character(site)))

df_compare_site_summary %>%
  print(
    n = 100,
    width = Inf)
```

## Yearly lambda comparison diagnostic
```{r}
df_lambda_year_check <- lambda_year %>%
  mutate(
    min_state_lambda = pmin(
      lambda_undisturbed,
      lambda_current_fire,
      lambda_previous_fire,
      lambda_consecutive_fire),
    max_state_lambda = pmax(
      lambda_undisturbed,
      lambda_current_fire,
      lambda_previous_fire,
      lambda_consecutive_fire),
    env_hist_minus_undisturbed =
      lambda_environmental_history -
      lambda_undisturbed) %>%
  dplyr::select(
    year,
    lambda_undisturbed,
    lambda_current_fire,
    lambda_previous_fire,
    lambda_consecutive_fire,
    lambda_environmental_history,
    min_state_lambda,
    max_state_lambda,
    env_hist_minus_undisturbed)

df_lambda_year_check %>%
  as_tibble() %>%
  print(
    n = 100,
    width = Inf)
```
