# Yearly historical envi projection model with only disturbance in recruit model {#yr-ipm-hist-envi-re_dist_only}


## Introduction

*Polygala lewtonii* is a short-lived perennial herb occurring in fire-prone Florida scrub. Fire is an important component of the species' environment and affects multiple demographic processes, including survival, growth, flowering, and recruitment. The objective of this integral projection model (IPM) is to describe annual population dynamics while explicitly accounting for both current and previous disturbance conditions and for year-to-year demographic variation.

The IPM is parameterized using annual demographic observations from Archbold Biological Station from 2001 through 2024. Vital-rate models are fitted separately for survival, growth, flowering probability, number of flowering stems, and recruitment. Annual random effects are retained where supported by the data so that the IPM represents observed temporal variation rather than reducing all years to a single mean demographic environment.

### Recruitment model development

Recruitment required particular attention during development of the IPM. The initial model treated recruitment as a conventional fecundity process in which flowering stems produced in year $t$ determined the number of recruits observed in year $t+1$. Recruitment was originally modeled as

$$
\log(1 + R_{q,t+1}),
$$

using a Gaussian mixed-effects model, where $R_{q,t+1}$ is the number of recruits observed in quadrat $q$ in year $t+1$.

Inspection of the raw recruitment data showed that this was not an ideal distributional assumption. Recruitment consisted of non-negative integer counts, contained many zeros, and included occasional very large recruitment events. Recruit abundance ranged from zero to more than 100 individuals per quadrat in some years. Recruitment was therefore modeled directly on the count scale using a negative-binomial generalized linear mixed model with a log link.

Initial negative-binomial models included current disturbance, previous disturbance, flowering-stem abundance, and interactions between flowering and disturbance. The best-supported simple recruitment model included current and previous disturbance,

$$
\log E(R_{q,t+1}) =
\beta_0 +
\beta_D D_{q,t} +
\beta_{D-1} D_{q,t-1} +
u_{0,t+1},
$$

where $D_{q,t}$ represents current disturbance, $D_{q,t-1}$ represents disturbance in the previous year, and $u_{0,t+1}$ is a recruitment-year random intercept.

However, this model still substantially underestimated the largest recruitment events. In particular, recruitment following fire varied enormously among years. Some fires were followed by extremely large recruitment pulses, whereas other fires produced very little recruitment. A single fixed effect of current fire was therefore insufficient.

The recruitment model was subsequently extended so that the effect of current fire could vary among recruitment years:

$$
\log E(R_{q,t+1}) =
\beta_0 +
\beta_D D_{q,t} +
\beta_{D-1} D_{q,t-1} +
u_{0,t+1} +
u_{D,t+1}D_{q,t}.
$$

Here, $u_{D,t+1}$ is a year-specific random deviation from the average effect of current disturbance.

The resulting R model is:

```{r, eval=FALSE}
mod_re_yrfire <- glmer.nb(
  re_t1 ~ disturbance + disturbance_prev +
    (1 + disturbance || year_t1),
  data = df_fs2r_mod,
  control = ctrl_re)
```

This model was strongly supported relative to recruitment models containing a single fire effect. The standard deviation of the year-specific current-fire effect was approximately 1.69 on the log scale, indicating very strong temporal variation in recruitment following disturbance.

The estimated total current-fire multipliers differed substantially among recruitment years:

- 2002: 7.75
- 2008: 1.63
- 2010: 4.36
- 2016: 23.6
- 2019: 0.14
- 2020: 1.04
- 2022: 0.63

Thus, fire cannot be represented as producing one consistent recruitment response. The 2016 fire, for example, was associated with an extremely large positive recruitment response, whereas the 2019 fire was associated with very little recruitment.

## Investigation of alternative explanations for recruitment pulses

Several possible explanations for the strong variation among fire years were evaluated.

Current flowering-stem abundance did not improve the year-varying fire model. Models containing flowering stems or an interaction between flowering stems and current disturbance were less well supported. This indicated that the large recruitment pulses were not adequately explained by the number of flowering stems immediately preceding recruitment.

A hurdle formulation was also considered. This separated recruitment into the probability that any recruitment occurred and the number of recruits conditional on recruitment being greater than zero. Although the combined hurdle model had a lower AIC than the single negative-binomial model, it performed worse when evaluated against annual recruitment totals. Because annual recruitment is the quantity most directly relevant to the population projection, the single negative-binomial model was retained.

Fire severity was also investigated. Recruitment tended to increase with fire severity on average, but severity did not explain the extreme differences among fire years. The same high-severity fire category could be associated with very large recruitment in one year and little recruitment in another year.

Historical flowering was investigated as a possible proxy for the seed bank. Flowering one and two years before recruitment contained information about subsequent recruitment, and models containing historical flowering had lower AIC values on subsets of the data with sufficient flowering history. In particular, flowering approximately two years before recruitment was positively associated with subsequent recruit abundance.

However, these historical-flowering models performed worse when evaluated against annual recruitment totals. They substantially underestimated the 2016 recruitment pulse and overestimated recruitment in some low-recruitment fire years. Historical flowering therefore contains biologically relevant information but does not provide a sufficient explanation for the observed variation among fire years.

The year-varying negative-binomial fire model was consequently retained as the recruitment model.

### Historical reconstruction versus prediction

An important distinction emerged between historical reconstruction and prediction of an unknown future fire year.

When the fitted year-specific random effects were included, the recruitment model reproduced historical annual recruitment much better than a model containing a single fire effect. Annual recruitment RMSE declined from approximately 60.4 recruits to 21.8 recruits, and MAE declined from approximately 26.2 to 13.9 recruits.

The model therefore provides a useful historical description of realized recruitment conditions.

However, leave-one-fire-year-out tests showed that the model cannot determine whether an unknown fire year will behave like a high-recruitment event such as 2002 or 2016 or like a low-recruitment event such as 2019 or 2022. When each fire year was removed in turn and predicted as an unseen year, large recruitment events were strongly underestimated and some low-recruitment years were strongly overestimated.

For example, recruitment in 2002 was 722 individuals in the current-fire quadrats but was predicted to be approximately 61 when that year was omitted from model fitting. Recruitment in 2016 was 305 but was predicted to be only approximately 13. Conversely, recruitment in 2019 was only two individuals but was predicted to exceed 100 when that year was treated as unknown.

The year-specific random effects should therefore be interpreted as estimates of realized historical environmental variation rather than as predictors of the response to a future unknown fire.

For historical year-specific IPMs, the fitted random effect associated with each observed recruitment year is retained. For example, recruitment occurring in 2016 is associated with the 2015--2016 population transition. The estimated current-fire coefficient for this transition is

$$
0.6615 + 2.5013 = 3.1628,
$$

corresponding to a recruitment multiplier of

$$
\exp(3.1628) \approx 23.6.
$$

In contrast, the 2018--2019 transition has a total current-fire coefficient of approximately -1.97, corresponding to a multiplier of only approximately 0.14.

Future stochastic projections would require drawing the fire response from the estimated distribution of year-specific fire effects, or identifying additional environmental or seed-bank variables that explain this variation.

### Incorporating recruitment into the IPM

The statistical recruitment model created an additional issue when it was incorporated into the IPM.

The original IPM treated recruitment as part of a conventional fecundity kernel. The expected number of recruits at a reference level of flowering was first converted to recruits per flowering stem. This quantity was then multiplied by the flowering output predicted for each individual according to its size.

Conceptually, the original formulation was equivalent to

$$
R(x) =
FS(x)
\frac{R_{\mathrm{ref}}}{FS_{\mathrm{ref}}},
$$

where $FS(x)$ is the expected flowering-stem production of an individual of size $x$, $R_{\mathrm{ref}}$ is predicted recruitment at a reference level of flowering, and $FS_{\mathrm{ref}}$ is the reference number of flowering stems.

This created a contradiction with the selected recruitment model. Current flowering-stem abundance was not present in the best-supported recruitment model, yet the IPM subsequently forced recruitment to depend on current flowering output.

Diagnostics showed that this rescaling strongly distorted the recruitment predictions.

For the seven burned quadrats in the 2015--2016 transition, the recruitment model predicted approximately 33 recruits per quadrat. After rescaling through predicted flowering, individual quadrat recruitment contributions ranged from almost zero to more than 129 recruits.

Across historical fire years, the proportion of the direct recruitment prediction retained after flowering rescaling was approximately:

- 2001: 0.53
- 2007: 0.84
- 2009: 0.90
- 2015: 0.64
- 2018: 0.36
- 2019: 2.34
- 2021: 2.73

Thus, the flowering-based fecundity calculation could either strongly suppress or strongly inflate recruitment relative to the statistical recruitment model.

The historical projection was therefore changed so that recruitment enters directly as a quadrat-level recruitment pulse rather than being rescaled by current flowering.

### Direct-recruitment population projection

The historical one-year population projection is formulated as

$$
N_{t+1} = T_tN_t + R_t,
$$

where $N_t$ is the observed size distribution in year $t$, $T_t$ is the survival-growth kernel for that year and disturbance condition, and $R_t$ is the number of recruits predicted for the corresponding quadrat and recruitment year.

Existing individuals are projected using survival and growth:

$$
N_{\mathrm{surv+growth},t+1} = T_tN_t.
$$

Recruit abundance is predicted directly using the negative-binomial recruitment model and then distributed across the continuous size state using the observed recruit-size distribution.

The projected population is therefore

$$
N_{\mathrm{projected},t+1}
=
N_{\mathrm{surv+growth},t+1}
+
N_{\mathrm{recruits},t+1}.
$$

The corresponding one-year projected population growth rate is

$$
\lambda_{\mathrm{proj},t}
=
\frac{N_{\mathrm{projected},t+1}}
     {N_{\mathrm{observed},t}}.
$$

### Improvement in historical population projections

Allowing recruitment to enter the projection directly substantially improved agreement between observed and projected annual population growth.

Under the previous flowering-scaled recruitment formulation, the comparison between projected and observed annual population growth had approximately:

- RMSE = 0.652
- MAE = 0.382
- mean bias = -0.255

With direct recruitment, these values improved to approximately:

- RMSE = 0.275
- MAE = 0.155
- mean bias = -0.132

The improvement was particularly clear during several important disturbance transitions.

For the 2001--2002 transition, observed population growth was approximately 4.59 and projected population growth was approximately 4.34.

For the 2007--2008 transition, observed and projected growth were approximately 1.05 and 0.97.

For the 2009--2010 transition, observed and projected growth were approximately 1.07 and 1.04.

For the large 2015--2016 recruitment event, observed growth was approximately 2.14 and projected growth increased to approximately 1.76.

For the 2018--2019 transition, observed and projected growth were approximately 0.59 and 0.56.

For the 2019--2020 transition, observed growth was approximately 2.40 and projected growth was approximately 1.87.

For the 2021--2022 transition, observed and projected growth were approximately 0.79 and 0.76.

### Remaining model error

The annual projection was separated into recruitment and survival-growth components to determine the origin of the remaining prediction error.

Across all years, the recruitment component had approximately:

- RMSE = 21.8 individuals
- MAE = 13.9 individuals

The combined survival-growth component had approximately:

- RMSE = 66.0 individuals
- MAE = 32.9 individuals

Thus, across the full time series, survival and growth now represent a larger general source of absolute prediction error than recruitment.

The exceptional 2015--2016 transition was somewhat different. In that transition, 367 recruits were observed and approximately 298 were predicted. Among plants that were not new recruits, 76 were observed in 2016 and approximately 66 were predicted through survival and growth.

The total observed population in 2016 was therefore 443 individuals, compared with a projection of approximately 364 individuals. Of the approximately 79 missing individuals, roughly 69 resulted from underprediction of recruitment and roughly 10 resulted from underprediction of survival and growth.

The remaining failure to reproduce the complete 2016 population increase is therefore primarily associated with the exceptional magnitude of recruitment in that year rather than with further suppression of recruitment inside the IPM.

### Interpretation of the recruitment process

The resulting model changes the biological interpretation of recruitment in the IPM.

Recruitment is not well described as a simple function of the immediately preceding above-ground flowering population. Instead, recruitment behaves as a strongly temporally variable process in which fire can release very large recruitment pulses in some years but not others.

This pattern is compatible with recruitment being influenced by processes that are not completely represented by observed current flowering. *Polygala lewtonii* is amphicarpic, and the available flowering measurements do not represent all possible reproductive output. A persistent seed bank could therefore provide a biological mechanism connecting reproduction in previous years with episodic recruitment following disturbance.

Historical flowering analyses provided some support for an influence of past flowering, particularly flowering approximately two years before recruitment. However, these variables did not adequately explain the extreme differences among fire years and did not perform as well as the year-varying fire model for annual recruitment totals.

An explicit seed-bank state may ultimately provide a more mechanistic representation of this process, but the available data are currently insufficient to parameterize such a state with confidence. The current historical model therefore represents recruitment directly rather than imposing an unsupported dependence on current flowering.

### Projected and asymptotic population growth

The direct-recruitment formulation creates an important distinction between historical one-year projected population growth and conventional asymptotic population growth.

A standard linear IPM can be written as

$$
N_{t+1} = K_tN_t,
$$

where

$$
K_t = T_t + F_t.
$$

Because the projection is linear in $N_t$, the dominant eigenvalue of $K_t$ can be interpreted as the asymptotic population growth rate associated with that kernel.

The historical model developed here instead contains an additive recruitment component:

$$
N_{t+1} = T_tN_t + R_t.
$$

The recruitment pulse $R_t$ is predicted at the quadrat level and is not proportional to the current above-ground population. This makes the historical projection affine rather than strictly linear.

The annual projected population growth rate can still be calculated directly from observed and projected population abundance as

$$
\lambda_{\mathrm{proj},t}
=
\frac{N_{\mathrm{projected},t+1}}
     {N_{\mathrm{observed},t}},
$$

but the additive recruitment pulse cannot simply be inserted into a conventional fecundity kernel and interpreted through its dominant eigenvalue.

Consequently, the historical projected growth rate and the asymptotic eigenvalue-based growth rate represent different quantities under the current model formulation. The direct-recruitment model is used for historical one-year projections because it is consistent with the fitted recruitment model and provides substantially better agreement with observed population dynamics.

Development of a fully linear long-term IPM would require recruitment to arise from a state contained within the population vector, such as an explicitly modeled seed bank. Until such a state can be parameterized, historical one-year projections and long-term asymptotic projections should be treated as separate components of the demographic analysis.



## Working dataset
```{r}
df <- read.csv(
  file.path(
    dir_data,
    paste0('ab_', v_sp_abb, '_df_workdata_260519.csv'))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.integer(year),
    site = factor(site)) %>%
  filter(
    !is.na(year),
    !(year %in% v_years_re))
```



## Controls
```{r}
ctrl_glmer <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))
```

## Vital rates
### Survival
#### Survival data
```{r}
df_su <- df %>%
  filter(
    !is.na(survives),
    size_t0 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = factor(disturbance, levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage)
```

#### Survival model
```{r}
mod_su_0 <- glmer(
  survives ~ disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_10 <- glmer(
  survives ~ logvol_t0 + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_11 <- glmer(
  survives ~ logvol_t0 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_20 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_21 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_30 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)
mod_su_31 <- glmer(
  survives ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 + disturbance +
    (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)


mods_su <- list(
  mod_su_0, mod_su_10, mod_su_11, 
  mod_su_20, mod_su_21, mod_su_30, mod_su_31)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef <- coef(mod_su_best)$year

mod_su_best
summary(mod_su_best)
mods_su_dAIC
```

#### Survival plots by year
```{r, fig.width=10, fig.height=12}
make_su_year_plot <- function(year_i) {
  df_i <- df_su %>% filter(year == year_i)
  
  x <- seq(
    min(df_i$logvol_t0, na.rm = TRUE),
    max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance = levels(df_su$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance = factor(
        disturbance, levels = levels(df_su$disturbance)),
      year = factor(year_i, levels = levels(df_su$year)))
  
  pred_i$survives <- predict(
    mod_su_best, newdata = pred_i,
    type = "response", re.form = NULL)
  
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      survives = mean(survives, na.rm = TRUE),
      n = n(), .groups = "drop") %>%
    filter(!is.na(logvol_t0), n > 0)
  
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, survives, color = disturbance), size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, survives, color = disturbance,
          linetype = disturbance),
      linewidth = 0.7) +
    scale_color_manual(values = c("0" = "black", "1" = "red")) +
    labs(
      title = year_i,
      x = expression("log(volume)"[t0]),
      y = expression("Survival probability"[t1])) +
    ylim(0, 1) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = "none")
}

su_yrs <- lapply(levels(df_su$year), make_su_year_plot)
fig_su_years <- wrap_plots(su_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_su_years
```

### Growth
#### Growth data
```{r}
df_gr <- df %>%
  filter(
    size_t0 != 0,
    size_t1 != 0,
    is.finite(logvol_t0),
    is.finite(logvol_t1),
    is.finite(logvol_t0_2),
    is.finite(logvol_t0_3),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = factor(
      disturbance,
      levels = c(0, 1))) %>%
  dplyr::select(
    id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    volume_t0, volume_t1, logvol_t0, logvol_t1,
    logvol_t0_2, logvol_t0_3,
    disturbance, stage)
```

#### Growth model
```{r}
mod_gr_0 <- lmer(
  logvol_t1 ~ disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_10 <- lmer(
  logvol_t1 ~ logvol_t0 + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_11 <- lmer(
  logvol_t1 ~ logvol_t0 + disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_20 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_21 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + disturbance +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_30 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)
mod_gr_31 <- lmer(
  logvol_t1 ~ logvol_t0 + logvol_t0_2 + logvol_t0_3 +
    disturbance + (logvol_t0 | year),
  data = df_gr, REML = FALSE, control = ctrl_lmer)


mods_gr <- list(
  mod_gr_0, mod_gr_10, mod_gr_11, mod_gr_20, mod_gr_21, 
  mod_gr_30, mod_gr_31)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef <- coef(mod_gr_best)$year

mod_gr_best
summary(mod_gr_best)
mods_gr_dAIC
```

#### Growth plots by year
```{r, fig.width=10, fig.height=12}
make_gr_year_plot <- function(year_i) {
  df_i <- df_gr %>% filter(year == year_i)
  x <- seq(min(df_i$logvol_t0, na.rm = TRUE),
           max(df_i$logvol_t0, na.rm = TRUE), length.out = 100)
  pred_i <- tidyr::expand_grid(
    logvol_t0 = x,
    disturbance = levels(df_gr$disturbance)) %>%
    mutate(
      logvol_t0_2 = logvol_t0^2,
      logvol_t0_3 = logvol_t0^3,
      disturbance = factor(
        disturbance, levels = levels(df_gr$disturbance)),
      year = factor(year_i, levels = levels(df_gr$year)))
  
  pred_i <- pred_i %>%
    mutate(
      logvol_t1 = predict(
        mod_gr_best,
        newdata = pred_i,
        re.form = NULL))
  pts_i <- df_i %>%
    mutate(bin = cut(logvol_t0, breaks = 8)) %>%
    group_by(bin, disturbance) %>%
    summarise(
      logvol_t0 = mean(logvol_t0, na.rm = TRUE),
      logvol_t1 = mean(logvol_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(!is.na(logvol_t0), n > 0)
  ggplot() +
    geom_point(
      data = pts_i,
      aes(logvol_t0, logvol_t1, color = disturbance),
      size = 1.1) +
    geom_line(
      data = pred_i,
      aes(logvol_t0, logvol_t1, color = disturbance,
          linetype = disturbance),
      linewidth = 0.7) +
    geom_abline(intercept = 0, slope = 1, color = 'blue', lty = 2) +
    scale_color_manual(values = c('0' = 'black', '1' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(volume)'[t0]),
      y = expression('log(volume)'[t1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

gr_yrs <- lapply(levels(df_gr$year), make_gr_year_plot)
fig_gr_years <- wrap_plots(gr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Growth - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_gr_years
```

#### Growth variance
```{r}
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

### Recruitment
#### Recruitment data
```{r}
df_quad_year_sampled <- df %>%
  distinct(site, quad, year)

df_re <- df %>%
  group_by(site, quad, year) %>%
  summarise(
    disturbance = first(
      disturbance[!is.na(disturbance)], default = NA_real_),
    disturbance_prev = first(
      disturbance_prev[!is.na(disturbance_prev)], default = NA_real_),
    .groups = 'drop') %>%
  mutate(year_t1 = year + 1) %>%
  semi_join(
    df_quad_year_sampled %>%
      transmute(site, quad, year_t1 = year),
    by = c('site', 'quad', 'year_t1')) %>%
  left_join(
    df %>%
      filter(recruits == 1) %>%
      count(site, quad, year, name = 're_t1'),
    by = c('site', 'quad', 'year_t1' = 'year')) %>%
  mutate(re_t1 = replace_na(re_t1, 0L)) %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !is.na(disturbance),
    !is.na(disturbance_prev),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re)) %>%
  mutate(
    site = factor(site),
    year_t0 = factor(year),
    year_t1 = factor(year_t1))
```

#### Recruit model
```{r}
ctrl_re <- glmerControl(
  optimizer = 'bobyqa',
  optCtrl = list(maxfun = 2e5))

# Intercept only
mod_re_00 <- glmer.nb(
  re_t1 ~ 1 +
    (1 | year_t1),
  data = df_re,
  control = ctrl_re)


# Current disturbance only
mod_re_0 <- glmer.nb(
  re_t1 ~ disturbance +
    (1 | year_t1),
  data = df_re,
  control = ctrl_re)


# Current + previous disturbance
mod_re_01 <- glmer.nb(
  re_t1 ~ disturbance + disturbance_prev +
    (1 | year_t1),
  data = df_re,
  control = ctrl_re)


# Current fire effect varies among recruitment years
mod_re_yrfire <- glmer.nb(
  re_t1 ~ disturbance + disturbance_prev +
    (1 + disturbance || year_t1),
  data = df_re,
  control = ctrl_re)

mods_re <- list(
  mod_re_00,
  mod_re_0,
  mod_re_01,
  mod_re_yrfire)

mods_re_dAIC <- bbmle::AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAIC

mods_re_sorted <- order(mods_re_dAIC)

if (length(v_mod_set_re) == 0) {
  mod_re_index_bestfit <- mods_re_sorted[1]
} else {
  mod_re_index_bestfit <- v_mod_set_re + 1
}

v_mod_re_index <- mod_re_index_bestfit - 1
mod_re_best <- mods_re[[mod_re_index_bestfit]]

mod_re_best
summary(mod_re_best)
mods_re_dAIC
```

#### Recruitment plots by recruit year
```{r, fig.width=10, fig.height=12}
fire_levels <- c(
  'No fire',
  'Current-year fire',
  'Previous-year fire',
  'Fire in both years')

df_re_plot <- df_re %>%
  mutate(
    fire_group = case_when(
      disturbance == 0 & disturbance_prev == 0 ~ 'No fire',
      disturbance == 1 & disturbance_prev == 0 ~ 'Current-year fire',
      disturbance == 0 & disturbance_prev == 1 ~ 'Previous-year fire',
      disturbance == 1 & disturbance_prev == 1 ~ 'Fire in both years'),
    fire_group = factor(fire_group, levels = fire_levels),
    re_t1_log = log1p(re_t1))

df_re_pred <- expand_grid(
  year_t1 = levels(df_re$year_t1),
  disturbance = c(0, 1),
  disturbance_prev = c(0, 1)) %>%
  mutate(
    year_t1 = factor(year_t1, levels = levels(df_re$year_t1)),
    fire_group = case_when(
      disturbance == 0 & disturbance_prev == 0 ~ 'No fire',
      disturbance == 1 & disturbance_prev == 0 ~ 'Current-year fire',
      disturbance == 0 & disturbance_prev == 1 ~ 'Previous-year fire',
      disturbance == 1 & disturbance_prev == 1 ~ 'Fire in both years'),
    fire_group = factor(fire_group, levels = fire_levels),
    pred_re = predict(
      mod_re_best, newdata = ., type = 'response',
      re.form = NULL, allow.new.levels = TRUE),
    pred_re_log = log1p(pred_re))

fig_re_years <- ggplot(
  df_re_plot,
  aes(x = fire_group, y = re_t1_log, color = fire_group)) +
  geom_jitter(width = 0.15, height = 0, alpha = 0.4, size = 1) +
  geom_point(
    data = df_re_pred,
    aes(y = pred_re_log),
    shape = 4, size = 3, stroke = 1) +
  scale_color_manual(
    values = c(
      'No fire' = 'black',
      'Current-year fire' = 'red',
      'Previous-year fire' = 'purple',
      'Fire in both years' = 'magenta')) +
  facet_wrap(~ year_t1, ncol = 4) +
  labs(
    title = 'Recruitment - year specific',
    subtitle = paste(
      v_ggp_suffix,
      '- crosses are model predictions'),
    x = 'Fire history',
    y = 'log(1 + recruits)',
    color = 'Fire history') +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1, size = 5),
    axis.text.y = element_text(size = 5),
    strip.text = element_text(size = 6),
    legend.position = 'bottom')

fig_re_years
```

#### Recruit size distribution
```{r}
df_recr <- df %>%
  filter(recruits == 1, volume_t0 > 0, is.finite(logvol_t0))

rc_sz <- data.frame(
  mean = mean(df_recr$logvol_t0, na.rm = TRUE),
  sd = sd(df_recr$logvol_t0, na.rm = TRUE))
```

## Parameter extraction
### Parameter extraction helpers
```{r}
empty_coef_df <- function() {
  data.frame(coefficient = character(), value = numeric())
}

is_mixed_model <- function(model) {
  inherits(model, 'merMod')
}

get_fixef_safe <- function(model) {
  if (is_mixed_model(model)) {
    fixef(model)
  } else {
    coef(model)
  }
}

get_group_coef_safe <- function(model, group_var) {
  if (!is_mixed_model(model)) {
    return(NULL)
  }
  coef_list <- coef(model)
  if (!(group_var %in% names(coef_list))) {
    return(NULL)
  }
  coef_list[[group_var]]
}

find_model_term <- function(model_terms, aliases) {
  aliases <- unlist(aliases)
  for (alias in aliases) {
    if (grepl('^regex:', alias)) {
      pattern <- sub('^regex:', '', alias)
      hit <- grep(pattern, model_terms, value = TRUE)
    } else {
      hit <- model_terms[model_terms == alias]
    }
    if (length(hit) > 0) {
      return(hit[1])
    }
  }
  NA_character_
}

bind_coef_rows <- function(x) {
  x <- x[!vapply(x, is.null, logical(1))]
  if (length(x) == 0) {
    return(empty_coef_df())
  }
  out <- do.call(rbind, x)
  rownames(out) <- NULL
  out
}

check_duplicate_coefficients <- function(x, object_name = 'parameter table') {
  dup <- x$coefficient[duplicated(x$coefficient)]
  if (length(dup) > 0) {
    stop(
      object_name,
      ' contains duplicated coefficient names: ',
      paste(unique(dup), collapse = ', '))
  }
  invisible(TRUE)
}

coef_df_to_list <- function(x) {
  if (nrow(x) == 0) {
    return(list())
  }
  check_duplicate_coefficients(x)
  x %>%
    mutate(coefficient = as.character(coefficient)) %>%
    pivot_wider(names_from = coefficient, values_from = value) %>%
    as.list()
}

extract_fixed_pars <- function(model, term_map, fill_missing = TRUE) {
  model_coefs <- get_fixef_safe(model)
  model_terms <- names(model_coefs)
  out <- lapply(names(term_map), function(coef_name) {
    model_term <- find_model_term(model_terms, term_map[[coef_name]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- 0
      } else {
        return(NULL)
      }
    } else {
      value <- unname(model_coefs[model_term])
    }
    data.frame(coefficient = coef_name, value = value)
  })
  bind_coef_rows(out)
}

extract_group_pars <- function(model, group_var, term_map, fill_missing = TRUE) {
  coef_matrix <- get_group_coef_safe(model, group_var)
  if (is.null(coef_matrix)) {
    return(empty_coef_df())
  }
  model_terms <- colnames(coef_matrix)
  out <- lapply(names(term_map), function(coef_prefix) {
    model_term <- find_model_term(model_terms, term_map[[coef_prefix]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- rep(0, nrow(coef_matrix))
      } else {
        return(NULL)
      }
    } else {
      value <- coef_matrix[, model_term]
    }
    data.frame(
      coefficient = paste0(coef_prefix, '_', rownames(coef_matrix)),
      value = value)
  })
  bind_coef_rows(out)
}

extract_re_ranef_devs <- function(model, group_var, group_label) {
  if (!inherits(model, 'merMod')) {
    return(empty_coef_df())
  }
  
  ranef_list <- ranef(model)
  
  if (!(group_var %in% names(ranef_list))) {
    return(empty_coef_df())
  }
  
  ranef_matrix <- ranef_list[[group_var]]
  out <- list()
  
  if ('(Intercept)' %in% colnames(ranef_matrix)) {
    out[['intercept']] <- data.frame(
      coefficient = paste0(
        're_u0_', group_label, '_', rownames(ranef_matrix)),
      value = ranef_matrix[, '(Intercept)'])
  }
  
  if ('disturbance' %in% colnames(ranef_matrix)) {
    out[['disturbance']] <- data.frame(
      coefficient = paste0(
        're_ud_', group_label, '_', rownames(ranef_matrix)),
      value = ranef_matrix[, 'disturbance'])
  }
  
  bind_coef_rows(out)
}
```

### Term maps
```{r}
size_term_map <- list(
  b0 = c('(Intercept)'),
  b1 = c('logvol_t0'),
  b2 = c('logvol_t0_2', 'I(logvol_t0^2)'),
  b3 = c('logvol_t0_3', 'I(logvol_t0^3)'),
  bf = c('disturbance1', 'regex:^disturbance'))

make_term_map <- function(prefix, map) {
  out <- map
  names(out) <- paste0(prefix, names(out))
  out
}

su_term_map <- make_term_map('surv_', size_term_map)
gr_term_map <- make_term_map('grow_', size_term_map)

re_term_map <- list(
  re_b0 = c('(Intercept)'),
  re_bd = c('disturbance'),
  re_bdp = c('disturbance_prev'))


# Fixed parameters -------------------------------------------------------------
su_fe <- extract_fixed_pars(mod_su_best, su_term_map, fill_missing = TRUE)
gr_fe <- extract_fixed_pars(mod_gr_best, gr_term_map, fill_missing = TRUE)
re_fe <- extract_fixed_pars(mod_re_best, re_term_map, fill_missing = TRUE)


# Constants --------------------------------------------------------------------
gr_var_coef <- coef(mod_gr_var)

constants <- tibble::tribble(
  ~coefficient, ~value,
  'recr_sz', rc_sz$mean,
  'recr_sd', rc_sz$sd,
  'a', as.numeric(gr_var_coef[1]),
  'b', as.numeric(gr_var_coef[2]),
  'L', min(c(
    df_su$logvol_t0,
    df_gr$logvol_t0,
    df_recr$logvol_t0), na.rm = TRUE) - 0.1,
  'U', max(c(
    df_su$logvol_t0,
    df_gr$logvol_t0,
    df_recr$logvol_t0), na.rm = TRUE) + 0.1,
  'mat_siz', 200,
  'mod_su_index', v_mod_su_index,
  'mod_gr_index', v_mod_gr_index,
  'mod_re_index', v_mod_re_index) %>%
  mutate(coefficient = as.character(coefficient), value = as.numeric(value))

pars_cons <- bind_coef_rows(list(
  su_fe,
  gr_fe,
  re_fe,
  constants))

check_duplicate_coefficients(pars_cons, object_name = 'pars_cons')
pars_all_mean <- coef_df_to_list(pars_cons)
pars_mean <- pars_all_mean


# Year-varying parameters ------------------------------------------------------
su_out_yr <- extract_group_pars(
  mod_su_best, 'year', su_term_map, fill_missing = TRUE)

gr_out_yr <- extract_group_pars(
  mod_gr_best, 'year', gr_term_map, fill_missing = TRUE)

pars_var <- bind_coef_rows(list(
  su_out_yr,
  gr_out_yr))

check_duplicate_coefficients(pars_var, object_name = 'pars_var')
pars_all_year <- coef_df_to_list(pars_var)


# Recruitment-year random-effect deviations ---------------------------
pars_re_year_t1 <- extract_re_ranef_devs(
  mod_re_best, 'year_t1', 'year_t1') %>%
  mutate(coefficient = as.character(coefficient))
check_duplicate_coefficients(
  pars_re_year_t1, object_name = 'pars_re_year_t1')
pars_all_re_year_t1 <- coef_df_to_list(pars_re_year_t1)
```

### Disturbance transition probabilities
```{r}
# Current disturbance refers to t -> t + 1 and previous disturbance to
# t - 1 -> t.

df_disturbance_regime <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year),
    !is.na(disturbance),
    !is.na(disturbance_prev)) %>%
  distinct(
    site,
    quad,
    year,
    disturbance,
    disturbance_prev) %>%
  count(
    disturbance,
    disturbance_prev,
    name = 'n_quadrat_years') %>%
  complete(
    disturbance = c(0, 1),
    disturbance_prev = c(0, 1),
    fill = list(n_quadrat_years = 0)) %>%
  arrange(
    disturbance_prev,
    disturbance)

df_disturbance_regime


df_disturbance_transition <- df_disturbance_regime %>%
  group_by(disturbance_prev) %>%
  mutate(
    p_transition = n_quadrat_years /
      sum(n_quadrat_years)) %>%
  ungroup()

df_disturbance_transition


get_transition_probability <- function(
    disturbance_i,
    disturbance_prev_i) {
  
  df_disturbance_transition %>%
    filter(
      disturbance == disturbance_i,
      disturbance_prev == disturbance_prev_i) %>%
    pull(p_transition)
}


p_0_given_0 <- get_transition_probability(
  disturbance_i = 0,
  disturbance_prev_i = 0)

p_1_given_0 <- get_transition_probability(
  disturbance_i = 1,
  disturbance_prev_i = 0)

p_0_given_1 <- get_transition_probability(
  disturbance_i = 0,
  disturbance_prev_i = 1)

p_1_given_1 <- get_transition_probability(
  disturbance_i = 1,
  disturbance_prev_i = 1)


df_disturbance_transition
```

## IPM helper functions
```{r}
inv_logit <- function(x) {
  plogis(x)
}

get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    return(pars[[par_name]])
  }
  default
}

make_ipm_pars <- function(
    pars_mean,
    pars_year = NULL,
    pars_re_year_t1 = NULL,
    year = NULL,
    re_year_t1 = NULL) {
  pars <- pars_mean
  
  if (!is.null(year) && !is.null(pars_year)) {
    year <- as.character(year)
    year_hits <- grep(paste0('_', year, '$'), names(pars_year), value = TRUE)
    
    for (nm in year_hits) {
      base_nm <- sub(paste0('_', year, '$'), '', nm)
      pars[[base_nm]] <- pars_year[[nm]]
    }
  }
  
  if (is.null(re_year_t1) && !is.null(year)) {
    re_year_t1 <- as.numeric(as.character(year)) + 1
  }
  
  if (!is.null(re_year_t1) && !is.null(pars_re_year_t1)) {
    re_year_t1 <- as.character(re_year_t1)
    pars$re_b0 <- get_par(pars, 're_b0', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_u0_year_t1_', re_year_t1), 0)
    
    pars$re_bd <- get_par(pars, 're_bd', 0) +
      get_par(
        pars_re_year_t1,
        paste0('re_ud_year_t1_', re_year_t1), 0)
  }
  
  pars
}

surv_lp <- function(x, pars, disturbance = 0) {
  get_par(pars, 'surv_b0', 0) +
    get_par(pars, 'surv_b1', 0) * x +
    get_par(pars, 'surv_b2', 0) * x^2 +
    get_par(pars, 'surv_b3', 0) * x^3 +
    get_par(pars, 'surv_bf', 0) * disturbance
}

sx <- function(x, pars, disturbance = 0) {
  inv_logit(surv_lp(x, pars, disturbance))
}

grow_mu <- function(x, pars, disturbance = 0) {
  get_par(pars, 'grow_b0', 0) +
    get_par(pars, 'grow_b1', 0) * x +
    get_par(pars, 'grow_b2', 0) * x^2 +
    get_par(pars, 'grow_b3', 0) * x^3 +
    get_par(pars, 'grow_bf', 0) * disturbance
}

grow_sd <- function(x, pars) {
  sqrt(pars$a * exp(pars$b * x))
}

gxy <- function(x, y, pars, disturbance = 0) {
  dnorm(
    y,
    mean = grow_mu(x, pars, disturbance),
    sd = grow_sd(x, pars))
}


re_total_ref <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  
  re_eta <- get_par(pars, 're_b0', 0) +
    get_par(pars, 're_bd', 0) * disturbance +
    get_par(pars, 're_bdp', 0) * disturbance_prev
  
  pmax(exp(re_eta), 0)
}

re_y_dist <- function(y, pars, h = NULL) {
  dens <- dnorm(y, mean = pars$recr_sz, sd = pars$recr_sd)
  
  if (!is.null(h)) {
    dens <- dens / sum(dens * h)
  }
  
  dens
}
```

## Single-state size-structured kernel
```{r}
kernel <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  S <- sx(y, pars, disturbance = disturbance)
  
  G <- t(outer(
    y, y, Vectorize(function(x, y) {
      gxy(x, y, pars, disturbance = disturbance)
    }))) * h
  
  for (i in 1:n) {
    if (i <= n / 2) {
      G[1, i] <- G[1, i] + 1 - sum(G[, i])
    } else {
      G[n, i] <- G[n, i] + 1 - sum(G[, i])
    }
  }
  
  T <- sweep(G, 2, S, '*')
  
  re_total <- re_total_ref(
    pars = pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
  
  R <- re_total *
    re_y_dist(
      y,
      pars,
      h = h)
  
  A_aug <- rbind(
    cbind(T, R),
    c(rep(0, n), 1))
  
  list(
    A_aug = A_aug,
    T = T,
    R = R,
    G = G,
    S = S,
    meshpts = y,
    h = h,
    L = L,
    U = U)
}

lambda_ipm <- function(
    pars,
    disturbance = 0,
    disturbance_prev = 0) {
  
  A <- kernel(
    pars = pars,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)$A_aug
  
  max(Mod(
    eigen(
      A,
      only.values = TRUE)$values))
}


# Environmental-history kernel ------------------------------------------------
kernel_env_hist <- function(pars) {
  
  A_00 <- kernel(
    pars = pars,
    disturbance = 0,
    disturbance_prev = 0)$A_aug
  
  A_10 <- kernel(
    pars = pars,
    disturbance = 1,
    disturbance_prev = 0)$A_aug
  
  A_01 <- kernel(
    pars = pars,
    disturbance = 0,
    disturbance_prev = 1)$A_aug
  
  A_11 <- kernel(
    pars = pars,
    disturbance = 1,
    disturbance_prev = 1)$A_aug
  
  A_env <- rbind(
    cbind(
      p_0_given_0 * A_00,
      p_0_given_1 * A_01),
    cbind(
      p_1_given_0 * A_10,
      p_1_given_1 * A_11))
  
  list(
    A_env = A_env,
    A_00 = A_00,
    A_10 = A_10,
    A_01 = A_01,
    A_11 = A_11)
}


# Environmental-history asymptotic lambda -------------------------------------
lambda_env_hist <- function(pars) {
  
  A <- kernel_env_hist(
    pars = pars)$A_env
  
  max(Mod(
    eigen(
      A,
      only.values = TRUE)$values))
}
```


## Mean IPMs
```{r}
lambda_ipm(pars_mean, disturbance = 0, disturbance_prev = 0)
lambda_ipm(pars_mean, disturbance = 0, disturbance_prev = 1)
lambda_ipm(pars_mean, disturbance = 1, disturbance_prev = 1)
lambda_ipm(pars_mean, disturbance = 1, disturbance_prev = 0)

lambda_env_hist(pars_mean)
```

## Year-specific IPMs
```{r}
lambda_ipm_year <- function(
    year,
    disturbance = 0,
    disturbance_prev = 0,
    re_year_t1 = NULL) {
  pars_i <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = year,
    re_year_t1 = re_year_t1)
  
  lambda_ipm(
    pars_i,
    disturbance = disturbance,
    disturbance_prev = disturbance_prev)
}


# Year-specific environmental-history lambda -----------------------------------
lambda_env_hist_year <- function(
    year,
    re_year_t1 = NULL) {
  
  pars_i <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = year,
    re_year_t1 = re_year_t1)
  
  lambda_env_hist(
    pars = pars_i)
}


# Year-specific disturbance scenarios ------------------------------------------
ipm_years <- sort(unique(as.integer(as.character(df_su$year))))

lambda_year <- data.frame(
  year = ipm_years,
  lambda_undisturbed = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 0,
      disturbance_prev = 0)}),
  lambda_current_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 1,
      disturbance_prev = 0)}),
  lambda_previous_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 0,
      disturbance_prev = 1)}),
  lambda_consecutive_fire = sapply(ipm_years, function(i) {
    lambda_ipm_year(
      year = i,
      disturbance = 1,
      disturbance_prev = 1)}),
  lambda_environmental_history = sapply(ipm_years, function(i) {
    lambda_env_hist_year(
      year = i)})
)

fig_lambda_year <- lambda_year %>%
  pivot_longer(
    cols = -year,
    names_to = 'disturbance',
    values_to = 'lambda') %>%
  mutate(
    disturbance = recode(
      disturbance,
      lambda_undisturbed = 'Undisturbed',
      lambda_current_fire = 'Current-year fire',
      lambda_previous_fire = 'Previous-year fire',
      lambda_consecutive_fire = 'Fire in both years',
      lambda_environmental_history = 'Environmental-history regime')) %>%
  ggplot(aes(year, lambda, color = disturbance)) +
  geom_hline(yintercept = 1, linetype = 'dashed') +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(
    title = 'Year-specific asymptotic lambda',
    subtitle = v_ggp_suffix,
    x = 'Year',
    y = expression(lambda),
    color = 'Disturbance')

fig_lambda_year
```

## Observed and projected population growth
```{r}
# Observed growth is calculated from the same quadrats represented in both
# consecutive years. Each quadrat is projected with its own current- and
# previous-year disturbance state.


# Quadrat-year disturbance lookup ---------------------------------------------
fire_lookup_quad <- df %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    disturbance_num = first(
      disturbance[!is.na(disturbance)],
      default = NA_real_),
    disturbance_prev_num = first(
      disturbance_prev[!is.na(disturbance_prev)],
      default = NA_real_),
    .groups = 'drop')


# Quadrat-year abundance -------------------------------------------------------
# Counts include sized plants and recruits without a usable size measurement.
df_counts_quad_year <- df %>%
  filter(
    !is.na(site),
    !is.na(quad),
    !is.na(year)) %>%
  group_by(
    site,
    quad,
    year) %>%
  summarise(
    n_sized = sum(
      is.finite(logvol_t0),
      na.rm = TRUE),
    n_unsized_recruits = sum(
      !is.finite(logvol_t0) & recruits == 1,
      na.rm = TRUE),
    n_ipm_state = n_sized + n_unsized_recruits,
    .groups = 'drop')


# Matched consecutive quadrat transitions -------------------------------------
df_obs_pgr_quad <- df_counts_quad_year %>%
  arrange(
    site,
    quad,
    year) %>%
  group_by(
    site,
    quad) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n_ipm_state),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(
    year_gap == 1,
    n_ipm_state > 0) %>%
  rename(
    n_t0 = n_ipm_state) %>%
  left_join(
    fire_lookup_quad,
    by = c(
      'site',
      'quad',
      'year')) %>%
  filter(
    !is.na(disturbance_num),
    !is.na(disturbance_prev_num)) %>%
  mutate(
    obs_pgr = n_t1 / n_t0)


# Initial quadrat size distribution -------------------------------------------
make_initial_n_quad <- function(
    year0,
    site_i,
    quad_i,
    pars) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  breaks <- seq(
    L,
    U,
    length.out = n + 1)
  
  b <- L + c(0:n) * h
  
  y <- 0.5 *
    (b[1:n] + b[2:(n + 1)])
  
  df0 <- df %>%
    filter(
      year == year0,
      site == site_i,
      quad == quad_i)
  
  sizes <- df0 %>%
    filter(
      is.finite(logvol_t0)) %>%
    pull(logvol_t0)
  
  size_counts <- hist(
    pmin(
      pmax(sizes, L),
      U),
    breaks = breaks,
    plot = FALSE,
    include.lowest = TRUE)$counts
  
  n_density <- size_counts / h
  
  n_unsized_recruits <- df0 %>%
    summarise(
      n = sum(
        !is.finite(logvol_t0) &
          recruits == 1,
        na.rm = TRUE)) %>%
    pull(n)
  
  if (n_unsized_recruits > 0) {
    n_density <- n_density +
      n_unsized_recruits *
      re_y_dist(
        y,
        pars,
        h = h)
  }
  
  n_density
}


# Project one quadrat-year -----------------------------------------------------
project_one_quad_year <- function(
    yr,
    site_i,
    quad_i,
    disturbance_y,
    disturbance_prev_y) {
  
  pars_y <- make_ipm_pars(
    pars_mean = pars_all_mean,
    pars_year = pars_all_year,
    pars_re_year_t1 = pars_all_re_year_t1,
    year = yr,
    re_year_t1 = yr + 1)
  
  n_obs <- make_initial_n_quad(
    year0 = yr,
    site_i = site_i,
    quad_i = quad_i,
    pars = pars_y)
  
  K_i <- kernel(
    pars = pars_y,
    disturbance = disturbance_y,
    disturbance_prev = disturbance_prev_y)
  
  h <- K_i$h
  
  n_initial <- sum(n_obs) * h
  
  if (n_initial <= 0) {
    return(
      data.frame(
        year = yr,
        site = site_i,
        quad = quad_i,
        disturbance_num = disturbance_y,
        disturbance_prev_num = disturbance_prev_y,
        n_obs_model = n_initial,
        n_surv_growth_model = NA_real_,
        n_recr_model = NA_real_,
        n_proj_model = NA_real_,
        state_asym_lambda = NA_real_,
        proj_lambda = NA_real_))
  }
  
  # Existing individuals
  n_surv_growth <- K_i$T %*% n_obs
  
  # Direct quadrat-level recruitment
  n_recruits <- K_i$R
  
  # Historical one-year projection
  n_proj <- n_surv_growth + n_recruits
  
  n_surv_growth_total <- sum(n_surv_growth) * h
  n_recruits_total <- sum(n_recruits) * h
  n_projected <- sum(n_proj) * h
  
  state_asym_lambda <- max(Mod(
    eigen(
      K_i$A_aug,
      only.values = TRUE)$values))
  
  data.frame(
    year = yr,
    site = site_i,
    quad = quad_i,
    disturbance_num = disturbance_y,
    disturbance_prev_num = disturbance_prev_y,
    n_obs_model = n_initial,
    n_surv_growth_model = n_surv_growth_total,
    n_recr_model = n_recruits_total,
    n_proj_model = n_projected,
    state_asym_lambda = state_asym_lambda,
    proj_lambda = n_projected / n_initial)
}


# Project all matched quadrat transitions -------------------------------------
df_proj_quad <- bind_rows(
  lapply(
    seq_len(nrow(df_obs_pgr_quad)),
    function(i) {
      
      project_one_quad_year(
        yr = df_obs_pgr_quad$year[i],
        site_i = df_obs_pgr_quad$site[i],
        quad_i = df_obs_pgr_quad$quad[i],
        disturbance_y =
          df_obs_pgr_quad$disturbance_num[i],
        disturbance_prev_y =
          df_obs_pgr_quad$disturbance_prev_num[i])
    }))


# Quadrat-level observed and modeled growth -----------------------------------
df_compare_quad <- df_obs_pgr_quad %>%
  left_join(
    df_proj_quad,
    by = c(
      'year',
      'site',
      'quad',
      'disturbance_num',
      'disturbance_prev_num')) %>%
  mutate(
    error_state_asymptotic_vs_obs =
      state_asym_lambda - obs_pgr,
    error_projected_vs_obs =
      proj_lambda - obs_pgr)
```

### Disturbance-state kernel used for each quadrat-year
```{r}
df_compare_quad <- df_compare_quad %>%
  mutate(
    kernel_state = case_when(
      disturbance_num == 0 & disturbance_prev_num == 0 ~ 'K_00',
      disturbance_num == 1 & disturbance_prev_num == 0 ~ 'K_10',
      disturbance_num == 0 & disturbance_prev_num == 1 ~ 'K_01',
      disturbance_num == 1 & disturbance_prev_num == 1 ~ 'K_11',
      TRUE ~ NA_character_))


df_kernel_state_use <- df_compare_quad %>%
  count(
    year,
    kernel_state,
    name = 'n_quadrat_transitions') %>%
  arrange(
    year,
    kernel_state)

df_kernel_state_use
```

### Site-year comparison from matched quadrats
```{r}
df_compare_site <- df_compare_quad %>%
  group_by(
    site,
    year) %>%
  summarise(
    n_t0 = sum(
      n_t0,
      na.rm = TRUE),
    n_t1 = sum(
      n_t1,
      na.rm = TRUE),
    n_obs_model = sum(
      n_obs_model,
      na.rm = TRUE),
    n_surv_growth_model = sum(
      n_surv_growth_model,
      na.rm = TRUE),
    n_recr_model = sum(
      n_recr_model,
      na.rm = TRUE),
    n_proj_model = sum(
      n_proj_model,
      na.rm = TRUE),
    n_quadrats = n(),
    n_burned_quadrats = sum(
      disturbance_num == 1,
      na.rm = TRUE),
    disturbance_num = as.numeric(
      any(
        disturbance_num == 1,
        na.rm = TRUE)),
    disturbance_prev_num = as.numeric(
      any(
        disturbance_prev_num == 1,
        na.rm = TRUE)),
    .groups = 'drop') %>%
  left_join(
    lambda_year %>%
      dplyr::select(
        year,
        lambda_environmental_history),
    by = 'year') %>%
  rename(
    asym_lambda = lambda_environmental_history) %>%
  mutate(
    obs_pgr = n_t1 / n_t0,
    proj_lambda = n_proj_model / n_obs_model,
    error_asymptotic_vs_obs =
      asym_lambda - obs_pgr,
    error_projected_vs_obs =
      proj_lambda - obs_pgr)

df_compare_site %>%
  print(
    n = 100,
    width = Inf)
```

### Whole-population annual comparison
```{r}
# Annual totals contain only quadrats represented in both t and t + 1.
df_compare <- df_compare_quad %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(
      n_t0,
      na.rm = TRUE),
    n_t1 = sum(
      n_t1,
      na.rm = TRUE),
    n_obs_model = sum(
      n_obs_model,
      na.rm = TRUE),
    n_surv_growth_model = sum(
      n_surv_growth_model,
      na.rm = TRUE),
    n_recr_model = sum(
      n_recr_model,
      na.rm = TRUE),
    n_proj_model = sum(
      n_proj_model,
      na.rm = TRUE),
    n_quadrats = n(),
    n_sites = n_distinct(site),
    n_burned_quadrats = sum(
      disturbance_num == 1,
      na.rm = TRUE),
    p_burned_quadrats = mean(
      disturbance_num == 1,
      na.rm = TRUE),
    disturbance = if_else(
      any(
        disturbance_num == 1,
        na.rm = TRUE),
      'Fire',
      'No fire'),
    .groups = 'drop') %>%
  left_join(
    lambda_year %>%
      dplyr::select(
        year,
        lambda_environmental_history),
    by = 'year') %>%
  rename(
    asym_lambda = lambda_environmental_history) %>%
  mutate(
    obs_pgr = n_t1 / n_t0,
    proj_lambda = n_proj_model / n_obs_model,
    disturbance = factor(
      disturbance,
      levels = c(
        'No fire',
        'Fire')))

# Observed vs modeled plot

df_plot <- df_compare %>%
  dplyr::select(year, obs_pgr, asym_lambda, proj_lambda, disturbance) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Environmental-history asymptotic lambda',
      proj_lambda =
        'Projected growth from observed size distributions + direct recruitment'))

fig_mod_vs_obs <- ggplot(
  df_plot,
  aes(x = lambda, y = obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = 'Observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs
```

#### Log-transformed observed vs modeled plot
```{r}
df_plot_log <- df_plot %>%
  filter(obs_pgr > 0, lambda > 0) %>%
  mutate(
    log_obs_pgr = log(obs_pgr),
    log_lambda = log(lambda))

fig_mod_vs_obs_log <- ggplot(
  df_plot_log,
  aes(x = log_lambda, y = log_obs_pgr, color = disturbance)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = paste(v_ggp_suffix, '- log-transformed lambda'),
    x = expression('log modeled ' * lambda),
    y = 'log observed population growth rate',
    color = 'Fire') +
  theme_classic()

fig_mod_vs_obs_log
```

## Performance
### Summary statistics
```{r}
df_compare_summary <- df_compare %>%
  summarise(
    n_years = n(),
    arithmetic_mean_obs_pgr =
      mean(obs_pgr, na.rm = TRUE),
    geometric_mean_obs_pgr =
      exp(mean(log(obs_pgr), na.rm = TRUE)),
    arithmetic_mean_env_hist_lambda =
      mean(asym_lambda, na.rm = TRUE),
    geometric_mean_env_hist_lambda =
      exp(mean(log(asym_lambda), na.rm = TRUE)),
    arithmetic_mean_proj_lambda =
      mean(proj_lambda, na.rm = TRUE),
    geometric_mean_proj_lambda =
      exp(mean(log(proj_lambda), na.rm = TRUE)),
    mean_error_env_hist_vs_obs =
      mean(
        asym_lambda - obs_pgr,
        na.rm = TRUE),
    mean_error_projected_vs_obs =
      mean(
        proj_lambda - obs_pgr,
        na.rm = TRUE),
    percent_bias_env_hist_vs_obs =
      100 * sum(
        asym_lambda - obs_pgr,
        na.rm = TRUE) /
      sum(
        obs_pgr,
        na.rm = TRUE),
    percent_bias_projected_vs_obs =
      100 * sum(
        proj_lambda - obs_pgr,
        na.rm = TRUE) /
      sum(
        obs_pgr,
        na.rm = TRUE),
    rmse_env_hist_vs_obs =
      sqrt(
        mean(
          (asym_lambda - obs_pgr)^2,
          na.rm = TRUE)),
    rmse_projected_vs_obs =
      sqrt(
        mean(
          (proj_lambda - obs_pgr)^2,
          na.rm = TRUE))) %>%
  pivot_longer(
    cols = everything(),
    names_to = 'statistic',
    values_to = 'value')

df_compare_summary

df_compare %>% print(n = 100)
```

### Historical projection components
```{r}
df_projection_components <- df_compare %>%
  transmute(
    year,
    observed_t0 = n_t0,
    observed_t1 = n_t1,
    predicted_surv_growth = n_surv_growth_model,
    predicted_recruits = n_recr_model,
    predicted_t1 = n_proj_model,
    observed_growth = obs_pgr,
    projected_growth = proj_lambda)

df_projection_components %>%
  print(n = 100)
```

### Recruitment and survival-growth diagnostic
```{r}
df_recruits_quad_year <- df %>%
  group_by(site, quad, year) %>%
  summarise(
    obs_recruits = sum(recruits == 1, na.rm = TRUE),
    .groups = 'drop')

df_components_quad <- df_obs_pgr_quad %>%
  left_join(
    df_recruits_quad_year %>%
      rename(year_t1 = year),
    by = c('site', 'quad', 'year_t1')) %>%
  mutate(
    obs_recruits = replace_na(obs_recruits, 0),
    obs_nonrecruits = pmax(n_t1 - obs_recruits, 0)) %>%
  left_join(
    df_proj_quad %>%
      dplyr::select(
        year,
        site,
        quad,
        disturbance_num,
        disturbance_prev_num,
        n_surv_growth_model,
        n_recr_model),
    by = c(
      'year',
      'site',
      'quad',
      'disturbance_num',
      'disturbance_prev_num'))

df_components_year <- df_components_quad %>%
  group_by(year) %>%
  summarise(
    obs_t1 = sum(n_t1),
    obs_recruits = sum(obs_recruits),
    pred_recruits = sum(n_recr_model),
    obs_nonrecruits = sum(obs_nonrecruits),
    pred_surv_growth = sum(n_surv_growth_model),
    recruit_error = pred_recruits - obs_recruits,
    surv_growth_error = pred_surv_growth - obs_nonrecruits,
    .groups = 'drop')

df_components_year %>%
  print(n = 100)

df_components_year %>%
  summarise(
    RMSE_recruitment =
      sqrt(mean(recruit_error^2)),
    RMSE_surv_growth =
      sqrt(mean(surv_growth_error^2)),
    MAE_recruitment =
      mean(abs(recruit_error)),
    MAE_surv_growth =
      mean(abs(surv_growth_error)))
```

### Site-level summary
```{r}
df_compare_site_summary <- df_compare_site %>%
  filter(
    !is.na(obs_pgr),
    !is.na(asym_lambda),
    !is.na(proj_lambda),
    obs_pgr > 0,
    asym_lambda > 0,
    proj_lambda > 0) %>%
  group_by(site) %>%
  summarise(
    n_year_transitions = n(),
    lambda_obs_geometric =
      exp(mean(log(obs_pgr), na.rm = TRUE)),
    lambda_obs_arithmetic =
      mean(obs_pgr, na.rm = TRUE),
    lambda_env_hist_geometric =
      exp(mean(log(asym_lambda), na.rm = TRUE)),
    lambda_env_hist_arithmetic =
      mean(asym_lambda, na.rm = TRUE),
    lambda_projected_geometric =
      exp(mean(log(proj_lambda), na.rm = TRUE)),
    lambda_projected_arithmetic =
      mean(proj_lambda, na.rm = TRUE),
    error_env_hist_geo_vs_obs_geo =
      lambda_env_hist_geometric -
      lambda_obs_geometric,
    error_projected_geo_vs_obs_geo =
      lambda_projected_geometric -
      lambda_obs_geometric,
    rmse_env_hist_vs_obs =
      sqrt(mean(
        (asym_lambda - obs_pgr)^2,
        na.rm = TRUE)),
    rmse_projected_vs_obs =
      sqrt(mean(
        (proj_lambda - obs_pgr)^2,
        na.rm = TRUE)),
    mean_n_initial =
      mean(n_obs_model, na.rm = TRUE),
    mean_disturbance =
      mean(disturbance_num, na.rm = TRUE),
    .groups = 'drop') %>%
  arrange(as.numeric(as.character(site)))

df_compare_site_summary %>%
  print(
    n = 100,
    width = Inf)
```

### Yearly lambda comparison diagnostic
```{r}
df_lambda_year_check <- lambda_year %>%
  mutate(
    min_state_lambda = pmin(
      lambda_undisturbed,
      lambda_current_fire,
      lambda_previous_fire,
      lambda_consecutive_fire),
    max_state_lambda = pmax(
      lambda_undisturbed,
      lambda_current_fire,
      lambda_previous_fire,
      lambda_consecutive_fire),
    env_hist_minus_undisturbed =
      lambda_environmental_history -
      lambda_undisturbed) %>%
  dplyr::select(
    year,
    lambda_undisturbed,
    lambda_current_fire,
    lambda_previous_fire,
    lambda_consecutive_fire,
    lambda_environmental_history,
    min_state_lambda,
    max_state_lambda,
    env_hist_minus_undisturbed)

df_lambda_year_check %>%
  as_tibble() %>%
  print(
    n = 100,
    width = Inf)
```

```{r}
# # Save key outputs -------------------------------------------------------------
# # Uncomment if needed.
# # saveRDS(pars_all_mean,
# #         file.path(dir_result, 'pole_year_fire_pars_mean.rds'))
# # saveRDS(pars_all_year,
# #         file.path(dir_result, 'pole_year_fire_pars_year.rds'))
# # saveRDS(df_compare,
# #         file.path(dir_result, 'pole_year_fire_lambda_compare.rds'))
# # saveRDS(df_compare_site,
# #         file.path(dir_result, 'pole_year_fire_lambda_compare_site.rds'))
```
