# Data {#data}

## Quadrat data 
We focus solely on the development and decision-making related to the quadrat data, as it is the only aspect that deviates from the approach used by Adler et al.

### Set-up {-}
#### Packages {-}
We used the following packages: 
`tidyverse`, `plantTracker`, `sf`, `janitor`

<details>
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(plantTracker) #ver 1.1.0
library(sf) #ver 1.0-1.2
library(janitor)
```
</details>

#### Key variables {-}
Once again, we worked with the Moore et al. (2021) dataset, which was collected in Arizona, USA.
```{r key variables}

# Define publication 
v_author_year <- c('moore_2021')
# Define region abbreviation
v_region_abb  <- c('az')
```


#### Directories {-}
```{r dir_main1, message=FALSE, warning=FALSE, include=FALSE}
dir_main <- file.path('C:', 'code', 'RUPDemo_IPMs')
```

```{r dir_main2, eval=FALSE, message=FALSE, warning=FALSE}
dir_main <- file.path('path', 'to', 'RUPDemo_IPMs')
```

```{r directories}

dir_publ        <- file.path(dir_main, paste0(v_author_year, '_', v_region_abb))
dir_data        <- file.path(dir_publ, 'data')
dir_data_ancill <- file.path(dir_data, 'Ancillary_Data_CSVs')
dir_data_quad   <- file.path(dir_data, 'quadrat_data')
dir_shp         <- file.path(dir_publ, 'data', 'Species_Shapefile_Extractions')
dir_geo         <- file.path(dir_data, 'Quadrat_Spatial_Data', 'Combined_by_Site.gdb')
```


### Inventory quadrats and years
We built the inventory by going through the data folder to extract the relevant demographic information.

```{r data quads and years, message=FALSE, warning=FALSE}
# Listing the shapefiles in the shapefile directory
shapefiles <- list.files(
  dir_shp, pattern = '\\.shp$', recursive = TRUE, full.names = TRUE)

# Function to extract year and quadrat information from file names
extract_info <- function(file_path) {
  # Extract the year (assuming it's a 4-digit number in the filename)
  year <- sub('.*_(\\d{4})\\.shp$', '\\1', basename(file_path))
  
  # Extract quadrat identifier (assuming it's before the year in the filename)
  quadrat <- sub('^(.*)_(\\d{4})\\.shp$', '\\1', basename(file_path))
  
  return(c(quadrat = quadrat, year = year))  # Return as a vector
}

# Apply the function to all shapefiles and store results
file_info <- lapply(shapefiles, extract_info)

# Convert to a data frame for easier manipulation
file_info_df <- do.call(rbind, file_info)
colnames(file_info_df) <- c('quadrat', 'year')
file_info_df <- as.data.frame(file_info_df)

file_info_df <- file_info_df %>% 
  mutate(year = str_sub(as.factor(year))) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(quadrat = gsub('Quadrat_', '', quadrat),
         quadrat = gsub('_bar_', ' / ', quadrat))

unique_quadrats_by_plot <- bind_rows(
  tibble(
    quadrat = 'year',  # New quadrat
    year = list(unique(file_info_df$year))),
  file_info_df %>%
  group_by(quadrat) %>%
  summarise(year = list(unique(year))) %>%
  arrange(quadrat)
  )

unique_year_list <- setNames(
  unique_quadrats_by_plot$year, unique_quadrats_by_plot$quadrat)
inv_sgs <- unique_year_list

str(inv_sgs[1:10])
```

## Geo-data
We obtained the geographic data directly from the data repository, where it is divided into cover data and density data. These served as the raw inputs for our analysis.
```{r data geo, message=FALSE, warning=FALSE}
cover_all <- sf::st_read( 
  dsn = dir_geo, layer = 'Cover_All') %>% 
  rename(geometry = Shape) %>% 
  clean_names()

density_all <- sf::st_read(
  dsn = dir_geo, layer = 'Density_All' ) %>% 
  rename(geometry = Shape) %>% 
  clean_names()
```
### Centroids
Because some of the geographic data points represent very small areas, we also reviewed the associated .shp files to verify their accuracy—and confirmed they matched the original data. Notably, approximately 5% of individuals are expected to have an area of 0.00003 (I think this should be meters2)

```{r too samll}
hist(log(cover_all$shape_area))
quantile(cover_all$shape_area, 0.05)
```
#### Centroids {-}

```{r centroids, message=FALSE, warning=FALSE}
# Centroids for Cover data
# Subset the data where Shape_Area < 2.5e-6 and Shape_Area > 0
subset_to_update <- cover_all[
  cover_all$shape_area < 2.5e-6 & cover_all$shape_area > 0, ]

# Apply centroid and buffer transformations to the subset
subset_to_update <- st_buffer(st_centroid(subset_to_update), dist = 0.001)

# Remove the updated subset from the original data
cover_all_remaining <- cover_all[
  cover_all$shape_area >= 2.5e-6 | cover_all$shape_area == 0, ]

# Combine the updated subset with the remaining original data
cover_all_updated <- rbind(cover_all_remaining, subset_to_update)


# Centroids for density data
density_all1 <- density_all %>% 
  mutate(area = NA,
         n_flower = NA) %>%
  rename(x = coords_x1,
         y = coords_x2)

density_all2 <- st_centroid(density_all1) %>%  
#  st_buffer(dist = 0.0003120444) # what we decided with aspen
  st_buffer(dist = 0.001) # this gives us the lowest in the graphs 
 # which is 3.14*10^-6 m2 -- log --> ~ -13
```

#### Working data {-}
By combining the density and cover data, we created our working dataset, which has the following structure:
```{r df, message=FALSE, warning=FALSE}
# Bind density and cover
df0 <- rbind(density_all2, cover_all_updated) %>% 
  select(-c('area')) %>% 
  mutate(year = as.numeric(year))

head(df0)
```

## Species list
```{r data species, message=FALSE, warning=FALSE}
df_species <- st_drop_geometry(df0) %>%
  group_by(species, type) %>% 
  summarise(
    quads  = length(unique(quadrat)),
    years  = length(unique(year)),
    counts = n()) %>%
  arrange(desc(counts))

head(df_species)
```

## Data preparation
Using the inventory and the working data, we first checked the data for bugs and then performed the necessary cleaning with the `checkDat` funciton.
```{r eval=FALSE, message=FALSE, warning=FALSE}
checkDat(
  dat = df0, inv = inv_sgs, species = 'species', site = 'site', 
  quad = 'quadrat', year = 'year', geometry = 'geometry')
```

