# Investigations {#investigation}

We will now investigate several key aspects of the data to maximize its utility, given the constraints. Specifically, we aim to address the following questions:

1. How does the size of recruits develop over time?
2. What is the size trajectory following the exit from dormancy?
3. Is it possible to distinguish between true recruits and individuals exiting dormancy when their first appearance coincides with the beginning of the observation period?
4. Fires
5. Spatial relationships between sites or plots

## Issue 1. - Size of recruits

To understand how the plants develop over time, we first examine the size at their first appearance. We can confidently identify this as their first appearance only if there is a record of three consecutive years with no presence prior to that year.

```{r i1 plots, fig.width=10, fig.height=4}
i1_hist <- ggplot(df_mean %>% filter(recruit == 1), aes(x = size_t0)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black") +
  theme_minimal()

df_mean_r <- df_mean %>%
  filter(plant_id %in% df_mean$plant_id[df_mean$recruit == 1]) %>% 
  # 1_4_17 has a problem
  filter(plant_id != '1_4_17')

mod_i1_glm_size_age_quad <- lmer(logsize_t0 ~ age + I(age^2) + (1| plant_id), data = df_mean_r)
df_mean_r$pred_glm_size_age_quad <- predict(mod_i1_glm_size_age_quad, newdata = df_mean_r, re.form = NULL)

mod_i1_lm_size_age_quad  <- lm  (logsize_t0 ~ age + I(age^2), data = df_mean_r)
df_mean_r$pred_lm_size_age_quad <- predict(mod_i1_lm_size_age_quad, newdata = df_mean_r, re.form = NULL)


i1_size_age <- ggplot(df_mean_r, aes(x = age, y = logsize_t0)) +
  geom_jitter(aes(color = plant_id), width = 0.1, height = 0.1, alpha = 0.6) +  
  geom_line(aes(y = pred_glm_size_age_quad   , color = plant_id), alpha = 0.5) +  
  geom_line(aes(y = pred_lm_size_age_quad), alpha = 1, color = 'black', size = 0.8) +  
  theme_minimal() + 
  labs(x = "Age", y = "Log Size", title = "Log Size vs Age of Recruits") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_x_continuous(
    limits = c(0, 12),
    breaks = seq(0, 12, by = 1)) +
  theme(legend.position = "none")


# Combine the plots
grid.arrange(i1_hist, i1_size_age, 
             ncol = 2)
```


## Issue 2 - Size after Dormancy
This plot shows the distribution of size_t1 for dormant individuals. Most certainly, there are very large individuals that go into dormancy  Unfortunately we don't find any clear pattern that sets the dormant individuals apart form recruits. 

```{r i2 plots, fig.width=10, fig.height=4}
i2_hist <- ggplot(df_mean %>% filter(dormancy == 1), aes(x = size_t1)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black") +
  labs(title = "Histogram of Size After Dormancy", x = "Size", y = "Frequency") +
  theme_minimal() +
  annotate("text", x = Inf, y = Inf, label = paste("n:", nrow(df_mean %>% 
                                                                  filter(dormancy == 1) %>% 
                                                                  filter(!is.na(size_t1)))), 
           hjust = 2, vjust = 2, size = 4, color = "black") +
  xlim(0, 41) + 
  ylim(0, 41) 

# Create a vector to store the size values before dormancy == 1
size_before_dormancy <- c()

# Loop through the rows and capture the size value before dormancy == 1
for(i in 2:nrow(df_mean)) {
  if(!is.na(df_mean$dormancy[i]) && df_mean$dormancy[i] == 1) {
    size_before_dormancy <- c(size_before_dormancy, df_mean$size_t0[i - 1])
  }
}

# Convert to a data frame for ggplot
df_size_before_dormancy <- data.frame(size_t0 = size_before_dormancy)

# Create ggplot histogram
i2_hist_1 <- ggplot(df_size_before_dormancy, aes(x = size_t0)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black") +
  labs(title = "Histogram of Size Before Dormancy", x = "Size", y = "Frequency") +
  theme_minimal() +
  annotate("text", x = Inf, y = Inf, label = paste("n:", sum(!is.na(size_before_dormancy))), 
           hjust = 2, vjust = 2, size = 4, color = "black") +
  xlim(0, 41) + 
  ylim(0, 41) 

# Combine the plots
grid.arrange(i2_hist_1, i2_hist,  
             ncol = 2)
```

<details>
<summary>Expand to see the mean dataframe of the very large individuals after dormancy</summary>
```{r MDFLaAfDo}

print(df_mean %>% 
  filter(plant_id ==  df_mean %>% 
  filter(dormancy == 1, size_t1 > 15) %>% 
  select(plant_id) %>% 
  .[[1]]) %>% 
    select(plant_id, size_t0), n = 100)

```



<details>
<summary>Expand to see the full dataframe of the very large individuals after dormancy</summary>
```{r FDFLaAfDo}

print(df %>% 
  filter(plant_id ==  df_mean %>% 
  filter(dormancy == 1, size_t1 > 15) %>% 
  select(plant_id) %>% 
  .[[1]]) %>% 
    select(plant_id, br), n = 100)

```
</details>


## Issue 3 - Flagged

To facilitate direct comparison of observations, we create a new data frame that flags entries based on two criteria. First, individuals with missing size_t0 values for the years 1999, 2000, or 2001 are flagged. Second, individuals with non-missing size_t1 values for any of these years (1999, 2000, or 2001) are also flagged. After applying these flags, we assign an age to each individual.


```{r i3 data}
df_mean_f <- df_mean %>%
  group_by(site, quad, plant, plant_id) %>%
  mutate(
    flag_1 = if_else(
      any(year == 1999 & is.na(size_t0)) |
        (any(year %in% c(1999, 2000)) & all(is.na(size_t0[year %in% c(1999, 2000)]))) |
        (any(year %in% c(1999, 2000, 2001)) & all(is.na(size_t0[year %in% c(1999, 2000, 2001)]))),
      1, 0
    ),
    flag_2 = if_else(any(year %in% c(1999, 2000, 2001) & !is.na(size_t1)), 1, 0)
  ) %>%
  ungroup() %>% 
  filter(flag_1 == 1 & flag_2 == 1) 

# Initialize the age column
df_mean_f$age <- NA

# Loop through each unique combination of site, quad, and plant
for (i in 1:nrow(df_mean_f)) {
  
  current_site <- df_mean_f$site[i]
  current_quad <- df_mean_f$quad[i]
  current_plant <- df_mean_f$plant[i]
  
  # Subset the data for the current combination of site, quad, and plant
  group_data <- df_mean_f %>%
    filter(site == current_site, quad == current_quad, plant == current_plant) %>%
    arrange(year)  # Ensure the rows are ordered by year
  
  # Initialize the current age starting at NA
  current_age <- 1  # We will start at 1 when the first non-NA size_t0 is found
  
  # Loop through the rows within this group
  for (j in 1:nrow(group_data)) {
    if (!is.na(group_data$size_t0[j])) {
      # If the size_t0 is not NA, assign the current age
      df_mean_f$age[df_mean_f$site == current_site & 
                      df_mean_f$quad == current_quad & 
                      df_mean_f$plant == current_plant & 
                      df_mean_f$year == group_data$year[j]] <- current_age
      # Increment the age for the next year
      current_age <- current_age + 1
    } else if (j > 1) {
      # If size_t0 is NA, carry the age from the previous row
      previous_year_age <- df_mean_f$age[df_mean_f$site == current_site &
                                           df_mean_f$quad == current_quad &
                                           df_mean_f$plant == current_plant &
                                           df_mean_f$year == group_data$year[j - 1]]
      df_mean_f$age[df_mean_f$site == current_site & 
                      df_mean_f$quad == current_quad & 
                      df_mean_f$plant == current_plant & 
                      df_mean_f$year == group_data$year[j]] <- previous_year_age
    }
  }
}
```


The following plots display key patterns in the data: 1. Histogram of Size at t0 showing the distribution of size_t0 for plants at age 1, with bins of width 1. 2. Scatter Plot of Log Size at t0 vs Age visualizing the relationship between age and log size.

```{r i3 plots1, fig.width=10, fig.height=4}
i3_hist <- ggplot(df_mean_f %>% filter(age == 1), aes(x = size_t0)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black") +
  theme_minimal()


i3_size_age <- ggplot() +
  geom_jitter(
    data = df_mean_f, 
    aes(x = age, y = logsize_t0)) +
  theme_minimal()


# Combine the plots
grid.arrange(i3_hist, i3_size_age, 
             ncol = 2)
```


### Modeling Growth of the Flags

In this section, we fit four different linear mixed models to examine the relationship between age and log size at time zero (logsize_t0), incorporating random effects to account for variability among individual plant IDs.

1. Linear Model: The first model uses a linear relationship between age and log size, with a random intercept for plant-id.
2. Quadratic Model: The second model introduces a quadratic term for log size to capture potential non-linear effects of age on log size
3. Linear Model with Age: The third model reverses the dependent and independent variables, modeling log size as a function of age, with a random intercept for plant-id.
4. Quadratic Model with Age: The fourth model extends the third model by including a quadratic term for age.

```{r i3 models, echo=TRUE, code_folding="none", results='markup'}
# Fit the first model: logsize_t0 ~ age + (1 | plant_id) (linear model)
mod_flag_linear <- lmer(age ~ logsize_t0 + (1 | plant_id), data = df_mean_f)
#summary(mod_flag_linear)

# Predict values using the linear model
df_mean_f$predicted_linear <- predict(mod_flag_linear, newdata = df_mean_f, re.form = NULL)

# Fit the second model: logsize_t0 ~ age + I(age^2) + (1 | plant_id) (quadratic model)
mod_flag_quad <- lmer(age ~ logsize_t0 + I(logsize_t0^2) + (1 | plant_id), data = df_mean_f)
#summary(mod_flag_quad)

# Predict values using the quadratic model
df_mean_f$predicted_quad <- predict(mod_flag_quad, newdata = df_mean_f, re.form = NULL)

# Fit the third model: logsize_t0 ~ age + (1 | plant_id) (reversed model)
mod_flag_logsize_age <- lmer(logsize_t0 ~ age + (1 | plant_id), data = df_mean_f)
#summary(mod_flag_logsize_age)

# Predict values using the logsize_t0 ~ age model
df_mean_f$predicted_logsize_age <- predict(mod_flag_logsize_age, newdata = df_mean_f, re.form = NULL)

# Fit the fourth model: logsize_t0 ~ age + I(age^2) + (1 | plant_id) (quadratic term for age)
mod_flag_quad_age <- lmer(logsize_t0 ~ age + I(age^2) + (1 | plant_id), data = df_mean_f)
#summary(mod_flag_quad_age)

# Predict values using the logsize_t0 ~ age + I(age^2) model
df_mean_f$predicted_quad_age <- predict(mod_flag_quad_age, newdata = df_mean_f, re.form = NULL)



lm_flag_quad_age <- lm(logsize_t0 ~ age + I(age^2), data = df_mean_f)
df_mean_f$predicted_lm_quad_age <- predict(lm_flag_quad_age, newdata = df_mean_f, re.form = NULL)



```


The following plots visualize the relationships captured by the second, third, and fourth models

```{r i3 plots2, fig.width=10, fig.height=8}
# Plot the second model (quadratic relationship for logsize_t0 ~ logsize_t0^2)
p_i3_mod_1 <- ggplot(df_mean_f, aes(x = logsize_t0, y = age)) +
  geom_jitter(aes(color = plant_id), width = 0.1, height = 0.1, alpha = 0.6) +  
  geom_line(aes(y = predicted_quad, color = plant_id), alpha = 0.5) +  
  theme_minimal() +
  labs(title = "Quadratic Model with Random Intercepts: logsize_t0 vs Age",
       y = "Age",
       x = "Log Size at t0") +
  theme(legend.position = "none")

# Plot the third model (logsize_t0 ~ age relationship)
p_i3_mod_2 <- ggplot(df_mean_f, aes(x = age, y = logsize_t0)) +
  geom_jitter(aes(color = plant_id), width = 0.1, height = 0.1, alpha = 0.6) +  
  geom_line(aes(y = predicted_logsize_age, color = plant_id), alpha = 0.5) +  
  theme_minimal() +
  labs(title = "Log Size Model with Random Intercepts: Age vs Log Size at t0",
       y = "Log Size at t0",
       x = "Age") +
  theme(legend.position = "none")

# Plot the fourth model (logsize_t0 ~ age + I(age^2) relationship)
p_i3_mod_3 <- ggplot(df_mean_f, aes(x = age, y = logsize_t0)) +
  geom_jitter(aes(color = plant_id), width = 0.1, height = 0.1, alpha = 0.6) +  
  geom_line(aes(y = predicted_quad_age   , color = plant_id), alpha = 0.5) +  
  geom_line(aes(y = predicted_lm_quad_age), alpha = 1, color = 'black', size = 0.8) + 
  theme_minimal() +
  labs(title = "Quadratic Model with Random Intercepts: Age vs Log Size at t0\nand quadratic LM",
       y = "Log Size at t0",
       x = "Age") +
  theme(legend.position = "none")

grid.arrange(p_i3_mod_1, p_i3_mod_2,
             p_i3_mod_3,
             ncol = 2)
```

## Issue 4 - Fires
The data from Carter Creek Tract, Lake Wales Ridge Wildlife Environmental Area, 1998-2017, includes information on the severity of six fire outbreaks during this time. In the data we found the severity of ´0´ for some of the individuals. We raise the question of what a severity of ´0´ means? 

Once again, as of March 2025, we have not been in contact with the corresponding author. 

Menges et al. (2016) describe how fire affects Avon Park Harebells, emphasizing the plant’s ability to resprout and thrive after fire. They hypothesize that the plant’s long taproot helps them readily resprout post-fire, supported by their observation of high survival and growth after fire and mechanical disturbances (Menges and Kohfeldt 1995). Fire is crucial for maintaining the biodiversity of Florida scrub and sandhill ecosystems, which rely on periodic fires at intervals of 5 to 60 years (Menges 2007, Menges and Gordon 2010).

Menges suggests that protected sites, which are managed with fire, have more positive population trends compared to unprotected sites that do not use fire management (Menges 1999). For example, the Rosemary Path population underwent a prescribed burn in 2005, and fire did not negatively affect initial **survival**. In fact, survival rates were higher in burned quadrats in 2006-2007 (95.6%) compared to control quadrats (71.4%), and continued to be better in the treated areas in the following years.

**Growth** was influenced by fire and mechanical treatments, with the combination of Gyro-Trac and fire producing higher initial growth than fire alone or the control. 

Fire also helped with **recruitment**, increasing germination speed, although herbivory had a negative effect. Avon Park Harebells are strong resprouters after fire, herbivory, and mechanical disturbances that remove their aboveground parts. Menges recommends management practices that encourage pollinators, such as prescribed fire.

Unfortunately, neither Menges' publication nor the data webpage provide information on how fire severity was measured, leaving the level 0 severity fire a mystery, for now.


```{r fire data 1}
df_fire1 <- df_mean %>% 
  group_by(quad_id, year) %>% 
  summarise(fire_sev = mean(fire_sev, na.rm = T))
```

```{r fire plot 1}
ggplot(df_fire1, aes(x = quad_id, y = year, size = fire_sev, color = fire_sev)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(1, 10), name = "Severity") +
  scale_color_continuous(name = "") +
  labs(x = "Quadrat ID", y = "Year", title = "Fire Severity by Year and Plot") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r fire on growth}
df_fire_grow <- df_mean %>% 
  subset(size_t0 != 0) %>%
  subset(size_t1 != 0) %>% 
  select(plant_id, year, size_t0, size_t1, age,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire_event)

ggplot(
  data  = df_fire_grow, aes(x = logsize_t0, y = logsize_t1)) +
  geom_point(alpha = 0.5, pch = 16, size = 0.7, color = 'red') +
  geom_abline(aes(slope = 1, intercept = 0)) + 
  geom_smooth(method = 'lm') +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Fire on Growth',
       subtitle = v_ggp_suffix,
       x        = expression('log(size) ' [t0]),
       y        = expression('log(size)  '[t1])) +
  theme(plot.subtitle = element_text(size = 8)) +
  facet_wrap(~ fire_event)
```

```{r fire on survival}
df_fire_surv_0 <- df_mean %>% 
  filter(!is.na(survives)) %>%
  filter(size_t0 != 0) %>%
  select(plant_id, year, size_t0, survives, size_t1, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire_event) %>% 
  filter(fire_event == 0)

df_fire_surv_1 <- df_mean %>% 
  filter(!is.na(survives)) %>%
  filter(size_t0 != 0) %>%
  select(plant_id, year, size_t0, survives, size_t1, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire_event) %>% 
  filter(fire_event == 1)

p_fire_suv_0 <- ggplot(
  data = plot_binned_prop(df_fire_surv_0, 10, logsize_t0, survives)) +
  geom_point(aes(x = logsize_t0,
                 y = survives),
             alpha = 1, pch = 16, color = 'red' ) +
  geom_errorbar(aes(x = logsize_t0, ymin = lwr, ymax = upr),
                size = 0.5, width = 0.5) +
  scale_y_continuous(breaks = c(0.1, 0.5, 0.9)) +
  ylim(0, 1.01) +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Fire on Survival',
       subtitle = v_ggp_suffix,
       x        = expression('log(size)'[t0]),
       y        = expression('Survival to time t1')) +
  theme(plot.subtitle = element_text(size = 8))

p_fire_suv_1 <- ggplot(
  data = plot_binned_prop(df_fire_surv_1, 10, logsize_t0, survives)) +
  geom_point(aes(x = logsize_t0,
                 y = survives),
             alpha = 1, pch = 16, color = 'red' ) +
  geom_errorbar(aes(x = logsize_t0, ymin = lwr, ymax = upr),
                size = 0.5, width = 0.5) +
  scale_y_continuous(breaks = c(0.1, 0.5, 0.9)) +
  ylim(0, 1.01) +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Fire on Survival',
       subtitle = v_ggp_suffix,
       x        = expression('log(size)'[t0]),
       y        = expression('Survival to time t1')) +
  theme(plot.subtitle = element_text(size = 8))

p_fire_suv <- p_fire_suv_0 + p_fire_suv_1 + plot_layout()
p_fire_suv
```


```{r fire on recruits}
df_quad <- df_mean %>%  
  group_by (year, site, quad) %>% 
  summarise(tot_p_area = sum(size_t0, na.rm = T)) %>% 
  ungroup

df_group <- df_quad %>% 
  group_by (year) %>% 
  summarise(g_cov = mean(tot_p_area)) %>% 
  ungroup

df_cover <- left_join(df_quad, df_group) %>%
  mutate(year = year + 1) %>% 
  mutate(year = as.integer(year)) %>% 
  drop_na()

df_fire_recr <- df_mean %>%
  group_by (year, site, quad) %>% 
  summarise(nr_quad = sum(recruit, na.rm = T), 
            fire_event = max(fire_event, na.rm = T)) %>% 
  ungroup

df_fire_recr <- left_join(df_cover, df_fire_recr) %>% 
  filter(!is.na(fire_event))

ggplot(
  df_fire_recr, aes(x = tot_p_area, y = nr_quad)) + 
  geom_point(alpha = 0.5, pch = 16, size = 1, color = 'red') +  
  theme_bw() + 
  labs(title    = 'Fire on Recruitment',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),   
       y        = expression('Number of recruits '     [t1])) +
  theme(plot.subtitle = element_text(size = 8)) + 
  facet_wrap(~ fire_event)
```



## Issue 5 - Spatial relationship

Is there any spatial relationship regarding size, recruits, flowering, fruiting, ect. In the data we have two chategories of sizes, which are *site* and *plot*.
