# Models {#models}

For this IPM we will model the following processes:

1. Growth
2. Survival (binmial)
3. Recruits (negative binomial)
1. Fruiting (negative binomial)
2. Flowering (binomial)

## Data preparation
### Survival 
```{r su data}
df_surv <- df_mean %>% 
  filter(!is.na(survives)) %>%
  filter(size_t0 != 0) %>%
  select(plant_id, year, size_t0, survives, size_t1, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3)
```


```{r su plot1, fig.width=10, fig.height=4}
p_su_1.1 <- ggplot(data = df_surv) +
  geom_jitter(aes(x = logsize_t0, y = survives), 
              position = position_jitter(width = 0.1, height = 0.3)) +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Survival',
       subtitle = v_ggp_suffix,
       x        = expression('log(size) ' [t0]),
       y        = expression('Survival to time t1')) +
  theme(plot.subtitle = element_text(size = 8))

p_su_1.2 <- ggplot(
  data = plot_binned_prop(df_mean, 10, logsize_t0, survives)) +
  geom_point(aes(x = logsize_t0,
                 y = survives),
             alpha = 1, pch = 16, color = 'red' ) +
  geom_errorbar(aes(x = logsize_t0, ymin = lwr, ymax = upr),
                size = 0.5, width = 0.5) +
  scale_y_continuous(breaks = c(0.1, 0.5, 0.9)) +
  ylim(0, 1.01) +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Survival',
       subtitle = v_ggp_suffix,
       x        = expression('log(size)'[t0]),
       y        = expression('Survival to time t1')) +
  theme(plot.subtitle = element_text(size = 8))

grid.arrange(p_su_1.1, p_su_1.2,  
             ncol = 2)
```


### Growth
```{r gr data}
df_grow <- df_mean %>% 
  subset(size_t0 != 0) %>%
  subset(size_t1 != 0) %>% 
  select(plant_id, year, size_t0, size_t1, age,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3)
```

Since we have maximum count of branches in a year on an individual as growth, therefore count (integer) data, we find a binned structure of the logsizes. This effect is more pronounced in the on the lower end.  

```{r gr plot1}
ggplot(
  data  = df_grow, aes(x = logsize_t0, y = logsize_t1)) +
  geom_point(alpha = 0.5, pch = 16, size = 0.7, color = 'red') +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Growth',
       subtitle = v_ggp_suffix,
       x        = expression('log(size) ' [t0]),
       y        = expression('log(size)  '[t1])) +
  theme(plot.subtitle = element_text(size = 8))
```


### Flowering

We model flowering in a binomial fashion, thus we reduce the counts of flowers only to if the individual has flowered or not.

```{r fl data}
df_flower <- df_mean %>% 
  filter(!is.na(flower)) %>%
  filter(size_t0 != 0) %>%
  select(plant_id, year, size_t0, flower, size_t1, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3) %>% 
  mutate(flower = if_else(flower > 0, 1, flower))
```


```{r fl plot1, fig.width=10, fig.height=4}
p_fl_1.1 <- ggplot(data = df_flower) +
  geom_jitter(aes(x = logsize_t0, y = flower), 
              position = position_jitter(width = 0.1, height = 0.3)) +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Flowering',
       subtitle = v_ggp_suffix,
       x        = expression('log(size) ' [t0]),
       y        = expression('Flowering to time t1')) +
  theme(plot.subtitle = element_text(size = 8))

p_fl_1.2 <- ggplot(
  data = plot_binned_prop(df_flower, 10, logsize_t0, flower)) +
  geom_point(aes(x = logsize_t0,
                 y = flower),
             alpha = 1, pch = 16, color = 'red' ) +
  geom_errorbar(aes(x = logsize_t0, ymin = lwr, ymax = upr),
                size = 0.5, width = 0.5) +
  scale_y_continuous(breaks = c(0.1, 0.5, 0.9)) +
  ylim(0, 1.01) +
  theme_bw() +
  theme(axis.text = element_text(size = 8),
        title     = element_text(size = 10)) +
  labs(title    = 'Flowering',
       subtitle = v_ggp_suffix,
       x        = expression('log(size)'[t0]),
       y        = expression('Flowering to time t1')) +
  theme(plot.subtitle = element_text(size = 8))

grid.arrange(p_fl_1.1, p_fl_1.2, 
             ncol = 2)
```


### Fruting
```{r fr data}
df_quad <- df_mean %>%  
  group_by (year, site, quad) %>% 
  summarise(tot_p_area = sum(size_t0, na.rm = T)) %>% 
  ungroup

df_group <- df_quad %>% 
  group_by (year) %>% 
  summarise(g_cov = mean(tot_p_area)) %>% 
  ungroup

df_cover <- left_join(df_quad, df_group) %>%
  mutate(year = year + 1) %>% 
  mutate(year = as.integer(year)) %>% 
  drop_na()

df_fruit <- df_mean %>%
  group_by (year, site, quad) %>% 
  summarise(nr_quad = sum(fruit, na.rm = T)) %>% 
  ungroup

df_fruit <- left_join(df_cover, df_fruit)
```


```{r fr plot1, fig.width=10, fig.height=4}
p_fr_1.1 <- ggplot() +
  geom_histogram(data = df %>% filter(!is.na(fl)), 
                 aes(fl)) +
  theme_bw() + 
  labs(title    = 'Fruting',
       subtitle = v_ggp_suffix,
       x        = expression('Number of fruits'),   
       y        = expression('Count')) +
  theme(plot.subtitle = element_text(size = 8))

p_fr_1.2 <- ggplot(
  df_fruit, aes(x = tot_p_area, y = nr_quad)) + 
  geom_point(alpha = 0.5, pch = 16, size = 1, color = 'red') +  
  theme_bw() + 
  labs(title    = 'Fruitment',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),   
       y        = expression('Number of fruits '     [t1])) +
  theme(plot.subtitle = element_text(size = 8))

grid.arrange(p_fr_1.1, p_fr_1.2, 
             ncol = 2)
```


### Recruits
```{r re data}
df_quad <- df_mean %>%  
  group_by (year, site, quad) %>% 
  summarise(tot_p_area = sum(size_t0, na.rm = T)) %>% 
  ungroup

df_group <- df_quad %>% 
  group_by (year) %>% 
  summarise(g_cov = mean(tot_p_area)) %>% 
  ungroup

df_cover <- left_join(df_quad, df_group) %>%
  mutate(year = year + 1) %>% 
  mutate(year = as.integer(year)) %>% 
  drop_na()

df_recr <- df_mean %>%
  group_by (year, site, quad) %>% 
  summarise(nr_quad = sum(recruit, na.rm = T)) %>% 
  ungroup

df_recr <- left_join(df_cover, df_recr)
```


```{r re plot1}
ggplot(
  df_recr, aes(x = tot_p_area, y = nr_quad)) + 
  geom_point(alpha = 0.5, pch = 16, size = 1, color = 'red') +  
  theme_bw() + 
  labs(title    = 'Recruitment',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),   
       y        = expression('Number of recruits '     [t1])) +
  theme(plot.subtitle = element_text(size = 8))
```


## Models
### Survival
```{r su mod}
mod_su_0 <- glm(survives ~ 1,
                data = df_surv, family = 'binomial') 
# Logistic regression
mod_su_1 <- glm(survives ~ logsize_t0,
                data = df_surv, family = 'binomial') 
# Quadratic logistic model
mod_su_2 <- glm(survives ~ logsize_t0 + logsize_t0_2,
                data = df_surv, family = 'binomial')  
# Cubic logistic model
mod_su_3 <- glm(survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
                data = df_surv, family = 'binomial')  


# Compare models using AIC
mods_su      <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAIC <- AICtab(mods_su, weights = T, sort = F)$dAIC

# Get the sorted indices of dAIC values
mods_su_sorted <- order(mods_su_dAIC)

# Establish the index of model complexity
if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index       <- mod_su_index_bestfit - 1 
} else {
  mod_su_index_bestfit <- v_mod_set_su +1
  v_mod_su_index       <- v_mod_set_su
}


mod_su_bestfit   <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef         <- coef(mod_su_bestfit)

# Generate predictions for survival across a range of sizes
mod_su_x <- seq(
  min(df_surv$logsize_t0, na.rm = T),
  max(df_surv$logsize_t0, na.rm = T), length.out = 100)

# Prepare data for survival plot
df_surv_pred <- predictor_fun(mod_su_x, mod_su_ranef) %>% 
  # Inverse logit for predictions
  boot::inv.logit() %>% 
  data.frame(logsize_t0 = mod_su_x, survives = .)
```


```{r su plot2, fig.width=10, fig.height=4}
p_su_2.1 <- ggplot() +
  geom_jitter(data = df_surv, aes(x = logsize_t0, 
                                  y = survives),
              alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(data = df_surv_pred, aes(x = logsize_t0, 
                                     y = survives),
            color = line_color_pred_fun(mod_su_ranef), 
            lwd   = 2) +  
  theme_bw() + 
  labs(title    = 'Survival prediction',
       subtitle = v_ggp_suffix) +
  theme(plot.subtitle = element_text(size = 8))

p_su_2.2 <- ggplot() +
  geom_point(data =  plot_binned_prop(
    df_mean, 10, logsize_t0, survives), 
    aes(x = logsize_t0, 
        y = survives) ) +
  geom_errorbar(
    data = plot_binned_prop(df_mean, 10, logsize_t0, survives), 
    aes(x = logsize_t0, 
        ymin = lwr,
        ymax = upr) ) +
  geom_line(data = df_surv_pred, aes(x = logsize_t0, 
                                     y = survives),
            color = 'red', lwd   = 2) + 
  theme_bw() +
  ylim(0, 1)

p_su_2.3 <- p_su_2.1 + p_su_2.2 + plot_layout()
p_su_2.3

```


### Growth

```{r gr mod}
mod_gr_0 <- lm(logsize_t1 ~ 1, 
               data = df_grow)
# Linear model
mod_gr_1   <- lm(logsize_t1 ~ logsize_t0, 
                 data = df_grow)
# Quadratic model
mod_gr_2 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2, 
               data = df_grow)  
# Cubic model
mod_gr_3 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3, 
               data = df_grow)

mods_gr      <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAIC <- AICtab(mods_gr, weights = T, sort = F)$dAIC

# Get the sorted indices of dAIC values
mods_gr_sorted <- order(mods_gr_dAIC)

# Establish the index of model complexity
if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index       <- mod_gr_index_bestfit - 1 
} else {
  mod_gr_index_bestfit <- v_mod_set_gr +1
  v_mod_gr_index       <- v_mod_set_gr
}

mod_gr_bestfit         <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef           <- coef(mod_gr_bestfit)

# Predict size at time t1 using the mean growth model
df_grow$pred <- predict(mod_gr_bestfit, type = 'response')
```


```{r gr plot2, fig.width=10, fig.height=4}
p_gr_2.1 <- ggplot(
  df_grow, aes(x = logsize_t0, y = logsize_t1)) +
  # Plot observed data
  geom_point() +
  geom_function(fun = function(x) predictor_fun(x, mod_gr_ranef), 
                color = line_color_pred_fun(mod_gr_ranef), 
                lwd = 2) +
  theme_bw() + 
  labs(title    = 'Growth prediction',
       subtitle = v_ggp_suffix) +
  theme(plot.subtitle = element_text(size = 8))

p_gr_2.2 <- ggplot(
  df_grow, aes(x = pred, y = logsize_t1)) +
  geom_point() +  
  geom_abline(aes(intercept = 0, slope = 1),  
              color = 'red', lwd = 2) + 
  theme_bw()

grid.arrange(p_gr_2.1, p_gr_2.2, 
             ncol = 2)
```


```{r gr mod var}
# Fit a model to assess variance in growth
# Fitted values from growth model
mod_gr_x   <- fitted(mod_gr_bestfit)  
# Squared residuals
mod_gr_y   <- resid(mod_gr_bestfit)^2  
# Non-linear model for variance
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x), start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE) ) 
```


### Flowering
```{r fl mod}
mod_fl_0 <- glm(flower ~ 1,
                data = df_flower, family = 'binomial') 
# Logistic regression
mod_fl_1 <- glm(flower ~ logsize_t0,
                data = df_flower, family = 'binomial') 
# Quadratic logistic model
mod_fl_2 <- glm(flower ~ logsize_t0 + logsize_t0_2,
                data = df_flower, family = 'binomial')  
# Cubic logistic model
mod_fl_3 <- glm(flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
                data = df_flower, family = 'binomial')  


# Compare models using AIC
mods_fl      <- list(mod_fl_0, mod_fl_1, mod_fl_2, mod_fl_3)
mods_fl_dAIC <- AICtab(mods_fl, weights = T, sort = F)$dAIC

# Get the sorted indices of dAIC values
mods_fl_sorted <- order(mods_fl_dAIC)

# Establish the index of model complexity
if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index       <- mod_fl_index_bestfit - 1 
} else {
  mod_fl_index_bestfit <- v_mod_set_fl +1
  v_mod_fl_index       <- v_mod_set_fl
}


mod_fl_bestfit   <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef     <- coef(mod_fl_bestfit)

# Generate predictions for survival across a range of sizes
mod_fl_x <- seq(
  min(df_flower$logsize_t0, na.rm = T),
  max(df_flower$logsize_t0, na.rm = T), length.out = 100)

# Prepare data for survival plot
df_flow_pred <- predictor_fun(mod_fl_x, mod_fl_ranef) %>% 
  # Inverse logit for predictions
  boot::inv.logit() %>% 
  data.frame(logsize_t0 = mod_fl_x, flower = .)
```


```{r fl plot2}
p_fl_2.1 <- ggplot() +
  geom_jitter(data = df_flower, 
              aes(x = logsize_t0, y = flower),
              alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(data = df_flow_pred, 
            aes(x = logsize_t0, y = flower),
            color = line_color_pred_fun(mod_fl_ranef), 
            lwd   = 2) +  
  theme_bw() + 
  labs(title    = 'Survival prediction',
       subtitle = v_ggp_suffix) +
  theme(plot.subtitle = element_text(size = 8))

p_fl_2.2 <- ggplot() +
  geom_point(data =  plot_binned_prop(
    df_flower, 10, logsize_t0, flower), 
    aes(x = logsize_t0, y = flower) ) +
  geom_errorbar(
    data = plot_binned_prop(df_flower, 10, logsize_t0, flower), 
    aes(x = logsize_t0, ymin = lwr, ymax = upr) ) +
  geom_line(data = df_flow_pred, 
            aes(x = logsize_t0, y = flower),
            color = 'red', lwd   = 2) + 
  theme_bw() +
  ylim(0, 1)

p_fl_2.3 <- p_fl_2.1 + p_fl_2.2 + plot_layout()
p_fl_2.3
```


### Fruting
```{r fr mod}
df_fruit_nona_nr_quad <- df_fruit %>% filter(!is.na(nr_quad))
# Fit a negative binomial model
mod_fru <- MASS::glm.nb(nr_quad ~ 1, data = df_fruit_nona_nr_quad)

# Generate predictions
df_fruit_nona_nr_quad <- df_fruit_nona_nr_quad %>% 
  mutate(mod_pred = predict(mod_fru, type = 'response')) 

# Summarize total number of fruits and predictions
df_fruit_sums_m <- df_fruit_nona_nr_quad %>%
  summarize(nr_quad = sum(nr_quad),
            mod_pred = sum(mod_pred))

# Count number of adult individuals
indiv_m <- df_surv %>%
  summarize(n_adults = n())

# Calculate reproduction per capita (both observed and predicted)
fruit_pc_m <- indiv_m %>%
  bind_cols(df_fruit_sums_m) %>%
  mutate(fruit_pc_mean = mod_pred / n_adults) %>%
  mutate(fruit_pc_obs  = nr_quad  / n_adults) %>%
  drop_na 

fruit_pc_m
```


### Recruits
```{r re mod}
df_recr_nona_nr_quad <- df_recr %>% filter(!is.na(nr_quad))
# Fit a negative binomial model for recruitment
mod_rec <- MASS::glm.nb(nr_quad ~ 1, data = df_recr_nona_nr_quad)

# Generate predictions for recruitment
df_recr_nona_nr_quad <- df_recr_nona_nr_quad %>% 
  mutate(mod_pred = predict(mod_rec, type = 'response')) 

# Summarize total number of recruits and predictions
df_rec_sums_m <- df_recr_nona_nr_quad %>%
  summarize(nr_quad = sum(nr_quad),
            mod_pred = sum(mod_pred))

# Count number of adult individuals
indiv_m <- df_surv %>%
  summarize(n_adults = n())

# Calculate reproduction per capita (both observed and predicted)
repr_pc_m <- indiv_m %>%
  bind_cols(df_rec_sums_m) %>%
  mutate(repr_pc_mean = mod_pred / n_adults) %>%
  mutate(repr_pc_obs = nr_quad / n_adults) %>%
  drop_na 

repr_pc_m
```


## Parameter estimates

```{r}
# Survival
surv_fe  <- data.frame(coefficient = names(coef(mod_su_bestfit)),
                       value       = coef(mod_su_bestfit))

surv_out <- Reduce(function(...) rbind(...), list(surv_fe)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))


# Growth
grow_fe  <- data.frame(coefficient = names(coef(mod_gr_bestfit)),
                       value       = coef(mod_gr_bestfit))
grow_var <- data.frame(coefficient = names(coef(mod_gr_var)),
                       value       = coef(mod_gr_var))

grow_out <- Reduce(function(...) rbind(...), list(grow_fe, grow_var)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))


# Flower
flwr_fe  <- data.frame(coefficient = names(coef(mod_fl_bestfit)),
                       value       = coef(mod_fl_bestfit))

flwr_out <- Reduce(function(...) rbind(...), list(flwr_fe)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))


# Fruiting & Recruitment & Others
rec_size <- df_mean %>% subset(recruit == 1)
fru_size <- df_mean %>% subset(fruit   == 1)

others   <- data.frame(coefficient = c('rec_siz', 'rec_sd', 
                                       'fru_siz', 'fru_sd', 
                                       'max_siz', 'min_siz',
                                       'fecu_b0',
                                       'frui_b0'),
                       value       = c(mean(log(rec_size$size_t0), na.rm = T), 
                                       sd(  log(rec_size$size_t0), na.rm = T),
                                       mean(log(fru_size$size_t0), na.rm = T), 
                                       sd(  log(fru_size$size_t0), na.rm = T),
                                       df_grow$logsize_t0 %>% max, 
                                       df_grow$logsize_t0 %>% min,
                                       repr_pc_m$repr_pc_mean,
                                       fruit_pc_m$fruit_pc_mean))


bind_rows(grow_out %>% mutate(estimate = rep('growth', nrow(grow_out))), 
          surv_out %>% mutate(estimate = rep('survival', nrow(surv_out))),
          flwr_out %>% mutate(estimate = rep('flower', nrow(flwr_out))),
          others   %>% mutate(estimate = c(
            'recuitment', 'recuitment', 'fruit', 'fruit', 
            'other', 'other', 'recuitment', 'fruit'))) %>% 
  `rownames<-`(NULL) %>% 
  select(c(3, 1, 2))
```

