# Data NEW {#data_n}

This dataset provides demographic information from a long-term study of *Crotalaria avonensis*, an endangered small perennial herbaceous legume native to the Lake Wales Ridge in central Florida. The [study](https://portal.edirepository.org/nis/metadataviewer?packageid=edi.219.1) spans from 1998 to 2017 and monthly during the growing season tracks individual plant survival, size, and reproduction, recorded within permanent quadrats - 25 cm radius circles in stratified random positions within subjectively located belt transects. Additionally, the dataset contains information on fires, herbivory reduction through caging, and other events that might be of effect. Individuals were defined as groups of stems separated by 5 cm or more and marked with plastic toothpicks. Transplant data suggests this species has clonal growth. 

## New data! 

In the following we will explore the new, unreleased, extended data provided by the authors. 
First we notice that the data format is different. We now have year to year data, instead of some complex monthly sencor. 

The dataset includes a unique id for each plant and records from `site`, `mp` (macroplot), `quad` (circular quadrat), and `year` to identify where and when each observation was collected. It also includes survival and status variables such as `s` (survival code: 0 = dead or dormant, 1 = alive, 2 = missing, 3 = new adult, 5 = new seedling, 6 = dead/run over, 8= dormant), `alive` (alive or dead), and `stage` (life stage: 0=dead, 1=seedling, 2=vegetative and not flowering, 3=flowering), along with reproductive measures `maxbr`, `maxfl`, and `maxfr`, which capture the maximum number of branches, flowers, and fruits observed across the three sampling dates in a year.



```{r}
df_og <- read_delim("C:/code/RUPDemo_IPMs/archbold/data/crav/crotalaria_avonensis_data_v260515.csv", 
                    delim = ";", escape_double = FALSE, trim_ws = TRUE)
skim(df_og)
```


We found 3 plant individuals that seem to never have been recorded alive. Their IDs are the following `517`, `902`, `922`

```{r}
df_og %>% 
  filter(id %in% c(517, 902, 922))
```

## Disturbance
A disturbance history table was created from the quadrat dataset by extracting all burn-related variables and converting them into a long-format structure using `pivot_longer()`. Each burn variable was mapped to its corresponding disturbance year (2005, 2015, 2016, 2017, 2018, 2021, 2022, and 2023), and quadrats with recorded disturbance values greater than zero were assigned a disturbance indicator of 1. The resulting table contains one row per disturbance event and quadrat combination, including the columns `site`, `mp`, `quad`, `year`, and `disturbance.`

```{r}
df_og_quad <- read_delim('C:/code/RUPDemo_IPMs/archbold/data/crav/crotalaria_avonensis_data_quad_v260515.csv',
                         delim = ';', escape_double = FALSE, trim_ws = TRUE)
skim(df_og_quad)
```

```{r}
df_disturbance <- df_og_quad %>%
    # keep only ID columns + burn columns
  select(site, mp, quad,
         burn221, burn1222, burn0123, burn05, burn2014_2015, burn2016,
         burn0217, burnoct2017, burnjun2018) %>%
  pivot_longer(
    cols = starts_with('burn'), names_to = 'burn_event', values_to = 'value') %>%
  mutate(
    disturbance = case_when(
      is.na(value) ~ 0,
      value == FALSE ~ 0,
      value == TRUE ~ 1,
      suppressWarnings(as.numeric(value)) > 0 ~ 1,
      TRUE ~ 0)) %>%
  filter(disturbance == 1) %>%
  # assign actual years
  mutate(
    year = case_when(
      burn_event == 'burn221' ~ 2021,
      burn_event == 'burn1222' ~ 2022,
      burn_event == 'burn0123' ~ 2023,
      burn_event == 'burn05' ~ 2005,
      burn_event == 'burn2014_2015' ~ 2015,
      burn_event == 'burn2016' ~ 2016,
      burn_event == 'burn0217' ~ 2017,
      burn_event == 'burnoct2017' ~ 2017,
      burn_event == 'burnjun2018' ~ 2018),
    disturbance = 1) %>%
  mutate(site = as.factor(site),
         mp   = as.factor(mp),
         quad = as.factor(quad)) %>%
  # final columns
  select(site, mp, quad, year, disturbance) %>%
  distinct() %>%
  arrange(site, mp, quad, year)
```


### Off-meta fire
There is another fire record that is not mentioned in the meta sheet, i.e. `burn221`, most likely this is a fire in 2021. 



## Work data frames
The original demographic dataset was cleaned and standardized using `janitor::clean_names()`, after which several variables were renamed for clarity, including plant identity (`plant_id`), survival status (`survives`), and initial plant size (`size_t0`). Unique quadrat identifiers (`quad_id`) were created by combining site, macroplot, and quadrat information, and all grouping variables were converted to factors to facilitate downstream analyses. Records were then ordered chronologically by site, quadrat, plant, and year. For surviving individuals, subsequent-year size (`size_t1`) was generated by assigning the following year’s recorded size using `lead()`, while non-surviving individuals were assigned missing values. Recruitment status was derived from life stage and first observation year, with seedlings (`stage == 1)` and newly appearing individuals after the initial census period classified as recruits. Dormancy was defined for individuals recorded as alive but lacking size measurements, whereas individuals with measured size were considered non-dormant. A consecutive dormancy counter (`dormancy_count`) was additionally generated using lagged dormancy states to quantify repeated years of aboveground absence. Survival records from the final four years of the dataset were adjusted to missing values when plants were recorded as dead, thereby accounting for the possibility of unresolved dormancy rather than confirmed mortality. Plant size variables were log-transformed (`logsize_t0`, `logsize_t1`) and expanded into quadratic and cubic terms to support non-linear modeling of size-dependent demographic responses. Finally, the demographic dataset was joined with the quadrat-level disturbance history table using shared spatial and temporal identifiers (`site`, `mp`, `quad`, and `year`). Missing disturbance values were interpreted as undisturbed conditions and recoded to zero. The resulting dataset retained variables describing plant identity, survival, size, reproduction, recruitment, dormancy, and disturbance history for subsequent demographic analyses.

**Survival** We took the alive column and transformed it into the survives witht the lead() function.

**Growth** Yet again, this plant species looks at maximum branches in a year as the size measure and therefore is not a real continous measure. Which would make it more suitable for a MPM. Nevertheless, we will apply a IPM to this dataset. **Log-transformation** is without adding a small number which asigns a `-Inf` to size at t0 = 0 observations. The subsequent script excludes all observations with a non finite size.

**Disturbance** takes fire and any other measured disturbance to the plot into account with we got as an extra data set alongside its meta data.

**Flower** is the maximum flower count provided by the authors.


```{r}
df <- df_og %>% 
  janitor::clean_names() %>%  
  rename(plant_id = id,
         size_t0  = maxbr,
         flower   = maxfl,
         fruit    = maxfr) %>% 
  mutate(
    plant_id = as.factor(plant_id),
    quad_id  = as.factor(paste(site, mp, quad, sep = '_')),
    site     = as.factor(site),
    quad     = as.factor(quad),
    mp       = as.factor(mp)) %>%
  arrange(site, quad, quad_id, plant_id, year) %>%
  
  # Survival
  group_by(plant_id) %>% 
  mutate(survives = lead(alive)) %>% 
  ungroup %>% 
  
  # Growth based on survival; propagate size_t0 if survives, otherwise set to NA
  group_by(plant_id) %>%
  mutate(
    size_t1 = case_when(
      survives == 1 ~ lead(size_t0), 
      TRUE          ~ NA_real_)) %>%
  ungroup() %>% 
  
  # Recruits
  mutate(
    latest_alive_date      = max(year[(s > 0 & s != 2 & s < 6)], na.rm = TRUE),
    earliest_recorded_date = min(year[(s > 0 & s != 2 | 6)]    , na.rm = TRUE)) %>%
  mutate(
    recruit = case_when(
      stage == 1 ~ 1,
      year  == earliest_recorded_date & year(earliest_recorded_date) > 1999 + 4 ~ 1,
      TRUE ~ NA_real_)) %>% 
  
  # Dormancy
  mutate(
    dormancy = case_when(
      survives == 1 & is.na(size_t0)              ~ 1,
      size_t0  >  0 & !is.na(survives)            ~ 0,
      survives == 0 & year <= max(df_og$year) - 4 ~ 0, 
      TRUE                                        ~ NA_real_ ),
    # Generate a new column 'dormancy_count' that counts consecutive 1s
    dormancy_count = case_when(
      dormancy == 1 & lag(dormancy, 1) == 1 & lag(dormancy, 2) == 1 & 
        lag(dormancy, 3) == 1 & lag(dormancy, 4) == 1 & lag(dormancy, 5) == 1 ~ 6,
      dormancy == 1 & lag(dormancy, 1) == 1 & lag(dormancy, 2) == 1 
      & lag(dormancy, 3) == 1 & lag(dormancy, 4) == 1                         ~ 5,
      dormancy == 1 & lag(dormancy, 1) == 1 & lag(dormancy, 2) == 1 
      & lag(dormancy, 3) == 1                                                 ~ 4,
      dormancy == 1 & lag(dormancy, 1) == 1 & lag(dormancy, 2) == 1           ~ 3,
      dormancy == 1 & lag(dormancy, 1) == 1 & lag(dormancy, 2) == 0           ~ 2,
      dormancy == 1                                                           ~ 1,
      TRUE ~ dormancy)) %>%
  # adjust the survival for potential dormancy
  mutate(survives = if_else(survives == 0 & year > max(df_og$year) - 4, NA, survives)) %>%
  
    # Log-transformed sizes
  mutate(
    logsize_t0   = log(size_t0),     
    logsize_t1   = log(size_t1),    
    logsize_t0_2 = logsize_t0^2,     
    logsize_t0_3 = logsize_t0^3) %>% 
  
  # Disturbance form the quad data
  full_join(df_disturbance, by = c('site', 'mp', 'quad', 'year')) %>%
  mutate(disturbance = ifelse(is.na(disturbance), 0, disturbance)) %>% 
  
  select(site, mp, quad, quad_id, plant_id, year, 
         s, survives, size_t0, size_t1, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
         flower, fruit, recruit, dormancy, disturbance)
```

