# Models {#models_n}

For this IPM we will model the following processes:

1. Survival (binmial)
2. Growth
3. Flowering (binomial)
4. Fruiting (negative binomial)
5. Recruits (negative binomial)

## Survival

```{r su_data}
df_su <- df %>%
  filter(size_t0 != 0) %>%
  mutate(dist_label = ifelse(disturbance == 1, 'Disturbance', 'No disturbance')) %>%
   select(plant_id, year, size_t0, survives, size_t1,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
         disturbance, dist_label)
```


```{r su_model}
# Logistic regression
mod_su_0 <- glm(survives ~ disturbance,
                data = df_su, family = 'binomial')
# Logistic regression
mod_su_1 <- glm(survives ~ logsize_t0 + disturbance,
                data = df_su, family = 'binomial')
# Quadratic logistic model
mod_su_2 <- glm(survives ~ logsize_t0 + logsize_t0_2 + disturbance,
                data = df_su, family = 'binomial')
# Cubic logistic model
mod_su_3 <- glm(survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
                data = df_su, family = 'binomial')


# Compare models using AICc
mods_su      <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAICc <- AICctab(mods_su, weights = T, sort = F)$dAICc

# Get the sorted indices of dAICc values
mods_su_sorted <- order(mods_su_dAICc)

# Establish the index of model complexity
if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index       <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su +1
  v_mod_su_index       <- v_mod_set_su
}

mod_su_bestfit <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef   <- coef(mod_su_bestfit)
```

```{r su_predictions}
# Create prediction data frames for disturbance = 0 and disturbance = 1
df_su_pred <- data.frame(
  logsize_t0 = rep(seq(min(df_su$logsize_t0), max(df_su$logsize_t0), length.out = 100), 2),
  disturbance = rep(c(0, 1), each = 100)) %>%
  mutate(dist_label = ifelse(disturbance == 1, 'Disturbance', 'No disturbance'),
         survives = predict(mod_su_bestfit, newdata = ., type = 'response'))

# Binned observed data for both disturbance levels
df_su_binned <- bind_rows(
  plot_binned_prop(filter(df_su, disturbance == 0), 10, logsize_t0, survives) %>%
    mutate(dist_label = 'No disturbance'),
  plot_binned_prop(filter(df_su, disturbance == 1), 10, logsize_t0, survives) %>%
    mutate(dist_label = 'Disturbance'))
```

```{r su_plot, warning=FALSE}
# Plot 1: Raw data + prediction lines
fig_su_line_combined <- ggplot() +
  geom_jitter(data = df_su, aes(x = logsize_t0, y = survives, color = dist_label),
              alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(data = df_su_pred, aes(x = logsize_t0, y = survives, color = dist_label),
            linewidth = 0.9) +
  scale_color_manual(values = c('No disturbance' = 'black', 'Disturbance' = 'red')) +
  theme_bw() +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Survival to time t1') +
  theme(legend.position = 'none')

# Plot 2: Binned proportions + prediction lines
fig_su_bin_combined <- ggplot() +
  geom_point(data = df_su_binned, aes(x = logsize_t0, y = survives, color = dist_label)) +
  geom_errorbar(data = df_su_binned, aes(x = logsize_t0, ymin = lwr, ymax = upr, color = dist_label),
                width = 0.2) +
  geom_line(data = df_su_pred, aes(x = logsize_t0, y = survives, color = dist_label),
            linewidth = 0.9) +
  scale_color_manual(values = c('No disturbance' = 'black', 'Disturbance' = 'red')) +
  theme_bw() +
  ylim(0, 1) +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Survival to time t1') +
  theme(legend.title = element_blank(), legend.position = 'top')

# Combine the two plots
fig_su_all <- fig_su_line_combined + fig_su_bin_combined +
  plot_annotation(
    title = 'Survival',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_su_all
```

## Growth
```{r gr_data}
df_gr <- df %>%
  filter(size_t0 != 0,
         is.finite(logsize_t1)) %>%
  mutate(dist_label = ifelse(disturbance == 1, 'disturbance', 'No disturbance')) %>%
  select(plant_id, year, size_t0, size_t1,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
         disturbance, dist_label)
```

```{r gr_model}
# Intercept model
mod_gr_0 <- lm(logsize_t1 ~ disturbance,
               data = df_gr, na.action = na.omit)
# Linear model
mod_gr_1 <- lm(logsize_t1 ~ logsize_t0 + disturbance,
               data = df_gr)
# Quadratic model
mod_gr_2 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + disturbance,
               data = df_gr)
# Cubic model
mod_gr_3 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
               data = df_gr)

mods_gr      <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAICc <- AICctab(mods_gr, weights = T, sort = F)$dAICc

# Get the sorted indices of dAICc values
mods_gr_sorted <- order(mods_gr_dAICc)

# Establish the index of model complexity
if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index       <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr +1
  v_mod_gr_index       <- v_mod_set_gr
}

mod_gr_bestfit         <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef           <- coef(mod_gr_bestfit)
```

```{r gr_preditions}
# Create prediction data for disturbance = 0 and disturbance = 1
df_gr_pred <- data.frame(
  logsize_t0 = rep(seq(min(df_gr$logsize_t0, na.rm = TRUE),
                       max(df_gr$logsize_t0, na.rm = TRUE), length.out = 100), 2),
  disturbance = rep(c(0, 1), each = 100))

df_gr_pred <- df_gr_pred %>%
  mutate(logsize_t0_2 = logsize_t0^2,
         logsize_t0_3 = logsize_t0^3,
         dist_label   = ifelse(disturbance == 1, 'disturbance', 'No disturbance'))

df_gr_pred$logsize_t1 <- predict(mod_gr_bestfit, newdata = df_gr_pred)
```

```{r gr_plot}
# Plot 1: Observed points + prediction lines
fig_gr_line_combined <- ggplot(df_gr, aes(x = logsize_t0, y = logsize_t1, color = dist_label)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(aes(x = logsize_t0, y = logsize_t1, color = dist_label),
            data = df_gr_pred, linewidth = 1) +
  scale_color_manual(values = c('No disturbance' = 'black', 'disturbance' = 'red')) +
  theme_bw() +
  labs(x = expression('log(size)'[t0]),
       y = expression('log(size)'[t1])) +
  theme(
    legend.position = 'top',
    legend.title = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank())

# Plot 2: Predicted vs observed
fig_gr_pred_combined <- ggplot(df_gr, aes(x = predict(mod_gr_bestfit, newdata = df_gr),
                                          y = logsize_t1, color = dist_label)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No disturbance' = 'black', 'disturbance' = 'red')) +
  theme_bw() +
  labs(x = 'Predicted', y = 'Observed') +
  theme(
    legend.position = 'top',
    legend.title = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank())

# Combine plots
fig_gr_all_combined <- fig_gr_line_combined + fig_gr_pred_combined +
  plot_annotation(
    title = 'Growth Prediction',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_gr_all_combined
```


### Growth variance

```{r gr_variance}
# Fitted values from growth model
mod_gr_x   <- fitted(mod_gr_bestfit)
# Squared residuals
mod_gr_y   <- resid(mod_gr_bestfit)^2
# Non-linear model for variance
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x), start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```


## Flowering
```{r fl_data}
df_fl <- df %>%
  filter(!is.na(flower), is.finite(logsize_t0), 
         !is.na(disturbance) & !is.na(size_t0) & !is.na(survives)) %>%
  select(plant_id, year, size_t0, flower, size_t1,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, disturbance) %>%
  mutate(
    flower = if_else(flower > 0, 1, flower),
    dist_label = if_else(disturbance == 1, 'disturbance', 'No disturbance'))
```

```{r fl_model}
# Logistic regression
mod_fl_0 <- glm(flower ~ disturbance,
                data = df_fl, family = 'binomial')
# Logistic regression
mod_fl_1 <- glm(flower ~ logsize_t0 + disturbance,
                data = df_fl, family = 'binomial')
# Quadratic logistic model
mod_fl_2 <- glm(flower ~ logsize_t0 + logsize_t0_2 + disturbance,
                data = df_fl, family = 'binomial')
# Cubic logistic model
mod_fl_3 <- glm(flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
                data = df_fl, family = 'binomial')

# Compare models using AICc
mods_fl      <- list(mod_fl_0, mod_fl_1, mod_fl_2, mod_fl_3)
mods_fl_dAICc <- AICctab(mods_fl, weights = T, sort = F)$dAICc

# Get the sorted indices of dAICc values
mods_fl_sorted <- order(mods_fl_dAICc)

# Establish the index of model complexity
if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index       <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl +1
  v_mod_fl_index       <- v_mod_set_fl
}

mod_fl_bestfit <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef   <- coef(mod_fl_bestfit)
```

```{r fl_predictions}
# Create prediction data frames for disturbance = 0 and disturbance = 1
df_fl_pred_0 <- data.frame(
  logsize_t0 = seq(
    min(df_fl$logsize_t0, na.rm = T), max(df_fl$logsize_t0, na.rm = T), length.out = 100),
  disturbance = 0) %>% mutate(logsize_t0_2 = logsize_t0^2)

df_fl_pred_1 <- data.frame(
  logsize_t0 = seq(
    min(df_fl$logsize_t0, na.rm = T), max(df_fl$logsize_t0, na.rm = T), length.out = 100),
  disturbance = 1) %>% mutate(logsize_t0_2 = logsize_t0^2)

# Add predictions using model
df_fl_pred <- data.frame(
  logsize_t0 = rep(seq(min(df_fl$logsize_t0, na.rm = TRUE), max(df_fl$logsize_t0, na.rm = TRUE), length.out = 100), 2),
  disturbance = rep(c(0, 1), each = 100)) %>%
  mutate(
    dist_label = ifelse(disturbance == 1, 'disturbance', 'No disturbance'),
    logsize_t0_2 = logsize_t0^2)

df_fl_pred$flower <- predict(mod_fl_bestfit, newdata = df_fl_pred, type = 'response')

# Binned observed data for both disturbance levels
df_fl_binned <- bind_rows(
  plot_binned_prop(filter(df_fl, disturbance == 0), 10, logsize_t0, flower) %>%
    mutate(dist_label = 'No disturbance'),
  plot_binned_prop(filter(df_fl, disturbance == 1), 10, logsize_t0, flower) %>%
    mutate(dist_label = 'disturbance'))
```

```{r fl_plot}
# Plot 1: Raw jitter + prediction lines
fig_fl_line_combined <- ggplot() +
  geom_jitter(data = df_fl, aes(x = logsize_t0, y = flower, color = dist_label),
              alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(data = df_fl_pred, aes(x = logsize_t0, y = flower, color = dist_label),
            linewidth = 0.9) +
  scale_color_manual(values = c('No disturbance' = 'black', 'disturbance' = 'red')) +
  theme_bw() +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Flowering Probability') +
  theme(legend.position = 'none')

# Plot 2: Binned points + error bars + prediction lines
fig_fl_bin_combined <- ggplot() +
  geom_point(data = df_fl_binned, aes(x = logsize_t0, y = flower, color = dist_label)) +
  geom_errorbar(data = df_fl_binned, aes(x = logsize_t0, ymin = lwr, ymax = upr, color = dist_label),
                width = 0.2) +
  geom_line(data = df_fl_pred, aes(x = logsize_t0, y = flower, color = dist_label),
            linewidth = 0.9) +
  scale_color_manual(values = c('No disturbance' = 'black', 'disturbance' = 'red')) +
  theme_bw() +
  ylim(0, 1) +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Flowering Probability') +
  theme(legend.title = element_blank(), legend.position = 'top')

# Combine the two plots with patchwork
fig_fl_all <- fig_fl_line_combined + fig_fl_bin_combined +
  plot_annotation(
    title = 'Flowering',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_fl_all
```


## Fruiting

```{r fr_data}
df_fr <- df %>%
  filter(is.finite(fruit), is.finite(logsize_t0))
```

### Fruiting disperion
Variance > Mean (especially > 1.5× or 2×) -> Overdispersed -> negative binomial
Variance = Mean                                            -> Poisson
```{r fr_dispertion}
mean(df_fr$fruit)
var(df_fr$fruit)
```

```{r fr_model}
fr_mod_f0 <- glm.nb(fruit ~ disturbance, data = df_fr)
fr_mod_1  <- glm.nb(fruit ~ logsize_t0, data = df_fr)
fr_mod_f1 <- glm.nb(fruit ~ logsize_t0 + disturbance, data = df_fr)
fr_mod_f2 <- glm.nb(fruit ~ logsize_t0 + logsize_t0_2 + disturbance, data = df_fr)
fr_mod_f3 <- glm.nb(fruit ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance, data = df_fr)

fr_mods      <- list(fr_mod_f0, fr_mod_1, fr_mod_f1, fr_mod_f2, fr_mod_f3)
fr_mods_dAICc <- AICctab(fr_mods, weights = T, sort = F)$dAICc

# Get the sorted indices of dAICc values
fr_mods_i_sort <- order(fr_mods_dAICc)

# Establish the index of model complexity
if (length(v_mod_set_fr) == 0) {
  fr_mod_i_best <- fr_mods_i_sort[1]
  v_mod_fr_i    <- fr_mod_i_best - 1
} else {
  fr_mod_i_best <- v_mod_set_fr +1
  v_mod_fr_i    <- v_mod_set_fr
}

fr_mod_best  <- fr_mods[[fr_mod_i_best]]
fr_mod_ranef <- coef(fr_mod_best)
```

```{r fr_predictions}
# Prediction data for disturbance = 0
df_pred_0 <- data.frame(
  logsize_t0  = seq(
    min(df_fr$logsize_t0, na.rm = TRUE), max(df_fr$logsize_t0, na.rm = TRUE), length.out = 200),
  disturbance = 0)

df_pred_0$logsize_t0_2 <- df_pred_0$logsize_t0^2
df_pred_0$pred         <- predict(fr_mod_best, newdata = df_pred_0, type = 'response')

# Prediction data for disturbance = 1
df_pred_1 <- data.frame(
  logsize_t0   = df_pred_0$logsize_t0,  # same x values
  disturbance  = 1)
df_pred_1$logsize_t0_2 <- df_pred_1$logsize_t0^2
df_pred_1$pred         <- predict(fr_mod_best, newdata = df_pred_1, type = 'response')

```

```{r fr_plot}
fig_fr <- ggplot(df_fr, aes(x = logsize_t0, y = fruit)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_line(data = df_pred_0, aes(x = logsize_t0, y = pred),
            color = '#1b9e77', linewidth = 1.2, linetype = 'solid') +
  geom_line(data = df_pred_1, aes(x = logsize_t0, y = pred),
            color = '#d95f02', linewidth = 1.2, linetype = 'dashed') +
  theme_bw() +
  labs(
    title    = 'Fruit Production by log(Size) and disturbance',
    subtitle = v_ggp_suffix,
    x        = 'log(Size at t0)',
    y        = 'Number of Fruits') +
  annotate('text', x = Inf, y = Inf, label = 'Solid = no disturbance; Dashed = disturbance',
           hjust = 1.1, vjust = 1.5, size = 3, color = 'gray40')
fig_fr
```


## Fruit to recruit transition
```{r f2r_data}
repr_pc_by_year <- {
  fruit_by_year <- df %>%
    filter(!is.na(fruit)) %>%
    group_by(year) %>%
    summarise(total_fruit = sum(fruit, na.rm = TRUE)) %>%
    mutate(year = year + 1)

  recruits_by_year <- df %>%
    filter(recruit == 1) %>%
    group_by(year) %>%
    summarise(n_recruits = n())

  recruits_by_year %>%
    left_join(fruit_by_year, by = 'year') %>%
    mutate(repr_pc_mean = n_recruits / total_fruit)
}
```

```{r f2r_coefficient}
repr_pc_mean   <- mean(  repr_pc_by_year$repr_pc_mean, na.rm = T)
repr_pc_median <- median(repr_pc_by_year$repr_pc_mean, na.rm = T)

repr_pc_by_year %>%
  summarise(mean(repr_pc_mean),
            sd  (repr_pc_mean),
            median(repr_pc_mean))
```


## Recruits

*** NEEDS WORK AND SUPERVISION ***

```{r re_data}
df_re <- df %>%
  group_by(year, site, quad_id) %>%
  summarise(disturbance = if_else(max(disturbance, na.rm = TRUE) == 1, 1, 0),
            tot_p_area = sum(size_t0, na.rm = TRUE),
            .groups = 'drop') %>%
  {
    df_quad <- .
    df_group <- df_quad %>%
      group_by(year) %>%
      summarise(g_cov = mean(tot_p_area), .groups = 'drop')

    df_cover <- left_join(df_quad, df_group, by = 'year') %>%
      mutate(year = as.integer(year + 1)) %>%
      drop_na()

    df_re <- df %>%
      group_by(year, site, quad_id) %>%
      summarise(nr_quad = sum(recruit, na.rm = TRUE), .groups = 'drop')

    left_join(df_cover, df_re, by = c('year', 'site', 'quad_id'))
  }
```

```{r re_plot}
ggplot(
  df_re, aes(x = tot_p_area, y = nr_quad, colour = as.factor(disturbance))) +
  geom_point(alpha = 0.5, pch = 16, size = 1) +
  theme_bw() +
  labs(title    = 'Recruitment',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),
       y        = expression('Number of recruits '     [t1]),
       colour   = expression('Disturbance')) +
  theme(plot.subtitle = element_text(size = 8))
```

**
```{r re_model n predictions}
df_re_mod <- df_re %>% filter(!is.na(nr_quad))
# Fit a negative binomial model for recruitment
mod_rec <- MASS::glm.nb(nr_quad ~ disturbance, data = df_re_mod)

# Generate predictions for recruitment
df_re_mod <- df_re_mod %>%
  mutate(mod_pred = predict(mod_rec, type = 'response'))
```

### Per-capita reproduction

```{r re_pc}
df_repr_pc <- df %>%
  filter(!is.na(size_t0)) %>%
  summarize(n_adults = n()) %>%
  bind_cols(
    df_re_mod %>%
      summarize(nr_quad = sum(nr_quad, na.rm = TRUE),
                mod_pred = sum(mod_pred, na.rm = TRUE))) %>%
  mutate(
    repr_pc_mean = mod_pred / n_adults,
    repr_pc_obs = nr_quad / n_adults) %>%
  drop_na()
```


### Recruitment: Density dependency
```{r re_dd_data}
df_re_qd <- df %>%
  group_by(site, quad_id, year) %>%
  dplyr::select(recruit) %>%
  summarise(rec_qd_t1 = sum(recruit, na.rm = T)) %>%
  left_join(df %>%
              group_by(site, quad_id, year) %>%
              summarise(nr_ind = sum(!is.na(size_t0)),
                        disturbance = as.factor(
                          if_else(max(disturbance, na.rm = TRUE) == 1, 1, 0))) %>%
              mutate(year = year - 1),
            by = c('site', 'quad_id', 'year'))
```
```{r}
fig_re_dens <- ggplot(data = df_re_qd) +
  geom_jitter(aes(y = rec_qd_t1, x = nr_ind, color = disturbance)) +
  geom_smooth(aes(y = rec_qd_t1, x = nr_ind, color = disturbance), method = 'lm') +
  theme_bw() +
  labs(title    = 'Recruitment - desity dependence: Quad level',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),
       y        = expression('Number of recruits '     [t1])) +
  theme(plot.subtitle = element_text(size = 8))
fig_re_dens
```

## Extracting parameter estimates
```{r parameter}
# Survival
coef_su_fe  <- data.frame(coefficient = names(coef(mod_su_bestfit)),
                          value       =       coef(mod_su_bestfit))

coef_su <- Reduce(function(...) rbind(...), list(coef_su_fe)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))

# Growth
coef_gr_fe  <- data.frame(coefficient = names(coef(mod_gr_bestfit)),
                          value       =       coef(mod_gr_bestfit))
coef_gr_var <- data.frame(coefficient = names(coef(mod_gr_var)),
                          value       =       coef(mod_gr_var))

coef_gr <- Reduce(function(...) rbind(...), list(coef_gr_fe, coef_gr_var)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))

# Flower
coef_fl_fe  <- data.frame(coefficient = names(coef(mod_fl_bestfit)),
                          value       =       coef(mod_fl_bestfit))

coef_fl <- Reduce(function(...) rbind(...), list(coef_fl_fe)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))

# Fruit
coef_fr_fe  <- data.frame(coefficient = names(coef(fr_mod_best)),
                          value       =       coef(fr_mod_best))

coef_fr <- Reduce(function(...) rbind(...), list(coef_fr_fe)) %>%
  mutate(coefficient = as.character(coefficient)) %>%
  mutate(coefficient = replace(
    coefficient, grepl('Intercept', coefficient), 'b0'))

# Recruitment
df_re_size <- df %>% subset(recruit == 1)

# Miscellany
coef_misc   <- data.frame(
  coefficient = c('rec_siz', 'rec_sd', 'fecu_b0', 'max_siz', 'min_siz'),
  value       = c(mean(log(df_re_size$size_t0), na.rm = T),
                  sd(  log(df_re_size$size_t0), na.rm = T),
                  repr_pc_median,
                  df_gr$logsize_t0 %>% max,
                  df_gr$logsize_t0 %>% min))

extr_value <- function(x, field){
  subset(x, coefficient == field)$value
}

pars <- Filter(function(x) length(x) > 0, list(
  prefix  = v_script_prefix,
  species = v_species,
  surv_b0 = extr_value(coef_su, 'b0'),
  surv_b1 = extr_value(coef_su, 'logsize_t0'),
  surv_b2 = extr_value(coef_su, 'logsize_t0_2'),
  surv_b3 = extr_value(coef_su, 'logsize_t0_3'),
  surv_bf = extr_value(coef_su, 'disturbance'),
  grow_b0 = extr_value(coef_gr, 'b0'),
  grow_b1 = extr_value(coef_gr, 'logsize_t0'),
  grow_b2 = extr_value(coef_gr, 'logsize_t0_2'),
  grow_b3 = extr_value(coef_gr, 'logsize_t0_3'),
  grow_bf = extr_value(coef_gr, 'disturbance'),
  a       = extr_value(coef_gr, 'a'),
  b       = extr_value(coef_gr, 'b'),
  fl_b0   = extr_value(coef_fl, 'b0'),
  fl_b1   = extr_value(coef_fl, 'logsize_t0'),
  fl_b2   = extr_value(coef_fl, 'logsize_t0_2'),
  fl_b3   = extr_value(coef_fl, 'logsize_t0_3'),
  fl_bf   = extr_value(coef_fl, 'disturbance'),
  fr_b0   = extr_value(coef_fl, 'b0'),
  fr_b1   = extr_value(coef_fr, 'logsize_t0'),
  fr_b2   = extr_value(coef_fr, 'logsize_t0_2'),
  fr_b3   = extr_value(coef_fr, 'logsize_t0_3'),
  fr_bf   = extr_value(coef_fr, 'disturbance'),
  fecu_b0 = extr_value(coef_misc, 'fecu_b0'),
  recr_sz = extr_value(coef_misc, 'rec_siz'),
  recr_sd = extr_value(coef_misc, 'rec_sd'),
  L       = extr_value(coef_misc, 'min_siz'),
  U       = extr_value(coef_misc, 'max_siz'),
  mat_siz = 200,
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_gr_index = v_mod_fl_index))
```


## IPM building

### Functions
```{r}
# Function describing the invert logit
inv_logit <- function(x) {exp(x) / (1 + exp(x))}

# Survival of x-sized individual to time t1
sx <- function(x, pars, num_pars = v_mod_su_index) {
  survival_value <- pars$surv_b0
  for (i in 1:num_pars) {
    param_name <- paste0('surv_b', i)
    if (!is.null(pars[[param_name]])) {
      survival_value <- survival_value + pars[[param_name]] * x^(i)
    }
  }
  return(inv_logit(survival_value))
}

# Function describing standard deviation of growth model
grow_sd <- function(x, pars) {
  pars$a * (exp(pars$b* x)) %>% sqrt
}

# Growth from size x to size y
gxy <- function(x, y, pars, num_pars = v_mod_gr_index) {
  mean_value <- 0
  for (i in 0:num_pars) {
    param_name <- paste0('grow_b', i)
    if (!is.null(pars[[param_name]])) {
      mean_value <- mean_value + pars[[param_name]] * x^i
    }
  }
  sd_value <- grow_sd(x, pars)
  return(dnorm(y, mean = mean_value, sd = sd_value))
}

# Function describing the transition kernel
pxy <- function(x, y, pars) {
  return(sx(x, pars) * gxy(x, y, pars))
}

# Flowering of x-sized individual at time t0
fl_x <- function(x, pars, num_pars = v_mod_fl_index) {
  val <- pars$fl_b0
  for (i in 1:num_pars) {
    param <- paste0('fl_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  inv_logit(val)
}

# Fruiting of x-sized individuals at time t0
fr_x <- function(x, pars, num_pars = v_mod_fr_i) {
  val <- pars$fr_b0
  for (i in 1:num_pars) {
    param <- paste0('fr_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  exp(val)  # Negative binomial uses log link
}

# Recruitment size distribution at time t1
re_y_dist <- function(y, pars) {
  dnorm(y, mean = pars$recr_sz, sd = pars$recr_sd)
}

# F-kernel
fyx <- function(y, x, pars) {
  fl_x(x, pars) *
    fr_x(x, pars) *
    pars$fecu_b0 *
    re_y_dist(y, pars)
}
```


### Kernel
```{r}
# Kernel
kernel <- function(pars) {

  # number of bins over which to integrate
  n   <- pars$mat_siz
  # lower limit of integration
  L   <- pars$L
  # upper limit of integration
  U   <- pars$U
  # bin size
  h   <- (U - L) / n
  # lower boundaries of bins
  b   <- L + c(0:n) * h
  # midpoints of bins
  y   <- 0.5 * (b[1:n] + b[2:(n + 1)])

  # Survival vector
  Smat   <- c()
  Smat   <- sx(y, pars)

  # Growth matrix
  Gmat   <- matrix(0, n, n)
  Gmat[] <- t(outer(y, y, gxy, pars)) * h

  # Growth/survival transition matrix
  Tmat   <- matrix(0, n, n)

  # Correct for eviction of offspring
  for(i in 1:(n / 2)) {
    Gmat[1,i] <- Gmat[1,i] + 1 - sum(Gmat[,i])
    Tmat[,i]  <- Gmat[,i] * Smat[i]
  }

  # Correct eviction of large adults
  for(i in (n / 2 + 1):n) {
    Gmat[n,i] <- Gmat[n,i] + 1 - sum(Gmat[,i])
    Tmat[,i]  <- Gmat[,i] * Smat[i]
  }

  # Fertility matrix
  Fmat <- outer(y, y, Vectorize(function(x, y) fyx(x, y, pars))) * h

  # Full Kernel is simply a summation of fertility and transition matrices
  k_yx <- Fmat + Tmat

  return(list(k_yx    = k_yx,
              Fmat    = Fmat,
              Tmat    = Tmat,
              Gmat    = Gmat,
              meshpts = y))
}

lambda_ipm <- function(i) {
  return(Re(eigen(kernel(i)$k_yx)$value[1]))
}
```

### Lambda
#### Mean population growth rate
```{r mpgr}
lam_mean <- lambda_ipm(pars)
lam_mean
```

### Expected growth under disturbance regime (mean exposure per year)
```{r mpgr_disturbance}
# Build no-disturbance version of parameters
pars_no_disturbance <- pars
pars_no_disturbance$surv_b0 <- pars$surv_b0 + pars$surv_bf * 0
pars_no_disturbance$grow_b0 <- pars$grow_b0 + pars$grow_bf * 0
pars_no_disturbance$fl_b0   <- pars$fl_b0   + pars$fl_bf   * 0
pars_no_disturbance$fr_b0   <- pars$fr_b0   + pars$fr_bf   * 0

# Build disturbance version of parameters
pars_disturbance <- pars
pars_disturbance$surv_b0 <- pars$surv_b0 + pars$surv_bf * 1
pars_disturbance$grow_b0 <- pars$grow_b0 + pars$grow_bf * 1
pars_disturbance$fl_b0   <- pars$fl_b0   + pars$fl_bf * 1
pars_disturbance$fr_b0   <- pars$fr_b0   + pars$fr_bf * 1

# Compute deterministic lambdas
lam_nodisturbance <- lambda_ipm(pars_no_disturbance)
lam_disturbance   <- lambda_ipm(pars_disturbance)

# Expected growth under disturbance regime (mean exposure per plant)
p_disturbance <- mean(df %>%
                 filter(!is.na(survives)) %>%
                 .$disturbance, na.rm = TRUE)
# Expected growth under disturbance regime (mean exposure per year)
p_disturbance2 <- df %>%
  group_by(year) %>%
  summarise(disturbance = max(disturbance, na.rm = TRUE)) %>%
  pull(disturbance) %>%
  mean()
lam_avg <- (1 - p_disturbance2) * lam_nodisturbance + p_disturbance2 * lam_disturbance
lam_avg 

```

#### Observed population growth
```{r opg}
df_counts_year <- df %>%
  group_by(year) %>%
  filter(!is.na(size_t0)) %>%
  summarise(n = n())

# Then compute observed lambda
lam_obs_y <- df_counts_year$n[-1] / df_counts_year$n[-nrow(df_counts_year)]
lam_obs_mean <- mean(lam_obs_y, na.rm = TRUE)
lam_obs_mean
```

