# by Site {#by_site}

This section extends the mean IPM from the previous file. The vital-rate models, mean IPM functions already exist from **Models NEW**.

The difference from the mean IPM is that recruitment is no longer represented by one global recruits-per-fruit scalar. Instead, we fit a site-level fruit-to-recruit transition and use it to build one IPM per site. Disturbance is included as a site-specific disturbance frequency that modifies survival, growth, flowering, and fruiting through the disturbance coefficients already estimated in the mean model.

The biological path in the fertility kernel is:

```r
Pr(flowering | size, disturbance) *
  E(fruits | flowering, size, disturbance) *
  E(recruits | site-level fruit production)
```

2026.7.9: Forcing the log1p(fruit_t0) to log1p(recruit_t1) model to go through 0 as a fixed intercept.

## Required objects

This section assumes that the previous R Markdown file has already run.

```{r by_site_required_objects}
required_objects <- c(
  "df",
  "pars",
  "fl_x",
  "fr_x",
  "re_y_dist",
  "kernel",
  "lambda_ipm",
  "v_ggp_suffix")

missing_objects <- required_objects[!purrr::map_lgl(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop(
    "The following objects are missing. Run the mean-IPM script first: ",
    paste(missing_objects, collapse = ", "))
}

if (!exists("v_mod_set_fr2re_site")) {
  v_mod_set_fr2re_site <- c()
}
```

## Site-level fruit-to-recruit transition

The mean IPM used one mean recruits-per-fruit value. Here, we first summarize total fruits per site-year and total recruits in the following year. Recruitment is counted with `recruit == 1` because `recruit` is coded as `NA` or `1`.

```{r by_site_fruit_to_recruit_data}
df_fr2re_site <- df %>%
  mutate(
    site = as.factor(site),
    year = as.numeric(year)
  ) %>%
  group_by(site, year) %>%
  summarise(
    fruits_site_t0 = sum(fruit, na.rm = TRUE),
    recruits_site_t0 = sum(recruit == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(site, year) %>%
  group_by(site) %>%
  mutate(
    recruits_site_t1 = lead(recruits_site_t0),
    year_t1 = lead(year),
    year_gap = year_t1 - year
  ) %>%
  ungroup() %>%
  filter(year_gap == 1) %>%
  mutate(
    logfruits_site_t0 = log1p(fruits_site_t0),
    logrecruits_site_t1 = log1p(recruits_site_t1)
  )

df_fr2re_site %>%
  summarise(
    n_site_year_transitions = n(),
    n_positive_recruit_transitions = sum(recruits_site_t1 > 0),
    n_zero_recruit_transitions = sum(recruits_site_t1 == 0),
    total_fruits_t0 = sum(fruits_site_t0),
    total_recruits_t1 = sum(recruits_site_t1),
    max_fruits_site_t0 = max(fruits_site_t0),
    max_recruits_site_t1 = max(recruits_site_t1),
    proportion_zero_recruit = mean(recruits_site_t1 == 0)
  )
```

## Candidate fruit-to-recruit models

We fit the same basic relationship in all candidate models:

```r
log1p(recruits at site in year t + 1) ~ log1p(fruits at site in year t)
```

The candidate set has a forced intercept at 0 and differs only in whether the slope are allowed to vary by site.

```{r by_site_fruit_to_recruit_models}
# These models are forced through 0 on the log1p scale:
# log1p(recruits) = b1 * log1p(fruits).
#
# This means:
# fruits = 0 -> log1p(fruits) = 0
# predicted log1p(recruits) = 0
# predicted recruits = expm1(0) = 0

# 1. Fixed-effect slope-only model
fr2re_site_mod_1 <- glmmTMB(
  logrecruits_site_t1 ~ 0 + logfruits_site_t0,
  family = gaussian,
  REML = FALSE,
  data = df_fr2re_site)

# 2. Site-specific random slope only
fr2re_site_mod_2 <- glmmTMB(
  logrecruits_site_t1 ~ 0 + logfruits_site_t0 +
    (0 + logfruits_site_t0 | site),
  family = gaussian,
  REML = FALSE,
  data = df_fr2re_site)

fr2re_site_mods <- list(
  fr2re_site_mod_1,
  fr2re_site_mod_2)

fr2re_site_mods_dAICc <- AICctab(
  fr2re_site_mods,
  weights = TRUE,
  sort = FALSE)$dAICc

fr2re_site_mods_i_sort <- order(fr2re_site_mods_dAICc)

if (length(v_mod_set_fr2re_site) == 0) {
  fr2re_site_mod_i_best <- fr2re_site_mods_i_sort[1]
  v_mod_fr2re_site_index <- fr2re_site_mod_i_best
} else {
  fr2re_site_mod_i_best <- v_mod_set_fr2re_site
  v_mod_fr2re_site_index <- v_mod_set_fr2re_site
}

fr2re_site_mod_best <- fr2re_site_mods[[fr2re_site_mod_i_best]]
summary(fr2re_site_mod_best)
```

```{r by_site_fruit_to_recruit_convergence}
lapply(fr2re_site_mods, function(m) {
  list(
    convergence = m$fit$convergence,
    message = m$fit$message
  )
})
```

## Extract site-specific fruit-to-recruit coefficients

The extraction below works for both candidate models. If the selected model does not include a random intercept or random slope, that missing random effect is set to zero.

```{r by_site_fruit_to_recruit_coefficients}
fr2re_site_fixef <- fixef(fr2re_site_mod_best)$cond

fr2re_site_ranef <- tryCatch(
  {
    ranef(fr2re_site_mod_best)$cond$site %>%
      as.data.frame() %>%
      tibble::rownames_to_column("site")
  },
  error = function(e) {
    tibble(site = levels(df_fr2re_site$site))
  })

# Add missing random slope column as zero.
# There should be no intercept column because the model is forced through 0.
if (!"logfruits_site_t0" %in% names(fr2re_site_ranef)) {
  fr2re_site_ranef[["logfruits_site_t0"]] <- 0
}

df_fr2re_site_coef <- tibble(site = levels(df_fr2re_site$site)) %>%
  left_join(fr2re_site_ranef, by = "site") %>%
  mutate(
    logfruits_site_t0 = replace_na(logfruits_site_t0, 0),
    
    # Keep this only for compatibility with the rest of the IPM.
    # It is forced to zero.
    fr2re_b0_site = 0,
    
    fr2re_b1_site =
      unname(fr2re_site_fixef["logfruits_site_t0"]) + logfruits_site_t0) %>%
  select(site, fr2re_b0_site, fr2re_b1_site)
```

```{r by_site_fruit_to_recruit_prediction_function}
predict_site_recruits <- function(logfruits_site_t0, fr2re_b0, fr2re_b1) {
  
  pred_log_recruits <- fr2re_b1 * logfruits_site_t0
  
  # Model was fit to log1p(recruits), so backtransform with expm1()
  pred_recruits <- expm1(pred_log_recruits)
  
  # Avoid impossible negative predictions
  pmax(pred_recruits, 0)
}
```

```{r by_site_fruit_to_recruit_plot, warning=FALSE}
df_fr2re_site_plot <- df_fr2re_site %>%
  left_join(df_fr2re_site_coef, by = "site") %>%
  mutate(
    pred_logrecruits_site_t1 = fr2re_b1_site * logfruits_site_t0
  ) %>%
  arrange(site, logfruits_site_t0)

fig_fr2re_site <- ggplot(
  df_fr2re_site_plot,
  aes(x = logfruits_site_t0, y = logrecruits_site_t1)
) +
  geom_point(
    alpha = 0.45,
    size = 1
  ) +
  geom_line(
    aes(y = pred_logrecruits_site_t1),
    linewidth = 0.9
  ) +
  facet_wrap(~ site, scales = "free_y") +
  labs(
    x = "log1p(total fruits at site in year t)",
    y = "log1p(total recruits at site in year t + 1)",
    title = "Site-level fruit-to-recruit transition",
    subtitle = "log1p(recruits) ~ 0 + log1p(fruits), forced through zero"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 9),
    panel.grid.minor = element_blank()
  )

fig_fr2re_site
```


## Pars
```{r by_site_zero_intercept_pars}

pars_mean <- pars

if (!exists("v_mod_fr_index")) {
  v_mod_fr_index <- pars_mean$mod_fr_index
}


pars <- Filter(function(x) length(x) > 0, list(
  prefix  = v_script_prefix,
  species = v_species,
  surv_b0 = extr_value(coef_su, 'b0'),
  surv_b1 = extr_value(coef_su, 'logsize_t0'),
  surv_b2 = extr_value(coef_su, 'logsize_t0_2'),
  surv_b3 = extr_value(coef_su, 'logsize_t0_3'),
  surv_bf = extr_value(coef_su, 'disturbance'),
  grow_b0 = extr_value(coef_gr, 'b0'),
  grow_b1 = extr_value(coef_gr, 'logsize_t0'),
  grow_b2 = extr_value(coef_gr, 'logsize_t0_2'),
  grow_b3 = extr_value(coef_gr, 'logsize_t0_3'),
  grow_bf = extr_value(coef_gr, 'disturbance'),
  a       = extr_value(coef_gr, 'a'),
  b       = extr_value(coef_gr, 'b'),
  fl_b0   = extr_value(coef_fl, 'b0'),
  fl_b1   = extr_value(coef_fl, 'logsize_t0'),
  fl_b2   = extr_value(coef_fl, 'logsize_t0_2'),
  fl_b3   = extr_value(coef_fl, 'logsize_t0_3'),
  fl_bf   = extr_value(coef_fl, 'disturbance'),
  fr_b0   = extr_value(coef_fr, 'b0'),
  fr_b1   = extr_value(coef_fr, 'logsize_t0'),
  fr_b2   = extr_value(coef_fr, 'logsize_t0_2'),
  fr_b3   = extr_value(coef_fr, 'logsize_t0_3'),
  fr_bf   = extr_value(coef_fr, 'disturbance'),
  fr2re_b0 = 0,
  fr2re_b1 = unname(fr2re_site_fixef["logfruits_site_t0"]),
  recr_sz = extr_value(coef_misc, 'rec_siz'),
  recr_sd = extr_value(coef_misc, 'rec_sd'),
  L       = extr_value(coef_misc, 'min_siz'),
  U       = extr_value(coef_misc, 'max_siz'),
  mat_siz = 200,
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_fl_index = v_mod_fl_index,
  mod_fr_index = v_mod_fr_index,
  mod_fr2re_site_index = v_mod_fr2re_site_index))
```



## By-site IPM functions

The mean IPM already defined survival, growth, flowering, fruiting, recruitment size, and the IPM kernel. Here we only redefine the fertility part so that recruitment is controlled by the site-specific fruit-to-recruit transition instead of the mean recruits-per-fruit scalar.

This means all chunks below use the by-site fertility kernel. If you need the original mean-IPM fertility kernel again, re-run the mean-IPM file.

```{r by_site_ipm_functions}
fruit_x <- function(x, pars) {
  fl_x(x, pars) * fr_x(x, pars)
}

make_initial_n_site <- function(pars, site_i, df_init = df) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  breaks <- seq(L, U, length.out = n + 1)
  
  years_i <- df_init %>%
    filter(site == site_i) %>%
    distinct(year) %>%
    arrange(year) %>%
    pull(year)
  
  if (length(years_i) == 0) {
    return(rep(0, n))
  }
  
  counts_by_year <- purrr::map(
    years_i,
    function(year_i) {
      
      x <- df_init %>%
        filter(
          site == site_i,
          year == year_i,
          !is.na(logsize_t0),
          is.finite(logsize_t0)
        ) %>%
        pull(logsize_t0)
      
      if (length(x) == 0) {
        return(rep(0, n))
      }
      
      hist(
        pmin(pmax(x, L), U),
        breaks = breaks,
        plot = FALSE,
        include.lowest = TRUE
      )$counts
    }
  )
  
  adult_counts_mean <- Reduce(`+`, counts_by_year) / length(counts_by_year)
  
  adult_density <- adult_counts_mean / h
  
  adult_density
}

expected_site_fruits <- function(pars, site_i, df_init = df) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  n_site <- make_initial_n_site(
    pars = pars, site_i = site_i, df_init = df_init)
  
  # n_site is a density, so n_site * h gives counts per size bin
  sum(fruit_x(y, pars) * n_site * h, na.rm = TRUE)
}

predict_site_recruits_from_fruits <- function(site_fruits, pars) {
  
  if (is.na(site_fruits) || site_fruits <= 0) {
    return(0)
  }
  
  pred_log_recruits <- pars$fr2re_b1 * log1p(site_fruits)
  
  # Model was fit to log1p(recruits), so backtransform with expm1()
  pred_recruits <- expm1(pred_log_recruits)
  
  pmax(pred_recruits, 0)
}

site_recruits_per_fruit <- function(pars) {
  site_fruits <- pars$site_fruits_ref
  if (is.null(site_fruits) || is.na(site_fruits) || site_fruits <= 0) {
    return(0)
  }
  site_recruits <- predict_site_recruits_from_fruits(
    site_fruits = site_fruits,
    pars = pars)
  site_recruits / site_fruits
}

fyx <- function(y, x, pars) {
  
  fruit_x(x, pars) *
    site_recruits_per_fruit(pars) *
    re_y_dist(y, pars)
}
```

## Site-specific disturbance frequency

For the site-level IPM, disturbance is summarized as the proportion of years in which each site experienced disturbance. This gives one disturbance-regime kernel per site.

```{r by_site_disturbance_frequency}
p_disturbance_site <- df %>%
  group_by(site, year) %>%
  summarise(
    disturbance = max(disturbance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(site) %>%
  summarise(
    p_disturbance = mean(disturbance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(site = factor(site, levels = levels(df$site)))

p_disturbance_site
```

```{r by_site_apply_disturbance}
apply_site_disturbance <- function(p, p_disturbance) {
  
  p2 <- p
  
  if (!is.null(p$surv_bf)) {
    p2$surv_b0 <- p$surv_b0 + p$surv_bf * p_disturbance
  }
  
  if (!is.null(p$grow_bf)) {
    p2$grow_b0 <- p$grow_b0 + p$grow_bf * p_disturbance
  }
  
  if (!is.null(p$fl_bf)) {
    p2$fl_b0 <- p$fl_b0 + p$fl_bf * p_disturbance
  }
  
  if (!is.null(p$fr_bf)) {
    p2$fr_b0 <- p$fr_b0 + p$fr_bf * p_disturbance
  }
  
  p2
}
```

## Build one parameter object per site

Each site receives:

1. Its site-specific fruit-to-recruit coefficients.
2. Its observed disturbance frequency.
3. Its average annual observed size distribution.
4. A site-specific recruits-per-fruit scalar derived from expected site-level fruit production.

```{r by_site_parameter_objects}
pars_by_site <- df_fr2re_site_coef %>%
  mutate(site = factor(site, levels = levels(df$site))) %>%
  left_join(p_disturbance_site, by = "site") %>%
  mutate(
    p_disturbance = replace_na(p_disturbance, 0),

    site_fruits_ref = purrr::map2_dbl(
      site,
      p_disturbance,
      function(site_i, p_dist_i) {
        p <- apply_site_disturbance(
          p = pars,
          p_disturbance = p_dist_i
        )

        expected_site_fruits(
          pars = p,
          site_i = site_i,
          df_init = df
        )
      }
    ),

    site_recruits_ref = purrr::pmap_dbl(
      list(fr2re_b0_site, fr2re_b1_site, site_fruits_ref, p_disturbance),
      function(fr2re_b0_site, fr2re_b1_site, site_fruits_ref, p_disturbance) {
        p <- apply_site_disturbance(
          p = pars,
          p_disturbance = p_disturbance
        )

        p$fr2re_b0 <- fr2re_b0_site
        p$fr2re_b1 <- fr2re_b1_site
        p$site_fruits_ref <- site_fruits_ref

        predict_site_recruits_from_fruits(
          site_fruits = site_fruits_ref,
          pars = p
        )
      }
    ),

    recruits_per_fruit_ref = if_else(
      site_fruits_ref > 0,
      site_recruits_ref / site_fruits_ref,
      0
    ),

    pars_site = purrr::pmap(
      list(site, fr2re_b0_site, fr2re_b1_site, site_fruits_ref, p_disturbance),
      function(site, fr2re_b0_site, fr2re_b1_site, site_fruits_ref, p_disturbance) {
        p <- apply_site_disturbance(
          p = pars,
          p_disturbance = p_disturbance
        )

        p$site <- site
        p$p_disturbance <- p_disturbance
        p$fr2re_b0 <- fr2re_b0_site
        p$fr2re_b1 <- fr2re_b1_site
        p$site_fruits_ref <- site_fruits_ref

        p
      }
    )
  )

pars_by_site %>%
  select(
    site,
    p_disturbance,
    fr2re_b0_site,
    fr2re_b1_site,
    site_fruits_ref,
    site_recruits_ref,
    recruits_per_fruit_ref
  ) %>%
  print(n = 100)
```

## Project site-specific IPMs

For each site, we calculate asymptotic lambda from that site's kernel and projected lambda from that site's average annual observed size distribution.

```{r by_site_project_ipm}
project_site_ipm <- function(p, df_init = df) {
  n_obs <- make_initial_n_site(
    pars = p,
    site_i = p$site,
    df_init = df_init
  )

  K <- kernel(p)$k_yx
  h <- (p$U - p$L) / p$mat_siz
  n_proj <- K %*% n_obs

  tibble(
    lambda_asymptotic = Re(eigen(K)$values[1]),
    lambda_projected = as.numeric((sum(n_proj) * h) / (sum(n_obs) * h)),
    n_initial = sum(n_obs) * h,
    n_projected = sum(n_proj) * h,
    p_disturbance = p$p_disturbance,
    site_fruits_ref = p$site_fruits_ref,
    recruits_per_fruit_ref = site_recruits_per_fruit(p)
  )
}

df_lambda_model <- pars_by_site %>%
  transmute(
    site,
    lambda_data = purrr::map(pars_site, project_site_ipm)
  ) %>%
  tidyr::unnest(lambda_data)

df_lambda_model %>%
  print(n = 100)
```

## Observed population growth by site

Observed growth is calculated from the same adult definition used in the IPM: plants with finite `logsize_t0`.

```{r by_site_observed_growth}
df_obs_counts_site_year <- df %>%
  filter(
    !is.na(logsize_t0),
    is.finite(logsize_t0)
  ) %>%
  group_by(site, year) %>%
  summarise(
    n_adults = n(),
    .groups = "drop"
  ) %>%
  arrange(site, year) %>%
  group_by(site) %>%
  mutate(
    n_adults_t1 = lead(n_adults),
    year_t1 = lead(year),
    year_gap = year_t1 - year,
    lambda_obs_year = n_adults_t1 / n_adults
  ) %>%
  ungroup() %>%
  filter(
    year_gap == 1,
    n_adults > 0,
    n_adults_t1 > 0
  )

df_lambda_obs <- df_obs_counts_site_year %>%
  group_by(site) %>%
  summarise(
    lambda_obs_arithmetic = mean(lambda_obs_year, na.rm = TRUE),
    lambda_obs_geometric = exp(mean(log(lambda_obs_year), na.rm = TRUE)),
    n_year_transitions = n(),
    mean_n_adults = mean(n_adults, na.rm = TRUE),
    .groups = "drop"
  )

df_lambda_obs %>%
  print(n = 100)
```

## Compare modeled and observed site growth

The main comparison is projected lambda against observed geometric lambda. Projected lambda uses each site's average observed size distribution, while asymptotic lambda is the dominant eigenvalue of the site-specific kernel.

```{r by_site_compare_lambda}
df_lambda_compare <- df_lambda_model %>%
  left_join(df_lambda_obs, by = "site") %>%
  mutate(
    error_asymptotic_vs_obs = lambda_asymptotic - lambda_obs_geometric,
    error_projected_vs_obs = lambda_projected - lambda_obs_geometric
  )

df_lambda_compare %>%
  select(
    site,
    p_disturbance,
    lambda_asymptotic,
    lambda_projected,
    lambda_obs_geometric,
    error_projected_vs_obs,
    n_year_transitions,
    mean_n_adults,
    site_fruits_ref,
    recruits_per_fruit_ref
  ) %>%
  arrange(site) %>%
  print(n = 100, width = Inf)
```

```{r by_site_fit_statistics}
df_lambda_fit <- df_lambda_compare %>%
  summarise(
    cor_asymptotic = cor(lambda_asymptotic, lambda_obs_geometric, use = "complete.obs"),
    cor_projected = cor(lambda_projected, lambda_obs_geometric, use = "complete.obs"),
    rmse_asymptotic = sqrt(mean((lambda_asymptotic - lambda_obs_geometric)^2, na.rm = TRUE)),
    rmse_projected = sqrt(mean((lambda_projected - lambda_obs_geometric)^2, na.rm = TRUE)),
    mae_asymptotic = mean(abs(lambda_asymptotic - lambda_obs_geometric), na.rm = TRUE),
    mae_projected = mean(abs(lambda_projected - lambda_obs_geometric), na.rm = TRUE),
    bias_asymptotic = mean(lambda_asymptotic - lambda_obs_geometric, na.rm = TRUE),
    bias_projected = mean(lambda_projected - lambda_obs_geometric, na.rm = TRUE)
  )

df_lambda_fit
```

```{r by_site_lambda_plot, fig.width=8, fig.height=5}
df_lambda_compare_plot <- df_lambda_compare %>%
  select(
    site,
    lambda_obs_geometric,
    lambda_asymptotic,
    lambda_projected
  ) %>%
  pivot_longer(
    cols = c(lambda_asymptotic, lambda_projected),
    names_to = "modeled_lambda_type",
    values_to = "modeled_lambda"
  ) %>%
  mutate(
    modeled_lambda_type = recode(
      modeled_lambda_type,
      lambda_asymptotic = "Asymptotic lambda",
      lambda_projected = "Projected lambda"
    )
  )

fig_lambda_compare <- ggplot(
  df_lambda_compare_plot,
  aes(x = modeled_lambda, y = lambda_obs_geometric)
) +
  geom_point(size = 2.5) +
  geom_text(aes(label = site), nudge_y = 0.015, size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~ modeled_lambda_type) +
  theme_bw() +
  labs(
    title = "Observed versus modeled population growth",
    subtitle = "By-site IPM with site-specific disturbance and fruit-to-recruit transition",
    x = expression("Modeled " * lambda),
    y = expression("Observed geometric " * lambda)
  )

fig_lambda_compare
```
