# by Year by Site {#by_year_by_site}

This section builds one site-year IPM for each observed site-year transition. It
uses year-specific vital-rate parameters, a site-year fruit-to-recruit transition,
and the observed site-year disturbance exposure.

The fruit-to-recruit model is forced through zero on the `log1p` scale:

```r
log1p(recruits) = b1 * log1p(fruits)
```

Therefore, if fruits are zero, predicted recruits are also zero.

## Required objects

This section assumes that the data section has already created `df`.

```{r by_year_by_site_required_objects}
required_objects <- c("df", "v_ggp_suffix")
missing_objects <- required_objects[!purrr::map_lgl(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop(
    "The following objects are missing. Run the earlier files first: ",
    paste(missing_objects, collapse = ", "))
}

if (!exists("v_mod_set_su")) v_mod_set_su <- c()
if (!exists("v_mod_set_gr")) v_mod_set_gr <- c()
if (!exists("v_mod_set_fl")) v_mod_set_fl <- c()
if (!exists("v_mod_set_fr")) v_mod_set_fr <- c()
if (!exists("v_mod_set_fr2re_site")) v_mod_set_fr2re_site <- c()

ctrl_glmer <- lme4::glmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lme4::lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5))
```

## Year-specific vital-rate models

These models are fitted with year random effects. The disturbance coefficient is
kept as a fixed effect.

### Survival

```{r by_year_by_site_survival_model}
df_su <- df %>%
  filter(
    size_t0 != 0,
    !is.na(survives),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)

mod_su_0 <- lme4::glmer(
  survives ~ disturbance + (1 | year),
  data = df_su,
  family = binomial,
  control = ctrl_glmer)

mod_su_1 <- lme4::glmer(
  survives ~ logsize_t0 + disturbance + (1 | year),
  data = df_su,
  family = binomial,
  control = ctrl_glmer)

mod_su_2 <- lme4::glmer(
  survives ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_su,
  family = binomial,
  control = ctrl_glmer)

mod_su_3 <- lme4::glmer(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 +
    disturbance + (1 | year),
  data = df_su,
  family = binomial,
  control = ctrl_glmer)

mods_su <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
summary(mod_su_best)
mods_su_dAIC
```

### Growth

```{r by_year_by_site_growth_model}
df_gr <- df %>%
  filter(
    size_t0 != 0,
    survives == 1,
    is.finite(logsize_t0),
    is.finite(logsize_t1),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)

mod_gr_0 <- lme4::lmer(
  logsize_t1 ~ disturbance + (logsize_t0 | year),
  data = df_gr,
  control = ctrl_lmer)

mod_gr_1 <- lme4::lmer(
  logsize_t1 ~ logsize_t0 + disturbance + (logsize_t0 | year),
  data = df_gr,
  control = ctrl_lmer)

mod_gr_2 <- lme4::lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + disturbance +
    (logsize_t0 | year),
  data = df_gr,
  control = ctrl_lmer)

mod_gr_3 <- lme4::lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 +
    disturbance + (logsize_t0 | year),
  data = df_gr,
  control = ctrl_lmer)

mods_gr <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
summary(mod_gr_best)
mods_gr_dAIC
```

```{r by_year_by_site_growth_variance}
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

### Flowering

```{r by_year_by_site_flowering_model}
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance),
    flower = if_else(flower > 0, 1, 0)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)

mod_fl_0 <- lme4::glmer(
  flower ~ disturbance + (1 | year),
  data = df_fl,
  family = binomial,
  control = ctrl_glmer)

mod_fl_1 <- lme4::glmer(
  flower ~ logsize_t0 + disturbance + (1 | year),
  data = df_fl,
  family = binomial,
  control = ctrl_glmer)

mod_fl_2 <- lme4::glmer(
  flower ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fl,
  family = binomial,
  control = ctrl_glmer)

mod_fl_3 <- lme4::glmer(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 +
    disturbance + (1 | year),
  data = df_fl,
  family = binomial,
  control = ctrl_glmer)

mods_fl <- list(mod_fl_0, mod_fl_1, mod_fl_2, mod_fl_3)
mods_fl_dAIC <- bbmle::AICctab(mods_fl, weights = TRUE, sort = FALSE)$dAIC
mods_fl_sorted <- order(mods_fl_dAIC)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_best <- mods_fl[[mod_fl_index_bestfit]]
summary(mod_fl_best)
mods_fl_dAIC
```

### Fruiting

The fruit model is conditional on flowering plants.

```{r by_year_by_site_fruiting_model}
df_fr <- df %>%
  filter(
    flower > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, fruit,
    logsize_t0, logsize_t0_2, logsize_t0_3,
    disturbance)

mod_fr_0 <- lme4::glmer.nb(
  fruit ~ disturbance + (1 | year),
  data = df_fr,
  control = ctrl_glmer)

mod_fr_1 <- lme4::glmer.nb(
  fruit ~ logsize_t0 + disturbance + (1 | year),
  data = df_fr,
  control = ctrl_glmer)

mod_fr_2 <- lme4::glmer.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fr,
  control = ctrl_glmer)

mods_fr <- list(mod_fr_0, mod_fr_1, mod_fr_2)
mods_fr_dAIC <- bbmle::AICctab(mods_fr, weights = TRUE, sort = FALSE)$dAIC
mods_fr_i_sort <- order(mods_fr_dAIC)

if (length(v_mod_set_fr) == 0) {
  mod_fr_i_best <- mods_fr_i_sort[1]
  v_mod_fr_i <- mod_fr_i_best - 1
} else {
  mod_fr_i_best <- v_mod_set_fr + 1
  v_mod_fr_i <- v_mod_set_fr
}

mod_fr_best <- mods_fr[[mod_fr_i_best]]
summary(mod_fr_best)
mods_fr_dAIC
```

## Site-year fruit-to-recruit transition

Fruits at site `i` in year `t` are matched to recruits at the same site in year
`t + 1`.

```{r by_year_by_site_fr2re_data}
df_fr2re_site_year <- df %>%
  mutate(
    site = factor(site),
    year = as.integer(as.character(year))) %>%
  group_by(site, year) %>%
  summarise(
    fruits_site_t0 = sum(fruit, na.rm = TRUE),
    recruits_site_t0 = sum(recruit == 1, na.rm = TRUE),
    .groups = "drop") %>%
  arrange(site, year) %>%
  group_by(site) %>%
  mutate(
    recruits_site_t1 = lead(recruits_site_t0),
    year_t1 = lead(year),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1) %>%
  mutate(
    year_t0 = factor(year),
    year_t1 = factor(year_t1),
    site = factor(site),
    logfruits_site_t0 = log1p(fruits_site_t0),
    logrecruits_site_t1 = log1p(recruits_site_t1))

df_fr2re_site_year %>%
  summarise(
    n_site_year_transitions = n(),
    n_sites = n_distinct(site),
    n_year_t1 = n_distinct(year_t1),
    n_positive_recruit_transitions = sum(recruits_site_t1 > 0),
    n_zero_recruit_transitions = sum(recruits_site_t1 == 0),
    total_fruits_t0 = sum(fruits_site_t0, na.rm = TRUE),
    total_recruits_t1 = sum(recruits_site_t1, na.rm = TRUE),
    max_fruits_site_t0 = max(fruits_site_t0, na.rm = TRUE),
    max_recruits_site_t1 = max(recruits_site_t1, na.rm = TRUE),
    proportion_zero_recruit = mean(recruits_site_t1 == 0, na.rm = TRUE))
```

## Candidate fruit-to-recruit models

All candidate models are forced through zero. Only the slope can vary by site
and/or year.

```{r by_year_by_site_fr2re_models}
ctrl_re <- lme4::lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 2e5))

fr2re_mod_01 <- lm(
  logrecruits_site_t1 ~ 0 + logfruits_site_t0,
  data = df_fr2re_site_year)

fr2re_mod_02 <- lme4::lmer(
  logrecruits_site_t1 ~ 0 + logfruits_site_t0 +
    (0 + logfruits_site_t0 | site),
  data = df_fr2re_site_year,
  REML = FALSE,
  control = ctrl_re)

fr2re_mod_03 <- lme4::lmer(
  logrecruits_site_t1 ~ 0 + logfruits_site_t0 +
    (0 + logfruits_site_t0 | year_t1),
  data = df_fr2re_site_year,
  REML = FALSE,
  control = ctrl_re)

fr2re_mod_04 <- lme4::lmer(
  logrecruits_site_t1 ~ 0 + logfruits_site_t0 +
    (0 + logfruits_site_t0 | site) +
    (0 + logfruits_site_t0 | year_t1),
  data = df_fr2re_site_year,
  REML = FALSE,
  control = ctrl_re)

fr2re_mods <- list(
  fr2re_mod_01,
  fr2re_mod_02,
  fr2re_mod_03,
  fr2re_mod_04)

fr2re_mods_dAIC <- bbmle::AICctab(
  fr2re_mods,
  weights = TRUE,
  sort = FALSE)$dAIC

fr2re_mods_i_sort <- order(fr2re_mods_dAIC)

if (length(v_mod_set_fr2re_site) == 0) {
  fr2re_mod_i_best <- fr2re_mods_i_sort[1]
  v_mod_fr2re_site_index <- fr2re_mod_i_best
} else {
  if (!v_mod_set_fr2re_site %in% seq_along(fr2re_mods)) {
    stop("v_mod_set_fr2re_site must be between 1 and ", length(fr2re_mods))
  }
  fr2re_mod_i_best <- v_mod_set_fr2re_site
  v_mod_fr2re_site_index <- v_mod_set_fr2re_site
}

fr2re_mod_best <- fr2re_mods[[fr2re_mod_i_best]]
summary(fr2re_mod_best)
fr2re_mods_dAIC
```

```{r by_year_by_site_fr2re_diagnostics}
fr2re_model_diagnostics <- tibble(
  model_id = seq_along(fr2re_mods),
  dAIC = as.numeric(fr2re_mods_dAIC),
  singular = purrr::map_lgl(
    fr2re_mods,
    ~ if (inherits(.x, "merMod")) lme4::isSingular(.x) else FALSE),
  convergence_message = purrr::map_chr(
    fr2re_mods,
    function(m) {
      if (!inherits(m, "merMod")) return(NA_character_)
      msg <- summary(m)$optinfo$conv$lme4$messages
      if (is.null(msg)) return(NA_character_)
      paste(msg, collapse = "; ")
    }))

fr2re_model_diagnostics
```

## Extract fixed and random effects

```{r by_year_by_site_extract_effects}
get_fixef_re <- function(model) {
  if (inherits(model, "merMod")) {
    lme4::fixef(model)
  } else {
    coef(model)
  }
}

fr2re_fixef <- get_fixef_re(fr2re_mod_best)
fr2re_b1_mean <- unname(fr2re_fixef["logfruits_site_t0"])

get_ranef_re <- function(model, group_var, levels_vec) {
  
  if (!inherits(model, "merMod")) {
    out <- tibble(!!group_var := levels_vec)
  } else if (!group_var %in% names(lme4::ranef(model))) {
    out <- tibble(!!group_var := levels_vec)
  } else {
    out <- lme4::ranef(model)[[group_var]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column(group_var)
  }
  
  if (!"(Intercept)" %in% names(out)) {
    out[["(Intercept)"]] <- 0
  }
  
  if (!"logfruits_site_t0" %in% names(out)) {
    out[["logfruits_site_t0"]] <- 0
  }
  
  out %>%
    mutate(
      across(c(`(Intercept)`, logfruits_site_t0), ~ replace_na(.x, 0)))
}

fr2re_site_ranef <- get_ranef_re(
  model = fr2re_mod_best,
  group_var = "site",
  levels_vec = levels(df_fr2re_site_year$site)) %>%
  rename(
    site_u0 = `(Intercept)`,
    site_u1 = logfruits_site_t0)

fr2re_year_ranef <- get_ranef_re(
  model = fr2re_mod_best,
  group_var = "year_t1",
  levels_vec = levels(df_fr2re_site_year$year_t1)) %>%
  rename(
    year_u0 = `(Intercept)`,
    year_u1 = logfruits_site_t0)

df_fr2re_site_year_coef <- tidyr::expand_grid(
  site = levels(df_fr2re_site_year$site),
  year_t1 = levels(df_fr2re_site_year$year_t1)) %>%
  mutate(
    site = factor(site, levels = levels(df_fr2re_site_year$site)),
    year_t1 = factor(year_t1, levels = levels(df_fr2re_site_year$year_t1))) %>%
  left_join(fr2re_site_ranef, by = "site") %>%
  left_join(fr2re_year_ranef, by = "year_t1") %>%
  mutate(
    site_u0 = replace_na(site_u0, 0),
    site_u1 = replace_na(site_u1, 0),
    year_u0 = replace_na(year_u0, 0),
    year_u1 = replace_na(year_u1, 0),
    fr2re_b0_site_year = 0,
    fr2re_b1_site_year = fr2re_b1_mean + site_u1 + year_u1) %>%
  select(
    site,
    year_t1,
    fr2re_b0_site_year,
    fr2re_b1_site_year,
    site_u0,
    site_u1,
    year_u0,
    year_u1)

df_fr2re_site_year_coef %>%
  summarise(
    min_b0 = min(fr2re_b0_site_year),
    max_b0 = max(fr2re_b0_site_year),
    mean_b1 = mean(fr2re_b1_site_year),
    min_b1 = min(fr2re_b1_site_year),
    max_b1 = max(fr2re_b1_site_year))
```

## Fruit-to-recruit diagnostic plot

```{r by_year_by_site_fr2re_prediction_plot, warning=FALSE}
predict_site_year_recruits <- function(logfruits_site_t0, fr2re_b0, fr2re_b1) {
  pred_log_recruits <- fr2re_b0 + fr2re_b1 * logfruits_site_t0
  pred_recruits <- expm1(pred_log_recruits)
  pmax(pred_recruits, 0)
}

df_fr2re_site_year_plot <- df_fr2re_site_year %>%
  left_join(df_fr2re_site_year_coef, by = c("site", "year_t1")) %>%
  mutate(
    pred_recruits_site_t1 = predict_site_year_recruits(
      logfruits_site_t0 = logfruits_site_t0,
      fr2re_b0 = fr2re_b0_site_year,
      fr2re_b1 = fr2re_b1_site_year))

fig_fr2re_site_year <- ggplot(
  df_fr2re_site_year_plot,
  aes(x = fruits_site_t0, y = recruits_site_t1)
) +
  geom_jitter(width = 0, height = 0.08, alpha = 0.45, size = 1) +
  geom_point(
    aes(y = pred_recruits_site_t1),
    colour = "red",
    alpha = 0.6,
    size = 1.2) +
  facet_wrap(~ site, scales = "free_y") +
  scale_x_continuous(
    trans = scales::pseudo_log_trans(sigma = 1),
    name = "Total fruits at site in year t") +
  theme_bw() +
  labs(
    title = "Site-year fruit-to-recruit transition",
    subtitle = "Observed recruits and model predictions by site",
    y = "Total recruits at site in year t + 1")

fig_fr2re_site_year
```

## Export parameter objects

```{r by_year_by_site_parameter_helpers}
empty_coef_df <- function() {
  data.frame(coefficient = character(), value = numeric())
}

is_mixed_model <- function(model) {
  inherits(model, "merMod")
}

get_fixef_safe <- function(model) {
  if (is_mixed_model(model)) {
    lme4::fixef(model)
  } else {
    coef(model)
  }
}

get_group_coef_safe <- function(model, group_var) {
  if (!is_mixed_model(model)) {
    return(NULL)
  }
  coef_list <- coef(model)
  if (!(group_var %in% names(coef_list))) {
    return(NULL)
  }
  coef_list[[group_var]]
}

find_model_term <- function(model_terms, aliases) {
  aliases <- unlist(aliases)
  for (alias in aliases) {
    if (grepl("^regex:", alias)) {
      pattern <- sub("^regex:", "", alias)
      hit <- grep(pattern, model_terms, value = TRUE)
    } else {
      hit <- model_terms[model_terms == alias]
    }
    if (length(hit) > 0) {
      return(hit[1])
    }
  }
  NA_character_
}

bind_coef_rows <- function(x) {
  x <- x[!vapply(x, is.null, logical(1))]
  if (length(x) == 0) {
    return(empty_coef_df())
  }
  out <- do.call(rbind, x)
  rownames(out) <- NULL
  out
}

check_duplicate_coefficients <- function(x, object_name = "parameter table") {
  dup <- x$coefficient[duplicated(x$coefficient)]
  if (length(dup) > 0) {
    stop(
      object_name,
      " contains duplicated coefficient names: ",
      paste(unique(dup), collapse = ", "))
  }
  invisible(TRUE)
}

coef_df_to_list <- function(x) {
  if (nrow(x) == 0) {
    return(list())
  }
  check_duplicate_coefficients(x)
  x %>%
    mutate(coefficient = as.character(coefficient)) %>%
    tidyr::pivot_wider(names_from = coefficient, values_from = value) %>%
    as.list()
}

extract_fixed_pars <- function(model, term_map, fill_missing = TRUE) {
  model_coefs <- get_fixef_safe(model)
  model_terms <- names(model_coefs)
  out <- lapply(names(term_map), function(coef_name) {
    model_term <- find_model_term(
      model_terms = model_terms,
      aliases = term_map[[coef_name]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- 0
      } else {
        return(NULL)
      }
    } else {
      value <- unname(model_coefs[model_term])
    }
    data.frame(coefficient = coef_name, value = value)
  })
  bind_coef_rows(out)
}

extract_year_pars <- function(model, group_var, term_map, fill_missing = TRUE) {
  coef_matrix <- get_group_coef_safe(model = model, group_var = group_var)
  if (is.null(coef_matrix)) {
    return(empty_coef_df())
  }
  model_terms <- colnames(coef_matrix)
  out <- lapply(names(term_map), function(coef_prefix) {
    model_term <- find_model_term(
      model_terms = model_terms,
      aliases = term_map[[coef_prefix]])
    if (is.na(model_term)) {
      if (fill_missing) {
        value <- rep(0, nrow(coef_matrix))
      } else {
        return(NULL)
      }
    } else {
      value <- coef_matrix[, model_term]
    }
    data.frame(
      coefficient = paste0(coef_prefix, "_", rownames(coef_matrix)),
      value = value)
  })
  bind_coef_rows(out)
}
```

```{r by_year_by_site_parameter_objects}
size_term_map <- list(
  b0 = c("(Intercept)"),
  b1 = c("logsize_t0"),
  b2 = c("logsize_t0_2", "I(logsize_t0^2)"),
  b3 = c("logsize_t0_3", "I(logsize_t0^3)"),
  bf = c("disturbance", "regex:^disturbance"))

make_size_term_map <- function(prefix) {
  out <- size_term_map
  names(out) <- paste0(prefix, names(out))
  out
}

su_term_map <- make_size_term_map("surv_")
gr_term_map <- make_size_term_map("grow_")
fl_term_map <- make_size_term_map("fl_")
fr_term_map <- make_size_term_map("fr_")

su_fe <- extract_fixed_pars(mod_su_best, su_term_map, fill_missing = TRUE)
gr_fe <- extract_fixed_pars(mod_gr_best, gr_term_map, fill_missing = TRUE)
fl_fe <- extract_fixed_pars(mod_fl_best, fl_term_map, fill_missing = TRUE)
fr_fe <- extract_fixed_pars(mod_fr_best, fr_term_map, fill_missing = TRUE)

df_re_size <- df %>%
  filter(
    recruit == 1,
    size_t0 > 0,
    is.finite(logsize_t0))

recr_sz_mean <- mean(df_re_size$logsize_t0, na.rm = TRUE)
recr_sz_sd <- sd(df_re_size$logsize_t0, na.rm = TRUE)
gr_var_coef <- coef(mod_gr_var)

constants <- tibble::tribble(
  ~coefficient, ~value,
  "recr_sz", recr_sz_mean,
  "recr_sd", recr_sz_sd,
  "a", as.numeric(gr_var_coef[1]),
  "b", as.numeric(gr_var_coef[2]),
  "L", min(df_gr$logsize_t0, na.rm = TRUE),
  "U", max(df_gr$logsize_t0, na.rm = TRUE),
  "mat_siz", 200,
  "mod_su_index", v_mod_su_index,
  "mod_gr_index", v_mod_gr_index,
  "mod_fl_index", v_mod_fl_index,
  "mod_fr_index", v_mod_fr_i,
  "mod_fr2re_site_year_index", v_mod_fr2re_site_index) %>%
  mutate(coefficient = as.character(coefficient), value = as.numeric(value))

fr2re_fe <- tibble::tribble(
  ~coefficient, ~value,
  "fr2re_b0", 0,
  "fr2re_b1", fr2re_b1_mean) %>%
  mutate(coefficient = as.character(coefficient), value = as.numeric(value))

pars_cons <- bind_coef_rows(
  list(su_fe, gr_fe, fl_fe, fr_fe, fr2re_fe, constants))

check_duplicate_coefficients(pars_cons, object_name = "pars_cons")
pars_mean <- coef_df_to_list(pars_cons)

su_out_yr <- extract_year_pars(mod_su_best, "year", su_term_map, TRUE)
gr_out_yr <- extract_year_pars(mod_gr_best, "year", gr_term_map, TRUE)
fl_out_yr <- extract_year_pars(mod_fl_best, "year", fl_term_map, TRUE)
fr_out_yr <- extract_year_pars(mod_fr_best, "year", fr_term_map, TRUE)

pars_var <- bind_coef_rows(list(su_out_yr, gr_out_yr, fl_out_yr, fr_out_yr))
check_duplicate_coefficients(pars_var, object_name = "pars_var")
pars_var_wide <- coef_df_to_list(pars_var)

make_year_pars <- function(year_i, pars_mean, pars_var_wide) {
  p <- pars_mean
  yr <- as.character(year_i)
  possible_names <- names(pars_var_wide)
  for (nm in possible_names) {
    suffix <- paste0("_", yr)
    if (endsWith(nm, suffix)) {
      base_nm <- sub(paste0("_", yr, "$"), "", nm)
      p[[base_nm]] <- pars_var_wide[[nm]]
    }
  }
  p$year <- yr
  p
}

v_years_model <- sort(unique(as.character(df$year)))

pars_year <- tibble(
  year = v_years_model,
  pars_year = purrr::map(
    year,
    ~ make_year_pars(
      year_i = .x,
      pars_mean = pars_mean,
      pars_var_wide = pars_var_wide)))

pars_site_year_recruit <- df_fr2re_site_year_coef %>%
  mutate(
    site = factor(site),
    year_t1 = factor(year_t1),
    year = as.character(as.integer(as.character(year_t1)) - 1),
    fr2re_b0 = fr2re_b0_site_year,
    fr2re_b1 = fr2re_b1_site_year) %>%
  select(
    site,
    year,
    year_t1,
    fr2re_b0,
    fr2re_b1,
    site_u0,
    site_u1,
    year_u0,
    year_u1)
```

## Site-year IPM functions

```{r by_year_by_site_ipm_functions}
inv_logit <- function(x) {
  plogis(x)
}

apply_disturbance_state <- function(p, disturbance_value) {
  p2 <- p
  if (!is.null(p$surv_bf)) {
    p2$surv_b0 <- p$surv_b0 + p$surv_bf * disturbance_value
  }
  if (!is.null(p$grow_bf)) {
    p2$grow_b0 <- p$grow_b0 + p$grow_bf * disturbance_value
  }
  if (!is.null(p$fl_bf)) {
    p2$fl_b0 <- p$fl_b0 + p$fl_bf * disturbance_value
  }
  if (!is.null(p$fr_bf)) {
    p2$fr_b0 <- p$fr_b0 + p$fr_bf * disturbance_value
  }
  p2
}

sx <- function(x, pars, num_pars = pars$mod_su_index) {
  survival_value <- pars$surv_b0
  for (i in seq_len(num_pars)) {
    param_name <- paste0("surv_b", i)
    if (!is.null(pars[[param_name]])) {
      survival_value <- survival_value + pars[[param_name]] * x^i
    }
  }
  inv_logit(survival_value)
}

grow_mu <- function(x, pars, num_pars = pars$mod_gr_index) {
  mean_value <- 0
  for (i in 0:num_pars) {
    param_name <- paste0("grow_b", i)
    if (!is.null(pars[[param_name]])) {
      mean_value <- mean_value + pars[[param_name]] * x^i
    }
  }
  mean_value
}

grow_sd <- function(x, pars) {
  sqrt(pars$a * exp(pars$b * x))
}

gxy <- function(x, y, pars) {
  dnorm(y, mean = grow_mu(x, pars), sd = grow_sd(x, pars))
}

fl_x <- function(x, pars, num_pars = pars$mod_fl_index) {
  flower_value <- pars$fl_b0
  for (i in seq_len(num_pars)) {
    param_name <- paste0("fl_b", i)
    if (!is.null(pars[[param_name]])) {
      flower_value <- flower_value + pars[[param_name]] * x^i
    }
  }
  inv_logit(flower_value)
}

fr_x <- function(x, pars, num_pars = pars$mod_fr_index) {
  fruit_value <- pars$fr_b0
  for (i in seq_len(num_pars)) {
    param_name <- paste0("fr_b", i)
    if (!is.null(pars[[param_name]])) {
      fruit_value <- fruit_value + pars[[param_name]] * x^i
    }
  }
  exp(fruit_value)
}

fruit_x <- function(x, pars) {
  fl_x(x, pars) * fr_x(x, pars)
}

re_y_dist <- function(y, pars) {
  dens <- dnorm(y, mean = pars$recr_sz, sd = pars$recr_sd)
  norm_const <- pnorm(pars$U, mean = pars$recr_sz, sd = pars$recr_sd) -
    pnorm(pars$L, mean = pars$recr_sz, sd = pars$recr_sd)
  dens / norm_const
}

predict_site_year_recruits_from_fruits <- function(site_fruits, pars) {
  if (is.na(site_fruits) || site_fruits <= 0) {
    return(0)
  }
  pred_log_recruits <- pars$fr2re_b0 + pars$fr2re_b1 * log1p(site_fruits)
  pred_recruits <- expm1(pred_log_recruits)
  pmax(pred_recruits, 0)
}

site_year_recruits_per_fruit <- function(pars) {
  site_fruits <- pars$site_fruits_ref
  if (is.null(site_fruits) || is.na(site_fruits) || site_fruits <= 0) {
    return(0)
  }
  site_recruits <- predict_site_year_recruits_from_fruits(
    site_fruits = site_fruits,
    pars = pars)
  site_recruits / site_fruits
}

fyx <- function(y, x, pars) {
  fruit_x(x, pars) * site_year_recruits_per_fruit(pars) * re_y_dist(y, pars)
}

kernel <- function(pars) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  Smat <- sx(y, pars)
  Gmat <- matrix(0, n, n)
  Gmat[] <- t(outer(y, y, gxy, pars)) * h
  Tmat <- matrix(0, n, n)
  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }
  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }
  Fmat <- outer(
    y, y,
    Vectorize(function(x, y) fyx(x, y, pars))) * h
  k_yx <- Fmat + Tmat
  list(k_yx = k_yx, Fmat = Fmat, Tmat = Tmat, Gmat = Gmat, meshpts = y)
}

lambda_ipm <- function(pars) {
  Re(eigen(kernel(pars)$k_yx)$values[1])
}
```

## Site-year projections

```{r by_year_by_site_projection_helpers}
p_disturbance_site_year <- df %>%
  filter(
    !is.na(logsize_t0),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(site = factor(site), year = as.character(year)) %>%
  group_by(site, year) %>%
  summarise(p_disturbance = mean(disturbance, na.rm = TRUE), .groups = "drop")

make_initial_n_site_year <- function(pars, site_i, year_i, df_init = df) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  breaks <- seq(L, U, length.out = n + 1)
  x <- df_init %>%
    filter(
      site == site_i,
      as.character(year) == as.character(year_i),
      !is.na(logsize_t0),
      is.finite(logsize_t0)) %>%
    pull(logsize_t0)
  if (length(x) == 0) {
    return(rep(0, n))
  }
  adult_counts <- hist(
    pmin(pmax(x, L), U),
    breaks = breaks,
    plot = FALSE,
    include.lowest = TRUE)$counts
  adult_counts / h
}

expected_site_year_fruits <- function(pars, site_i, year_i, df_init = df) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  n_site_year <- make_initial_n_site_year(
    pars = pars,
    site_i = site_i,
    year_i = year_i,
    df_init = df_init)
  sum(fruit_x(y, pars) * n_site_year * h, na.rm = TRUE)
}

get_year_pars <- function(year_i) {
  out <- pars_year %>%
    filter(year == as.character(year_i)) %>%
    pull(pars_year)
  if (length(out) != 1) {
    stop("Could not find exactly one parameter object for year: ", year_i)
  }
  out[[1]]
}
```

```{r by_year_by_site_project_function}
project_site_year_ipm <- function(
    site_i,
    year_i,
    year_t1_i,
    p_disturbance_i,
    fr2re_b0_i,
    fr2re_b1_i,
    df_init = df
) {
  
  p_base <- get_year_pars(year_i)
  p_base$site <- site_i
  p_base$year <- as.character(year_i)
  p_base$year_t1 <- as.character(year_t1_i)
  p_base$p_disturbance <- p_disturbance_i
  
  p_no <- apply_disturbance_state(p = p_base, disturbance_value = 0)
  p_dist <- apply_disturbance_state(p = p_base, disturbance_value = 1)
  
  site_fruits_no <- expected_site_year_fruits(
    pars = p_no,
    site_i = site_i,
    year_i = year_i,
    df_init = df_init)
  
  site_fruits_dist <- expected_site_year_fruits(
    pars = p_dist,
    site_i = site_i,
    year_i = year_i,
    df_init = df_init)
  
  site_fruits_ref <-
    (1 - p_disturbance_i) * site_fruits_no +
    p_disturbance_i       * site_fruits_dist
  
  p_no$fr2re_b0 <- fr2re_b0_i
  p_no$fr2re_b1 <- fr2re_b1_i
  p_no$site_fruits_ref <- site_fruits_ref
  
  p_dist$fr2re_b0 <- fr2re_b0_i
  p_dist$fr2re_b1 <- fr2re_b1_i
  p_dist$site_fruits_ref <- site_fruits_ref
  
  n_obs <- make_initial_n_site_year(
    pars = p_base,
    site_i = site_i,
    year_i = year_i,
    df_init = df_init)
  
  h <- (p_base$U - p_base$L) / p_base$mat_siz
  n_initial <- sum(n_obs) * h
  
  if (n_initial <= 0) {
    return(
      tibble(
        lambda_asymptotic = NA_real_,
        lambda_projected = NA_real_,
        n_initial = n_initial,
        n_projected = NA_real_,
        p_disturbance = p_disturbance_i,
        site_fruits_no = site_fruits_no,
        site_fruits_dist = site_fruits_dist,
        site_fruits_ref = site_fruits_ref,
        site_recruits_ref = NA_real_,
        recruits_per_fruit_ref = NA_real_))
  }
  
  K_no <- kernel(p_no)$k_yx
  K_dist <- kernel(p_dist)$k_yx
  K_regime <-
    (1 - p_disturbance_i) * K_no +
    p_disturbance_i       * K_dist
  
  n_proj <- K_regime %*% n_obs
  site_recruits_ref <- predict_site_year_recruits_from_fruits(
    site_fruits = site_fruits_ref,
    pars = p_no)
  
  recruits_per_fruit_ref <- ifelse(
    site_fruits_ref > 0,
    site_recruits_ref / site_fruits_ref,
    0)
  
  tibble(
    lambda_asymptotic = Re(eigen(K_regime)$values[1]),
    lambda_projected = as.numeric((sum(n_proj) * h) / n_initial),
    n_initial = n_initial,
    n_projected = sum(n_proj) * h,
    p_disturbance = p_disturbance_i,
    site_fruits_no = site_fruits_no,
    site_fruits_dist = site_fruits_dist,
    site_fruits_ref = site_fruits_ref,
    site_recruits_ref = site_recruits_ref,
    recruits_per_fruit_ref = recruits_per_fruit_ref)
}
```

```{r by_year_by_site_run_projections}
df_site_year_projection_grid <- pars_site_year_recruit %>%
  mutate(
    site = factor(site),
    year = as.character(year),
    year_t1 = as.character(year_t1)) %>%
  left_join(p_disturbance_site_year, by = c("site", "year")) %>%
  mutate(p_disturbance = replace_na(p_disturbance, 0))

df_lambda_site_year <- df_site_year_projection_grid %>%
  mutate(
    lambda_data = purrr::pmap(
      list(site, year, year_t1, p_disturbance, fr2re_b0, fr2re_b1),
      function(site, year, year_t1, p_disturbance, fr2re_b0, fr2re_b1) {
        project_site_year_ipm(
          site_i = site,
          year_i = year,
          year_t1_i = year_t1,
          p_disturbance_i = p_disturbance,
          fr2re_b0_i = fr2re_b0,
          fr2re_b1_i = fr2re_b1,
          df_init = df)
      })) %>%
  select(site, year, year_t1, fr2re_b0, fr2re_b1, lambda_data) %>%
  tidyr::unnest(lambda_data)

df_lambda_site_year %>%
  print(n = 100)
```

## Observed site-year population growth

```{r by_year_by_site_observed_lambda}
df_obs_counts_site_year <- df %>%
  filter(
    !is.na(logsize_t0),
    is.finite(logsize_t0)) %>%
  mutate(site = factor(site), year = as.character(year)) %>%
  group_by(site, year) %>%
  summarise(n_obs = n_distinct(plant_id), .groups = "drop") %>%
  arrange(site, year) %>%
  group_by(site) %>%
  mutate(
    year_t1 = lead(year),
    n_obs_t1 = lead(n_obs),
    year_gap = as.integer(year_t1) - as.integer(year),
    lambda_obs = n_obs_t1 / n_obs) %>%
  ungroup() %>%
  filter(year_gap == 1, n_obs > 0, n_obs_t1 > 0)

df_lambda_site_year_compare <- df_lambda_site_year %>%
  left_join(
    df_obs_counts_site_year %>%
      select(site, year, year_t1, n_obs, n_obs_t1, lambda_obs),
    by = c("site", "year", "year_t1")) %>%
  mutate(
    error_asymptotic_vs_obs = lambda_asymptotic - lambda_obs,
    error_projected_vs_obs = lambda_projected - lambda_obs)

df_lambda_site_year_compare_summary <- df_lambda_site_year_compare %>%
  summarise(
    n_rows = n(),
    n_with_observed_lambda = sum(!is.na(lambda_obs)),
    n_missing_modeled_projected = sum(is.na(lambda_projected)),
    n_missing_observed = sum(is.na(lambda_obs)),
    mean_lambda_obs = mean(lambda_obs, na.rm = TRUE),
    geometric_lambda_obs = exp(mean(log(lambda_obs), na.rm = TRUE)),
    mean_lambda_asymptotic = mean(lambda_asymptotic, na.rm = TRUE),
    mean_lambda_projected = mean(lambda_projected, na.rm = TRUE),
    mean_error_asymptotic_vs_obs = mean(error_asymptotic_vs_obs, na.rm = TRUE),
    mean_error_projected_vs_obs = mean(error_projected_vs_obs, na.rm = TRUE),
    rmse_asymptotic_vs_obs = sqrt(mean(error_asymptotic_vs_obs^2, na.rm = TRUE)),
    rmse_projected_vs_obs = sqrt(mean(error_projected_vs_obs^2, na.rm = TRUE)))

df_lambda_site_year_compare_summary
```

## Investigation: observed lambda exactly one

This checks whether stable observed counts are caused by the same individuals
persisting or by losses being balanced by gains.

```{r by_year_by_site_lambda_one_turnover}
df_obs_turnover_site_year <- df %>%
  filter(
    !is.na(logsize_t0),
    is.finite(logsize_t0)) %>%
  mutate(year = as.character(year)) %>%
  group_by(site, year) %>%
  summarise(
    plants = list(unique(plant_id)),
    n_obs = n_distinct(plant_id),
    .groups = "drop") %>%
  arrange(site, year) %>%
  group_by(site) %>%
  mutate(
    year_t1 = lead(year),
    plants_t1 = lead(plants),
    n_obs_t1 = lead(n_obs),
    year_gap = as.integer(year_t1) - as.integer(year)) %>%
  ungroup() %>%
  filter(year_gap == 1) %>%
  rowwise() %>%
  mutate(
    n_same = length(intersect(plants, plants_t1)),
    n_lost = length(setdiff(plants, plants_t1)),
    n_gained = length(setdiff(plants_t1, plants)),
    lambda_obs = n_obs_t1 / n_obs) %>%
  ungroup()

df_obs_turnover_site_year %>%
  filter(lambda_obs == 1) %>%
  select(site, year, year_t1, n_obs, n_obs_t1, n_same, n_lost, n_gained) %>%
  arrange(site, year) %>%
  print(n = Inf, width = Inf)

df_obs_turnover_site_year %>%
  filter(lambda_obs == 1) %>%
  mutate(
    reason = case_when(
      n_lost == 0 & n_gained == 0 ~ "same individuals",
      n_lost == n_gained          ~ "losses balanced by gains",
      TRUE                        ~ "check")) %>%
  count(reason)
```

## Investigation: sample size by site-year

```{r by_year_by_site_sample_size}
df_sample_site_year <- df %>%
  filter(
    !is.na(logsize_t0),
    is.finite(logsize_t0)) %>%
  mutate(
    site = factor(site),
    year = as.integer(as.character(year))) %>%
  group_by(site, year) %>%
  summarise(
    n_plants = n_distinct(plant_id),
    n_records = n(),
    .groups = "drop") %>%
  arrange(site, year)

df_sample_site_year %>%
  group_by(site) %>%
  summarise(
    n_years = n(),
    min_n = min(n_plants, na.rm = TRUE),
    mean_n = mean(n_plants, na.rm = TRUE),
    median_n = median(n_plants, na.rm = TRUE),
    max_n = max(n_plants, na.rm = TRUE),
    .groups = "drop") %>%
  arrange(as.numeric(as.character(site)))
```

```{r by_year_by_site_sample_size_plot}
fig_sample_site_year <- ggplot(
  df_sample_site_year,
  aes(x = year, y = site, fill = n_plants)
) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  theme_bw() +
  labs(
    title = "Sample size by site and year",
    subtitle = "Number of distinct sized plants",
    x = "Year",
    y = "Site",
    fill = "N plants")

fig_sample_site_year
```

## Investigation: mean fruit-to-recruit ratio

This compares the baseline fruit-to-recruit model to the simple observed recruits
per fruit ratio.

```{r by_year_by_site_fr2re_mean_comparison}
df_fr2re_mean_compare <- df_fr2re_site_year %>%
  filter(fruits_site_t0 > 0) %>%
  mutate(
    obs_recruits_per_fruit = recruits_site_t1 / fruits_site_t0,
    pred_recruits_mean_model = expm1(fr2re_b1_mean * log1p(fruits_site_t0)),
    pred_recruits_per_fruit_mean_model =
      pred_recruits_mean_model / fruits_site_t0) %>%
  summarise(
    fr2re_b0 = 0,
    fr2re_b1_log1p = fr2re_b1_mean,
    simple_total_recruits_per_fruit =
      sum(recruits_site_t1, na.rm = TRUE) / sum(fruits_site_t0, na.rm = TRUE),
    mean_site_year_recruits_per_fruit =
      mean(obs_recruits_per_fruit, na.rm = TRUE),
    median_site_year_recruits_per_fruit =
      median(obs_recruits_per_fruit, na.rm = TRUE),
    model_total_recruits_per_fruit =
      sum(pred_recruits_mean_model, na.rm = TRUE) / sum(fruits_site_t0,
                                                        na.rm = TRUE),
    mean_model_site_year_recruits_per_fruit =
      mean(pred_recruits_per_fruit_mean_model, na.rm = TRUE),
    median_model_site_year_recruits_per_fruit =
      median(pred_recruits_per_fruit_mean_model, na.rm = TRUE),
    n_site_year = n())

df_fr2re_mean_compare

df_fr2re_mean_compare %>%
  transmute(
    simple_total = simple_total_recruits_per_fruit,
    model_total = model_total_recruits_per_fruit,
    difference = model_total - simple_total,
    ratio = model_total / simple_total)
```

```{r by_year_by_site_fr2re_site_comparison}
df_fr2re_site_compare <- df_fr2re_site_year %>%
  filter(fruits_site_t0 > 0) %>%
  mutate(
    pred_recruits_mean_model = expm1(fr2re_b1_mean * log1p(fruits_site_t0))) %>%
  group_by(site) %>%
  summarise(
    total_fruits = sum(fruits_site_t0, na.rm = TRUE),
    total_recruits = sum(recruits_site_t1, na.rm = TRUE),
    total_pred_recruits = sum(pred_recruits_mean_model, na.rm = TRUE),
    obs_recruits_per_fruit = total_recruits / total_fruits,
    model_recruits_per_fruit = total_pred_recruits / total_fruits,
    .groups = "drop") %>%
  arrange(as.numeric(as.character(site)))

df_fr2re_site_compare

df_fr2re_site_compare %>%
  summarise(
    mean_obs_site_ratio = mean(obs_recruits_per_fruit, na.rm = TRUE),
    median_obs_site_ratio = median(obs_recruits_per_fruit, na.rm = TRUE),
    mean_model_site_ratio = mean(model_recruits_per_fruit, na.rm = TRUE),
    median_model_site_ratio = median(model_recruits_per_fruit, na.rm = TRUE))
```

## Observed versus modeled lambda

```{r by_year_by_site_lambda_compare_plot}
df_lambda_site_year_compare_plot <- df_lambda_site_year_compare %>%
  select(
    site,
    year,
    year_t1,
    lambda_obs,
    lambda_asymptotic,
    lambda_projected) %>%
  filter(!is.na(lambda_obs)) %>%
  pivot_longer(
    cols = c(lambda_asymptotic, lambda_projected),
    names_to = "modeled_lambda_type",
    values_to = "modeled_lambda") %>%
  mutate(
    modeled_lambda_type = recode(
      modeled_lambda_type,
      lambda_asymptotic = "Asymptotic lambda",
      lambda_projected = "Projected lambda"))

fig_lambda_site_year_compare <- ggplot(
  df_lambda_site_year_compare_plot,
  aes(x = modeled_lambda, y = lambda_obs)
) +
  geom_point(alpha = 0.65, size = 1.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~ modeled_lambda_type) +
  theme_bw() +
  labs(
    title = "Observed versus modeled site-year population growth",
    subtitle = v_ggp_suffix,
    x = expression("Modeled " * lambda),
    y = expression("Observed " * lambda))

fig_lambda_site_year_compare
```

## Observed versus modeled lambda on the log scale

```{r by_year_by_site_lambda_compare_plot_log}
df_lambda_site_year_compare_plot_log <- df_lambda_site_year_compare %>%
  select(
    site,
    year,
    year_t1,
    lambda_obs,
    lambda_asymptotic,
    lambda_projected) %>%
  filter(!is.na(lambda_obs), lambda_obs > 0) %>%
  pivot_longer(
    cols = c(lambda_asymptotic, lambda_projected),
    names_to = "modeled_lambda_type",
    values_to = "modeled_lambda") %>%
  filter(!is.na(modeled_lambda), modeled_lambda > 0) %>%
  mutate(
    log_lambda_obs = log(lambda_obs),
    log_modeled_lambda = log(modeled_lambda),
    modeled_lambda_type = recode(
      modeled_lambda_type,
      lambda_asymptotic = "Asymptotic lambda",
      lambda_projected = "Projected lambda"))

fig_lambda_site_year_compare_log <- ggplot(
  df_lambda_site_year_compare_plot_log,
  aes(x = log_modeled_lambda, y = log_lambda_obs)
) +
  geom_point(alpha = 0.65, size = 1.8) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  facet_wrap(~ modeled_lambda_type) +
  theme_bw() +
  labs(
    title = "Observed versus modeled site-year population growth",
    subtitle = paste(v_ggp_suffix, "- log-transformed lambda"),
    x = expression("log modeled " * lambda),
    y = expression("log observed " * lambda))

fig_lambda_site_year_compare_log
```

## Site-level summary

```{r by_year_by_site_site_summary}
df_lambda_site_summary <- df_lambda_site_year_compare %>%
  filter(
    !is.na(lambda_obs),
    !is.na(lambda_asymptotic),
    !is.na(lambda_projected),
    lambda_obs > 0,
    lambda_asymptotic > 0,
    lambda_projected > 0) %>%
  group_by(site) %>%
  summarise(
    n_year_transitions = n(),
    lambda_obs_geometric = exp(mean(log(lambda_obs), na.rm = TRUE)),
    lambda_obs_arithmetic = mean(lambda_obs, na.rm = TRUE),
    lambda_asymptotic_geometric = exp(mean(log(lambda_asymptotic),
                                           na.rm = TRUE)),
    lambda_asymptotic_arithmetic = mean(lambda_asymptotic, na.rm = TRUE),
    lambda_projected_geometric = exp(mean(log(lambda_projected),
                                          na.rm = TRUE)),
    lambda_projected_arithmetic = mean(lambda_projected, na.rm = TRUE),
    error_projected_geo_vs_obs_geo =
      lambda_projected_geometric - lambda_obs_geometric,
    rmse_projected_vs_obs =
      sqrt(mean((lambda_projected - lambda_obs)^2, na.rm = TRUE)),
    mean_n_initial = mean(n_initial, na.rm = TRUE),
    mean_p_disturbance = mean(p_disturbance, na.rm = TRUE),
    .groups = "drop") %>%
  arrange(as.numeric(as.character(site)))

df_lambda_site_summary %>%
  print(n = 100, width = Inf)
```
