# by Year: two-part reproduction and fruit-to-recruit {#byyear_fl2x_fr2x_fr2re}

This chapter builds year-specific integral projection models for *Crotalaria
avonensis*. Survival, growth, flowering, flower production, fruiting, fruit
production, and fruit-to-recruit conversion vary among years. Disturbance is a
fixed effect and is assigned to the transition year $t_0$.

No site random effects are fitted or extracted. Site and quadrat identifiers are
used only to pair observations through time and to compare observed and projected
population growth.

The reproductive models are split into two parts:

1. flowering probability and flower number conditional on flowering;
2. fruiting probability conditional on flowering and fruit number conditional on
   fruiting.

The fertility kernel uses

```r
Pr(flowering) * Pr(fruiting | flowering) *
  E(number of fruits | fruiting) * recruits per fruit
```

Flower number is retained as a separate biological process and diagnostic, but
it is not multiplied into fecundity because the fruit-number model already
predicts total fruits per fruiting plant.

## Required objects

This chapter assumes that the extended-data chapter has created `df`, `df_og`,
and `v_ggp_suffix`.


```{r byyear_fl2x_required_objects}
required_objects <- c('df', 'df_og', 'v_ggp_suffix')
missing_objects <- required_objects[!purrr::map_lgl(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop(
    'The following objects are missing. Run the data chapter first: ',
    paste(missing_objects, collapse = ', '))
}

if (!exists('v_years_re')) v_years_re <- c()
if (!exists('v_mod_set_su')) v_mod_set_su <- c()
if (!exists('v_mod_set_gr')) v_mod_set_gr <- c()
if (!exists('v_mod_set_fl')) v_mod_set_fl <- c()
if (!exists('v_mod_set_fl_n')) v_mod_set_fl_n <- c()
if (!exists('v_mod_set_fr')) v_mod_set_fr <- c()
if (!exists('v_mod_set_fr_n')) v_mod_set_fr_n <- c()
if (!exists('v_mod_set_fr2re')) v_mod_set_fr2re <- c()

ctrl_glmer <- lme4::glmerControl(
  optimizer = 'bobyqa', optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lme4::lmerControl(
  optimizer = 'bobyqa', optCtrl = list(maxfun = 2e5))
```

## Correct recruitment timing

The first-observation year must be calculated within each plant. The earlier data
chapter calculated it without grouping by plant, which can misclassify recruits.
The following block replaces `recruit` and rebuilds the previous-year quadrat
sampling check before fitting the fruit-to-recruit model.


```{r byyear_fl2x_recruit_correction}
df_recruit_corrected <- df_og %>%
  janitor::clean_names() %>%
  transmute(
    plant_id = as.character(id),
    year = as.integer(year),
    s,
    stage) %>%
  group_by(plant_id) %>%
  mutate(
    earliest_recorded_date = suppressWarnings(
      min(year[s > 0 & s != 2 & s != 6], na.rm = TRUE)),
    earliest_recorded_date = if_else(
      is.infinite(earliest_recorded_date),
      NA_real_, earliest_recorded_date),
    recruit = case_when(
      stage == 1 ~ 1,
      year == earliest_recorded_date & earliest_recorded_date > 2003 ~ 1,
      TRUE ~ NA_real_)) %>%
  ungroup() %>%
  select(plant_id, year, recruit)

df <- df %>%
  mutate(
    plant_id = as.character(plant_id),
    year = as.integer(as.character(year))) %>%
  select(-any_of(c('recruit', 'recruit_plot_valid'))) %>%
  left_join(df_recruit_corrected, by = c('plant_id', 'year')) %>%
  mutate(plant_id = factor(plant_id))

# Identify quadrats sampled in each year.
df_plot_year_sampled <- df %>%
  group_by(site, mp, quad, quad_id, year) %>%
  summarise(
    n_records_plot_year = n(),
    n_plants_plot_year = n_distinct(plant_id),
    .groups = 'drop')

# Retain recruits only when their quadrat was sampled in the previous year.
df_recruit_valid_ids <- df %>%
  filter(recruit == 1) %>%
  mutate(year_prev = year - 1) %>%
  left_join(
    df_plot_year_sampled %>%
      transmute(
        quad_id,
        year_prev = year,
        sampled_prev_year = TRUE),
    by = c('quad_id', 'year_prev')) %>%
  filter(replace_na(sampled_prev_year, FALSE)) %>%
  select(site, mp, quad, quad_id, plant_id, year) %>%
  mutate(recruit_plot_valid = 1)

df <- df %>%
  left_join(
    df_recruit_valid_ids,
    by = c('site', 'mp', 'quad', 'quad_id', 'plant_id', 'year')) %>%
  mutate(
    recruit_plot_valid = case_when(
      recruit == 1 & recruit_plot_valid == 1 ~ 1,
      TRUE ~ NA_real_))
```

## Process-specific data


```{r byyear_fl2x_model_data}
# Survival data ----------------------------------------------------------------
df_su <- df %>%
  filter(
    size_t0 != 0,
    !is.na(survives),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)


# Growth data ------------------------------------------------------------------
df_gr <- df %>%
  filter(
    size_t0 != 0,
    survives == 1,
    is.finite(logsize_t0),
    is.finite(logsize_t1),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, disturbance)


# Flower data ------------------------------------------------------------------
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance),
    flower = if_else(flower > 0, 1, 0)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)


# Flower-number data -----------------------------------------------------------
# Flower number is conditional on flowering because flower > 0.
# Expected flowers per plant in the IPM:
# Pr(flowering) * E(number of flowers | flowering).

df_fl_n <- df %>%
  filter(
    flower > 0,
    is.finite(flower),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower,
    logsize_t0, logsize_t0_2, logsize_t0_3,
    disturbance)


# Fruiting-probability data ----------------------------------------------------
# Fruiting probability is conditional on flowering because flower > 0.

df_fr <- df %>%
  filter(
    flower > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance),
    fruiting = if_else(fruit > 0, 1, 0)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, fruit, fruiting,
    logsize_t0, logsize_t0_2, logsize_t0_3, disturbance)


# Fruit-number data ------------------------------------------------------------
# Fruit number is conditional on fruiting because fruit > 0.

df_fr_n <- df %>%
  filter(
    fruit > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, fruit,
    logsize_t0, logsize_t0_2, logsize_t0_3, disturbance)


# Fruit-to-recruit data --------------------------------------------------------
# Total fruits in year t predict recruits in year t + 1.
# Only quadrats sampled in consecutive years are retained.

df_fr2re_quad <- df %>%
  mutate(year = as.integer(as.character(year))) %>%
  group_by(site, quad_id, year) %>%
  summarise(
    fruits_t0 = sum(fruit, na.rm = TRUE),
    recruits_t0 = sum(recruit_plot_valid == 1, na.rm = TRUE),
    .groups = "drop") %>%
  arrange(site, quad_id, year) %>%
  group_by(site, quad_id) %>%
  mutate(
    recruits_t1 = lead(recruits_t0),
    year_t1 = lead(year),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_fr2re <- df_fr2re_quad %>%
  group_by(site, year, year_t1) %>%
  summarise(
    fruits_t0 = sum(fruits_t0, na.rm = TRUE),
    recruits_t1 = sum(recruits_t1, na.rm = TRUE),
    n_quads = n(),
    .groups = "drop") %>%
  mutate(
    site = factor(site),
    year_t1 = factor(year_t1),
    logfruits_t0 = log1p(fruits_t0),
    logrecruits_t1 = log1p(recruits_t1))


# Removing years with too few data --------------------------------------------
v_years_og <- sort(unique(df$year))

df <- df %>% filter(!is.na(year), !(year %in% v_years_re))
df_su <- df_su %>% filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()
df_gr <- df_gr %>% filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()
df_fl <- df_fl %>%
  filter(!is.na(year), !(year %in% v_years_re)) %>% droplevels()
df_fl_n <- df_fl_n %>% filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()
df_fr <- df_fr %>% filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()
df_fr_n <- df_fr_n %>% filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()
df_fr2re <- df_fr2re %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !(year %in% v_years_re),
    !(as.integer(as.character(year_t1)) %in% v_years_re)) %>%
  droplevels()

v_years <- sort(unique(df$year))
```

## Survival


```{r byyear_fl2x_survival_model}
# Survival model ---------------------------------------------------------------
# GLMM; binomial
mod_su_0 <- glmer(
  survives ~ disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mod_su_1 <- glmer(
  survives ~ logsize_t0 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mod_su_2 <- glmer(
  survives ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mod_su_3 <- glmer(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mods_su <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef <- coef(mod_su_best)$year

summary(mod_su_best)
```

```{r byyear_fl2x_survival_plot, fig.width=10, fig.height=12}
# Survival plots by year -------------------------------------------------------
make_su_year_plot <- function(year_i) {
  df_i <- df_su %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_su$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      survives = predict(
        mod_su_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      survives = mean(survives, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, survives, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, survives, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, survives, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Survival probability') +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

su_yrs <- lapply(levels(df_su$year), make_su_year_plot)

fig_su_years <- wrap_plots(su_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_su_years
```

## Growth


```{r byyear_fl2x_growth_model}
# Growth model -----------------------------------------------------------------
mod_gr_0 <- lmer(
  logsize_t1 ~ disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mod_gr_1 <- lmer(
  logsize_t1 ~ logsize_t0 + disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mod_gr_2 <- lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mod_gr_3 <- lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mods_gr <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef <- coef(mod_gr_best)$year

summary(mod_gr_best)
```

```{r byyear_fl2x_growth_plot, fig.width=10, fig.height=12}
# Growth plots by year ---------------------------------------------------------
make_gr_year_plot <- function(year_i) {
  df_i <- df_gr %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_gr$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      logsize_t1 = predict(
        mod_gr_best, newdata = pred_i, re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      logsize_t1 = mean(logsize_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, logsize_t1, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, logsize_t1, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, logsize_t1, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = expression('log(size)'[t1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

gr_yrs <- lapply(levels(df_gr$year), make_gr_year_plot)

fig_gr_years <- wrap_plots(gr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Growth - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_gr_years
```

```{r byyear_fl2x_growth_variance}
# Growth variance --------------------------------------------------------------
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

## Flowering probability


```{r byyear_fl2x_flower_model}
# Flower model -----------------------------------------------------------------
mod_fl_0 <- glmer(
  flower ~ disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mod_fl_1 <- glmer(
  flower ~ logsize_t0 + disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mod_fl_2 <- glmer(
  flower ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mod_fl_3 <- glmer(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mods_fl <- list(mod_fl_0, mod_fl_1, mod_fl_2, mod_fl_3)
mods_fl_dAIC <- bbmle::AICctab(mods_fl, weights = TRUE, sort = FALSE)$dAIC
mods_fl_sorted <- order(mods_fl_dAIC)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_best <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef <- coef(mod_fl_best)$year

summary(mod_fl_best)
```

```{r byyear_fl2x_flower_plot, fig.width=10, fig.height=12}
# Flowering plots by year ------------------------------------------------------
make_fl_year_plot <- function(year_i) {
  df_i <- df_fl %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fl$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      flower = predict(
        mod_fl_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      flower = mean(flower, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, flower, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, flower, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, flower, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Flowering probability') +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_yrs <- lapply(levels(df_fl$year), make_fl_year_plot)

fig_fl_years <- wrap_plots(fl_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flowering probability - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_years
```

## Flower number conditional on flowering


```{r byyear_fl2x_flower_number_model}
# Flower-number model ----------------------------------------------------------
mod_fl_n_0 <- glmer.nb(
  flower ~ disturbance + (1 | year),
  data = df_fl_n, control = ctrl_glmer)

mod_fl_n_1 <- glmer.nb(
  flower ~ logsize_t0 + disturbance + (1 | year),
  data = df_fl_n, control = ctrl_glmer)

mod_fl_n_2 <- glmer.nb(
  flower ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fl_n, control = ctrl_glmer)

mod_fl_n_3 <- glmer.nb(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance +
    (1 | year),
  data = df_fl_n, control = ctrl_glmer)

mods_fl_n <- list(
  mod_fl_n_0, mod_fl_n_1, mod_fl_n_2, mod_fl_n_3)

mods_fl_n_dAIC <- bbmle::AICctab(
  mods_fl_n, weights = TRUE, sort = FALSE)$dAIC

mods_fl_n_sorted <- order(mods_fl_n_dAIC)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_bestfit <- mods_fl_n_sorted[1]
  v_mod_fl_n_index <- mod_fl_n_index_bestfit - 1
} else {
  mod_fl_n_index_bestfit <- v_mod_set_fl_n + 1
  v_mod_fl_n_index <- v_mod_set_fl_n
}

mod_fl_n_best <- mods_fl_n[[mod_fl_n_index_bestfit]]
mod_fl_n_ranef <- coef(mod_fl_n_best)$year

summary(mod_fl_n_best)
```

```{r byyear_fl2x_flower_number_plot, fig.width=10, fig.height=12}
# Flower-number plots by year --------------------------------------------------
make_fl_n_year_plot <- function(year_i) {
  df_i <- df_fl_n %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fl_n$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      flower = predict(
        mod_fl_n_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      flower = mean(flower, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, flower, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, flower, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, flower, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Number of flowers') +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_n_yrs <- lapply(levels(df_fl_n$year), make_fl_n_year_plot)

fig_fl_n_years <- wrap_plots(fl_n_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flower number given flowering - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_n_years
```

## Fruiting probability conditional on flowering


```{r byyear_fl2x_fruiting_model}
# Fruiting-probability model ---------------------------------------------------
mod_fr_0 <- glmer(
  fruiting ~ disturbance + (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mod_fr_1 <- glmer(
  fruiting ~ logsize_t0 + disturbance + (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mod_fr_2 <- glmer(
  fruiting ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mod_fr_3 <- glmer(
  fruiting ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance +
    (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mods_fr <- list(mod_fr_0, mod_fr_1, mod_fr_2, mod_fr_3)
mods_fr_dAIC <- bbmle::AICctab(
  mods_fr, weights = TRUE, sort = FALSE)$dAIC
mods_fr_sorted <- order(mods_fr_dAIC)

if (length(v_mod_set_fr) == 0) {
  mod_fr_index_bestfit <- mods_fr_sorted[1]
  v_mod_fr_index <- mod_fr_index_bestfit - 1
} else {
  mod_fr_index_bestfit <- v_mod_set_fr + 1
  v_mod_fr_index <- v_mod_set_fr
}

mod_fr_best <- mods_fr[[mod_fr_index_bestfit]]
mod_fr_ranef <- coef(mod_fr_best)$year

summary(mod_fr_best)
```

```{r byyear_fl2x_fruiting_plot, fig.width=10, fig.height=12}
# Fruiting-probability plots by year ------------------------------------------
make_fr_year_plot <- function(year_i) {
  df_i <- df_fr %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fr$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      fruiting = predict(
        mod_fr_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      fruiting = mean(fruiting, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, fruiting, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, fruiting, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, fruiting, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Fruiting probability') +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fr_yrs <- lapply(levels(df_fr$year), make_fr_year_plot)

fig_fr_years <- wrap_plots(fr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Fruiting probability given flowering - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fr_years
```

## Fruit number conditional on fruiting


```{r byyear_fl2x_fruit_number_model}
# Fruit-number model -----------------------------------------------------------
mod_fr_n_0 <- glmer.nb(
  fruit ~ disturbance + (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mod_fr_n_1 <- glmer.nb(
  fruit ~ logsize_t0 + disturbance + (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mod_fr_n_2 <- glmer.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mod_fr_n_3 <- glmer.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance +
    (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mods_fr_n <- list(mod_fr_n_0, mod_fr_n_1, mod_fr_n_2, mod_fr_n_3)
mods_fr_n_dAIC <- bbmle::AICctab(
  mods_fr_n, weights = TRUE, sort = FALSE)$dAIC
mods_fr_n_sorted <- order(mods_fr_n_dAIC)

if (length(v_mod_set_fr_n) == 0) {
  mod_fr_n_index_bestfit <- mods_fr_n_sorted[1]
  v_mod_fr_n_index <- mod_fr_n_index_bestfit - 1
} else {
  mod_fr_n_index_bestfit <- v_mod_set_fr_n + 1
  v_mod_fr_n_index <- v_mod_set_fr_n
}

mod_fr_n_best <- mods_fr_n[[mod_fr_n_index_bestfit]]
mod_fr_n_ranef <- coef(mod_fr_n_best)$year

summary(mod_fr_n_best)
```

```{r byyear_fl2x_fruit_number_plot, fig.width=10, fig.height=12}
# Fruit-number plots by year ---------------------------------------------------
make_fr_n_year_plot <- function(year_i) {
  df_i <- df_fr_n %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fr_n$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      fruit = predict(
        mod_fr_n_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      fruit = mean(fruit, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, fruit, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, fruit, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, fruit, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Number of fruits') +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fr_n_yrs <- lapply(levels(df_fr_n$year), make_fr_n_year_plot)

fig_fr_n_years <- wrap_plots(fr_n_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Fruit number given fruiting - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fr_n_years
```

## Fruit-to-recruit transition


```{r byyear_fl2x_fr2re_model}
# Fruit-to-recruit models ------------------------------------------------------
# Zero fruits predict zero recruits.
# Recruitment varies among years, but not among sites.

mod_fr2re_0 <- glmmTMB(
  logrecruits_t1 ~ 0 + logfruits_t0,
  data = df_fr2re, family = gaussian, REML = FALSE)

mod_fr2re_1 <- glmmTMB(
  logrecruits_t1 ~ 0 + logfruits_t0 +
    (0 + logfruits_t0 | year_t1),
  data = df_fr2re, family = gaussian, REML = FALSE)

mods_fr2re <- list(mod_fr2re_0, mod_fr2re_1)

mods_fr2re_dAIC <- bbmle::AICctab(
  mods_fr2re, weights = TRUE, sort = FALSE)$dAIC
mods_fr2re_sorted <- order(mods_fr2re_dAIC)

if (length(v_mod_set_fr2re) == 0) {
  mod_fr2re_index_bestfit <- mods_fr2re_sorted[1]
  v_mod_fr2re_index <- mod_fr2re_index_bestfit - 1
} else {
  mod_fr2re_index_bestfit <- v_mod_set_fr2re + 1
  v_mod_fr2re_index <- v_mod_set_fr2re
}

mod_fr2re_best <- mods_fr2re[[mod_fr2re_index_bestfit]]
mod_fr2re_fixef <- fixef(mod_fr2re_best)$cond
mod_fr2re_ranef <- ranef(mod_fr2re_best)$cond

summary(mod_fr2re_best)
```

```{r byyear_fl2x_fr2re_plot, fig.width=10, fig.height=12}
# Fruit-to-recruit plots by year ----------------------------------------------
make_fr2re_year_plot <- function(year_i) {
  df_i <- df_fr2re %>%
    filter(as.character(year_t1) == year_i)
  
  x <- seq(
    min(df_i$logfruits_t0, na.rm = TRUE),
    max(df_i$logfruits_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- tibble(
    logfruits_t0 = x,
    year_t1 = factor(year_i, levels = levels(df_fr2re$year_t1)))
  
  pred_i <- pred_i %>%
    mutate(
      logrecruits_t1 = predict(
        mod_fr2re_best, newdata = pred_i, type = 'response',
        re.form = NULL, allow.new.levels = TRUE),
      recruits_t1 = pmax(0, expm1(logrecruits_t1)))
  
  pts_i <- df_i %>%
    mutate(bin = cut(logfruits_t0, breaks = 8)) %>%
    group_by(bin) %>%
    summarise(
      logfruits_t0 = mean(logfruits_t0, na.rm = TRUE),
      recruits_t1 = mean(recruits_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logfruits_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i,
      aes(logfruits_t0, recruits_t1),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logfruits_t0, recruits_t1),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logfruits_t0, recruits_t1),
      linewidth = 0.7) +
    labs(
      title = year_i,
      x = expression('log(number of fruits + 1)'[t]),
      y = expression('Number of recruits'[t+1])) +
    theme_bw() +
    theme(text = element_text(size = 5))
}

fr2re_yrs <- lapply(
  levels(df_fr2re$year_t1), make_fr2re_year_plot)

fig_fr2re_years <- wrap_plots(fr2re_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Fruit-to-recruit relationship - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fr2re_years
```

## Recruit size distribution


```{r byyear_fl2x_recruit_size}
# Recruit size distribution ----------------------------------------------------
df_recr <- df %>%
  filter(
    recruit_plot_valid == 1,
    size_t0 > 0,
    is.finite(logsize_t0))

rc_sz <- df_recr %>%
  summarise(
    mean = mean(logsize_t0, na.rm = TRUE),
    sd = sd(logsize_t0, na.rm = TRUE))
```

## Parameter extraction

The fixed parameters are stored in `pars_mean`. The year-specific conditional coefficients are stored as one row per transition year in `pars_year`.


```{r byyear_fl2x_parameter_helpers}
# Parameter extraction ---------------------------------------------------------
# pars_mean contains fixed-effect coefficients.
# pars_year contains complete conditional coefficients for each year.

get_model_fixef <- function(model) {
  if (inherits(model, 'glmmTMB')) {
    return(glmmTMB::fixef(model)$cond)
  }
  
  if (inherits(model, 'merMod')) {
    return(lme4::fixef(model))
  }
  
  stats::coef(model)
}


get_group_coef <- function(model, group_var) {
  if (inherits(model, 'glmmTMB')) {
    coef_list <- stats::coef(model)$cond
  } else if (inherits(model, 'merMod')) {
    coef_list <- stats::coef(model)
  } else {
    return(NULL)
  }
  
  if (is.null(coef_list) || !(group_var %in% names(coef_list))) {
    return(NULL)
  }
  
  as.data.frame(coef_list[[group_var]])
}


size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logsize_t0',
  b2 = 'logsize_t0_2',
  b3 = 'logsize_t0_3',
  bd = 'disturbance')


extract_terms <- function(coefs, prefix, term_map = size_term_map) {
  values <- vapply(term_map, function(term_i) {
    if (term_i %in% names(coefs)) {
      unname(coefs[[term_i]])
    } else {
      0
    }
  }, numeric(1))
  
  names(values) <- paste0(prefix, names(term_map))
  values
}


extract_fixed_rate <- function(model, prefix) {
  extract_terms(
    coefs = get_model_fixef(model),
    prefix = prefix)
}


extract_year_rate <- function(
    model, prefix, years, group_var = 'year') {
  
  fixed_coef <- get_model_fixef(model)
  group_coef <- get_group_coef(model, group_var)
  
  purrr::map_dfr(years, function(year_i) {
    level_i <- as.character(year_i)
    coef_i <- fixed_coef
    
    if (!is.null(group_coef) &&
        level_i %in% rownames(group_coef)) {
      coef_i <- unlist(
        group_coef[level_i, , drop = FALSE],
        use.names = FALSE)
      
      names(coef_i) <- colnames(group_coef)
    }
    
    tibble::as_tibble_row(c(
      year = year_i,
      extract_terms(coef_i, prefix)))
  })
}


extract_fr2re_year <- function(model, years) {
  fixed_coef <- get_model_fixef(model)
  group_coef <- get_group_coef(model, 'year_t1')
  
  purrr::map_dfr(years, function(year_i) {
    year_t1_i <- year_i + 1
    level_i <- as.character(year_t1_i)
    coef_i <- fixed_coef
    
    if (!is.null(group_coef) &&
        level_i %in% rownames(group_coef)) {
      coef_i <- unlist(
        group_coef[level_i, , drop = FALSE],
        use.names = FALSE)
      
      names(coef_i) <- colnames(group_coef)
    }
    
    fr2re_b1 <- if ('logfruits_t0' %in% names(coef_i)) {
      unname(coef_i[['logfruits_t0']])
    } else {
      0
    }
    
    tibble(
      year = year_i,
      year_t1 = year_t1_i,
      fr2re_b1 = fr2re_b1)
  })
}
```

```{r byyear_fl2x_mean_parameters}
# Mean parameters --------------------------------------------------------------
grow_var_coef <- stats::coef(mod_gr_var)

mesh_limits <- range(
  c(
    df_gr$logsize_t0,
    df_gr$logsize_t1,
    df_recr$logsize_t0),
  na.rm = TRUE,
  finite = TRUE)

pars_mean <- as.list(c(
  extract_fixed_rate(mod_su_best, 'surv_'),
  extract_fixed_rate(mod_gr_best, 'grow_'),
  extract_fixed_rate(mod_fl_best, 'fl_'),
  extract_fixed_rate(mod_fl_n_best, 'fln_'),
  extract_fixed_rate(mod_fr_best, 'fr_'),
  extract_fixed_rate(mod_fr_n_best, 'frn_'),
  fr2re_b1 = unname(
    get_model_fixef(mod_fr2re_best)[['logfruits_t0']]),
  recr_mean = rc_sz$mean,
  recr_sd = rc_sz$sd,
  grow_var_a = unname(grow_var_coef[['a']]),
  grow_var_b = unname(grow_var_coef[['b']]),
  fruits_ref = mean(df_fr2re$fruits_t0, na.rm = TRUE),
  fr2re_sigma = stats::sigma(mod_fr2re_best),
  L = mesh_limits[1],
  U = mesh_limits[2],
  mat_siz = 200))


pars_mean_table <- tibble(
  parameter = names(pars_mean),
  value = as.numeric(unlist(pars_mean)))

pars_mean_table
```

```{r byyear_fl2x_year_parameters}
# Years with all required vital-rate models -----------------------------------
get_year_values <- function(x) {
  sort(unique(as.integer(as.character(x))))
}

ipm_years <- Reduce(
  intersect,
  list(
    get_year_values(df_su$year),
    get_year_values(df_gr$year),
    get_year_values(df_fl$year),
    get_year_values(df_fl_n$year),
    get_year_values(df_fr$year),
    get_year_values(df_fr_n$year),
    get_year_values(df_fr2re$year_t1) - 1))

if (length(ipm_years) == 0) {
  stop('No years have complete data for all IPM processes')
}

ipm_years


# Year-specific parameter extraction ------------------------------------------
pars_su_year <- extract_year_rate(
  model = mod_su_best, prefix = 'surv_', years = ipm_years)

pars_gr_year <- extract_year_rate(
  model = mod_gr_best, prefix = 'grow_', years = ipm_years)

pars_fl_year <- extract_year_rate(
  model = mod_fl_best, prefix = 'fl_', years = ipm_years)

pars_fl_n_year <- extract_year_rate(
  model = mod_fl_n_best, prefix = 'fln_', years = ipm_years)

pars_fr_year <- extract_year_rate(
  model = mod_fr_best, prefix = 'fr_', years = ipm_years)

pars_fr_n_year <- extract_year_rate(
  model = mod_fr_n_best, prefix = 'frn_', years = ipm_years)

pars_fr2re_year <- extract_fr2re_year(
  model = mod_fr2re_best, years = ipm_years)

pars_year <- list(
  pars_su_year,
  pars_gr_year,
  pars_fl_year,
  pars_fl_n_year,
  pars_fr_year,
  pars_fr_n_year,
  pars_fr2re_year) %>%
  purrr::reduce(left_join, by = 'year') %>%
  arrange(year)

pars_year


# Create mean or year-specific parameter list ---------------------------------
make_ipm_pars <- function(year_i = NULL) {
  pars <- pars_mean
  
  if (is.null(year_i)) {
    return(pars)
  }
  
  year_i <- as.integer(year_i)
  
  pars_i <- pars_year %>%
    filter(.data$year == .env$year_i)
  
  if (nrow(pars_i) != 1) {
    stop(
      'Expected one parameter row for year ', year_i,
      ', but found ', nrow(pars_i))
  }
  
  pars_i <- pars_i %>%
    select(-year, -any_of('year_t1')) %>%
    as.list()
  
  for (par_name in names(pars_i)) {
    pars[[par_name]] <- pars_i[[par_name]]
  }
  
  pars
}
```

## IPM functions


```{r byyear_fl2x_ipm_functions}
# IPM functions ----------------------------------------------------------------
get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    pars[[par_name]]
  } else {
    default
  }
}


vital_lp <- function(x, pars, prefix, disturbance = 0) {
  get_par(pars, paste0(prefix, 'b0')) +
    get_par(pars, paste0(prefix, 'b1')) * x +
    get_par(pars, paste0(prefix, 'b2')) * x^2 +
    get_par(pars, paste0(prefix, 'b3')) * x^3 +
    get_par(pars, paste0(prefix, 'bd')) * disturbance
}


# Survival ---------------------------------------------------------------------
sx <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'surv_',
    disturbance = disturbance))
}


# Growth -----------------------------------------------------------------------
grow_mu <- function(x, pars, disturbance = 0) {
  vital_lp(
    x = x,
    pars = pars,
    prefix = 'grow_',
    disturbance = disturbance)
}


grow_sd <- function(mean_value, pars) {
  variance <- get_par(pars, 'grow_var_a') *
    exp(get_par(pars, 'grow_var_b') * mean_value)
  
  sqrt(pmax(variance, .Machine$double.eps))
}


gxy <- function(x, y, pars, disturbance = 0) {
  mean_value <- grow_mu(
    x = x,
    pars = pars,
    disturbance = disturbance)
  
  sd_value <- grow_sd(
    mean_value = mean_value,
    pars = pars)
  
  dnorm(y, mean = mean_value, sd = sd_value)
}


pxy <- function(x, y, pars, disturbance = 0) {
  sx(x, pars, disturbance) *
    gxy(x, y, pars, disturbance)
}


# Flowering probability --------------------------------------------------------
fl_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fl_',
    disturbance = disturbance))
}


# Flower number conditional on flowering --------------------------------------
fln_x <- function(x, pars, disturbance = 0) {
  exp(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fln_',
    disturbance = disturbance))
}


flowers_x <- function(x, pars, disturbance = 0) {
  fl_x(x, pars, disturbance) *
    fln_x(x, pars, disturbance)
}


# Fruiting probability conditional on flowering -------------------------------
fr_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fr_',
    disturbance = disturbance))
}


# Fruit number conditional on fruiting ----------------------------------------
frn_x <- function(x, pars, disturbance = 0) {
  exp(vital_lp(
    x = x,
    pars = pars,
    prefix = 'frn_',
    disturbance = disturbance))
}


# Expected fruits per individual ----------------------------------------------
# Flower number is not multiplied here because the fruit models describe
# plant-level fruiting and total fruit number.

fruits_x <- function(x, pars, disturbance = 0) {
  fl_x(x, pars, disturbance) *
    fr_x(x, pars, disturbance) *
    frn_x(x, pars, disturbance)
}


# Recruitment at the reference fruit production -------------------------------
re_total_ref <- function(pars) {
  eta <- get_par(pars, 'fr2re_b1') *
    log1p(get_par(pars, 'fruits_ref'))
  
  recruits <- expm1(
    eta + 0.5 * get_par(pars, 'fr2re_sigma')^2)
  
  pmax(recruits, 0)
}


re_per_fruit <- function(pars) {
  re_total_ref(pars) /
    pmax(
      get_par(pars, 'fruits_ref'),
      .Machine$double.eps)
}


# Expected recruits contributed by an individual ------------------------------
rx <- function(x, pars, disturbance = 0) {
  fruits_x(
    x = x,
    pars = pars,
    disturbance = disturbance) *
    re_per_fruit(pars)
}


# Recruit size distribution ----------------------------------------------------
recr_y <- function(y, pars) {
  dnorm(
    y,
    mean = get_par(pars, 'recr_mean'),
    sd = get_par(pars, 'recr_sd'))
}


# Fertility matrix -------------------------------------------------------------
fy <- function(y, x, pars, h, disturbance = 0) {
  Pvec <- recr_y(y, pars)
  
  recruit_mass <- sum(Pvec * h)
  
  if (!is.finite(recruit_mass) || recruit_mass <= 0) {
    stop('Recruit size distribution has no mass inside the mesh')
  }
  
  Pvec <- Pvec / recruit_mass
  
  Rvec <- rx(
    x = x,
    pars = pars,
    disturbance = disturbance)
  
  outer(Pvec, Rvec) * h
}
```

## Kernel


```{r byyear_fl2x_kernel}
# Kernel -----------------------------------------------------------------------
kernel <- function(pars, disturbance = 0) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  # Fertility matrix
  Fmat <- fy(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance)
  
  # Survival vector
  Smat <- sx(
    x = y,
    pars = pars,
    disturbance = disturbance)
  
  # Growth matrix
  Gmat <- matrix(0, n, n)
  
  Gmat[] <- t(
    outer(
      y,
      y,
      Vectorize(function(x, y) {
        gxy(
          x = x,
          y = y,
          pars = pars,
          disturbance = disturbance)
      }))) * h
  
  # Growth-survival matrix
  Tmat <- matrix(0, n, n)
  
  # Correct eviction of small individuals
  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }
  
  # Correct eviction of large individuals
  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }
  
  k_yx <- Fmat + Tmat
  
  list(
    k_yx = k_yx,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Smat = Smat,
    meshpts = y,
    h = h)
}
```

## Mean and year-specific population growth


```{r byyear_fl2x_lambda}
# Population growth rate -------------------------------------------------------
lambda_ipm <- function(pars, disturbance = 0) {
  K <- kernel(
    pars = pars,
    disturbance = disturbance)$k_yx
  
  Re(eigen(K, only.values = TRUE)$values[1])
}


# Mean IPM ---------------------------------------------------------------------
lambda_mean_no_fire <- lambda_ipm(
  pars = pars_mean,
  disturbance = 0)

lambda_mean_fire <- lambda_ipm(
  pars = pars_mean,
  disturbance = 1)

lambda_mean_no_fire
lambda_mean_fire


# Years with complete year-specific parameter sets ----------------------------
ipm_years <- pars_year %>%
  filter(!is.na(year)) %>%
  distinct(year) %>%
  arrange(year) %>%
  pull(year)

ipm_years


# Year-specific IPM ------------------------------------------------------------
lambda_ipm_year <- function(year, disturbance = 0) {
  pars_i <- make_ipm_pars(year_i = year)
  
  lambda_ipm(
    pars = pars_i,
    disturbance = disturbance)
}


lambda_year <- tibble(
  year = ipm_years,
  lambda_no_fire = purrr::map_dbl(
    ipm_years,
    ~ lambda_ipm_year(.x, disturbance = 0)),
  lambda_fire = purrr::map_dbl(
    ipm_years,
    ~ lambda_ipm_year(.x, disturbance = 1)))

lambda_year
```

## Observed population growth versus modeled lambda

The validation uses quadrats observed in both $t_0$ and $t_1$. Each quadrat is
projected with its own disturbance status in $t_0$, but all quadrats share the
same year-specific demographic parameters. This prevents changing quadrat
sampling and partially disturbed sites from being interpreted as demographic
change.


```{r byyear_fl2x_lambda_validation, fig.width=8, fig.height=5}
# Observed quadrat-year abundance and disturbance ---------------------------
df_counts_quad <- df %>%
  filter(
    size_t0 > 0,
    is.finite(logsize_t0)) %>%
  group_by(site, quad_id, year) %>%
  summarise(
    n = n(),
    disturbance = as.integer(any(disturbance == 1, na.rm = TRUE)),
    .groups = 'drop')

# Pair each quadrat in t0 with the same quadrat in t1.
df_obs_pgr_quad <- df_counts_quad %>%
  rename(
    n_t0 = n,
    disturbance_t0 = disturbance) %>%
  left_join(
    df_counts_quad %>%
      transmute(
        site,
        quad_id,
        year = year - 1,
        n_t1 = n),
    by = c('site', 'quad_id', 'year')) %>%
  filter(
    !is.na(n_t1),
    year %in% ipm_years) %>%
  mutate(obs_pgr_quad = n_t1 / n_t0)

# Convert the observed quadrat size distribution to a mesh density.
make_initial_n_quad <- function(year_i, quad_i, pars) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  df_i <- df %>%
    filter(
      year == year_i,
      quad_id == quad_i,
      size_t0 > 0,
      is.finite(logsize_t0))

  counts_i <- hist(
    pmin(pmax(df_i$logsize_t0, L), U),
    breaks = seq(L, U, length.out = n + 1),
    include.lowest = TRUE,
    plot = FALSE)$counts

  counts_i / h
}

# Project one paired quadrat from t0 to t1.
project_one_quad_year <- function(year_i, quad_i, disturbance_i) {
  pars_i <- make_ipm_pars(year_i = year_i)

  n_obs <- make_initial_n_quad(
    year_i = year_i,
    quad_i = quad_i,
    pars = pars_i)

  kernel_i <- kernel(
    pars = pars_i,
    disturbance = disturbance_i)

  K <- kernel_i$k_yx
  h <- kernel_i$h
  n_proj <- K %*% n_obs

  n_obs_total <- sum(n_obs) * h
  n_proj_total <- sum(n_proj) * h

  tibble(
    year = year_i,
    quad_id = quad_i,
    disturbance_t0 = disturbance_i,
    n_obs_model = n_obs_total,
    n_proj_model = n_proj_total,
    asym_lambda = Re(eigen(K, only.values = TRUE)$values[1]),
    proj_lambda = n_proj_total / n_obs_total)
}

# Project all paired quadrat-years.
df_proj_quad <- bind_rows(
  lapply(seq_len(nrow(df_obs_pgr_quad)), function(i) {
    project_one_quad_year(
      year_i = df_obs_pgr_quad$year[i],
      quad_i = df_obs_pgr_quad$quad_id[i],
      disturbance_i = df_obs_pgr_quad$disturbance_t0[i])
  }))

# Combine observed and modeled quadrat transitions.
df_compare_quad <- df_obs_pgr_quad %>%
  left_join(
    df_proj_quad,
    by = c('year', 'quad_id', 'disturbance_t0'))

# Aggregate to whole-population annual growth.
df_compare <- df_compare_quad %>%
  group_by(year) %>%
  summarise(
    asym_lambda = weighted.mean(
      asym_lambda, w = n_obs_model, na.rm = TRUE),
    n_t0 = sum(n_t0, na.rm = TRUE),
    n_t1 = sum(n_t1, na.rm = TRUE),
    obs_pgr = n_t1 / n_t0,
    n_obs_model = sum(n_obs_model, na.rm = TRUE),
    n_proj_model = sum(n_proj_model, na.rm = TRUE),
    proj_lambda = n_proj_model / n_obs_model,
    disturbance = if_else(
      any(disturbance_t0 == 1, na.rm = TRUE), 1, 0),
    .groups = 'drop') %>%
  mutate(
    disturbance = factor(
      disturbance, levels = c(0, 1),
      labels = c('No fire', 'Fire')))

# Long format for asymptotic and one-step projected lambda.
df_plot <- df_compare %>%
  select(year, obs_pgr, asym_lambda, proj_lambda, disturbance) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Asymptotic lambda',
      proj_lambda = 'Projected lambda'))

g_mod_vs_obs <- ggplot(
  df_plot,
  aes(x = lambda, y = obs_pgr, color = disturbance)) +
  geom_point(size = 2.5) +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(
    values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = 'Observed population growth rate',
    color = NULL) +
  theme_classic()

g_mod_vs_obs
```

