# by Year by Site: one-part flowering, two-part fruiting, and fruit-to-recruit {#byyear_bysite_fl1x_fr2x_fr2re}

This chapter builds year- and site-specific integral projection models for
*Crotalaria avonensis*. Survival, growth, flowering probability, fruiting
probability, and fruit number vary among years. The fruit-to-recruit slope can
vary among both recruitment years and sites. Disturbance is a fixed effect and
is assigned to the transition year $t_0$.

The flowering component has one part: the probability that an individual
flowers. A separate flower-number model is not fitted because flower number does
not enter the current fertility kernel. The fruit component has two parts:

1. fruiting probability conditional on flowering;
2. fruit number conditional on fruiting.

The fertility pathway is

```r
Pr(flowering) * Pr(fruiting | flowering) *
  E(number of fruits | fruiting) * recruits per fruit
```

Site is used as a random effect only in the fruit-to-recruit relationship. The
other vital-rate models contain year random effects but no site random effects.
The site-year fruit-to-recruit model is forced through zero on the `log1p`
scale, so zero fruits predict zero recruits.

## Required objects

This chapter assumes that the extended-data chapter has created `df`, `df_og`,
and `v_ggp_suffix`. The quadrat-sampling table is rebuilt below after recruit
timing is corrected.

```{r byyear_bysite_fl1x_required_objects}
required_objects <- c('df', 'df_og', 'v_ggp_suffix')
missing_objects <- required_objects[!purrr::map_lgl(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop(
    'The following objects are missing. Run the data chapter first: ',
    paste(missing_objects, collapse = ', '))
}

if (!exists('v_years_re')) v_years_re <- c()
if (!exists('v_mod_set_su')) v_mod_set_su <- c()
if (!exists('v_mod_set_gr')) v_mod_set_gr <- c()
if (!exists('v_mod_set_fl')) v_mod_set_fl <- c()
if (!exists('v_mod_set_fr')) v_mod_set_fr <- c()
if (!exists('v_mod_set_fr_n')) v_mod_set_fr_n <- c()

# Fruit-to-recruit model index:
# 0 = fixed slope; 1 = year; 2 = site; 3 = year + site.
# The full year-and-site model is used unless this object was defined earlier.
if (!exists('v_mod_set_fr2re')) v_mod_set_fr2re <- 3

ctrl_glmer <- lme4::glmerControl(
  optimizer = 'bobyqa', optCtrl = list(maxfun = 2e5))

ctrl_lmer <- lme4::lmerControl(
  optimizer = 'bobyqa', optCtrl = list(maxfun = 2e5))
```

## Correct recruitment timing

The first-observation year must be calculated within each plant. The earlier data
chapter calculated it without grouping by plant, which can misclassify recruits.
The following block replaces `recruit` and rebuilds the previous-year quadrat
sampling check before fitting the fruit-to-recruit model.


```{r byyear_bysite_fl1x_recruit_correction}
df_recruit_corrected <- df_og %>%
  janitor::clean_names() %>%
  transmute(
    plant_id = as.character(id),
    year = as.integer(year),
    s,
    stage) %>%
  group_by(plant_id) %>%
  mutate(
    earliest_recorded_date = suppressWarnings(
      min(year[s > 0 & s != 2 & s != 6], na.rm = TRUE)),
    earliest_recorded_date = if_else(
      is.infinite(earliest_recorded_date),
      NA_real_, earliest_recorded_date),
    recruit = case_when(
      stage == 1 ~ 1,
      year == earliest_recorded_date & earliest_recorded_date > 2003 ~ 1,
      TRUE ~ NA_real_)) %>%
  ungroup() %>%
  select(plant_id, year, recruit)

df <- df %>%
  mutate(
    plant_id = as.character(plant_id),
    year = as.integer(as.character(year))) %>%
  select(-any_of(c('recruit', 'recruit_plot_valid'))) %>%
  left_join(df_recruit_corrected, by = c('plant_id', 'year')) %>%
  mutate(plant_id = factor(plant_id))

# Identify quadrats sampled in each year.
df_plot_year_sampled <- df %>%
  group_by(site, mp, quad, quad_id, year) %>%
  summarise(
    n_records_plot_year = n(),
    n_plants_plot_year = n_distinct(plant_id),
    .groups = 'drop')

# Retain recruits only when their quadrat was sampled in the previous year.
df_recruit_valid_ids <- df %>%
  filter(recruit == 1) %>%
  mutate(year_prev = year - 1) %>%
  left_join(
    df_plot_year_sampled %>%
      transmute(
        quad_id,
        year_prev = year,
        sampled_prev_year = TRUE),
    by = c('quad_id', 'year_prev')) %>%
  filter(replace_na(sampled_prev_year, FALSE)) %>%
  select(site, mp, quad, quad_id, plant_id, year) %>%
  mutate(recruit_plot_valid = 1)

df <- df %>%
  left_join(
    df_recruit_valid_ids,
    by = c('site', 'mp', 'quad', 'quad_id', 'plant_id', 'year')) %>%
  mutate(
    recruit_plot_valid = case_when(
      recruit == 1 & recruit_plot_valid == 1 ~ 1,
      TRUE ~ NA_real_))
```

## Process-specific data

The raw flower count is retained to identify flowering individuals, but it is
not modelled as a separate count response.

```{r byyear_bysite_fl1x_model_data}
# Survival data ----------------------------------------------------------------
df_su <- df %>%
  filter(
    size_t0 != 0,
    !is.na(survives),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, survives, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)


# Growth data ------------------------------------------------------------------
df_gr <- df %>%
  filter(
    size_t0 != 0,
    survives == 1,
    is.finite(logsize_t0),
    is.finite(logsize_t1),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)


# Flowering-probability data ---------------------------------------------------
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance),
    flower = if_else(flower > 0, 1, 0)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, size_t1,
    logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3,
    disturbance)


# Fruiting-probability data ----------------------------------------------------
# Fruiting probability is conditional on flowering because flower > 0.
df_fr <- df %>%
  filter(
    flower > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance),
    fruiting = if_else(fruit > 0, 1, 0)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, fruit, fruiting,
    logsize_t0, logsize_t0_2, logsize_t0_3, disturbance)


# Fruit-number data ------------------------------------------------------------
# Fruit number is conditional on fruiting because fruit > 0.
df_fr_n <- df %>%
  filter(
    fruit > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    year = factor(year),
    disturbance = as.numeric(disturbance)) %>%
  dplyr::select(
    plant_id, site, year, size_t0, flower, fruit,
    logsize_t0, logsize_t0_2, logsize_t0_3, disturbance)


# Fruit-to-recruit data --------------------------------------------------------
# Total fruits in year t predict recruits in year t + 1. Only quadrats sampled
# in consecutive years contribute to the site-year totals.
df_fr2re_quad <- df %>%
  mutate(year = as.integer(as.character(year))) %>%
  group_by(site, quad_id, year) %>%
  summarise(
    fruits_t0 = sum(fruit, na.rm = TRUE),
    recruits_t0 = sum(recruit_plot_valid == 1, na.rm = TRUE),
    .groups = 'drop') %>%
  arrange(site, quad_id, year) %>%
  group_by(site, quad_id) %>%
  mutate(
    recruits_t1 = lead(recruits_t0),
    year_t1 = lead(year),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_fr2re <- df_fr2re_quad %>%
  group_by(site, year, year_t1) %>%
  summarise(
    fruits_t0 = sum(fruits_t0, na.rm = TRUE),
    recruits_t1 = sum(recruits_t1, na.rm = TRUE),
    n_quads = n(),
    .groups = 'drop') %>%
  mutate(
    site = factor(site),
    year_t1 = factor(year_t1),
    logfruits_t0 = log1p(fruits_t0),
    logrecruits_t1 = log1p(recruits_t1))


# Remove excluded years --------------------------------------------------------
v_years_og <- sort(unique(df$year))

df <- df %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_su <- df_su %>%
  filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()

df_gr <- df_gr %>%
  filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()

df_fl <- df_fl %>%
  filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()

df_fr <- df_fr %>%
  filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()

df_fr_n <- df_fr_n %>%
  filter(!is.na(year), !(year %in% v_years_re)) %>%
  droplevels()

df_fr2re <- df_fr2re %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !(year %in% v_years_re),
    !(as.integer(as.character(year_t1)) %in% v_years_re)) %>%
  droplevels()

v_years <- sort(unique(df$year))
```

## Survival


```{r byyear_bysite_fl1x_survival_model}
# Survival model ---------------------------------------------------------------
# GLMM; binomial
mod_su_0 <- glmer(
  survives ~ disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mod_su_1 <- glmer(
  survives ~ logsize_t0 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mod_su_2 <- glmer(
  survives ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mod_su_3 <- glmer(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance + (1 | year),
  data = df_su, family = binomial, control = ctrl_glmer)

mods_su <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_best <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef <- coef(mod_su_best)$year

summary(mod_su_best)
```

```{r byyear_bysite_fl1x_survival_plot, fig.width=10, fig.height=12}
# Survival plots by year -------------------------------------------------------
make_su_year_plot <- function(year_i) {
  df_i <- df_su %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_su$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      survives = predict(
        mod_su_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      survives = mean(survives, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, survives, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, survives, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, survives, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Survival probability') +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

su_yrs <- lapply(levels(df_su$year), make_su_year_plot)

fig_su_years <- wrap_plots(su_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_su_years
```

## Growth


```{r byyear_bysite_fl1x_growth_model}
# Growth model -----------------------------------------------------------------
mod_gr_0 <- lmer(
  logsize_t1 ~ disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mod_gr_1 <- lmer(
  logsize_t1 ~ logsize_t0 + disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mod_gr_2 <- lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mod_gr_3 <- lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance + (logsize_t0 | year),
  data = df_gr, control = ctrl_lmer)

mods_gr <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAIC <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_best <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef <- coef(mod_gr_best)$year

summary(mod_gr_best)
```

```{r byyear_bysite_fl1x_growth_plot, fig.width=10, fig.height=12}
# Growth plots by year ---------------------------------------------------------
make_gr_year_plot <- function(year_i) {
  df_i <- df_gr %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_gr$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      logsize_t1 = predict(
        mod_gr_best, newdata = pred_i, re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      logsize_t1 = mean(logsize_t1, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, logsize_t1, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, logsize_t1, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, logsize_t1, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = expression('log(size)'[t1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

gr_yrs <- lapply(levels(df_gr$year), make_gr_year_plot)

fig_gr_years <- wrap_plots(gr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Growth - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_gr_years
```

```{r byyear_bysite_fl1x_growth_variance}
# Growth variance --------------------------------------------------------------
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
```

## Flowering probability


```{r byyear_bysite_fl1x_flower_model}
# Flower model -----------------------------------------------------------------
mod_fl_0 <- glmer(
  flower ~ disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mod_fl_1 <- glmer(
  flower ~ logsize_t0 + disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mod_fl_2 <- glmer(
  flower ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mod_fl_3 <- glmer(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance + (1 | year),
  data = df_fl, family = binomial, control = ctrl_glmer)

mods_fl <- list(mod_fl_0, mod_fl_1, mod_fl_2, mod_fl_3)
mods_fl_dAIC <- bbmle::AICctab(mods_fl, weights = TRUE, sort = FALSE)$dAIC
mods_fl_sorted <- order(mods_fl_dAIC)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_best <- mods_fl[[mod_fl_index_bestfit]]
mod_fl_ranef <- coef(mod_fl_best)$year

summary(mod_fl_best)
```

```{r byyear_bysite_fl1x_flower_plot, fig.width=10, fig.height=12}
# Flowering plots by year ------------------------------------------------------
make_fl_year_plot <- function(year_i) {
  df_i <- df_fl %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fl$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      flower = predict(
        mod_fl_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      flower = mean(flower, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, flower, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, flower, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, flower, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Flowering probability') +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fl_yrs <- lapply(levels(df_fl$year), make_fl_year_plot)

fig_fl_years <- wrap_plots(fl_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Flowering probability - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fl_years
```

## Fruiting probability conditional on flowering


```{r byyear_bysite_fl1x_fruiting_model}
# Fruiting-probability model ---------------------------------------------------
mod_fr_0 <- glmer(
  fruiting ~ disturbance + (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mod_fr_1 <- glmer(
  fruiting ~ logsize_t0 + disturbance + (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mod_fr_2 <- glmer(
  fruiting ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mod_fr_3 <- glmer(
  fruiting ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance +
    (1 | year),
  data = df_fr, family = binomial, control = ctrl_glmer)

mods_fr <- list(mod_fr_0, mod_fr_1, mod_fr_2, mod_fr_3)
mods_fr_dAIC <- bbmle::AICctab(
  mods_fr, weights = TRUE, sort = FALSE)$dAIC
mods_fr_sorted <- order(mods_fr_dAIC)

if (length(v_mod_set_fr) == 0) {
  mod_fr_index_bestfit <- mods_fr_sorted[1]
  v_mod_fr_index <- mod_fr_index_bestfit - 1
} else {
  mod_fr_index_bestfit <- v_mod_set_fr + 1
  v_mod_fr_index <- v_mod_set_fr
}

mod_fr_best <- mods_fr[[mod_fr_index_bestfit]]
mod_fr_ranef <- coef(mod_fr_best)$year

summary(mod_fr_best)
```

```{r byyear_bysite_fl1x_fruiting_plot, fig.width=10, fig.height=12}
# Fruiting-probability plots by year ------------------------------------------
make_fr_year_plot <- function(year_i) {
  df_i <- df_fr %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fr$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      fruiting = predict(
        mod_fr_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      fruiting = mean(fruiting, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, fruiting, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, fruiting, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, fruiting, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Fruiting probability') +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fr_yrs <- lapply(levels(df_fr$year), make_fr_year_plot)

fig_fr_years <- wrap_plots(fr_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Fruiting probability given flowering - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fr_years
```

## Fruit number conditional on fruiting


```{r byyear_bysite_fl1x_fruit_number_model}
# Fruit-number model -----------------------------------------------------------
mod_fr_n_0 <- glmer.nb(
  fruit ~ disturbance + (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mod_fr_n_1 <- glmer.nb(
  fruit ~ logsize_t0 + disturbance + (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mod_fr_n_2 <- glmer.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + disturbance + (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mod_fr_n_3 <- glmer.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance +
    (1 | year),
  data = df_fr_n, control = ctrl_glmer)

mods_fr_n <- list(mod_fr_n_0, mod_fr_n_1, mod_fr_n_2, mod_fr_n_3)
mods_fr_n_dAIC <- bbmle::AICctab(
  mods_fr_n, weights = TRUE, sort = FALSE)$dAIC
mods_fr_n_sorted <- order(mods_fr_n_dAIC)

if (length(v_mod_set_fr_n) == 0) {
  mod_fr_n_index_bestfit <- mods_fr_n_sorted[1]
  v_mod_fr_n_index <- mod_fr_n_index_bestfit - 1
} else {
  mod_fr_n_index_bestfit <- v_mod_set_fr_n + 1
  v_mod_fr_n_index <- v_mod_set_fr_n
}

mod_fr_n_best <- mods_fr_n[[mod_fr_n_index_bestfit]]
mod_fr_n_ranef <- coef(mod_fr_n_best)$year

summary(mod_fr_n_best)
```

```{r byyear_bysite_fl1x_fruit_number_plot, fig.width=10, fig.height=12}
# Fruit-number plots by year ---------------------------------------------------
make_fr_n_year_plot <- function(year_i) {
  df_i <- df_fr_n %>%
    filter(as.character(year) == year_i)
  
  x <- seq(
    min(df_i$logsize_t0, na.rm = TRUE),
    max(df_i$logsize_t0, na.rm = TRUE), length.out = 100)
  
  pred_i <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      year = factor(year_i, levels = levels(df_fr_n$year)),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire')))
  
  pred_i <- pred_i %>%
    mutate(
      fruit = predict(
        mod_fr_n_best, newdata = pred_i, type = 'response',
        re.form = NULL))
  
  pts_i <- df_i %>%
    mutate(
      bin = cut(logsize_t0, breaks = 8),
      disturbance_plot = factor(
        disturbance, levels = c(0, 1),
        labels = c('No fire', 'Fire'))) %>%
    group_by(bin, disturbance_plot) %>%
    summarise(
      logsize_t0 = mean(logsize_t0, na.rm = TRUE),
      fruit = mean(fruit, na.rm = TRUE),
      n = n(), .groups = 'drop') %>%
    filter(is.finite(logsize_t0), n > 0)
  
  ggplot() +
    geom_jitter(
      data = df_i %>%
        mutate(
          disturbance_plot = factor(
            disturbance, levels = c(0, 1),
            labels = c('No fire', 'Fire'))),
      aes(logsize_t0, fruit, color = disturbance_plot),
      alpha = 0.2, size = 0.7, width = 0.2, height = 0.2) +
    geom_point(
      data = pts_i,
      aes(logsize_t0, fruit, color = disturbance_plot),
      shape = 15, size = 1.5) +
    geom_line(
      data = pred_i,
      aes(logsize_t0, fruit, color = disturbance_plot),
      linewidth = 0.7) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    labs(
      title = year_i,
      x = expression('log(size)'[t0]),
      y = 'Number of fruits') +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fr_n_yrs <- lapply(levels(df_fr_n$year), make_fr_n_year_plot)

fig_fr_n_years <- wrap_plots(fr_n_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Fruit number given fruiting - year specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fr_n_years
```

## Fruit-to-recruit transition

The response is total recruits in site $i$ and year $t+1$, and the predictor is
total fruits from the consecutively sampled quadrats at the same site in year
$t$. All candidates are forced through zero on the `log1p` scale. Random effects
are random slopes rather than random intercepts.

```{r byyear_bysite_fl1x_fr2re_model}
# Fixed slope only.
mod_fr2re_0 <- glmmTMB(
  logrecruits_t1 ~ 0 + logfruits_t0,
  data = df_fr2re, family = gaussian, REML = FALSE)

# Year-specific random slope.
mod_fr2re_1 <- glmmTMB(
  logrecruits_t1 ~ 0 + logfruits_t0 +
    (0 + logfruits_t0 | year_t1),
  data = df_fr2re, family = gaussian, REML = FALSE)

# Site-specific random slope.
mod_fr2re_2 <- glmmTMB(
  logrecruits_t1 ~ 0 + logfruits_t0 +
    (0 + logfruits_t0 | site),
  data = df_fr2re, family = gaussian, REML = FALSE)

# Year- and site-specific random slopes.
mod_fr2re_3 <- glmmTMB(
  logrecruits_t1 ~ 0 + logfruits_t0 +
    (0 + logfruits_t0 | year_t1) +
    (0 + logfruits_t0 | site),
  data = df_fr2re, family = gaussian, REML = FALSE)

mods_fr2re <- list(
  mod_fr2re_0, mod_fr2re_1, mod_fr2re_2, mod_fr2re_3)

mods_fr2re_dAIC <- bbmle::AICctab(
  mods_fr2re, weights = TRUE, sort = FALSE)$dAIC
mods_fr2re_sorted <- order(mods_fr2re_dAIC)

if (length(v_mod_set_fr2re) == 0) {
  mod_fr2re_index_bestfit <- mods_fr2re_sorted[1]
  v_mod_fr2re_index <- mod_fr2re_index_bestfit - 1
} else {
  mod_fr2re_index_bestfit <- v_mod_set_fr2re + 1
  v_mod_fr2re_index <- v_mod_set_fr2re
}

mod_fr2re_best <- mods_fr2re[[mod_fr2re_index_bestfit]]
mod_fr2re_fixef <- fixef(mod_fr2re_best)$cond
mod_fr2re_ranef <- ranef(mod_fr2re_best)$cond

summary(mod_fr2re_best)

fr2re_model_diagnostics <- tibble(
  model_index = 0:3,
  model = c('fixed', 'year', 'site', 'year + site'),
  dAIC = as.numeric(mods_fr2re_dAIC),
  positive_definite_hessian = purrr::map_lgl(
    mods_fr2re, ~ isTRUE(.x$sdr$pdHess)))

fr2re_model_diagnostics
```

```{r byyear_bysite_fl1x_fr2re_plot, fig.width=10, fig.height=12}
make_fr2re_year_plot <- function(year_i) {
  df_i <- df_fr2re %>%
    filter(as.character(year_t1) == year_i)

  x <- seq(
    min(df_i$logfruits_t0, na.rm = TRUE),
    max(df_i$logfruits_t0, na.rm = TRUE), length.out = 100)

  pred_i <- expand_grid(
    logfruits_t0 = x,
    site = levels(df_fr2re$site)) %>%
    mutate(
      site = factor(site, levels = levels(df_fr2re$site)),
      year_t1 = factor(year_i, levels = levels(df_fr2re$year_t1)))

  pred_i$logrecruits_t1 <- predict(
    mod_fr2re_best,
    newdata = pred_i,
    type = 'response',
    re.form = NULL,
    allow.new.levels = TRUE)

  pred_i <- pred_i %>%
    mutate(recruits_t1 = pmax(0, expm1(logrecruits_t1)))

  ggplot() +
    geom_point(
      data = df_i,
      aes(logfruits_t0, recruits_t1, color = site),
      alpha = 0.55, size = 1.2) +
    geom_line(
      data = pred_i,
      aes(logfruits_t0, recruits_t1, color = site, group = site),
      linewidth = 0.7) +
    scale_color_viridis_d() +
    labs(
      title = year_i,
      x = expression('log(number of fruits + 1)'[t]),
      y = expression('Number of recruits'[t+1])) +
    theme_bw() +
    theme(text = element_text(size = 5), legend.position = 'none')
}

fr2re_yrs <- lapply(
  levels(df_fr2re$year_t1), make_fr2re_year_plot)

fig_fr2re_years <- wrap_plots(fr2re_yrs) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = 'Fruit-to-recruit relationship - year and site specific',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 13, face = 'bold'),
      plot.subtitle = element_text(size = 9)))

fig_fr2re_years
```

## Recruit size distribution

```{r byyear_bysite_fl1x_recruit_size}
df_recr <- df %>%
  filter(
    recruit_plot_valid == 1,
    size_t0 > 0,
    is.finite(logsize_t0))

rc_sz <- df_recr %>%
  summarise(
    mean = mean(logsize_t0, na.rm = TRUE),
    sd = sd(logsize_t0, na.rm = TRUE))

rc_sz
```

## Parameter extraction

The parameter flow has three levels:

1. `pars_mean` contains fixed effects and IPM constants;
2. `pars_year` contains conditional year coefficients and the recruitment-year
   deviation with the site deviation set to zero;
3. `pars_site_year_recruit` adds the site and recruitment-year deviations to the
   fixed fruit-to-recruit slope.

```{r byyear_bysite_fl1x_parameter_helpers}
get_model_fixef <- function(model) {
  if (inherits(model, 'glmmTMB')) {
    return(fixef(model)$cond)
  }

  if (inherits(model, 'merMod')) {
    return(lme4::fixef(model))
  }

  stats::coef(model)
}

get_group_coef <- function(model, group_var) {
  if (inherits(model, 'glmmTMB')) {
    coef_list <- coef(model)$cond
  } else if (inherits(model, 'merMod')) {
    coef_list <- coef(model)
  } else {
    return(NULL)
  }

  if (is.null(coef_list) || !(group_var %in% names(coef_list))) {
    return(NULL)
  }

  as.data.frame(coef_list[[group_var]])
}

get_group_ranef <- function(model, group_var) {
  if (inherits(model, 'glmmTMB')) {
    ranef_list <- ranef(model)$cond
  } else if (inherits(model, 'merMod')) {
    ranef_list <- lme4::ranef(model)
  } else {
    return(NULL)
  }

  if (is.null(ranef_list) || !(group_var %in% names(ranef_list))) {
    return(NULL)
  }

  as.data.frame(ranef_list[[group_var]])
}

size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logsize_t0',
  b2 = 'logsize_t0_2',
  b3 = 'logsize_t0_3',
  bd = 'disturbance')

extract_terms <- function(coefs, prefix, term_map = size_term_map) {
  values <- vapply(term_map, function(term_i) {
    if (term_i %in% names(coefs)) {
      unname(coefs[[term_i]])
    } else {
      0
    }
  }, numeric(1))

  names(values) <- paste0(prefix, names(term_map))
  values
}

extract_fixed_rate <- function(model, prefix) {
  extract_terms(
    coefs = get_model_fixef(model),
    prefix = prefix)
}

extract_year_rate <- function(
    model, prefix, years, group_var = 'year') {
  fixed_coef <- get_model_fixef(model)
  group_coef <- get_group_coef(model, group_var)

  purrr::map_dfr(years, function(year_i) {
    level_i <- as.character(year_i)
    coef_i <- fixed_coef

    if (!is.null(group_coef) && level_i %in% rownames(group_coef)) {
      coef_i <- unlist(
        group_coef[level_i, , drop = FALSE],
        use.names = FALSE)
      names(coef_i) <- colnames(group_coef)
    }

    tibble::as_tibble_row(c(
      year = year_i,
      extract_terms(coef_i, prefix)))
  })
}

extract_fr2re_deviation <- function(
    model, group_var, levels_vec, output_name) {
  out <- tibble(
    level = as.character(levels_vec),
    deviation = 0)

  ranef_i <- get_group_ranef(model, group_var)

  if (!is.null(ranef_i) && 'logfruits_t0' %in% names(ranef_i)) {
    ranef_i <- tibble(
      level = rownames(ranef_i),
      deviation = ranef_i[['logfruits_t0']])

    out <- out %>%
      select(-deviation) %>%
      left_join(ranef_i, by = 'level') %>%
      mutate(deviation = replace_na(deviation, 0))
  }

  names(out) <- c(group_var, output_name)
  out
}
```

```{r byyear_bysite_fl1x_mean_parameters}
grow_var_coef <- stats::coef(mod_gr_var)

mesh_limits <- range(
  c(
    df_gr$logsize_t0,
    df_gr$logsize_t1,
    df_recr$logsize_t0),
  na.rm = TRUE,
  finite = TRUE)

pars_mean <- as.list(c(
  extract_fixed_rate(mod_su_best, 'surv_'),
  extract_fixed_rate(mod_gr_best, 'grow_'),
  extract_fixed_rate(mod_fl_best, 'fl_'),
  extract_fixed_rate(mod_fr_best, 'fr_'),
  extract_fixed_rate(mod_fr_n_best, 'frn_'),
  fr2re_b0 = 0,
  fr2re_b1 = unname(
    get_model_fixef(mod_fr2re_best)[['logfruits_t0']]),
  recr_mean = rc_sz$mean,
  recr_sd = rc_sz$sd,
  grow_var_a = unname(grow_var_coef[['a']]),
  grow_var_b = unname(grow_var_coef[['b']]),
  fruits_ref = mean(df_fr2re$fruits_t0, na.rm = TRUE),
  fr2re_sigma = stats::sigma(mod_fr2re_best),
  L = mesh_limits[1],
  U = mesh_limits[2],
  mat_siz = 200))

pars_mean_table <- tibble(
  parameter = names(pars_mean),
  value = as.numeric(unlist(pars_mean)))

pars_mean_table
```

```{r byyear_bysite_fl1x_year_site_parameters}
get_year_values <- function(x) {
  sort(unique(as.integer(as.character(x))))
}

ipm_years <- Reduce(
  intersect,
  list(
    get_year_values(df_su$year),
    get_year_values(df_gr$year),
    get_year_values(df_fl$year),
    get_year_values(df_fr$year),
    get_year_values(df_fr_n$year),
    get_year_values(df_fr2re$year_t1) - 1))

if (length(ipm_years) == 0) {
  stop('No years have complete data for all IPM processes')
}

# Conditional year coefficients for plant vital rates.
pars_su_year <- extract_year_rate(
  model = mod_su_best, prefix = 'surv_', years = ipm_years)

pars_gr_year <- extract_year_rate(
  model = mod_gr_best, prefix = 'grow_', years = ipm_years)

pars_fl_year <- extract_year_rate(
  model = mod_fl_best, prefix = 'fl_', years = ipm_years)

pars_fr_year <- extract_year_rate(
  model = mod_fr_best, prefix = 'fr_', years = ipm_years)

pars_fr_n_year <- extract_year_rate(
  model = mod_fr_n_best, prefix = 'frn_', years = ipm_years)

# Site and recruitment-year random-slope deviations.
pars_fr2re_site <- extract_fr2re_deviation(
  model = mod_fr2re_best,
  group_var = 'site',
  levels_vec = levels(df_fr2re$site),
  output_name = 'site_u1') %>%
  mutate(site = factor(site, levels = levels(df_fr2re$site)))

pars_fr2re_year <- extract_fr2re_deviation(
  model = mod_fr2re_best,
  group_var = 'year_t1',
  levels_vec = levels(df_fr2re$year_t1),
  output_name = 'year_u1') %>%
  mutate(
    year_t1 = as.integer(as.character(year_t1)),
    year = year_t1 - 1,
    fr2re_b1 = pars_mean$fr2re_b1 + year_u1) %>%
  select(year, year_t1, year_u1, fr2re_b1)

# One readable parameter row per transition year.
pars_year <- list(
  pars_su_year,
  pars_gr_year,
  pars_fl_year,
  pars_fr_year,
  pars_fr_n_year,
  pars_fr2re_year %>% select(-year_t1)) %>%
  purrr::reduce(left_join, by = 'year') %>%
  arrange(year)

# One fruit-to-recruit slope per site-year transition.
pars_site_year_recruit <- expand_grid(
  site = levels(df_fr2re$site),
  year = ipm_years) %>%
  mutate(
    site = factor(site, levels = levels(df_fr2re$site)),
    year_t1 = year + 1) %>%
  left_join(pars_fr2re_site, by = 'site') %>%
  left_join(
    pars_fr2re_year %>% select(year, year_t1, year_u1),
    by = c('year', 'year_t1')) %>%
  mutate(
    site_u1 = replace_na(site_u1, 0),
    year_u1 = replace_na(year_u1, 0),
    fr2re_b0 = 0,
    fr2re_b1 = pars_mean$fr2re_b1 + site_u1 + year_u1) %>%
  select(
    site, year, year_t1,
    fr2re_b0, fr2re_b1,
    site_u1, year_u1)

make_ipm_pars <- function(year_i = NULL, site_i = NULL) {
  pars <- pars_mean

  if (!is.null(year_i)) {
    year_i <- as.integer(year_i)

    pars_i <- pars_year %>%
      filter(.data$year == .env$year_i)

    if (nrow(pars_i) != 1) {
      stop(
        'Expected one parameter row for year ', year_i,
        ', but found ', nrow(pars_i))
    }

    pars_i <- pars_i %>%
      select(-year, -any_of('year_u1')) %>%
      as.list()

    for (par_name in names(pars_i)) {
      pars[[par_name]] <- pars_i[[par_name]]
    }
  }

  if (!is.null(site_i)) {
    if (is.null(year_i)) {
      stop('site_i requires year_i')
    }

    site_i <- as.character(site_i)

    pars_re_i <- pars_site_year_recruit %>%
      filter(
        as.character(.data$site) == .env$site_i,
        .data$year == .env$year_i)

    if (nrow(pars_re_i) != 1) {
      stop(
        'Expected one recruitment row for site ', site_i,
        ' and year ', year_i, ', but found ', nrow(pars_re_i))
    }

    pars$fr2re_b0 <- pars_re_i$fr2re_b0
    pars$fr2re_b1 <- pars_re_i$fr2re_b1
    pars$site_u1 <- pars_re_i$site_u1
    pars$year_u1 <- pars_re_i$year_u1
  }

  pars
}

pars_year
pars_site_year_recruit
```

## IPM functions

### Back-transformation of the fruit-to-recruit relationship

The relationship between fruit production and recruitment was fitted after
transforming both variables using `log(1 + count)`. Predictions from this model
therefore had to be converted back to the original scale before they could be
used in the fertility component of the IPM.

We initially included a correction for the residual variation of the model
during this back-transformation. However, this correction produced a small
positive number of recruits even when predicted fruit production was close to
zero. The IPM subsequently converts total recruitment into the number of
recruits produced per fruit. Dividing a positive recruitment estimate by an
extremely small number of fruits resulted in unrealistically high estimates of
recruitment per fruit and strongly inflated estimates of population growth.

We therefore used the direct inverse transformation,

\[
\widehat{R} = \exp(\widehat{\eta}) - 1,
\]

where \(\widehat{\eta}\) is the predicted value on the transformed scale and
\(\widehat{R}\) is the predicted number of recruits. Predictions below zero
were set to zero.

This approach preserves the biologically necessary relationship that no fruit
production results in no recruitment. It also ensures that predicted
recruitment approaches zero as fruit production approaches zero. The resulting
values should be interpreted as the central predictions from the fitted
fruit-to-recruit relationship rather than as residual-variance-adjusted means.
The fitted relationship itself, including its site and year effects, was not
changed; only the conversion of its predictions back to the original count
scale was modified.

```{r byyear_bysite_fl1x_ipm_functions}
get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    pars[[par_name]]
  } else {
    default
  }
}

vital_lp <- function(x, pars, prefix, disturbance = 0) {
  get_par(pars, paste0(prefix, 'b0')) +
    get_par(pars, paste0(prefix, 'b1')) * x +
    get_par(pars, paste0(prefix, 'b2')) * x^2 +
    get_par(pars, paste0(prefix, 'b3')) * x^3 +
    get_par(pars, paste0(prefix, 'bd')) * disturbance
}

# Survival.
sx <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x, pars = pars,
    prefix = 'surv_', disturbance = disturbance))
}

# Growth mean and standard deviation.
grow_mu <- function(x, pars, disturbance = 0) {
  vital_lp(
    x = x, pars = pars,
    prefix = 'grow_', disturbance = disturbance)
}

grow_sd <- function(mean_value, pars) {
  variance <- get_par(pars, 'grow_var_a') *
    exp(get_par(pars, 'grow_var_b') * mean_value)

  sqrt(pmax(variance, .Machine$double.eps))
}

gxy <- function(x, y, pars, disturbance = 0) {
  mean_value <- grow_mu(
    x = x, pars = pars, disturbance = disturbance)

  sd_value <- grow_sd(
    mean_value = mean_value, pars = pars)

  dnorm(y, mean = mean_value, sd = sd_value)
}

# Flowering probability.
fl_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x, pars = pars,
    prefix = 'fl_', disturbance = disturbance))
}

# Fruiting probability conditional on flowering.
fr_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x, pars = pars,
    prefix = 'fr_', disturbance = disturbance))
}

# Fruit number conditional on fruiting.
frn_x <- function(x, pars, disturbance = 0) {
  exp(vital_lp(
    x = x, pars = pars,
    prefix = 'frn_', disturbance = disturbance))
}

# Expected total fruits per individual.
fruits_x <- function(x, pars, disturbance = 0) {
  fl_x(x, pars, disturbance) *
    fr_x(x, pars, disturbance) *
    frn_x(x, pars, disturbance)
}

# Site-year fruit-to-recruit conversion.
predict_recruits_from_fruits <- function(fruits, pars) {
  if (length(fruits) == 1 &&
      (!is.finite(fruits) || fruits <= 0)) {
    return(0)
  }

  eta <- get_par(pars, 'fr2re_b0') +
    get_par(pars, 'fr2re_b1') * log1p(pmax(fruits, 0))

  recruits <- pmax(expm1(eta), 0)

  ifelse(fruits <= 0, 0, recruits)
}

site_year_recruits_per_fruit <- function(pars) {
  fruits_ref <- get_par(
    pars, 'site_fruits_ref',
    get_par(pars, 'fruits_ref', 0))

  if (!is.finite(fruits_ref) || fruits_ref <= 0) {
    return(0)
  }

  predict_recruits_from_fruits(fruits_ref, pars) / fruits_ref
}

rx <- function(x, pars, disturbance = 0) {
  fruits_x(x, pars, disturbance) *
    site_year_recruits_per_fruit(pars)
}

# Recruit-size distribution.
recr_y <- function(y, pars) {
  dnorm(
    y,
    mean = get_par(pars, 'recr_mean'),
    sd = get_par(pars, 'recr_sd'))
}

# Fertility matrix.
fy <- function(y, x, pars, h, disturbance = 0) {
  Pvec <- recr_y(y, pars)
  recruit_mass <- sum(Pvec * h)

  if (!is.finite(recruit_mass) || recruit_mass <= 0) {
    stop('Recruit size distribution has no mass inside the mesh')
  }

  Pvec <- Pvec / recruit_mass
  Rvec <- rx(x, pars, disturbance)

  outer(Pvec, Rvec) * h
}
```

## Kernel

```{r byyear_bysite_fl1x_kernel}
kernel <- function(pars, disturbance = 0) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  Fmat <- fy(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance)

  Smat <- sx(
    x = y,
    pars = pars,
    disturbance = disturbance)

  Gmat <- matrix(0, n, n)
  Gmat[] <- t(
    outer(
      y,
      y,
      Vectorize(function(x, y) {
        gxy(
          x = x,
          y = y,
          pars = pars,
          disturbance = disturbance)
      }))) * h

  Tmat <- matrix(0, n, n)

  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  k_yx <- Fmat + Tmat

  list(
    k_yx = k_yx,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Smat = Smat,
    meshpts = y,
    h = h)
}

lambda_ipm <- function(pars, disturbance = 0) {
  K <- kernel(pars, disturbance)$k_yx
  Re(eigen(K, only.values = TRUE)$values[1])
}
```

## Mean and year-specific population growth

These values use the mean site, meaning that the site recruitment deviation is
zero. Site-specific values are calculated in the next section.

```{r byyear_bysite_fl1x_lambda}
lambda_mean_no_fire <- lambda_ipm(
  pars = pars_mean, disturbance = 0)

lambda_mean_fire <- lambda_ipm(
  pars = pars_mean, disturbance = 1)

lambda_ipm_year <- function(year, disturbance = 0) {
  pars_i <- make_ipm_pars(year_i = year)
  lambda_ipm(pars_i, disturbance)
}

lambda_year <- tibble(
  year = ipm_years,
  lambda_no_fire = purrr::map_dbl(
    ipm_years,
    ~ lambda_ipm_year(.x, disturbance = 0)),
  lambda_fire = purrr::map_dbl(
    ipm_years,
    ~ lambda_ipm_year(.x, disturbance = 1)))

lambda_mean_no_fire
lambda_mean_fire
lambda_year
```

## Site-year projections

The validation retains only quadrats sampled in both $t_0$ and $t_1$. Within a
site-year transition, the initial size distribution and observed abundance are
therefore based on the same quadrats. The proportion disturbed is calculated
from those paired quadrats in $t_0$.

For partially disturbed sites, the chapter builds no-fire and fire kernels and
combines them using the proportion of paired quadrats disturbed. This treats the
site as a disturbance regime while retaining its site-specific recruitment
slope.

```{r byyear_bysite_fl1x_site_year_projection}
# Sampled quadrat-year combinations, including sampled quadrats with no sized
# individuals.
df_sampled_quad_year <- df_plot_year_sampled %>%
  transmute(
    site = factor(site),
    quad_id = factor(quad_id),
    year = as.integer(year)) %>%
  distinct()

# Counts of sized plants, completed with zeros for sampled quadrats.
df_sized_quad_year <- df %>%
  filter(size_t0 > 0, is.finite(logsize_t0)) %>%
  mutate(year = as.integer(as.character(year))) %>%
  group_by(site, quad_id, year) %>%
  summarise(n = n_distinct(plant_id), .groups = 'drop')

df_quad_counts <- df_sampled_quad_year %>%
  left_join(
    df_sized_quad_year,
    by = c('site', 'quad_id', 'year')) %>%
  mutate(n = replace_na(n, 0))

# Pair the same quadrat in consecutive years.
df_paired_quad_year <- df_quad_counts %>%
  rename(n_t0_quad = n) %>%
  inner_join(
    df_quad_counts %>%
      transmute(
        site,
        quad_id,
        year_t1 = year,
        year = year_t1 - 1,
        n_t1_quad = n),
    by = c('site', 'quad_id', 'year')) %>%
  filter(year %in% ipm_years)

# Observed site-year population growth from paired quadrats.
df_obs_pgr_site <- df_paired_quad_year %>%
  group_by(site, year, year_t1) %>%
  summarise(
    n_t0 = sum(n_t0_quad),
    n_t1 = sum(n_t1_quad),
    n_paired_quads = n_distinct(quad_id),
    .groups = 'drop') %>%
  filter(n_t0 > 0) %>%
  mutate(obs_pgr_site = n_t1 / n_t0)

# Disturbance status of paired quadrats in t0.
df_disturbance_paired <- df %>%
  mutate(year = as.integer(as.character(year))) %>%
  group_by(site, quad_id, year) %>%
  summarise(
    disturbance = as.integer(any(disturbance == 1, na.rm = TRUE)),
    .groups = 'drop') %>%
  inner_join(
    df_paired_quad_year %>%
      select(site, quad_id, year, year_t1),
    by = c('site', 'quad_id', 'year')) %>%
  group_by(site, year, year_t1) %>%
  summarise(
    p_disturbance = mean(disturbance, na.rm = TRUE),
    .groups = 'drop')

get_paired_quads <- function(site_i, year_i, year_t1_i) {
  df_paired_quad_year %>%
    filter(
      as.character(site) == as.character(site_i),
      year == as.integer(year_i),
      year_t1 == as.integer(year_t1_i)) %>%
    pull(quad_id) %>%
    as.character()
}

make_initial_n_site_year <- function(
    pars, site_i, year_i, year_t1_i, df_init = df) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  paired_quads <- get_paired_quads(
    site_i = site_i,
    year_i = year_i,
    year_t1_i = year_t1_i)

  x <- df_init %>%
    filter(
      as.character(site) == as.character(site_i),
      as.integer(as.character(year)) == as.integer(year_i),
      as.character(quad_id) %in% paired_quads,
      size_t0 > 0,
      is.finite(logsize_t0)) %>%
    pull(logsize_t0)

  if (length(x) == 0) {
    return(rep(0, n))
  }

  counts_i <- hist(
    pmin(pmax(x, L), U),
    breaks = seq(L, U, length.out = n + 1),
    include.lowest = TRUE,
    plot = FALSE)$counts

  counts_i / h
}

expected_site_year_fruits <- function(
    pars, site_i, year_i, year_t1_i,
    disturbance = 0, df_init = df) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  n_obs <- make_initial_n_site_year(
    pars = pars,
    site_i = site_i,
    year_i = year_i,
    year_t1_i = year_t1_i,
    df_init = df_init)

  sum(
    fruits_x(y, pars, disturbance) * n_obs * h,
    na.rm = TRUE)
}

project_site_year_ipm <- function(
    site_i, year_i, year_t1_i,
    p_disturbance_i, df_init = df) {
  pars_i <- make_ipm_pars(
    year_i = year_i,
    site_i = site_i)

  n_obs <- make_initial_n_site_year(
    pars = pars_i,
    site_i = site_i,
    year_i = year_i,
    year_t1_i = year_t1_i,
    df_init = df_init)

  h <- (pars_i$U - pars_i$L) / pars_i$mat_siz
  n_initial <- sum(n_obs) * h

  if (!is.finite(n_initial) || n_initial <= 0) {
    return(tibble(
      asym_lambda = NA_real_,
      proj_lambda = NA_real_,
      n_obs_model = n_initial,
      n_proj_model = NA_real_,
      site_fruits_no_fire = NA_real_,
      site_fruits_fire = NA_real_,
      site_fruits_ref = NA_real_,
      recruits_per_fruit = NA_real_))
  }

  site_fruits_no_fire <- expected_site_year_fruits(
    pars = pars_i,
    site_i = site_i,
    year_i = year_i,
    year_t1_i = year_t1_i,
    disturbance = 0,
    df_init = df_init)

  site_fruits_fire <- expected_site_year_fruits(
    pars = pars_i,
    site_i = site_i,
    year_i = year_i,
    year_t1_i = year_t1_i,
    disturbance = 1,
    df_init = df_init)

  site_fruits_ref <-
    (1 - p_disturbance_i) * site_fruits_no_fire +
    p_disturbance_i * site_fruits_fire

  pars_i$site_fruits_ref <- site_fruits_ref

  K_no_fire <- kernel(pars_i, disturbance = 0)$k_yx
  K_fire <- kernel(pars_i, disturbance = 1)$k_yx

  K_regime <-
    (1 - p_disturbance_i) * K_no_fire +
    p_disturbance_i * K_fire

  n_proj <- K_regime %*% n_obs
  n_projected <- sum(n_proj) * h

  tibble(
    asym_lambda = Re(
      eigen(K_regime, only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial,
    n_obs_model = n_initial,
    n_proj_model = n_projected,
    site_fruits_no_fire = site_fruits_no_fire,
    site_fruits_fire = site_fruits_fire,
    site_fruits_ref = site_fruits_ref,
    recruits_per_fruit = site_year_recruits_per_fruit(pars_i))
}

# Site-year projection grid.
df_site_year_projection_grid <- df_obs_pgr_site %>%
  left_join(
    pars_site_year_recruit,
    by = c('site', 'year', 'year_t1')) %>%
  left_join(
    df_disturbance_paired,
    by = c('site', 'year', 'year_t1')) %>%
  mutate(p_disturbance = replace_na(p_disturbance, 0)) %>%
  filter(!is.na(fr2re_b1))

# Run all site-year IPMs.
df_proj_site <- df_site_year_projection_grid %>%
  mutate(
    projection = purrr::pmap(
      list(site, year, year_t1, p_disturbance),
      function(site, year, year_t1, p_disturbance) {
        project_site_year_ipm(
          site_i = site,
          year_i = year,
          year_t1_i = year_t1,
          p_disturbance_i = p_disturbance,
          df_init = df)
      })) %>%
  select(
    site, year, year_t1,
    n_t0, n_t1, n_paired_quads, obs_pgr_site,
    fr2re_b0, fr2re_b1, site_u1, year_u1,
    p_disturbance, projection) %>%
  tidyr::unnest(projection)

df_compare_site <- df_proj_site %>%
  mutate(
    error_asymptotic = asym_lambda - obs_pgr_site,
    error_projected = proj_lambda - obs_pgr_site)

df_compare_site
```

## Observed population growth versus modeled lambda

The final figure compares observed and modeled population growth for each
site-year transition. Each point represents the change in population size at
one site between two consecutive years. Sites are distinguished by color.

Observed population growth was calculated from the change in the number of
plants between year \(t\) and year \(t+1\). Only quadrats represented in both
years were included so that differences in sampling effort were not mistaken
for changes in population size. A value of \(\lambda = 1\) indicates that
population size remained stable. Values greater than one indicate population
growth, whereas values below one indicate population decline.

The horizontal axis shows the population growth rate predicted by the IPM, and
the vertical axis shows the corresponding observed growth rate. The dashed
one-to-one line represents perfect agreement between the model and the
observations. Points close to this line indicate that the modeled and observed
growth rates were similar. Points above the line indicate that observed growth
was greater than predicted, meaning that the model underestimated population
growth. Points below the line indicate that the model predicted greater growth
than was observed.

The two panels show different measures of modeled population growth.
Asymptotic population growth describes the long-term growth rate expected if
the demographic conditions estimated for that site and year remained constant
and the population eventually reached a stable size distribution. It should
therefore be interpreted as a long-term property of the site-year kernel rather
than as a direct prediction of the next census.

Projected population growth describes the expected change over a single year
after applying the site-year kernel to the size distribution that was actually
observed at the beginning of the transition. Because it retains the observed
population structure, projected growth provides the more direct comparison
with the observed change from year \(t\) to year \(t+1\).

Some disagreement between observed and modeled growth is expected. The IPM
represents the demographic processes included in the fitted vital-rate models,
whereas observed population change may also be influenced by demographic
chance, observation error, temporary dormancy, and environmental conditions
not measured in the study. The figure is therefore used to assess whether the
model captures the general direction and magnitude of variation in population
growth among sites and years, rather than whether every modeled value exactly
matches its corresponding observation.

```{r byyear_bysite_fl1x_lambda_validation, fig.width=8, fig.height=5}
# Aggregate site projections into one whole-population result per year.
df_compare <- df_compare_site %>%
  group_by(year) %>%
  summarise(
    asym_lambda = weighted.mean(
      asym_lambda, w = n_obs_model, na.rm = TRUE),
    n_t0 = sum(n_t0, na.rm = TRUE),
    n_t1 = sum(n_t1, na.rm = TRUE),
    obs_pgr = n_t1 / n_t0,
    n_obs_model = sum(n_obs_model, na.rm = TRUE),
    n_proj_model = sum(n_proj_model, na.rm = TRUE),
    proj_lambda = n_proj_model / n_obs_model,
    disturbance = if_else(
      any(p_disturbance > 0, na.rm = TRUE), 'Fire', 'No fire'),
    .groups = 'drop') %>%
  mutate(
    disturbance = factor(
      disturbance, levels = c('No fire', 'Fire')))

df_plot <- df_compare %>%
  select(year, obs_pgr, asym_lambda, proj_lambda, disturbance) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Asymptotic lambda',
      proj_lambda = 'Projected lambda'))

g_mod_vs_obs <- ggplot(
  df_plot,
  aes(x = lambda, y = obs_pgr, color = disturbance)) +
  geom_point(size = 2.5) +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  facet_wrap(~ lambda_type, scales = 'free_x') +
  scale_color_manual(
    values = c('No fire' = 'black', 'Fire' = 'red')) +
  labs(
    title = 'Observed population growth vs modeled lambda',
    subtitle = paste(v_ggp_suffix, '- site-specific recruitment'),
    x = expression('Modeled ' * lambda),
    y = 'Observed population growth rate',
    color = NULL) +
  theme_classic()

g_mod_vs_obs
```

```{r byyear_bysite_fl1x_site_validation, fig.width=8, fig.height=5}
df_plot_site <- df_compare_site %>%
  select(
    site, year, year_t1,
    obs_pgr_site,
    asym_lambda,
    proj_lambda) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Asymptotic lambda',
      proj_lambda = 'Projected lambda'))

fig_mod_vs_obs_site <- ggplot(
  df_plot_site,
  aes(x = lambda, y = obs_pgr_site, color = site)) +
  geom_point(alpha = 0.65, size = 1.8) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  facet_wrap(~ lambda_type, scales = 'free') +
  scale_color_viridis_d() +
  theme_bw() +
  labs(
    title = 'Observed versus modeled site-year population growth',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = expression('Observed ' * lambda),
    color = 'Site')

fig_mod_vs_obs_site
```

