# Mean IPM: one-part flowering, two-part fruiting, fruit-to-recruit, and disturbance {#mean_flx1_frx2_fr2re_dist}

This chapter builds a mean integral projection model for *Crotalaria
avonensis*. All vital-rate models contain fixed effects only: no year random
effects and no site random effects are included. Disturbance remains a fixed
predictor of survival, growth, flowering probability, fruiting probability, and
fruit number. The model is evaluated for no fire, fire, and the observed mean
disturbance regime.

The flowering component has one part, the probability that an individual
flowers. Flower number is retained in the raw data only to distinguish flowering
from nonflowering plants; a separate flower-number model is not fitted. Fruit
production has two parts:

1. fruiting probability conditional on flowering;
2. fruit number conditional on fruiting.

The expected reproductive contribution of an individual is therefore

```r
Pr(flowering) * Pr(fruiting | flowering) *
  E(number of fruits | fruiting) * recruits per fruit
```

The fruit-to-recruit model relates total fruits in year $t$ to valid recruits in
year $t+1$. It is fitted on the `log1p` scale, forced through zero, and contains
no site or year effects.

## Required objects

This chapter assumes that the data chapter has created `df`, `df_og`, and
`v_ggp_suffix`.

```{r mean_flx1_frx2_required_objects}
required_objects <- c('df', 'df_og', 'v_ggp_suffix')
missing_objects <- required_objects[!purrr::map_lgl(required_objects, exists)]

if (length(missing_objects) > 0) {
  stop(
    'The following objects are missing. Run the data chapter first: ',
    paste(missing_objects, collapse = ', '))
}

if (!exists('v_years_re')) v_years_re <- c()

# Empty vectors select the best-supported candidate by AICc. To force a
# candidate, provide its zero-based position in the corresponding model list.
v_mod_set_su <- c()
v_mod_set_gr <- c()
v_mod_set_fl <- c()
v_mod_set_fr <- c()
v_mod_set_fr_n <- c()
```

## Correct recruitment timing

The first recorded year is calculated separately for each plant. Recruits are
then retained for the fruit-to-recruit analysis only when their quadrat was
represented in the preceding year. This reduces the risk of treating the first
observation from a newly sampled quadrat as a known recruit.

```{r mean_flx1_frx2_recruit_correction}
df_recruit_corrected <- df_og %>%
  janitor::clean_names() %>%
  transmute(
    plant_id = as.character(id),
    year = as.integer(year),
    s,
    stage) %>%
  group_by(plant_id) %>%
  mutate(
    earliest_recorded_date = suppressWarnings(
      min(year[s > 0 & s != 2 & s != 6], na.rm = TRUE)),
    earliest_recorded_date = if_else(
      is.infinite(earliest_recorded_date),
      NA_real_, earliest_recorded_date),
    recruit = case_when(
      stage == 1 ~ 1,
      year == earliest_recorded_date & earliest_recorded_date > 2003 ~ 1,
      TRUE ~ NA_real_)) %>%
  ungroup() %>%
  select(plant_id, year, recruit)

df <- df %>%
  mutate(
    plant_id = as.character(plant_id),
    year = as.integer(as.character(year))) %>%
  select(-any_of(c('recruit', 'recruit_plot_valid'))) %>%
  left_join(df_recruit_corrected, by = c('plant_id', 'year')) %>%
  mutate(plant_id = factor(plant_id))

df_plot_year_sampled <- df %>%
  group_by(site, mp, quad, quad_id, year) %>%
  summarise(
    n_records_plot_year = n(),
    n_plants_plot_year = n_distinct(plant_id),
    .groups = 'drop')

df_recruit_valid_ids <- df %>%
  filter(recruit == 1) %>%
  mutate(year_prev = year - 1) %>%
  left_join(
    df_plot_year_sampled %>%
      transmute(
        quad_id,
        year_prev = year,
        sampled_prev_year = TRUE),
    by = c('quad_id', 'year_prev')) %>%
  filter(replace_na(sampled_prev_year, FALSE)) %>%
  select(site, mp, quad, quad_id, plant_id, year) %>%
  mutate(recruit_plot_valid = 1)

df <- df %>%
  left_join(
    df_recruit_valid_ids,
    by = c('site', 'mp', 'quad', 'quad_id', 'plant_id', 'year')) %>%
  mutate(
    recruit_plot_valid = case_when(
      recruit == 1 & recruit_plot_valid == 1 ~ 1,
      TRUE ~ NA_real_))
```

## Process-specific data

The same biological definitions used in the year-by-site chapter are retained.
Flowering is binary, fruiting is conditional on flowering, and fruit number is
conditional on fruiting.

```{r mean_flx1_frx2_model_data}
df_su <- df %>%
  filter(
    size_t0 > 0,
    !is.na(survives),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(disturbance = as.numeric(disturbance))

df_gr <- df %>%
  filter(
    size_t0 > 0,
    survives == 1,
    is.finite(logsize_t0),
    is.finite(logsize_t1),
    !is.na(disturbance)) %>%
  mutate(disturbance = as.numeric(disturbance))

# Flowering is one binary process. Flower number is not modeled separately.
df_fl <- df %>%
  filter(
    !is.na(flower),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    disturbance = as.numeric(disturbance),
    flower = if_else(flower > 0, 1, 0))

# Fruiting probability is conditional on flowering.
df_fr <- df %>%
  filter(
    flower > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(
    disturbance = as.numeric(disturbance),
    fruiting = if_else(fruit > 0, 1, 0))

# Fruit number is conditional on fruiting.
df_fr_n <- df %>%
  filter(
    fruit > 0,
    is.finite(fruit),
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  mutate(disturbance = as.numeric(disturbance))


# Fruit-to-recruit data --------------------------------------------------------
# Total fruits in year t predict valid recruits in year t + 1. The mean model
# is forced through zero and has no year or site random effects.
df_fr2re_quad <- df %>%
  group_by(site, quad_id, year) %>%
  summarise(
    fruits_t0 = sum(fruit, na.rm = TRUE),
    recruits_t0 = sum(recruit_plot_valid == 1, na.rm = TRUE),
    .groups = 'drop') %>%
  arrange(site, quad_id, year) %>%
  group_by(site, quad_id) %>%
  mutate(
    recruits_t1 = lead(recruits_t0),
    year_t1 = lead(year),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_fr2re <- df_fr2re_quad %>%
  group_by(site, year, year_t1) %>%
  summarise(
    fruits_t0 = sum(fruits_t0, na.rm = TRUE),
    recruits_t1 = sum(recruits_t1, na.rm = TRUE),
    n_quads = n(),
    .groups = 'drop') %>%
  mutate(
    site = factor(site),
    logfruits_t0 = log1p(fruits_t0),
    logrecruits_t1 = log1p(recruits_t1))
```

## Remove selected years

```{r mean_flx1_frx2_remove_years}
v_years_og <- sort(unique(df$year))

df <- df %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_su <- df_su %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_gr <- df_gr %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_fl <- df_fl %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_fr <- df_fr %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_fr_n <- df_fr_n %>%
  filter(!is.na(year), !(year %in% v_years_re))

df_fr2re <- df_fr2re %>%
  filter(
    !is.na(year),
    !is.na(year_t1),
    !(year %in% v_years_re),
    !(year_t1 %in% v_years_re))

v_years <- sort(unique(df$year))
```

## Mean survival model

Candidate models compare intercept-only, linear, quadratic, and cubic size
relationships, each with and without the fixed disturbance effect. AICc selects
the best-supported candidate unless a model index is supplied above.

```{r mean_flx1_frx2_survival_model}
mod_su_00 <- glm(
  survives ~ 1,
  data = df_su, family = binomial)

mod_su_0 <- glm(
  survives ~ disturbance,
  data = df_su, family = binomial)

mod_su_10 <- glm(
  survives ~ logsize_t0,
  data = df_su, family = binomial)

mod_su_1 <- glm(
  survives ~ logsize_t0 + disturbance,
  data = df_su, family = binomial)

mod_su_20 <- glm(
  survives ~ logsize_t0 + logsize_t0_2,
  data = df_su, family = binomial)

mod_su_2 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_su, family = binomial)

mod_su_30 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_su, family = binomial)

mod_su_3 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_su, family = binomial)

mods_su <- list(mod_su_00, mod_su_0, mod_su_10, mod_su_1, 
                mod_su_20, mod_su_2, mod_su_30, mod_su_3)
mods_su_dAIC <- bbmle::AICctab(mods_su, weights = TRUE, sort = FALSE)$dAIC
mods_su_sorted <- order(mods_su_dAIC)

if (length(v_mod_set_su) == 0) {
  mod_su_index_best <- mods_su_sorted[1]
} else {
  mod_su_index_best <- v_mod_set_su + 1
}

v_mod_su_index <- floor((mod_su_index_best - 1) / 2)
mod_su_best <- mods_su[[mod_su_index_best]]
mod_su_coef <- coef(mod_su_best)
summary(mod_su_best)
```

## Mean growth model

Growth is modeled on the log-size scale with the same candidate structure used
for survival.

```{r mean_flx1_frx2_growth_model}
mod_gr_00 <- lm(logsize_t1 ~ 1,
               data = df_gr)

mod_gr_0 <- lm(logsize_t1 ~ disturbance,
               data = df_gr)

mod_gr_10 <- lm(logsize_t1 ~ logsize_t0,
               data = df_gr)

mod_gr_1 <- lm(logsize_t1 ~ logsize_t0 + disturbance,
               data = df_gr)

mod_gr_20 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2,
               data = df_gr)

mod_gr_2 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + disturbance,
               data = df_gr)

mod_gr_30 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
               data = df_gr)

mod_gr_3 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + 
                 logsize_t0_3 + disturbance,
               data = df_gr)

mods_gr <- list(mod_gr_00, mod_gr_0, mod_gr_10, mod_gr_1, 
                mod_gr_20, mod_gr_2, mod_gr_30, mod_gr_3)
mods_gr_dAICc <- bbmle::AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAICc
mods_gr_sorted <- order(mods_gr_dAICc)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_best <- mods_gr_sorted[1]
} else {
  mod_gr_index_best <- v_mod_set_gr +1
}

v_mod_gr_index <- floor((mod_gr_index_best - 1) / 2)
mod_gr_best  <- mods_gr[[mod_gr_index_best]]
mod_gr_coef <- coef(mod_gr_best)
summary(mod_gr_best)
```

### Growth variance

Squared growth residuals are modeled as an exponential function of the predicted
growth mean. The IPM uses the same mean-scale predictor when calculating the
growth standard deviation.

```{r mean_flx1_frx2_growth_variance}
# Variance is modeled against the predicted growth mean. grow_sd() below uses
# the same mean-scale predictor.
mod_gr_x <- fitted(mod_gr_best)
mod_gr_y <- resid(mod_gr_best)^2

mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x),
  start = list(a = 1, b = 0),
  control = nls.control(
    maxiter = 1000,
    tol = 1e-6,
    warnOnly = TRUE))
```

## Mean flowering-probability model

This is the only flower-related model entering the fertility kernel. Flower
number is not analyzed as a separate vital rate.

```{r mean_flx1_frx2_flowering_model}
mod_fl_00 <- glm(
  flower ~ 1,
  data = df_fl, family = binomial)

mod_fl_0 <- glm(
  flower ~ disturbance,
  data = df_fl, family = binomial)

mod_fl_10 <- glm(
  flower ~ logsize_t0,
  data = df_fl, family = binomial)

mod_fl_1 <- glm(
  flower ~ logsize_t0 + disturbance,
  data = df_fl, family = binomial)

mod_fl_20 <- glm(
  flower ~ logsize_t0 + logsize_t0_2,
  data = df_fl, family = binomial)

mod_fl_2 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_fl, family = binomial)

mod_fl_30 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fl, family = binomial)

mod_fl_3 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_fl, family = binomial)

mods_fl <- list(
  mod_fl_00, mod_fl_0, mod_fl_10, mod_fl_1,
  mod_fl_20, mod_fl_2, mod_fl_30, mod_fl_3)

mods_fl_dAICc <- bbmle::AICctab(
  mods_fl, weights = TRUE, sort = FALSE)$dAICc
mods_fl_sorted <- order(mods_fl_dAICc)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_best <- mods_fl_sorted[1]
} else {
  mod_fl_index_best <- v_mod_set_fl + 1
}

v_mod_fl_index <- floor((mod_fl_index_best - 1) / 2)
mod_fl_best <- mods_fl[[mod_fl_index_best]]
mod_fl_coef <- coef(mod_fl_best)

summary(mod_fl_best)
mods_fl_dAICc
```

## Mean fruiting-probability model

Fruiting probability is estimated only among plants that flowered.

```{r mean_flx1_frx2_fruiting_model}
mod_fr_00 <- glm(
  fruiting ~ 1,
  data = df_fr, family = binomial)

mod_fr_0 <- glm(
  fruiting ~ disturbance,
  data = df_fr, family = binomial)

mod_fr_10 <- glm(
  fruiting ~ logsize_t0,
  data = df_fr, family = binomial)

mod_fr_1 <- glm(
  fruiting ~ logsize_t0 + disturbance,
  data = df_fr, family = binomial)

mod_fr_20 <- glm(
  fruiting ~ logsize_t0 + logsize_t0_2,
  data = df_fr, family = binomial)

mod_fr_2 <- glm(
  fruiting ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_fr, family = binomial)

mod_fr_30 <- glm(
  fruiting ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fr, family = binomial)

mod_fr_3 <- glm(
  fruiting ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_fr, family = binomial)

mods_fr <- list(
  mod_fr_00, mod_fr_0, mod_fr_10, mod_fr_1,
  mod_fr_20, mod_fr_2, mod_fr_30, mod_fr_3)

mods_fr_dAICc <- bbmle::AICctab(
  mods_fr, weights = TRUE, sort = FALSE)$dAICc
mods_fr_sorted <- order(mods_fr_dAICc)

if (length(v_mod_set_fr) == 0) {
  mod_fr_index_best <- mods_fr_sorted[1]
} else {
  mod_fr_index_best <- v_mod_set_fr + 1
}

v_mod_fr_index <- floor((mod_fr_index_best - 1) / 2)
mod_fr_best <- mods_fr[[mod_fr_index_best]]
mod_fr_coef <- coef(mod_fr_best)

summary(mod_fr_best)
mods_fr_dAICc
```

## Mean fruit-number model

Fruit number is estimated only among plants that produced at least one fruit.
A negative-binomial model accommodates overdispersion in the count response.

```{r mean_flx1_frx2_fruit_number_model}
mod_fr_n_00 <- MASS::glm.nb(
  fruit ~ 1,
  data = df_fr_n)

mod_fr_n_0 <- MASS::glm.nb(
  fruit ~ disturbance,
  data = df_fr_n)

mod_fr_n_10 <- MASS::glm.nb(
  fruit ~ logsize_t0,
  data = df_fr_n)

mod_fr_n_1 <- MASS::glm.nb(
  fruit ~ logsize_t0 + disturbance,
  data = df_fr_n)

mod_fr_n_20 <- MASS::glm.nb(
  fruit ~ logsize_t0 + logsize_t0_2,
  data = df_fr_n)

mod_fr_n_2 <- MASS::glm.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_fr_n)

mod_fr_n_30 <- MASS::glm.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fr_n)

mod_fr_n_3 <- MASS::glm.nb(
  fruit ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_fr_n)

mods_fr_n <- list(
  mod_fr_n_00, mod_fr_n_0, mod_fr_n_10, mod_fr_n_1,
  mod_fr_n_20, mod_fr_n_2, mod_fr_n_30, mod_fr_n_3)

mods_fr_n_dAICc <- bbmle::AICctab(
  mods_fr_n, weights = TRUE, sort = FALSE)$dAICc
mods_fr_n_sorted <- order(mods_fr_n_dAICc)

if (length(v_mod_set_fr_n) == 0) {
  mod_fr_n_index_best <- mods_fr_n_sorted[1]
} else {
  mod_fr_n_index_best <- v_mod_set_fr_n + 1
}

v_mod_fr_n_index <- floor((mod_fr_n_index_best - 1) / 2)
mod_fr_n_best <- mods_fr_n[[mod_fr_n_index_best]]
mod_fr_n_coef <- coef(mod_fr_n_best)

summary(mod_fr_n_best)
mods_fr_n_dAICc
```

## Mean fruit-to-recruit model

Total fruit production at a site in year $t$ predicts the number of valid
recruits at that site in year $t+1$. The relationship is constrained to pass
through the origin on the `log1p` scale:

\[
\log(1 + R_{i,t+1}) =
\beta_1\log(1 + F_{i,t}).
\]

Thus, zero fruits predict zero recruits. The slope is shared across all sites
and years.

```{r mean_flx1_frx2_fr2re_model}
# Zero fruits predict zero recruits. No site or year random effects are fitted.
mod_fr2re <- lm(
  logrecruits_t1 ~ 0 + logfruits_t0,
  data = df_fr2re)

summary(mod_fr2re)
```

## Mean vital-rate plots

The fitted lines show the mean fixed-effect relationships under no fire and
fire. The figures summarize the relationships used by the mean IPM rather than
year- or site-specific deviations.

```{r mean_flx1_frx2_vital_rate_plots, fig.width=10, fig.height=12}
make_mean_prediction <- function(data, model) {
  x <- seq(
    min(data$logsize_t0, na.rm = TRUE),
    max(data$logsize_t0, na.rm = TRUE),
    length.out = 200)

  pred <- expand_grid(
    logsize_t0 = x,
    disturbance = c(0, 1)) %>%
    mutate(
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3,
      disturbance_plot = factor(
        disturbance,
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  pred$prediction <- predict(
    model,
    newdata = pred,
    type = 'response')

  pred
}

make_mean_plot <- function(
    data, model, response, title, y_lab,
    probability = FALSE) {
  pred <- make_mean_prediction(data, model)

  data_plot <- data %>%
    mutate(
      disturbance_plot = factor(
        disturbance,
        levels = c(0, 1),
        labels = c('No fire', 'Fire')))

  point_layer <- if (probability) {
    geom_jitter(
      data = data_plot,
      aes(
        x = logsize_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7,
      width = 0.05,
      height = 0.05)
  } else {
    geom_point(
      data = data_plot,
      aes(
        x = logsize_t0,
        y = .data[[response]],
        color = disturbance_plot),
      alpha = 0.15,
      size = 0.7)
  }

  p <- ggplot() +
    point_layer +
    geom_line(
      data = pred,
      aes(
        x = logsize_t0,
        y = prediction,
        color = disturbance_plot),
      linewidth = 0.8) +
    scale_color_manual(
      values = c('No fire' = 'black', 'Fire' = 'red')) +
    theme_bw() +
    labs(
      title = title,
      subtitle = v_ggp_suffix,
      x = expression('log(size)'[t0]),
      y = y_lab,
      color = NULL)

  if (probability) {
    p <- p + coord_cartesian(ylim = c(0, 1))
  }

  p
}

fig_su <- make_mean_plot(
  data = df_su,
  model = mod_su_best,
  response = 'survives',
  title = 'Mean survival',
  y_lab = 'Survival probability',
  probability = TRUE)

fig_gr <- make_mean_plot(
  data = df_gr,
  model = mod_gr_best,
  response = 'logsize_t1',
  title = 'Mean growth',
  y_lab = expression('log(size)'[t1]))

fig_fl <- make_mean_plot(
  data = df_fl,
  model = mod_fl_best,
  response = 'flower',
  title = 'Mean flowering probability',
  y_lab = 'Flowering probability',
  probability = TRUE)

fig_fr <- make_mean_plot(
  data = df_fr,
  model = mod_fr_best,
  response = 'fruiting',
  title = 'Mean fruiting probability given flowering',
  y_lab = 'Fruiting probability',
  probability = TRUE)

fig_fr_n <- make_mean_plot(
  data = df_fr_n,
  model = mod_fr_n_best,
  response = 'fruit',
  title = 'Mean fruit number given fruiting',
  y_lab = 'Number of fruits')

fig_vital_rates <-
  fig_su + fig_gr + fig_fl +
  fig_fr + fig_fr_n +
  plot_layout(ncol = 2, guides = 'collect') &
  theme(legend.position = 'bottom')

fig_vital_rates
```

## Mean fruit-to-recruit relationship

```{r mean_flx1_frx2_fr2re_plot, fig.width=7, fig.height=5}
df_fr2re_pred <- tibble(
  logfruits_t0 = seq(
    0,
    max(df_fr2re$logfruits_t0, na.rm = TRUE),
    length.out = 200)) %>%
  mutate(
    logrecruits_t1 = predict(
      mod_fr2re,
      newdata = .),
    fruits_t0 = expm1(logfruits_t0),
    recruits_t1 = pmax(0, expm1(logrecruits_t1)))

fig_fr2re <- ggplot(
  df_fr2re,
  aes(x = fruits_t0, y = recruits_t1)) +
  geom_point(alpha = 0.4, size = 1.3) +
  geom_line(
    data = df_fr2re_pred,
    aes(x = fruits_t0, y = recruits_t1),
    linewidth = 0.8) +
  scale_x_continuous(
  trans = scales::pseudo_log_trans(sigma = 1)) +
  theme_bw() +
  labs(
    title = 'Mean fruit-to-recruit relationship',
    subtitle = v_ggp_suffix,
    x = expression('Total fruits'[t]),
    y = expression('Valid recruits'[t+1]))

fig_fr2re
```

## Recruit size distribution

Recruit size is represented by a normal distribution on the log-size scale. Only
recruits whose quadrats were represented in the previous year contribute to the
estimated mean and standard deviation.

```{r mean_flx1_frx2_recruit_size}
df_recr <- df %>%
  filter(
    recruit_plot_valid == 1,
    size_t0 > 0,
    is.finite(logsize_t0))

rc_sz <- df_recr %>%
  summarise(
    mean = mean(logsize_t0, na.rm = TRUE),
    sd = sd(logsize_t0, na.rm = TRUE))
```

## Mean parameter extraction

Missing polynomial or disturbance terms are filled with zero, allowing the same
IPM functions to work with any AICc-selected candidate.

```{r mean_flx1_frx2_parameters}
size_term_map <- c(
  b0 = '(Intercept)',
  b1 = 'logsize_t0',
  b2 = 'logsize_t0_2',
  b3 = 'logsize_t0_3',
  bd = 'disturbance')

extract_terms <- function(model, prefix, term_map = size_term_map) {
  coefs <- stats::coef(model)

  values <- vapply(term_map, function(term_i) {
    if (term_i %in% names(coefs)) {
      unname(coefs[[term_i]])
    } else {
      0
    }
  }, numeric(1))

  names(values) <- paste0(prefix, names(term_map))
  values
}

grow_var_coef <- stats::coef(mod_gr_var)

mesh_limits <- range(
  c(
    df_gr$logsize_t0,
    df_gr$logsize_t1,
    df_recr$logsize_t0),
  na.rm = TRUE,
  finite = TRUE)

pars_mean <- as.list(c(
  extract_terms(mod_su_best, 'surv_'),
  extract_terms(mod_gr_best, 'grow_'),
  extract_terms(mod_fl_best, 'fl_'),
  extract_terms(mod_fr_best, 'fr_'),
  extract_terms(mod_fr_n_best, 'frn_'),
  fr2re_b0 = 0,
  fr2re_b1 = unname(
    stats::coef(mod_fr2re)[['logfruits_t0']]),
  recr_mean = rc_sz$mean,
  recr_sd = rc_sz$sd,
  grow_var_a = unname(grow_var_coef[['a']]),
  grow_var_b = unname(grow_var_coef[['b']]),
  fruits_ref = mean(df_fr2re$fruits_t0, na.rm = TRUE),
  L = mesh_limits[1],
  U = mesh_limits[2],
  mat_siz = 200))

pars_mean_table <- tibble(
  parameter = names(pars_mean),
  value = as.numeric(unlist(pars_mean)))

pars_mean_table
```

## IPM functions

Survival and flowering probabilities use inverse-logit links. Expected fruits
per individual combine flowering probability, fruiting probability, and
conditional fruit number. The raw number of flowers is not multiplied into this
expression because the fruit-number model already predicts total fruits for a
fruiting individual.

### Fruit-to-recruit back-transformation

The fruit-to-recruit model is returned to the original count scale using the
direct inverse of the `log1p` transformation:

\[
\widehat{R}_{i,t+1} =
\exp\left(\widehat{\eta}_{i,t+1}\right)-1.
\]

A residual-variance correction of $0.5\sigma^2$ is not added inside the
exponential. With a model constrained through zero, that correction creates a
positive predicted number of recruits as fruit production approaches zero.
Dividing that positive baseline by very small fruit totals can produce
unrealistically large estimates of recruits per fruit and inflated fertility
values. The direct inverse transformation preserves

\[
F_{i,t}=0 \quad\Longrightarrow\quad \widehat{R}_{i,t+1}=0,
\]

and predicted recruitment approaches zero continuously as fruit production
approaches zero.

```{r mean_flx1_frx2_ipm_functions}
get_par <- function(pars, par_name, default = 0) {
  if (!is.null(pars[[par_name]])) {
    pars[[par_name]]
  } else {
    default
  }
}

vital_lp <- function(x, pars, prefix, disturbance = 0) {
  get_par(pars, paste0(prefix, 'b0')) +
    get_par(pars, paste0(prefix, 'b1')) * x +
    get_par(pars, paste0(prefix, 'b2')) * x^2 +
    get_par(pars, paste0(prefix, 'b3')) * x^3 +
    get_par(pars, paste0(prefix, 'bd')) * disturbance
}


# Survival ---------------------------------------------------------------------
sx <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'surv_',
    disturbance = disturbance))
}


# Growth -----------------------------------------------------------------------
grow_mu <- function(x, pars, disturbance = 0) {
  vital_lp(
    x = x,
    pars = pars,
    prefix = 'grow_',
    disturbance = disturbance)
}

grow_sd <- function(mean_value, pars) {
  variance <- get_par(pars, 'grow_var_a') *
    exp(get_par(pars, 'grow_var_b') * mean_value)

  sqrt(pmax(variance, .Machine$double.eps))
}

gxy <- function(x, y, pars, disturbance = 0) {
  mean_value <- grow_mu(
    x = x,
    pars = pars,
    disturbance = disturbance)

  sd_value <- grow_sd(
    mean_value = mean_value,
    pars = pars)

  dnorm(y, mean = mean_value, sd = sd_value)
}

pxy <- function(x, y, pars, disturbance = 0) {
  sx(x, pars, disturbance) *
    gxy(x, y, pars, disturbance)
}


# Flowering probability --------------------------------------------------------
fl_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fl_',
    disturbance = disturbance))
}


# Fruiting probability conditional on flowering -------------------------------
fr_x <- function(x, pars, disturbance = 0) {
  plogis(vital_lp(
    x = x,
    pars = pars,
    prefix = 'fr_',
    disturbance = disturbance))
}


# Fruit number conditional on fruiting ----------------------------------------
frn_x <- function(x, pars, disturbance = 0) {
  exp(vital_lp(
    x = x,
    pars = pars,
    prefix = 'frn_',
    disturbance = disturbance))
}


# Expected fruits per individual ----------------------------------------------
# Flower count is not multiplied into this expression. The fruit-number model
# predicts total fruits for fruiting individuals.
fruits_x <- function(x, pars, disturbance = 0) {
  fl_x(x, pars, disturbance) *
    fr_x(x, pars, disturbance) *
    frn_x(x, pars, disturbance)
}


# Fruit-to-recruit conversion --------------------------------------------------
predict_recruits_from_fruits <- function(fruits, pars) {
  fruits_nonnegative <- pmax(fruits, 0)

  eta <- get_par(pars, 'fr2re_b0') +
    get_par(pars, 'fr2re_b1') * log1p(fruits_nonnegative)

  # Directly invert the log1p prediction. A 0.5 * sigma^2 correction is not
  # added because it creates positive recruitment as fruits approach zero,
  # resulting in unrealistically large estimates of recruits per fruit.
  recruits <- pmax(expm1(eta), 0)

  ifelse(fruits_nonnegative <= 0, 0, recruits)
}

recruits_per_fruit <- function(pars) {
  fruits_ref <- get_par(pars, 'fruits_ref', 0)

  if (!is.finite(fruits_ref) || fruits_ref <= 0) {
    return(0)
  }

  predict_recruits_from_fruits(fruits_ref, pars) / fruits_ref
}

rx <- function(x, pars, disturbance = 0) {
  fruits_x(
    x = x,
    pars = pars,
    disturbance = disturbance) *
    recruits_per_fruit(pars)
}


# Recruit size distribution ----------------------------------------------------
recr_y <- function(y, pars) {
  dnorm(
    y,
    mean = get_par(pars, 'recr_mean'),
    sd = get_par(pars, 'recr_sd'))
}


# Fertility matrix -------------------------------------------------------------
fy <- function(y, x, pars, h, disturbance = 0) {
  Pvec <- recr_y(y, pars)
  recruit_mass <- sum(Pvec * h)

  if (!is.finite(recruit_mass) || recruit_mass <= 0) {
    stop('Recruit size distribution has no mass inside the mesh')
  }

  Pvec <- Pvec / recruit_mass

  Rvec <- rx(
    x = x,
    pars = pars,
    disturbance = disturbance)

  outer(Pvec, Rvec) * h
}


# Kernel -----------------------------------------------------------------------
kernel <- function(pars, disturbance = 0) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  Fmat <- fy(
    y = y,
    x = y,
    pars = pars,
    h = h,
    disturbance = disturbance)

  Smat <- sx(
    x = y,
    pars = pars,
    disturbance = disturbance)

  Gmat <- matrix(0, n, n)
  Gmat[] <- t(
    outer(
      y,
      y,
      Vectorize(function(x, y) {
        gxy(
          x = x,
          y = y,
          pars = pars,
          disturbance = disturbance)
      }))) * h

  Tmat <- matrix(0, n, n)

  for (i in 1:(n / 2)) {
    Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  for (i in (n / 2 + 1):n) {
    Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    Tmat[, i] <- Gmat[, i] * Smat[i]
  }

  k_yx <- Fmat + Tmat

  list(
    k_yx = k_yx,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Smat = Smat,
    meshpts = y,
    h = h)
}

lambda_ipm <- function(pars, disturbance = 0) {
  K <- kernel(
    pars = pars,
    disturbance = disturbance)$k_yx

  Re(eigen(K, only.values = TRUE)$values[1])
}
```

## Mean observed size distribution

The initial size distribution is the arithmetic mean of the annual
size-frequency vectors. Each year therefore contributes equally, rather than
plant-rich years dominating the reference distribution.

```{r mean_flx1_frx2_mean_size_distribution}
# The vector is the arithmetic mean of annual size-frequency vectors.
make_initial_n_mean <- function(pars, df_init = df) {
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  breaks <- seq(L, U, length.out = n + 1)

  years_i <- df_init %>%
    filter(size_t0 > 0, is.finite(logsize_t0)) %>%
    distinct(year) %>%
    arrange(year) %>%
    pull(year)

  counts_by_year <- purrr::map(years_i, function(year_i) {
    x <- df_init %>%
      filter(
        year == year_i,
        size_t0 > 0,
        is.finite(logsize_t0)) %>%
      pull(logsize_t0)

    hist(
      pmin(pmax(x, L), U),
      breaks = breaks,
      plot = FALSE,
      include.lowest = TRUE)$counts
  })

  mean_counts <- Reduce(`+`, counts_by_year) / length(counts_by_year)
  mean_counts / h
}
```

## Expected mean fruit production

The mean size distribution is combined with the fitted reproductive functions
to calculate total expected fruit production under no fire and fire. These
population-level fruit totals provide the reference at which the nonlinear
fruit-to-recruit relationship is converted to recruits per fruit.

```{r mean_flx1_frx2_expected_fruits}
expected_mean_fruits <- function(
    pars, disturbance = 0, df_init = df) {
  n_obs <- make_initial_n_mean(pars, df_init)
  n <- as.integer(pars$mat_siz)
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  sum(
    fruits_x(y, pars, disturbance) * n_obs * h,
    na.rm = TRUE)
}

mean_fruits_no_fire <- expected_mean_fruits(
  pars = pars_mean,
  disturbance = 0)

mean_fruits_fire <- expected_mean_fruits(
  pars = pars_mean,
  disturbance = 1)
```

## Mean disturbance exposure

Disturbance exposure is averaged across unique quadrat-year combinations. This
gives each sampled quadrat-year equal weight and prevents quadrats containing
many plants from contributing disproportionately to the mean disturbance
regime.

```{r mean_flx1_frx2_disturbance_exposure}
# Each sampled quadrat-year receives equal weight; plant-rich quadrats do not
# receive greater weight merely because they contain more plant records.
p_disturbance_mean <- df %>%
  filter(
    !is.na(year),
    !is.na(quad_id),
    !is.na(disturbance)) %>%
  distinct(year, quad_id, disturbance) %>%
  summarise(
    p_disturbance = mean(disturbance, na.rm = TRUE)) %>%
  pull(p_disturbance)

mean_fruits_regime <-
  (1 - p_disturbance_mean) * mean_fruits_no_fire +
  p_disturbance_mean * mean_fruits_fire
```

## State-specific and disturbance-regime kernels

Three mean scenarios are constructed:

1. a no-fire kernel;
2. a fire kernel;
3. an expected disturbance-regime kernel.

The expected-regime kernel is a weighted mixture of the no-fire and fire
kernels, using the observed mean disturbance exposure. Both components of this
mixture use the same expected total fruit production under the mean disturbance
regime so that the fruit-to-recruit conversion is consistent across the two
components.

```{r mean_flx1_frx2_kernels}
pars_no_fire <- pars_mean
pars_no_fire$fruits_ref <- mean_fruits_no_fire

pars_fire <- pars_mean
pars_fire$fruits_ref <- mean_fruits_fire

# For the mixture kernel, both component kernels use the same expected total
# fruit production under the observed mean disturbance regime.
pars_regime <- pars_mean
pars_regime$fruits_ref <- mean_fruits_regime

K_no_fire <- kernel(
  pars = pars_no_fire,
  disturbance = 0)$k_yx

K_fire <- kernel(
  pars = pars_fire,
  disturbance = 1)$k_yx

K_regime_no_fire <- kernel(
  pars = pars_regime,
  disturbance = 0)$k_yx

K_regime_fire <- kernel(
  pars = pars_regime,
  disturbance = 1)$k_yx

K_regime <-
  (1 - p_disturbance_mean) * K_regime_no_fire +
  p_disturbance_mean * K_regime_fire
```

## Asymptotic and projected population growth

Asymptotic population growth is the dominant eigenvalue of each kernel. It
represents the long-term growth rate if the corresponding mean vital rates and
disturbance state remained constant and the population reached its stable size
distribution.

Projected population growth is obtained by applying the kernel once to the mean
observed size distribution. It therefore retains the transient influence of the
observed population structure.

```{r mean_flx1_frx2_lambda}
project_kernel <- function(K, pars, label, df_init = df) {
  n_obs <- make_initial_n_mean(pars, df_init)
  h <- (pars$U - pars$L) / pars$mat_siz
  n_initial <- sum(n_obs) * h

  n_proj <- K %*% n_obs
  n_projected <- sum(n_proj) * h

  tibble(
    lambda_type = label,
    asym_lambda = Re(eigen(K, only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial,
    n_initial = n_initial,
    n_projected = n_projected)
}

df_lambda_mean <- bind_rows(
  project_kernel(
    K = K_no_fire,
    pars = pars_no_fire,
    label = 'No fire'),
  project_kernel(
    K = K_fire,
    pars = pars_fire,
    label = 'Fire'),
  project_kernel(
    K = K_regime,
    pars = pars_regime,
    label = 'Expected disturbance regime')) %>%
  mutate(
    p_disturbance = c(0, 1, p_disturbance_mean),
    fruits_ref = c(
      mean_fruits_no_fire,
      mean_fruits_fire,
      mean_fruits_regime),
    recruits_ref = purrr::map2_dbl(
      fruits_ref,
      list(pars_no_fire, pars_fire, pars_regime),
      predict_recruits_from_fruits),
    recruits_per_fruit = if_else(
      fruits_ref > 0,
      recruits_ref / fruits_ref,
      0))

df_lambda_mean
```

## Observed annual population growth

Observed annual growth is calculated from quadrats with plant records in
consecutive years. Because true sampled zero counts cannot always be identified
from the individual-level table, transitions involving a sampled quadrat with no
plants may be absent. The observed series should therefore be interpreted with
this sampling limitation in mind.

```{r mean_flx1_frx2_observed_growth}
# Counts are restricted to quadrats represented in consecutive years. Because
# quadrats containing no plant records cannot be identified from the individual
# table, transitions to a true zero count may remain unavailable.
df_counts_quad_year <- df %>%
  filter(size_t0 > 0, is.finite(logsize_t0)) %>%
  group_by(quad_id, year) %>%
  summarise(
    n_t0 = n_distinct(plant_id),
    .groups = 'drop') %>%
  arrange(quad_id, year) %>%
  group_by(quad_id) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n_t0),
    year_gap = year_t1 - year) %>%
  ungroup() %>%
  filter(year_gap == 1)

df_obs_growth_year <- df_counts_quad_year %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(n_t0, na.rm = TRUE),
    n_t1 = sum(n_t1, na.rm = TRUE),
    n_quads = n(),
    obs_lambda = n_t1 / n_t0,
    .groups = 'drop') %>%
  filter(n_t0 > 0, n_t1 > 0)

df_obs_growth_year


df_obs_growth_summary <- df_obs_growth_year %>%
  summarise(
    n_transitions = n(),
    lambda_arithmetic = mean(obs_lambda, na.rm = TRUE),
    lambda_geometric = exp(mean(log(obs_lambda), na.rm = TRUE)),
    lambda_median = median(obs_lambda, na.rm = TRUE),
    lambda_min = min(obs_lambda, na.rm = TRUE),
    lambda_max = max(obs_lambda, na.rm = TRUE))

df_obs_growth_summary
```

## Observed and modeled population growth through time

The connected points show observed annual population growth. The horizontal
lines show the asymptotic and one-step projected growth rates from the three mean
IPM scenarios. A value of $\lambda=1$ indicates a stationary population;
values above one indicate growth and values below one indicate decline.

The modeled lines are constant through time because this chapter contains no
year-specific effects. They summarize the growth rates expected under average
vital-rate relationships, whereas the observed series includes year-to-year
environmental variation, demographic stochasticity, observation error, and
changes in population structure. Agreement is therefore assessed by whether the
mean-IPM estimates are of a similar magnitude to the center of the observed
annual variation, not by expecting the model to reproduce every annual value.

The projected estimates are based on the mean observed size distribution and are
the more direct short-term comparison with observed annual growth. The
asymptotic estimates instead describe growth after convergence to the stable size
distribution under each fixed scenario.

```{r mean_flx1_frx2_growth_plot, fig.width=10, fig.height=6}
df_model_lines <- df_lambda_mean %>%
  select(lambda_type, asym_lambda, proj_lambda) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'metric',
    values_to = 'lambda') %>%
  mutate(
    metric = recode(
      metric,
      asym_lambda = 'Asymptotic',
      proj_lambda = 'Projected'),
    line_type = paste(lambda_type, metric, sep = ' - '))

fig_lambda_mean <- ggplot(
  df_obs_growth_year,
  aes(x = year, y = obs_lambda)) +
  geom_hline(
    yintercept = 1,
    linetype = 'dotted') +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(
    data = df_model_lines,
    aes(
      yintercept = lambda,
      linetype = line_type),
    linewidth = 0.6) +
  theme_bw() +
  labs(
    title = 'Observed annual growth and mean-IPM lambda',
    subtitle = v_ggp_suffix,
    x = 'Year t0',
    y = expression(lambda),
    linetype = NULL)

fig_lambda_mean
```

## Kernel diagnostics

The kernel diagnostic table checks for nonfinite entries and reports the range of
matrix values for each disturbance scenario.

```{r mean_flx1_frx2_kernel_diagnostics}
df_kernel_diagnostics <- tibble(
  kernel = c('No fire', 'Fire', 'Expected regime'),
  min_entry = c(
    min(K_no_fire),
    min(K_fire),
    min(K_regime)),
  max_entry = c(
    max(K_no_fire),
    max(K_fire),
    max(K_regime)),
  any_nonfinite = c(
    any(!is.finite(K_no_fire)),
    any(!is.finite(K_fire)),
    any(!is.finite(K_regime))))

df_kernel_diagnostics
```

