# Data {#data}

## From publication
### Useful demographic estimates for *Chrysopsis highlandsensis*

This section summarizes useful demographic estimates for *Chrysopsis highlandsensis* from David et al. (2025). The estimates are grouped by vital-rate category so they can be used to guide IPM structure, parameter choices, and interpretation of model results.

### Survival

#### Seedling and recruit survival

Recruitment occurred throughout the year, but most new seedlings appeared during the December--March census interval. This interval produced the largest number of recruits but also had the lowest short-term and one-year survival.

| Census interval | New recruits | Percent of annual total | Survival to next quarter | One-year survival |
| --------------- | -----------: | ----------------------: | -----------------------: | ----------------: |
| March           |        2,794 |                   53.4% |                    49.8% |             29.2% |
| June            |          669 |                   12.8% |                    81.2% |             61.1% |
| September       |          527 |                   10.1% |                    82.2% |             54.5% |
| December        |        1,240 |                   23.7% |                    83.4% |             46.5% |
| Total / mean    |        5,230 |                    100% |                    64.4% |             40.0% |

The most important implication is that recruitment timing matters. Winter recruitment is numerically dominant, but survival of these recruits is comparatively low.

#### Yearling survival

Yearling survival increased strongly with rosette diameter. Habitat also mattered: yearling survival was higher in scrub than roadside habitat.

| Stage    | Roadside survival | Scrub survival | Notes                        |
| -------- | ----------------: | -------------: | ---------------------------- |
| Yearling |    68.4 ± 7.6% SE | 74.8 ± 6.4% SE | Survival increased with size |

The size effect was strong, and the habitat effect was significant. There was no significant habitat-by-size interaction.

#### Vegetative survival

Vegetative survival also increased strongly with size. Scrub plants again had higher survival than roadside plants.

| Stage      | Roadside survival | Scrub survival | Notes                        |
| ---------- | ----------------: | -------------: | ---------------------------- |
| Vegetative |    70.4 ± 6.7% SE | 81.7 ± 4.8% SE | Survival increased with size |

This supports using size-dependent survival in the IPM.

#### Reproductive survival

Reproductive plants had low survival, consistent with a largely monocarpic life cycle.

| Reproductive fate                               | Estimate |
| ----------------------------------------------- | -------: |
| Reproductive individuals surviving to next year |    19.2% |
| Mortality after reproduction                    |    80.8% |
| Regressed to vegetative stage                   |     6.7% |
| Remained reproductive next year                 |    12.5% |

Among individuals that ever reached reproduction, most died in the same year they flowered. However, reproduction was not strictly fatal: some individuals survived and either regressed to vegetative status or remained reproductive in the following year.

This is important for the IPM. Flowering should impose a strong survival cost, but survival after flowering should not be forced to zero.

#### Fire-related survival

Fire reduced survival, especially when plants were fully consumed.

| Burn status | Survival |
| ----------- | -------: |
| Unburned    |    84.5% |
| Scorched    |    52.2% |
| Consumed    |    22.0% |

Across non-yearling plants, 38.7% of individuals that experienced fire damage survived to the next quarterly census. Seedlings younger than one year did not survive fire damage, although many seedlings within burned sites remained unburned.

The modeling implication is that fire should affect survival, but fire presence alone is coarse. Burn severity is biologically more informative than a simple fire/no-fire classification.

### Growth

#### Yearling size and growth

Yearlings were small but variable in size.

| Quantity                |      Estimate |
| ----------------------- | ------------: |
| Mean rosette diameter   | 2.6 ± 0.04 cm |
| Median rosette diameter |        1.9 cm |
| Range                   |  0.2--18.3 cm |

Relative growth rate from yearling to vegetative stage was:

| Transition             | Relative growth rate |
| ---------------------- | -------------------: |
| Yearling to vegetative |        0.661 ± 0.084 |

There was no significant habitat effect on yearling-to-vegetative relative growth rate.

#### Vegetative size and growth

Vegetative plants were larger and grew more slowly than yearlings.

| Quantity                         |     Estimate |
| -------------------------------- | -----------: |
| Mean vegetative rosette diameter | 8.2 ± 0.1 cm |

Vegetative-to-vegetative relative growth differed by habitat.

| Transition               |  Roadside RGR |      Scrub RGR |
| ------------------------ | ------------: | -------------: |
| Vegetative to vegetative | 0.210 ± 0.037 | 0.154 ± 0.0346 |

Roadside vegetative plants grew faster than scrub vegetative plants. Among vegetative individuals that remained vegetative, 76.3% grew or stayed the same size, whereas 23.7% shrank.

This supports modeling growth as a size-dependent continuous process, with potential habitat or disturbance effects.

#### Size threshold for flowering

Vegetative diameter strongly predicted flowering probability in the following year. Roadside plants reached a 50% probability of flowering at a slightly smaller size than scrub plants.

| Habitat  | Diameter for 50% flowering probability |
| -------- | -------------------------------------: |
| Roadside |                                16.6 cm |
| Scrub    |                                17.8 cm |

This supports a strongly size-dependent flowering model.

### Reproductive

#### Probability and timing of reproduction

Only a minority of individuals reproduced.

| Quantity                                                  |    Estimate |
| --------------------------------------------------------- | ----------: |
| Plants with full life cycle that reproduced at least once |         17% |
| Median age at first reproduction                          |     3 years |
| Interquartile range                                       |  2--4 years |
| Range                                                     | 1--11 years |
| First reproduction by age 4                               |         85% |
| First reproduction at age 7 or older                      |        2.8% |

There were 84 cases of one-year-old plants reaching reproduction, representing 15.7% of first reproductive events.

#### Monocarpy and repeated reproduction

The species is mostly monocarpic, but not strictly monocarpic.

| Number of reproductive years | Number of plants | Percent |
| ---------------------------- | ---------------: | ------: |
| Once                         |               -- |   87.0% |
| Twice                        |              109 |   10.8% |
| Three times                  |               19 |    1.9% |
| Four times                   |                2 |    0.2% |

This supports including flowering status in the survival model. A flowering plant should have much lower survival than a non-flowering plant, but survival after reproduction should remain possible.

#### Flower-head production

Flower-head number varied strongly among reproductive individuals.

| Quantity                                 | Estimate |
| ---------------------------------------- | -------: |
| Mean flower heads per reproductive plant |     63.4 |
| Standard deviation                       |     71.7 |
| Range                                    |   1--525 |

Flower-head number did not differ significantly between scrub and roadside habitat.

#### Branch number and height

| Trait                         |      Estimate |
| ----------------------------- | ------------: |
| Mean branch number            |           3.0 |
| Branch number SD              |           2.9 |
| Branch number range           |         1--33 |
| Mean height of tallest branch |       53.9 cm |
| Height SD                     |       21.8 cm |
| Height range                  | 3.8--182.5 cm |

Branch number did not differ by habitat. Tallest branch height was greater in scrub than roadside habitat.

#### Seed production

Seed production per head was estimated as follows:

| Quantity                    | Estimate |
| --------------------------- | -------: |
| Mean total seeds per head   |     78.5 |
| SD total seeds per head     |     24.0 |
| Range total seeds per head  |   3--291 |
| Mean intact seeds per head  |     50.9 |
| SD intact seeds per head    |     30.3 |
| Range intact seeds per head |   0--182 |
| Percent seeds intact        |    64.8% |

Using flower-head and seed-production estimates, the authors estimated average seed production per reproductive plant as:

| Quantity                            |                     Estimate |
| ----------------------------------- | ---------------------------: |
| Total seeds per reproductive plant  |                        4,906 |
| Intact seeds per reproductive plant |                        3,181 |
| Minimum estimated seed production   |            3 seeds, 0 intact |
| Maximum estimated seed production   | 152,775 seeds, 95,813 intact |

For the IPM, the most useful value is **50.9 intact seeds per head**, because the model already predicts flower-head number separately. Using 3,181 intact seeds per reproductive plant would double-count flower-head production.

#### Seed viability

Seed viability was low, even among visually intact seeds.

| Seed class   | Laboratory germination |
| ------------ | ---------------------: |
| Intact       |         19.8 ± 0.1% SE |
| Intermediate |          2.7 ± 0.1% SE |
| Empty        |                     0% |

Because only 64.8% of seeds were intact, the estimated viable fraction of all seeds produced was approximately 12.8%.

This is important when converting raw seed production into effective seed-bank entry or recruitment.

### Disturbances

#### Fire and recruitment

Large recruitment events were not clearly associated with fire or mechanical disturbance. Across 15 disturbance events, including 12 fires and 3 mechanical treatments, seedling recruitment was not significantly related to time since disturbance.

Recruitment was higher in roadside habitat than scrub habitat.

This means that *C. highlandsensis* does not show the simple post-fire recruitment boom expected for many Florida scrub reseeders. Fire may maintain open habitat, but it does not automatically generate large recruitment pulses.

#### Fire and survival

Fire reduced survival depending on burn severity.

| Burn severity | Survival |
| ------------- | -------: |
| Unburned      |    84.5% |
| Scorched      |    52.2% |
| Consumed      |    22.0% |

Individual burn events varied in survival response.

| Burn year / property | Unburned | Scorched | Consumed |
| -------------------- | -------: | -------: | -------: |
| 2004 ABS             |    83.3% |    28.9% |    12.5% |
| 2008 LWRSF           |    96.6% |    66.7% |       -- |
| 2015 ABS             |     100% |     100% |    42.9% |
| 2017 HHSP            |    73.9% |     100% |    22.2% |
| 2020 ABS             |    91.7% |    66.7% |    22.2% |

The main implication is that fire effects are not uniform. Survival depends strongly on whether plants are unburned, scorched, or consumed.

#### Fire and future reproduction

Fire also reduced the probability that surviving non-seedling plants eventually reproduced.

| Burn status | Eventually reproduced |
| ----------- | --------------------: |
| Unburned    |                 40.8% |
| Scorched    |                 28.6% |
| Consumed    |                 22.6% |

This suggests that fire can have delayed demographic effects beyond immediate mortality.

#### Mechanical disturbance

Mechanical disturbance had mixed evidence. The HHSP Site 10 population increased beginning in 2019 after roller-chopping, but the broader analysis did not find a general recruitment response to disturbance timing.

This suggests that mechanical treatments may sometimes improve local conditions, but their effects are not consistently predictable from the available data.

#### Microdisturbance, litter, and shade

Seed ecology experiments showed that germination was sensitive to microhabitat conditions.

In one experiment, soil disturbance decreased germination from 6.8% in undisturbed soil to 2.4% in disturbed soil. Shade increased germination, with partial shade and full shade producing much higher germination than full sun. Litter removal increased germination in the HHSP seed-addition experiment.

This suggests that open but not fully exposed microsites may be most favorable for germination.

### Else

#### Monitoring scope

The demographic dataset is large and long-term.

| Quantity                                           |   Estimate |
| -------------------------------------------------- | ---------: |
| Total plants followed                              |      6,456 |
| Seedling recruits tracked                          |      5,230 |
| Full life cycles tracked from germination to death |      4,759 |
| Study period                                       | 1999--2022 |
| Number of demographic sites                        |          9 |

This is a strong dataset for estimating vital rates and building an IPM.

#### Lifespan

| Group                             | Mean lifespan |  SD | Range / IQR                  |
| --------------------------------- | ------------: | --: | ---------------------------- |
| All individuals                   |     1.3 years | 1.7 | 0.25--12 years               |
| Individuals reaching reproduction |     4.3 years | 1.7 | IQR 3--5 years; max 12 years |

The species is short-lived overall, but individuals that reach reproduction typically persist for several years before flowering.

#### Seed-bank and germination estimates

Field germination was low across experiments.

| Experiment   | Overall germination | First-year germination | Maximum observed seed-bank duration |
| ------------ | ------------------: | ---------------------: | ----------------------------------: |
| Experiment 1 |                2.0% |                    83% |                             4 years |
| Experiment 2 |                1.0% |                    20% |                             4 years |
| Experiment 3 |                2.1% |                    89% |                             5 years |
| Experiment 4 |                8.6% |                    97% |                             5 years |
| Experiment 5 |                7.2% |                    92% |                             2 years |

For modeling, the useful field germination or emergence range is approximately 0.01--0.086.

The maximum observed seed-bank duration was five years, but late germination was uncommon. This supports including a seed-bank stage, but not treating the seed bank as a very large, long-lived, nearly immortal reservoir.

#### Seed predation

Seed predation and seed removal were substantial.

| Habitat / treatment | Seeds remaining after 48 h |
| ------------------- | -------------------------: |
| Roadside, exclusion |                        49% |
| Roadside, open      |                        32% |
| Scrub, exclusion    |                        68% |
| Scrub, open         |                        40% |

Across treatments, exclusion dishes retained more seeds than open dishes, and scrub dishes retained more seeds than roadside dishes. The abstract summarizes vertebrate seed predation as accounting for a 23% decline in seed numbers.

This supports using an effective seed-bank entry probability rather than assuming all intact seeds enter the seed bank.

#### Rangewide trends

Most populations declined in rangewide surveys, although APAFR increased.

| Population      |                                   Trend |
| --------------- | --------------------------------------: |
| APAFR           |                    +7.0 ± 1.9% per year |
| LWRSF           |                    -0.7 ± 3.7% per year |
| HHSP            |                   -10.0 ± 2.9% per year |
| ABS             |                    -6.9 ± 3.5% per year |
| CR731           |                    -7.2 ± 2.2% per year |
| Red Beach       | no reproductive plants observed in 2024 |
| Sheriff's Tower | no reproductive plants observed in 2024 |

The best estimate of total rangewide population size in 2020 was 7,444 reproductive individuals.

#### Habitat and soils

Disturbed habitats were common across the species' range. Approximately 43% of subpopulations occurred in disturbed habitat, including roadsides, sandy roads, firelanes, planted pine, canals, and ditches. Scrubby flatwoods accounted for 25% of subpopulations, followed by sandhill, oak scrub, dry-mesic flatwoods/savannah, and flatwoods.

More than 75% of plants occurred on grey sands, and about 21% occurred on white sands.

#### Main modeling implications

The main implications for demographic modeling are:

1. Survival should be size-dependent.
2. Flowering should carry a strong survival cost.
3. Growth should be size-dependent.
4. Flowering probability should be strongly size-dependent.
5. Fecundity should be modeled as flowering probability times flower-head number times intact seeds per head.
6. If flower-head number is already modeled, use 50.9 intact seeds per head rather than 3,181 intact seeds per reproductive plant.
7. The seed bank should be included, but not treated as a huge immortal pool.
8. A reasonable field germination or emergence range is 0.01--0.086.
9. Seed-bank persistence can last up to five years, but late germination is uncommon.
10. Disturbance effects are complex: fire reduces survival, does not reliably trigger recruitment booms, but may help maintain suitable open microhabitat.


## Dataset
This dataset provides demographic information on *Chrysopsis highlandsensis*.
Four datasets were provided by source. 
1. ‘demography_cleaned’ - transition between years
2. ‘site_history_data’ - site level disturbances
3. ‘plot_data’ - info on quadrats and transects
4. ‘demog_data’ - raw data


```{r data ab chhi}
df_og <- read_csv(file.path(dir_data, 'chrysopsis_highlandsensis_data.csv')) %>% 
  janitor::clean_names()

df_meta <- read_csv(file.path(dir_data, 'chrysopsis_highlandsensis_data_meta.csv')) %>% 
  janitor::clean_names()

df_fire <- read_csv(file.path(dir_data, 'chrysopsis_highlandsensis_fire.csv')) %>% 
  janitor::clean_names()
```


## Data exploration
The original dataset (`df_og`) contains 11,031 observations and 24 variables, primarily numeric, representing repeated measurements of individual plants across multiple sites and years. Each row corresponds to a plant observation at a given census interval, with identifiers linking individuals across time.

1. The temporal coverage spans from 1999 to 2024 with a relatively balanced distribution across the study period.

2. Plant size is recorded as diameter (`dia`), which serves as our key state variable for demographic analyses. More on that in **chapter 03XXX**. Diameter values range from very small individuals (≈0.2 cm) up to about 50.7, with a median of 5.4, indicating a right-skewed size distribution. It contains a substantial proportion of missing data (~46%).

3. Survival (`survival_1`) is a binary variable (0 = dead, 1 = alive in the next census) and is nearly complete across the dataset. The mean survival probability is approximately 0.59, indicating that just over half of the individuals survive between consecutive census intervals.

4. Flowering (`hd`) represents reproductive output (e.g. number of flower heads) and is highly zero-inflated and sparse, with a large proportion of missing values (~88%). When present, values range widely (0 to 594), suggesting strong variability in reproductive effort among individuals.


<details>
<summary>Expand for metadata</summary>
```{r data meta}
df_meta %>% print(n = nrow(.))
```
</details>


```{r data original}
df_og
```

<details>
<summary>Expand for data summary</summary> 
```{r data skim}
skimr::skim(df_og)
```
</details>

## Work dataframe/ adjustments

The original dataset (`df_og`) was systematically cleaned and restructured to produce a consistent, analysis-ready dataset for demographic modelling.

First, key variables were renamed to improve clarity and standardisation: the individual identifier (`identifier`) became `plant_id`, the census year (`year0`) became `year`, and the survival indicator (`survival_1`) was renamed to `survival.` The `plant_id` variable was then converted to a categorical (factor) variable to ensure proper handling of individual identities in subsequent analyses.

The data were ordered by site, plant identity, year, and survival status to ensure a consistent structure for grouped operations. A new binary variable, recruit, was created to indicate whether an individual was a recruit (astg == 1) or not.

Next, the dataset was aggregated to the level of site × plant × year. Within each group, summary variables were constructed:

1. Survival (`survives`) was defined as the minimum observed survival value, capturing mortality if it occurred, while returning NA if all values were missing.
2. Size at time t (`size_t0`) and size at time t+1 (`size_t1`) were defined as the maximum recorded diameters (`dia`,`dia_1`) within each group.
3. Flowering (`flower`) was summarized as the maximum observed flowering status (`hd`), indicating whether flowering occurred.
4. Recruitment (`recruit`) was summarized as the minimum value within the group, preserving the presence of recruits.

After summarization, the flowering variable was further processed: a copy (`fl_nr`) retained the original counts, while the main flower variable was converted to a binary indicator (flowering vs. non-flowering).

To support modelling within an integral projection framework, size variables were log-transformed. Specifically, the natural logarithm of size at times t and t+1 was calculated, and polynomial terms (squared and cubed) of log-size at time t were added to allow for nonlinear size effects in subsequent models.

The dataset was then augmented with external fire-treatment information (`df_fire`). This involved renaming variables for consistency, removing unnecessary columns, and joining the fire data to the main dataset by site and year.

Finally, the fire variable was recoded into a clear categorical factor with two levels: “No fire” (including missing values) and “Fire” (corresponding to burn treatments). T



```{r data extention}
df <- df_og %>% 
  rename(
    plant_id = identifier,
    year     = year0,
    survival = survival_1) %>%  
  mutate(
    plant_id = as.factor(plant_id)) %>%
  arrange(site, plant_id, year, survival) %>%
  mutate(recruit = ifelse(astg == 1, 1, 0)) %>%
  group_by(site , plant_id, year) %>%
  summarise(
    survives = if (all(is.na(survival))) NA_real_ else min(survival, na.rm = TRUE),
    size_t0  = if (all(is.na(dia)))      NA_real_ else max(dia,      na.rm = TRUE),
    size_t1  = if (all(is.na(dia_1)))    NA_real_ else max(dia_1,    na.rm = TRUE),
    flower   = if (all(is.na(hd)))       NA_real_ else max(hd,       na.rm = TRUE),
    recruit  = if (all(is.na(recruit)))  NA_real_ else min(recruit,  na.rm = TRUE),
    .groups = 'drop') %>%
  mutate(
    fl_nr  = flower,
    flower = if_else(flower > 0, 1, flower)) %>% 
  mutate(
    logsize_t0   = log(size_t0),
    logsize_t1   = log(size_t1),
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  full_join(
    
    df_fire %>%
      rename(year = burn_yr, fire = treatment) %>%
      select(!c(year0, notes)), 
    
    by = c('site', 'year')) %>%
  
  mutate(
    fire = case_when(
      is.na(fire)    ~ 'No fire',
      fire == 'burn' ~ 'Fire',
      TRUE           ~ NA_character_),
    fire = factor(fire, levels = c('No fire', 'Fire')))
```


## Graphical data exploration

```{r message=FALSE, warning=FALSE}
ggpairs(df %>%
          mutate(fire = case_when(
            fire == 'No fire' ~ 0,
            fire == 'Fire'    ~ 1)) %>%
          dplyr::select(survives, logsize_t0, logsize_t1, fl_nr, recruit, fire))
```

