# Data {#data}

This dataset provides demographic information from a long-term study of *Crotalaria avonensis*, an endangered small perennial herbaceous legume native to the Lake Wales Ridge in central Florida. The [study](https://portal.edirepository.org/nis/metadataviewer?packageid=edi.219.1) spans from 1998 to 2017 and monthly during the growing season tracks individual plant survival, size, and reproduction, recorded within permanent quadrats - 25 cm radius circles in stratified random positions within subjectively located belt transects. Additionally, the dataset contains information on fires, herbivory reduction through caging, and other events that might be of effect. Individuals were defined as groups of stems separated by 5 cm or more and marked with plastic toothpicks. Transplant data suggests this species has clonal growth. 

```{r data ab crav}
df_og <- read_csv(file.path(dir_data, 'chrysopsis_highlandsensis_data.csv')) %>% 
  janitor::clean_names()

df_meta <- read_csv(file.path(dir_data, 'chrysopsis_highlandsensis_data_meta.csv')) %>% 
  janitor::clean_names()
```

<details>
<summary>Expand to skim the raw dataframe</summary>
```{r data ab crav skim}
skimr::skim(df_og)
```
</details>

## Data exploration


```{r data mean original}
df_og
```

<details>
<summary>Expand to skim the condenced raw dataframe</summary> 
```{r}
skimr::skim(df_og_extended)
```
</details>

## Work dataframe/ adjustments


```{r data mean updated}
df <- 
```

<details>
<summary>Expand to skim the working dataframe</summary>
```{r}
skimr::skim(df)
```
</details>
