# Investigations {#investigation}


## Flower number, non integer

There are two observation that have a non integer flower numbers. 
The first one is disregarded in the analysis simply by not having size measures. The other we drop for the model of number of flowers depended on flowering. 

```{r 3 flower data}
df_fl <- df %>%
  filter(!is.na(flower),
         !is.na(fire),
         !is.na(size_t0),
         !is.na(survives)) %>%
  dplyr::select(plant_id, year, size_t0, size_t1, flower, fl_nr, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire)
```

```{r non integer flower}
df %>% dplyr::filter(fl_nr %% 1 != 0) %>% 
  dplyr::select(!c(logsize_t0, logsize_t1, logsize_t0_2,logsize_t0_3))
```

## Flower 

Flowering appears to be strongly structured by both size and fire history. However, preliminary inspection revealed an apparent inconsistency: individuals recorded under fire conditions seem to flower almost exclusively, which is biologically unlikely and suggests either sampling structure effects or missing data bias. 

Question: Would it make more sense to link fire to the probability of flowering the following year? This might be a more accurate link, since thus far we have not taken in to account when the fire event occurs, i.e. before of after censoring. 

```{r}
# Create binned flowering probability by fire group
df_fl_binned <- df_fl %>%
  group_split(fire) %>%
  map_df(~ plot_binned_prop(.x, 10, logsize_t0, flower) %>%
           mutate(fire = unique(.x$fire)))

# Plot overlapped flowering probability
ggplot(df_fl_binned, aes(x = logsize_t0, y = flower, color = fire)) +
  geom_jitter(
    data = df_fl,
    aes(x = logsize_t0, y = flower, color = fire),
    position = position_jitter(width = 0.1, height = 0.3),
    alpha = 0.1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, linewidth = 0.5) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8),
    title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    legend.title = element_blank(),
    legend.position = 'top') +
  labs(
    title = 'Flowering probability by fire status',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Flowering Probability')
```



```{r}
df3_flower <- df %>%
  mutate(
    flower = factor(flower),
    survives = factor(survives),
    fire = factor(fire),
    logsize_t0 = log(size_t0 + 0.001))

# Summarise flowering prevalence across site-year-fire combinations
df3_flower_sum <- df %>%
  mutate(flower = ifelse(is.na(flower), 0, flower)) %>%
  group_by(site, year, fire) %>%
  summarise(
    flowering = mean(flower),
    n = n(),
    .groups = "drop")

```





The raw dataset shows that flowering is heavily incomplete. Missing flowering values dominate the dataset (NA = 9102 in No fire, 477 in Fire). Fire subset is extremely small compared to non-fire conditions. Observed flowering under fire is only 28 individuals

```{r}
df %>%
  group_by(fire, flower) %>%
  summarise(n = n(), .groups = 'drop')
```
Apparent flowering frequency (biased by missingness). Fire shows 100% flowering rate, but only because: only flowering individuals are present in observed fire records. This is not a population-level estimate

```{r}
df %>%
  filter(!is.na(flower)) %>%
  group_by(fire) %>%
  summarise(
    n = n(),
    flowering_events = sum(flower),
    flowering_rate = mean(flower)
  )
```

Fire sample structure is extremely small.
Fire condition: only 17 individuals total, only 5 years of observations.

```{r}
df %>%
  group_by(fire) %>%
  summarise(
    n = n(),
    n_individuals = n_distinct(plant_id),
    n_years = n_distinct(year),
    min_year = min(year),
    max_year = max(year)
  )
```




```{r}
p_flowering_frequency <- df3_flower %>%
  count(flower) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = flower, y = prop, fill = flower)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = percent(prop, accuracy = 0.1)),
    vjust = -0.5,
    size = 4) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_viridis_d(end = 0.8) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8),
    title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    legend.position = 'none') +
  labs(
    title = 'Overall flowering frequency',
    subtitle = v_ggp_suffix,
    x = 'Flowering status',
    y = 'Proportion')

p_flowering_size_distribution <- ggplot(
  df3_flower,
  aes(x = logsize_t0, fill = flower)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(end = 0.85) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8),
    title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    legend.title = element_blank(),
    legend.position = 'top') +
  labs(
    title = 'Size distributions by flowering status',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Density')


(p_flowering_frequency | p_flowering_size_distribution)
```

## Density dependency

```{r}
df_re_qd <- df %>% 
  group_by(site, year) %>%
  summarise(rec_qd_t1 = sum(recruit, na.rm = TRUE), .groups = 'drop') %>%
  left_join(
    df %>% 
      group_by(site, year) %>% 
      summarise(
        nr_ind = sum(!is.na(size_t0)),
        fire = if_else(any(fire == 'Fire', na.rm = TRUE), 'Fire', 'No fire'),
        .groups = 'drop') %>% 
      mutate(year = year - 1),
    by = c('site', 'year')) %>%
  mutate(fire = factor(fire, levels = c('No fire', 'Fire')))

fig_re_dens <- ggplot(data = df_re_qd) + 
  geom_jitter(aes(y = rec_qd_t1, x = nr_ind, color = fire)) + 
  geom_smooth(aes(y = rec_qd_t1, x = nr_ind, color = fire), method = 'lm') + 
  theme_bw() + 
  labs(title    = 'Recruitment - density dependence: Quad level',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),   
       y        = expression('Number of recruits '     [t1])) +
  theme(plot.subtitle = element_text(size = 8))

fig_re_dens
```

## Check vital models with GAM
### Survival

```{r}
df_su <- df %>%
  filter(!is.na(survives), size_t0 != 0, !is.na(fire)) %>%
  dplyr::select(site, plant_id, year, size_t0, survives, size_t1,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire)

mod_su_0 <- glm(survives ~ fire, data = df_su, family = 'binomial') 
mod_su_1 <- glm(survives ~ logsize_t0 + fire, data = df_su, family = 'binomial') 
mod_su_2 <- glm(survives ~ logsize_t0 + logsize_t0_2 + fire, data = df_su, family = 'binomial')  
mod_su_3 <- glm(survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + fire, data = df_su, family = 'binomial')  
mods_su      <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAIC <- AICctab(mods_su, weights = T, sort = F)$dAIC
mods_su_sorted <- order(mods_su_dAIC)
if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index       <- mod_su_index_bestfit - 1 
} else {
  mod_su_index_bestfit <- v_mod_set_su +1
  v_mod_su_index       <- v_mod_set_su
}
mod_su_bestfit <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef   <- coef(mod_su_bestfit)


df_su_pred <- data.frame(
  logsize_t0   = rep(seq(min(df_su$logsize_t0  ), max(df_su$logsize_t0  ), length.out = 100), 2),
  logsize_t0_2 = rep(seq(min(df_su$logsize_t0_2), max(df_su$logsize_t0_2), length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         survives = predict(mod_su_bestfit, newdata = ., type = 'response'))
df_su_pre1 <- data.frame(
  logsize_t0   = rep(seq(min(df_su$logsize_t0  ), max(df_su$logsize_t0  ), length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         survives = predict(mod_su_1, newdata = ., type = 'response'))
df_su_pre3 <- data.frame(
  logsize_t0   = rep(seq(min(df_su$logsize_t0  ), max(df_su$logsize_t0  ), length.out = 100), 2),
  logsize_t0_2 = rep(seq(min(df_su$logsize_t0_2), max(df_su$logsize_t0_2), length.out = 100), 2),
  logsize_t0_3 = rep(seq(min(df_su$logsize_t0_3), max(df_su$logsize_t0_3), length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         survives = predict(mod_su_3, newdata = ., type = 'response'))


df_su_binned <- bind_rows(
  plot_binned_prop(filter(df_su, fire == 'No fire'), 10, logsize_t0, survives) %>% mutate(fire = 'No fire'),
  plot_binned_prop(filter(df_su, fire == 'Fire'   ), 10, logsize_t0, survives) %>% mutate(fire = 'Fire'))


mod_su_1_gam <- gam(survives ~ s(logsize_t0) + fire, data = df_su, family = binomial(), method = "REML")
df_su_pred_gam <- data.frame(
  logsize_t0 = rep(seq(min(df_su$logsize_t0), max(df_su$logsize_t0), length.out = 100), 2),
  fire       = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         survives = predict(mod_su_1_gam, newdata = ., type = 'response'))


```


```{r}
fig_su_line_combined <- ggplot() +
  geom_jitter(data = df_su, aes(x = logsize_t0, y = survives, color = fire), alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(data = df_su_pred,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "solid") +
  geom_line(data = df_su_pre1,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = df_su_pre3,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "twodash") +
  geom_line(data = df_su_pred_gam, aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "dotted") +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Survival to time t1') +
  theme(legend.position = 'none')

fig_su_bin_combined <- ggplot() +
  geom_point(data = df_su_binned, aes(x = logsize_t0, y = survives, color = fire)) +
  geom_errorbar(data = df_su_binned, aes(x = logsize_t0, ymin = lwr, ymax = upr, color = fire), width = 0.2) +
  geom_line(data = df_su_pred,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "solid") +
  geom_line(data = df_su_pre1,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = df_su_pre3,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "twodash") +
  geom_line(data = df_su_pred_gam, aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "dotted") +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  ylim(0, 1) +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Survival to time t1') +
  theme(legend.title = element_blank(), legend.position = 'top')

fig_su_all <- fig_su_line_combined + fig_su_bin_combined +
  plot_annotation(title = 'Survival', subtitle = v_ggp_suffix, theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_su_all
```

### Growth
```{r}
df_gr <- df %>%
  filter(size_t0 != 0, !is.na(size_t1), size_t1 > 0, !is.na(fire)) %>%
  mutate(logsize_t0 = log(size_t0), logsize_t1 = log(size_t1), logsize_t0_2 = logsize_t0^2, logsize_t0_3 = logsize_t0^3) %>%
  dplyr::select(plant_id, year, size_t0, size_t1, logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire)

mod_gr_0 <- lm(logsize_t1 ~ fire, data = df_gr)
mod_gr_1 <- lm(logsize_t1 ~ logsize_t0 + fire, data = df_gr)
mod_gr_2 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + fire, data = df_gr)
mod_gr_3 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + fire, data = df_gr)
mods_gr      <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAIC <- AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index       <- mod_gr_index_bestfit - 1 
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index       <- v_mod_set_gr
}

mod_gr_bestfit <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef   <- coef(mod_gr_bestfit)

df_gr_pred <- data.frame(
  logsize_t0 = rep(seq(min(df_gr$logsize_t0, na.rm = TRUE), max(df_gr$logsize_t0, na.rm = TRUE), length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire = factor(fire, levels = c('No fire', 'Fire')), logsize_t0_2 = logsize_t0^2, logsize_t0_3 = logsize_t0^3)

df_gr_pred$logsize_t1   <- predict(mod_gr_bestfit, newdata = df_gr_pred)
df_gr_pred$logsize_t1_1 <- predict(mod_gr_1,       newdata = df_gr_pred)
df_gr_pred$logsize_t1_2 <- predict(mod_gr_2,       newdata = df_gr_pred)

mod_gr_1_gam <- gam(logsize_t1 ~ s(logsize_t0) + fire, data = df_gr, method = "REML")
df_gr_pr_g <- data.frame(
  logsize_t0 = rep(seq(min(df_su$logsize_t0), max(df_su$logsize_t0), length.out = 100), 2),
  fire       = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire       = factor(fire, levels = c('No fire', 'Fire')),
         logsize_t1 = predict(mod_gr_1_gam, newdata = ., type = 'response'))

```


```{r}
fig_gr_line_combined <- ggplot(df_gr, aes(x = logsize_t0, y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = df_gr_pred, aes(x = logsize_t0, y = logsize_t1,   color = fire), linewidth = 1,   linetype = "solid") +
  geom_line(data = df_gr_pred, aes(x = logsize_t0, y = logsize_t1_1, color = fire), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = df_gr_pred, aes(x = logsize_t0, y = logsize_t1_2, color = fire), linewidth = 0.5, linetype = "twodash") +
  geom_line(data = df_gr_pr_g, aes(x = logsize_t0, y = logsize_t1,   color = fire), linewidth = 1,   linetype = "dotted") +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = expression('log(size)'[t0]), y = expression('log(size)'[t1])) +
  theme(legend.position = 'top', legend.title = element_blank(), plot.title = element_blank(), plot.subtitle = element_blank())

# Plot 2: Predicted vs observed
fig_gr_pred_combined <- ggplot(df_gr, aes(x = predict(mod_gr_bestfit, newdata = df_gr), y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = '', y = 'Observed', subtitle = 'cubic') +
  theme(legend.position = 'none', legend.title = element_blank(), plot.title = element_blank())

fig_gr_pred_1 <- ggplot(df_gr, aes(x = predict(mod_gr_1, newdata = df_gr), y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = 'Predicted', y = '', subtitle = 'linear') +
  theme(legend.position = 'none', legend.title = element_blank(), plot.title = element_blank())

fig_gr_pred_2 <- ggplot(df_gr, aes(x = predict(mod_gr_2, newdata = df_gr), y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = '', y = '', subtitle = 'quadratic') +
  theme(legend.position = 'right', legend.title = element_blank(), plot.title = element_blank())


fig_gr_line_combined +
  plot_annotation(
    title = 'Growth Model Prediction',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))


# Combine plots
fig_gr_all_combined <- fig_gr_pred_combined + fig_gr_pred_1 + fig_gr_pred_2 +
  plot_annotation(
    title = 'Growth Prediction',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_gr_all_combined
```

