# Investigations {#investigation}


## Flower number, non integer

There are two observation that have a non integer flower numbers. 
The first one is disregarded in the analysis simply by not having size measures. The other we drop for the model of number of flowers depended on flowering. 

```{r 3 flower data}
df_fl <- df %>%
  filter(!is.na(flower),
         !is.na(fire),
         !is.na(size_t0),
         !is.na(survives)) %>%
  dplyr::select(plant_id, year, size_t0, size_t1, flower, fl_nr, 
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire)
```

```{r non integer flower}
df %>% dplyr::filter(fl_nr %% 1 != 0) %>% 
  dplyr::select(!c(logsize_t0, logsize_t1, logsize_t0_2,logsize_t0_3))
```

## Flower 

Flowering appears to be strongly structured by both size and fire history. However, preliminary inspection revealed an apparent inconsistency: individuals recorded under fire conditions seem to flower almost exclusively, which is biologically unlikely and suggests either sampling structure effects or missing data bias. 

Question: Would it make more sense to link fire to the probability of flowering the following year? This might be a more accurate link, since thus far we have not taken in to account when the fire event occurs, i.e. before of after censoring. 

```{r}
# Create binned flowering probability by fire group
df_fl_binned <- df_fl %>%
  group_split(fire) %>%
  map_df(~ plot_binned_prop(.x, 10, logsize_t0, flower) %>%
           mutate(fire = unique(.x$fire)))

# Plot overlapped flowering probability
ggplot(df_fl_binned, aes(x = logsize_t0, y = flower, color = fire)) +
  geom_jitter(
    data = df_fl,
    aes(x = logsize_t0, y = flower, color = fire),
    position = position_jitter(width = 0.1, height = 0.3),
    alpha = 0.1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, linewidth = 0.5) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8),
    title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    legend.title = element_blank(),
    legend.position = 'top') +
  labs(
    title = 'Flowering probability by fire status',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Flowering Probability')
```



```{r}
df3_flower <- df %>%
  mutate(
    flower = factor(flower),
    survives = factor(survives),
    fire = factor(fire),
    logsize_t0 = log(size_t0 + 0.001))

# Summarise flowering prevalence across site-year-fire combinations
df3_flower_sum <- df %>%
  mutate(flower = ifelse(is.na(flower), 0, flower)) %>%
  group_by(site, year, fire) %>%
  summarise(
    flowering = mean(flower),
    n = n(),
    .groups = "drop")

```





The raw dataset shows that flowering is heavily incomplete. Missing flowering values dominate the dataset (NA = 9102 in No fire, 477 in Fire). Fire subset is extremely small compared to non-fire conditions. Observed flowering under fire is only 28 individuals. 

Attention! Over 9000 observations show an NA for flowering. This will be looked at in the following, i.e. **Chapter XXXX** 

```{r}
df %>%
  group_by(fire, flower) %>%
  summarise(n = n(), .groups = 'drop')
```
Apparent flowering frequency (biased by missingness). Fire shows 100% flowering rate, but only because: only flowering individuals are present in observed fire records. This is not a population-level estimate

```{r}
df %>%
  filter(!is.na(flower)) %>%
  group_by(fire) %>%
  summarise(
    n = n(),
    flowering_events = sum(flower),
    flowering_rate = mean(flower)
  )
```

Fire sample structure is small.

```{r}
df %>%
  group_by(fire) %>%
  summarise(
    n = n(),
    n_individuals = n_distinct(plant_id),
    n_years = n_distinct(year),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE))
```

### NA in flowering

Looking at the flowering structure, we find that only 3% of observations show flowering. More strikingly -and the focus of this investigation- is that just under 90% of observations have NA recorded for flowering. We find this highly suspicious. Furthermore, when examining the size distribution and density across the different flowering statuses, these NA values occur across the full range of plant sizes, ruling out the possibility that they are simply associated with new recruits. 

```{r}
p_flowering_frequency <- df3_flower %>%
  count(flower) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = flower, y = prop, fill = flower)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = percent(prop, accuracy = 0.1)),
    vjust = -0.5,
    size = 4) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_viridis_d(end = 0.8) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8),
    title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    legend.position = 'none') +
  labs(
    title = 'Overall flowering frequency',
    subtitle = v_ggp_suffix,
    x = 'Flowering status',
    y = 'Proportion')

p_flowering_size_distribution <- ggplot(
  df3_flower,
  aes(x = logsize_t0, fill = flower)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(end = 0.85) +
  theme_bw() +
  theme(
    axis.text = element_text(size = 8),
    title = element_text(size = 10),
    plot.subtitle = element_text(size = 8),
    legend.title = element_blank(),
    legend.position = 'top') +
  labs(
    title = 'Size distributions by flowering status',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Density')


(p_flowering_frequency | p_flowering_size_distribution)
```

The flowering structure reveals that only a very small proportion of observations are recorded as flowering (1), while the overwhelming majority are coded as NA. Specifically, flowering records (1) span the period from 1999–2024 and account for 1,038 observations, whereas explicit non-flowering records (0) are only present from 2021–2024 and comprise just 331 observations. In contrast, NA values dominate the dataset with 9,673 observations spanning nearly the entire monitoring period (1985–2025), suggesting that flowering status was not systematically recorded for much of the study.
```{r}
df3_flower %>%
  mutate(flower_status = case_when(
    is.na(flower) ~ "NA",
    as.character(flower) == "0" ~ "0",
    as.character(flower) == "1" ~ "1"
  )) %>%
  group_by(flower_status) %>%
  summarise(
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE),
    n = n(),
    .groups = "drop")
```


Examining the temporal distribution of missing flowering values further supports this interpretation. Of the 9,673 NA observations, 7,981 occur between 1985 and 2020, while 1,691 occur between 2021 and 2025. This pattern contrasts with the introduction of explicit 0 records after 2020 especially since there were only 331 observations of no flower occurrence within the newer time frame, yet still raises the possibility that some historical NA values may in fact correspond to non-flowering individuals (0) rather than truly missing information. However, this pattern leaves a drastic measure such as converting all NAs to 0 rather suspicious.
```{r}
df3_flower %>%
  filter(is.na(flower)) %>%
  mutate(period = case_when(
    year >= 1985 & year <= 2020 ~ "1985–2020",
    year >= 2021 & year <= 2025 ~ "2021–2025")) %>%
  group_by(period) %>%
  summarise(
    n_NA = n(),
    min_year = min(year),
    max_year = max(year),
    .groups = "drop")
```
Comparing the size distributions of observations coded as 0 and NA between 2021 and 2024 reveals substantial overlap, although plants recorded as non-flowering (0) tend to be somewhat larger on average. Observations with flower = 0 had a mean log-size of 1.75, compared to 1.07 for NA observations, while both groups covered a very similar overall size range. This suggests that NA values are not restricted to a distinct subset of very small or newly recruited individuals and supports the possibility that at least some NA observations in this timer period and especially the more historic data may correspond to non-flowering plants rather than entirely missing flowering information.
```{r}
# subset to recent years and compare flower = 0 vs NA
df_flower_compare <- df3_flower %>%
  filter(year >= 2021 & year <= 2024) %>%
  filter(is.na(flower) | as.character(flower) == "0") %>%
  mutate(flower_status = case_when(
    is.na(flower) ~ "NA",
    as.character(flower) == "0" ~ "0"))

# summary statistics
df_flower_compare %>%
  group_by(flower_status) %>%
  summarise(
    n = n(),
    mean_logsize = mean(logsize_t0, na.rm = TRUE),
    median_logsize = median(logsize_t0, na.rm = TRUE),
    sd_logsize = sd(logsize_t0, na.rm = TRUE),
    min_logsize = min(logsize_t0, na.rm = TRUE),
    max_logsize = max(logsize_t0, na.rm = TRUE),
    .groups = "drop"  )
```


```{r}
# density plot
ggplot(df_flower_compare,
       aes(x = logsize_t0, fill = flower_status)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(end = 0.85) +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = "top") +
  labs(
    title = "Size distribution of flower = 0 and flower = NA (2021–2024)",
    x = expression(log(size[t0])),
    y = "Density")
```


## Density dependency

```{r}
df_re_qd <- df %>% 
  group_by(site, year) %>%
  summarise(rec_qd_t1 = sum(recruit, na.rm = TRUE), .groups = 'drop') %>%
  left_join(
    df %>% 
      group_by(site, year) %>% 
      summarise(
        nr_ind = sum(!is.na(size_t0)),
        fire = if_else(any(fire == 'Fire', na.rm = TRUE), 'Fire', 'No fire'),
        .groups = 'drop') %>% 
      mutate(year = year - 1),
    by = c('site', 'year')) %>%
  mutate(fire = factor(fire, levels = c('No fire', 'Fire')))

fig_re_dens <- ggplot(data = df_re_qd) + 
  geom_jitter(aes(y = rec_qd_t1, x = nr_ind, color = fire)) + 
  geom_smooth(aes(y = rec_qd_t1, x = nr_ind, color = fire), method = 'lm') + 
  theme_bw() + 
  labs(title    = 'Recruitment - density dependence: Quad level',
       subtitle = v_ggp_suffix,
       x        = expression('Total parent plant area '[t0]),   
       y        = expression('Number of recruits '     [t1])) +
  theme(plot.subtitle = element_text(size = 8))

fig_re_dens
```

## Check vital models with GAM
### Survival

```{r}
df_su <- df %>%
  filter(!is.na(survives), size_t0 != 0, !is.na(fire)) %>%
  dplyr::select(site, plant_id, year, size_t0, survives, size_t1,
         logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire)

mod_su_0 <- glm(survives ~ fire, data = df_su, family = 'binomial') 
mod_su_1 <- glm(survives ~ logsize_t0 + fire, data = df_su, family = 'binomial') 
mod_su_2 <- glm(survives ~ logsize_t0 + logsize_t0_2 + fire, data = df_su, family = 'binomial')  
mod_su_3 <- glm(survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + fire, data = df_su, family = 'binomial')  
mods_su      <- list(mod_su_0, mod_su_1, mod_su_2, mod_su_3)
mods_su_dAIC <- AICctab(mods_su, weights = T, sort = F)$dAIC
mods_su_sorted <- order(mods_su_dAIC)
if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index       <- mod_su_index_bestfit - 1 
} else {
  mod_su_index_bestfit <- v_mod_set_su +1
  v_mod_su_index       <- v_mod_set_su
}
mod_su_bestfit <- mods_su[[mod_su_index_bestfit]]
mod_su_ranef   <- coef(mod_su_bestfit)


df_su_pred <- data.frame(
  logsize_t0   = rep(seq(min(df_su$logsize_t0  ), max(df_su$logsize_t0  ), 
                         length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire         = factor(fire, levels = c('No fire', 'Fire')), 
         logsize_t0_2 = logsize_t0^2 ) %>% 
  mutate(survives     = predict(mod_su_bestfit, newdata = ., type = 'response')) 
df_su_pre1 <- data.frame(
  logsize_t0   = rep(seq(min(df_su$logsize_t0  ),
                         max(df_su$logsize_t0  ), length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         survives = predict(mod_su_1, newdata = ., type = 'response'))
df_su_pre3 <- data.frame(
  logsize_t0   = rep(seq(min(df_su$logsize_t0  ), 
                         max(df_su$logsize_t0  ), length.out = 100), 2),
  fire         = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         logsize_t0_2 = logsize_t0^2,
         logsize_t0_3 = logsize_t0^3) %>% 
  mutate(survives = predict(mod_su_3, newdata = ., type = 'response'))


df_su_binned <- bind_rows(
  plot_binned_prop(filter(df_su, fire == 'No fire'), 10, logsize_t0, survives) %>% mutate(fire = 'No fire'),
  plot_binned_prop(filter(df_su, fire == 'Fire'   ), 10, logsize_t0, survives) %>% mutate(fire = 'Fire'))


mod_su_1_gam <- gam(survives ~ s(logsize_t0) + fire, data = df_su, family = binomial(), method = "REML")
df_su_pred_gam <- data.frame(
  logsize_t0 = rep(seq(min(df_su$logsize_t0), max(df_su$logsize_t0), length.out = 100), 2),
  fire       = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire     = factor(fire, levels = c('No fire', 'Fire')),
         survives = predict(mod_su_1_gam, newdata = ., type = 'response'))


```


```{r}
fig_su_line_combined <- ggplot() +
  geom_jitter(data = df_su, aes(x = logsize_t0, y = survives, color = fire), alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(data = df_su_pred,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "solid") +
  geom_line(data = df_su_pre1,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = df_su_pre3,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "twodash") +
  geom_line(data = df_su_pred_gam, aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "dotted") +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Survival to time t1') +
  theme(legend.position = 'none')

fig_su_bin_combined <- ggplot() +
  geom_point(data = df_su_binned, aes(x = logsize_t0, y = survives, color = fire)) +
  geom_errorbar(data = df_su_binned, aes(x = logsize_t0, ymin = lwr, ymax = upr, color = fire), width = 0.2) +
  geom_line(data = df_su_pred,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "solid") +
  geom_line(data = df_su_pre1,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = df_su_pre3,     aes(x = logsize_t0, y = survives, color = fire), linewidth = 0.5, linetype = "twodash") +
  geom_line(data = df_su_pred_gam, aes(x = logsize_t0, y = survives, color = fire), linewidth = 1,   linetype = "dotted") +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  ylim(0, 1) +
  labs(title = NULL, x = 'Size at time t0 (log())', y = 'Survival to time t1') +
  theme(legend.title = element_blank(), legend.position = 'top')

fig_su_all <- fig_su_line_combined + fig_su_bin_combined +
  plot_annotation(title = 'Survival', subtitle = v_ggp_suffix, theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_su_all
```

### Growth
```{r}
df_gr <- df %>%
  filter(size_t0 != 0, !is.na(size_t1), size_t1 > 0, !is.na(fire)) %>%
  mutate(logsize_t0 = log(size_t0), logsize_t1 = log(size_t1), logsize_t0_2 = logsize_t0^2, logsize_t0_3 = logsize_t0^3) %>%
  dplyr::select(plant_id, year, size_t0, size_t1, logsize_t0, logsize_t1, logsize_t0_2, logsize_t0_3, fire)

mod_gr_0 <- lm(logsize_t1 ~ fire, data = df_gr)
mod_gr_1 <- lm(logsize_t1 ~ logsize_t0 + fire, data = df_gr)
mod_gr_2 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + fire, data = df_gr)
mod_gr_3 <- lm(logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + fire, data = df_gr)
mods_gr      <- list(mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)
mods_gr_dAIC <- AICctab(mods_gr, weights = TRUE, sort = FALSE)$dAIC
mods_gr_sorted <- order(mods_gr_dAIC)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index       <- mod_gr_index_bestfit - 1 
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index       <- v_mod_set_gr
}

mod_gr_bestfit <- mods_gr[[mod_gr_index_bestfit]]
mod_gr_ranef   <- coef(mod_gr_bestfit)

df_gr_pred <- data.frame(
  logsize_t0 = rep(seq(min(df_gr$logsize_t0, na.rm = TRUE), max(df_gr$logsize_t0, na.rm = TRUE), length.out = 100), 2),
  fire = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire = factor(fire, levels = c('No fire', 'Fire')), logsize_t0_2 = logsize_t0^2, logsize_t0_3 = logsize_t0^3)

df_gr_pred$logsize_t1   <- predict(mod_gr_bestfit, newdata = df_gr_pred)
df_gr_pred$logsize_t1_1 <- predict(mod_gr_1,       newdata = df_gr_pred)
df_gr_pred$logsize_t1_2 <- predict(mod_gr_2,       newdata = df_gr_pred)

mod_gr_1_gam <- gam(logsize_t1 ~ s(logsize_t0) + fire, data = df_gr, method = "REML")
df_gr_pr_g <- data.frame(
  logsize_t0 = rep(seq(min(df_su$logsize_t0), max(df_su$logsize_t0), length.out = 100), 2),
  fire       = rep(c('No fire', 'Fire'), each = 100)) %>%
  mutate(fire       = factor(fire, levels = c('No fire', 'Fire')),
         logsize_t1 = predict(mod_gr_1_gam, newdata = ., type = 'response'))

```


```{r}
fig_gr_line_combined <- ggplot(df_gr, aes(x = logsize_t0, y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = df_gr_pred, aes(x = logsize_t0, y = logsize_t1,   color = fire), linewidth = 1,   linetype = "solid") +
  geom_line(data = df_gr_pred, aes(x = logsize_t0, y = logsize_t1_1, color = fire), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = df_gr_pred, aes(x = logsize_t0, y = logsize_t1_2, color = fire), linewidth = 0.5, linetype = "twodash") +
  geom_line(data = df_gr_pr_g, aes(x = logsize_t0, y = logsize_t1,   color = fire), linewidth = 1,   linetype = "dotted") +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = expression('log(size)'[t0]), y = expression('log(size)'[t1])) +
  theme(legend.position = 'top', legend.title = element_blank(), plot.title = element_blank(), plot.subtitle = element_blank())

# Plot 2: Predicted vs observed
fig_gr_pred_combined <- ggplot(df_gr, aes(x = predict(mod_gr_bestfit, newdata = df_gr), y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = '', y = 'Observed', subtitle = 'cubic') +
  theme(legend.position = 'none', legend.title = element_blank(), plot.title = element_blank())

fig_gr_pred_1 <- ggplot(df_gr, aes(x = predict(mod_gr_1, newdata = df_gr), y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = 'Predicted', y = '', subtitle = 'linear') +
  theme(legend.position = 'none', legend.title = element_blank(), plot.title = element_blank())

fig_gr_pred_2 <- ggplot(df_gr, aes(x = predict(mod_gr_2, newdata = df_gr), y = logsize_t1, color = fire)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', linewidth = 1) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(x = '', y = '', subtitle = 'quadratic') +
  theme(legend.position = 'right', legend.title = element_blank(), plot.title = element_blank())


fig_gr_line_combined +
  plot_annotation(
    title = 'Growth Model Prediction',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))


# Combine plots
fig_gr_all_combined <- fig_gr_pred_combined + fig_gr_pred_1 + fig_gr_pred_2 +
  plot_annotation(
    title = 'Growth Prediction',
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = 'bold'),
      plot.subtitle = element_text(size = 10, face = 'italic')))

fig_gr_all_combined
```

