# Year-specific disturbance IPM {#dist-year-flx2-fl2re}

## Overview

This chapter describes the year-specific integral projection model (IPM) for
*Chrysopsis highlandsensis*. The model follows individuals from year `t0` to
`t1` and contains five demographic processes:

1. survival;
2. growth in log rosette diameter;
3. flowering probability;
4. flowerhead number conditional on flowering; and
5. recruitment from flowerheads in `t0` to recruits in `t1`.

Fire is included as a fixed effect in the four size-dependent vital rates.
Survival, growth, flowering, and flowerhead production vary among years.
Recruitment may also vary among recruit years, but it has **no site random
effect**. Sites are retained when constructing observed population sizes and
initial size distributions, not as recruitment coefficients.

The hidden chunk below runs the standalone model script. The following sections
show only the code needed to understand the main decisions and results.

```{r dist-year-run, include=FALSE}
script_candidates <- c(
  file.path(
    'C:', 'code', 'RUPDemo_IPMs',
    'archbold', 'R', 'chhi',
    'archbold_chhi_ipm_year_dist_flx2_fl2re.R'),
  'archbold_chhi_ipm_year_dist_flx2_fl2re.R')

script_path <- script_candidates[file.exists(script_candidates)][1]

if (is.na(script_path)) {
  stop('Could not locate archbold_chhi_ipm_year_dist_flx2_fl2re.R')
}

source(script_path, local = knitr::knit_global())
```

```{r dist-year-summaries, include=FALSE}
model_formula <- function(model) {
  paste(deparse(stats::formula(model)), collapse = ' ')
}

fixed_effect <- function(model, term) {
  coefs <- if (inherits(model, 'merMod')) {
    lme4::fixef(model)
  } else {
    stats::coef(model)
  }

  if (term %in% names(coefs)) {
    unname(coefs[[term]])
  } else {
    NA_real_
  }
}

selected_models <- tibble::tibble(
  process = c(
    'Survival',
    'Growth',
    'Flowering probability',
    'Flowerheads given flowering',
    'Flowerheads to recruits'),
  selected_formula = c(
    model_formula(mod_su_best),
    model_formula(mod_gr_best),
    model_formula(mod_fl_best),
    model_formula(mod_fl_n_best),
    model_formula(mod_re_best)))

lambda_year_long <- lambda_year %>%
  tidyr::pivot_longer(
    cols = c(lambda_no_fire, lambda_fire),
    names_to = 'fire_state',
    values_to = 'lambda') %>%
  mutate(
    fire_state = dplyr::recode(
      fire_state,
      lambda_no_fire = 'No fire',
      lambda_fire = 'Fire'))

validation_summary <- df_compare %>%
  summarise(
    n_years = n(),
    cor_asym = cor(obs_pgr, asym_lambda, use = 'complete.obs'),
    cor_proj = cor(obs_pgr, proj_lambda, use = 'complete.obs'),
    mae_asym = mean(abs(obs_pgr - asym_lambda), na.rm = TRUE),
    mae_proj = mean(abs(obs_pgr - proj_lambda), na.rm = TRUE))
```

## How recruitment was defined

Recruitment was initially identified only from annual seedling stage
(`astg == 1`). Console checks showed that seedlings were not counted in
multiple years, but the quarterly censuses contained additional plants marked
as `new plant`. Most first observations of this kind occurred in the first year
that a site was monitored, when prior presence was unknown; these were not
accepted as recruits. The remaining 280 plants appeared after site monitoring
had begun and were therefore added as recruits.

The final rule classifies a plant as a recruit when it is either a seedling or a
first observation after site establishment with a quarterly `new plant` status.

```{r dist-year-recruit-code, eval=FALSE}
df_recruit_flag <- df_og %>%
  rename(
    plant_id = identifier,
    year = year0) %>%
  mutate(
    new_plant_status = if_any(
      c(s_03, s_06, s_09, s_12), ~ .x == 3)) %>%
  group_by(site) %>%
  mutate(site_first_year = min(year, na.rm = TRUE)) %>%
  group_by(site, plant_id) %>%
  mutate(first_plant_year = min(year, na.rm = TRUE)) %>%
  group_by(site, plant_id, year) %>%
  summarise(
    seedling = any(astg == 1, na.rm = TRUE),
    new_plant_status = any(new_plant_status, na.rm = TRUE),
    site_first_year = first(site_first_year),
    first_plant_year = first(first_plant_year),
    .groups = 'drop') %>%
  mutate(
    recruit = as.integer(
      seedling |
        (new_plant_status &
           year == first_plant_year &
           year > site_first_year)))
```

After this classification, the annual dataset contains
`r sum(df$recruit == 1, na.rm = TRUE)` recruit records across
`r dplyr::n_distinct(df$site)` sites and years
`r min(df$year, na.rm = TRUE)`--`r max(df$year, na.rm = TRUE)`.

## Annual demographic data

Repeated observations are collapsed to one record per site, plant, and year.
The annual record retains survival, maximum rosette diameter, next-year
rosette diameter, and maximum flowerhead count. Diameter is log-transformed for
all size-dependent models. Fire is attached to the transition beginning in
`t0`.

```{r dist-year-data-code, eval=FALSE}
df <- df_og %>%
  rename(
    plant_id = identifier,
    year = year0,
    survival = survival_1) %>%
  group_by(site, plant_id, year) %>%
  summarise(
    survives = if (all(is.na(survival))) {
      NA_real_
    } else {
      min(survival, na.rm = TRUE)
    },
    size_t0 = if (all(is.na(dia))) NA_real_ else max(dia, na.rm = TRUE),
    size_t1 = if (all(is.na(dia_1))) NA_real_ else max(dia_1, na.rm = TRUE),
    flower = if (all(is.na(hd))) NA_real_ else max(hd, na.rm = TRUE),
    .groups = 'drop') %>%
  left_join(df_recruit_flag, by = c('site', 'plant_id', 'year')) %>%
  mutate(
    fl_nr = flower,
    flower = if_else(flower > 0, 1, flower),
    logsize_t0 = log(size_t0),
    logsize_t1 = log(size_t1),
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)
```

Keeping `fl_nr` and `flower` separately is important. `flower` is the binary
response for flowering probability, whereas `fl_nr` is the count response for
flowerhead production among flowering plants.

## Vital-rate models

Candidate models range from intercept-only to cubic size dependence. AICc
selects the supported polynomial complexity. Fire remains a fixed effect, while
year enters as a random effect.

```{r dist-year-selected-models}
knitr::kable(
  selected_models,
  col.names = c('Process', 'Selected model'),
  caption = 'Vital-rate models selected by AICc.')
```

### Survival

Survival is binary and is fitted with a binomial mixed model. The year random
intercept allows generally favorable and unfavorable survival years, while the
fire coefficient shifts the curve between burned and unburned transitions.

```{r dist-year-survival-code, eval=FALSE}
mod_su_0 <- glmer(
  survives ~ fire + (1 | year),
  data = df_su, family = binomial)

mod_su_1 <- glmer(
  survives ~ logsize_t0 + fire + (1 | year),
  data = df_su, family = binomial)

# Quadratic and cubic size candidates are fitted in the same way.
mod_su_best <- mods_su[[order(mods_su_dAIC)[1]]]
```

The selected model is
`r model_formula(mod_su_best)`. The fixed fire coefficient is
`r round(fixed_effect(mod_su_best, 'fireFire'), 3)` on the logit scale. In the
figure, separation between the black and red curves is the modeled fire effect;
vertical changes among panels are year effects.

```{r dist-year-survival-figure, fig.width=12, fig.height=15, out.width='100%'}
g_surv_years +
  patchwork::plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix)
```

### Growth

Growth predicts log size in `t1` from log size in `t0`. The model includes a
random year intercept and size slope, allowing both average growth and the
size--growth relationship to change among years. Residual variance is modeled
separately because growth variability changes along the fitted growth curve.

```{r dist-year-growth-code, eval=FALSE}
mod_gr_1 <- lmer(
  logsize_t1 ~ logsize_t0 + fire + (logsize_t0 | year),
  data = df_gr)

mod_gr_2 <- lmer(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + fire +
    (logsize_t0 | year),
  data = df_gr)

mod_gr_best <- mods_gr[[order(mods_gr_dAIC)[1]]]

gr_var <- nls(
  resid(mod_gr_best)^2 ~ a * exp(b * fitted(mod_gr_best)),
  start = list(a = 1, b = 0))
```

The selected growth model is
`r model_formula(mod_gr_best)`. The blue dashed line in each panel represents
no change in size. Curves above this line indicate expected growth, whereas
curves below it indicate expected shrinkage.

```{r dist-year-growth-figure, fig.width=12, fig.height=15, out.width='100%'}
g_grow_years
```

### Flowering probability and flowerhead number

Reproduction is separated into two processes. First, a binomial model estimates
the probability that a plant flowers. Second, a negative-binomial model
estimates flowerhead number among plants that flowered. This avoids treating all
zero flowerhead observations as part of the flowerhead count distribution.

```{r dist-year-flowering-code, eval=FALSE}
mod_fl_1 <- glmer(
  flower ~ logsize_t0 + fire + (1 | year),
  data = df_fl, family = binomial)

mod_fl_n_1 <- glmer.nb(
  fl_nr ~ logsize_t0 + fire + (1 | year),
  data = df_fl_cond)
```

The selected flowering model is
`r model_formula(mod_fl_best)`, and the selected conditional flowerhead model is
`r model_formula(mod_fl_n_best)`. Their expected product is the mean number of
flowerheads produced by an individual of a given size:

\[
E(FH \mid x) = \Pr(\text{flowering} \mid x)\,
E(FH \mid \text{flowering},x).
\]

The following plots reproduce the year-panel logic used for survival and growth.
Points are binned observations and lines are conditional year predictions.

```{r dist-year-flowering-plots, fig.width=12, fig.height=15, out.width='100%'}
make_year_prediction_grid <- function(data, n = 100) {
  data %>%
    group_by(year) %>%
    summarise(
      xmin = min(logsize_t0, na.rm = TRUE),
      xmax = max(logsize_t0, na.rm = TRUE),
      .groups = 'drop') %>%
    tidyr::crossing(
      fire = factor(
        c('No fire', 'Fire'),
        levels = c('No fire', 'Fire'))) %>%
    rowwise() %>%
    mutate(logsize_t0 = list(seq(xmin, xmax, length.out = n))) %>%
    tidyr::unnest(logsize_t0) %>%
    ungroup() %>%
    mutate(
      year = factor(year, levels = levels(data$year)),
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3)
}

fl_pred <- make_year_prediction_grid(df_fl) %>%
  mutate(
    prediction = predict(
      mod_fl_best,
      newdata = ., type = 'response', re.form = NULL))

fl_points <- df_fl %>%
  group_by(year, fire) %>%
  mutate(bin = dplyr::ntile(logsize_t0, 8)) %>%
  group_by(year, fire, bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    response = mean(flower, na.rm = TRUE),
    .groups = 'drop')

fig_fl_years <- ggplot() +
  geom_point(
    data = fl_points,
    aes(logsize_t0, response, color = fire),
    size = 1) +
  geom_line(
    data = fl_pred,
    aes(logsize_t0, prediction, color = fire),
    linewidth = 0.8) +
  facet_wrap(~ year, ncol = 4) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  labs(
    title = 'Flowering probability - year specific',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Flowering probability',
    color = NULL) +
  theme(legend.position = 'top')

fig_fl_years
```

```{r dist-year-flowerhead-plots, fig.width=12, fig.height=15, out.width='100%'}
fln_pred <- make_year_prediction_grid(df_fl_cond) %>%
  mutate(
    prediction = predict(
      mod_fl_n_best,
      newdata = ., type = 'response', re.form = NULL))

fln_points <- df_fl_cond %>%
  group_by(year, fire) %>%
  mutate(bin = dplyr::ntile(logsize_t0, 8)) %>%
  group_by(year, fire, bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    response = mean(fl_nr, na.rm = TRUE),
    .groups = 'drop')

fig_fln_years <- ggplot() +
  geom_point(
    data = fln_points,
    aes(logsize_t0, response, color = fire),
    size = 1) +
  geom_line(
    data = fln_pred,
    aes(logsize_t0, prediction, color = fire),
    linewidth = 0.8) +
  facet_wrap(~ year, ncol = 4, scales = 'free_y') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(
    title = 'Flowerhead number given flowering - year specific',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Expected flowerheads',
    color = NULL) +
  theme(legend.position = 'top')

fig_fln_years
```

## Flowerhead-to-recruit transition

Recruitment is fitted at the site-year level. Flowerheads produced in year `t0`
are paired with recruits observed in `t1`, and both quantities are transformed
with `log1p`. The candidate models allow recruitment to vary among recruit years
but deliberately omit site random effects.

```{r dist-year-recruitment-model-code, eval=FALSE}
df_re <- df_site_year %>%
  transmute(site, year_t1 = year, re_t1 = re) %>%
  left_join(
    df_site_year %>%
      transmute(
        site,
        year_t0 = year,
        year_t1 = year + 1,
        fh_t0 = fh),
    by = c('site', 'year_t1')) %>%
  mutate(
    re_t1_log = log1p(re_t1),
    fh_t0_log = log1p(fh_t0))

mod_re_0 <- glm(
  re_t1_log ~ fh_t0_log,
  data = df_re)

mod_re_2 <- lmer(
  re_t1_log ~ fh_t0_log + (fh_t0_log | year_t1),
  data = df_re, REML = FALSE)
```

The selected recruitment model is
`r model_formula(mod_re_best)`, fitted to `r nrow(df_re)` site-year
transitions. Colored symbols identify sites, but each panel has one black
recruitment relationship because recruitment coefficients vary by year rather
than by site.

The intercept is retained. Consequently, the relationship is an **effective
flowerhead-to-recruit transition**, not a claim that all recruits originate
only from flowerheads counted in the immediately preceding census. A non-zero
intercept can absorb carry-over from persistent seeds, unobserved flowerheads,
and other mismatch between annual flowerhead and recruit censuses.

```{r dist-year-recruitment-figure, fig.width=12, fig.height=15, out.width='100%'}
g_re_years
```

## From vital rates to the IPM

The parameter-extraction section separates population-level fixed effects from
year-specific conditional coefficients. `make_ipm_pars()` begins with the mean
parameter list and replaces coefficients with those for the requested year.
Recruitment-year effects are indexed by `t1`, because they describe the year in
which recruits are observed.

```{r dist-year-parameter-code, eval=FALSE}
pars_i <- make_ipm_pars(
  pars_mean = pars_all_mean,
  pars_year = pars_all_year,
  pars_re_year_t1 = pars_all_re_year_t1,
  year = year,
  re_year_t1 = year + 1)
```

The annual kernel is the sum of two components:

\[
K(y,x) = T(y,x) + F(y,x).
\]

`T` combines survival and growth. `F` combines flowering probability,
flowerhead number, the effective recruits-per-flowerhead conversion, and the
observed recruit-size distribution.

```{r dist-year-kernel-code, eval=FALSE}
# Survival-growth transition
Tmat[, i] <- Gmat[, i] * Smat[i]

# Fertility contribution
Rvec <- fhx(x, pars, fire) * recruits_per_flowerhead
Fmat <- outer(recruit_size_density, Rvec) * h

# Complete kernel
K <- Tmat + Fmat
```

The mesh contains `r pars_all_mean$mat_siz` size classes, ranging from
`r round(pars_all_mean$L, 2)` to `r round(pars_all_mean$U, 2)` on the log-size
scale. Eviction correction places growth probability outside the mesh into the
smallest or largest size class, preventing loss of probability mass at the
boundaries.

## Year-specific population growth

For every year, the model builds one no-fire kernel and one fire kernel. The
dominant eigenvalue is the asymptotic population growth rate, \(\lambda\), for
that year's vital rates.

```{r dist-year-lambda-code, eval=FALSE}
lambda_year <- data.frame(
  year = ipm_years,
  lambda_no_fire = sapply(
    ipm_years,
    function(i) lambda_ipm_year(i, fire = 0)),
  lambda_fire = sapply(
    ipm_years,
    function(i) lambda_ipm_year(i, fire = 1)))
```

Across `r nrow(lambda_year)` modeled years, no-fire \(\lambda\) ranges from
`r round(min(lambda_year$lambda_no_fire, na.rm = TRUE), 3)` to
`r round(max(lambda_year$lambda_no_fire, na.rm = TRUE), 3)`, while fire
\(\lambda\) ranges from
`r round(min(lambda_year$lambda_fire, na.rm = TRUE), 3)` to
`r round(max(lambda_year$lambda_fire, na.rm = TRUE), 3)`. Fire produces the
lower value in
`r sum(lambda_year$lambda_fire < lambda_year$lambda_no_fire, na.rm = TRUE)` of
`r nrow(lambda_year)` years. Variation shared by both lines reflects changing
year effects; the gap between the lines reflects the fixed fire effects carried
through all vital rates.

```{r dist-year-lambda-figure, fig.width=8, fig.height=4.5, out.width='90%'}
fig_lambda_year <- ggplot(
  lambda_year_long,
  aes(x = year, y = lambda, color = fire_state)) +
  geom_hline(yintercept = 1, linetype = 'dotted') +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(
    title = 'Year-specific asymptotic population growth',
    subtitle = v_ggp_suffix,
    x = 'Projection year',
    y = expression(lambda),
    color = NULL) +
  theme(legend.position = 'top')

fig_lambda_year
```

Values above one indicate asymptotic increase under that year's kernel, and
values below one indicate decline. These are long-run rates for a fixed annual
kernel; they are not the same as the observed one-year change from the current
population structure.

## Observed versus modeled growth

Validation retains only site-year pairs observed in both `t0` and `t1`. Each
site is projected from its observed size distribution using the corresponding
year and fire kernel. Site projections are then combined to obtain:

- **projected lambda**, the modeled one-year change from the observed size
  distribution; and
- **asymptotic lambda**, the dominant eigenvalue of the kernel, averaged across
  sites with abundance weights.

```{r dist-year-validation-code, eval=FALSE}
n_proj <- K %*% n_obs

proj_lambda <- sum(n_proj) / sum(n_obs)
asym_lambda <- Re(eigen(K, only.values = TRUE)$values[1])
obs_pgr <- n_t1 / n_t0
```

The comparison contains `r validation_summary$n_years` annual transitions. The
correlation with observed growth is
`r round(validation_summary$cor_proj, 2)` for projected lambda and
`r round(validation_summary$cor_asym, 2)` for asymptotic lambda. Their mean
absolute errors are `r round(validation_summary$mae_proj, 2)` and
`r round(validation_summary$mae_asym, 2)`, respectively.

```{r dist-year-validation-figure, fig.width=9, fig.height=4.5, out.width='95%'}
g_mod_vs_obs
```

Points on the dashed line would represent exact agreement. Differences are not
necessarily model failures: observed growth includes demographic stochasticity,
sampling variation, and recruitment pulses, while the IPM describes expected
rates from fitted vital-rate relationships. Projected lambda is usually the
fairer one-year comparison because it starts from the observed size structure;
asymptotic lambda answers a longer-term question.

## Main conclusions

- The model represents reproduction as flowering probability multiplied by
  flowerhead number, followed by an empirical flowerhead-to-recruit transition.
- The revised recruitment definition captures both seedlings and verified new
  plants appearing after site monitoring began.
- Year variation is included throughout the vital-rate pathway, whereas site
  recruitment effects are intentionally excluded.
- Fire affects the IPM through fixed effects on survival, growth, flowering,
  and flowerhead production.
- The observed-versus-modeled comparison distinguishes a one-year projection
  from the asymptotic growth rate of the annual kernel.
