# Year- and site-specific disturbance IPM {#dist-year-site-flx2-fl2re}

## Overview

This chapter extends the year-specific IPM from the previous chapter by
allowing the flowerhead-to-recruit relationship to vary among sites as well as
among recruit years. The vital-rate pathway remains:

1. survival;
2. growth in log rosette diameter;
3. flowering probability;
4. flowerhead number conditional on flowering; and
5. recruitment from flowerheads in `t0` to recruits in `t1`.

Fire is a fixed effect in the four size-dependent vital rates. Survival,
growth, flowering, and flowerhead production vary among years. Recruitment is
fitted from site-year totals, and the candidate set allows random intercepts
and slopes for site, recruit year, or both.

The hidden chunk below runs the standalone script. Later chunks show only the
parts needed to understand the model and its results.

```{r dist-year-site-run, include=FALSE}
script_candidates <- c(
  file.path(
    'archbold', 'R', 'chhi',
    'archbold_chhi_ipm_dist_year_site_flx2_fl2re.R'),
  'archbold_chhi_ipm_dist_year_site_flx2_fl2re.R')

script_path <- script_candidates[file.exists(script_candidates)][1]

if (is.na(script_path)) {
  stop('Could not locate the year- and site-specific IPM script')
}

source(script_path, local = knitr::knit_global())
```

```{r dist-year-site-summaries, include=FALSE}
model_formula <- function(model) {
  paste(deparse(stats::formula(model)), collapse = ' ')
}

fixed_effect <- function(model, term) {
  coefs <- if (inherits(model, 'merMod')) {
    lme4::fixef(model)
  } else {
    stats::coef(model)
  }

  if (term %in% names(coefs)) {
    unname(coefs[[term]])
  } else {
    NA_real_
  }
}

random_groups <- function(model) {
  if (!inherits(model, 'merMod')) {
    return('None')
  }

  groups <- names(lme4::ranef(model))
  if (length(groups) == 0) 'None' else paste(groups, collapse = ', ')
}

selected_models <- tibble::tibble(
  process = c(
    'Survival',
    'Growth',
    'Flowering probability',
    'Flowerheads given flowering',
    'Flowerheads to recruits'),
  selected_formula = c(
    model_formula(mod_su_best),
    model_formula(mod_gr_best),
    model_formula(mod_fl_best),
    model_formula(mod_fl_n_best),
    model_formula(mod_re_best)))

recruitment_candidates <- tibble::tibble(
  candidate = paste0('Model ', seq_along(mods_re) - 1),
  formula = vapply(mods_re, model_formula, character(1)),
  delta_aic = as.numeric(mods_re_dAIC)) %>%
  arrange(delta_aic)

recruitment_groups <- random_groups(mod_re_best)

site_effect_present <- inherits(mod_re_best, 'merMod') &&
  'site' %in% names(lme4::ranef(mod_re_best))

year_effect_present <- inherits(mod_re_best, 'merMod') &&
  'year_t1' %in% names(lme4::ranef(mod_re_best))

# Recompute projections with site-specific recruitment enabled. The switch in
# the standalone script is changed only after its first projections are built,
# so the objects below ensure that site effects are actually used here.
use_site_specific_recruitment <- TRUE

df_proj_site_specific <- dplyr::bind_rows(
  lapply(seq_len(nrow(df_obs_pgr_site)), function(i) {
    project_one_site_year(
      yr = df_obs_pgr_site$year[i],
      site_i = df_obs_pgr_site$site[i])
  }))

df_compare_site_specific <- df_obs_pgr_site %>%
  left_join(
    df_proj_site_specific,
    by = c('year', 'site', 'fire_num')) %>%
  mutate(obs_pgr = n_t1 / n_t0)

df_compare_specific <- df_compare_site_specific %>%
  group_by(year) %>%
  summarise(
    n_t0 = sum(n_t0, na.rm = TRUE),
    n_t1 = sum(n_t1, na.rm = TRUE),
    obs_pgr = n_t1 / n_t0,
    n_obs_model = sum(n_obs, na.rm = TRUE),
    n_proj_model = sum(n_proj, na.rm = TRUE),
    proj_lambda = n_proj_model / n_obs_model,
    asym_lambda = weighted.mean(
      asym_lambda, w = n_obs, na.rm = TRUE),
    fire = if_else(
      any(fire == 'Fire', na.rm = TRUE), 'Fire', 'No fire'),
    .groups = 'drop') %>%
  mutate(fire = factor(fire, levels = c('No fire', 'Fire')))

validation_summary <- df_compare_specific %>%
  summarise(
    n_years = n(),
    cor_asym = cor(obs_pgr, asym_lambda, use = 'complete.obs'),
    cor_proj = cor(obs_pgr, proj_lambda, use = 'complete.obs'),
    mae_asym = mean(abs(obs_pgr - asym_lambda), na.rm = TRUE),
    mae_proj = mean(abs(obs_pgr - proj_lambda), na.rm = TRUE))
```

## Recruitment definition

The recruit definition is unchanged from the previous model. A plant is counted
as a recruit when it is either recorded as a seedling or first appears after
site monitoring began with a quarterly `new plant` status. Plants first seen in
a site's initial census are excluded because their prior presence is unknown.

This rule was adopted after checking that seedling recruits were not counted in
multiple years and finding 280 additional new plants that appeared during
established monitoring.

```{r dist-year-site-recruit-code, eval=FALSE}
df_recruit_flag <- df_og %>%
  rename(
    plant_id = identifier,
    year = year0) %>%
  mutate(
    new_plant_status = if_any(
      c(s_03, s_06, s_09, s_12), ~ .x == 3)) %>%
  group_by(site) %>%
  mutate(site_first_year = min(year, na.rm = TRUE)) %>%
  group_by(site, plant_id) %>%
  mutate(first_plant_year = min(year, na.rm = TRUE)) %>%
  group_by(site, plant_id, year) %>%
  summarise(
    seedling = any(astg == 1, na.rm = TRUE),
    new_plant_status = any(new_plant_status, na.rm = TRUE),
    site_first_year = first(site_first_year),
    first_plant_year = first(first_plant_year),
    .groups = 'drop') %>%
  mutate(
    recruit = as.integer(
      seedling |
        (new_plant_status &
           year == first_plant_year &
           year > site_first_year)))
```

The annual dataset contains `r sum(df$recruit == 1, na.rm = TRUE)` recruit
records from `r dplyr::n_distinct(df$site)` sites.

## Annual demographic data

Repeated observations are collapsed to one plant-year record. The model keeps
annual survival, rosette diameter in `t0` and `t1`, flowerhead number, recruit
status, and fire exposure. Diameter is log-transformed before fitting the
size-dependent vital rates.

The flowerhead variable is stored twice: `flower` is converted to a binary
flowering indicator, while `fl_nr` retains the original flowerhead count.

```{r dist-year-site-data-code, eval=FALSE}
df <- df_og %>%
  rename(
    plant_id = identifier,
    year = year0,
    survival = survival_1) %>%
  group_by(site, plant_id, year) %>%
  summarise(
    survives = if (all(is.na(survival))) {
      NA_real_
    } else {
      min(survival, na.rm = TRUE)
    },
    size_t0 = if (all(is.na(dia))) NA_real_ else max(dia, na.rm = TRUE),
    size_t1 = if (all(is.na(dia_1))) NA_real_ else max(dia_1, na.rm = TRUE),
    flower = if (all(is.na(hd))) NA_real_ else max(hd, na.rm = TRUE),
    .groups = 'drop') %>%
  left_join(df_recruit_flag, by = c('site', 'plant_id', 'year')) %>%
  mutate(
    fl_nr = flower,
    flower = if_else(flower > 0, 1, flower),
    logsize_t0 = log(size_t0),
    logsize_t1 = log(size_t1),
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)
```

## Vital-rate models

The size-based vital-rate models are the same as in the year-specific model.
Candidate models range from intercept-only to cubic size dependence, and AICc
selects the supported polynomial order. Fire is a fixed effect and year is a
random effect.

```{r dist-year-site-selected-models}
knitr::kable(
  selected_models,
  col.names = c('Process', 'Selected model'),
  caption = 'Vital-rate models selected by AICc.')
```

### Survival and growth

Survival is modeled with a binomial mixed model. Growth is modeled on the
log-size scale with a random year intercept and size slope. The selected
survival model is `r model_formula(mod_su_best)`, and the selected growth model
is `r model_formula(mod_gr_best)`.

In the survival panels, black and red curves compare no-fire and fire
transitions. In the growth panels, the blue dashed line represents no change in
size.

```{r dist-year-site-survival-figure, fig.width=12, fig.height=15, out.width='100%'}
g_surv_years +
  patchwork::plot_annotation(
    title = 'Survival - year specific',
    subtitle = v_ggp_suffix)
```

```{r dist-year-site-growth-figure, fig.width=12, fig.height=15, out.width='100%'}
g_grow_years
```

### Flowering and flowerhead production

Reproduction is separated into flowering probability and flowerhead number
conditional on flowering. Their product gives expected flowerhead production
for a plant of size `x`:

\[
E(FH \mid x) = \Pr(\text{flowering} \mid x)
E(FH \mid \text{flowering},x).
\]

The selected flowering model is `r model_formula(mod_fl_best)`. The selected
conditional flowerhead model is `r model_formula(mod_fl_n_best)`.

```{r dist-year-site-flowering-plots, fig.width=12, fig.height=15, out.width='100%'}
make_year_prediction_grid <- function(data, n = 100) {
  data %>%
    group_by(year) %>%
    summarise(
      xmin = min(logsize_t0, na.rm = TRUE),
      xmax = max(logsize_t0, na.rm = TRUE),
      .groups = 'drop') %>%
    tidyr::crossing(
      fire = factor(
        c('No fire', 'Fire'),
        levels = c('No fire', 'Fire'))) %>%
    rowwise() %>%
    mutate(logsize_t0 = list(seq(xmin, xmax, length.out = n))) %>%
    tidyr::unnest(logsize_t0) %>%
    ungroup() %>%
    mutate(
      year = factor(year, levels = levels(data$year)),
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3)
}

fl_pred <- make_year_prediction_grid(df_fl) %>%
  mutate(
    prediction = predict(
      mod_fl_best,
      newdata = ., type = 'response', re.form = NULL))

fl_points <- df_fl %>%
  group_by(year, fire) %>%
  mutate(bin = dplyr::ntile(logsize_t0, 8)) %>%
  group_by(year, fire, bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    response = mean(flower, na.rm = TRUE),
    .groups = 'drop')

fig_fl_years <- ggplot() +
  geom_point(
    data = fl_points,
    aes(logsize_t0, response, color = fire),
    size = 1) +
  geom_line(
    data = fl_pred,
    aes(logsize_t0, prediction, color = fire),
    linewidth = 0.8) +
  facet_wrap(~ year, ncol = 4) +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw() +
  labs(
    title = 'Flowering probability - year specific',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Flowering probability',
    color = NULL) +
  theme(legend.position = 'top')

fig_fl_years
```

```{r dist-year-site-flowerhead-plots, fig.width=12, fig.height=15, out.width='100%'}
fln_pred <- make_year_prediction_grid(df_fl_cond) %>%
  mutate(
    prediction = predict(
      mod_fl_n_best,
      newdata = ., type = 'response', re.form = NULL))

fln_points <- df_fl_cond %>%
  group_by(year, fire) %>%
  mutate(bin = dplyr::ntile(logsize_t0, 8)) %>%
  group_by(year, fire, bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    response = mean(fl_nr, na.rm = TRUE),
    .groups = 'drop')

fig_fln_years <- ggplot() +
  geom_point(
    data = fln_points,
    aes(logsize_t0, response, color = fire),
    size = 1) +
  geom_line(
    data = fln_pred,
    aes(logsize_t0, prediction, color = fire),
    linewidth = 0.8) +
  facet_wrap(~ year, ncol = 4, scales = 'free_y') +
  scale_color_manual(values = c('No fire' = 'black', 'Fire' = 'red')) +
  theme_bw() +
  labs(
    title = 'Flowerhead number given flowering - year specific',
    subtitle = v_ggp_suffix,
    x = expression('log(size)'[t0]),
    y = 'Expected flowerheads',
    color = NULL) +
  theme(legend.position = 'top')

fig_fln_years
```

## Why add site effects to recruitment?

Recruitment is measured as a site-year transition: total flowerheads in `t0`
are paired with recruits counted in `t1`. Even after accounting for flowerhead
production, sites may differ in seed availability, microsite conditions,
detection, or establishment success. The candidate model set therefore allows
site and recruit-year random effects in both the intercept and the slope.

```{r dist-year-site-recruitment-code, eval=FALSE}
mod_re_00 <- glm(
  re_t1_log ~ fh_t0_log,
  data = df_re)

mod_re_12 <- lmer(
  re_t1_log ~ fh_t0_log + (1 | site) + (1 | year_t1),
  data = df_re, REML = FALSE)

mod_re_22 <- lmer(
  re_t1_log ~ fh_t0_log +
    (fh_t0_log | site) +
    (fh_t0_log | year_t1),
  data = df_re, REML = FALSE)

mod_re_32 <- lmer(
  re_t1_log ~ fh_t0_log +
    (0 + fh_t0_log | site) +
    (0 + fh_t0_log | year_t1),
  data = df_re, REML = FALSE)
```

The full candidate set compares fixed recruitment with site effects, year
effects, and combined site-year effects. AIC determines whether the added
variation is supported rather than forcing it into the final IPM.

```{r dist-year-site-recruitment-table}
knitr::kable(
  recruitment_candidates,
  digits = 2,
  col.names = c('Candidate', 'Recruitment formula', 'Delta AIC'),
  caption = 'Flowerhead-to-recruit candidate models, ordered by support.')
```

The selected model is `r model_formula(mod_re_best)`. Its random-effect groups
are **`r recruitment_groups`**. A site random effect is
`r if (site_effect_present) 'retained' else 'not retained'`, and a recruit-year
random effect is
`r if (year_effect_present) 'retained' else 'not retained'`.

In the recruitment figure, colored lines are conditional site predictions and
the black line is the population-level fixed-effect relationship. Separation
among colored lines shows the amount of site variation supported by the model.

```{r dist-year-site-recruitment-figure, fig.width=12, fig.height=15, out.width='100%'}
g_re_years
```

The intercept is retained, so this is an effective flowerhead-to-recruit
relationship. It can absorb seed-bank carry-over, unobserved flowerheads, and
other mismatch between annual flowerhead and recruit censuses.

## Adding year and site coefficients to the IPM

Fixed effects form the baseline parameter list. Conditional coefficients from
the annual vital-rate models replace the corresponding mean coefficients for a
chosen year. Recruitment deviations are then added for the selected site and
recruit year `t1`.

```{r dist-year-site-parameter-code, eval=FALSE}
pars_i <- make_ipm_pars(
  pars_mean = pars_all_mean,
  pars_year = pars_all_year,
  pars_site = pars_all_site,
  pars_re_year_t1 = pars_all_re_year_t1,
  year = year,
  site = site,
  re_year_t1 = year + 1)
```

This ordering matters. The year supplied to the survival, growth, flowering,
and flowerhead models is `t0`, while the recruitment random effect belongs to
the year in which recruits are observed, `t1`.

## Kernel construction

The annual IPM kernel is

\[
K(y,x) = T(y,x) + F(y,x).
\]

`T` combines survival and growth. `F` multiplies flowering probability,
conditional flowerhead number, the site- and year-specific effective
recruits-per-flowerhead conversion, and the recruit-size distribution.

```{r dist-year-site-kernel-code, eval=FALSE}
# Expected flowerheads from an individual of size x
fh_x <- fl_x(x, pars, fire) * fln_x(x, pars, fire)

# Recruitment conversion after adding site and year deviations
re_per_fh <- re_total_ref(pars) / pars$fh_t0_ref
Rvec <- fh_x * re_per_fh

# Fertility and full kernel
Fmat <- outer(recruit_size_density, Rvec) * h
K <- Tmat + Fmat
```

The mesh contains `r pars_all_mean$mat_siz` size classes from
`r round(pars_all_mean$L, 2)` to `r round(pars_all_mean$U, 2)` on the log-size
scale. Eviction correction returns probability outside the mesh to its nearest
boundary class.

## Site-specific projections

For validation, each site-year starts from its observed size distribution. The
model adds the corresponding year coefficients, site recruitment deviations,
recruit-year deviations, and observed fire state before projecting one year.

```{r dist-year-site-projection-code, eval=FALSE}
use_site_specific_recruitment <- TRUE

pars_y <- make_ipm_pars(
  pars_mean = pars_all_mean,
  pars_year = pars_all_year,
  pars_site = pars_all_site,
  pars_re_year_t1 = pars_all_re_year_t1,
  year = yr,
  site = site_i,
  re_year_t1 = yr + 1)

n_proj <- kernel(pars_y, fire = fire_y)$k_yx %*% n_obs
```

The standalone script originally builds its first projection table while
`use_site_specific_recruitment` is `FALSE` and changes the switch only later.
The hidden setup chunk in this chapter therefore recomputes all site-year
projections after setting the switch to `TRUE`. Without that recomputation, the
site-specific graph would still contain population-level recruitment
coefficients.

## Observed versus modeled growth

Observed population growth is `n_t1 / n_t0`. The model produces two related
quantities:

- **projected lambda**, the one-year change from the observed site size
  distribution; and
- **asymptotic lambda**, the dominant eigenvalue of that site's annual kernel.

The comparison includes `r nrow(df_compare_site_specific)` site-year
transitions across `r validation_summary$n_years` years. After aggregating site
projections by year, the correlation with observed growth is
`r round(validation_summary$cor_proj, 2)` for projected lambda and
`r round(validation_summary$cor_asym, 2)` for asymptotic lambda. Mean absolute
errors are `r round(validation_summary$mae_proj, 2)` and
`r round(validation_summary$mae_asym, 2)`, respectively.

```{r dist-year-site-validation-figure, fig.width=9, fig.height=5, out.width='95%'}
df_plot_site_specific <- df_compare_site_specific %>%
  select(
    site, year, fire, obs_pgr,
    asym_lambda, proj_lambda) %>%
  pivot_longer(
    cols = c(asym_lambda, proj_lambda),
    names_to = 'lambda_type',
    values_to = 'lambda') %>%
  mutate(
    lambda_type = recode(
      lambda_type,
      asym_lambda = 'Asymptotic lambda',
      proj_lambda = 'Projected lambda'))

fig_site_validation <- ggplot(
  df_plot_site_specific,
  aes(x = lambda, y = obs_pgr, color = factor(year))) +
  geom_point(alpha = 0.75, size = 2) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = 'dashed') +
  facet_wrap(~ lambda_type, scales = 'free') +
  scale_color_viridis_d() +
  theme_bw() +
  labs(
    title = 'Observed versus modeled site-year population growth',
    subtitle = v_ggp_suffix,
    x = expression('Modeled ' * lambda),
    y = expression('Observed ' * lambda),
    color = 'Year')

fig_site_validation
```

Points on the dashed line indicate exact agreement. Large departures can arise
from demographic stochasticity, sampling differences, and recruitment pulses,
not only from misspecified vital-rate functions. Projected lambda is the more
direct one-year comparison because it begins with the observed size structure.

## Main conclusions

- The demographic pathway is unchanged: survival and growth determine the
  transition kernel, while flowering, flowerhead production, and recruitment
  determine fertility.
- Recruitment candidates allow variation among sites, recruit years, or both;
  AIC decides which structure is retained.
- Site effects enter only the flowerhead-to-recruit conversion. Other vital
  rates vary among years but not among sites.
- Fire remains a fixed effect on survival, growth, flowering, and conditional
  flowerhead production.
- Site-specific validation must be recomputed after enabling site recruitment
  effects; changing the switch after projection does not alter existing model
  outputs.
