# Introduction {#intro}

```{r species-image, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="*Eriogonum longifolium var. gnaphalifolium. Image from Wikipedia.*"}

wiki_page <- "Eriogonum longifolium var. gnaphalifolium"

wiki_api <- paste0(
  "https://en.wikipedia.org/w/api.php?",
  "action=query&format=json&prop=pageimages&",
  "piprop=thumbnail&pithumbsize=1400&titles=",
  URLencode(wiki_page, reserved = TRUE))

wiki_data <- jsonlite::fromJSON(
  wiki_api,
  simplifyVector = FALSE)

img_url <- wiki_data$query$pages[[1]]$thumbnail$source

dir.create("images", showWarnings = FALSE)

img_ext <- tools::file_ext(
  sub("\\?.*$", "", img_url))

img_file <- file.path(
  "images",
  paste0("eriogonum_wikipedia.", img_ext))

download.file(
  img_url,
  destfile = img_file,
  mode = "wb",
  quiet = TRUE)

knitr::include_graphics(img_file)
```

The supplementary material to the *Eriogonum longifolium* demographic dataset
from Archbold Biological Station.

The GIS and Data Management program at
[Archbold Biological Station](https://www.archbold-station.org/gis-and-data-mgmt/)
maintains a comprehensive collection of ecological and environmental data
gathered through research initiatives and long-term monitoring projects. One
key platform for accessing these data is the
[Environmental Data Initiative (EDI)](https://portal.edirepository.org/nis/home.jsp),
which hosts long-term ecological datasets from Archbold Biological Station.

For our model, we utilized 24 years (1990--2013) of demographic data on scrub
buckwheat, [*Eriogonum longifolium* var. *gnaphalifolium*](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.226.1),
a threatened perennial herb endemic to periodically burned xeric uplands in
central peninsular Florida. The long-term dataset includes observations of
individual survival, growth, flowering, dormancy, recruitment, and responses to
prescribed fire, allowing us to incorporate both continuous plant size and
fire history into an integral projection model (IPM). :contentReference[oaicite:0]{index=0}

These demographic data have previously been used by Satterthwaite, Menges, and
Quintana-Ascencio (2002) in *Assessing scrub buckwheat population viability in
relation to fire using multiple modeling techniques*. Their study used
stage-structured matrix models and population viability analyses to evaluate
the effects of fire and fire-return intervals on the population dynamics of
*E. longifolium* var. *gnaphalifolium*. In the present analysis, we revisit the
long-term demographic record using an IPM framework to describe individual
demographic rates and their relationship with fire history.


## Setup

We use the following packages: `tidyverse`, `patchwork`, `skimr`, `ipmr`, `binom`, `bbmle`, `gridExtra`, `lme4`

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Packages
packages <- c('MASS', 'patchwork', 'skimr', 'ipmr', 'binom', 'bbmle', 'gridExtra', 
              'lme4', 'janitor', 'readxl', 'GGally', 'tidyverse')
lapply(packages, library, character.only = TRUE)


# Specification 
# Define head-directory
v_head <- c('archbold')
# Define species
v_species <- c('Eriogonum longifolium')
# Customized delimiter for `read_delim` function, comma is predefined
custom_delimiter <- c()


# Create a unique species abbreviation for file naming
v_sp_abb <- tolower(
  gsub(' ', '', paste(
    substr(unlist(strsplit(v_species, ' ')), 1, 2), collapse = '')))

# Define script prefix
v_script_prefix <- str_c(v_head)

# Plot subtitle
v_ggp_suffix <- paste(
  tools::toTitleCase(v_head), '-', v_species)


# Directory 
dir_wd     <- file.path('C:', 'code', 'RUPDemo_IPMs')
dir_pub    <- file.path(dir_wd,  paste0(v_head))
dir_R      <- file.path(dir_pub, 'R',       v_sp_abb)
dir_data   <- file.path(dir_pub, 'data',    v_sp_abb)
dir_result <- file.path(dir_pub, 'results', v_sp_abb)
dir_func   <- file.path(dir_wd,  'helper_functions')


# Functions
# plot your survival data 'binned' (instead of 'jittered')
source(file.path(dir_func, 'plot_binned_prop.R'))
# 
source(file.path(dir_func, 'line_color_pred_fun.R'))
# 
source(file.path(dir_func, 'predictor_fun.R'))
```

