# Environmental mean IPM {#ipm_mean_env}
## Introduction

This analysis develops a mean integral projection model (IPM) for
*Eriogonum longifolium* using annual demographic observations from Archbold
Biological Station from 1990 to 2013. The model describes transitions among
active plants distributed continuously by size and a discrete dormant state.
Demographic processes include survival, growth, entry into dormancy,
reactivation from dormancy, flowering, flowering-scape production, and
recruitment.

The main objective is to incorporate fire while retaining its temporal position
relative to the annual demographic transition. Disturbance during the focal
transition from year $t$ to $t+1$ is termed **current disturbance**, whereas
disturbance during the preceding transition from $t-1$ to $t$ is termed
**previous disturbance**. These two disturbance variables are not assumed to
affect all vital rates in the same way. Instead, each vital rate is evaluated
using the disturbance timing that is biologically associated with that process.

Survival, growth, dormancy entry, and reactivation are evaluated with current
disturbance because they describe the fate of individuals during the transition
from $t$ to $t+1$. Flowering probability and the number of flowering scapes are
evaluated with previous disturbance because reproductive condition at census
$t$ may reflect disturbance experienced during the preceding annual interval.
Recruitment links flowering scapes at $t$ to recruits observed at $t+1$ and is
therefore evaluated with current disturbance.

For each size-dependent vital rate, alternative models ranging from an
intercept-only model through cubic size dependence are compared using AICc.
For each polynomial degree, models with and without the corresponding
disturbance effect are compared on the same set of observations. The
best-supported models are then combined into a single mean parameter set for
the IPM.

Because current and previous disturbance can occur in four combinations, four
demographic kernels are constructed:

1. no current or previous disturbance,
2. current disturbance only,
3. previous disturbance only, and
4. disturbance in both intervals.

These kernels describe fixed disturbance scenarios, but the historical fire
regime is represented separately. Observed transitions between previous and
current disturbance states are used to estimate
$P(D_t \mid D_{t-1})$. The four demographic kernels are then combined with
these transition probabilities in an environmental-history megamatrix. This
allows population structure and disturbance history to evolve jointly rather
than averaging the four disturbance-specific population growth rates.

Population growth is evaluated in two complementary ways. Asymptotic
population growth, $\lambda$, is calculated from the dominant eigenvalue of
each projection matrix. One-step projected growth is additionally calculated
by applying each kernel to the mean observed population structure. The
environmental-history projection begins with the observed population divided
between the two previous-disturbance states according to their observed
frequencies. These model-based growth rates are finally compared with observed
annual changes in population abundance.

## Specifications
### Vital model leaverage
```{r}
v_mod_set_su   <- c()
v_mod_set_gr   <- c()
v_mod_set_do   <- c()
v_mod_set_fl   <- c()
v_mod_set_fl_n <- c()
```

## Data
```{r}
disturbance_case_levels <- c(
  "Undisturbed",
  "Current-year fire",
  "Previous-year fire",
  "Fire in both years")

disturbance_case_cols <- c(
  "Undisturbed" = "black",
  "Current-year fire" = "red",
  "Previous-year fire" = "purple",
  "Fire in both years" = "magenta")

df <- read.csv(
  file.path(
    dir_data,
    paste0("ab_", v_sp_abb, "_df_workdata_260820.csv"))) %>%
  mutate(
    disturbance = as.numeric(dist_transition),
    disturbance_prev = as.numeric(disturbance_prev),
    year = as.numeric(year),
    row_type = as.character(row_type),
    disturbance_case = factor(
      paste0(disturbance, disturbance_prev),
      levels = c("00", "10", "01", "11"),
      labels = disturbance_case_levels))

df_ind <- df %>%
  filter(row_type == "individual")

df_re_all <- df %>%
  filter(row_type == "recruitment")

df_re <- df_re_all %>%
  filter(
    recruitment_complete,
    !is.na(recruits_simple))
```

## Vital rates
### Survival
#### Data
```{r}
df_su <- df %>%
  filter(
    state == "active",
    !is.na(survives),
    size_t0 > 0,
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, year, size_t0, survives, disturbance,
    logsize_t0, logsize_t0_2, logsize_t0_3)
```

#### Model
```{r}
mod_su_00 <- glm(
  survives ~ 1,
  data = df_su, family = binomial)

mod_su_0 <- glm(
  survives ~ disturbance,
  data = df_su, family = binomial)

mod_su_10 <- glm(
  survives ~ logsize_t0,
  data = df_su, family = binomial)

mod_su_1 <- glm(
  survives ~ logsize_t0 + disturbance,
  data = df_su, family = binomial)

mod_su_20 <- glm(
  survives ~ logsize_t0 + logsize_t0_2,
  data = df_su, family = binomial)

mod_su_2 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_su, family = binomial)

mod_su_30 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_su, family = binomial)

mod_su_3 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_su, family = binomial)

mods_su <- list(
  mod_su_00, mod_su_0, mod_su_10, mod_su_1,
  mod_su_20, mod_su_2, mod_su_30, mod_su_3)

mods_su_dAICc <- AICctab(
  mods_su, weights = TRUE, sort = FALSE)$dAICc

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- which.min(mods_su_dAICc)
} else {
  keep <- 2 * v_mod_set_su + c(1, 2)
  mod_su_index_bestfit <- keep[which.min(mods_su_dAICc[keep])]
}

v_mod_su_index <- floor((mod_su_index_bestfit - 1) / 2)
mod_su_bestfit <- mods_su[[mod_su_index_bestfit]]

mod_su_bestfit
mods_su_dAICc
```

#### Plot
```{r}
df_su_newdata <- df_su %>%
  summarise(
    x = list(seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)),
    .groups = "drop") %>%
  unnest(x) %>%
  mutate(
    logsize_t0 = x,
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    predicted = predict(
      mod_su_bestfit,
      newdata = .,
      type = "response"))

fig_su_line <- ggplot(
  df_su,
  aes(x = logsize_t0, y = survives)) +
  geom_jitter(
    height = 0.05, width = 0,
    alpha = 0.3) +
  geom_line(
    data = df_su_newdata,
    aes(y = predicted),
    linewidth = 1) +
  labs(
    title = "Survival probability by size",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = "Probability of survival") +
  theme_bw()


df_su_bindata <- plot_binned_prop(
  df_su, 10, logsize_t0, survives)

df_su_pred <- df_su %>%
  reframe(
    logsize_t0 = seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    survives = predict(
      mod_su_bestfit,
      newdata = .,
      type = "response"))

fig_su_bin <- ggplot() +
  geom_point(
    data = df_su_bindata,
    aes(x = logsize_t0, y = survives)) +
  geom_errorbar(
    data = df_su_bindata,
    aes(
      x = logsize_t0,
      ymin = lwr,
      ymax = upr),
    width = 0.1) +
  geom_line(
    data = df_su_pred,
    aes(x = logsize_t0, y = survives),
    linewidth = 1.2) +
  labs(
    x = expression("log(diameter)"[t0]),
    y = "") +
  theme_bw() +
  ylim(0, 1)

fig_su <- fig_su_line + fig_su_bin + plot_layout()
fig_su
```

### Growth 
#### Data
```{r}
df_gr <- df %>%
  filter(
    state == "active",
    state_t1 == "active",
    size_t0 > 0,
    size_t1 > 0,
    is.finite(logsize_t0),
    is.finite(logsize_t1),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, year, size_t0, size_t1,
    logsize_t0, logsize_t1,
    logsize_t0_2, logsize_t0_3, disturbance)
```

#### Model
```{r}
mod_gr_00 <- lm(
  logsize_t1 ~ 1,
  data = df_gr)

mod_gr_0 <- lm(
  logsize_t1 ~ disturbance,
  data = df_gr)

mod_gr_10 <- lm(
  logsize_t1 ~ logsize_t0,
  data = df_gr)

mod_gr_1 <- lm(
  logsize_t1 ~ logsize_t0 + disturbance,
  data = df_gr)

mod_gr_20 <- lm(
  logsize_t1 ~ logsize_t0 + logsize_t0_2,
  data = df_gr)

mod_gr_2 <- lm(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_gr)

mod_gr_30 <- lm(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_gr)

mod_gr_3 <- lm(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_gr)

mods_gr <- list(
  mod_gr_00, mod_gr_0, mod_gr_10, mod_gr_1,
  mod_gr_20, mod_gr_2, mod_gr_30, mod_gr_3)

mods_gr_dAICc <- AICctab(
  mods_gr, weights = TRUE, sort = FALSE)$dAICc

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- which.min(mods_gr_dAICc)
} else {
  keep <- 2 * v_mod_set_gr + c(1, 2)
  mod_gr_index_bestfit <- keep[which.min(mods_gr_dAICc[keep])]
}

v_mod_gr_index <- floor((mod_gr_index_bestfit - 1) / 2)
mod_gr_bestfit <- mods_gr[[mod_gr_index_bestfit]]

mod_gr_bestfit
mods_gr_dAICc
```

#### Plot
```{r}
df_gr_newdata <- expand_grid(
  logsize_t0 = seq(
    min(df_gr$logsize_t0),
    max(df_gr$logsize_t0),
    length.out = 100),
  disturbance = c(0, 1)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)

df_gr_newdata <- cbind(
  df_gr_newdata,
  as.data.frame(
    predict(
      mod_gr_bestfit,
      newdata = df_gr_newdata,
      interval = "confidence")))

fig_gr <- ggplot(
  df_gr,
  aes(
    x = logsize_t0,
    y = logsize_t1,
    color = factor(disturbance))) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = df_gr_newdata,
    aes(y = fit),
    linewidth = 1) +
  geom_abline(intercept = 0, slope = 1) +
  scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("0" = "No disturbance", "1" = "Disturbance"),
    name = NULL) +
  labs(
    title = "Growth prediction",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = expression("log(diameter)"[t1])) +
  theme_bw()

fig_gr
```

#### Variance
```{r}
# Fitted values from growth model
mod_gr_x   <- fitted(mod_gr_bestfit)  
# Squared residuals
mod_gr_y   <- resid(mod_gr_bestfit)^2  
# Non-linear model for variance
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x), start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE)) 
```

### Dormancy entry 
#### Data
```{r}
df_do <- df %>%
  filter(
    !is.na(enter_dormancy),
    size_t0 > 0,
    is.finite(logsize_t0),
    !is.na(disturbance)) %>%
  dplyr::select(
    id, year, size_t0, enter_dormancy, disturbance,
    logsize_t0, logsize_t0_2, logsize_t0_3)
```

#### Model
```{r}
mod_do_00 <- glm(
  enter_dormancy ~ 1,
  data = df_do, family = binomial)

mod_do_0 <- glm(
  enter_dormancy ~ disturbance,
  data = df_do, family = binomial)

mod_do_10 <- glm(
  enter_dormancy ~ logsize_t0,
  data = df_do, family = binomial)

mod_do_1 <- glm(
  enter_dormancy ~ logsize_t0 + disturbance,
  data = df_do, family = binomial)

mod_do_20 <- glm(
  enter_dormancy ~ logsize_t0 + logsize_t0_2,
  data = df_do, family = binomial)

mod_do_2 <- glm(
  enter_dormancy ~ logsize_t0 + logsize_t0_2 + disturbance,
  data = df_do, family = binomial)

mod_do_30 <- glm(
  enter_dormancy ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_do, family = binomial)

mod_do_3 <- glm(
  enter_dormancy ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 + disturbance,
  data = df_do, family = binomial)

mods_do <- list(
  mod_do_00, mod_do_0, mod_do_10, mod_do_1,
  mod_do_20, mod_do_2, mod_do_30, mod_do_3)

mods_do_dAICc <- AICctab(
  mods_do, weights = TRUE, sort = FALSE)$dAICc

if (length(v_mod_set_do) == 0) {
  mod_do_index_bestfit <- which.min(mods_do_dAICc)
} else {
  keep <- 2 * v_mod_set_do + c(1, 2)
  mod_do_index_bestfit <- keep[which.min(mods_do_dAICc[keep])]
}

v_mod_do_index <- floor((mod_do_index_bestfit - 1) / 2)
mod_do_bestfit <- mods_do[[mod_do_index_bestfit]]

mod_do_bestfit
mods_do_dAICc
```

#### Dormancy entry plot
```{r}
df_do_newdata <- df_do %>%
  summarise(
    x = list(seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)),
    .groups = "drop") %>%
  unnest(x) %>%
  mutate(
    logsize_t0 = x,
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    predicted = predict(
      mod_do_bestfit,
      newdata = .,
      type = "response"))

fig_do_line <- ggplot(
  df_do,
  aes(x = logsize_t0, y = enter_dormancy)) +
  geom_jitter(
    height = 0.05, width = 0,
    alpha = 0.3) +
  geom_line(
    data = df_do_newdata,
    aes(y = predicted),
    linewidth = 1) +
  labs(
    title = "Dormancy probability by size",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = "Probability of entering dormancy") +
  theme_bw()


df_do_bindata <- plot_binned_prop(
  df_do, 10, logsize_t0, enter_dormancy)

df_do_pred <- df_do %>%
  reframe(
    logsize_t0 = seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    enter_dormancy = predict(
      mod_do_bestfit,
      newdata = .,
      type = "response"))

fig_do_bin <- ggplot() +
  geom_point(
    data = df_do_bindata,
    aes(x = logsize_t0, y = enter_dormancy)) +
  geom_errorbar(
    data = df_do_bindata,
    aes(
      x = logsize_t0,
      ymin = lwr,
      ymax = upr),
    width = 0.1) +
  geom_line(
    data = df_do_pred,
    aes(x = logsize_t0, y = enter_dormancy),
    linewidth = 1.2) +
  labs(
    x = expression("log(diameter)"[t0]),
    y = "") +
  theme_bw() +
  ylim(0, 1)

fig_do <- fig_do_line + fig_do_bin + plot_layout()
fig_do
```

### Reactivation
#### Data
```{r}
df_ra <- df %>%
  filter(!is.na(reactivate),
         !is.na(disturbance))
```

#### Model
```{r}
mod_ra_0 <- glm(
  reactivate ~ 1,
  data = df_ra, family = binomial)

mod_ra_1 <- glm(
  reactivate ~ disturbance,
  data = df_ra, family = binomial)

mods_ra <- list(
  mod_ra_0, mod_ra_1)

mods_ra_dAICc <- AICctab(
  mods_ra, weights = TRUE, sort = FALSE)$dAICc

mod_ra_index_bestfit <- which.min(mods_ra_dAICc)
mod_ra_bestfit <- mods_ra[[mod_ra_index_bestfit]]

mod_ra_bestfit
mods_ra_dAICc

p_reactivate_0 <- predict(
  mod_ra_bestfit,
  newdata = data.frame(disturbance = 0),
  type = "response")[1]

p_reactivate_1 <- predict(
  mod_ra_bestfit,
  newdata = data.frame(disturbance = 1),
  type = "response")[1]

p_reactivate_0
p_reactivate_1

df_ra_size <- df %>%
  filter(
    reactivate == 1,
    size_reactivate_t1 > 0) %>%
  mutate(
    logsize_reactivate =
      log(size_reactivate_t1))

react_sz <- mean(
  df_ra_size$logsize_reactivate,
  na.rm = TRUE)

react_sd <- sd(
  df_ra_size$logsize_reactivate,
  na.rm = TRUE)
```

#### Plot
```{r}
df_ra_plot <- df_ra %>%
  group_by(disturbance) %>%
  summarise(
    n = n(),
    nr_reactivate = sum(reactivate == 1),
    reactivation = mean(reactivate), .groups = "drop") %>%
  mutate(
    lwr = binom.confint(
      nr_reactivate, n, methods = "wilson")$lower,
    upr = binom.confint(
      nr_reactivate, n, methods = "wilson")$upper,
    disturbance_lab = factor(
      disturbance,
      levels = c(0, 1),
      labels = c("No disturbance", "Disturbance")),
    pred = c(p_reactivate_0, p_reactivate_1))

fig_ra <- ggplot(
  df_ra_plot,
  aes(
    x = disturbance_lab,
    y = reactivation,
    color = disturbance_lab)) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(ymin = lwr, ymax = upr),
    width = 0.1) +
  geom_point(
    aes(y = pred),
    shape = 95,
    size = 10) +
  scale_color_manual(
    values = c("No disturbance" = "black",
               "Disturbance" = "red"),
    guide = "none") +
  scale_y_continuous(
    limits = c(0, 1)) +
  labs(
    title = "Reactivation probability",
    subtitle = v_ggp_suffix,
    x = "",
    y = "Probability of reactivation") +
  theme_bw()

fig_ra
```

### Flowering probability
#### Data
```{r}
df_fl <- df %>%
  filter(
    state == "active",
    !is.na(flower),
    size_t0 > 0,
    is.finite(logsize_t0),
    !is.na(disturbance_prev))
```

#### Model
```{r}
mod_fl_00 <- glm(
  flower ~ 1,
  data = df_fl, family = binomial)

mod_fl_0 <- glm(
  flower ~ disturbance_prev,
  data = df_fl, family = binomial)

mod_fl_10 <- glm(
  flower ~ logsize_t0,
  data = df_fl, family = binomial)

mod_fl_1 <- glm(
  flower ~ logsize_t0 + disturbance_prev,
  data = df_fl, family = binomial)

mod_fl_20 <- glm(
  flower ~ logsize_t0 + logsize_t0_2,
  data = df_fl, family = binomial)

mod_fl_2 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + disturbance_prev,
  data = df_fl, family = binomial)

mod_fl_30 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fl, family = binomial)

mod_fl_3 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 +
    disturbance_prev,
  data = df_fl, family = binomial)

mods_fl <- list(
  mod_fl_00, mod_fl_0, mod_fl_10, mod_fl_1,
  mod_fl_20, mod_fl_2, mod_fl_30, mod_fl_3)

mods_fl_dAICc <- AICctab(
  mods_fl, weights = TRUE, sort = FALSE)$dAICc

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- which.min(mods_fl_dAICc)
} else {
  keep <- 2 * v_mod_set_fl + c(1, 2)
  mod_fl_index_bestfit <- keep[which.min(mods_fl_dAICc[keep])]
}

v_mod_fl_index <- floor((mod_fl_index_bestfit - 1) / 2)
mod_fl_bestfit <- mods_fl[[mod_fl_index_bestfit]]

mod_fl_bestfit
mods_fl_dAICc
```

#### Plot
```{r}
df_fl_newdata <- expand_grid(
  logsize_t0 = seq(
    min(df_fl$logsize_t0),
    max(df_fl$logsize_t0),
    length.out = 100),
  disturbance_prev = c(0, 1)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    predicted = predict(
      mod_fl_bestfit,
      newdata = .,
      type = "response"),
    disturbance_plot = factor(
      disturbance_prev,
      levels = c(0, 1),
      labels = c("No disturbance", "Disturbance")))

fig_fl <- ggplot(
  df_fl,
  aes(x = logsize_t0, y = flower)) +
  geom_jitter(
    aes(
      color = factor(
        disturbance_prev,
        levels = c(0, 1),
        labels = c("No disturbance", "Disturbance"))),
    height = 0.05, width = 0,
    alpha = 0.15) +
  geom_line(
    data = df_fl_newdata,
    aes(y = predicted, color = disturbance_plot),
    linewidth = 1) +
  scale_color_manual(
    values = c(
      "No disturbance" = "black",
      "Disturbance" = "red"),
    name = NULL) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Flowering probability by size",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = "Probability of flowering") +
  theme_bw()

fig_fl
```

### Number of scapes conditional on flowering 
#### Data
```{r}
df_fl_cond <- df_fl %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr %% 1 == 0,
    !is.na(disturbance_prev))
```

#### Model 
```{r}
mod_fl_n_00 <- glm.nb(
  fl_nr ~ 1,
  data = df_fl_cond)

mod_fl_n_0 <- glm.nb(
  fl_nr ~ disturbance_prev,
  data = df_fl_cond)

mod_fl_n_10 <- glm.nb(
  fl_nr ~ logsize_t0,
  data = df_fl_cond)

mod_fl_n_1 <- glm.nb(
  fl_nr ~ logsize_t0 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_20 <- glm.nb(
  fl_nr ~ logsize_t0 + logsize_t0_2,
  data = df_fl_cond)

mod_fl_n_2 <- glm.nb(
  fl_nr ~ logsize_t0 + logsize_t0_2 + disturbance_prev,
  data = df_fl_cond)

mod_fl_n_30 <- glm.nb(
  fl_nr ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fl_cond)

mod_fl_n_3 <- glm.nb(
  fl_nr ~ logsize_t0 + logsize_t0_2 + logsize_t0_3 +
    disturbance_prev,
  data = df_fl_cond)

mods_fl_n <- list(
  mod_fl_n_00, mod_fl_n_0, mod_fl_n_10, mod_fl_n_1,
  mod_fl_n_20, mod_fl_n_2, mod_fl_n_30, mod_fl_n_3)

mods_fl_n_dAICc <- AICctab(
  mods_fl_n, weights = TRUE, sort = FALSE)$dAICc

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_bestfit <- which.min(mods_fl_n_dAICc)
} else {
  keep <- 2 * v_mod_set_fl_n + c(1, 2)
  mod_fl_n_index_bestfit <- keep[
    which.min(mods_fl_n_dAICc[keep])]
}

v_mod_fl_n_index <- floor((mod_fl_n_index_bestfit - 1) / 2)
mod_fl_n_bestfit <- mods_fl_n[[mod_fl_n_index_bestfit]]

mod_fl_n_bestfit
mods_fl_n_dAICc
```

#### Plot
```{r}
# Predictions for flower number[]
# Create prediction grid
df_fl_n_pred <- expand.grid(
  logsize_t0 = seq(
    min(df_fl_cond$logsize_t0),
    max(df_fl_cond$logsize_t0),
    length.out = 100))

df_fl_n_pred <- df_fl_n_pred %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)

# Predict
df_fl_n_pred$fl_nr <- predict(
  mod_fl_n_bestfit,
  newdata = df_fl_n_pred,
  type = "response")


# Binned observed data
df_fl_n_binned <- df_fl_cond %>%
  mutate(bin = cut(logsize_t0, breaks = 10)) %>%
  group_by(bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    fl_nr = mean(fl_nr, na.rm = TRUE),
    se = sd(fl_nr, na.rm = TRUE) / sqrt(n()),
    .groups = "drop") %>%
  mutate(
    lwr = fl_nr - 1.96 * se,
    upr = fl_nr + 1.96 * se)


# Flower number plots
# Plot 1: Raw jitter + prediction
fig_fl_n_line <- ggplot() +
  geom_jitter(
    data = df_fl_cond,
    aes(x = logsize_t0, y = fl_nr),
    alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(
    data = df_fl_n_pred,
    aes(x = logsize_t0, y = fl_nr),
    linewidth = 0.9) +
  theme_bw() +
  labs(
    title = NULL,
    x = expression("log(diameter)"[t0]),
    y = "Number of flowering scapes")


# Plot 2: Binned + prediction
fig_fl_n_bin <- ggplot() +
  geom_point(
    data = df_fl_n_binned,
    aes(x = logsize_t0, y = fl_nr)) +
  geom_errorbar(
    data = df_fl_n_binned,
    aes(x = logsize_t0, ymin = lwr, ymax = upr),
    width = 0.2) +
  geom_line(
    data = df_fl_n_pred,
    aes(x = logsize_t0, y = fl_nr),
    linewidth = 0.9) +
  theme_bw() +
  labs(
    title = NULL,
    x = expression("log(diameter)"[t0]),
    y = "Number of flowering scapes")


# Combine
fig_fl_n <- fig_fl_n_line + fig_fl_n_bin +
  plot_annotation(
    title = "Flower number",
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10, face = "italic")))

fig_fl_n
```

### Recruitment 
#### Data
```{r}
df_sc2re <- df_re %>%
  filter(
    !is.na(nr_scapes),
    !is.na(recruits_simple)) %>%
  group_by(site, pop, year) %>%
  summarise(
    total_scapes = sum(nr_scapes),
    recruit_count = sum(recruits_simple),
    disturbance = case_when(
      any(disturbance == 1, na.rm = TRUE) ~ 1,
      any(is.na(disturbance)) ~ NA_real_,
      TRUE ~ 0),
    disturbance_prev = case_when(
      any(disturbance_prev == 1, na.rm = TRUE) ~ 1,
      any(is.na(disturbance_prev)) ~ NA_real_,
      TRUE ~ 0),
    .groups = "drop")

df_sc2re_mod <- df_sc2re %>%
  filter(
    !is.na(total_scapes),
    !is.na(recruit_count),
    !is.na(disturbance))
```

#### Model
```{r}
mod_re_00 <- glm.nb(
  recruit_count ~ 1,
  data = df_sc2re_mod)

mod_re_0 <- glm.nb(
  recruit_count ~ disturbance,
  data = df_sc2re_mod)

mod_re_sc0 <- glm.nb(
  recruit_count ~ total_scapes,
  data = df_sc2re_mod)

mod_re_sc1 <- glm.nb(
  recruit_count ~ total_scapes + disturbance,
  data = df_sc2re_mod)

mod_re_sc2 <- glm.nb(
  recruit_count ~ total_scapes * disturbance,
  data = df_sc2re_mod)

mod_re_logsc0 <- glm.nb(
  recruit_count ~ log1p(total_scapes),
  data = df_sc2re_mod)

mod_re_logsc1 <- glm.nb(
  recruit_count ~ log1p(total_scapes) + disturbance,
  data = df_sc2re_mod)

mod_re_logsc2 <- glm.nb(
  recruit_count ~ log1p(total_scapes) * disturbance,
  data = df_sc2re_mod)

mods_re <- list(
  mod_re_00, mod_re_0,
  mod_re_sc0, mod_re_sc1, mod_re_sc2,
  mod_re_logsc0, mod_re_logsc1, mod_re_logsc2)

mods_re_dAICc <- AICctab(
  mods_re, weights = TRUE, sort = FALSE)$dAICc

mods_re_sorted <- order(mods_re_dAICc)
mod_re_index_bestfit <- mods_re_sorted[1]
mod_re_bestfit <- mods_re[[mod_re_index_bestfit]]

mod_re_bestfit
mods_re_dAICc
```

#### Mean recruitment conversion for the IPM
```{r}
# Predict recruitment for the same observed scape distribution under
# no disturbance and disturbance.

df_sc2re_pred_0 <- df_sc2re_mod %>%
  mutate(
    disturbance = 0)

df_sc2re_pred_0 <- df_sc2re_pred_0 %>%
  mutate(
    recruits_pred = predict(
      mod_re_bestfit,
      newdata = df_sc2re_pred_0,
      type = "response"))


df_sc2re_pred_1 <- df_sc2re_mod %>%
  mutate(
    disturbance = 1)

df_sc2re_pred_1 <- df_sc2re_pred_1 %>%
  mutate(
    recruits_pred = predict(
      mod_re_bestfit,
      newdata = df_sc2re_pred_1,
      type = "response"))


recr_per_scape_0 <- sum(df_sc2re_pred_0$recruits_pred) /
  sum(df_sc2re_pred_0$total_scapes)

recr_per_scape_1 <- sum(df_sc2re_pred_1$recruits_pred) /
  sum(df_sc2re_pred_1$total_scapes)

recr_per_scape_0
recr_per_scape_1

recr_per_scape_0
recr_per_scape_1
```

#### Plot
```{r}
df_sc2re_pred <- expand_grid(
  total_scapes = seq(
    min(df_sc2re_mod$total_scapes),
    max(df_sc2re_mod$total_scapes),
    length.out = 100),
  disturbance = c(0, 1))

mod_sc2re_preds <- predict(
  mod_re_bestfit,
  newdata = df_sc2re_pred,
  type = "link",
  se.fit = TRUE)

df_sc2re_pred <- df_sc2re_pred %>%
  mutate(
    fit = exp(mod_sc2re_preds$fit),
    lower = exp(
      mod_sc2re_preds$fit -
        1.96 * mod_sc2re_preds$se.fit),
    upper = exp(
      mod_sc2re_preds$fit +
        1.96 * mod_sc2re_preds$se.fit),
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c("No disturbance", "Disturbance")))

df_sc2re_plot <- df_sc2re_mod %>%
  mutate(
    disturbance_plot = factor(
      disturbance,
      levels = c(0, 1),
      labels = c("No disturbance", "Disturbance")))

fig_sc2re <- ggplot(
  df_sc2re_plot,
  aes(
    x = total_scapes,
    y = recruit_count,
    color = disturbance_plot)) +
  geom_jitter(
    height = 0.2, width = 0.5,
    alpha = 0.4) +
  geom_line(
    data = df_sc2re_pred,
    aes(
      y = fit,
      color = disturbance_plot),
    linewidth = 1.2) +
  scale_color_manual(
    values = c(
      "No disturbance" = "black",
      "Disturbance" = "red"),
    name = NULL) +
  labs(
    title = "Recruits t1 by flowering scapes t0",
    subtitle = v_ggp_suffix,
    x = expression("Total flowering scapes "[t0]),
    y = expression("Number of recruits "[t1])) +
  theme_bw()

fig_sc2re
```

#### Recruit size
```{r}
df_re_size <- df_ind %>%
  filter(
    recruit_type == "seedling",
    size_t0 > 0,
    is.finite(logsize_t0))

recr_sz <- mean(
  df_re_size$logsize_t0,
  na.rm = TRUE)

recr_sd <- sd(
  df_re_size$logsize_t0,
  na.rm = TRUE)
```

## Parameters
### Extracting parameter estimates
```{r}
# Survival
coef_su <- data.frame(
  coefficient = names(coef(mod_su_bestfit)),
  value = coef(mod_su_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Growth
coef_gr_fe <- data.frame(
  coefficient = names(coef(mod_gr_bestfit)),
  value = coef(mod_gr_bestfit))

coef_gr_var <- data.frame(
  coefficient = names(coef(mod_gr_var)),
  value = coef(mod_gr_var))

coef_gr <- bind_rows(coef_gr_fe, coef_gr_var) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Dormancy entry
coef_do <- data.frame(
  coefficient = names(coef(mod_do_bestfit)),
  value = coef(mod_do_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Reactivation
coef_ra <- data.frame(
  coefficient = names(coef(mod_ra_bestfit)),
  value = coef(mod_ra_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Flowering probability
coef_fl <- data.frame(
  coefficient = names(coef(mod_fl_bestfit)),
  value = coef(mod_fl_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Scape number conditional on flowering
coef_fln <- data.frame(
  coefficient = names(coef(mod_fl_n_bestfit)),
  value = coef(mod_fl_n_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

extr_value <- function(x, field) {
  subset(x, coefficient == field)$value
}
```

### IPM constants
```{r}
mesh_limits <- range(
  c(
    df_gr$logsize_t0,
    df_gr$logsize_t1,
    df_re_size$logsize_t0,
    df_ra_size$logsize_reactivate),
  na.rm = TRUE,
  finite = TRUE)

coef_misc <- data.frame(
  coefficient = c(
    "recr_sz",
    "recr_sd",
    "react_sz",
    "react_sd",
    "recr_0",
    "recr_1",
    "max_siz",
    "min_siz"),
  value = c(
    recr_sz,
    recr_sd,
    react_sz,
    react_sd,
    recr_per_scape_0,
    recr_per_scape_1,
    mesh_limits[2],
    mesh_limits[1]))
```

### Parameter list
```{r}
pars <- Filter(function(x) length(x) > 0, list(
  prefix = v_script_prefix,
  species = v_species,
  
  surv_b0 = extr_value(coef_su, "b0"),
  surv_b1 = extr_value(coef_su, "logsize_t0"),
  surv_b2 = extr_value(coef_su, "logsize_t0_2"),
  surv_b3 = extr_value(coef_su, "logsize_t0_3"),
  surv_bd = extr_value(coef_su, "disturbance"),
  
  grow_b0 = extr_value(coef_gr, "b0"),
  grow_b1 = extr_value(coef_gr, "logsize_t0"),
  grow_b2 = extr_value(coef_gr, "logsize_t0_2"),
  grow_b3 = extr_value(coef_gr, "logsize_t0_3"),
  grow_bd = extr_value(coef_gr, "disturbance"),
  a = extr_value(coef_gr, "a"),
  b = extr_value(coef_gr, "b"),
  
  dorm_b0 = extr_value(coef_do, "b0"),
  dorm_b1 = extr_value(coef_do, "logsize_t0"),
  dorm_b2 = extr_value(coef_do, "logsize_t0_2"),
  dorm_b3 = extr_value(coef_do, "logsize_t0_3"),
  dorm_bd = extr_value(coef_do, "disturbance"),
  
  react_b0 = extr_value(coef_ra, "b0"),
  react_bd = extr_value(coef_ra, "disturbance"),
  
  fl_b0 = extr_value(coef_fl, "b0"),
  fl_b1 = extr_value(coef_fl, "logsize_t0"),
  fl_b2 = extr_value(coef_fl, "logsize_t0_2"),
  fl_b3 = extr_value(coef_fl, "logsize_t0_3"),
  fl_bd = extr_value(coef_fl, "disturbance_prev"),
  
  fln_b0 = extr_value(coef_fln, "b0"),
  fln_b1 = extr_value(coef_fln, "logsize_t0"),
  fln_b2 = extr_value(coef_fln, "logsize_t0_2"),
  fln_b3 = extr_value(coef_fln, "logsize_t0_3"),
  fln_bd = extr_value(coef_fln, "disturbance_prev"),
  
  recr_0 = extr_value(coef_misc, "recr_0"),
  recr_1 = extr_value(coef_misc, "recr_1"),
  recr_sz = extr_value(coef_misc, "recr_sz"),
  recr_sd = extr_value(coef_misc, "recr_sd"),
  react_sz = extr_value(coef_misc, "react_sz"),
  react_sd = extr_value(coef_misc, "react_sd"),
  
  L = extr_value(coef_misc, "min_siz"),
  U = extr_value(coef_misc, "max_siz"),
  mat_siz = 200,
  
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_do_index = v_mod_do_index,
  mod_fl_index = v_mod_fl_index,
  mod_fl_n_index = v_mod_fl_n_index))
```

## Building the IPM
The demographic IPM consists of 200 continuous active-size classes and one
discrete dormant state. The resulting demographic kernel therefore has
dimensions $201 \times 201$.

For active individuals, survival is divided between remaining active and
entering dormancy. Individuals that remain active transition among size classes
according to the fitted growth distribution. Dormant individuals either remain
dormant or reactivate and return to the active size distribution. Fecundity
adds new recruits to the active portion of the kernel according to the fitted
recruit-size distribution.

The full demographic kernel can therefore be represented schematically as

$$
K =
\begin{bmatrix}
T + F & R \\
D & Z
\end{bmatrix},
$$

where $T$ contains survival and growth among active individuals, $F$ is the
fecundity kernel, $D$ represents entry into dormancy, $R$ represents
reactivation into the active state, and $Z$ represents continued dormancy.

### IPM functions
```{r}
inv_logit <- function(x) {
  exp(x) / (1 + exp(x))
}


# Survival ---------------------------------------------------------------------
sx <- function(
    x, pars, disturbance = 0,
    num_pars = pars$mod_su_index) {
  
  val <- pars$surv_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("surv_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  if (!is.null(pars$surv_bd)) {
    val <- val + pars$surv_bd * disturbance
  }
  
  inv_logit(val)
}


# Growth -----------------------------------------------------------------------
grow_mu <- function(
    x, pars, disturbance = 0,
    num_pars = pars$mod_gr_index) {
  
  val <- pars$grow_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("grow_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  if (!is.null(pars$grow_bd)) {
    val <- val + pars$grow_bd * disturbance
  }
  
  val
}


# Growth variation -------------------------------------------------------------
grow_sd <- function(mu, pars) {
  sqrt(pars$a * exp(pars$b * mu))
}


# Growth transition ------------------------------------------------------------
gxy <- function(x, y, pars, disturbance = 0) {
  mu <- grow_mu(
    x, pars,
    disturbance = disturbance)
  
  dnorm(
    y,
    mean = mu,
    sd = grow_sd(mu, pars))
}


# Dormancy entry ---------------------------------------------------------------
dorm_x <- function(
    x, pars, disturbance = 0,
    num_pars = pars$mod_do_index) {
  
  val <- pars$dorm_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("dorm_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  if (!is.null(pars$dorm_bd)) {
    val <- val + pars$dorm_bd * disturbance
  }
  
  inv_logit(val)
}


# Reactivation -----------------------------------------------------------------
react_p <- function(pars, disturbance = 0) {
  val <- pars$react_b0
  
  if (!is.null(pars$react_bd)) {
    val <- val + pars$react_bd * disturbance
  }
  
  inv_logit(val)
}


# Flowering probability --------------------------------------------------------
fl_x <- function(
    x, pars, disturbance_prev = 0,
    num_pars = pars$mod_fl_index) {
  
  val <- pars$fl_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("fl_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  if (!is.null(pars$fl_bd)) {
    val <- val + pars$fl_bd * disturbance_prev
  }
  
  inv_logit(val)
}


# Scape number conditional on flowering ---------------------------------------
fl_n_x <- function(
    x, pars, disturbance_prev = 0,
    num_pars = pars$mod_fl_n_index) {
  
  val <- pars$fln_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("fln_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  if (!is.null(pars$fln_bd)) {
    val <- val + pars$fln_bd * disturbance_prev
  }
  
  exp(val)
}


# Recruitment conversion -------------------------------------------------------
recruits_per_scape <- function(pars, disturbance = 0) {
  if (disturbance == 0) {
    pars$recr_0
  } else {
    pars$recr_1
  }
}

# Recruitment size distribution ------------------------------------------------
re_y_dist <- function(y, pars) {
  norm <- pnorm(
    pars$U, mean = pars$recr_sz, sd = pars$recr_sd) -
    pnorm(
      pars$L, mean = pars$recr_sz, sd = pars$recr_sd)
  
  dnorm(
    y,
    mean = pars$recr_sz,
    sd = pars$recr_sd) / norm
}


# Reactivation size distribution -----------------------------------------------
react_y_dist <- function(y, pars) {
  norm <- pnorm(
    pars$U, mean = pars$react_sz, sd = pars$react_sd) -
    pnorm(
      pars$L, mean = pars$react_sz, sd = pars$react_sd)
  
  dnorm(
    y,
    mean = pars$react_sz,
    sd = pars$react_sd) / norm
}


# F-kernel ---------------------------------------------------------------------
fyx <- function(
    y, x, pars,
    disturbance = 0,
    disturbance_prev = 0) {
  
  fl_x(
    x, pars,
    disturbance_prev = disturbance_prev) *
    fl_n_x(
      x, pars,
      disturbance_prev = disturbance_prev) *
    recruits_per_scape(
      pars,
      disturbance = disturbance) *
    re_y_dist(y, pars)
}
```

### Kernel
```{r}
kernel <- function(
    pars, disturbance = 0, disturbance_prev = 0) {
  
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  # Active survival and dormancy
  Smat <- sx(
    y, pars,
    disturbance = disturbance)
  
  Dmat <- dorm_x(
    y, pars,
    disturbance = disturbance)
  
  # Growth
  Gmat <- matrix(0, n, n)
  
  Gmat[] <- t(outer(
    y, y,
    Vectorize(function(x, y) {
      gxy(
        x, y, pars,
        disturbance = disturbance)
    }))) * h
  
  # Correct growth eviction
  for (i in seq_len(n)) {
    if (i <= n / 2) {
      Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    } else {
      Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    }
  }
  
  # Active -> active
  Tmat <- sweep(
    Gmat,
    2,
    Smat * (1 - Dmat),
    "*")
  
  # Active -> dormant
  Dorm_row <- matrix(
    Smat * Dmat,
    nrow = 1)
  
  # Dormant -> active
  React_vec <- react_y_dist(y, pars) * h
  React_vec <- React_vec / sum(React_vec)
  
  p_react <- react_p(
    pars,
    disturbance = disturbance)
  
  React_col <- matrix(
    p_react * React_vec,
    ncol = 1)
  
  # Dormant -> dormant
  Dorm_stasis <- 1 - p_react
  
  # Fertility
  Fmat <- outer(
    y, y,
    Vectorize(function(y, x) {
      fyx(
        y, x, pars,
        disturbance = disturbance,
        disturbance_prev = disturbance_prev)
    })) * h
  
  # Full 201 x 201 demographic kernel
  K <- rbind(
    cbind(Tmat + Fmat, React_col),
    cbind(Dorm_row, Dorm_stasis))
  
  return(list(
    k_yx = K,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Dorm_row = Dorm_row,
    React_col = React_col,
    Smat = Smat,
    Dmat = Dmat,
    p_reactivate = p_react,
    meshpts = y,
    h = h))
}
```

### Population growth rate
```{r}
lambda_ipm <- function(
    pars, disturbance = 0, disturbance_prev = 0) {
  
  Re(eigen(
    kernel(
      pars,
      disturbance = disturbance,
      disturbance_prev = disturbance_prev)$k_yx,
    only.values = TRUE)$values[1])
}
```

## Disturbance history
The four fixed disturbance kernels describe demographic outcomes under
specific disturbance histories but do not describe how frequently those
histories occur. The observed disturbance record was therefore used to
estimate transition probabilities between environmental states.

Across observed quadrat-year transitions, the probability of remaining
undisturbed following an undisturbed interval was 0.943, whereas the
probability of current disturbance was 0.057. Following a previously disturbed
interval, the probability of another disturbance was higher, at 0.205, while
the probability of returning to an undisturbed interval was 0.795.

Previous disturbance was uncommon overall: approximately 93.1% of observed
quadrat-years had no disturbance in the preceding interval and 6.9% had
previous disturbance. These frequencies are used only to initialize the
environmental-history population vector, whereas the conditional transition
probabilities determine movement between environmental states during repeated
```{r}
df_disturbance_regime <- df_re_all %>%
  filter(
    !is.na(disturbance),
    !is.na(disturbance_prev)) %>%
  distinct(
    site, pop, qu, year,
    disturbance, disturbance_prev) %>%
  count(
    disturbance,
    disturbance_prev,
    name = "n_quadrat_years") %>%
  complete(
    disturbance = c(0, 1),
    disturbance_prev = c(0, 1),
    fill = list(n_quadrat_years = 0)) %>%
  arrange(
    disturbance_prev,
    disturbance)

df_disturbance_regime
```

### Disturbance transition probabilities
```{r}
df_disturbance_transition <- df_disturbance_regime %>%
  group_by(disturbance_prev) %>%
  mutate(
    p_transition = n_quadrat_years /
      sum(n_quadrat_years)) %>%
  ungroup()

df_disturbance_transition

get_transition_probability <- function(
    disturbance_i, disturbance_prev_i) {
  
  df_disturbance_transition %>%
    filter(
      disturbance == disturbance_i,
      disturbance_prev == disturbance_prev_i) %>%
    pull(p_transition)
}

p_0_given_0 <- get_transition_probability(0, 0)
p_1_given_0 <- get_transition_probability(1, 0)
p_0_given_1 <- get_transition_probability(0, 1)
p_1_given_1 <- get_transition_probability(1, 1)
```

### Previous-disturbance probabilities
```{r}
df_previous_disturbance <- df_disturbance_regime %>%
  group_by(disturbance_prev) %>%
  summarise(
    n_quadrat_years = sum(n_quadrat_years),
    .groups = "drop") %>%
  mutate(
    p_previous = n_quadrat_years /
      sum(n_quadrat_years))

df_previous_disturbance

p_prev_0 <- df_previous_disturbance %>%
  filter(disturbance_prev == 0) %>%
  pull(p_previous)

p_prev_1 <- df_previous_disturbance %>%
  filter(disturbance_prev == 1) %>%
  pull(p_previous)
```

## State-specific demographic kernels
Current and previous disturbance enter different components of the demographic
kernel. Current disturbance can affect demographic transitions occurring from
$t$ to $t+1$, including growth, reactivation, and recruitment. Previous
disturbance can affect reproduction observed at time $t$, most importantly
flowering probability.

The full IPM must therefore distinguish four combinations of disturbance
history. We denote a demographic kernel by $K_{ij}$, where the first index
indicates current disturbance during $t$ to $t+1$ and the second indicates
previous disturbance during $t-1$ to $t$. Thus,

$$
K_{00}
$$

represents no fire in either interval,

$$
K_{10}
$$

represents current fire only,

$$
K_{01}
$$

represents previous fire only, and

$$
K_{11}
$$

represents fire in both intervals.

Although individual vital-rate models generally contain either current or
previous disturbance, both disturbance variables occur together in the full
fecundity and demographic kernels. This allows the immediate recruitment
response to fire and the lagged flowering response to fire to operate
simultaneously.
```{r}
K_00_obj <- kernel(
  pars,
  disturbance = 0,
  disturbance_prev = 0)

K_10_obj <- kernel(
  pars,
  disturbance = 1,
  disturbance_prev = 0)

K_01_obj <- kernel(
  pars,
  disturbance = 0,
  disturbance_prev = 1)

K_11_obj <- kernel(
  pars,
  disturbance = 1,
  disturbance_prev = 1)

K_00 <- K_00_obj$k_yx
K_10 <- K_10_obj$k_yx
K_01 <- K_01_obj$k_yx
K_11 <- K_11_obj$k_yx
```

## Environmental historical megamatrix
```{r}
K_env_hist <- rbind(
  cbind(
    p_0_given_0 * K_00,
    p_0_given_1 * K_01),
  cbind(
    p_1_given_0 * K_10,
    p_1_given_1 * K_11))
```

## Environmental historical asymptotic lambda
To retain disturbance history through time, the demographic state space is
duplicated for the two possible previous-disturbance states. Each environmental
state contains the complete 201-state demographic IPM, giving an
environmental-history projection matrix of dimension $402 \times 402$.

The environmental-history matrix is

$$
K_{\mathrm{env}} =
\begin{bmatrix}
P(0 \mid 0)K_{00} & P(0 \mid 1)K_{01} \\
P(1 \mid 0)K_{10} & P(1 \mid 1)K_{11}
\end{bmatrix}.
$$

Columns identify the disturbance state inherited from the preceding interval,
whereas rows identify disturbance during the current interval. Current
disturbance then becomes the previous-disturbance state for the following
projection. In this way, demographic dynamics and the observed temporal
dependence of disturbance are propagated simultaneously.
```{r}
lambda_env_hist <- Re(
  eigen(
    K_env_hist,
    only.values = TRUE)$values[1])

lambda_env_hist
```

## Mean observed population structure
```{r}
make_initial_n_mean <- function(pars, df_init = df_ind) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  
  breaks <- seq(
    L, U,
    length.out = n + 1)
  
  years_i <- df_init %>%
    filter(state %in% c("active", "dormant")) %>%
    distinct(year) %>%
    arrange(year) %>%
    pull(year)
  
  counts_by_year <- map(years_i, function(year_i) {
    df_i <- df_init %>%
      filter(year == year_i)
    
    active_size <- df_i %>%
      filter(
        state == "active",
        is.finite(logsize_t0)) %>%
      pull(logsize_t0)
    
    active_counts <- hist(
      pmin(pmax(active_size, L), U),
      breaks = breaks,
      plot = FALSE,
      include.lowest = TRUE)$counts
    
    dormant_count <- sum(
      df_i$state == "dormant",
      na.rm = TRUE)
    
    c(active_counts, dormant_count)
  })
  
  Reduce(`+`, counts_by_year) / length(counts_by_year)
}

n_obs <- make_initial_n_mean(pars)
```

## Initial environmental-history population
```{r}
n_env_initial <- c(
  p_prev_0 * n_obs,
  p_prev_1 * n_obs)
```

## One-step projection
```{r}
project_kernel <- function(K, n_initial, label) {
  n_initial_total <- sum(n_initial)
  
  n_proj <- as.numeric(
    K %*% n_initial)
  
  n_projected <- sum(n_proj)
  
  tibble(
    lambda_type = label,
    asym_lambda = Re(
      eigen(
        K,
        only.values = TRUE)$values[1]),
    proj_lambda = n_projected / n_initial_total,
    n_initial = n_initial_total,
    n_projected = n_projected)
}
```

## Fixed disturbance-state lambdas
```{r}
df_lambda_states <- bind_rows(
  project_kernel(
    K_00,
    n_obs,
    "Undisturbed") %>%
    mutate(
      disturbance = 0,
      disturbance_prev = 0),
  
  project_kernel(
    K_10,
    n_obs,
    "Current-year fire") %>%
    mutate(
      disturbance = 1,
      disturbance_prev = 0),
  
  project_kernel(
    K_01,
    n_obs,
    "Previous-year fire") %>%
    mutate(
      disturbance = 0,
      disturbance_prev = 1),
  
  project_kernel(
    K_11,
    n_obs,
    "Fire in both years") %>%
    mutate(
      disturbance = 1,
      disturbance_prev = 1))

df_lambda_states
```

## Environmental-history projection
```{r}
df_lambda_env_hist <- project_kernel(
  K_env_hist,
  n_env_initial,
  "Environmental historical regime") %>%
  mutate(
    disturbance = NA_real_,
    disturbance_prev = NA_real_)

df_lambda_env_hist
```

## Combined lambda table
```{r}
df_lambda_env <- bind_rows(
  df_lambda_states,
  df_lambda_env_hist)

df_lambda_env
```

## Observed population growth
```{r}
# Count known living individuals. New adults are additionally counted one year
# earlier, matching the simplified recruitment assumption used in this IPM.

df_counts_alive <- df_ind %>%
  filter(state %in% c("active", "dormant", "alive_or_dormant")) %>%
  count(site, pop, qu, year, name = "n_alive")

df_newadult_backfill <- df_ind %>%
  filter(recruit_type == "new_adult") %>%
  transmute(site, pop, qu, year = year - 1) %>%
  count(site, pop, qu, year, name = "n_newadult_backfill")

# Use recruitment rows as the monitored quadrat-year inventory
df_counts_quad <- df %>%
  filter(row_type == "recruitment") %>%
  distinct(site, pop, qu, year) %>%
  left_join(
    df_counts_alive,
    by = c("site", "pop", "qu", "year")) %>%
  left_join(
    df_newadult_backfill,
    by = c("site", "pop", "qu", "year")) %>%
  mutate(
    n_alive = replace_na(n_alive, 0L),
    n_newadult_backfill = replace_na(n_newadult_backfill, 0L),
    n = n_alive + n_newadult_backfill) %>%
  group_by(site, pop, qu) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n),
    annual_transition = year_t1 == year + 1) %>%
  ungroup() %>%
  filter(annual_transition)

# The final three transitions cannot fully resolve terminal dormancy
dormancy_cutoff <- 3
last_complete_transition <- max(df_ind$year, na.rm = TRUE) -
  dormancy_cutoff - 1

df_counts_year <- df_counts_quad %>%
  filter(year <= last_complete_transition) %>%
  group_by(year) %>%
  summarise(
    n_quads = n(),
    n_t0 = sum(n),
    n_t1 = sum(n_t1), .groups = "drop") %>%
  mutate(lambda_obs = n_t1 / n_t0)

df_counts_year

# Mean observed population growth
lam_obs_y <- df_counts_year$lambda_obs

lam_obs_mean <- mean(lam_obs_y, na.rm = TRUE)
lam_obs_geo <- exp(mean(log(lam_obs_y), na.rm = TRUE))

lam_obs_mean
lam_obs_geo
```

## Modeled versus observed mean population growth
```{r}
c(
  asym_lambda_env = lambda_env_hist,
  projected_lambda_env = df_lambda_env_hist$proj_lambda,
  lambda_obs_arithmetic = lam_obs_mean,
  lambda_obs_geometric = lam_obs_geo)


# Modeled versus observed population growth -----------------------------------
df_lambda_mean <- tibble(
  estimate = c(
    "IPM asymptotic",
    "IPM one-step projected",
    "Observed arithmetic",
    "Observed geometric"),
  lambda = c(
    lambda_env_hist,
    df_lambda_env_hist$proj_lambda,
    lam_obs_mean,
    lam_obs_geo))

df_lambda_mean
```

```{r}
# # Save data --------------------------------------------------------------------
# write.csv(df_og, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_original.csv')))
# write.csv(df_meta, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_meta.csv')))
# write.csv(df, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_workdata.csv')))
# write.csv(df_su, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_survival.csv')))
# write.csv(df_gr, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_growth.csv')))
# write.csv(df_fl, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_flower.csv')))
# write.csv(df_fr, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_fruit.csv')))
# write.csv(df_re, row.names = F,
#           file.path(dir_data,  paste0('ab_', v_sp_abb, '_df_recruit.csv')))
# 
```
