# Investigations 01 {#invest01}

## Presumed second-year recruits

Recruitment is not always observed at the seedling stage. Individuals first
recorded with `s = 5` are explicitly identified as seedlings, whereas individuals
first appearing with `s = 3` enter the demographic record as new adults. In the
recruitment reconstruction, these new adults are interpreted as recruits that
were not detected during their seedling year and are therefore assigned to an
earlier recruitment cohort.

The clearest cases supporting this interpretation are new adults that enter as
stage 2. These individuals are treated here as *presumed second-year recruits*.
Before using this interpretation in the IPM, we examined whether their sizes at
entry are consistent with the sizes actually reached by known seedlings one
year after their observed seedling census.

The purpose of this analysis is diagnostic only. No recruit classifications or
recorded sizes are changed.


```{r presumed-recruits-run, include=FALSE}
# Raw demographic data --------------------------------------------------------
df_recruit_og <- read_csv(
  file.path(
    dir_data,
    "eriogonum_longifolium_data.csv"),
  col_types = cols(
    comment = col_character())) %>%
  janitor::clean_names() %>%
  mutate(
    year = as.numeric(
      str_sub(date, 1, 4)),
    month = as.numeric(
      str_sub(date, 6, 7)),
    id = str_c(
      site, pop, qu, plant,
      sep = "_"),
    record_type = case_when(
      !is.na(s) ~ "demography",
      is.na(s) & !is.na(burn) ~ "burn",
      TRUE ~ "other"))


df_recruit_demog <- df_recruit_og %>%
  filter(
    record_type == "demography")


df_recruit_june <- df_recruit_demog %>%
  filter(
    month == 6)


# First year in which each quadrat was monitored -------------------------------
df_recruit_quad_start <- df_recruit_june %>%
  group_by(
    site, pop, qu) %>%
  summarise(
    first_quad_year = min(year),
    .groups = "drop")


# First recruitment record for each individual --------------------------------
df_recruit_first_diag <- df_recruit_demog %>%
  filter(
    s %in% c(3, 5)) %>%
  arrange(
    id, year, month) %>%
  distinct(
    id,
    .keep_all = TRUE) %>%
  mutate(
    recruit_year = if_else(
      month > 6,
      year + 1,
      year)) %>%
  left_join(
    df_recruit_quad_start,
    by = c(
      "site",
      "pop",
      "qu")) %>%
  mutate(
    baseline =
      recruit_year ==
      first_quad_year,
    recruit_type = case_when(
      s == 5 ~ "seedling",
      s == 3 ~ "new_adult"),
    recruit_class = case_when(
      s == 5 ~
        "observed_seedling",
      s == 3 & stage == 2 ~
        "presumed_2yr",
      s == 3 ~
        "new_adult_other")) %>%
  filter(
    !baseline)


# Annual active-to-active transitions -----------------------------------------
df_recruit_transition <- df_recruit_june %>%
  group_by(id) %>%
  arrange(
    year,
    .by_group = TRUE) %>%
  mutate(
    year_t1 = lead(year),
    s_t1 = lead(s),
    dia_t1 = lead(dia),
    annual_transition =
      year_t1 == year + 1,
    active_t0 =
      s %in% c(1, 3, 5),
    active_t1 =
      s_t1 %in% c(1, 3, 5),
    size_t0 = if_else(
      active_t0,
      dia,
      NA_real_),
    size_t1 = if_else(
      annual_transition &
        active_t0 &
        active_t1,
      dia_t1,
      NA_real_)) %>%
  ungroup()


# Recruit entries linked to annual size histories -----------------------------
df_recruit_entry_diag <- df_recruit_first_diag %>%
  transmute(
    site,
    pop,
    qu,
    plant,
    id,
    recruit_year,
    recruit_type,
    recruit_class,
    stage_entry = stage,
    raw_entry_dia = dia) %>%
  left_join(
    df_recruit_transition %>%
      dplyr::select(
        id,
        year,
        size_t0,
        size_t1),
    by = c(
      "id",
      "recruit_year" = "year"))


# Recruit-class counts ---------------------------------------------------------
df_recruit_class_summary <- df_recruit_entry_diag %>%
  count(
    recruit_class,
    name = "n") %>%
  mutate(
    Recruit_class = recode(
      recruit_class,
      observed_seedling =
        "Observed seedlings",
      presumed_2yr =
        "Presumed second-year recruits",
      new_adult_other =
        "Other new-adult entries")) %>%
  dplyr::select(
    Recruit_class,
    N = n)


# Empirical second-year sizes --------------------------------------------------
# Known seedlings:
# size at t + 1 after an observed seedling census.
#
# Presumed second-year recruits:
# size at first annual census as a stage-2 new adult.

df_year2_size <- bind_rows(
  df_recruit_entry_diag %>%
    filter(
      recruit_class ==
        "observed_seedling",
      size_t1 > 0) %>%
    transmute(
      id,
      source =
        "Observed seedlings at t + 1",
      year2_size = size_t1),

  df_recruit_entry_diag %>%
    filter(
      recruit_class ==
        "presumed_2yr",
      size_t0 > 0) %>%
    transmute(
      id,
      source =
        "Presumed second-year recruits",
      year2_size = size_t0)) %>%
  mutate(
    log_year2_size =
      log(year2_size))


# Size-distribution summary ---------------------------------------------------
df_year2_size_summary <- df_year2_size %>%
  group_by(source) %>%
  summarise(
    n = n(),
    mean = mean(year2_size),
    median = median(year2_size),
    sd = sd(year2_size),
    q25 = quantile(
      year2_size, 0.25),
    q75 = quantile(
      year2_size, 0.75),
    min = min(year2_size),
    max = max(year2_size),
    .groups = "drop")


# Size comparison figure ------------------------------------------------------
fig_year2_recruit_size <- ggplot(
  df_year2_size,
  aes(
    x = source,
    y = log_year2_size)) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.5) +
  geom_jitter(
    width = 0.12,
    alpha = 0.20,
    size = 1) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Log diameter") +
  theme(
    axis.text.x =
      element_text(
        angle = 15,
        hjust = 1))


# Growth after presumed second-year entry -------------------------------------
df_presumed_growth <- df_recruit_entry_diag %>%
  filter(
    recruit_class ==
      "presumed_2yr",
    size_t0 > 0,
    size_t1 > 0) %>%
  mutate(
    logsize_t0 = log(size_t0),
    logsize_t1 = log(size_t1))


df_growth_background <- df_recruit_transition %>%
  filter(
    annual_transition,
    active_t0,
    active_t1,
    size_t0 > 0,
    size_t1 > 0) %>%
  mutate(
    logsize_t0 = log(size_t0),
    logsize_t1 = log(size_t1))


fig_presumed_growth <- ggplot(
  df_growth_background,
  aes(
    x = logsize_t0,
    y = logsize_t1)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed") +
  geom_point(
    alpha = 0.10,
    size = 0.9) +
  geom_point(
    data = df_presumed_growth,
    shape = 21,
    size = 2.5,
    stroke = 0.8) +
  theme_bw() +
  labs(
    x = expression(log(Diameter[t])),
    y = expression(log(Diameter[t+1])))


# Compact numbers for text ----------------------------------------------------
n_presumed_2yr <- df_recruit_entry_diag %>%
  filter(
    recruit_class ==
      "presumed_2yr") %>%
  nrow()


n_presumed_growth <- nrow(
  df_presumed_growth)


n_presumed_lt1 <- df_recruit_entry_diag %>%
  filter(
    recruit_class ==
      "presumed_2yr",
    size_t0 > 0,
    size_t0 < 1) %>%
  nrow()
```


The first-record classifications are summarized below. Baseline individuals
already present when a quadrat entered monitoring are excluded because their
recruitment timing cannot be inferred from their first database record.

```{r presumed-recruit-counts, echo=FALSE}
df_recruit_class_summary %>%
  knitr::kable()
```


### Are presumed second-year recruits plausible second-year plants?

For observed seedlings with a subsequent active annual census, the size at the
following census provides an empirical distribution of second-year plant sizes.
This distribution can be compared directly with the entry sizes of stage-2 new
adults interpreted as presumed second-year recruits.

```{r presumed-recruit-size-table, echo=FALSE}
df_year2_size_summary %>%
  mutate(
    across(
      c(
        mean,
        median,
        sd,
        q25,
        q75,
        min,
        max),
      ~ round(.x, 2))) %>%
  rename(
    Group = source,
    N = n,
    Mean = mean,
    Median = median,
    SD = sd,
    `25%` = q25,
    `75%` = q75,
    Minimum = min,
    Maximum = max) %>%
  knitr::kable()
```


```{r presumed-recruit-size-figure, echo=FALSE, fig.width=7, fig.height=5}
fig_year2_recruit_size
```


The size distributions provide no indication that the stage-2 new adults form
an artificial group of exceptionally small observations. Instead, their entry
sizes overlap the size range reached by individuals that were directly observed
as seedlings in the preceding year. This supports the interpretation of these
stage-2 new adults as plausible second-year recruits, while recognizing that
their actual seedling census was not observed.


### Subsequent growth

There were `r n_presumed_2yr` presumed second-year recruit entries. Of these,
`r n_presumed_growth` had a positive active size measurement in the following
annual census and could therefore be compared with the complete annual growth
dataset.

```{r presumed-recruit-growth-figure, echo=FALSE, fig.width=7, fig.height=6}
fig_presumed_growth
```


The highlighted observations show the growth transitions beginning at presumed
second-year recruit entry. Their subsequent growth falls within the broader
annual growth distribution rather than forming a clearly separate growth
process.

Only `r n_presumed_lt1` presumed second-year recruits entered the annual data
below 1 cm. Thus, the later issue of extremely small non-seedling measurements
is not primarily generated by the reconstruction of presumed second-year
recruits.

Overall, the available size and growth information is consistent with retaining
stage-2 new adults as presumed second-year recruits. The assumption nevertheless
remains inferential because their seedling stage was not directly observed.
A separate sensitivity analysis can therefore test how recruitment and
population growth change when these reconstructed recruits are included or
excluded.



## Dormancy and uncertainty in terminal survival

Dormancy is an important component of the life history of *Eriogonum
longifolium*. Individuals may remain belowground for one or more annual censuses
before reappearing aboveground. Consequently, the disappearance of an
individual from a census cannot automatically be interpreted as mortality.

During construction of the working data, absences occurring between two known
living observations were retrospectively classified as dormancy. These cases
are unambiguous because subsequent reappearance demonstrates that the individual
had remained alive. In total, 409 reconstructed dormancy spells were identified
among 344 individuals. Most were short: 76.5% lasted one year, 90.5% no more
than two years, and 94.9% no more than three years. Nevertheless, substantially
longer dormant periods occurred, including a maximum reconstructed interval of
20 years.

The main uncertainty therefore concerns individuals disappearing near the end
of the study. Unlike earlier absences, these records cannot be resolved
retrospectively because there are insufficient subsequent censuses to determine
whether an individual died or remained dormant. In the working data, three
years was adopted as the default operational cutoff. An absence of up to three
years at the end of the record is treated as unresolved rather than as mortality.
The three-year threshold captures 388 of the 409 observed dormancy spells
(94.9%) and is therefore intended as a pragmatic classification rule rather than
as a biological maximum duration of dormancy.

### Sensitivity to the dormancy cutoff

Because the treatment of terminal absences directly affects estimated survival
and the years for which observed population growth can be considered complete,
we evaluated the sensitivity of the results to this assumption. The maximum
allowable period of unresolved dormancy was varied from 0 to 20 years, with
three years retained as the reference scenario.

For each cutoff, active individuals followed by an unresolved absence spell
longer than the specified cutoff were classified as dead, whereas shorter
terminal absences remained unresolved. The survival model was then refitted and
the mean IPM rebuilt. Independently, observed annual population growth was
recalculated after excluding terminal transitions for which the selected
dormancy allowance prevented mortality from being resolved. Confirmed dormancy
and reactivation observations, as well as observed growth, flowering and
recruitment data, were retained throughout the analysis.


```{r}
# Sensitivity specification ----------------------------------------------------
v_mod_set_su <- c()
v_mod_set_gr <- c()
v_mod_set_do <- c()
v_mod_set_fl <- c()
v_mod_set_fl_n <- c()

dormancy_cutoffs <- 0:20
reference_cutoff <- 3

v_ggp_suffix <- "Archbold - Eriogonum longifolium"
```


```{r dormancy-sensitivity-run, include=FALSE}
# Raw data ---------------------------------------------------------------------
df_og <- read_csv(
  file.path(dir_data, "eriogonum_longifolium_data.csv"),
  col_types = cols(comment = col_character())) %>%
  janitor::clean_names() %>%
  mutate(
    year = as.numeric(str_sub(date, 1, 4)),
    month = as.numeric(str_sub(date, 6, 7)),
    id = str_c(site, pop, qu, plant, sep = "_"),
    record_type = case_when(
      !is.na(s) ~ "demography",
      is.na(s) & !is.na(burn) ~ "burn",
      TRUE ~ "other"))

df_demog <- df_og %>%
  filter(record_type == "demography")

df_demog_june <- df_demog %>%
  filter(month == 6)


# Clean annual demographic history --------------------------------------------
# Absences bracketed by known living observations are confirmed dormancy.
# These remain dormant under every sensitivity scenario.

df_alive_range <- df_demog_june %>%
  filter(s %in% c(1, 3, 5)) %>%
  group_by(id) %>%
  summarise(
    first_alive = min(year),
    last_alive = max(year), .groups = "drop")

df_annual <- df_demog_june %>%
  left_join(df_alive_range, by = "id") %>%
  mutate(
    dormancy_repair = s %in% c(0, 9) &
      !is.na(first_alive) & year > first_alive & year < last_alive,
    s_clean = if_else(dormancy_repair, 8, s),
    stage_clean = if_else(dormancy_repair, 5, stage),
    state_clean = case_when(
      s_clean %in% c(1, 3, 5) ~ "active",
      s_clean == 8 ~ "dormant",
      s_clean == 10 ~ "alive_or_dormant",
      s_clean == 0 ~ "absent",
      s_clean == 9 ~ "previous_absent",
      s_clean %in% c(2, 6, 7) ~ "missing",
      s_clean == 11 ~ "discontinued",
      TRUE ~ NA_character_))


# Unresolved absence spells ----------------------------------------------------
df_absence_spell <- df_annual %>%
  group_by(id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    absent_state = state_clean %in% c("absent", "previous_absent"),
    new_spell = row_number() == 1 |
      coalesce(absent_state != lag(absent_state), TRUE) |
      coalesce(year != lag(year) + 1, TRUE),
    spell = cumsum(new_spell)) %>%
  group_by(id, spell) %>%
  summarise(
    absent_state = first(absent_state),
    absence_start = min(year),
    absence_end = max(year),
    absence_years = n(), .groups = "drop") %>%
  filter(absent_state)

df_absence_t1 <- df_absence_spell %>%
  transmute(
    id,
    year_t1 = absence_start,
    absence_years)


# Base annual transitions ------------------------------------------------------
df_transition_base <- df_annual %>%
  group_by(id) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    year_t1 = lead(year),
    annual_transition = coalesce(year_t1 == year + 1, FALSE),
    stage_t1 = if_else(
      annual_transition, lead(stage_clean), NA_real_),
    state_t1 = if_else(
      annual_transition, lead(state_clean), NA_character_),
    size_t0 = if_else(state_clean == "active", dia, NA_real_),
    size_t1 = if_else(
      annual_transition & state_clean == "active" &
        lead(state_clean) == "active",
      lead(dia), NA_real_),
    size_reactivate_t1 = if_else(
      annual_transition & state_clean == "dormant" &
        lead(state_clean) == "active",
      lead(dia), NA_real_)) %>%
  ungroup() %>%
  left_join(df_absence_t1, by = c("id", "year_t1"))


# Recruitment structure --------------------------------------------------------
df_quad_start <- df_demog_june %>%
  group_by(site, pop, qu) %>%
  summarise(first_quad_year = min(year), .groups = "drop")

df_recruit_first <- df_demog %>%
  filter(s %in% c(3, 5)) %>%
  arrange(id, year, month) %>%
  distinct(id, .keep_all = TRUE) %>%
  mutate(
    recruit_year = if_else(month > 6, year + 1, year)) %>%
  left_join(df_quad_start, by = c("site", "pop", "qu")) %>%
  mutate(
    baseline = recruit_year == first_quad_year,
    recruit_type = case_when(
      s == 5 ~ "seedling",
      s == 3 ~ "new_adult"))

df_quad_monitor <- df_annual %>%
  group_by(site, pop, qu, year) %>%
  summarise(
    monitored = any(state_clean != "discontinued", na.rm = TRUE),
    .groups = "drop") %>%
  filter(monitored) %>%
  select(-monitored)

df_recruit_followup <- df_quad_monitor %>%
  left_join(
    df_quad_monitor %>%
      transmute(
        site, pop, qu, year = year - 1,
        monitored_t1 = TRUE),
    by = c("site", "pop", "qu", "year")) %>%
  left_join(
    df_quad_monitor %>%
      transmute(
        site, pop, qu, year = year - 2,
        monitored_t2 = TRUE),
    by = c("site", "pop", "qu", "year")) %>%
  mutate(
    monitored_t1 = replace_na(monitored_t1, FALSE),
    monitored_t2 = replace_na(monitored_t2, FALSE))

df_recruit_count_simple <- df_recruit_first %>%
  filter(!baseline) %>%
  mutate(
    fecundity_year_simple = case_when(
      recruit_type == "seedling" ~ recruit_year - 1,
      recruit_type == "new_adult" ~ recruit_year - 2)) %>%
  semi_join(
    df_quad_monitor,
    by = c(
      "site", "pop", "qu",
      "fecundity_year_simple" = "year")) %>%
  count(
    site, pop, qu, fecundity_year_simple, recruit_type,
    name = "nr_recruits") %>%
  pivot_wider(
    names_from = recruit_type,
    values_from = nr_recruits,
    values_fill = 0) %>%
  rename(
    observed_seedling_simple = seedling,
    new_adult_simple = new_adult)

df_repro_quad <- df_quad_monitor %>%
  left_join(
    df_annual %>%
      filter(state_clean != "discontinued") %>%
      group_by(site, pop, qu, year) %>%
      summarise(
        nr_active = sum(state_clean == "active", na.rm = TRUE),
        nr_flower_obs = sum(
          state_clean == "active" & !is.na(scape), na.rm = TRUE),
        scape_sum = sum(
          if_else(state_clean == "active", scape, NA_real_),
          na.rm = TRUE), .groups = "drop") %>%
      mutate(
        nr_scapes = case_when(
          nr_active == 0 ~ 0,
          nr_flower_obs == 0 ~ NA_real_,
          TRUE ~ scape_sum)) %>%
      select(site, pop, qu, year, nr_active, nr_scapes),
    by = c("site", "pop", "qu", "year"))

df_recruit_quad <- df_recruit_followup %>%
  left_join(
    df_recruit_count_simple,
    by = c(
      "site", "pop", "qu",
      "year" = "fecundity_year_simple")) %>%
  left_join(
    df_repro_quad,
    by = c("site", "pop", "qu", "year")) %>%
  mutate(
    observed_seedling_simple = case_when(
      monitored_t1 ~ replace_na(observed_seedling_simple, 0L),
      TRUE ~ NA_integer_),
    new_adult_simple = case_when(
      monitored_t2 ~ replace_na(new_adult_simple, 0L),
      TRUE ~ NA_integer_),
    recruitment_complete = monitored_t1 & monitored_t2,
    recruits_simple = case_when(
      recruitment_complete ~
        observed_seedling_simple + new_adult_simple,
      TRUE ~ NA_integer_))

df_recruit <- df_recruit_quad %>%
  transmute(
    row_type = "recruitment",
    site, pop, qu, year,
    recruitment_complete,
    recruits_simple,
    nr_scapes)

df_recruit_ind <- df_recruit_first %>%
  filter(!baseline) %>%
  transmute(
    id,
    year = recruit_year,
    recruit_type)


# Scenario-specific workdata ---------------------------------------------------
make_scenario_data <- function(dormancy_cutoff) {
  
  df_transition <- df_transition_base %>%
    mutate(
      survives = case_when(
        !annual_transition ~ NA_real_,
        !state_clean %in% c("active", "dormant") ~ NA_real_,
        state_t1 %in%
          c("active", "dormant", "alive_or_dormant") ~ 1,
        state_t1 %in% c("absent", "previous_absent") &
          absence_years > dormancy_cutoff ~ 0,
        state_t1 %in% c("absent", "previous_absent") &
          absence_years <= dormancy_cutoff ~ NA_real_,
        TRUE ~ NA_real_),
      
      enter_dormancy = case_when(
        state_clean != "active" | survives != 1 ~ NA_real_,
        state_t1 == "dormant" ~ 1,
        state_t1 == "active" ~ 0,
        TRUE ~ NA_real_),
      
      reactivate = case_when(
        state_clean != "dormant" ~ NA_real_,
        state_t1 == "active" ~ 1,
        state_t1 == "dormant" ~ 0,
        TRUE ~ NA_real_),
      
      flower = case_when(
        state_clean != "active" ~ NA_real_,
        !is.na(scape) & scape > 0 ~ 1,
        !is.na(scape) & scape == 0 ~ 0,
        TRUE ~ NA_real_),
      
      fl_nr = if_else(
        state_clean == "active", scape, NA_real_),
      logsize_t0 = log(size_t0),
      logsize_t1 = log(size_t1),
      logsize_t0_2 = logsize_t0^2,
      logsize_t0_3 = logsize_t0^3)
  
  df_ind <- df_transition %>%
    left_join(df_recruit_ind, by = c("id", "year")) %>%
    transmute(
      row_type = "individual",
      site, pop, qu, id, year,
      state = state_clean,
      state_t1,
      stage = stage_clean,
      stage_t1,
      survives,
      enter_dormancy,
      recruit_type,
      size_reactivate_t1,
      reactivate,
      size_t0,
      size_t1,
      logsize_t0,
      logsize_t1,
      logsize_t0_2,
      logsize_t0_3,
      flower,
      fl_nr)
  
  bind_rows(df_ind, df_recruit) %>%
    mutate(
      row_type = as.character(row_type),
      year = as.numeric(year))
}


# Model helpers ----------------------------------------------------------------
poly_formulas <- function(response) {
  list(
    as.formula(paste(response, '~ 1')),
    as.formula(paste(response, '~ logsize_t0')),
    as.formula(paste(
      response, '~ logsize_t0 + logsize_t0_2')),
    as.formula(paste(
      response,
      '~ logsize_t0 + logsize_t0_2 + logsize_t0_3')))
}

get_daicc <- function(mods) {
  aicc <- sapply(mods, function(mod) {
    k <- attr(logLik(mod), "df")
    n <- nobs(mod)
    AIC(mod) + 2 * k * (k + 1) / (n - k - 1)
  })
  
  aicc - min(aicc)
}

pick_model <- function(mods, manual = c()) {
  d_aicc <- get_daicc(mods)
  index <- if (length(manual) == 0) {
    which.min(d_aicc)
  } else {
    manual[1] + 1
  }

  list(
    model = mods[[index]],
    index = index - 1,
    d_aicc = d_aicc)
}

fit_poly_glm <- function(data, response, manual = c()) {
  forms <- poly_formulas(response)
  mods <- lapply(
    forms,
    function(form) glm(form, data = data, family = 'binomial'))
  pick_model(mods, manual)
}

fit_poly_lm <- function(data, response, manual = c()) {
  forms <- poly_formulas(response)
  mods <- lapply(forms, function(form) lm(form, data = data))
  pick_model(mods, manual)
}

fit_poly_nb <- function(data, response, manual = c()) {
  forms <- poly_formulas(response)
  mods <- lapply(forms, function(form) MASS::glm.nb(form, data = data))
  pick_model(mods, manual)
}

get_coef <- function(model, term) {
  values <- coef(model)
  if (term %in% names(values)) {
    unname(values[[term]])
  } else {
    NULL
  }
}


# IPM functions ----------------------------------------------------------------
inv_logit <- function(x) {
  exp(x) / (1 + exp(x))
}

sx <- function(x, pars, num_pars = pars$mod_su_index) {
  val <- pars$surv_b0

  for (i in seq_len(num_pars)) {
    param <- paste0('surv_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }

  inv_logit(val)
}

grow_mu <- function(x, pars, num_pars = pars$mod_gr_index) {
  val <- pars$grow_b0

  for (i in seq_len(num_pars)) {
    param <- paste0('grow_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }

  val
}

grow_sd <- function(mu, pars) {
  sqrt(pars$a * exp(pars$b * mu))
}

gxy <- function(x, y, pars) {
  mu <- grow_mu(x, pars)
  dnorm(y, mean = mu, sd = grow_sd(mu, pars))
}

dorm_x <- function(x, pars, num_pars = pars$mod_do_index) {
  val <- pars$dorm_b0

  for (i in seq_len(num_pars)) {
    param <- paste0('dorm_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }

  inv_logit(val)
}

fl_x <- function(x, pars, num_pars = pars$mod_fl_index) {
  val <- pars$fl_b0

  for (i in seq_len(num_pars)) {
    param <- paste0('fl_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }

  inv_logit(val)
}

fl_n_x <- function(x, pars, num_pars = pars$mod_fl_n_index) {
  val <- pars$fln_b0

  for (i in seq_len(num_pars)) {
    param <- paste0('fln_b', i)
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }

  exp(val)
}

re_y_dist <- function(y, pars) {
  norm <- pnorm(
    pars$U, mean = pars$recr_sz, sd = pars$recr_sd) -
    pnorm(
      pars$L, mean = pars$recr_sz, sd = pars$recr_sd)

  dnorm(
    y, mean = pars$recr_sz, sd = pars$recr_sd) / norm
}

react_y_dist <- function(y, pars) {
  norm <- pnorm(
    pars$U, mean = pars$react_sz, sd = pars$react_sd) -
    pnorm(
      pars$L, mean = pars$react_sz, sd = pars$react_sd)

  dnorm(
    y, mean = pars$react_sz, sd = pars$react_sd) / norm
}

fyx <- function(y, x, pars) {
  fl_x(x, pars) *
    fl_n_x(x, pars) *
    pars$fecu_b0 *
    re_y_dist(y, pars)
}

kernel <- function(pars) {
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n

  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])

  Smat <- sx(y, pars)
  Dmat <- dorm_x(y, pars)

  Gmat <- matrix(0, n, n)
  Gmat[] <- t(outer(y, y, gxy, pars)) * h

  for (i in seq_len(n)) {
    if (i <= n / 2) {
      Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    } else {
      Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    }
  }

  Tmat <- sweep(
    Gmat, 2, Smat * (1 - Dmat), '*')

  Dorm_row <- matrix(
    Smat * Dmat, nrow = 1)

  React_vec <- react_y_dist(y, pars) * h
  React_vec <- React_vec / sum(React_vec)

  React_col <- matrix(
    pars$p_reactivate * React_vec, ncol = 1)

  Dorm_stasis <- 1 - pars$p_reactivate

  Fmat <- outer(
    y, y,
    Vectorize(function(y, x) {
      fyx(y, x, pars)
    })) * h

  K <- rbind(
    cbind(Tmat + Fmat, React_col),
    cbind(Dorm_row, Dorm_stasis))

  K
}

lambda_ipm <- function(pars) {
  Re(eigen(
    kernel(pars), only.values = TRUE)$values[1])
}


# Reference data for scenario-independent population counts -------------------
df_reference <- make_scenario_data(reference_cutoff)

df_ind_reference <- df_reference %>%
  filter(row_type == "individual")

max_ind_year <- max(df_annual$year, na.rm = TRUE)


# Observed population counts ---------------------------------------------------
df_counts_alive <- df_ind_reference %>%
  filter(state %in% c("active", "dormant", "alive_or_dormant")) %>%
  count(site, pop, qu, year, name = "n_alive")

df_newadult_backfill <- df_ind_reference %>%
  filter(recruit_type == "new_adult") %>%
  transmute(site, pop, qu, year = year - 1) %>%
  count(site, pop, qu, year, name = "n_newadult_backfill")

df_counts_quad <- df_reference %>%
  filter(row_type == "recruitment") %>%
  distinct(site, pop, qu, year) %>%
  left_join(
    df_counts_alive,
    by = c("site", "pop", "qu", "year")) %>%
  left_join(
    df_newadult_backfill,
    by = c("site", "pop", "qu", "year")) %>%
  mutate(
    n_alive = replace_na(n_alive, 0L),
    n_newadult_backfill = replace_na(n_newadult_backfill, 0L),
    n = n_alive + n_newadult_backfill) %>%
  group_by(site, pop, qu) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n),
    annual_transition = year_t1 == year + 1) %>%
  ungroup() %>%
  filter(annual_transition)


# One sensitivity scenario -----------------------------------------------------
run_dormancy_scenario <- function(dormancy_cutoff) {
  
  df_i <- make_scenario_data(dormancy_cutoff)
  
  df_ind_i <- df_i %>%
    filter(row_type == "individual")
  
  df_re_i <- df_i %>%
    filter(
      row_type == "recruitment",
      recruitment_complete,
      !is.na(recruits_simple))
  
  
  # Survival -------------------------------------------------------------------
  df_su_i <- df_ind_i %>%
    filter(
      state == "active",
      !is.na(survives),
      size_t0 > 0,
      is.finite(logsize_t0))
  
  fit_su_i <- fit_poly_glm(
    df_su_i, "survives", v_mod_set_su)
  
  mod_su_i <- fit_su_i$model
  
  
  # Growth ---------------------------------------------------------------------
  df_gr_i <- df_ind_i %>%
    filter(
      state == "active",
      state_t1 == "active",
      size_t0 > 0,
      size_t1 > 0,
      is.finite(logsize_t0),
      is.finite(logsize_t1))
  
  fit_gr_i <- fit_poly_lm(
    df_gr_i, "logsize_t1", v_mod_set_gr)
  
  mod_gr_i <- fit_gr_i$model
  
  mod_gr_x_i <- fitted(mod_gr_i)
  mod_gr_y_i <- resid(mod_gr_i)^2
  
  mod_gr_var_i <- nls(
    mod_gr_y_i ~ a * exp(b * mod_gr_x_i),
    start = list(a = 1, b = 0),
    control = nls.control(
      maxiter = 1000, tol = 1e-6, warnOnly = TRUE))
  
  
  # Dormancy entry -------------------------------------------------------------
  df_do_i <- df_ind_i %>%
    filter(
      !is.na(enter_dormancy),
      size_t0 > 0,
      is.finite(logsize_t0))
  
  fit_do_i <- fit_poly_glm(
    df_do_i, "enter_dormancy", v_mod_set_do)
  
  mod_do_i <- fit_do_i$model
  
  
  # Reactivation ---------------------------------------------------------------
  df_ra_i <- df_ind_i %>%
    filter(!is.na(reactivate))
  
  mod_ra_i <- glm(
    reactivate ~ 1,
    data = df_ra_i, family = "binomial")
  
  p_reactivate_i <- predict(
    mod_ra_i,
    newdata = data.frame(x = 1),
    type = "response")[1]
  
  df_ra_size_i <- df_ind_i %>%
    filter(
      reactivate == 1,
      size_reactivate_t1 > 0) %>%
    mutate(
      logsize_reactivate = log(size_reactivate_t1))
  
  react_sz_i <- mean(
    df_ra_size_i$logsize_reactivate, na.rm = TRUE)
  
  react_sd_i <- sd(
    df_ra_size_i$logsize_reactivate, na.rm = TRUE)
  
  
  # Flowering ------------------------------------------------------------------
  df_fl_i <- df_ind_i %>%
    filter(
      state == "active",
      !is.na(flower),
      size_t0 > 0,
      is.finite(logsize_t0))
  
  fit_fl_i <- fit_poly_glm(
    df_fl_i, "flower", v_mod_set_fl)
  
  mod_fl_i <- fit_fl_i$model
  
  
  # Flower number --------------------------------------------------------------
  df_fl_cond_i <- df_fl_i %>%
    filter(
      flower == 1,
      !is.na(fl_nr),
      fl_nr > 0,
      fl_nr %% 1 == 0)
  
  fit_fln_i <- fit_poly_nb(
    df_fl_cond_i, "fl_nr", v_mod_set_fl_n)
  
  mod_fln_i <- fit_fln_i$model
  
  
  # Recruitment ----------------------------------------------------------------
  df_sc2re_i <- df_re_i %>%
    filter(
      !is.na(nr_scapes),
      !is.na(recruits_simple)) %>%
    group_by(site, pop, year) %>%
    summarise(
      total_scapes = sum(nr_scapes),
      recruit_count = sum(recruits_simple), .groups = "drop")
  
  mods_re_i <- list(
    MASS::glm.nb(recruit_count ~ 1, data = df_sc2re_i),
    MASS::glm.nb(recruit_count ~ total_scapes, data = df_sc2re_i),
    MASS::glm.nb(
      recruit_count ~ log1p(total_scapes), data = df_sc2re_i))
  
  mod_re_i <- mods_re_i[[which.min(get_daicc(mods_re_i))]]
  
  df_sc2re_i <- df_sc2re_i %>%
    mutate(
      recruits_pred = predict(
        mod_re_i, type = "response"))
  
  recr_per_scape_i <- sum(df_sc2re_i$recruits_pred) /
    sum(df_sc2re_i$total_scapes)
  
  
  # Recruit size ---------------------------------------------------------------
  df_re_size_i <- df_ind_i %>%
    filter(
      recruit_type == "seedling",
      size_t0 > 0,
      is.finite(logsize_t0))
  
  recr_sz_i <- mean(
    df_re_size_i$logsize_t0, na.rm = TRUE)
  
  recr_sd_i <- sd(
    df_re_size_i$logsize_t0, na.rm = TRUE)
  
  
  # IPM parameters -------------------------------------------------------------
  pars_i <- Filter(function(x) length(x) > 0, list(
    surv_b0 = get_coef(mod_su_i, "(Intercept)"),
    surv_b1 = get_coef(mod_su_i, "logsize_t0"),
    surv_b2 = get_coef(mod_su_i, "logsize_t0_2"),
    surv_b3 = get_coef(mod_su_i, "logsize_t0_3"),
    
    grow_b0 = get_coef(mod_gr_i, "(Intercept)"),
    grow_b1 = get_coef(mod_gr_i, "logsize_t0"),
    grow_b2 = get_coef(mod_gr_i, "logsize_t0_2"),
    grow_b3 = get_coef(mod_gr_i, "logsize_t0_3"),
    a = unname(coef(mod_gr_var_i)[["a"]]),
    b = unname(coef(mod_gr_var_i)[["b"]]),
    
    dorm_b0 = get_coef(mod_do_i, "(Intercept)"),
    dorm_b1 = get_coef(mod_do_i, "logsize_t0"),
    dorm_b2 = get_coef(mod_do_i, "logsize_t0_2"),
    dorm_b3 = get_coef(mod_do_i, "logsize_t0_3"),
    
    fl_b0 = get_coef(mod_fl_i, "(Intercept)"),
    fl_b1 = get_coef(mod_fl_i, "logsize_t0"),
    fl_b2 = get_coef(mod_fl_i, "logsize_t0_2"),
    fl_b3 = get_coef(mod_fl_i, "logsize_t0_3"),
    
    fln_b0 = get_coef(mod_fln_i, "(Intercept)"),
    fln_b1 = get_coef(mod_fln_i, "logsize_t0"),
    fln_b2 = get_coef(mod_fln_i, "logsize_t0_2"),
    fln_b3 = get_coef(mod_fln_i, "logsize_t0_3"),
    
    fecu_b0 = recr_per_scape_i,
    recr_sz = recr_sz_i,
    recr_sd = recr_sd_i,
    react_sz = react_sz_i,
    react_sd = react_sd_i,
    p_reactivate = p_reactivate_i,
    
    L = min(df_gr_i$logsize_t0),
    U = max(df_gr_i$logsize_t0),
    mat_siz = 200,
    
    mod_su_index = fit_su_i$index,
    mod_gr_index = fit_gr_i$index,
    mod_do_index = fit_do_i$index,
    mod_fl_index = fit_fl_i$index,
    mod_fl_n_index = fit_fln_i$index))
  
  lambda_ipm_i <- lambda_ipm(pars_i)
  
  
  # Observed population growth -------------------------------------------------
  last_complete_transition <- max_ind_year - dormancy_cutoff - 1
  
  df_counts_year_i <- df_counts_quad %>%
    filter(year <= last_complete_transition) %>%
    group_by(year) %>%
    summarise(
      n_quads = n(),
      n_t0 = sum(n),
      n_t1 = sum(n_t1), .groups = "drop") %>%
    filter(n_t0 > 0) %>%
    mutate(lambda_obs = n_t1 / n_t0)
  
  lambda_obs_arithmetic <- mean(
    df_counts_year_i$lambda_obs, na.rm = TRUE)
  
  lambda_obs_geometric <- exp(mean(
    log(df_counts_year_i$lambda_obs), na.rm = TRUE))
  
  
  # Scenario summary -----------------------------------------------------------
  n_absence_dead <- df_ind_i %>%
    filter(
      state == "active",
      state_t1 %in% c("absent", "previous_absent"),
      survives == 0) %>%
    nrow()
  
  n_absence_unresolved <- df_ind_i %>%
    filter(
      state == "active",
      state_t1 %in% c("absent", "previous_absent"),
      is.na(survives)) %>%
    nrow()
  
  tibble(
    dormancy_cutoff = dormancy_cutoff,
    last_complete_transition = last_complete_transition,
    n_observed_transitions = nrow(df_counts_year_i),
    n_absence_dead = n_absence_dead,
    n_absence_unresolved = n_absence_unresolved,
    n_survival = nrow(df_su_i),
    n_survival_dead = sum(df_su_i$survives == 0),
    n_survival_alive = sum(df_su_i$survives == 1),
    survival_model_degree = fit_su_i$index,
    lambda_ipm = lambda_ipm_i,
    lambda_obs_arithmetic = lambda_obs_arithmetic,
    lambda_obs_geometric = lambda_obs_geometric)
}


# Run 0-20 years ----------------------------------------------------------------
df_sensitivity <- map_dfr(
  dormancy_cutoffs, run_dormancy_scenario)


# Summary relative to the current 3-year assumption ----------------------------
ref <- df_sensitivity %>%
  filter(dormancy_cutoff == reference_cutoff) %>%
  transmute(
    ref_ipm = lambda_ipm,
    ref_obs_arithmetic = lambda_obs_arithmetic,
    ref_obs_geometric = lambda_obs_geometric)

df_sensitivity_summary <- df_sensitivity %>%
  crossing(ref) %>%
  mutate(
    delta_ipm = lambda_ipm - ref_ipm,
    delta_obs_arithmetic =
      lambda_obs_arithmetic - ref_obs_arithmetic,
    delta_obs_geometric =
      lambda_obs_geometric - ref_obs_geometric,
    pct_delta_ipm = 100 * delta_ipm / ref_ipm,
    pct_delta_obs_arithmetic =
      100 * delta_obs_arithmetic / ref_obs_arithmetic,
    pct_delta_obs_geometric =
      100 * delta_obs_geometric / ref_obs_geometric) %>%
  select(-starts_with("ref_"))


# Final sensitivity plot -------------------------------------------------------
df_sensitivity_plot <- df_sensitivity_summary %>%
  select(
    dormancy_cutoff,
    `Mean IPM` = lambda_ipm,
    `Observed arithmetic` = lambda_obs_arithmetic,
    `Observed geometric` = lambda_obs_geometric) %>%
  pivot_longer(
    -dormancy_cutoff,
    names_to = "estimate",
    values_to = "lambda")

fig_dormancy_sensitivity <- ggplot(
  df_sensitivity_plot,
  aes(
    x = dormancy_cutoff,
    y = lambda,
    linetype = estimate,
    shape = estimate)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_vline(
    xintercept = reference_cutoff,
    linetype = "dotted") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.4) +
  scale_x_continuous(breaks = dormancy_cutoffs) +
  theme_bw() +
  labs(
    title = "Sensitivity to terminal dormancy assumption",
    subtitle = v_ggp_suffix,
    x = "Maximum unresolved dormancy (years)",
    y = expression(lambda),
    linetype = NULL,
    shape = NULL)
```

The effect of the assumption on the number of classified deaths and unresolved
terminal absences, together with the resulting estimates of population growth,
is summarized below.

```{r dormancy-sensitivity-table, echo=FALSE}
df_sensitivity %>%
  select(
    dormancy_cutoff,
    n_absence_dead,
    n_absence_unresolved,
    n_survival_dead,
    lambda_ipm,
    lambda_obs_arithmetic,
    lambda_obs_geometric) %>%
  rename(
    `Dormancy cutoff (years)` = dormancy_cutoff,
    `Absences classified as dead` = n_absence_dead,
    `Unresolved absences` = n_absence_unresolved,
    `Deaths in survival model` = n_survival_dead,
    `Mean IPM lambda` = lambda_ipm,
    `Observed arithmetic lambda` = lambda_obs_arithmetic,
    `Observed geometric lambda` = lambda_obs_geometric) %>%
  knitr::kable(digits = 3)
```

The three-year reference assumption is indicated by the vertical dotted line.
The horizontal dashed line marks $\lambda = 1$, separating projected population
decline from population growth.

```{r dormancy-sensitivity-figure, echo=FALSE, fig.width=8, fig.height=5}
fig_dormancy_sensitivity
```


## Small individuals and extreme size reversals

Inspection of the diameter data identified a small number of unusually small
measurements among established plants. Small diameters are not inherently
problematic because genuine seedlings and young plants occur at these sizes.
The relevant concern is therefore restricted to active non-seedlings, especially
individuals showing an abrupt one-year reduction to a very small diameter
followed by a return to substantially larger size.

No diameter observations were corrected or removed. The following analyses are
used only to identify unusual observations and assess whether they can be
explained by other information in the demographic records.


```{r small-individuals-growth-plot, echo=FALSE, fig.width=7, fig.height=6}
# Annual growth transitions ----------------------------------------------------
df_growth_plot <- df_annual %>%
  arrange(id, year) %>%
  group_by(id) %>%
  mutate(
    year_t1 = lead(year),
    state_t1 = lead(state_clean),
    stage_t1 = lead(stage_clean),
    dia_t1 = lead(dia)) %>%
  ungroup() %>%
  filter(
    year_t1 == year + 1,
    state_clean == "active",
    state_t1 == "active",
    dia > 0,
    dia_t1 > 0) %>%
  mutate(
    logdia_t0 = log(dia),
    logdia_t1 = log(dia_t1),
    flagged_small =
      (stage_clean != 1 & dia < 1) |
      (stage_t1 != 1 & dia_t1 < 1))


fig_small_growth <- ggplot(
  df_growth_plot,
  aes(
    x = logdia_t0,
    y = logdia_t1)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed") +
  geom_point(
    alpha = 0.25,
    size = 1.2) +
  geom_point(
    data = df_growth_plot %>%
      filter(flagged_small),
    shape = 21,
    size = 3,
    stroke = 1) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted") +
  geom_hline(
    yintercept = 0,
    linetype = "dotted") +
  theme_bw() +
  labs(
    x = expression(log(Diameter[t])),
    y = expression(log(Diameter[t+1])))

fig_small_growth
```


```{r small-individuals-run, include=FALSE}
# Active non-seedlings below 1 cm ----------------------------------------------
df_annual <- df_annual %>%
  mutate(
    flag_low_nonseedling =
      state_clean == "active" &
      !is.na(stage_clean) &
      stage_clean != 1 &
      !is.na(dia) &
      dia < 1)


df_low_nonseedling <- df_annual %>%
  filter(flag_low_nonseedling) %>%
  dplyr::select(
    site, pop, qu, plant, id, year,
    stage_clean,
    dia, scape, invol, nb, nr, herb,
    comment) %>%
  arrange(year, site, pop, qu, plant)


# Three-year active histories --------------------------------------------------
df_three_year <- df_annual %>%
  arrange(id, year) %>%
  group_by(id) %>%
  mutate(
    year_prev = lag(year),
    year_next = lead(year),
    state_prev = lag(state_clean),
    state_next = lead(state_clean),
    dia_prev = lag(dia),
    dia_next = lead(dia),
    stage_prev = lag(stage_clean),
    stage_next = lead(stage_clean),
    scape_prev = lag(scape),
    scape_next = lead(scape),
    herb_prev = lag(herb),
    herb_next = lead(herb)) %>%
  ungroup() %>%
  filter(
    year_prev == year - 1,
    year_next == year + 1,
    state_prev == "active",
    state_clean == "active",
    state_next == "active",
    dia_prev > 0,
    dia > 0,
    dia_next > 0) %>%
  mutate(
    logdia_prev = log(dia_prev),
    logdia = log(dia),
    logdia_next = log(dia_next),
    valley_depth =
      pmin(logdia_prev, logdia_next) - logdia,
    local_valley =
      dia < dia_prev &
      dia < dia_next,
    strict_small_valley =
      dia < 1 &
      dia_prev >= 1 &
      dia_next >= 1)


# Strict large-small-large observations ---------------------------------------
df_suspicious <- df_three_year %>%
  filter(strict_small_valley) %>%
  arrange(desc(valley_depth))


df_suspicious_table <- df_suspicious %>%
  transmute(
    ID = id,
    Year = year,
    `Previous stage` = stage_prev,
    `Focal stage` = stage_clean,
    `Next stage` = stage_next,
    `Previous diameter` = dia_prev,
    `Focal diameter` = dia,
    `Next diameter` = dia_next,
    Scapes = scape,
    Herbivory = herb,
    `Valley depth` = valley_depth)


# Trajectories -----------------------------------------------------------------
df_suspicious_long <- df_suspicious %>%
  dplyr::select(
    id, year,
    dia_prev, dia, dia_next) %>%
  pivot_longer(
    cols = c(dia_prev, dia, dia_next),
    names_to = "position",
    values_to = "diameter") %>%
  mutate(
    census_year = case_when(
      position == "dia_prev" ~ year - 1,
      position == "dia" ~ year,
      position == "dia_next" ~ year + 1))


fig_suspicious_trajectories <- ggplot(
  df_suspicious_long,
  aes(
    x = census_year,
    y = diameter,
    group = id)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.2) +
  facet_wrap(~ id, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Year",
    y = "Diameter")


# Flowering at small size ------------------------------------------------------
df_small_flowering <- df_annual %>%
  filter(
    state_clean == "active",
    !is.na(dia),
    dia < 3,
    !is.na(scape),
    scape > 0) %>%
  dplyr::select(
    id, year,
    stage_clean,
    dia, scape,
    herb) %>%
  arrange(dia)


# Flowering and three-year size deviations ------------------------------------
df_fl_valley <- df_three_year %>%
  filter(!is.na(scape)) %>%
  mutate(
    flower = scape > 0,
    expected_dia =
      sqrt(dia_prev * dia_next),
    deviation_middle =
      log(dia / expected_dia))


df_flowering_valley_summary <- df_fl_valley %>%
  group_by(flower) %>%
  summarise(
    n = n(),
    mean_deviation = mean(deviation_middle),
    median_deviation = median(deviation_middle),
    prop_local_valley = mean(local_valley),
    .groups = "drop")


# Herbivory check --------------------------------------------------------------
df_herb_test <- df_three_year %>%
  filter(herb %in% c(0, 1))


mod_herb_shrink <- lme4::lmer(
  log(dia) ~ log(dia_prev) + herb +
    (1 | year) +
    (1 | id),
  data = df_herb_test)


mod_herb_rebound <- lme4::lmer(
  log(dia_next) ~ log(dia) + herb +
    (1 | year) +
    (1 | id),
  data = df_herb_test)


herb_shrink_coef <-
  unname(lme4::fixef(mod_herb_shrink)[["herb"]])

herb_rebound_coef <-
  unname(lme4::fixef(mod_herb_rebound)[["herb"]])


# Summary ----------------------------------------------------------------------
df_small_summary <- tibble(
  Diagnostic = c(
    "Active non-seedling observations below 1 cm",
    "Strict >=1 -> <1 -> >=1 size reversals",
    "Flowering observations below 1 cm",
    "Flowering observations below 3 cm"),
  N = c(
    sum(df_annual$flag_low_nonseedling, na.rm = TRUE),
    nrow(df_suspicious),
    sum(
      df_annual$state_clean == "active" &
        !is.na(df_annual$dia) &
        df_annual$dia < 1 &
        !is.na(df_annual$scape) &
        df_annual$scape > 0,
      na.rm = TRUE),
    nrow(df_small_flowering)))
```


There were only a small number of active non-seedling observations at extremely
small sizes. The most informative cases are those in which an established plant
was at least 1 cm in the preceding year, fell below 1 cm in the focal census,
and returned to at least 1 cm in the following year.

```{r small-individuals-summary, echo=FALSE}
df_small_summary %>%
  knitr::kable()
```


Seven observations met this strict definition. Their complete three-year size
sequences are shown below.

```{r small-individuals-table, echo=FALSE}
df_suspicious_table %>%
  mutate(
    across(
      c(
        `Previous diameter`,
        `Focal diameter`,
        `Next diameter`,
        `Valley depth`),
      ~ round(.x, 2))) %>%
  knitr::kable()
```


```{r small-individuals-figure, echo=FALSE, fig.width=9, fig.height=7}
fig_suspicious_trajectories
```


The strongest reversals were substantially more extreme than ordinary annual
size variation. For example, the most pronounced sequence was 24.0, 0.5 and
19.9 cm in consecutive years. The focal observation was nevertheless recorded
as an active and flowering plant, with three scapes and recorded herbivory.
Other extreme reversals occurred without the same associated information, and
no common explanatory variable was identified across all seven observations.

## Flowering at very small size

Flowering at very small recorded diameter was itself rare.

```{r flowering-number-size-plot, echo=FALSE, fig.width=7, fig.height=5.5}
# Number of flowers against size -----------------------------------------------
df_flowering_plot <- df_annual %>%
  filter(
    state_clean == "active",
    !is.na(dia),
    dia > 0,
    !is.na(scape)) %>%
  mutate(
    logdia = log(dia),
    small_flowering = scape > 0 & dia < 3)

fig_flowering_number_size <- ggplot(
  df_flowering_plot,
  aes(
    x = logdia,
    y = scape)) +
  geom_jitter(
    height = 0.15,
    width = 0,
    alpha = 0.18,
    size = 1) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "poisson"),
    se = TRUE,
    linewidth = 1) +
  geom_point(
    data = df_flowering_plot %>%
      filter(small_flowering),
    shape = 21,
    size = 3,
    stroke = 1) +
  geom_vline(
    xintercept = 0,
    linetype = "dotted") +
  theme_bw() +
  labs(
    x = "Log diameter",
    y = "Number of scapes")

fig_flowering_number_size
```


```{r small-flowering-table, echo=FALSE}
df_small_flowering %>%
  rename(
    ID = id,
    Year = year,
    Stage = stage_clean,
    Diameter = dia,
    Scapes = scape,
    Herbivory = herb) %>%
  knitr::kable()
```


The 0.5-cm observation was the only flowering record below 1 cm. However,
flowering itself does not appear to produce a general temporary reduction in
diameter. Across complete three-year histories, flowering observations had a
slightly positive rather than negative mean deviation from the size expected
from the preceding and following censuses.

```{r flowering-valley-summary, echo=FALSE}
df_flowering_valley_summary %>%
  mutate(
    Flowering = if_else(flower, "Yes", "No"),
    mean_deviation = round(mean_deviation, 3),
    median_deviation = round(median_deviation, 3),
    prop_local_valley = round(prop_local_valley, 3)) %>%
  dplyr::select(
    Flowering,
    N = n,
    `Mean log deviation` = mean_deviation,
    `Median log deviation` = median_deviation,
    `Proportion local valleys` = prop_local_valley) %>%
  knitr::kable()
```


Recorded herbivory may therefore provide a biologically relevant explanation
for the single most extreme observation, but it does not provide a general
explanation for the unusual size reversals. In the available observations with
herbivory coded as either absent or present, its estimated effects on size
reduction (`r round(herb_shrink_coef, 3)`) and subsequent recovery
(`r round(herb_rebound_coef, 3)`) were both small.

The unusual diameter records are therefore retained unchanged. Their primary
importance is that they represent a small set of potentially influential
observations that can later be examined explicitly in sensitivity analyses of
the growth model.



## Diameter measurement precision and heaping

A separate feature of the diameter data is the high frequency of exact integer
and half-unit measurements. This is distinct from the issue of unusually small
individuals. Here the question is whether particular numerical values indicate
isolated suspicious records or instead reflect the precision with which
diameter was recorded during different parts of the study.

```{r growth-heaping-plot, echo=FALSE, fig.width=7, fig.height=6}
# Growth plot with heaped diameter observations highlighted --------------------
df_growth_heap <- df_annual %>%
  arrange(id, year) %>%
  group_by(id) %>%
  mutate(
    year_t1 = lead(year),
    state_t1 = lead(state_clean),
    dia_t1 = lead(dia)) %>%
  ungroup() %>%
  filter(
    year_t1 == year + 1,
    state_clean == "active",
    state_t1 == "active",
    dia > 0,
    dia_t1 > 0) %>%
  mutate(
    logdia_t0 = log(dia),
    logdia_t1 = log(dia_t1),

    integer_t0 = dia %% 1 == 0,
    integer_t1 = dia_t1 %% 1 == 0,

    half_t0 = (dia * 2) %% 1 == 0 & !integer_t0,
    half_t1 = (dia_t1 * 2) %% 1 == 0 & !integer_t1,

    heap_class = case_when(
      integer_t0 | integer_t1 ~ "Exact integer",
      half_t0 | half_t1 ~ "Half-unit only",
      TRUE ~ "Other"))


fig_growth_heaping <- ggplot(
  df_growth_heap,
  aes(
    x = logdia_t0,
    y = logdia_t1)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed") +
  geom_point(
    data = df_growth_heap %>%
      filter(heap_class == "Other"),
    alpha = 0.18,
    size = 1) +
  geom_point(
    data = df_growth_heap %>%
      filter(heap_class != "Other"),
    aes(color = heap_class),
    size = 1.5,
    alpha = 0.5) +
  theme_bw() +
  labs(
    x = expression(log(Diameter[t])),
    y = expression(log(Diameter[t+1])),
    shape = "Heaping")

fig_growth_heaping
```


```{r diameter-heaping-run, include=FALSE}
# Overall heaping --------------------------------------------------------------
df_heaping_summary <- df_annual %>%
  filter(
    state_clean == "active",
    !is.na(dia),
    dia > 0) %>%
  summarise(
    n = n(),
    exact_integer = sum(dia %% 1 == 0),
    prop_integer = mean(dia %% 1 == 0),
    exact_half = sum((dia * 2) %% 1 == 0),
    prop_half = mean((dia * 2) %% 1 == 0))


# Heaping through time ---------------------------------------------------------
df_heap_year <- df_annual %>%
  filter(
    state_clean == "active",
    !is.na(dia),
    dia > 0) %>%
  group_by(year) %>%
  summarise(
    n = n(),
    prop_integer = mean(dia %% 1 == 0),
    prop_half = mean((dia * 2) %% 1 == 0),
    n_unique = n_distinct(dia),
    .groups = "drop")


# Display summary --------------------------------------------------------------
df_heaping_display <- tibble(
  Diagnostic = c(
    "Positive active diameter observations",
    "Exact integer observations",
    "Multiples of 0.5, including integers"),
  N = c(
    df_heaping_summary$n,
    df_heaping_summary$exact_integer,
    df_heaping_summary$exact_half),
  Proportion = c(
    1,
    df_heaping_summary$prop_integer,
    df_heaping_summary$prop_half))


# Figure -----------------------------------------------------------------------
fig_heaping_year <- df_heap_year %>%
  dplyr::select(
    year,
    `Exact integer` = prop_integer,
    `Multiple of 0.5` = prop_half) %>%
  pivot_longer(
    -year,
    names_to = "precision",
    values_to = "proportion") %>%
  ggplot(
    aes(
      x = year,
      y = proportion,
      linetype = precision,
      shape = precision)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  theme_bw() +
  labs(
    x = "Year",
    y = "Proportion of active diameter records",
    linetype = NULL,
    shape = NULL)
```


Across all positive active diameter observations, exact integer and half-unit
values were common.

```{r diameter-heaping-summary, echo=FALSE}
df_heaping_display %>%
  mutate(
    Proportion = round(Proportion, 3)) %>%
  knitr::kable()
```


However, this heaping was strongly structured through time.

```{r diameter-heaping-figure, echo=FALSE, fig.width=8, fig.height=5}
fig_heaping_year
```


In the earliest censuses almost all recorded diameters were integers. Integer
recording remained dominant through much of the 1990s, whereas from 2000 onward
the proportion of integer observations declined sharply and measurements at
finer decimal resolution became common. The same temporal pattern is evident
for measurements falling exactly on half-unit values.

This indicates that integer and half-unit values are primarily a feature of
changing measurement or recording resolution rather than a set of isolated
suspicious observations. In particular, there is no basis for treating values
such as exactly 1 or 2 cm as erroneous solely because they are integers.
Consequently, these observations are retained unchanged. The temporal change in
measurement precision should instead be kept in mind when interpreting the
growth data and its residual variation.



### Sensitivity of the growth model

The unusually small diameter observations may have disproportionate influence
on the growth model because a single extreme measurement contributes both a
large apparent shrinkage into the observation and a large apparent increase
from it in the following year. We therefore refitted the growth model under
three scenarios:

1. all observations retained;
2. transitions involving one of the seven strict
   `>= 1 -> < 1 -> >= 1 cm` reversals removed; and
3. transitions involving any active non-seedling diameter below 1 cm removed.

For a flagged diameter observation, both the transition entering and the
transition leaving that observation were excluded. No recorded diameter was
modified.


```{r growth-sensitivity-run, include=FALSE}
# Flag strict focal observations ----------------------------------------------
df_strict_key <- df_suspicious %>%
  dplyr::select(id, year) %>%
  distinct() %>%
  mutate(flag_strict_focal = TRUE)


# Growth data with endpoint flags ---------------------------------------------
df_growth_sensitivity <- df_annual %>%
  left_join(
    df_strict_key,
    by = c("id", "year")) %>%
  mutate(
    flag_strict_focal =
      replace_na(flag_strict_focal, FALSE)) %>%
  arrange(id, year) %>%
  group_by(id) %>%
  mutate(
    year_t1 = lead(year),
    state_t1 = lead(state_clean),
    dia_t1 = lead(dia),

    flag_low_t0 = flag_low_nonseedling,
    flag_low_t1 = lead(flag_low_nonseedling),

    flag_strict_t0 = flag_strict_focal,
    flag_strict_t1 = lead(flag_strict_focal)) %>%
  ungroup() %>%
  filter(
    year_t1 == year + 1,
    state_clean == "active",
    state_t1 == "active",
    dia > 0,
    dia_t1 > 0) %>%
  mutate(
    logsize_t0 = log(dia),
    logsize_t1 = log(dia_t1),
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3,

    touches_strict =
      flag_strict_t0 |
      replace_na(flag_strict_t1, FALSE),

    touches_low_nonseedling =
      flag_low_t0 |
      replace_na(flag_low_t1, FALSE))


# Sensitivity datasets --------------------------------------------------------
growth_data_list <- list(
  `All observations` =
    df_growth_sensitivity,

  `Exclude strict reversals` =
    df_growth_sensitivity %>%
    filter(!touches_strict),

  `Exclude all <1 cm non-seedlings` =
    df_growth_sensitivity %>%
    filter(!touches_low_nonseedling))


# Same growth candidate models as the IPM -------------------------------------
growth_formulas <- list(
  logsize_t1 ~ 1,
  logsize_t1 ~ logsize_t0,
  logsize_t1 ~ logsize_t0 + logsize_t0_2,
  logsize_t1 ~
    logsize_t0 + logsize_t0_2 + logsize_t0_3)


fit_growth_scenario <- function(data, scenario) {

  mods <- lapply(
    growth_formulas,
    function(form) {
      lm(
        form,
        data = data)
    })

  aicc <- sapply(
    mods,
    function(mod) {
      k <- attr(logLik(mod), "df")
      n <- nobs(mod)

      AIC(mod) +
        2 * k * (k + 1) /
        (n - k - 1)
    })

  best_index <- which.min(aicc)
  mod <- mods[[best_index]]

  mod_gr_x <- fitted(mod)
  mod_gr_y <- resid(mod)^2

  mod_var <- nls(
    mod_gr_y ~ a * exp(b * mod_gr_x),
    start = list(
      a = 1,
      b = 0),
    control = nls.control(
      maxiter = 1000,
      tol = 1e-6,
      warnOnly = TRUE))

  list(
    scenario = scenario,
    data = data,
    model = mod,
    var_model = mod_var,
    degree = best_index - 1,
    aicc = aicc)
}


growth_fits <- imap(
  growth_data_list,
  ~ fit_growth_scenario(
    .x,
    .y))


# Model summary ---------------------------------------------------------------
get_growth_coef <- function(mod, term) {

  value <- coef(mod)[term]

  if (length(value) == 0) {
    return(NA_real_)
  }

  unname(value)
}


df_growth_sensitivity_summary <- map_dfr(
  growth_fits,
  function(x) {

    tibble(
      Scenario = x$scenario,
      N = nrow(x$data),
      `Excluded transitions` =
        nrow(df_growth_sensitivity) -
        nrow(x$data),
      `Selected degree` = x$degree,
      Intercept =
        get_growth_coef(
          x$model,
          "(Intercept)"),
      Linear =
        get_growth_coef(
          x$model,
          "logsize_t0"),
      Quadratic =
        get_growth_coef(
          x$model,
          "logsize_t0_2"),
      Cubic =
        get_growth_coef(
          x$model,
          "logsize_t0_3"),
      RMSE =
        sqrt(mean(resid(x$model)^2)),
      `Variance a` =
        unname(
          coef(x$var_model)[["a"]]),
      `Variance b` =
        unname(
          coef(x$var_model)[["b"]]))
  })


# Predicted growth curves -----------------------------------------------------
growth_grid <- tibble(
  logsize_t0 = seq(
    min(
      df_growth_sensitivity$logsize_t0),
    max(
      df_growth_sensitivity$logsize_t0),
    length.out = 300)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)


df_growth_predictions <- map_dfr(
  growth_fits,
  function(x) {

    growth_grid %>%
      mutate(
        Scenario = x$scenario,
        pred_logsize_t1 =
          predict(
            x$model,
            newdata = growth_grid))
  })


# Difference from the complete-data model -------------------------------------
df_growth_pred_wide <- df_growth_predictions %>%
  dplyr::select(
    Scenario,
    logsize_t0,
    pred_logsize_t1) %>%
  pivot_wider(
    names_from = Scenario,
    values_from = pred_logsize_t1)


df_growth_sensitivity_difference <-
  df_growth_pred_wide %>%
  summarise(
    `Strict reversals: max absolute difference` =
      max(
        abs(
          `Exclude strict reversals` -
            `All observations`)),
    `<1 cm non-seedlings: max absolute difference` =
      max(
        abs(
          `Exclude all <1 cm non-seedlings` -
            `All observations`))) %>%
  pivot_longer(
    everything(),
    names_to = "Comparison",
    values_to = "Max_difference")


# Growth sensitivity figure ---------------------------------------------------
fig_growth_sensitivity <- ggplot() +
  geom_point(
    data = df_growth_sensitivity,
    aes(
      x = logsize_t0,
      y = logsize_t1),
    alpha = 0.08,
    size = 0.8) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed") +
  geom_line(
    data = df_growth_predictions,
    aes(
      x = logsize_t0,
      y = pred_logsize_t1,
      linetype = Scenario),
    linewidth = 1) +
  theme_bw() +
  labs(
    x = expression(log(Diameter[t])),
    y = expression(log(Diameter[t+1])),
    linetype = NULL)
```


The sensitivity analysis changes only which observations are used to estimate
growth; all other demographic data remain unchanged.

```{r growth-sensitivity-table, echo=FALSE}
df_growth_sensitivity_summary %>%
  mutate(
    across(
      c(
        Intercept,
        Linear,
        Quadratic,
        Cubic,
        RMSE,
        `Variance a`,
        `Variance b`),
      ~ round(.x, 4))) %>%
  knitr::kable()
```


```{r growth-sensitivity-figure, echo=FALSE, fig.width=7, fig.height=6}
fig_growth_sensitivity
```


In addition to changes in individual model coefficients, the fitted curves were
compared directly over the complete observed size range. The maximum absolute
difference in predicted log diameter after removing transitions involving the
seven strict reversals was
`r round(df_growth_sensitivity_difference$Max_difference[1], 4)`,
whereas removing transitions involving any of the active non-seedling
observations below 1 cm produced a maximum difference of
`r round(df_growth_sensitivity_difference$Max_difference[2], 4)` log-diameter
units.

Because the extreme observations can also affect the estimated spread of the
growth distribution, the table reports the parameters of the same
size-dependent residual-variance model used in the IPM. Thus, the sensitivity
analysis evaluates both the fitted mean growth relationship and the variance
around it rather than considering only changes in the regression line.




### Sensitivity to the treatment of new adults

The reconstruction of recruitment assumes that individuals first recorded as
new adults represent recruits that were missed during their seedling census.
The current analysis assigns these individuals to the recruitment cohort two
years before their first adult census. Although the size analysis supports this
interpretation particularly for stage-2 new adults, recruitment timing remains
inferred rather than observed.

We therefore compared five alternative treatments of new adults. All
non-recruit vital rates and the 3-year dormancy assumption were held constant,
so differences among scenarios isolate the effect of the recruitment
reconstruction.


```{r recruit-sensitivity-run, include=FALSE}
# Recruit classes --------------------------------------------------------------
df_recruit_first_sens <- df_recruit_first %>%
  mutate(
    recruit_class = case_when(
      recruit_type == "seedling" ~
        "observed_seedling",
      recruit_type == "new_adult" &
        stage == 2 ~
        "presumed_2yr",
      recruit_type == "new_adult" ~
        "new_adult_other"))


# Scenario names ---------------------------------------------------------------
recruit_scenarios <- c(
  "Current: all new adults = year-2 recruits",
  "Stage-2 only = year-2 recruits",
  "New adults enter only as adults",
  "All new adults = year-1 recruits",
  "Stage-2 only = year-1 recruits")


# Fixed reference demographic data --------------------------------------------
# Recruitment treatment does not alter survival, growth, dormancy, flowering
# or reactivation. These components are therefore fitted once at the reference
# 3-year dormancy cutoff.

df_recruit_reference <- make_scenario_data(
  reference_cutoff)

df_ind_recruit_reference <- df_recruit_reference %>%
  filter(
    row_type == "individual")


# Survival ---------------------------------------------------------------------
df_su_rs <- df_ind_recruit_reference %>%
  filter(
    state == "active",
    !is.na(survives),
    size_t0 > 0,
    is.finite(logsize_t0))

fit_su_rs <- fit_poly_glm(
  df_su_rs,
  "survives",
  v_mod_set_su)

mod_su_rs <- fit_su_rs$model


# Growth -----------------------------------------------------------------------
df_gr_rs <- df_ind_recruit_reference %>%
  filter(
    state == "active",
    state_t1 == "active",
    size_t0 > 0,
    size_t1 > 0,
    is.finite(logsize_t0),
    is.finite(logsize_t1))

fit_gr_rs <- fit_poly_lm(
  df_gr_rs,
  "logsize_t1",
  v_mod_set_gr)

mod_gr_rs <- fit_gr_rs$model

mod_gr_x_rs <- fitted(
  mod_gr_rs)

mod_gr_y_rs <- resid(
  mod_gr_rs)^2

mod_gr_var_rs <- nls(
  mod_gr_y_rs ~ a * exp(b * mod_gr_x_rs),
  start = list(
    a = 1,
    b = 0),
  control = nls.control(
    maxiter = 1000,
    tol = 1e-6,
    warnOnly = TRUE))


# Dormancy entry ---------------------------------------------------------------
df_do_rs <- df_ind_recruit_reference %>%
  filter(
    !is.na(enter_dormancy),
    size_t0 > 0,
    is.finite(logsize_t0))

fit_do_rs <- fit_poly_glm(
  df_do_rs,
  "enter_dormancy",
  v_mod_set_do)

mod_do_rs <- fit_do_rs$model


# Reactivation -----------------------------------------------------------------
df_ra_rs <- df_ind_recruit_reference %>%
  filter(
    !is.na(reactivate))

mod_ra_rs <- glm(
  reactivate ~ 1,
  data = df_ra_rs,
  family = "binomial")

p_reactivate_rs <- predict(
  mod_ra_rs,
  newdata = data.frame(x = 1),
  type = "response")[1]


df_ra_size_rs <- df_ind_recruit_reference %>%
  filter(
    reactivate == 1,
    size_reactivate_t1 > 0) %>%
  mutate(
    logsize_reactivate =
      log(size_reactivate_t1))

react_sz_rs <- mean(
  df_ra_size_rs$logsize_reactivate,
  na.rm = TRUE)

react_sd_rs <- sd(
  df_ra_size_rs$logsize_reactivate,
  na.rm = TRUE)


# Flowering --------------------------------------------------------------------
df_fl_rs <- df_ind_recruit_reference %>%
  filter(
    state == "active",
    !is.na(flower),
    size_t0 > 0,
    is.finite(logsize_t0))

fit_fl_rs <- fit_poly_glm(
  df_fl_rs,
  "flower",
  v_mod_set_fl)

mod_fl_rs <- fit_fl_rs$model


# Number of scapes -------------------------------------------------------------
df_fln_rs <- df_fl_rs %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr %% 1 == 0)

fit_fln_rs <- fit_poly_nb(
  df_fln_rs,
  "fl_nr",
  v_mod_set_fl_n)

mod_fln_rs <- fit_fln_rs$model


# Recruit size distribution ----------------------------------------------------
# This remains based on directly observed seedlings in every scenario.

df_re_size_rs <- df_ind_recruit_reference %>%
  filter(
    recruit_type == "seedling",
    size_t0 > 0,
    is.finite(logsize_t0))

recr_sz_rs <- mean(
  df_re_size_rs$logsize_t0,
  na.rm = TRUE)

recr_sd_rs <- sd(
  df_re_size_rs$logsize_t0,
  na.rm = TRUE)


# Base IPM parameters ----------------------------------------------------------
pars_recruit_base <- Filter(
  function(x) length(x) > 0,
  list(
    surv_b0 =
      get_coef(
        mod_su_rs,
        "(Intercept)"),
    surv_b1 =
      get_coef(
        mod_su_rs,
        "logsize_t0"),
    surv_b2 =
      get_coef(
        mod_su_rs,
        "logsize_t0_2"),
    surv_b3 =
      get_coef(
        mod_su_rs,
        "logsize_t0_3"),

    grow_b0 =
      get_coef(
        mod_gr_rs,
        "(Intercept)"),
    grow_b1 =
      get_coef(
        mod_gr_rs,
        "logsize_t0"),
    grow_b2 =
      get_coef(
        mod_gr_rs,
        "logsize_t0_2"),
    grow_b3 =
      get_coef(
        mod_gr_rs,
        "logsize_t0_3"),
    a =
      unname(
        coef(
          mod_gr_var_rs)[["a"]]),
    b =
      unname(
        coef(
          mod_gr_var_rs)[["b"]]),

    dorm_b0 =
      get_coef(
        mod_do_rs,
        "(Intercept)"),
    dorm_b1 =
      get_coef(
        mod_do_rs,
        "logsize_t0"),
    dorm_b2 =
      get_coef(
        mod_do_rs,
        "logsize_t0_2"),
    dorm_b3 =
      get_coef(
        mod_do_rs,
        "logsize_t0_3"),

    fl_b0 =
      get_coef(
        mod_fl_rs,
        "(Intercept)"),
    fl_b1 =
      get_coef(
        mod_fl_rs,
        "logsize_t0"),
    fl_b2 =
      get_coef(
        mod_fl_rs,
        "logsize_t0_2"),
    fl_b3 =
      get_coef(
        mod_fl_rs,
        "logsize_t0_3"),

    fln_b0 =
      get_coef(
        mod_fln_rs,
        "(Intercept)"),
    fln_b1 =
      get_coef(
        mod_fln_rs,
        "logsize_t0"),
    fln_b2 =
      get_coef(
        mod_fln_rs,
        "logsize_t0_2"),
    fln_b3 =
      get_coef(
        mod_fln_rs,
        "logsize_t0_3"),

    recr_sz = recr_sz_rs,
    recr_sd = recr_sd_rs,
    react_sz = react_sz_rs,
    react_sd = react_sd_rs,
    p_reactivate = p_reactivate_rs,

    L =
      min(
        df_gr_rs$logsize_t0),
    U =
      max(
        df_gr_rs$logsize_t0),
    mat_siz = 200,

    mod_su_index =
      fit_su_rs$index,
    mod_gr_index =
      fit_gr_rs$index,
    mod_do_index =
      fit_do_rs$index,
    mod_fl_index =
      fit_fl_rs$index,
    mod_fl_n_index =
      fit_fln_rs$index))


# Reference population counts -------------------------------------------------
df_counts_alive_rs <- df_ind_recruit_reference %>%
  filter(
    state %in%
      c(
        "active",
        "dormant",
        "alive_or_dormant")) %>%
  count(
    site, pop, qu, year,
    name = "n_alive")


last_complete_transition_rs <-
  max_ind_year -
  reference_cutoff -
  1


# One recruitment-treatment scenario -----------------------------------------
run_recruit_scenario <- function(scenario) {

  # Assign each observed entry to a recruitment cohort, or leave it as an
  # observed adult only.
  df_events_i <- df_recruit_first_sens %>%
    filter(
      !baseline) %>%
    mutate(
      fecundity_year = case_when(

        recruit_class ==
          "observed_seedling" ~
          recruit_year - 1,

        scenario ==
          "Current: all new adults = year-2 recruits" &
          recruit_type ==
          "new_adult" ~
          recruit_year - 2,

        scenario ==
          "Stage-2 only = year-2 recruits" &
          recruit_class ==
          "presumed_2yr" ~
          recruit_year - 2,

        scenario ==
          "All new adults = year-1 recruits" &
          recruit_type ==
          "new_adult" ~
          recruit_year - 1,

        scenario ==
          "Stage-2 only = year-1 recruits" &
          recruit_class ==
          "presumed_2yr" ~
          recruit_year - 1,

        TRUE ~ NA_real_),

      backfill_previous_year = case_when(

        scenario ==
          "Current: all new adults = year-2 recruits" &
          recruit_type ==
          "new_adult" ~
          TRUE,

        scenario ==
          "Stage-2 only = year-2 recruits" &
          recruit_class ==
          "presumed_2yr" ~
          TRUE,

        TRUE ~ FALSE))


  # Scenarios using year-2 reconstruction require monitoring through t + 2.
  requires_t2 <- scenario %in% c(
    "Current: all new adults = year-2 recruits",
    "Stage-2 only = year-2 recruits")


  # Count reconstructed recruits by fecundity year ----------------------------
  df_recruit_count_i <- df_events_i %>%
    filter(
      !is.na(fecundity_year)) %>%
    count(
      site,
      pop,
      qu,
      fecundity_year,
      name = "recruits")


  # Construct complete quadrat-year recruitment data --------------------------
  df_recruit_quad_i <- df_recruit_followup %>%
    mutate(
      recruitment_complete =
        monitored_t1 &
        if_else(
          requires_t2,
          monitored_t2,
          TRUE)) %>%
    filter(
      recruitment_complete) %>%
    left_join(
      df_recruit_count_i,
      by = c(
        "site",
        "pop",
        "qu",
        "year" =
          "fecundity_year")) %>%
    left_join(
      df_repro_quad %>%
        dplyr::select(
          site,
          pop,
          qu,
          year,
          nr_scapes),
      by = c(
        "site",
        "pop",
        "qu",
        "year")) %>%
    mutate(
      recruits =
        replace_na(
          recruits,
          0L))


  # Recruitment model ---------------------------------------------------------
  df_sc2re_i <- df_recruit_quad_i %>%
    filter(
      !is.na(nr_scapes)) %>%
    group_by(
      site,
      pop,
      year) %>%
    summarise(
      total_scapes =
        sum(nr_scapes),
      recruit_count =
        sum(recruits),
      .groups = "drop")


  mods_re_i <- list(
    MASS::glm.nb(
      recruit_count ~ 1,
      data = df_sc2re_i),

    MASS::glm.nb(
      recruit_count ~ total_scapes,
      data = df_sc2re_i),

    MASS::glm.nb(
      recruit_count ~
        log1p(total_scapes),
      data = df_sc2re_i))


  re_daicc_i <- get_daicc(
    mods_re_i)

  re_model_index_i <-
    which.min(
      re_daicc_i)

  mod_re_i <-
    mods_re_i[[
      re_model_index_i]]


  df_sc2re_i <- df_sc2re_i %>%
    mutate(
      recruits_pred =
        predict(
          mod_re_i,
          type = "response"))


  recr_per_scape_i <-
    sum(
      df_sc2re_i$recruits_pred) /
    sum(
      df_sc2re_i$total_scapes)


  # IPM lambda ----------------------------------------------------------------
  pars_i <-
    pars_recruit_base

  pars_i$fecu_b0 <-
    recr_per_scape_i

  lambda_ipm_i <-
    lambda_ipm(
      pars_i)


  # Scenario-specific observed population counts ------------------------------
  # Backfilling is used only when a new adult is assumed to have already been
  # present, but unobserved, in the preceding annual census.

  df_backfill_i <- df_events_i %>%
    filter(
      backfill_previous_year) %>%
    transmute(
      site,
      pop,
      qu,
      year =
        recruit_year - 1) %>%
    count(
      site,
      pop,
      qu,
      year,
      name =
        "n_backfill")


  df_counts_quad_i <- df_quad_monitor %>%
    left_join(
      df_counts_alive_rs,
      by = c(
        "site",
        "pop",
        "qu",
        "year")) %>%
    left_join(
      df_backfill_i,
      by = c(
        "site",
        "pop",
        "qu",
        "year")) %>%
    mutate(
      n_alive =
        replace_na(
          n_alive,
          0L),
      n_backfill =
        replace_na(
          n_backfill,
          0L),
      n =
        n_alive +
        n_backfill) %>%
    group_by(
      site,
      pop,
      qu) %>%
    arrange(
      year,
      .by_group = TRUE) %>%
    mutate(
      year_t1 =
        lead(year),
      n_t1 =
        lead(n),
      annual_transition =
        year_t1 ==
        year + 1) %>%
    ungroup() %>%
    filter(
      annual_transition,
      year <=
        last_complete_transition_rs)


  df_counts_year_i <- df_counts_quad_i %>%
    group_by(year) %>%
    summarise(
      n_t0 =
        sum(n),
      n_t1 =
        sum(n_t1),
      .groups = "drop") %>%
    filter(
      n_t0 > 0) %>%
    mutate(
      lambda_obs =
        n_t1 / n_t0)


  lambda_obs_arithmetic_i <-
    mean(
      df_counts_year_i$lambda_obs,
      na.rm = TRUE)

  lambda_obs_geometric_i <-
    exp(
      mean(
        log(
          df_counts_year_i$lambda_obs),
        na.rm = TRUE))


  # Scenario summary ----------------------------------------------------------
  tibble(
    Scenario = scenario,

    `Recruit events used` =
      sum(
        !is.na(
          df_events_i$fecundity_year)),

    `Presumed 2yr used` =
      sum(
        df_events_i$recruit_class ==
          "presumed_2yr" &
          !is.na(
            df_events_i$fecundity_year)),

    `Other new adults used` =
      sum(
        df_events_i$recruit_class ==
          "new_adult_other" &
          !is.na(
            df_events_i$fecundity_year)),

    `Recruitment model` =
      re_model_index_i - 1,

    `Recruits per scape` =
      recr_per_scape_i,

    lambda_ipm =
      lambda_ipm_i,

    lambda_obs_arithmetic =
      lambda_obs_arithmetic_i,

    lambda_obs_geometric =
      lambda_obs_geometric_i)
}


# Run sensitivity --------------------------------------------------------------
df_recruit_sensitivity <- map_dfr(
  recruit_scenarios,
  run_recruit_scenario)


# Difference from current reconstruction --------------------------------------
ref_recruit <- df_recruit_sensitivity %>%
  filter(
    Scenario ==
      recruit_scenarios[1]) %>%
  transmute(
    ref_recr =
      `Recruits per scape`,
    ref_ipm =
      lambda_ipm,
    ref_obs_arithmetic =
      lambda_obs_arithmetic,
    ref_obs_geometric =
      lambda_obs_geometric)


df_recruit_sensitivity <- df_recruit_sensitivity %>%
  crossing(
    ref_recruit) %>%
  mutate(
    delta_recruits_per_scape =
      `Recruits per scape` -
      ref_recr,

    delta_ipm =
      lambda_ipm -
      ref_ipm,

    delta_obs_arithmetic =
      lambda_obs_arithmetic -
      ref_obs_arithmetic,

    delta_obs_geometric =
      lambda_obs_geometric -
      ref_obs_geometric) %>%
  dplyr::select(
    -starts_with("ref_"))


# Figure -----------------------------------------------------------------------
df_recruit_sensitivity_plot <- df_recruit_sensitivity %>%
  dplyr::select(
    Scenario,
    `Mean IPM` =
      lambda_ipm,
    `Observed arithmetic` =
      lambda_obs_arithmetic,
    `Observed geometric` =
      lambda_obs_geometric) %>%
  pivot_longer(
    -Scenario,
    names_to = "Estimate",
    values_to = "lambda") %>%
  mutate(
    Scenario = factor(
      Scenario,
      levels =
        recruit_scenarios))


fig_recruit_sensitivity <- ggplot(
  df_recruit_sensitivity_plot,
  aes(
    x = Scenario,
    y = lambda,
    shape = Estimate)) +
  geom_hline(
    yintercept = 1,
    linetype = "dashed") +
  geom_point(
    size = 2.8,
    position =
      position_dodge(
        width = 0.4)) +
  theme_bw() +
  labs(
    x = NULL,
    y = expression(lambda),
    shape = NULL) +
  theme(
    axis.text.x =
      element_text(
        angle = 25,
        hjust = 1))
```


The scenarios differ only in how first-record new adults contribute to
recruitment and whether their inferred presence is backfilled into the preceding
population census. Observed seedlings are treated identically in every
scenario, and all directly observed new adults remain in the demographic data.


```{r recruit-sensitivity-table, echo=FALSE}
df_recruit_sensitivity %>%
  mutate(
    across(
      c(
        `Recruits per scape`,
        lambda_ipm,
        lambda_obs_arithmetic,
        lambda_obs_geometric,
        delta_recruits_per_scape,
        delta_ipm,
        delta_obs_arithmetic,
        delta_obs_geometric),
      ~ round(.x, 4))) %>%
  knitr::kable()
```


```{r recruit-sensitivity-figure, echo=FALSE, fig.width=9, fig.height=5.5}
fig_recruit_sensitivity
```


The comparison separates two uncertainties. Contrasting the current
reconstruction with the stage-2-only scenario tests whether the inference should
be restricted to the new adults whose stage and size are most consistent with
second-year recruits. The "new adults enter only as adults" scenario provides
the conservative alternative in which no new adult is used to infer recruitment
before its first observation. Finally, the year-1 scenarios test whether the
results depend specifically on assigning new adults to a two-year rather than a
one-year recruitment lag.
