# Mean IPM {#ipm_mean}

## Introduction

This analysis develops a mean integral projection model (IPM) for
*Eriogonum longifolium* var. *gnaphalifolium* using the long-term demographic
data from Archbold Biological Station and Carter Creek. The purpose of this
first IPM is to provide a relatively simple description of the principal
demographic processes and their consequences for long-term population growth.
Vital rates are therefore pooled across years, populations, and disturbance
histories rather than allowing demographic relationships to vary through time
or among sites. The resulting kernel represents the demographic conditions
averaged across the observations contained in the dataset and should not be
interpreted specifically as an unburned or post-fire kernel.

The population is represented by two demographic states. Above-ground plants
are described continuously by log-transformed basal rosette diameter, whereas
dormant plants form one additional discrete state. With 200 continuous size
classes, the complete projection kernel therefore contains 201 states. This
structure retains the principal advantage of an IPM, namely continuous
size-dependent demographic rates, while explicitly accommodating prolonged
below-ground dormancy, which is an important feature of the life history of
*E. longifolium* var. *gnaphalifolium*.

The mean kernel can be written conceptually as

$$
\mathbf{K} =
\begin{pmatrix}
\mathbf{T} + \mathbf{F} & \mathbf{R} \\
\mathbf{D}              & P_{DD}
\end{pmatrix},
$$

where $\mathbf{T}$ describes survival and growth of active plants,
$\mathbf{F}$ describes recruitment from active plants, $\mathbf{D}$ describes
entry into dormancy, $\mathbf{R}$ describes reactivation into the continuous
active-size distribution, and $P_{DD}$ is the probability that a dormant plant
remains dormant.

### Survival, growth, and entry into dormancy

Survival is modeled for plants that are active at census $t$. Candidate
binomial models contain an intercept and linear, quadratic, or cubic functions
of log basal diameter, with model complexity selected using AICc. Survival
includes plants that remain active as well as plants that subsequently enter
dormancy. Consequently, dormancy is not treated as mortality.

Growth is estimated only from transitions in which a plant is active at both
$t$ and $t+1$. Log diameter at $t+1$ is modeled as an intercept, linear,
quadratic, or cubic function of log diameter at $t$, again with model
complexity selected by AICc. Growth variance is heteroscedastic. Squared
residuals from the selected growth model are modeled as an exponential
function of the fitted growth mean, so that the standard deviation used in the
growth kernel varies with expected size at $t+1$.

Entry into dormancy is estimated separately for active plants known to have
survived the annual transition. A binomial model describes the conditional
probability that a surviving active individual enters dormancy as a function
of its size. This conditional formulation is important for the construction of
the kernel. For an active plant of size $x$, the probability of remaining in
the continuous active state is

$$
S(x)[1-D(x)],
$$

where $S(x)$ is survival and $D(x)$ is the probability of entering dormancy
conditional on survival. The probability of moving from the active state into
the dormant state is correspondingly

$$
S(x)D(x).
$$

Thus, survival is first allocated between active and dormant destinations
rather than dormancy being modeled as an additional loss from the population.

### Dormancy and reactivation

Dormant plants have no observable above-ground size, so their subsequent
reactivation cannot be modeled as a function of current size. Reactivation is
therefore represented by an intercept-only binomial model that estimates one
mean annual probability of returning to the active state. Plants that
reactivate are distributed across the continuous size axis according to the
observed log-diameter distribution of plants when they reappear above ground.

No independent mortality probability is estimated for the dormant state in
this first mean IPM. A dormant individual therefore either reactivates with
probability $p_R$ or remains dormant with probability $1-p_R$. The dormant
column of the kernel consequently sums to one. This is a simplifying
assumption arising from the difficulty of distinguishing mortality during
dormancy from continued unobserved dormancy in the census data.

### Flowering and flowering-scape production

Reproduction is divided into two sequential vital rates. First, the
probability that an active individual flowers is modeled with a binomial
regression as a function of log diameter. Second, among flowering individuals,
the number of flowering scapes is modeled using a negative-binomial
regression. Intercept-only, linear, quadratic, and cubic size relationships
are considered for both components where appropriate and selected using AICc.

The expected number of flowering scapes produced by an active individual of
size $x$ is therefore

$$
P_F(x)N_F(x),
$$

where $P_F(x)$ is its probability of flowering and $N_F(x)$ is its expected
number of scapes conditional on flowering. This decomposition avoids treating
the large number of non-flowering individuals as structural zeros in the
scape-number model.

### Recruitment

Recruitment requires additional assumptions because newly observed plants
cannot always be assigned unambiguously to a particular reproductive event.
Observed seedlings are treated directly as recruits. In addition, for this
simple mean IPM, all post-baseline individuals first observed as new adults
are assumed to have been seedlings that were missed during the preceding
census. These individuals are therefore back-assigned to the preceding
recruitment cohort. This follows the general rationale used previously for
presumed two-year-old plants in this species, but represents a deliberately
inclusive recruitment assumption.

Recruitment abundance is related to flowering-scape abundance at the
population-year level using a negative-binomial model. The selected model
allows expected recruitment to increase sublinearly with the number of
contemporaneous flowering scapes,

$$
\log E(R) =
\beta_0 +
\beta_1\log(1 + N_{\mathrm{scape}}).
$$

The non-zero intercept allows recruitment to occur even when no flowering
scapes are recorded in the current census. It should therefore be interpreted
as recruitment not explained by contemporaneous observed flowering, which may
include reproductive carry-over, incomplete observation of reproductive
output, dispersal, or other unmeasured processes.

This population-level recruitment relationship cannot be inserted directly
into a conventional linear individual-based IPM because it contains both an
intercept and a nonlinear dependence on total population-level scape
abundance. For the mean IPM, the fitted recruitment relationship is therefore
collapsed to a single overall fitted recruit-to-scape conversion. This scalar
is used only to construct the linear fertility kernel and represents an
average conversion over the demographic conditions observed in the dataset.

The size distribution of recruits is estimated separately using individuals
actually observed as seedlings. New adults contribute to reconstructed
recruit abundance but do not determine the recruit-size distribution. All
recruits entering the IPM are consequently distributed across the continuous
size axis according to the observed log-diameter distribution of true
seedlings.

### Fertility and construction of the complete kernel

For an active individual of size $x$, the fertility contribution to recruits
of size $y$ is

$$
F(y,x) =
P_F(x)
N_F(x)
c_R
f_R(y),
$$

where $c_R$ is the mean fitted recruitment conversion and $f_R(y)$ is the
recruit-size distribution. Fertility is not multiplied by survival because
reproduction during the census interval can contribute recruits at $t+1$
independently of whether the parent subsequently survives to that census.

The active-to-active transition kernel is

$$
T(y,x) =
S(x)
[1-D(x)]
G(y,x),
$$

where $G(y,x)$ is the size-transition distribution. Growth probabilities that
fall beyond the lower or upper integration limits are returned to the nearest
boundary size class to avoid loss of probability mass through eviction.
Recruit and reactivation size distributions are similarly normalized over the
modeled size interval.

The complete kernel therefore combines continuous active-plant dynamics with
one discrete dormant state. Active individuals may remain active, enter
dormancy, or contribute new recruits; dormant individuals may reactivate into
the continuous size distribution or remain dormant.

Because all vital-rate relationships are pooled across years, the resulting
kernel is a single time-invariant mean kernel. Its dominant eigenvalue is the
asymptotic population growth rate associated with these mean demographic
conditions. The current model gives an asymptotic growth rate of approximately

$$
\lambda = 0.990,
$$

indicating a slowly declining population under the demographic conditions
represented by the mean IPM. This is close to the reconstructed realized
long-term population trajectory, for which the geometric mean annual growth
rate is approximately $0.998$. The two quantities are not expected to be
identical: the first is the asymptotic growth rate of a population subjected
repeatedly to one mean kernel, whereas the second reflects the sequence of
annual population changes actually observed.



## Specifications
### Vital model leaverage
```{r}
v_mod_set_su   <- c()
v_mod_set_gr   <- c()
v_mod_set_do   <- c()
v_mod_set_fl   <- c()
v_mod_set_fl_n <- c()
```

## Data
```{r}
df <- read.csv(
  file.path(
    dir_data,
    paste0("ab_", v_sp_abb, "_df_workdata_260820.csv"))) %>%
  mutate(
    year = as.numeric(year),
    row_type = as.character(row_type))

df_ind <- df %>%
  filter(row_type == "individual")

df_re <- df %>%
  filter(
    row_type == "recruitment",
    recruitment_complete,
    !is.na(recruits_simple))
```

## Vital rates
### Survival
#### Data
```{r}
df_su <- df %>%
  filter(
    state == "active",
    !is.na(survives),
    size_t0 > 0,
    is.finite(logsize_t0)) %>%
  dplyr::select(
    id, year, size_t0, survives,
    logsize_t0, logsize_t0_2, logsize_t0_3)
```

#### Model
```{r}
mod_su_0 <- glm(
  survives ~ 1,
  data = df_su, family = "binomial")

mod_su_1 <- glm(
  survives ~ logsize_t0,
  data = df_su, family = "binomial")

mod_su_2 <- glm(
  survives ~ logsize_t0 + logsize_t0_2,
  data = df_su, family = "binomial")

mod_su_3 <- glm(
  survives ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_su, family = "binomial")

mods_su <- list(
  mod_su_0, mod_su_1, mod_su_2, mod_su_3)

mods_su_dAICc <- AICctab(
  mods_su, weights = TRUE, sort = FALSE)$dAICc

mods_su_sorted <- order(mods_su_dAICc)

if (length(v_mod_set_su) == 0) {
  mod_su_index_bestfit <- mods_su_sorted[1]
  v_mod_su_index <- mod_su_index_bestfit - 1
} else {
  mod_su_index_bestfit <- v_mod_set_su + 1
  v_mod_su_index <- v_mod_set_su
}

mod_su_bestfit <- mods_su[[mod_su_index_bestfit]]

mod_su_bestfit
mods_su_dAICc
```

#### Plot
```{r}
df_su_newdata <- df_su %>%
  summarise(
    x = list(seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)),
    .groups = "drop") %>%
  unnest(x) %>%
  mutate(
    logsize_t0 = x,
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    predicted = predict(
      mod_su_bestfit,
      newdata = .,
      type = "response"))

fig_su_line <- ggplot(
  df_su,
  aes(x = logsize_t0, y = survives)) +
  geom_jitter(
    height = 0.05, width = 0,
    alpha = 0.3) +
  geom_line(
    data = df_su_newdata,
    aes(y = predicted),
    linewidth = 1) +
  labs(
    title = "Survival probability by size",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = "Probability of survival") +
  theme_bw()


df_su_bindata <- plot_binned_prop(
  df_su, 10, logsize_t0, survives)

df_su_pred <- df_su %>%
  reframe(
    logsize_t0 = seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    survives = predict(
      mod_su_bestfit,
      newdata = .,
      type = "response"))

fig_su_bin <- ggplot() +
  geom_point(
    data = df_su_bindata,
    aes(x = logsize_t0, y = survives)) +
  geom_errorbar(
    data = df_su_bindata,
    aes(
      x = logsize_t0,
      ymin = lwr,
      ymax = upr),
    width = 0.1) +
  geom_line(
    data = df_su_pred,
    aes(x = logsize_t0, y = survives),
    linewidth = 1.2) +
  labs(
    x = expression("log(diameter)"[t0]),
    y = "") +
  theme_bw() +
  ylim(0, 1)

fig_su <- fig_su_line + fig_su_bin + plot_layout()
fig_su
```

### Growth 
#### Data
```{r}
df_gr <- df %>%
  filter(
    state == "active",
    state_t1 == "active",
    size_t0 > 0,
    size_t1 > 0,
    is.finite(logsize_t0),
    is.finite(logsize_t1)) %>%
  dplyr::select(
    id, year, size_t0, size_t1,
    logsize_t0, logsize_t1,
    logsize_t0_2, logsize_t0_3)
```

#### Model
```{r}
mod_gr_0 <- lm(
  logsize_t1 ~ 1,
  data = df_gr)

mod_gr_1 <- lm(
  logsize_t1 ~ logsize_t0,
  data = df_gr)

mod_gr_2 <- lm(
  logsize_t1 ~ logsize_t0 + logsize_t0_2,
  data = df_gr)

mod_gr_3 <- lm(
  logsize_t1 ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_gr)

mods_gr <- list(
  mod_gr_0, mod_gr_1, mod_gr_2, mod_gr_3)

mods_gr_dAICc <- AICctab(
  mods_gr, weights = TRUE, sort = FALSE)$dAICc

mods_gr_sorted <- order(mods_gr_dAICc)

if (length(v_mod_set_gr) == 0) {
  mod_gr_index_bestfit <- mods_gr_sorted[1]
  v_mod_gr_index <- mod_gr_index_bestfit - 1
} else {
  mod_gr_index_bestfit <- v_mod_set_gr + 1
  v_mod_gr_index <- v_mod_set_gr
}

mod_gr_bestfit <- mods_gr[[mod_gr_index_bestfit]]

mod_gr_bestfit
mods_gr_dAICc
```

#### Plot
```{r}
df_gr_newdata <- tibble(
  logsize_t0 = seq(
    min(df_gr$logsize_t0),
    max(df_gr$logsize_t0),
    length.out = 100)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)

df_gr_newdata <- cbind(
  df_gr_newdata,
  as.data.frame(
    predict(
      mod_gr_bestfit,
      newdata = df_gr_newdata,
      interval = "confidence")))

fig_gr <- ggplot(
  df_gr,
  aes(x = logsize_t0, y = logsize_t1)) +
  geom_point(alpha = 0.4) +
  geom_line(
    data = df_gr_newdata,
    aes(y = fit),
    linewidth = 1) +
  geom_ribbon(
    data = df_gr_newdata,
    aes(y = fit, ymin = lwr, ymax = upr),
    alpha = 0.2) +
  geom_abline(intercept = 0, slope = 1) +
  labs(
    title = "Growth prediction",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = expression("log(diameter)"[t1])) +
  theme_bw()

fig_gr
```

#### Growth variance
```{r}
# Fitted values from growth model
mod_gr_x   <- fitted(mod_gr_bestfit)  
# Squared residuals
mod_gr_y   <- resid(mod_gr_bestfit)^2  
# Non-linear model for variance
mod_gr_var <- nls(
  mod_gr_y ~ a * exp(b * mod_gr_x), start = list(a = 1, b = 0),
  control = nls.control(maxiter = 1000, tol = 1e-6, warnOnly = TRUE)) 
```

### Dormancy entry
#### Data
```{r}
df_do <- df %>%
  filter(
    !is.na(enter_dormancy),
    size_t0 > 0,
    is.finite(logsize_t0)) %>%
  dplyr::select(
    id, year, size_t0, enter_dormancy,
    logsize_t0, logsize_t0_2, logsize_t0_3)
```

#### Model
```{r}
mod_do_0 <- glm(
  enter_dormancy ~ 1,
  data = df_do, family = "binomial")

mod_do_1 <- glm(
  enter_dormancy ~ logsize_t0,
  data = df_do, family = "binomial")

mod_do_2 <- glm(
  enter_dormancy ~ logsize_t0 + logsize_t0_2,
  data = df_do, family = "binomial")

mod_do_3 <- glm(
  enter_dormancy ~
    logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_do, family = "binomial")

mods_do <- list(
  mod_do_0, mod_do_1, mod_do_2, mod_do_3)

mods_do_dAICc <- AICctab(
  mods_do, weights = TRUE, sort = FALSE)$dAICc

mods_do_sorted <- order(mods_do_dAICc)

if (length(v_mod_set_do) == 0) {
  mod_do_index_bestfit <- mods_do_sorted[1]
  v_mod_do_index <- mod_do_index_bestfit - 1
} else {
  mod_do_index_bestfit <- v_mod_set_do + 1
  v_mod_do_index <- v_mod_set_do
}

mod_do_bestfit <- mods_do[[mod_do_index_bestfit]]

mod_do_bestfit
mods_do_dAICc
```

#### Plot
```{r}
df_do_newdata <- df_do %>%
  summarise(
    x = list(seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)),
    .groups = "drop") %>%
  unnest(x) %>%
  mutate(
    logsize_t0 = x,
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    predicted = predict(
      mod_do_bestfit,
      newdata = .,
      type = "response"))

fig_do_line <- ggplot(
  df_do,
  aes(x = logsize_t0, y = enter_dormancy)) +
  geom_jitter(
    height = 0.05, width = 0,
    alpha = 0.3) +
  geom_line(
    data = df_do_newdata,
    aes(y = predicted),
    linewidth = 1) +
  labs(
    title = "Dormancy probability by size",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = "Probability of entering dormancy") +
  theme_bw()


df_do_bindata <- plot_binned_prop(
  df_do, 10, logsize_t0, enter_dormancy)

df_do_pred <- df_do %>%
  reframe(
    logsize_t0 = seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    enter_dormancy = predict(
      mod_do_bestfit,
      newdata = .,
      type = "response"))

fig_do_bin <- ggplot() +
  geom_point(
    data = df_do_bindata,
    aes(x = logsize_t0, y = enter_dormancy)) +
  geom_errorbar(
    data = df_do_bindata,
    aes(
      x = logsize_t0,
      ymin = lwr,
      ymax = upr),
    width = 0.1) +
  geom_line(
    data = df_do_pred,
    aes(x = logsize_t0, y = enter_dormancy),
    linewidth = 1.2) +
  labs(
    x = expression("log(diameter)"[t0]),
    y = "") +
  theme_bw() +
  ylim(0, 1)

fig_do <- fig_do_line + fig_do_bin + plot_layout()
fig_do
```

### Dormancy reactivation - Mean and variance
```{r}
df_ra <- df %>%
  filter(!is.na(reactivate))

mod_ra <- glm(
  reactivate ~ 1,
  data = df_ra, family = "binomial")

p_reactivate <- predict(
  mod_ra,
  newdata = data.frame(x = 1),
  type = "response")[1]

p_reactivate

df_ra_size <- df %>%
  filter(
    reactivate == 1,
    size_reactivate_t1 > 0) %>%
  mutate(
    logsize_reactivate =
      log(size_reactivate_t1))

react_sz <- mean(
  df_ra_size$logsize_reactivate,
  na.rm = TRUE)

react_sd <- sd(
  df_ra_size$logsize_reactivate,
  na.rm = TRUE)
```


### Flowering probability
#### Data
```{r}
df_fl <- df %>%
  filter(
    state == "active",
    !is.na(flower),
    size_t0 > 0,
    is.finite(logsize_t0))
```

#### Model
```{r}
mod_fl_0 <- glm(
  flower ~ 1,
  data = df_fl, family = "binomial")

mod_fl_1 <- glm(
  flower ~ logsize_t0,
  data = df_fl, family = "binomial")

mod_fl_2 <- glm(
  flower ~ logsize_t0 + logsize_t0_2,
  data = df_fl, family = "binomial")

mod_fl_3 <- glm(
  flower ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fl, family = "binomial")

mods_fl <- list(
  mod_fl_0, mod_fl_1, mod_fl_2, mod_fl_3)

mods_fl_dAICc <- AICctab(
  mods_fl, weights = TRUE, sort = FALSE)$dAICc

mods_fl_sorted <- order(mods_fl_dAICc)

if (length(v_mod_set_fl) == 0) {
  mod_fl_index_bestfit <- mods_fl_sorted[1]
  v_mod_fl_index <- mod_fl_index_bestfit - 1
} else {
  mod_fl_index_bestfit <- v_mod_set_fl + 1
  v_mod_fl_index <- v_mod_set_fl
}

mod_fl_bestfit <- mods_fl[[mod_fl_index_bestfit]]

mod_fl_bestfit
mods_fl_dAICc
```

#### Plot
```{r}
df_fl_newdata <- df_fl %>%
  summarise(
    x = list(seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)),
    .groups = "drop") %>%
  unnest(x) %>%
  mutate(
    logsize_t0 = x,
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    predicted = predict(
      mod_fl_bestfit,
      newdata = .,
      type = "response"))

fig_fl_line <- ggplot(
  df_fl,
  aes(x = logsize_t0, y = flower)) +
  geom_jitter(
    height = 0.05, width = 0,
    alpha = 0.3) +
  geom_line(
    data = df_fl_newdata,
    aes(y = predicted),
    linewidth = 1) +
  labs(
    title = "Flowering probability by size",
    subtitle = v_ggp_suffix,
    x = expression("log(diameter)"[t0]),
    y = "Probability of flowering") +
  theme_bw()


df_fl_bindata <- plot_binned_prop(
  df_fl, 10, logsize_t0, flower)

df_fl_pred <- df_fl %>%
  reframe(
    logsize_t0 = seq(
      min(logsize_t0),
      max(logsize_t0),
      length.out = 100)) %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3) %>%
  mutate(
    flower = predict(
      mod_fl_bestfit,
      newdata = .,
      type = "response"))

fig_fl_bin <- ggplot() +
  geom_point(
    data = df_fl_bindata,
    aes(x = logsize_t0, y = flower)) +
  geom_errorbar(
    data = df_fl_bindata,
    aes(
      x = logsize_t0,
      ymin = lwr,
      ymax = upr),
    width = 0.1) +
  geom_line(
    data = df_fl_pred,
    aes(x = logsize_t0, y = flower),
    linewidth = 1.2) +
  labs(
    x = expression("log(diameter)"[t0]),
    y = "") +
  theme_bw() +
  ylim(0, 1)

fig_fl <- fig_fl_line + fig_fl_bin + plot_layout()
fig_fl
```

### Number of scapes conditional on flowering
#### Data
```{r}
df_fl_cond <- df_fl %>%
  filter(
    flower == 1,
    !is.na(fl_nr),
    fl_nr > 0,
    fl_nr %% 1 == 0)
```

#### Models
```{r}
mod_fl_n_0 <- glm.nb(
  fl_nr ~ 1,
  data = df_fl_cond)

mod_fl_n_1 <- glm.nb(
  fl_nr ~ logsize_t0,
  data = df_fl_cond)

mod_fl_n_2 <- glm.nb(
  fl_nr ~ logsize_t0 + logsize_t0_2,
  data = df_fl_cond)

mod_fl_n_3 <- glm.nb(
  fl_nr ~ logsize_t0 + logsize_t0_2 + logsize_t0_3,
  data = df_fl_cond)

mods_fl_n <- list(
  mod_fl_n_0, mod_fl_n_1, mod_fl_n_2, mod_fl_n_3)

mods_fl_n_dAICc <- AICctab(
  mods_fl_n, weights = TRUE, sort = FALSE)$dAICc

mods_fl_n_sorted <- order(mods_fl_n_dAICc)

if (length(v_mod_set_fl_n) == 0) {
  mod_fl_n_index_bestfit <- mods_fl_n_sorted[1]
  v_mod_fl_n_index <- mod_fl_n_index_bestfit - 1
} else {
  mod_fl_n_index_bestfit <- v_mod_set_fl_n + 1
  v_mod_fl_n_index <- v_mod_set_fl_n
}

mod_fl_n_bestfit <- mods_fl_n[[mod_fl_n_index_bestfit]]

mod_fl_n_bestfit
mods_fl_n_dAICc
```

#### Plots
```{r}
# Predictions for flower number
# Create prediction grid
df_fl_n_pred <- expand.grid(
  logsize_t0 = seq(
    min(df_fl_cond$logsize_t0),
    max(df_fl_cond$logsize_t0),
    length.out = 100))

df_fl_n_pred <- df_fl_n_pred %>%
  mutate(
    logsize_t0_2 = logsize_t0^2,
    logsize_t0_3 = logsize_t0^3)

# Predict
df_fl_n_pred$fl_nr <- predict(
  mod_fl_n_bestfit,
  newdata = df_fl_n_pred,
  type = "response")


# Binned observed data
df_fl_n_binned <- df_fl_cond %>%
  mutate(bin = cut(logsize_t0, breaks = 10)) %>%
  group_by(bin) %>%
  summarise(
    logsize_t0 = mean(logsize_t0, na.rm = TRUE),
    fl_nr = mean(fl_nr, na.rm = TRUE),
    se = sd(fl_nr, na.rm = TRUE) / sqrt(n()),
    .groups = "drop") %>%
  mutate(
    lwr = fl_nr - 1.96 * se,
    upr = fl_nr + 1.96 * se)


# Flower number plots
# Plot 1: Raw jitter + prediction
fig_fl_n_line <- ggplot() +
  geom_jitter(
    data = df_fl_cond,
    aes(x = logsize_t0, y = fl_nr),
    alpha = 0.25, width = 0.08, height = 0.3) +
  geom_line(
    data = df_fl_n_pred,
    aes(x = logsize_t0, y = fl_nr),
    linewidth = 0.9) +
  theme_bw() +
  labs(
    title = NULL,
    x = expression("log(diameter)"[t0]),
    y = "Number of flowering scapes")


# Plot 2: Binned + prediction
fig_fl_n_bin <- ggplot() +
  geom_point(
    data = df_fl_n_binned,
    aes(x = logsize_t0, y = fl_nr)) +
  geom_errorbar(
    data = df_fl_n_binned,
    aes(x = logsize_t0, ymin = lwr, ymax = upr),
    width = 0.2) +
  geom_line(
    data = df_fl_n_pred,
    aes(x = logsize_t0, y = fl_nr),
    linewidth = 0.9) +
  theme_bw() +
  labs(
    title = NULL,
    x = expression("log(diameter)"[t0]),
    y = "Number of flowering scapes")


# Combine
fig_fl_n <- fig_fl_n_line + fig_fl_n_bin +
  plot_annotation(
    title = "Flower number",
    subtitle = v_ggp_suffix,
    theme = theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10, face = "italic")))

fig_fl_n
```

### Recruitment
#### Data
```{r}
df_sc2re <- df_re %>%
  filter(
    !is.na(nr_scapes),
    !is.na(recruits_simple)) %>%
  group_by(site, pop, year) %>%
  summarise(
    total_scapes = sum(nr_scapes),
    recruit_count = sum(recruits_simple),
    .groups = "drop")

df_sc2re_mod <- df_sc2re %>%
  filter(
    !is.na(total_scapes),
    !is.na(recruit_count))
```

#### Model
```{r}
# Recruitment can occur without current flowering, consistent with recruitment
# from previously accumulated seed.

mod_re_0 <- glm.nb(
  recruit_count ~ 1,
  data = df_sc2re_mod)

mod_re_sc <- glm.nb(
  recruit_count ~ total_scapes,
  data = df_sc2re_mod)

mod_re_logsc <- glm.nb(
  recruit_count ~ log1p(total_scapes),
  data = df_sc2re_mod)

mods_re <- list(
  mod_re_0,
  mod_re_sc,
  mod_re_logsc)

mods_re_dAICc <- AICctab(
  mods_re,
  weights = TRUE,
  sort = FALSE)$dAICc

mods_re_sorted <- order(mods_re_dAICc)
mod_re_bestfit <- mods_re[[mods_re_sorted[1]]]

mods_re_dAICc
summary(mod_re_bestfit)
```

#### Mean recruitment conversion for the IPM
```{r}
# The recruitment model includes baseline recruitment, which can represent
# recruitment from the seedbank. For the linear mean IPM, collapse the fitted
# recruitment relationship to one mean recruits-per-scape conversion.

df_sc2re_mod <- df_sc2re_mod %>%
  mutate(
    recruits_pred = predict(
      mod_re_bestfit,
      type = "response"))

recr_per_scape <- sum(df_sc2re_mod$recruits_pred) /
  sum(df_sc2re_mod$total_scapes)

recr_per_scape
```

#### Plot
```{r}
df_sc2re_pred <- data.frame(
  total_scapes = seq(
    min(df_sc2re_mod$total_scapes),
    max(df_sc2re_mod$total_scapes),
    length.out = 100))

mod_sc2re_preds <- predict(
  mod_re_bestfit,
  newdata = df_sc2re_pred,
  type = "link",
  se.fit = TRUE)

df_sc2re_pred <- df_sc2re_pred %>%
  mutate(
    fit = exp(mod_sc2re_preds$fit),
    lower = exp(
      mod_sc2re_preds$fit -
        1.96 * mod_sc2re_preds$se.fit),
    upper = exp(
      mod_sc2re_preds$fit +
        1.96 * mod_sc2re_preds$se.fit))

fig_sc2re <- ggplot(
  df_sc2re_mod,
  aes(x = total_scapes, y = recruit_count)) +
  geom_jitter(
    height = 0.2, width = 0.5,
    alpha = 0.4) +
  geom_ribbon(
    data = df_sc2re_pred,
    aes(
      x = total_scapes,
      ymin = lower,
      ymax = upper),
    inherit.aes = FALSE,
    alpha = 0.2) +
  geom_line(
    data = df_sc2re_pred,
    aes(x = total_scapes, y = fit),
    inherit.aes = FALSE,
    linewidth = 1.2) +
  labs(
    title = "Recruits t1 by flowering scapes t0",
    subtitle = v_ggp_suffix,
    x = expression("Total flowering scapes "[t0]),
    y = expression("Number of recruits "[t1])) +
  theme_bw()

fig_sc2re
```

### Recruit size and variance
```{r}
df_re_size <- df_ind %>%
  filter(
    recruit_type == "seedling",
    size_t0 > 0,
    is.finite(logsize_t0))

recr_sz <- mean(
  df_re_size$logsize_t0,
  na.rm = TRUE)

recr_sd <- sd(
  df_re_size$logsize_t0,
  na.rm = TRUE)
```

## Extracting parameter estimates
```{r}
# Survival
coef_su <- data.frame(
  coefficient = names(coef(mod_su_bestfit)),
  value = coef(mod_su_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Growth
coef_gr_fe <- data.frame(
  coefficient = names(coef(mod_gr_bestfit)),
  value = coef(mod_gr_bestfit))

coef_gr_var <- data.frame(
  coefficient = names(coef(mod_gr_var)),
  value = coef(mod_gr_var))

coef_gr <- bind_rows(coef_gr_fe, coef_gr_var) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Dormancy entry
coef_do <- data.frame(
  coefficient = names(coef(mod_do_bestfit)),
  value = coef(mod_do_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Flowering probability
coef_fl <- data.frame(
  coefficient = names(coef(mod_fl_bestfit)),
  value = coef(mod_fl_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

# Scape number conditional on flowering
coef_fln <- data.frame(
  coefficient = names(coef(mod_fl_n_bestfit)),
  value = coef(mod_fl_n_bestfit)) %>%
  mutate(
    coefficient = as.character(coefficient),
    coefficient = replace(
      coefficient, grepl("Intercept", coefficient), "b0"))

extr_value <- function(x, field) {
  subset(x, coefficient == field)$value
}
```

### Constants
```{r}
coef_misc <- data.frame(
  coefficient = c(
    "recr_sz",
    "recr_sd",
    "react_sz",
    "react_sd",
    "p_reactivate",
    "fecu_b0",
    "max_siz",
    "min_siz"),
  value = c(
    recr_sz,
    recr_sd,
    react_sz,
    react_sd,
    p_reactivate,
    recr_per_scape,
    max(df_gr$logsize_t0),
    min(df_gr$logsize_t0)))
```

### Parameter list
```{r}
pars <- Filter(function(x) length(x) > 0, list(
  prefix = v_script_prefix,
  species = v_species,
  
  surv_b0 = extr_value(coef_su, "b0"),
  surv_b1 = extr_value(coef_su, "logsize_t0"),
  surv_b2 = extr_value(coef_su, "logsize_t0_2"),
  surv_b3 = extr_value(coef_su, "logsize_t0_3"),
  
  grow_b0 = extr_value(coef_gr, "b0"),
  grow_b1 = extr_value(coef_gr, "logsize_t0"),
  grow_b2 = extr_value(coef_gr, "logsize_t0_2"),
  grow_b3 = extr_value(coef_gr, "logsize_t0_3"),
  a = extr_value(coef_gr, "a"),
  b = extr_value(coef_gr, "b"),
  
  dorm_b0 = extr_value(coef_do, "b0"),
  dorm_b1 = extr_value(coef_do, "logsize_t0"),
  dorm_b2 = extr_value(coef_do, "logsize_t0_2"),
  dorm_b3 = extr_value(coef_do, "logsize_t0_3"),
  
  fl_b0 = extr_value(coef_fl, "b0"),
  fl_b1 = extr_value(coef_fl, "logsize_t0"),
  fl_b2 = extr_value(coef_fl, "logsize_t0_2"),
  fl_b3 = extr_value(coef_fl, "logsize_t0_3"),
  
  fln_b0 = extr_value(coef_fln, "b0"),
  fln_b1 = extr_value(coef_fln, "logsize_t0"),
  fln_b2 = extr_value(coef_fln, "logsize_t0_2"),
  fln_b3 = extr_value(coef_fln, "logsize_t0_3"),
  
  fecu_b0 = extr_value(coef_misc, "fecu_b0"),
  recr_sz = extr_value(coef_misc, "recr_sz"),
  recr_sd = extr_value(coef_misc, "recr_sd"),
  react_sz = extr_value(coef_misc, "react_sz"),
  react_sd = extr_value(coef_misc, "react_sd"),
  p_reactivate = extr_value(coef_misc, "p_reactivate"),
  
  L = extr_value(coef_misc, "min_siz"),
  U = extr_value(coef_misc, "max_siz"),
  mat_siz = 200,
  
  mod_su_index = v_mod_su_index,
  mod_gr_index = v_mod_gr_index,
  mod_do_index = v_mod_do_index,
  mod_fl_index = v_mod_fl_index,
  mod_fl_n_index = v_mod_fl_n_index))
```


## Building IPM - Functions
```{r}
inv_logit <- function(x) {
  exp(x) / (1 + exp(x))
}


# Survival ---------------------------------------------------------------------
sx <- function(x, pars, num_pars = pars$mod_su_index) {
  val <- pars$surv_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("surv_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  inv_logit(val)
}


# Growth -----------------------------------------------------------------------
grow_mu <- function(x, pars, num_pars = pars$mod_gr_index) {
  val <- pars$grow_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("grow_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  val
}


# Growth variation -------------------------------------------------------------
grow_sd <- function(mu, pars) {
  sqrt(pars$a * exp(pars$b * mu))
}


# Growth transition ------------------------------------------------------------
gxy <- function(x, y, pars) {
  mu <- grow_mu(x, pars)
  
  dnorm(
    y,
    mean = mu,
    sd = grow_sd(mu, pars))
}


# Dormancy entry ---------------------------------------------------------------
dorm_x <- function(x, pars, num_pars = pars$mod_do_index) {
  val <- pars$dorm_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("dorm_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  inv_logit(val)
}


# Flowering probability --------------------------------------------------------
fl_x <- function(x, pars, num_pars = pars$mod_fl_index) {
  val <- pars$fl_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("fl_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  inv_logit(val)
}


# Scape number conditional on flowering ---------------------------------------
fl_n_x <- function(x, pars, num_pars = pars$mod_fl_n_index) {
  val <- pars$fln_b0
  
  for (i in seq_len(num_pars)) {
    param <- paste0("fln_b", i)
    
    if (!is.null(pars[[param]])) {
      val <- val + pars[[param]] * x^i
    }
  }
  
  exp(val)
}


# Recruitment size distribution ------------------------------------------------
re_y_dist <- function(y, pars) {
  norm <- pnorm(
    pars$U, mean = pars$recr_sz, sd = pars$recr_sd) -
    pnorm(
      pars$L, mean = pars$recr_sz, sd = pars$recr_sd)
  
  dnorm(
    y,
    mean = pars$recr_sz,
    sd = pars$recr_sd) / norm
}


# Reactivation size distribution -----------------------------------------------
react_y_dist <- function(y, pars) {
  norm <- pnorm(
    pars$U, mean = pars$react_sz, sd = pars$react_sd) -
    pnorm(
      pars$L, mean = pars$react_sz, sd = pars$react_sd)
  
  dnorm(
    y,
    mean = pars$react_sz,
    sd = pars$react_sd) / norm
}


# F-kernel ---------------------------------------------------------------------
fyx <- function(y, x, pars) {
  fl_x(x, pars) *
    fl_n_x(x, pars) *
    pars$fecu_b0 *
    re_y_dist(y, pars)
}
```

## Kernel
```{r}
kernel <- function(pars) {
  
  # Continuous active-size classes
  n <- pars$mat_siz
  L <- pars$L
  U <- pars$U
  h <- (U - L) / n
  
  b <- L + c(0:n) * h
  y <- 0.5 * (b[1:n] + b[2:(n + 1)])
  
  # Active survival and dormancy
  Smat <- sx(y, pars)
  Dmat <- dorm_x(y, pars)
  
  # Growth matrix
  Gmat <- matrix(0, n, n)
  Gmat[] <- t(outer(y, y, gxy, pars)) * h
  
  # Correct growth eviction
  for (i in seq_len(n)) {
    if (i <= n / 2) {
      Gmat[1, i] <- Gmat[1, i] + 1 - sum(Gmat[, i])
    } else {
      Gmat[n, i] <- Gmat[n, i] + 1 - sum(Gmat[, i])
    }
  }
  
  # Active -> active
  Tmat <- sweep(
    Gmat,
    2,
    Smat * (1 - Dmat),
    "*")
  
  # Active -> dormant
  Dorm_row <- matrix(
    Smat * Dmat,
    nrow = 1)
  
  # Dormant -> active
  React_vec <- react_y_dist(y, pars) * h
  React_vec <- React_vec / sum(React_vec)
  
  React_col <- matrix(
    pars$p_reactivate * React_vec,
    ncol = 1)
  
  # Dormant -> dormant
  Dorm_stasis <- 1 - pars$p_reactivate
  
  # Fertility from active individuals
  Fmat <- outer(
    y, y,
    Vectorize(function(y, x) {
      fyx(y, x, pars)
    })) * h
  
  # Full 201 x 201 kernel
  K <- rbind(
    cbind(Tmat + Fmat, React_col),
    cbind(Dorm_row, Dorm_stasis))
  
  return(list(
    k_yx = K,
    Fmat = Fmat,
    Tmat = Tmat,
    Gmat = Gmat,
    Dorm_row = Dorm_row,
    React_col = React_col,
    Smat = Smat,
    Dmat = Dmat,
    meshpts = y,
    h = h))
}
```

## Mean population growth rate
```{r}
lambda_ipm <- function(pars) {
  Re(eigen(
    kernel(pars)$k_yx,
    only.values = TRUE)$values[1])
}

lam_mean <- lambda_ipm(pars)
lam_mean
```

## Kernel checks
```{r}
K_mean <- kernel(pars)

dim(K_mean$k_yx)

range(K_mean$Smat)
range(K_mean$Dmat)

range(colSums(K_mean$Gmat))
sum(K_mean$React_col)
1 - pars$p_reactivate

range(K_mean$k_yx)
any(!is.finite(K_mean$k_yx))

lam_mean
```

## Stable population structure
```{r}
eig_mean <- eigen(K_mean$k_yx)

stable_dist <- Re(eig_mean$vectors[, 1])

if (sum(stable_dist) < 0) {
  stable_dist <- -stable_dist
}

stable_dist <- stable_dist / sum(stable_dist)

stable_active <- stable_dist[seq_len(pars$mat_siz)]
stable_dormant <- stable_dist[pars$mat_siz + 1]

df_stable <- tibble(
  logsize = K_mean$meshpts,
  size = exp(logsize),
  stable = stable_active,
  stable_density = stable_active / K_mean$h)

df_stable_summary <- tibble(
  state = c("Active", "Dormant"),
  proportion = c(
    sum(stable_active),
    stable_dormant))

df_stable_summary
```

## Stable size distribution
```{r}
fig_stable <- ggplot(
  df_stable,
  aes(x = size, y = stable_density)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    title = "Stable active size distribution",
    subtitle = v_ggp_suffix,
    x = "Rosette diameter (cm)",
    y = "Stable density")

fig_stable
```

## Observed population growth
```{r}
# Count known living individuals. New adults are additionally counted one year
# earlier, matching the simplified recruitment assumption used in this IPM.

df_counts_alive <- df_ind %>%
  filter(state %in% c("active", "dormant", "alive_or_dormant")) %>%
  count(site, pop, qu, year, name = "n_alive")

df_newadult_backfill <- df_ind %>%
  filter(recruit_type == "new_adult") %>%
  transmute(site, pop, qu, year = year - 1) %>%
  count(site, pop, qu, year, name = "n_newadult_backfill")

# Use recruitment rows as the monitored quadrat-year inventory
df_counts_quad <- df %>%
  filter(row_type == "recruitment") %>%
  distinct(site, pop, qu, year) %>%
  left_join(
    df_counts_alive,
    by = c("site", "pop", "qu", "year")) %>%
  left_join(
    df_newadult_backfill,
    by = c("site", "pop", "qu", "year")) %>%
  mutate(
    n_alive = replace_na(n_alive, 0L),
    n_newadult_backfill = replace_na(n_newadult_backfill, 0L),
    n = n_alive + n_newadult_backfill) %>%
  group_by(site, pop, qu) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    year_t1 = lead(year),
    n_t1 = lead(n),
    annual_transition = year_t1 == year + 1) %>%
  ungroup() %>%
  filter(annual_transition)

# The final three transitions cannot fully resolve terminal dormancy
dormancy_cutoff <- 3
last_complete_transition <- max(df_ind$year, na.rm = TRUE) -
  dormancy_cutoff - 1

df_counts_year <- df_counts_quad %>%
  filter(year <= last_complete_transition) %>%
  group_by(year) %>%
  summarise(
    n_quads = n(),
    n_t0 = sum(n),
    n_t1 = sum(n_t1), .groups = "drop") %>%
  mutate(lambda_obs = n_t1 / n_t0)

df_counts_year

# Mean observed population growth
lam_obs_y <- df_counts_year$lambda_obs

lam_obs_mean <- mean(lam_obs_y, na.rm = TRUE)
lam_obs_geo <- exp(mean(log(lam_obs_y), na.rm = TRUE))

lam_obs_mean
lam_obs_geo
```

## Modeled versus observed population growth
```{r}
df_lambda_mean <- tibble(
  estimate = c(
    "IPM asymptotic",
    "Observed arithmetic",
    "Observed geometric"),
  lambda = c(
    lam_mean,
    lam_obs_mean,
    lam_obs_geo))

df_lambda_mean
```
